### 2020-07-23 index.rmd
add some files

--- 
knit: "bookdown::render_book"
title: "Pre-processing high frequency animal tracking data"
author: "Pratik R. Gupte and others"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
# bibliography: [refs.bib]
biblio-style: apalike
link-citations: yes
github-repo: pratikunterwegs/atlas-best-practices
description: "TRES Tidyverse Tutorial"
---

# Outline {-}

An outline here.

### 2020-07-23 ms.Rmd
add some files

---
title: "ms"
output:
  pdf_document: 
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

# Background

Tracking individual movement has become a firmly established component of animal ecology, helped by the adoption of the movement ecology paradigm (Nathan et al. 2008; Holyoak et al. 2008) and the rapid proliferation of animal tracking technologies (Kays et al. 2015; Hussey et al. 2015). The animal tracking revolution has offered unprecedented insights into the causes and consequences of animal space-use, from the mechanisms behind collective behaviour (e.g. Strandburg-Peshkin et al. 2015), to how animal migrations engineer landscape level processes (Geremia et al. 2019).

# Methods

## Filtering point locations

### Basic spatio-temporal filtering

Filtering animal tracking data on existing variables is an easy first step in preparing data for further analysis. 
The majority of modern animal tracking studies consist at least of a time series coupled with spatial coordinates (hereafter referred to as X and Y for convenience), with other covariates for each observation. 
These three variables are already sufficient to eliminate serious positioning errors. 
In a basic first step, data can be filtered to exclude positions outside the bounding box of a known study area. 
This only involves comparing position coordinates with the range of acceptable coordinates, and removing those positions which fall outside of this range. 
Filtering on a bounding box is the fastest spatial filtering method and can quickly remove the most serious positioning errors without requiring knowledge of spatial computation. 
However, it can only select data within a right-angled quadrangular area.

Explicit spatial filtering is a finer method which examines whether the tracking data lie within a polygon, which may be irregular. 
This typically involves converting the data’s position coordinates to a multi-point spatial geometry, and finding the intersection with a polygon or multi-polygon geometry representing a study area. 
Combined with knowledge of species’ biology and the study site, explicit spatial filtering can be used to remove obviously erroneous positions such as those that would place an individual in an area it is physiologically incapable of traversing. 
Both bounding box and explicit spatial filters can be combined with the timestamp of a position to remove locations where time is an important determinant of accessibility. 
For example, terrestrial species may be able to access the bottom of a waterbody only during certain times of day or year, and data whose time and position coordinates do not meet these criteria can be removed.

Temporal filtering can also be applied independently of spatial filtering to exclude positions from certain time periods during tracking. 
These are typically periods when data are expected to be biased, such as where an animal has been fitted with a tracker but has not been released, or from the first day after fitting a tracker to avoid including movement behaviour biased by the stress of the capture. 
Such broad filters must always be applied with caution; they are especially susceptible to excluding data that represent novel space use of which researchers are unaware, such as a new movement mode or changed local conditions which make it possible for species to move across previously inaccessible regions.

### Filtering on metadata

Simple spatial and temporal filters are agnostic to any correlation between the status of a tracking system and the extent of error. 
However, both may be refined by consulting tracking system metadata, i.e., data on the status of the tracking system itself, which can be a good indicator of the reliability of calculated positions. 
For example, at a broad scale the number of active receivers in an ATLAS system can indicate time periods when a system suffers a malfunction and thus collects unreliable data; these ‘down times’ can form the basis of a temporal filter. 
Similarly, data from system tests may reveal regions of the study area in which the system’s localisations are error-prone, and can inform an explicit spatial filter. 
At smaller scales the observation-specific metadata may be used to filter out locations which fail to meet a certain cutoff. 
For example, ATLAS systems calculate animal positions using the times of arrival of a radio transmission from each tag to multiple receiver stations. 
The number of receiver stations involved in each localisation is considered to be an indicator of data quality, and positions localised using fewer than three receivers (the minimum required for an ATLAS localisation) can be excluded.

Further, tracking systems often yield internal measures of positioning error which are based on the dilution of precision. 
ATLAS and other time-of-arrival systems calculate positioning error in two dimensions (MacCurdy et al. 2009), while other systems may yield a three dimensional positioning error. 
Positioning error is an intuitive criterion by which to filter data, and this takes the form of locations with an estimated error above a certain threshold being excluded. 
There are two other ways of dealing with positioning errors, and of these the aggregation of positions over a time interval is discussed below.
The more demanding method is to model the animal’s movement, for example, in a continuous time movement model (Calabrese et al. 2016; Noonan et al. 2019). 
Movement model methods can use the positioning error from the tracking system, in conjunction with calibration data, to estimate the real individual position and movement distances between positions. 
However, movement model methods are computationally intensive, make several assumptions about animal movement, and can be challenging to implement for the beginning practitioner, and are not discussed further.

## Filtering trajectories

### Speed and turning angle filters

Since the phenomenon represented by data from animal tracking studies is movement, it is appropriate and necessary to examine whether individual trajectories obtained by joining positions remaining from previous filtering steps are realistic. 
This is often a subjective assessment based on expert judgement of the observed movement tracks, which presents two main problems: first, subjective assessments are difficult to reproduce, which hampers consistent analyses across tracking periods. 
The second issue is that manual identification and correction of problematic trajectories is prohibitively time consuming when dealing with high volumes of data. 
The solution to both problems is to employ a hybrid approach, in which experts manually examine a subset of movement tracks (rather than positions) and attempt to identify common problems with the trajectories. 
Experts can then conceptualise an algorithm by which problematic sections of a trajectory may be removed, and automate the application of this algorithm to the wider dataset, under the assumption that the subset used to define the algorithm is representative of the full data.

Algorithms for filtering trajectories are different from those for filtering point locations in that they commonly filter on derivative variables, i.e., the instantaneous change in some variable with respect to another variable. 
The two most common derivative variables are speed, the change in position in unit time, and turning angle, the change in heading in unit time.
Speed and turning angle filters should be applied in a stepwise fashion, and it is safest to first apply a simple speed filter that removes positions with instantaneous speeds above a certain threshold. 
This threshold should be informed by the tracked animal’s biology, but it is also important to consider that high frequency tracking typically overestimates the speed between positions due to positioning error (Ranacher et al. 2016; an explanation for animal tracking data is given in Noonan et al. 2019). 
Even assuming data with large positioning errors have been removed, it is advisable to begin with a liberal speed threshold that excludes only the most unlikely of movements.

Speed filters have the drawback of removing at twice as many positions as strictly necessary (the speed duplication problem). 
For example, in a trajectory of three positions (p1, p2, and p3), the outlier p2 has a high instantaneous speed, as does the ‘real’ position p3, whose displacement is calculated relative to the outlier p2; both will be removed by the same speed filter. 
Speed filtering with low thresholds also risks losing valid data on fast transit behaviour of a study species, and this is especially a risk for aerial transit which can be strongly assisted by winds. 
In such cases, trning angle filters can help distinguish between high instantaneous speeds arising from positioning error and those that comprise a fast transit segment. 
Most animals are physiologically constrained to move quickly only in relatively straight lines, i.e., with low turning angles, and large turning angles coupled with high speeds often represent positioning errors in the data, whereas high speed - low angle positions are likely to capture extreme yet real movements.

### Removing trajectory reflections

A more serious challenge to speed filters is the tracking system phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.
In such a ‘reflection’ comprised of the positions t1...tr, p1...pn, and tr+1...tn where positions denoted by t represent real trajectory positions, and those denoted p represent reflected data a great distance from the real trajectory, the positions p1 and tr+1 will have very high instantaneous speeds. 
A speed filter applied to such reflected data would only remove p1 and tr+1, leaving the positions p2...pn, which are erroneous and must be removed. 
Iteratively applying a speed filter to this data while recalculating speeds in each iteration until no positions with extreme speeds remain is an imperfect solution, since it will remove valid positions tr+i for each erroneous position pi; the true scale of the speed duplication problem is thus revealed.

A more heuristic approach to dealing with trajectory reflections involves using the phenomenon’s properties to identify its bounds and removing all positions between these bounds. 
The first step is identifying positions with anomalous (usually extremely high) speeds and turning angles, as in the point outlier case. 
The trajectory can then be split into pre- and post-anomaly segments, and the first position of the post-anomaly segment, which is the inner bound of the reflection, can be set as an anchor point. 
The first position after the anchor point with an instantaneous speed >= the anchor point can be considered to be the upper bound of the reflection, and lacking such a position, the position with the next greatest instantaneous speed can be considered the outer bound of the reflection. 
The positions between these bounds can be identified as a trajectory reflection and removed. 
This method is robust to variable displacement of the reflections from the real positions, and also to displacement in the real position of the individual itself.

### 2020-08-04 01_background.Rmd
add new sections and prelim bookdown

--- 2020-07-23 ms.Rmd
@@
----
-title: "ms"
-output:
-  pdf_document: 
-    latex_engine: xelatex
-editor_options: 
-  chunk_output_type: console
----
-
 # Background
@@
 
-# Methods
+## Install `atlastools`
 
-## Filtering point locations
+```{r install_package}
+# use either devtools or remotes to install
+# install.packages("devtools")
 
-### Basic spatio-temporal filtering
-
-Filtering animal tracking data on existing variables is an easy first step in preparing data for further analysis. 
-The majority of modern animal tracking studies consist at least of a time series coupled with spatial coordinates (hereafter referred to as X and Y for convenience), with other covariates for each observation.
-These three variables are already sufficient to eliminate serious positioning errors. 
-In a basic first step, data can be filtered to exclude positions outside the bounding box of a known study area. 
-This only involves comparing position coordinates with the range of acceptable coordinates, and removing those positions which fall outside of this range. 
-Filtering on a bounding box is the fastest spatial filtering method and can quickly remove the most serious positioning errors without requiring knowledge of spatial computation. 
-However, it can only select data within a right-angled quadrangular area.
-
-Explicit spatial filtering is a finer method which examines whether the tracking data lie within a polygon, which may be irregular. 
-This typically involves converting the data’s position coordinates to a multi-point spatial geometry, and finding the intersection with a polygon or multi-polygon geometry representing a study area. 
-Combined with knowledge of species’ biology and the study site, explicit spatial filtering can be used to remove obviously erroneous positions such as those that would place an individual in an area it is physiologically incapable of traversing.
-Both bounding box and explicit spatial filters can be combined with the timestamp of a position to remove locations where time is an important determinant of accessibility. 
-For example, terrestrial species may be able to access the bottom of a waterbody only during certain times of day or year, and data whose time and position coordinates do not meet these criteria can be removed.
-
-Temporal filtering can also be applied independently of spatial filtering to exclude positions from certain time periods during tracking. 
-These are typically periods when data are expected to be biased, such as where an animal has been fitted with a tracker but has not been released, or from the first day after fitting a tracker to avoid including movement behaviour biased by the stress of the capture.
-Such broad filters must always be applied with caution; they are especially susceptible to excluding data that represent novel space use of which researchers are unaware, such as a new movement mode or changed local conditions which make it possible for species to move across previously inaccessible regions.
-
-### Filtering on metadata
-
-Simple spatial and temporal filters are agnostic to any correlation between the status of a tracking system and the extent of error. 
-However, both may be refined by consulting tracking system metadata, i.e., data on the status of the tracking system itself, which can be a good indicator of the reliability of calculated positions. 
-For example, at a broad scale the number of active receivers in an ATLAS system can indicate time periods when a system suffers a malfunction and thus collects unreliable data; these ‘down times’ can form the basis of a temporal filter.
-Similarly, data from system tests may reveal regions of the study area in which the system’s localisations are error-prone, and can inform an explicit spatial filter. 
-At smaller scales the observation-specific metadata may be used to filter out locations which fail to meet a certain cutoff. 
-For example, ATLAS systems calculate animal positions using the times of arrival of a radio transmission from each tag to multiple receiver stations. 
-The number of receiver stations involved in each localisation is considered to be an indicator of data quality, and positions localised using fewer than three receivers (the minimum required for an ATLAS localisation) can be excluded.
-
-Further, tracking systems often yield internal measures of positioning error which are based on the dilution of precision. 
-ATLAS and other time-of-arrival systems calculate positioning error in two dimensions (MacCurdy et al. 2009), while other systems may yield a three dimensional positioning error. 
-Positioning error is an intuitive criterion by which to filter data, and this takes the form of locations with an estimated error above a certain threshold being excluded. 
-There are two other ways of dealing with positioning errors, and of these the aggregation of positions over a time interval is discussed below.
-The more demanding method is to model the animal’s movement, for example, in a continuous time movement model (Calabrese et al. 2016; Noonan et al. 2019). 
-Movement model methods can use the positioning error from the tracking system, in conjunction with calibration data, to estimate the real individual position and movement distances between positions. 
-However, movement model methods are computationally intensive, make several assumptions about animal movement, and can be challenging to implement for the beginning practitioner, and are not discussed further.
-
-## Filtering trajectories
-
-### Speed and turning angle filters
-
-Since the phenomenon represented by data from animal tracking studies is movement, it is appropriate and necessary to examine whether individual trajectories obtained by joining positions remaining from previous filtering steps are realistic.
-This is often a subjective assessment based on expert judgement of the observed movement tracks, which presents two main problems: first, subjective assessments are difficult to reproduce, which hampers consistent analyses across tracking periods.
-The second issue is that manual identification and correction of problematic trajectories is prohibitively time consuming when dealing with high volumes of data. 
-The solution to both problems is to employ a hybrid approach, in which experts manually examine a subset of movement tracks (rather than positions) and attempt to identify common problems with the trajectories.
-Experts can then conceptualise an algorithm by which problematic sections of a trajectory may be removed, and automate the application of this algorithm to the wider dataset, under the assumption that the subset used to define the algorithm is representative of the full data.
-
-Algorithms for filtering trajectories are different from those for filtering point locations in that they commonly filter on derivative variables, i.e., the instantaneous change in some variable with respect to another variable.
-The two most common derivative variables are speed, the change in position in unit time, and turning angle, the change in heading in unit time.
-Speed and turning angle filters should be applied in a stepwise fashion, and it is safest to first apply a simple speed filter that removes positions with instantaneous speeds above a certain threshold.
-This threshold should be informed by the tracked animal’s biology, but it is also important to consider that high frequency tracking typically overestimates the speed between positions due to positioning error (Ranacher et al.
-2016; an explanation for animal tracking data is given in Noonan et al.
-2019).
-Even assuming data with large positioning errors have been removed, it is advisable to begin with a liberal speed threshold that excludes only the most unlikely of movements.
-
-Speed filters have the drawback of removing at twice as many positions as strictly necessary (the speed duplication problem). 
-For example, in a trajectory of three positions (p1, p2, and p3), the outlier p2 has a high instantaneous speed, as does the ‘real’ position p3, whose displacement is calculated relative to the outlier p2; both will be removed by the same speed filter.
-Speed filtering with low thresholds also risks losing valid data on fast transit behaviour of a study species, and this is especially a risk for aerial transit which can be strongly assisted by winds.
-In such cases, trning angle filters can help distinguish between high instantaneous speeds arising from positioning error and those that comprise a fast transit segment. 
-Most animals are physiologically constrained to move quickly only in relatively straight lines, i.e., with low turning angles, and large turning angles coupled with high speeds often represent positioning errors in the data, whereas high speed - low angle positions are likely to capture extreme yet real movements.
-
-### Removing trajectory reflections
-
-A more serious challenge to speed filters is the tracking system phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.
-In such a ‘reflection’ comprised of the positions t1...tr, p1...pn, and tr+1...tn where positions denoted by t represent real trajectory positions, and those denoted p represent reflected data a great distance from the real trajectory, the positions p1 and tr+1 will have very high instantaneous speeds.
-A speed filter applied to such reflected data would only remove p1 and tr+1, leaving the positions p2...pn, which are erroneous and must be removed. 
-Iteratively applying a speed filter to this data while recalculating speeds in each iteration until no positions with extreme speeds remain is an imperfect solution, since it will remove valid positions tr+i for each erroneous position pi; the true scale of the speed duplication problem is thus revealed.
-
-A more heuristic approach to dealing with trajectory reflections involves using the phenomenon’s properties to identify its bounds and removing all positions between these bounds. 
-The first step is identifying positions with anomalous (usually extremely high) speeds and turning angles, as in the point outlier case. 
-The trajectory can then be split into pre- and post-anomaly segments, and the first position of the post-anomaly segment, which is the inner bound of the reflection, can be set as an anchor point. 
-The first position after the anchor point with an instantaneous speed >= the anchor point can be considered to be the upper bound of the reflection, and lacking such a position, the position with the next greatest instantaneous speed can be considered the outer bound of the reflection.
-The positions between these bounds can be identified as a trajectory reflection and removed. 
-This method is robust to variable displacement of the reflections from the real positions, and also to displacement in the real position of the individual itself.
+# install using devtools
+devtools::install_github("pratikunterwegs/atlastools")
+```

### 2020-08-04 02_filtering_on_spatio_temporal_bounds.Rmd
add new sections and prelim bookdown

---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Cleaning point data

## Filtering within spatio-temporal bounds

Here, we simulate some movement data using the `smoove` `R` package from [@gurarie2017].

```{r prep_libs_02_01}
library(data.table)
library(ggplot2)

library(atlastools)
```

```{r sim_data}
# uses smoove from Gurarie et al. (2017)
library(smoove)
nu <- 2
tau <- 5
dt <- .1 
ucvm1 <- simulateUCVM(nu=nu, tau=tau, T.max = 100, dt = dt)

# add a reflection to x_max + x_max / 10, y_max + y_max / 10
data <- as.data.table(ucvm1$XY)
data[, time := seq_len(nrow(data))]
data <- data[seq_len(1000), ]

fwrite(data, "data/sim_data_sec_02_01.csv")
```

Here we read the simulated data in, and plot it to observe its extent.

```{r read_sim_data, eval=TRUE}
# read in the data
data <- data.table::fread("data/sim_data_sec_02_01.csv")

# process the data to remove points in this range
data_excluded <- atlastools::atl_filter_bounds(data,
                                               x = "x", y = "y",
                                               x_range = c(-15, 0),
                                               y_range = c(-35, -15))
# process to keep points inside
data_included <- atlastools::atl_filter_bounds(data,
                                               x = "x", y = "y",
                                               x_range = c(-15, 0),
                                               y_range = c(-35, -15),
                                               remove_inside = FALSE)

# join the data
data[, type := "raw"]
data_excluded[, type := "bbox excluded"]
data_included[, type := "bbox included"]
data <- rbind(data, data_excluded, data_included)

# arrange type
data$type <- factor(data$type, 
                    levels = c("raw", "bbox excluded", "bbox included"))
```

```{r plot_sim_data, include=FALSE}
# plot the data
ggplot(data)+
  geom_path(aes(x, y),
            size = 0.1,
            col = "blue")+
  geom_rect(xmin = -12, xmax = 0,
            ymin = -35, ymax = -15,
            colour = "red",
            fill = NA)+
  facet_grid(~ type)+
  theme_test()+
  labs(x = "X coordinate",
       y = "Y coordinate")

# save figure
ggsave("figures/fig_filter_bbox.png", dpi = 300, width = 5, height = 2)
```

![Filtering by bounding box.](figures/fig_filter_bbox.png)

Explicit spatial filtering is a finer method which examines whether the tracking data lie within a polygon, which may be irregular. This typically involves converting the data’s position coordinates to a multi-point spatial geometry, and finding the intersection with a polygon or multi-polygon geometry representing a study area. 
Combined with knowledge of species’ biology and the study site, explicit spatial filtering can be used to remove obviously erroneous positions such as those that would place an individual in an area it is physiologically incapable of traversing. The main drawback to explicit spatial filtering is that it is computationally intensive, and it is advisable to apply a coarse bounding box filter first in order to reduce the computational load, and only then filter by an irregular spatial feature.

Both bounding box and explicit spatial filters can be combined with the timestamp of a position to remove locations where time is an important determinant of accessibility. For example, terrestrial species may be able to access the bottom of a waterbody only during certain times of day or year, and data whose time and position coordinates do not meet these criteria can be removed. 
Temporal filtering can also be applied independently of spatial filtering to exclude positions from certain time periods during tracking. These are typically periods when data are expected to be biased, such as where an animal has been fitted with a tracker but has not been released, or from the first day after fitting a tracker to avoid including movement behaviour biased by the stress of the capture. 
When combining spatial and temporal filters, it is better to apply the temporal filter first, since it is less computationally intensive than even a bounding box filter, and much less so than an explicit spatial filter, while reducing the number of positions that must be examined by these subsequent filters. 
In any case, broad filters must always be applied with caution; they are especially susceptible to excluding data that represent novel behaviour of which researchers are unaware, such as a new movement mode or changed local conditions which make it possible for species to move across previously inaccessible regions.

### 2020-08-04 03_filtering_on_covariates.Rmd
add new sections and prelim bookdown

--- 2020-07-23 ms.Rmd
@@
----
-title: "ms"
-output:
-  pdf_document: 
-    latex_engine: xelatex
-editor_options: 
-  chunk_output_type: console
----
-
-# Background
-
-Tracking individual movement has become a firmly established component of animal ecology, helped by the adoption of the movement ecology paradigm (Nathan et al.
-2008; Holyoak et al.
-2008) and the rapid proliferation of animal tracking technologies (Kays et al.
-2015; Hussey et al.
-2015).
-The animal tracking revolution has offered unprecedented insights into the causes and consequences of animal space-use, from the mechanisms behind collective behaviour (e.g.
-Strandburg-Peshkin et al.
-2015), to how animal migrations engineer landscape level processes (Geremia et al.
-2019).
-
-# Methods
-
-## Filtering point locations
-
-### Basic spatio-temporal filtering
-
-Filtering animal tracking data on existing variables is an easy first step in preparing data for further analysis. 
-The majority of modern animal tracking studies consist at least of a time series coupled with spatial coordinates (hereafter referred to as X and Y for convenience), with other covariates for each observation.
-These three variables are already sufficient to eliminate serious positioning errors. 
-In a basic first step, data can be filtered to exclude positions outside the bounding box of a known study area. 
-This only involves comparing position coordinates with the range of acceptable coordinates, and removing those positions which fall outside of this range. 
-Filtering on a bounding box is the fastest spatial filtering method and can quickly remove the most serious positioning errors without requiring knowledge of spatial computation. 
-However, it can only select data within a right-angled quadrangular area.
-
-Explicit spatial filtering is a finer method which examines whether the tracking data lie within a polygon, which may be irregular. 
-This typically involves converting the data’s position coordinates to a multi-point spatial geometry, and finding the intersection with a polygon or multi-polygon geometry representing a study area. 
-Combined with knowledge of species’ biology and the study site, explicit spatial filtering can be used to remove obviously erroneous positions such as those that would place an individual in an area it is physiologically incapable of traversing.
-Both bounding box and explicit spatial filters can be combined with the timestamp of a position to remove locations where time is an important determinant of accessibility. 
-For example, terrestrial species may be able to access the bottom of a waterbody only during certain times of day or year, and data whose time and position coordinates do not meet these criteria can be removed.
-
-Temporal filtering can also be applied independently of spatial filtering to exclude positions from certain time periods during tracking. 
-These are typically periods when data are expected to be biased, such as where an animal has been fitted with a tracker but has not been released, or from the first day after fitting a tracker to avoid including movement behaviour biased by the stress of the capture.
-Such broad filters must always be applied with caution; they are especially susceptible to excluding data that represent novel space use of which researchers are unaware, such as a new movement mode or changed local conditions which make it possible for species to move across previously inaccessible regions.
-
-### Filtering on metadata
+## Filtering on covariates
 
@@
 Further, tracking systems often yield internal measures of positioning error which are based on the dilution of precision.
-ATLAS and other time-of-arrival systems calculate positioning error in two dimensions (MacCurdy et al. 2009), while other systems may yield a three dimensional positioning error. 
+ATLAS and other time-of-arrival systems calculate positioning error in two dimensions (MacCurdy et al.
+2009), while other systems may yield a three dimensional positioning error.
 Positioning error is an intuitive criterion by which to filter data, and this takes the form of locations with an estimated error above a certain threshold being excluded.
 There are two other ways of dealing with positioning errors, and of these the aggregation of positions over a time interval is discussed below.
-The more demanding method is to model the animal’s movement, for example, in a continuous time movement model (Calabrese et al. 2016; Noonan et al. 2019). 
+The more demanding method is to model the animal’s movement, for example, in a continuous time movement model (Calabrese et al.
+2016; Noonan et al.
+2019).
 Movement model methods can use the positioning error from the tracking system, in conjunction with calibration data, to estimate the real individual position and movement distances between positions.
 However, movement model methods are computationally intensive, make several assumptions about animal movement, and can be challenging to implement for the beginning practitioner, and are not discussed further.
-
-## Filtering trajectories
-
-### Speed and turning angle filters
-
-Since the phenomenon represented by data from animal tracking studies is movement, it is appropriate and necessary to examine whether individual trajectories obtained by joining positions remaining from previous filtering steps are realistic.
-This is often a subjective assessment based on expert judgement of the observed movement tracks, which presents two main problems: first, subjective assessments are difficult to reproduce, which hampers consistent analyses across tracking periods.
-The second issue is that manual identification and correction of problematic trajectories is prohibitively time consuming when dealing with high volumes of data. 
-The solution to both problems is to employ a hybrid approach, in which experts manually examine a subset of movement tracks (rather than positions) and attempt to identify common problems with the trajectories.
-Experts can then conceptualise an algorithm by which problematic sections of a trajectory may be removed, and automate the application of this algorithm to the wider dataset, under the assumption that the subset used to define the algorithm is representative of the full data.
-
-Algorithms for filtering trajectories are different from those for filtering point locations in that they commonly filter on derivative variables, i.e., the instantaneous change in some variable with respect to another variable.
-The two most common derivative variables are speed, the change in position in unit time, and turning angle, the change in heading in unit time.
-Speed and turning angle filters should be applied in a stepwise fashion, and it is safest to first apply a simple speed filter that removes positions with instantaneous speeds above a certain threshold.
-This threshold should be informed by the tracked animal’s biology, but it is also important to consider that high frequency tracking typically overestimates the speed between positions due to positioning error (Ranacher et al.
-2016; an explanation for animal tracking data is given in Noonan et al.
-2019).
-Even assuming data with large positioning errors have been removed, it is advisable to begin with a liberal speed threshold that excludes only the most unlikely of movements.
-
-Speed filters have the drawback of removing at twice as many positions as strictly necessary (the speed duplication problem). 
-For example, in a trajectory of three positions (p1, p2, and p3), the outlier p2 has a high instantaneous speed, as does the ‘real’ position p3, whose displacement is calculated relative to the outlier p2; both will be removed by the same speed filter.
-Speed filtering with low thresholds also risks losing valid data on fast transit behaviour of a study species, and this is especially a risk for aerial transit which can be strongly assisted by winds.
-In such cases, trning angle filters can help distinguish between high instantaneous speeds arising from positioning error and those that comprise a fast transit segment. 
-Most animals are physiologically constrained to move quickly only in relatively straight lines, i.e., with low turning angles, and large turning angles coupled with high speeds often represent positioning errors in the data, whereas high speed - low angle positions are likely to capture extreme yet real movements.
-
-### Removing trajectory reflections
-
-A more serious challenge to speed filters is the tracking system phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.
-In such a ‘reflection’ comprised of the positions t1...tr, p1...pn, and tr+1...tn where positions denoted by t represent real trajectory positions, and those denoted p represent reflected data a great distance from the real trajectory, the positions p1 and tr+1 will have very high instantaneous speeds.
-A speed filter applied to such reflected data would only remove p1 and tr+1, leaving the positions p2...pn, which are erroneous and must be removed. 
-Iteratively applying a speed filter to this data while recalculating speeds in each iteration until no positions with extreme speeds remain is an imperfect solution, since it will remove valid positions tr+i for each erroneous position pi; the true scale of the speed duplication problem is thus revealed.
-
-A more heuristic approach to dealing with trajectory reflections involves using the phenomenon’s properties to identify its bounds and removing all positions between these bounds. 
-The first step is identifying positions with anomalous (usually extremely high) speeds and turning angles, as in the point outlier case. 
-The trajectory can then be split into pre- and post-anomaly segments, and the first position of the post-anomaly segment, which is the inner bound of the reflection, can be set as an anchor point. 
-The first position after the anchor point with an instantaneous speed >= the anchor point can be considered to be the upper bound of the reflection, and lacking such a position, the position with the next greatest instantaneous speed can be considered the outer bound of the reflection.
-The positions between these bounds can be identified as a trajectory reflection and removed. 
-This method is robust to variable displacement of the reflections from the real positions, and also to displacement in the real position of the individual itself.

### 2020-08-04 04_filtering_speed_angle.Rmd
add new sections and prelim bookdown

--- 2020-07-23 ms.Rmd
@@
----
-title: "ms"
-output:
-  pdf_document: 
-    latex_engine: xelatex
-editor_options: 
-  chunk_output_type: console
----
+# Cleaning movement tracks
 
-# Background
-
-Tracking individual movement has become a firmly established component of animal ecology, helped by the adoption of the movement ecology paradigm (Nathan et al.
-2008; Holyoak et al.
-2008) and the rapid proliferation of animal tracking technologies (Kays et al.
-2015; Hussey et al.
-2015).
-The animal tracking revolution has offered unprecedented insights into the causes and consequences of animal space-use, from the mechanisms behind collective behaviour (e.g.
-Strandburg-Peshkin et al.
-2015), to how animal migrations engineer landscape level processes (Geremia et al.
-2019).
-
-# Methods
-
-## Filtering point locations
-
-### Basic spatio-temporal filtering
-
-Filtering animal tracking data on existing variables is an easy first step in preparing data for further analysis. 
-The majority of modern animal tracking studies consist at least of a time series coupled with spatial coordinates (hereafter referred to as X and Y for convenience), with other covariates for each observation.
-These three variables are already sufficient to eliminate serious positioning errors. 
-In a basic first step, data can be filtered to exclude positions outside the bounding box of a known study area. 
-This only involves comparing position coordinates with the range of acceptable coordinates, and removing those positions which fall outside of this range. 
-Filtering on a bounding box is the fastest spatial filtering method and can quickly remove the most serious positioning errors without requiring knowledge of spatial computation. 
-However, it can only select data within a right-angled quadrangular area.
-
-Explicit spatial filtering is a finer method which examines whether the tracking data lie within a polygon, which may be irregular. 
-This typically involves converting the data’s position coordinates to a multi-point spatial geometry, and finding the intersection with a polygon or multi-polygon geometry representing a study area. 
-Combined with knowledge of species’ biology and the study site, explicit spatial filtering can be used to remove obviously erroneous positions such as those that would place an individual in an area it is physiologically incapable of traversing.
-Both bounding box and explicit spatial filters can be combined with the timestamp of a position to remove locations where time is an important determinant of accessibility. 
-For example, terrestrial species may be able to access the bottom of a waterbody only during certain times of day or year, and data whose time and position coordinates do not meet these criteria can be removed.
-
-Temporal filtering can also be applied independently of spatial filtering to exclude positions from certain time periods during tracking. 
-These are typically periods when data are expected to be biased, such as where an animal has been fitted with a tracker but has not been released, or from the first day after fitting a tracker to avoid including movement behaviour biased by the stress of the capture.
-Such broad filters must always be applied with caution; they are especially susceptible to excluding data that represent novel space use of which researchers are unaware, such as a new movement mode or changed local conditions which make it possible for species to move across previously inaccessible regions.
-
-### Filtering on metadata
-
-Simple spatial and temporal filters are agnostic to any correlation between the status of a tracking system and the extent of error. 
-However, both may be refined by consulting tracking system metadata, i.e., data on the status of the tracking system itself, which can be a good indicator of the reliability of calculated positions. 
-For example, at a broad scale the number of active receivers in an ATLAS system can indicate time periods when a system suffers a malfunction and thus collects unreliable data; these ‘down times’ can form the basis of a temporal filter.
-Similarly, data from system tests may reveal regions of the study area in which the system’s localisations are error-prone, and can inform an explicit spatial filter. 
-At smaller scales the observation-specific metadata may be used to filter out locations which fail to meet a certain cutoff. 
-For example, ATLAS systems calculate animal positions using the times of arrival of a radio transmission from each tag to multiple receiver stations. 
-The number of receiver stations involved in each localisation is considered to be an indicator of data quality, and positions localised using fewer than three receivers (the minimum required for an ATLAS localisation) can be excluded.
-
-Further, tracking systems often yield internal measures of positioning error which are based on the dilution of precision. 
-ATLAS and other time-of-arrival systems calculate positioning error in two dimensions (MacCurdy et al. 2009), while other systems may yield a three dimensional positioning error. 
-Positioning error is an intuitive criterion by which to filter data, and this takes the form of locations with an estimated error above a certain threshold being excluded. 
-There are two other ways of dealing with positioning errors, and of these the aggregation of positions over a time interval is discussed below.
-The more demanding method is to model the animal’s movement, for example, in a continuous time movement model (Calabrese et al. 2016; Noonan et al. 2019). 
-Movement model methods can use the positioning error from the tracking system, in conjunction with calibration data, to estimate the real individual position and movement distances between positions. 
-However, movement model methods are computationally intensive, make several assumptions about animal movement, and can be challenging to implement for the beginning practitioner, and are not discussed further.
-
-## Filtering trajectories
-
-### Speed and turning angle filters
+## Filtering on speed and turning angle
 
@@
 The solution to both problems is to employ a hybrid approach, in which experts manually examine a subset of movement tracks (rather than positions) and attempt to identify common problems with the trajectories.
-Experts can then conceptualise an algorithm by which problematic sections of a trajectory may be removed, and automate the application of this algorithm to the wider dataset, under the assumption that the subset used to define the algorithm is representative of the full data.
+Experts can then conceptualise an algorithm by which problematic sections of a trajectory may be removed, and automate the application of this algorithm to the wider dataset, under the assumption that the subset used to define the algorithm was representative of the full data.
 
@@
 The two most common derivative variables are speed, the change in position in unit time, and turning angle, the change in heading in unit time.
+Trajectories must always be checked for correct temporal ordering before calculating metrics with time interval as a denominator.
 Speed and turning angle filters should be applied in a stepwise fashion, and it is safest to first apply a simple speed filter that removes positions with instantaneous speeds above a certain threshold.
@@
 Speed filtering with low thresholds also risks losing valid data on fast transit behaviour of a study species, and this is especially a risk for aerial transit which can be strongly assisted by winds.
-In such cases, trning angle filters can help distinguish between high instantaneous speeds arising from positioning error and those that comprise a fast transit segment. 
+In such cases, turning angle filters can help distinguish between high instantaneous speeds arising from positioning error and those that comprise a fast transit segment.
 Most animals are physiologically constrained to move quickly only in relatively straight lines, i.e., with low turning angles, and large turning angles coupled with high speeds often represent positioning errors in the data, whereas high speed - low angle positions are likely to capture extreme yet real movements.
 
-### Removing trajectory reflections
+## Challenges to speed filters
 
-A more serious challenge to speed filters is the tracking system phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.
+A more serious challenge to speed filters is the tracking system phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to a location near the real position.
 In such a ‘reflection’ comprised of the positions t1...tr, p1...pn, and tr+1...tn where positions denoted by t represent real trajectory positions, and those denoted p represent reflected data a great distance from the real trajectory, the positions p1 and tr+1 will have very high instantaneous speeds.
 A speed filter applied to such reflected data would only remove p1 and tr+1, leaving the positions p2...pn, which are erroneous and must be removed.
-Iteratively applying a speed filter to this data while recalculating speeds in each iteration until no positions with extreme speeds remain is an imperfect solution, since it will remove valid positions tr+i for each erroneous position pi; the true scale of the speed duplication problem is thus revealed.
+Iteratively applying a speed filter to this data while recalculating speeds in each iteration until no extreme speeds remain is an imperfect solution, since it will remove valid positions tr+i for each erroneous position pi; the true scale of the speed duplication problem is thus revealed.
+Another approach to identifying reflections and especially to determining where they end is to identify the first position with extreme speeds (p1), and calculating the speed between the preceding position (tr) and all subsequent positions (p1...tn) until the first position with a realistic instantaneous speed relative to tr is located; in this case tr+1.
+The drawback here is that the method is sensitive to the duration of the reflection, since the speed between tr and subsequent (reflected) positions will decrease exponentially to a realistic value over time (as the denominator time increases), and the algorithm will prematurely consider the reflection to have ended.
 
-A more heuristic approach to dealing with trajectory reflections involves using the phenomenon’s properties to identify its bounds and removing all positions between these bounds. 
-The first step is identifying positions with anomalous (usually extremely high) speeds and turning angles, as in the point outlier case. 
+A more heuristic approach to dealing with trajectory reflections involves using the phenomenon’s properties to identify its bounds, and removing all positions between these bounds.
+The first step is identifying anomalous positions with extreme speeds and turning angles, as in the point outlier case.
 The trajectory can then be split into pre- and post-anomaly segments, and the first position of the post-anomaly segment, which is the inner bound of the reflection, can be set as an anchor point.
@@
 The positions between these bounds can be identified as a trajectory reflection and removed.
-This method is robust to variable displacement of the reflections from the real positions, and also to displacement in the real position of the individual itself.
+To prevent the algorithm from searching through many hundreds of thousands of data points, the algorithm can be set to consider a post-anomaly segment of K positions, and to look for the end of the reflection within this bound.
+
+While fast, the drawback with this method is in the simplicity of its assumptions that (1) the first position of the data is correct and not itself a reflection, (2) the reflection ends within K positions, and (3) the individual is not moving towards the reflected positions.
+Violation of any of these assumptions will cause the method to fail catastrophically: (1) if the first position is a reflection, it will be preserved and real positions will be removed, (2) if the reflection does not end within K positions the algorithm will erroneously assign the highest speed within K positions as the end, and (3) the individual’s real positions overlapping with the reflected positions will result in low speeds between erroneous and real positions, leaving the algorithm unable to find the correct reflection end.
+If reflections persist in the data after this procedure, or are so prevalent as to cause the algorithm to return erroneous data, there is a limited number of options: (1) attempt to identify whether reflections occur in a common area across individual tracks, and manually identify and remove positions from within the bounds of these so-called ‘attractor points’, (2) subset the data and discard subsets which have more than N anomalous speed-angle positions, as these may indicate many reflected segments, or (3) identify anomalies and remove positions within some time limit of each anomaly, in the hope that the reflection ends by that time.

### 2020-08-04 05_smoothing_tracks.Rmd
add new sections and prelim bookdown

# Smoothing positioning errors

## The median smooth

While the filters described so far are sufficient to remove serious positioning errors, careful examination of the resulting trajectories may reveal that smaller positioning errors remain – the trajectory appears ‘spiky’ at small scales. These smaller errors are usually more challenging to remove since they have covariates (such as speed and turning angles) which are within the expected range of movement behaviour for the study species, hence their persistence in the data. The high volumes of data generated by high frequency tracking studies can be brought to bear upon this issue, by smoothing the trajectory. A ‘smooth’ works by approximating the true value of an observation based on neighbouring values. A common and easily implemented smoothing technique is nearest neighbour averaging, in which the value of an observation X0 is the average of some K nearest neighbours.

The median smooth (confusingly also called the median filter) is a variant of nearest neighbour averaging which replaces X0 with the median of its neighbours, and this is recommended over simple averaging as it is robust to outliers. A median smooth with an appropriate smoothing window (usually 3 – 5 positions) centred on each observation can be applied to the X and Y coordinates of the trajectory to obtain a smoothed track. The median smooth will return as many smoothed coordinates as there are observations in the data, and does not interpolate between positions when data are missing. However, it is robust to even very large gaps, and it is not necessary to split the data into segments separated by periods of missing observations when applying the filter.

### 2020-08-04 08_reducing_volume.Rmd
add new sections and prelim bookdown

# Reducing data volume

## Thinning animal tracks

Most data at this stage is technically ‘clean’, yet its volume alone may pose challenges for older hardware (such as computers or storage) and software (data analysis functions) if these are not optimised for efficient computation. Examples in the R statistical environment favoured by movement ecologists (Joo et al. 2020) include the default function for importing data from a file read.csv, which is much slower than newer alternatives (such as data.table’s fread; Dowle et al. XXXX). This coupled with the requirement of uniform sampling intervals by many methods in the field (e.g. Hidden Markov Models from Michelot et al. 2016, and Step Selection Functions from Signer et al. 2019) makes evenly reducing data volumes worthwhile. The two approaches here are aggregation and resampling, and both begin with rounding the timestamp of the observations to the nearest multiple of the desired interval. When resampling the data, one position for each unique value of the rounded timestamp is retained, while aggregation involves computing the median of values at the positions sharing the same rounded timestamp. While the median aggregation method is less sensitive to position outliers than resampling, it may result in inappropriate fractional values for some position covariates (e.g. the number of base stations in ATLAS data), while others such as positioning error estimates cannot be aggregated using a simple median alone. It is thus advisable to evaluate whether position covariates will be required unaltered in further steps (e.g. positioning error for movement models), in which case resampling may be more appropriate, or whether it is more important to optimise the position estimate, for which aggregation may be better suited.

It is natural to consider evenly reducing data volume before implementing filters or smooths, since this is computationally more efficient than thinning after filtering-smoothing. While resampling data will preserve position covariates which may be passed to further filters, it may also sample outliers which could then be discarded by filters, resulting in uneven sampling once more. On the other hand, median aggregation may not correctly handle position covariates on which data can be filtered, rendering further filtering steps suspect. Even assuming median aggregation is applied to data without covariates, increasing the sampling interval by an order of magnitude or more (e.g. from 3 s to 30 s in ATLAS data) will result in significant underestimation of instantaneous speed (see Noonan et al. 2019). This could have knock-on effects on the efficacy of speed filters applied after aggregation, since the instantaneous speeds of point outliers or trajectory reflections may now be within the range of speeds realistically expected for the study species, making these artifacts more challenging to remove.

### 2020-08-04 10_synthesising_tracks.Rmd
add new sections and prelim bookdown

# Synthesising movement tracks

## Challenges in synthesising high-frequency trajectories

Having passed through the steps above, animal tracking data are ready to be processed to make biological inferences. Yet these data may still be too large to be handled efficiently by statistical modelling methods without resorting to operations unfamiliar to many biologists, such as parallel processing. Many of these methods attempt to identify correlates of movement behaviour, which can be broadly categorised into correlates of residency, and correlates of movement between residence areas, or more colloquially, why does the animal stay versus why does the animal go. Examples include relating residence time to proxies of resource density, or movement between foraging areas to a proxy of competition.

Why residence patches, why other methods are slow etc. WIP

### 2020-08-04 11_residence_patch.Rmd
add new sections and prelim bookdown

## Residence patches and their construction

We propose a simpler alternative which is to synthesise high frequency trajectories into ‘residence patches’ using an intuitive multi-step, multi-scale segmentation and clustering procedure based on first principles/prior knowledge of the species’ biology. The residence patch is defined as a spatio-temporally proximate sequence of consistent positions where the individual was stationary (as in Oudman et al. 2019). While there are sophisticated ways of identifying stationary points in a track (e.g. recursion analysis; Bracis et al. 2018), the simplest method is to subset data with low speeds. The residence patch method uses only three parameters to determine the appropriate clustering.

First, the distance between consecutive positions is compared against the buffer radius, which is the maximum distance between two temporally consecutive positions p1 and p2 for them to be considered to be part of the same sequence. Sequences of positions meeting this criterion are clustered into ‘proto-patches’. Next, recognising that proto-patches might represent discontinuous sequences of a larger residence patch (e.g. due to missing data), the time and distance between consecutive proto-patches is compared against the temporal independence limit and the spatial independence limit respectively. Both time difference and distance are calculated from the last position of proto-patch q1 to the first position of the next proto-patch q2. If either of these criteria for independence are met, i.e., proto-patch q2 is located at a distance greater than the spatial independence limit from q1, or begins more than temporal independence limit time units after q1 ends, the two are considered independent. Only q1 takes its final form as a residence patch, while q2 and q3 are then assessed for independence; a proto-patch is always merged with the proto-patch before it if they are found to be non-independent.

While this algorithm considers only spatio-temporal differences between proto-patches as criteria for being independent residence patches, it can be easily extended to compare differences in any variable against an appropriate parameter. This comparison may be between variables at the end and beginning of consecutive proto-patches, or it may be between the median values of the proto-patches as a whole. For example, the median residence time (Bracis et al. 2018) of consecutive proto-patches can be very different when individuals switch foraging modes, and this difference can be used to separate them into independent patches despite being spatially and temporally proximate.

Having classified the track into residence patches, it is useful to extract summary data which may be passed on to a statistical analysis. Simple metrics of interest to animal ecologists include the patch duration, the distance travelled within the patch, the displacement within the patch, the tortuosity of movement within the patch (the distance – displacement ratio), the displacement between consecutive patches, as well as the median coordinates of the patch. These may be supplemented by more advanced metrics, such as the total patch area (assuming the previously defined buffer radius around each position), and the patch circularity which is a measure of roundness (e.g. Polsby and Popper 1991).

The latter spatial metrics require the patch positions to be converted into spatial objects and merged to correct for overlapping buffers. These multi-polygon spatial objects representing each patch can be combined into a convenient spatial object representing the individual’s entire trajectory as a string of residence patches with summary statistics as covariates. Further environmental covariates for these patches are easily extracted from raster or vector spatial data.

### 2020-08-04 02_filtering_on_spatio_temporal_bounds.Rmd
add some code and to do

@@
 # plot the data
-ggplot(data)+
+fig_filter_bbox <- ggplot(data)+
   geom_path(aes(x, y),
-            size = 0.1,
             col = "blue")+
@@
   theme_test()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
   labs(x = "X coordinate",
@@
 # save figure
-ggsave("figures/fig_filter_bbox.png", dpi = 300, width = 5, height = 2)
+ggsave("figures/fig_filter_bbox.png", dpi = 300, width = 8, height = 2)
 ```
@@
 
+```{r make_explicit_spatial_filter}
+# read in data again
+# read in the data
+data <- data.table::fread("data/sim_data_sec_02_01.csv")
+
+# make circular buffer around centroid
+centroid <- c(median(data$x), median(data$y))
+
+library(sf)
+centroid <- st_point(centroid)
+buffer <- st_buffer(centroid, dist = 10)
+```
+
 Both bounding box and explicit spatial filters can be combined with the timestamp of a position to remove locations where time is an important determinant of accessibility.
@@
 These are typically periods when data are expected to be biased, such as where an animal has been fitted with a tracker but has not been released, or from the first day after fitting a tracker to avoid including movement behaviour biased by the stress of the capture.
-When combining spatial and temporal filters, it is better to apply the temporal filter first, since it is less computationally intensive than even a bounding box filter, and much less so than an explicit spatial filter, while reducing the number of positions that must be examined by these subsequent filters.
+
+> When combining spatial and temporal filters, it is better to apply the temporal filter first, since it is less computationally intensive than even a bounding box filter, and much less so than an explicit spatial filter, while reducing the number of positions that must be examined by these subsequent filters.
+
+**To do**
+
+- benchmark time filter
+- bmark spatial filter
+- bmark time and space filter (both indep, eg. filter before release)
+- bmark time filter (non-indep, both evald together)
+
 In any case, broad filters must always be applied with caution; they are especially susceptible to excluding data that represent novel behaviour of which researchers are unaware, such as a new movement mode or changed local conditions which make it possible for species to move across previously inaccessible regions.

### 2020-08-05 02_filtering_on_spatio_temporal_bounds.Rmd
some text edits'

--- 2020-08-04 02_filtering_on_spatio_temporal_bounds.Rmd
@@
 These are typically periods when data are expected to be biased, such as where an animal has been fitted with a tracker but has not been released, or from the first day after fitting a tracker to avoid including movement behaviour biased by the stress of the capture.
+
 When combining spatial and temporal filters, it is better to apply the temporal filter first, since it is less computationally intensive than even a bounding box filter, and much less so than an explicit spatial filter, while reducing the number of positions that must be examined by these subsequent filters.

### 2020-08-05 03_filtering_on_covariates.Rmd
some text edits'

@@
 For example, at a broad scale the number of active receivers in an ATLAS system can indicate time periods when a system suffers a malfunction and thus collects unreliable data; these ‘down times’ can form the basis of a temporal filter.
+
 Similarly, data from system tests may reveal regions of the study area in which the system’s localisations are error-prone, and can inform an explicit spatial filter.
@@
 Positioning error is an intuitive criterion by which to filter data, and this takes the form of locations with an estimated error above a certain threshold being excluded.
-There are two other ways of dealing with positioning errors, and of these the aggregation of positions over a time interval is discussed below.
-The more demanding method is to model the animal’s movement, for example, in a continuous time movement model (Calabrese et al.
-2016; Noonan et al.
-2019).
-Movement model methods can use the positioning error from the tracking system, in conjunction with calibration data, to estimate the real individual position and movement distances between positions.
-However, movement model methods are computationally intensive, make several assumptions about animal movement, and can be challenging to implement for the beginning practitioner, and are not discussed further.
+
+In `atlastools` filtering by covariates is easily implemented by passing a string representing a quoted `R` expression; multiple such experessions may be passed at once in a character vector.
+Rows of the data satisfying _all_ the conditions are returned.
+
+```{r show_covariate_filtering}
+# read in some data / or process previous data
+atl_filter_covariates(data = data,
+                      filters = c("NBS > 3",
+                                  "SD < 100",
+                                  "data.table::between(day, 5, 8)"))
+```

### 2020-08-13 01_background.Rmd
remove minor sections

@@
-# Background
+## Installing `atlastools`
 
-Tracking individual movement has become a firmly established component of animal ecology, helped by the adoption of the movement ecology paradigm (Nathan et al.
-2008; Holyoak et al.
-2008) and the rapid proliferation of animal tracking technologies (Kays et al.
-2015; Hussey et al.
-2015).
-The animal tracking revolution has offered unprecedented insights into the causes and consequences of animal space-use, from the mechanisms behind collective behaviour (e.g.
-Strandburg-Peshkin et al.
-2015), to how animal migrations engineer landscape level processes (Geremia et al.
-2019).
+This paper refers extensively to the `R` package `atlastools` (Gupte 2020), which can be installed from an online repository.
 
-## Install `atlastools`
+```{r install_package, eval=FALSE}
+# use either devtools or remotes to install
+install.packages("devtools")
 
-```{r install_package}
-# use either devtools or remotes to install
-# install.packages("devtools")
-
-# install using devtools
+# installation using devtools
 devtools::install_github("pratikunterwegs/atlastools")

### 2020-08-13 02_filtering_on_spatio_temporal_bounds.Rmd
remove minor sections

@@
 # save figure
-ggsave("figures/fig_filter_bbox.png", dpi = 300, width = 8, height = 2)
+ggsave("figures/fig-filter-bbox.png", dpi = 300, width = 8, height = 2)
 ```
@@
 
-Explicit spatial filtering is a finer method which examines whether the tracking data lie within a polygon, which may be irregular.
-This typically involves converting the data’s position coordinates to a multi-point spatial geometry, and finding the intersection with a polygon or multi-polygon geometry representing a study area.
-Combined with knowledge of species’ biology and the study site, explicit spatial filtering can be used to remove obviously erroneous positions such as those that would place an individual in an area it is physiologically incapable of traversing.
-The main drawback to explicit spatial filtering is that it is computationally intensive, and it is advisable to apply a coarse bounding box filter first in order to reduce the computational load, and only then filter by an irregular spatial feature.
+## Explicit spatial filtering
+
+```{r make_non-bbox_polygon}
+# make a polygon buffer around centroid
+library(sf)
+data <- data.table::fread("data/sim_data_sec_02_01.csv")
+centroid <- st_point(c(median(data$x), median(data$y)))
+buffer <- st_buffer(centroid, 5)
+```
+
+```{r spatial_filter_function}
+# an ad-hoc spatial filter function which assumes two coordinate columns
+# from a data table or dataframe
+do_spatial_filter <- function(data,
+                              spatial_object) {
+
+  data_coords <- data[, c("x", "y")]
+  data_coords <- st_sfc(st_multipoint(as.matrix(data_coords, byrow = TRUE)))
+  data_coords <- st_cast(data_coords, "POINT")
+  overlaps <- sf::st_intersects(data_coords, spatial_object)
+
+  # not length but lengths
+  data <- data[lengths(overlaps) > 0, ]
+
+  return(data)
+}
+```
+
+```{r do_filter_spatial}
+# filter the data
+filtered_data <- do_spatial_filter(data = data, 
+                                   spatial_object = buffer)
+
+# assign types and plot
+data$type <- "raw"
+filtered_data$type <- "filtered"
+
+# bind rows
+data <- rbind(data, filtered_data)
+
+# assign levels
+data$type <- factor(data$type,
+                    levels = c("raw", "filtered"))
+```
+
+```{r fig_spatial_filter}
+fig_spatial_filter <- 
+  ggplot(data)+
+  geom_path(aes(x, y),
+            col = "blue")+
+  theme_test()+
+  geom_sf(data = buffer, 
+          fill = "grey", 
+          alpha = 0.2)+
+  facet_grid(~ type)+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  labs(x = "X coordinate",
+       y = "Y coordinate")
+
+# save figure
+ggsave("figures/fig-spatial-filter.png", dpi = 300, width = 5, height = 2)
+```
+
+## Bounding box before spatial filter
+
+### Make a larger dataset
+
+Here we simulate a dataset of 50,000 positions, and save it to file for further use.
+
+```{r sim_data_large, eval=FALSE}
+# uses smoove from Gurarie et al. (2017)
+library(smoove)
+nu <- 2
+tau <- 5
+dt <- .1 
+ucvm1 <- simulateUCVM(nu=nu, tau=tau, T.max = 5e4L, dt = dt)
+
+# add a reflection to x_max + x_max / 10, y_max + y_max / 10
+data <- as.data.table(ucvm1$XY)
+data[, time := seq_len(nrow(data))]
+
+fwrite(data, "data/sim_data_large.csv")
+```
+
+### Resample large data
+
+```{r get_resample_prop}
+# powers of 10
+resample_rate <- c(
+  # 1, 2, 3, 10, 
+                   100)
+
+# get range of coordinates
+data_range <- sapply(data[, c("x", "y")], range)
+
+# get resample parameters
+resample_parameters <- CJ(rate = resample_rate,
+                          bbox_size = resample_rate)
+```
+
+```{r}
+do_resample_benchmark <- function(data,
+                                  resamp_rate, prop_bbox) {
+
+  bbox <- sapply(data[, c("x", "y")], range) / prop_bbox
+  data <- data[floor(sample(nrow(data) / resamp_rate)), ]
+
+  microbenchmark::microbenchmark(
+    atl_filter_bounds(data,
+                      x_range = bbox[, "x"],
+                      y_range = bbox[, "y"],
+                      remove_inside = FALSE),
+    unit = "s"
+  )
+}
+```
+
+### Benchmarking bounding box filters
+
+```{r benchmark_bbox_filter}
+library(microbenchmark)
+
+microbenchmark(
+  atlastools::atl_filter_bounds(data,
+                                x = "x", y = "y",
+                                x_range = c(-15, 0),
+                                y_range = c(-35, -15)),
+  unit = "s"
+)
+```
 

### 2020-08-13 index.rmd
remove minor sections

@@
 site: bookdown::bookdown_site
-output: bookdown::gitbook
+output: bookdown::pdf_book
 documentclass: book

### 2020-08-13 02_filtering_on_spatio_temporal_bounds.Rmd
write filtering methods

--- 2020-08-05 02_filtering_on_spatio_temporal_bounds.Rmd
@@
 
-# Cleaning point data
+This section covers:
 
-## Filtering within spatio-temporal bounds
+1. Simple spatio-temporal filtering of point data to remove gross positioning errors,
 
-Here, we simulate some movement data using the `smoove` `R` package from [@gurarie2017].
+2. Filtering positions on covariates that may be associated with increased positioning error,
+
+3. Filtering movement tracks using the 'non-movement' approach to remove implausible movement segments, and
+
+4. The challenge of reflected positions, and how it can be addressed.
+
+# Filtering data
+
+## Prepare libraries {-}
 
@@
 library(data.table)
-library(ggplot2)
-
 library(atlastools)
@@
 
-```{r sim_data}
-# uses smoove from Gurarie et al. (2017)
-library(smoove)
-nu <- 2
-tau <- 5
-dt <- .1 
-ucvm1 <- simulateUCVM(nu=nu, tau=tau, T.max = 100, dt = dt)
+## Introducing errors {-}
 
-# add a reflection to x_max + x_max / 10, y_max + y_max / 10
-data <- as.data.table(ucvm1$XY)
-data[, time := seq_len(nrow(data))]
-data <- data[seq_len(1000), ]
+Here we read the simulated data in.
+The limits of the coordinates are 0 and 1, so we add a few extraneous coordinates within a plausible distance to demonstrate the basic forms of filtering.
+We also assign a fictional _NBS_ value ranging between 2 and 6 to demonstrate filtering on covariates.
+Finally, we introduce a 'reflection', a phenomenon that results when a TOA system's position likelihood estimator shifts the real position by a large distance for an extended duration.
 
-fwrite(data, "data/sim_data_sec_02_01.csv")
+```{r read_sim_data}
+# read in the data
+data <- fread("data/data_sim.csv")
+
+# add some gross positioning errors
+data[sample(nrow(data), 100), c("x", "y") := lapply(.SD,
+                                                   function(z) {
+                                                     z + rnorm(100, sd = 0.1)
+                                                   }),
+     .SDcols = c("x", "y")]
+
+# add a fictional NBS covariate
+data[, NBS := floor(rnorm(n = nrow(data), mean = 4, sd = 1))]
 ```
 
-Here we read the simulated data in, and plot it to observe its extent.
+## Excluding gross positioning errors
 
-```{r read_sim_data, eval=TRUE}
-# read in the data
-data <- data.table::fread("data/sim_data_sec_02_01.csv")
+First we remove gross positioning errors using a bounding box filter, i.e., knowing or assuming the real bounds of the movement track (in this case, 0 -- 1), we remove all positions outside those bounds.
 
-# process the data to remove points in this range
-data_excluded <- atlastools::atl_filter_bounds(data,
-                                               x = "x", y = "y",
-                                               x_range = c(-15, 0),
-                                               y_range = c(-35, -15))
-# process to keep points inside
-data_included <- atlastools::atl_filter_bounds(data,
-                                               x = "x", y = "y",
-                                               x_range = c(-15, 0),
-                                               y_range = c(-35, -15),
+```{r}
+# remove positions outside a bounding box
+data_inside_bbox <- data.table::copy(data)
+
+# NB: set remove_inside to FALSE
+data_inside_bbox <- atlastools::atl_filter_bounds(data = data,
+                                                  x_range = c(0, 1),
+                                                  y_range = c(0, 1),
                                                   remove_inside = FALSE)
-
-# join the data
-data[, type := "raw"]
-data_excluded[, type := "bbox excluded"]
-data_included[, type := "bbox included"]
-data <- rbind(data, data_excluded, data_included)
-
-# arrange type
-data$type <- factor(data$type, 
-                    levels = c("raw", "bbox excluded", "bbox included"))
 ```
 
-```{r plot_sim_data, include=FALSE}
-# plot the data
-fig_filter_bbox <- ggplot(data)+
-  geom_path(aes(x, y),
-            col = "blue")+
-  geom_rect(xmin = -12, xmax = 0,
-            ymin = -35, ymax = -15,
-            colour = "red",
-            fill = NA)+
-  facet_grid(~ type)+
-  theme_test()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  labs(x = "X coordinate",
-       y = "Y coordinate")
+## Excluding data on covariate values
 
-# save figure
-ggsave("figures/fig_filter_bbox.png", dpi = 300, width = 8, height = 2)
+Some position covariates may be indicators of unreliable data, due to being linked with increased spatial uncertainty.
+The number of base-stations (NBS) used to localise a transmitting tag in ATLAS systems is one such covariate, with a minimum of three base-stations required for a two-dimensional localisation.
+Here, we remove all positions with < 3 base-stations.
+
+```{r}
+# remove positions with few base stations
+data_many_nbs <- copy(data_inside_bbox)
+data_many_nbs <- atl_filter_covariates(data_many_nbs, 
+                                       filters = c("NBS >= 3"))
 ```
 
-![Filtering by bounding box.](figures/fig_filter_bbox.png)
+## The non-movement approach
 
-Explicit spatial filtering is a finer method which examines whether the tracking data lie within a polygon, which may be irregular.
-This typically involves converting the data’s position coordinates to a multi-point spatial geometry, and finding the intersection with a polygon or multi-polygon geometry representing a study area.
-Combined with knowledge of species’ biology and the study site, explicit spatial filtering can be used to remove obviously erroneous positions such as those that would place an individual in an area it is physiologically incapable of traversing.
-The main drawback to explicit spatial filtering is that it is computationally intensive, and it is advisable to apply a coarse bounding box filter first in order to reduce the computational load, and only then filter by an irregular spatial feature.
+Bjorneraas et al.
+(2010) describe intuitive filtering steps for trajectories that are based in the biology of the study species (moose _Alces alces_ in their case).
+Briefly, this 'non-movement method' identifies implausible segments of a trajectory, and removes them.
+Here, we adopt elements of their method to identify and remove 'spikes' in the movement track, i.e., positions which indicate an improbably fast and short excursion away from the real trajectory.
 
-Both bounding box and explicit spatial filters can be combined with the timestamp of a position to remove locations where time is an important determinant of accessibility.
-For example, terrestrial species may be able to access the bottom of a waterbody only during certain times of day or year, and data whose time and position coordinates do not meet these criteria can be removed.
-Temporal filtering can also be applied independently of spatial filtering to exclude positions from certain time periods during tracking.
-These are typically periods when data are expected to be biased, such as where an animal has been fitted with a tracker but has not been released, or from the first day after fitting a tracker to avoid including movement behaviour biased by the stress of the capture.
+We begin by calculating the speed and turning angle along the track.
+For both metrics, the first value is set at `NA`, indicating that we do not know the position of the individual prior to the start of the trajectory.
+The turning angle is calculated using the cosine rule and requires three positions to yield a values; thus it has an `NA` at the very last position as well.
 
-When combining spatial and temporal filters, it is better to apply the temporal filter first, since it is less computationally intensive than even a bounding box filter, and much less so than an explicit spatial filter, while reducing the number of positions that must be examined by these subsequent filters.
-In any case, broad filters must always be applied with caution; they are especially susceptible to excluding data that represent novel behaviour of which researchers are unaware, such as a new movement mode or changed local conditions which make it possible for species to move across previously inaccessible regions.
+```{r}
+# get speed and turning angle
+data_many_nbs[, `:=`(in_speed = atl_get_speed(data_many_nbs),
+                     out_speed = c(atl_get_speed(data_many_nbs)[-1], NA),
+                     angle = atl_turning_angle(data_many_nbs))]
+
+```
+
+Next we idenfify the 90th and 95th percentile of speed and turning angle.
+This is more general than identifying the limits of implausibility for each individual (since there may be inter-individual differences), let alone each species in a large dataset.
+An alternative approach is to define a speed cutoff based on expert knowledge, and this is best suited to well studied species.
+
+```{r}
+# get 90 and 95 percentile of speed and turning angle
+sapply(data_many_nbs[, c("in_speed", "angle")], function(z) {
+  quantile(z, probs = c(0.9, 0.95), na.rm = TRUE)
+})
+```
+
+Finally, we remove positions whose incoming and outgoing speeds are both greater than the 95th speed percentile.
+
+```{r}
+# make a copy of the data
+data_non_move <- copy(data_many_nbs)
+
+# filter the copy
+data_non_move <- atl_filter_covariates(data_non_move,
+                                       filters = c("in_speed < 0.0012",
+                                                   "out_speed < 0.0012",
+                                                   "angle < 40"))
+```
+
+## Reflections: Challenges to the non-movement approach
+
+A curious issue affecting TOA tracking systems is the phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.
+
+These reflections cannot be resolved in the same way as spikes, since they consist of several points with very realistic speeds bookended by a pair with either an extreme incoming or outgoing speed.
+
+A reflection appears thus:
+
+```{r}
+# make a copy of data with spikes removed
+data_with_reflection <- copy(data_non_move)
+
+# add a reflection
+data_with_reflection[500:600, c("x", "y") := runif(101, min = 1, max = 1.1)]
+```
+
+Removing reflections involves using the phenomenon’s properties to first identify its bounds and then removing all positions between these bounds.
+
+The procedure is as follows:
+
+1. Identify positions with anomalous (usually extremely high) speeds and turning angles, as with spikes, and
+
+2. Split the trajectory into pre- and post-anomaly segments, and the first position of the post-anomaly segment, which is the inner bound of the reflection, can be set as an anchor point.
+
+3. The first position after the anchor point with an instantaneous outgoing speed >= the speed cutoff is set as the outer bound of the reflection.
+
+4. The positions between these bounds can be identified as a trajectory reflection and removed, along with one position before and after the bounds for safety's sake.
+
+This method is robust to variable displacement of the reflections from the real positions, and also to displacement in the real position of the individual itself.
+
+```{r}
+# attempt to remove reflections
+data_no_reflections <- copy(data_with_reflection)
+
+data_no_reflections <- atl_remove_reflections(data_with_reflection,
+                                              point_angle_cutoff = 40,
+                                              reflection_speed_cutoff = 0.0012)
+```

### 2020-08-13 01_background.Rmd

@@
+---
+output: html_document
+editor_options: 
+  chunk_output_type: console
+---
+
+# Getting started
+
+This section covers:
+
+1. Installing the `R` package `atlastools`,
+
+2. Simulating some realistic looking movement data using the `R` package `smoove` from [@gurarie2017], and
+
+3. Introducing positioning errors into the simulated movement data.
+
 ## Installing `atlastools`
 
-This paper refers extensively to the `R` package `atlastools` (Gupte 2020), which can be installed from an online repository.
+This paper refers extensively to the `R` package `atlastools` (Gupte 2020), which can be installed from [Github.](https://github.com/pratikunterwegs/atlastools). 
+See the repository for information on the build status.
+The code chunk below shows how to install `atlastools`.
 
@@
 ```
+
+## Simulating movement data
+
+Here, we simulate some movement data using the `smoove` `R` package from [@gurarie2017].
+
+```{r prep_libs_01_02}
+# load smoove and datatable
+library(smoove)
+library(data.table)
+```
+
+```{r sim_smoove_data}
+# define parameters
+nu <- 2
+tau <- 1
+dt <- .01 
+ucvm1 <- simulateUCVM(nu = nu, tau = tau, T.max = 100, dt = dt)
+
+# get the data in data.table format
+data <- as.data.table(ucvm1$XY)
+
+# rescale coordinates
+data[, c("x", "y") := (lapply(.SD, 
+                          scales::rescale)),
+     .SDcols = c("x", "y")]
+
+# add a time column
+data[, time := seq_len(nrow(data))]
+
+# save simulated data
+fwrite(data, "data/data_sim.csv")
+```
+
+Here, we write functions to add normally distributed errors to the position coordinates of the simulated data.
+
+```{r fun_add_error}
+# simple function to introduce normally distributed error
+do_add_error <- function(v, std_dev = 0.5) {
+  pos_error <- stats::rnorm(length(v), sd = std_dev)
+  return(v + pos_error)
+}
+```
+
+Example of introduced error.
+
+```{r}
+# load data
+data <- fread("data/data_sim.csv")
+
+# introduce errors to data
+data[, `:=`(x_error = do_add_error(x, std_dev = 0.01),
+            y_error = do_add_error(y, std_dev = 0.01))]
+
+# plot real and introduced error
+ggplot(d2)+
+
+  geom_path(aes(x_error, y_error),
+            col = "red", size = 0.2)+
+  geom_path(aes(x, y))##+
+  geom_path(data = data, aes(x, y),
+            col = "blue")
+
+atlastools::atl_compare_tracks(d2, data)
+```

### 2020-08-25 01_background.Rmd
update bg with do smoove fun

@@
 ```{r sim_smoove_data}
+do_smoove_data <- function(t_max = 100, dt = dt) {
   # define parameters
@@
   dt <- .01 
-ucvm1 <- simulateUCVM(nu = nu, tau = tau, T.max = 100, dt = dt)
+  ucvm1 <- smoove::simulateUCVM(nu = nu, tau = tau, T.max = 100, dt = dt)
 
   # get the data in data.table format
-data <- as.data.table(ucvm1$XY)
-
+  tmp_data <- data.table::as.data.table(ucvm1$XY)
   # rescale coordinates
-data[, c("x", "y") := (lapply(.SD, 
+  tmp_data[, c("x", "y") := (lapply(.SD, 
                                 scales::rescale)),
@@
   # add a time column
-data[, time := seq_len(nrow(data))]
+  tmp_data[, time := seq_len(nrow(tmp_data))]
+
+  return(tmp_data)
+}
+```
+
+```{r save_data}
+data <- do_smoove_data()
 
@@
 # plot real and introduced error
-ggplot(d2)+
-
-  geom_path(aes(x_error, y_error),
-            col = "red", size = 0.2)+
-  geom_path(aes(x, y))##+
+ggplot()+
   geom_path(data = data, aes(x, y),

### 2020-08-25 02_filtering_data.Rmd
rename and add work on section to compare

--- 2020-08-13 02_filtering_on_spatio_temporal_bounds.Rmd
@@
 ```{r prep_libs_02_01}
+# to handle movement data
 library(data.table)
 library(atlastools)
+
+# to simulate movement
+library(smoove)
+
+# to generate a landscape
+library(raster)
+library(NLMR)
+
+# to plot
+library(ggplot2)
 ```
@@
 data <- fread("data/data_sim.csv")
+```
 
-# add some gross positioning errors
-data[sample(nrow(data), 100), c("x", "y") := lapply(.SD,
+```{r function_add_outliers}
+# function for point outliers
+do_add_outliers <- function(data, p_data = 0.1, std_dev = 0.05) {
+  n_rows <- floor(p_data * nrow(data))
+  tmp_data <- copy(data)
+
+  tmp_data[sample(nrow(data), n_rows),
+            c("x", "y") := lapply(.SD,
                                   function(z) {
-                                                     z + rnorm(100, sd = 0.1)
+                                    z + stats::rnorm(n_rows, sd = std_dev)
                                   }),
             .SDcols = c("x", "y")]
-
-# add a fictional NBS covariate
-data[, NBS := floor(rnorm(n = nrow(data), mean = 4, sd = 1))]
+  return(tmp_data)
+}
 ```
 
-## Excluding gross positioning errors
+We add outliers at random to the data to demonstrate their removal.
+
+```{r add_outlier}
+# add 100 outliers
+data_copy <- do_add_outliers(data, p_data = 0.01, std_dev = 0.05)
+```
+
+## Removing gross positioning errors
 
@@
 
-```{r}
+```{r remove_outside_bbox, eval=FALSE}
 # remove positions outside a bounding box
-data_inside_bbox <- data.table::copy(data)
+data_inside_bbox <- data.table::copy(data_copy)
 
@@
 
-## Excluding data on covariate values
-
-Some position covariates may be indicators of unreliable data, due to being linked with increased spatial uncertainty.
-The number of base-stations (NBS) used to localise a transmitting tag in ATLAS systems is one such covariate, with a minimum of three base-stations required for a two-dimensional localisation.
-Here, we remove all positions with < 3 base-stations.
-
 ```{r}
-# remove positions with few base stations
-data_many_nbs <- copy(data_inside_bbox)
-data_many_nbs <- atl_filter_covariates(data_many_nbs, 
-                                       filters = c("NBS >= 3"))
+ggplot()+
+  geom_path(data = data_copy, aes(x, y), col = "red")+
+  geom_path(data = data_inside_bbox, aes(x, y))
 ```
 
-## The non-movement approach
+## Removing point outliers: The non-movement approach
 
@@
 Briefly, this 'non-movement method' identifies implausible segments of a trajectory, and removes them.
-Here, we adopt elements of their method to identify and remove 'spikes' in the movement track, i.e., positions which indicate an improbably fast and short excursion away from the real trajectory.
+Here, we adopt elements of their method to identify and remove 'spikes' or point outliers in the movement track, i.e., positions which indicate an improbably fast and short excursion away from the real trajectory.
 
@@
 
-```{r}
+```{r example_remove_outliers, eval=FALSE}
 # get speed and turning angle
-data_many_nbs[, `:=`(in_speed = atl_get_speed(data_many_nbs),
-                     out_speed = c(atl_get_speed(data_many_nbs)[-1], NA),
+data_many_nbs[, `:=`(in_speed = atl_get_speed(data_many_nbs,
+                                              type = "in"),
+                     out_speed = atl_get_speed(data_many_nbs,
+                                               type = "out"),
                      angle = atl_turning_angle(data_many_nbs))]
@@
 data_non_move <- atl_filter_covariates(data_non_move,
-                                       filters = c("in_speed < 0.0012",
-                                                   "out_speed < 0.0012",
+                                       filters = c("in_speed < 0.001",
+                                                   "out_speed < 0.001",
                                                    "angle < 40"))
+```
+
+```{r}
+ggplot()+
+  geom_path(data = data_inside_bbox,
+            aes(x, y),
+            colour = "black")+
+  geom_path(data = data_non_move,
+            aes(x, y),
+            colour = "red")
+```
+
+## Does removing point outliers help?
+
+### Making error combinations
+
+Point outliers in a trajectory can be boiled down to two parameters: (1) how many positions have been converted into point outliers or spikes, and (2) how large the spikes are relative to the scale of the movement track.
+
+Here we create combinations of these parameters and create 25 replicates of each combination using the error generator on the imported simulated data.
+
+```{r outlier_error_combinations}
+# define unique values
+p_outlier <- c(0.01, 0.05, 0.1, 0.25)
+ext_outlier <- c(0.01, 0.05, 0.1)
+replicate <- seq_len(1)
+
+# make combinations
+param_c <- CJ(p_outlier, ext_outlier, replicate)
+```
+
+### Procedural landscape generation
+
+Next we procedurally generate (simulate) a landscape from which to extract habitat choice.
+
+```{r make_landscape}
+# use NLMR
+landscape <- NLMR::nlm_randomcluster(100, 100, 
+                                     p = 0.2, 
+                                     ai = c(0.2, 0.3, 0.5),
+                                     rescale = TRUE)
+
+# save landscape as raster
+raster::writeRaster(landscape,
+                    filename = "data/data_landscape.tif",
+                    format = "GTiff", 
+                    overwrite = TRUE)
+```
+
+```{r read_landscape}
+# read landscape
+landscape <- raster("data/data_landscape.tif")
+
+# scale landscape?
+bounds <- c(0, 1, 0, 1)
+bound_scale <- c(1, 2, 5, 10)
+extent(landscape) <- bounds
+```
+
+### Make movement track replicates
+
+```{r}
+sim_data <- mapply(function(p_outlier, ext_outlier) {
+  # make simulated data
+  this_data <- do_smoove_data()
+
+  # extract 'true' movement metrics
+  speed <- atl_get_speed(this_data, type = "in")
+
+  # extract proportion of landscape used
+  land_vals <- extract(landscape,
+                       this_data[, c("x", "y")])
+
+  land_vals <- as.data.table(table(landscape))
+
+  # what to return
+
+  # WORK IN PROGRESS HERE
+
+  error_data <- do_add_outliers(this_data, 
+                                p_data = p_outlier,
+                                std_dev = ext_outlier)
+  return(error_data)
+}, 
+param_c$p_outlier, 
+param_c$ext_outlier, 
+SIMPLIFY = FALSE)
 ```
@@
 
-A curious issue affecting TOA tracking systems is the phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.
+A peculiar issue affecting TOA tracking systems is the phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.
 
@@
 
-A reflection appears thus:
+A reflection looks like this:
 
@@
 # add a reflection
-data_with_reflection[500:600, c("x", "y") := runif(101, min = 1, max = 1.1)]
+data_with_reflection[500:600, `:=`(x = runif(101, 1, 1.01),
+                                   y = runif(101, 1, 1.01))]
 ```
@@
 # attempt to remove reflections
-data_no_reflections <- copy(data_with_reflection)
+data_no_reflection <- copy(data_with_reflection)
 
-data_no_reflections <- atl_remove_reflections(data_with_reflection,
-                                              point_angle_cutoff = 40,
-                                              reflection_speed_cutoff = 0.0012)
+data_no_reflection <- atl_remove_reflections(data_with_reflection,
+                                              point_angle_cutoff = 10,
+                                              reflection_speed_cutoff = 0.0015)
 ```
+
+```{r}
+ggplot()+
+  geom_path(data = data_with_reflection,
+            aes(x, y))+
+  geom_path(data = data_no_reflection,
+            aes(x, y),
+            colour = "red")
+```
+
+## Assessing whether filtering helps

### 2020-08-25 03_smoothing_tracks.Rmd
rename some files and add one empty file

---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Smoothing positioning errors

## Prepare libraries {-}

```{r prep_libs_02_01}
library(data.table)
library(atlastools)
library(ggplot2)
```

## Introducing 'reasonable' psoitioning errors {-}

Here we read the simulated data in. The limits of the coordinates are 0 and 1, so we add a few extraneous coordinates within a plausible distance to demonstrate the basic forms of filtering. We also assign a fictional _NBS_ value ranging between 2 and 6 to demonstrate filtering on covariates. Finally, we introduce a 'reflection', a phenomenon that results when a TOA system's position likelihood estimator shifts the real position by a large distance for an extended duration.

```{r read_sim_data_2}
# read in the data
data <- fread("data/data_sim.csv")

# add small position errors to either X or Y
data_with_errors <- copy(data)
data_with_errors[, `:=`(x = x + rnorm(nrow(data), mean = 0, sd = 0.005),
                        y = y + rnorm(nrow(data), mean = 0, sd = 0.005))]

# plot for sanity
ggplot()+
  geom_point(data = data_with_errors,
             aes(x, y),
             shape = 16,
             size = 0.2)+
  theme_void()
```

## Applying a median smooth

```{r}
# make a copy
data_median_smooth <- copy(data_with_errors)
data_median_smooth <- atl_median_smooth(data_median_smooth,
                                        x = "x", 
                                        y = "y",
                                        time = "time",
                                        moving_window = 5)

# plot for sanity
ggplot()+
  geom_path(data = data,
            aes(x, y),
            colour = "blue",
            size = 2)+
  geom_point(data = data_with_errors,
             aes(x, y),
             shape = 16,
             size = 0.2)+
  geom_path(data = data_median_smooth,
            aes(x, y),
            colour = "red",
            size = 1)+
  theme_void()
```

### 2020-08-25 02_filtering_data.Rmd
add to simulation function

@@
   n_rows <- floor(p_data * nrow(data))
-  tmp_data <- copy(data)
+  tmp_data <- data.table::copy(data)
 
@@
   return(tmp_data)
+}
+
+# functions for extracting and tallying landscape values
+do_count_landcover <- function(land, track) {
+  land_values <- raster::extract(land, track[, c("x", "y")])
+  land_values <- data.table::as.data.table(table(land_values))
+  land_values[, p_land := N / sum(N)]
+  return(land_values)
 }
@@
   # make simulated data
-  this_data <- do_smoove_data()
+  ref_data <- do_smoove_data()
 
   # extract 'true' movement metrics
-  speed <- atl_get_speed(this_data, type = "in")
+  speed <- atl_get_speed(ref_data, type = "in")
+  land_vals <- do_count_landcover(landscape, ref_data)
+  lands_vals[, type := "reference"]
 
-  # extract proportion of landscape used
-  land_vals <- extract(landscape,
-                       this_data[, c("x", "y")])
-
-  land_vals <- as.data.table(table(landscape))
-
-  # what to return
-
-  # WORK IN PROGRESS HERE
-
+  # now add error
   error_data <- do_add_outliers(this_data, 
@@
                                 std_dev = ext_outlier)
-  return(error_data)
+
+  # get movement metrics with error
+  error_speed <- atl_get_speed(error_data, type = "in")
+  error_land_vals <- do_count_landcover(landscape, error_data)
+  error_land_vals[, type := "outliers"]
+
+  # now clean data
+  clean_data <- copy(error_data)
+  clean_data[, `:=`(in_speed = atl_get_speed(clean_data, type = "in"),
+                    out_speed = atl_get_speed(clean_data, type = "out"),
+                    angle = atl_turning_angle(clean_data))]
+  upper_speed_cutoff <- stats::quantile(clean_data$in_speed, probs = 0.9)
+  angle_cutoff <- 40
+  clean_data <- atl_filter_covariates(clean_data,
+                                      filters = c(glue('{c("in_speed", 
+                                                           "out_speed")} < 
+                                                         {upper_speed_cutoff}'),
+                                                  glue('angle < {angle_cutoff}')
+                                                  )
+                                      )
+
+  # get speed and land proportion
+  clean_speed <- atl_get_speed(clean_data, type = "in")
+  clean_land_vals <- do_count_landcover(landscape, clean_data)
+  clean_land_vals[, type := "cleaned"]
+
+  # synthesize data
+  land_props <- rbindlist(land_vals, error_land_vals, clean_land_vals)
+  setnames(land_props, old = "V1", new = "landcover")
+  # count proportion
+  land_props <- dcast(land_props, landcover ~ type, value.var = "p_land")
+
+  return(errorCondition("WORK IN PROGRESS"))
 }, 

### 2020-08-26 02_filtering_data.Rmd
add move metrics

--- 2020-08-25 02_filtering_data.Rmd
@@
   return(tmp_data)
+}
+
+# function for extracting and tallying landscape values
+do_count_landcover <- function(land, track) {
+  land_values <- raster::extract(land, track[, c("x", "y")])
+  land_values <- data.table::as.data.table(table(land_values))
+  land_values[, p_land := N / sum(N)]
+  return(land_values)
 }
@@
 # define unique values
-p_outlier <- c(0.01, 0.05, 0.1, 0.25)
+p_outlier <- c(0.05)
 ext_outlier <- c(0.01, 0.05, 0.1)
-replicate <- seq_len(1)
+land_scale <- c(1, 2, 5, 10)
+replicate <- seq_len(10)
 
 # make combinations
-param_c <- CJ(p_outlier, ext_outlier, replicate)
+param_c <- CJ(p_outlier, ext_outlier, land_scale, replicate)
 ```
@@
 landscape <- raster("data/data_landscape.tif")
-
-# scale landscape?
-bounds <- c(0, 1, 0, 1)
-bound_scale <- c(1, 2, 5, 10)
-extent(landscape) <- bounds
 ```
 
-### Make movement track replicates
+```{r scale_landscape}
+# a function to scale the landscape
+do_scale_landscape <- function(land_, scale_factor) {
+  bounds <- c(0, 1, 0, 1)
+  bounds <- bounds * scale_factor
+  raster::extent(land_) <- bounds
+  return(land_)
+}
+```
+
+### Simulate movement over a landscape
 
 ```{r}
-sim_data <- mapply(function(p_outlier, ext_outlier) {
+sim_data <- mapply(function(p_outlier, ext_outlier, scale) {
+
+  # scale the landscape
+  tmp_land <- do_scale_landscape(land_ = landscape,
+                                 scale_factor = scale)
+
   # make simulated data
-  this_data <- do_smoove_data()
+  ref_data <- do_smoove_data()
 
   # extract 'true' movement metrics
-  speed <- atl_get_speed(this_data, type = "in")
+  speed <- atl_get_speed(ref_data, type = "in")
+  speed
+  land_vals <- do_count_landcover(tmp_land, ref_data)
+  land_vals[, type := "reference"]
 
-  # extract proportion of landscape used
-  land_vals <- extract(landscape,
-                       this_data[, c("x", "y")])
-
-  land_vals <- as.data.table(table(landscape))
-
-  # what to return
-
-  # WORK IN PROGRESS HERE
-
-  error_data <- do_add_outliers(this_data, 
+  # now add error
+  error_data <- do_add_outliers(ref_data, 
                                 p_data = p_outlier,
                                 std_dev = ext_outlier)
-  return(error_data)
+
+  # get movement metrics with error
+  error_speed <- atl_get_speed(error_data, type = "in")
+  error_land_vals <- do_count_landcover(tmp_land, error_data)
+  error_land_vals[, type := "outliers"]
+
+  # now clean data
+  clean_data <- copy(error_data)
+  clean_data[, `:=`(in_speed = atl_get_speed(clean_data, type = "in"),
+                    out_speed = atl_get_speed(clean_data, type = "out"),
+                    angle = atl_turning_angle(clean_data))]
+  upper_speed_cutoff <- stats::quantile(clean_data$in_speed, 
+                                        probs = 0.9,
+                                        na.rm = TRUE)
+  angle_cutoff <- 40
+  clean_data <- atl_filter_covariates(clean_data,
+                                      filters = c(glue('{c("in_speed", 
+                                                           "out_speed")} < 
+                                                         {upper_speed_cutoff}'),
+                                                  glue('angle < {angle_cutoff}')
+                                                  )
+                                      )
+
+  # get speed and land proportion
+  clean_speed <- atl_get_speed(clean_data, type = "in")
+  clean_land_vals <- do_count_landcover(tmp_land, clean_data)
+  clean_land_vals[, type := "cleaned"]
+
+  # synthesize data
+  land_props <- rbindlist(list(land_vals, error_land_vals, clean_land_vals))
+  # count proportion
+  land_props <- dcast(land_props, land_values ~ type, value.var = "p_land")
+
+  land_props[, `:=`(scale = scale,
+                    p_outlier = p_outlier,
+                    ext_outlier = ext_outlier)]
+
+  # make speed results
+  speed_res <- data.table(type = c("reference", "outliers", "cleaned"),
+                          speed = c(speed, error_speed, clean_speed))
+  results <- list(land_props = land_props,
+                  speed_res = speed_res)
+
+  return(land_props)
 }, 
@@
 param_c$ext_outlier,
+param_c$land_scale,
 SIMPLIFY = FALSE)
+```
+
+```{r}
+# get bound list
+sim_data <- rbindlist(sim_data)
+# get differences between outliers and cleaned wrt ref
+#sim_data[, `:=`(d_cleaned = abs(cleaned - reference),
+ #               d_outliers = abs(outliers - reference))]
+
+sim_data_long <- copy(sim_data)
+
+sim_data_long <- melt(sim_data_long, 
+                      id.vars = c("land_values", "ext_outlier", 
+                                  "p_outlier", "scale"))
+
+ggplot(sim_data_long)+
+  geom_boxplot(aes(x = factor(land_values),
+                   y = value,
+                   col = variable))+
+  facet_grid(scale ~ ext_outlier, scales = "free")
 ```

### 2020-08-28 01_background.Rmd
remove functions

@@
 
-```{r sim_smoove_data}
-do_smoove_data <- function(t_max = 100, dt = dt) {
-  # define parameters
-  nu <- 2
-  tau <- 1
-  dt <- .01 
-  ucvm1 <- smoove::simulateUCVM(nu = nu, tau = tau, T.max = 100, dt = dt)
-
-  # get the data in data.table format
-  tmp_data <- data.table::as.data.table(ucvm1$XY)
-  # rescale coordinates
-  tmp_data[, c("x", "y") := (lapply(.SD, 
-                                scales::rescale)),
-       .SDcols = c("x", "y")]
-
-  # add a time column
-  tmp_data[, time := seq_len(nrow(tmp_data))]
-
-  return(tmp_data)
-}
-```
-
 ```{r save_data}
@@
 Here, we write functions to add normally distributed errors to the position coordinates of the simulated data.
-
-```{r fun_add_error}
-# simple function to introduce normally distributed error
-do_add_error <- function(v, std_dev = 0.5) {
-  pos_error <- stats::rnorm(length(v), sd = std_dev)
-  return(v + pos_error)
-}
-```
 

### 2020-08-28 02_filtering_data.Rmd
remove functions

@@
 data <- fread("data/data_sim.csv")
-```
-
-```{r function_add_outliers}
-# function for point outliers
-do_add_outliers <- function(data, p_data = 0.1, std_dev = 0.05) {
-  n_rows <- floor(p_data * nrow(data))
-  tmp_data <- data.table::copy(data)
-
-  tmp_data[sample(nrow(data), n_rows),
-            c("x", "y") := lapply(.SD,
-                                  function(z) {
-                                    z + stats::rnorm(n_rows, sd = std_dev)
-                                  }),
-            .SDcols = c("x", "y")]
-  return(tmp_data)
-}
-
-# function for extracting and tallying landscape values
-do_count_landcover <- function(land, track) {
-  land_values <- raster::extract(land, track[, c("x", "y")])
-  land_values <- data.table::as.data.table(table(land_values))
-  land_values[, p_land := N / sum(N)]
-  return(land_values)
-}
 ```

### 2020-08-28 helper_functions.R
add helper functions

@@ -1,0 +1,48 @@
+
+do_smoove_data <- function(t_max = 100, dt = dt) {
+  # define parameters
+  nu <- 2
+  tau <- 1
+  dt <- .01 
+  ucvm1 <- smoove::simulateUCVM(nu = nu, tau = tau, T.max = 100, dt = dt)
+  
+  # get the data in data.table format
+  tmp_data <- data.table::as.data.table(ucvm1$XY)
+  # rescale coordinates
+  tmp_data[, c("x", "y") := (lapply(.SD, 
+                                scales::rescale)),
+       .SDcols = c("x", "y")]
+  
+  # add a time column
+  tmp_data[, time := seq_len(nrow(tmp_data))]
+
+  return(tmp_data)
+}
+
+# simple function to introduce normally distributed error
+do_add_error <- function(v, std_dev = 0.5) {
+  pos_error <- stats::rnorm(length(v), sd = std_dev)
+  return(v + pos_error)
+}
+
+# function for point outliers
+do_add_outliers <- function(data, p_data = 0.1, std_dev = 0.05) {
+  n_rows <- floor(p_data * nrow(data))
+  tmp_data <- data.table::copy(data)
+  
+  tmp_data[sample(nrow(data), n_rows),
+            c("x", "y") := lapply(.SD,
+                                  function(z) {
+                                    z + stats::rnorm(n_rows, sd = std_dev)
+                                  }),
+            .SDcols = c("x", "y")]
+  return(tmp_data)
+}
+
+# function for extracting and tallying landscape values
+do_count_landcover <- function(land, track) {
+  land_values <- raster::extract(land, track[, c("x", "y")])
+  land_values <- data.table::as.data.table(table(land_values))
+  land_values[, p_land := N / sum(N)]
+  return(land_values)
+}

### 2020-08-28 helper_functions.R
add helper functions to folder

@@ -1,48 +1,0 @@
-
-do_smoove_data <- function(t_max = 100, dt = dt) {
-  # define parameters
-  nu <- 2
-  tau <- 1
-  dt <- .01 
-  ucvm1 <- smoove::simulateUCVM(nu = nu, tau = tau, T.max = 100, dt = dt)
-  
-  # get the data in data.table format
-  tmp_data <- data.table::as.data.table(ucvm1$XY)
-  # rescale coordinates
-  tmp_data[, c("x", "y") := (lapply(.SD, 
-                                scales::rescale)),
-       .SDcols = c("x", "y")]
-  
-  # add a time column
-  tmp_data[, time := seq_len(nrow(tmp_data))]
-
-  return(tmp_data)
-}
-
-# simple function to introduce normally distributed error
-do_add_error <- function(v, std_dev = 0.5) {
-  pos_error <- stats::rnorm(length(v), sd = std_dev)
-  return(v + pos_error)
-}
-
-# function for point outliers
-do_add_outliers <- function(data, p_data = 0.1, std_dev = 0.05) {
-  n_rows <- floor(p_data * nrow(data))
-  tmp_data <- data.table::copy(data)
-  
-  tmp_data[sample(nrow(data), n_rows),
-            c("x", "y") := lapply(.SD,
-                                  function(z) {
-                                    z + stats::rnorm(n_rows, sd = std_dev)
-                                  }),
-            .SDcols = c("x", "y")]
-  return(tmp_data)
-}
-
-# function for extracting and tallying landscape values
-do_count_landcover <- function(land, track) {
-  land_values <- raster::extract(land, track[, c("x", "y")])
-  land_values <- data.table::as.data.table(table(land_values))
-  land_values[, p_land := N / sum(N)]
-  return(land_values)
-}

### 2020-08-28 02_filtering_data.Rmd
add scale landscape
compare speed

@@
 library(NLMR)
+
+# strings
+library(glue)
 
@@
 
-```{r scale_landscape}
-# a function to scale the landscape
-do_scale_landscape <- function(land_, scale_factor) {
-  bounds <- c(0, 1, 0, 1)
-  bounds <- bounds * scale_factor
-  raster::extent(land_) <- bounds
-  return(land_)
-}
-```
-
 ### Simulate movement over a landscape
@@
   speed_res <- data.table(type = c("reference", "outliers", "cleaned"),
-                          speed = c(speed, error_speed, clean_speed))
+                          speed_mean = unlist(lapply(list(speed, 
+                                                          error_speed, 
+                                                          clean_speed),
+                                                     mean, na.rm = TRUE)),
+                          speed_sd = unlist(lapply(list(speed, 
+                                                          error_speed, 
+                                                          clean_speed),
+                                                     sd, na.rm = TRUE)),
+                          scale = scale,
+                          p_outlier = p_outlier,
+                          ext_outlier = ext_outlier
+                          )
   results <- list(land_props = land_props,
@@
 
-  return(land_props)
+  return(results)
 }, 
@@
 
-```{r}
+```{r outlier_removal_land_prop}
 # get bound list
-sim_data <- rbindlist(sim_data)
+land_prop_data <- rbindlist(lapply(sim_data, function(l) { l$land_props }))
 # get differences between outliers and cleaned wrt ref
-#sim_data[, `:=`(d_cleaned = abs(cleaned - reference),
+#land_prop_data[, `:=`(d_cleaned = abs(cleaned - reference),
  #               d_outliers = abs(outliers - reference))]
 
-sim_data_long <- copy(sim_data)
+land_prop_data_long <- copy(land_prop_data)
 
-sim_data_long <- melt(sim_data_long, 
+land_prop_data_long <- melt(land_prop_data_long, 
                             id.vars = c("land_values", "ext_outlier", 
@@
 
-ggplot(sim_data_long)+
-  geom_boxplot(aes(x = factor(land_values),
+ggplot(land_prop_data_long)+
+  # geom_boxplot(aes(x = factor(land_values),
+  #                  y = value,
+  #                  group = interaction(variable, scale, land_values),
+  #                  fill = variable))+
+  geom_line(aes(x = (land_values),
                 y = value,
-                   col = variable))+
-  facet_grid(scale ~ ext_outlier, scales = "free")
+                group = interaction(variable, scale),
+                col = variable,
+                lty = factor(scale)),
+            stat = "summary")+
+  geom_point(aes(x = (land_values),
+                 y = value,
+                 group = interaction(variable, scale),
+                 col = variable,
+                 shape = factor(scale)),
+             size = 2,
+             stat = "summary")+
+  facet_grid( ~ ext_outlier, 
+              scales = "free_y",
+              labeller = label_both)+
+  theme_classic()+
+  theme(strip.background = element_rect(fill = "grey90",
+                                        colour = NA))
+```
+
+```{r outlier_removal_speed}
+# get speed data
+speed_data <- rbindlist(lapply(sim_data, function(l) { l$speed_res }))
+
+# melt to
+speed_data_long <- copy(speed_data)
+
+speed_data_long <- melt(speed_data_long, 
+                            id.vars = c("ext_outlier", "type",
+                                        "p_outlier", "scale"))
+
+ggplot(speed_data_long)+
+  # geom_boxplot(aes(x = factor(land_values),
+  #                  y = value,
+  #                  group = interaction(variable, scale, land_values),
+  #                  fill = variable))+
+  geom_boxplot(aes(x = factor(ext_outlier),
+                   y = value,
+                   col = type))+
+  facet_grid(~ variable, 
+              scales = "free_y",
+              labeller = label_both)+
+  theme_classic()+
+  theme(strip.background = element_rect(fill = "grey90",
+                                        colour = NA))
 ```

### 2020-09-14 02_filtering_data.Rmd
add new filtering code

@@
 # NB: set remove_inside to FALSE
-data_inside_bbox <- atlastools::atl_filter_bounds(data = data,
+data_inside_bbox <- atlastools::atl_filter_bounds(data = data_copy,
                                                   x_range = c(0, 1),
@@
   geom_path(data = data_copy, aes(x, y), col = "red")+
-  geom_path(data = data_inside_bbox, aes(x, y))
+  geom_path(data = data_inside_bbox, aes(x, y))+
+  theme_test()
 ```
@@
 ```{r example_remove_outliers, eval=FALSE}
+# make a copy of the data
+data_many_nbs <- copy(data_inside_bbox)
+
 # get speed and turning angle
@@
 data_non_move <- atl_filter_covariates(data_non_move,
-                                       filters = c("in_speed < 0.001",
-                                                   "out_speed < 0.001",
-                                                   "angle < 40"))
+            filters = c("(in_speed < 0.001 & out_speed < 0.001) | angle < 40"))
 ```
@@
 ```{r}
-sim_data <- mapply(function(p_outlier, ext_outlier, scale) {
+sim_data <- mapply(function(p_outlier, ext_outlier, scale, replicate) {
 
@@
                     p_outlier = p_outlier,
-                    ext_outlier = ext_outlier)]
+                    ext_outlier = ext_outlier,
+                    replicate = replicate)]
 
@@
                           p_outlier = p_outlier,
-                          ext_outlier = ext_outlier
+                          ext_outlier = ext_outlier,
+                          replicate = replicate
                           )
@@
 param_c$land_scale,
+param_c$replicate,
 SIMPLIFY = FALSE)
@@
 
-# melt to
-speed_data_long <- copy(speed_data)
+# cast wide
+speed_data <- speed_data[, !"speed_sd"]
 
-speed_data_long <- melt(speed_data_long, 
-                            id.vars = c("ext_outlier", "type",
-                                        "p_outlier", "scale"))
+# find proportion of speed data
+speed_data <- dcast(speed_data, 
+      scale + p_outlier + ext_outlier + replicate ~ type, 
+      value.var = "speed_mean")
 
-ggplot(speed_data_long)+
-  # geom_boxplot(aes(x = factor(land_values),
-  #                  y = value,
-  #                  group = interaction(variable, scale, land_values),
-  #                  fill = variable))+
-  geom_boxplot(aes(x = factor(ext_outlier),
-                   y = value,
-                   col = type))+
-  facet_grid(~ variable, 
-              scales = "free_y",
-              labeller = label_both)+
-  theme_classic()+
+# get proportion of speeds
+speed_data[, `:=`(d_clean = cleaned * 100 / reference,
+                  d_outlier = cleaned * 100 / reference)]
+
+# melt
+speed_data <- melt(speed_data[, !c("replicate", "cleaned", 
+                                   "reference", "outliers")], 
+                   id.vars = c("scale", "p_outlier", "ext_outlier"))
+
+# summarise
+speed_data <- speed_data[, list(mean = mean(value),
+                                sd = sd(value)),
+                         by = list(scale, p_outlier, ext_outlier, variable)]
+
+ggplot(speed_data)+
+  geom_pointrange(aes(factor(ext_outlier), mean - 100,
+                      ymin = mean - sd - 100,
+                      ymax = mean + sd - 100,
+                      colour = scale,
+                      group = interaction(scale, ext_outlier)),
+                  position = position_dodge(width = 0.2))+
+  geom_line(aes(factor(ext_outlier), mean - 100,
+                      group = scale),
+            position = position_dodge(width = 0.2))+
+  theme_test()+
   theme(strip.background = element_rect(fill = "grey90",

### 2020-09-16 02_filtering_data.Rmd
new figures filter

@@
 library(ggplot2)
+library(patchwork)
+```
+
+```{r}
+# source helper functions
+source("R/helper_functions.R")
 ```
@@
 ```{r add_outlier}
+# make a dopy
+data_copy <- copy(data)
+
+# add normal error
+data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
+                 y = do_add_error(y, std_dev = 0.005))]
 # add 100 outliers
-data_copy <- do_add_outliers(data, p_data = 0.01, std_dev = 0.05)
+data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.05)
 ```
 
-## Removing gross positioning errors
+## Show data with errors
+
+```{r}
+# make list of data to plot
+data_plot <- list(data, data_copy)
+
+# make plot list
+plots <- lapply(data_plot, function(df) {
+  ggplot(data = df)+
+
+    geom_path(aes(x, y),
+              col = "red", 
+              alpha = 1)+
+    geom_point(aes(x, y),
+              col = "steelblue", 
+              alpha = 0.2,
+              pch = 16)+
+    theme_test()+
+    theme(axis.text = element_blank(),
+          axis.title = element_blank())
+})
+
+# wrap plots
+plot_figure <- wrap_plots(plots) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+# save figure
+ggsave(plot_figure, filename = "figures/fig_add_errors.png", dpi = 300,
+       width = 185, height = 90, units = "mm")
+```
+
+## Filtering by spatial bounds
 
@@
 # remove positions outside a bounding box
-data_inside_bbox <- data.table::copy(data_copy)
-
 # NB: set remove_inside to FALSE
 data_inside_bbox <- atlastools::atl_filter_bounds(data = data_copy,
-                                                  x_range = c(0, 1),
-                                                  y_range = c(0, 1),
+                                                  x_range = c(0.25, 0.75),
+                                                  y_range = c(0.25, 0.75),
                                                   remove_inside = FALSE)
+
+# data outside a box
+data_outside_bbox <- atlastools::atl_filter_bounds(data = data_copy,
+                                                  x_range = c(-1, 0.2),
+                                                  y_range = c(-1, 0.2),
+                                                  remove_inside = TRUE)
 ```
@@
 ```{r}
+# data to plot
+data_plot <- list(data_copy, data_inside_bbox, data_outside_bbox)
+
+# make plots
+plots <- lapply(data_plot, function(df) {
   ggplot()+
-  geom_path(data = data_copy, aes(x, y), col = "red")+
-  geom_path(data = data_inside_bbox, aes(x, y))+
-  theme_test()
+
+    geom_path(data = df,
+              aes(x, y),
+              col = "red", 
+              alpha = 1, lwd = 0.2)+
+    geom_point(data = df,
+               aes(x, y),
+              col = "steelblue", 
+              alpha = 0.2,
+              pch = 16)+
+    theme_test()+
+    coord_cartesian(xlim = c(0, 1),
+                    ylim = c(0, 1)) +
+    theme(axis.title = element_blank(),
+          axis.text = element_blank())
+})
+
+# add elements
+plots[[1]] <- plots[[1]] + 
+  annotate(xmin = 0.25, xmax = 0.75, ymin = 0.25, ymax = 0.75,
+              fill = NA, col = "black", lty = 1, geom = "rect")+
+  annotate(x = 0.25, y = 0.8, label = "(b)", 
+           fontface = "bold", geom = "text", size = 4)
+
+# add to plot
+plots[[1]] <- plots[[1]] + 
+  annotate(xmin = -0., xmax = 0.2, ymin = -0.1, ymax = 0.2,
+              fill = NA, col = "black", lty = 1, geom = "rect")+
+  annotate(x = 0, y = 0.25, label = "(c)", 
+           fontface = "bold", geom = "text", size = 4)
+
+# wrap plots
+plot_figure <- wrap_plots(plots) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+ggsave(plot_figure, filename = "figures/fig_filter_bounds.png", 
+       width = 185, height = 65, units = "mm")
 ```
@@
 ```{r example_remove_outliers, eval=FALSE}
-# make a copy of the data
-data_many_nbs <- copy(data_inside_bbox)
-
 # get speed and turning angle
-data_many_nbs[, `:=`(in_speed = atl_get_speed(data_many_nbs,
+data_copy[, `:=`(in_speed = atl_get_speed(data_copy,
                                           type = "in"),
-                     out_speed = atl_get_speed(data_many_nbs,
+                 out_speed = atl_get_speed(data_copy,
                                            type = "out"),
-                     angle = atl_turning_angle(data_many_nbs))]
+                 angle = atl_turning_angle(data_copy))]
 
@@
 # get 90 and 95 percentile of speed and turning angle
-sapply(data_many_nbs[, c("in_speed", "angle")], function(z) {
+sapply(data_copy[, c("in_speed", "angle")], function(z) {
   quantile(z, probs = c(0.9, 0.95), na.rm = TRUE)
@@
 ```{r}
-# make a copy of the data
-data_non_move <- copy(data_many_nbs)
-
-# filter the copy
-data_non_move <- atl_filter_covariates(data_non_move,
-            filters = c("(in_speed < 0.001 & out_speed < 0.001) | angle < 40"))
+# filter the copy by the 95th percentile
+data_filtered <- atl_filter_covariates(data_copy,
+            filters = c("(in_speed < 0.029 & out_speed < 0.029) | angle < 40"))
 ```
@@
 ```{r}
+# data plot
+data_plot <- list(data_copy, data_filtered)
+
+plots <- lapply(data_plot, function(df) {
   ggplot()+
-  geom_path(data = data_inside_bbox,
+     geom_path(data = df,
               aes(x, y),
-            colour = "black")+
-  geom_path(data = data_non_move,
-            aes(x, y),
-            colour = "red")
+              col = "red", 
+              alpha = 1, lwd = 0.2)+
+    geom_point(data = df,
+               aes(x, y, 
+                   col = (in_speed >= 0.029 & out_speed >= 0.029)),
+               shape = ifelse((df$in_speed >= 0.029 & df$out_speed >= 0.029), 
+                              16, 1),
+               show.legend = F)+
+    scale_colour_manual(values = c("steelblue", "red"))+
+    theme_test()+
+    coord_cartesian(xlim = c(0, 1),
+                    ylim = c(0, 1)) +
+    theme(axis.title = element_blank(),
+          axis.text = element_blank())
+})
+
+plot_figure <- wrap_plots(plots) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+plot_figure
+ggsave(plot_figure, filename = "figures/fig_filter_spikes.png", 
+       width = 185, height = 90, units = "mm")
 ```
@@
 ```{r}
-ggplot()+
-  geom_path(data = data_with_reflection,
-            aes(x, y))+
-  geom_path(data = data_no_reflection,
-            aes(x, y),
-            colour = "red")
+# data to plot
+plot_data <- list(data_with_reflection, data_no_reflection)
+
+# get plots
+plots <- lapply(plot_data, function(df) {
+  ggplot(data = df)+
+    geom_path(aes(x, y),
+              col = "red")+
+    theme_test()+
+    theme(axis.text = element_blank(),
+          axis.title = element_blank())
+})
+
+patchwork::wrap_plots(plots)+ 
+  patchwork::plot_annotation(tag_levels = "a",
+                             tag_prefix = "(",
+                             tag_suffix = ")")
 ```

### 2020-09-17 02_filtering_data.Rmd
updates

--- 2020-08-28 02_filtering_data.Rmd
@@
 # read in the data
-data <- fread("data/data_sim.csv")
+data <- fread("data/data_sim.csv")[5000:10000, ]
 ```
@@
 ```{r add_outlier}
+# make a dopy
+data_copy <- copy(data)
+
+# add normal error
+data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
+                 y = do_add_error(y, std_dev = 0.005))]
 # add 100 outliers
-data_copy <- do_add_outliers(data, p_data = 0.01, std_dev = 0.05)
+data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
 ```
 
-## Removing gross positioning errors
+## Show data with errors
+
+```{r}
+# make list of data to plot
+data_plot <- list(data, data_copy)
+
+# make plot list
+plots <- lapply(data_plot, function(df) {
+  ggplot(data = df)+
+
+    geom_path(aes(x, y),
+              col = "red", 
+              alpha = 1, lwd = 0.2)+
+    geom_point(aes(x, y),
+              col = "darkblue", 
+              alpha = 0.1,
+              shape = 16)+
+    # coord_cartesian(xlim = c(0.5, 1),
+    #                 ylim = c(0.5, 1)) +
+    coord_equal()+
+    theme_test()+
+    theme(axis.text = element_blank(),
+          axis.title = element_blank())
+})
+
+# wrap plots
+plot_figure <- wrap_plots(plots) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+plot_figure
+
+# save figure
+ggsave(plot_figure, filename = "figures/fig_add_errors.png",
+       width = 185, height = 90, units = "mm")
+```
+
+## Filtering by spatial bounds
 
@@
 # remove positions outside a bounding box
-data_inside_bbox <- data.table::copy(data_copy)
-
 # NB: set remove_inside to FALSE
-data_inside_bbox <- atlastools::atl_filter_bounds(data = data,
-                                                  x_range = c(0, 1),
-                                                  y_range = c(0, 1),
+data_inside_bbox <- atlastools::atl_filter_bounds(data = data_copy,
+                                                  y_range = c(0.5, 1),
                                                   remove_inside = FALSE)
@@
 ```{r}
+# data to plot
+data_plot <- list(data_copy, data_inside_bbox)
+
+# make plots
+plots <- lapply(data_plot, function(df) {
   ggplot()+
-  geom_path(data = data_copy, aes(x, y), col = "red")+
-  geom_path(data = data_inside_bbox, aes(x, y))
+
+    geom_path(data = df,
+              aes(x, y),
+              col = "red", 
+              alpha = 1, lwd = 0.2)+
+    geom_point(data = df,
+               aes(x, y),
+              col = "darkblue", 
+              alpha = 0.1,
+              pch = 16)+
+    theme_test()+
+    coord_equal() +
+    theme(axis.title = element_blank(),
+          axis.text = element_blank())
+})
+
+# add elements
+plots[[1]] <- plots[[1]] + 
+  annotate(xmin = 0.5, xmax = 1,
+           ymin = 0.5, ymax = 1,
+           fill = NA, col = "black", 
+           lty = 2, geom = "rect")+
+  annotate(x = 0.5, y = 1.1, label = "(b)", 
+           fontface = "bold", geom = "text", size = 4)
+
+# wrap plots
+plot_figure <- wrap_plots(plots) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+plot_figure
+
+ggsave(plot_figure, filename = "figures/fig_filter_bounds.png", 
+       width = 185, height = 90, units = "mm")
 ```
@@
 # get speed and turning angle
-data_many_nbs[, `:=`(in_speed = atl_get_speed(data_many_nbs,
+data_copy[, `:=`(in_speed = atl_get_speed(data_copy,
                                           type = "in"),
-                     out_speed = atl_get_speed(data_many_nbs,
+                 out_speed = atl_get_speed(data_copy,
                                            type = "out"),
-                     angle = atl_turning_angle(data_many_nbs))]
+                 angle = atl_turning_angle(data_copy))]
 
@@
 ```{r}
-# make a copy of the data
-data_non_move <- copy(data_many_nbs)
-
-# filter the copy
-data_non_move <- atl_filter_covariates(data_non_move,
-                                       filters = c("in_speed < 0.001",
-                                                   "out_speed < 0.001",
-                                                   "angle < 40"))
+# filter the copy by the 95th percentile
+data_filtered <- atl_filter_covariates(data_copy,
+            filters = c("(in_speed < 0.03 & out_speed < 0.03) | angle < 40"))
 ```
@@
 ```{r}
+# data plot
+data_plot <- list(data_copy, data_filtered)
+
+plots <- lapply(data_plot, function(df) {
   ggplot()+
-  geom_path(data = data_inside_bbox,
+     geom_path(data = df,
               aes(x, y),
-            colour = "black")+
-  geom_path(data = data_non_move,
-            aes(x, y),
-            colour = "red")
+              col = "red", 
+              alpha = 1, lwd = 0.2)+
+    geom_point(data = df,
+               alpha = 0.1,
+               aes(x, y, 
+                   col = (in_speed >= 0.03 & out_speed >= 0.03)),
+               shape = ifelse((df$in_speed >= 0.03 & df$out_speed >= 0.03), 
+                              16, 1),
+               show.legend = F)+
+    scale_colour_manual(values = c("darkblue", "red"))+
+    theme_test()+
+    coord_equal()+
+    theme(axis.title = element_blank(),
+          axis.text = element_blank())
+})
+
+plot_figure <- wrap_plots(plots) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+plot_figure
+ggsave(plot_figure, filename = "figures/fig_filter_spikes.png", 
+       width = 185, height = 90, units = "mm")
 ```
@@
 ext_outlier <- c(0.01, 0.05, 0.1)
-land_scale <- c(1, 2, 5, 10)
 replicate <- seq_len(10)
@@
 # make combinations
-param_c <- CJ(p_outlier, ext_outlier, land_scale, replicate)
-```
-
-### Procedural landscape generation
-
-Next we procedurally generate (simulate) a landscape from which to extract habitat choice.
-
-```{r make_landscape}
-# use NLMR
-landscape <- NLMR::nlm_randomcluster(100, 100, 
-                                     p = 0.2, 
-                                     ai = c(0.2, 0.3, 0.5),
-                                     rescale = TRUE)
-
-# save landscape as raster
-raster::writeRaster(landscape,
-                    filename = "data/data_landscape.tif",
-                    format = "GTiff", 
-                    overwrite = TRUE)
-```
-
-```{r read_landscape}
-# read landscape
-landscape <- raster("data/data_landscape.tif")
-```
-
-### Simulate movement over a landscape
-
-```{r}
-sim_data <- mapply(function(p_outlier, ext_outlier, scale) {
-
-  # scale the landscape
-  tmp_land <- do_scale_landscape(land_ = landscape,
-                                 scale_factor = scale)
-
-  # make simulated data
-  ref_data <- do_smoove_data()
-
-  # extract 'true' movement metrics
-  speed <- atl_get_speed(ref_data, type = "in")
-  speed
-  land_vals <- do_count_landcover(tmp_land, ref_data)
-  land_vals[, type := "reference"]
-
-  # now add error
-  error_data <- do_add_outliers(ref_data, 
-                                p_data = p_outlier,
-                                std_dev = ext_outlier)
-
-  # get movement metrics with error
-  error_speed <- atl_get_speed(error_data, type = "in")
-  error_land_vals <- do_count_landcover(tmp_land, error_data)
-  error_land_vals[, type := "outliers"]
-
-  # now clean data
-  clean_data <- copy(error_data)
-  clean_data[, `:=`(in_speed = atl_get_speed(clean_data, type = "in"),
-                    out_speed = atl_get_speed(clean_data, type = "out"),
-                    angle = atl_turning_angle(clean_data))]
-  upper_speed_cutoff <- stats::quantile(clean_data$in_speed, 
-                                        probs = 0.9,
-                                        na.rm = TRUE)
-  angle_cutoff <- 40
-  clean_data <- atl_filter_covariates(clean_data,
-                                      filters = c(glue('{c("in_speed", 
-                                                           "out_speed")} < 
-                                                         {upper_speed_cutoff}'),
-                                                  glue('angle < {angle_cutoff}')
-                                                  )
-                                      )
-
-  # get speed and land proportion
-  clean_speed <- atl_get_speed(clean_data, type = "in")
-  clean_land_vals <- do_count_landcover(tmp_land, clean_data)
-  clean_land_vals[, type := "cleaned"]
-
-  # synthesize data
-  land_props <- rbindlist(list(land_vals, error_land_vals, clean_land_vals))
-  # count proportion
-  land_props <- dcast(land_props, land_values ~ type, value.var = "p_land")
-
-  land_props[, `:=`(scale = scale,
-                    p_outlier = p_outlier,
-                    ext_outlier = ext_outlier)]
-
-  # make speed results
-  speed_res <- data.table(type = c("reference", "outliers", "cleaned"),
-                          speed_mean = unlist(lapply(list(speed, 
-                                                          error_speed, 
-                                                          clean_speed),
-                                                     mean, na.rm = TRUE)),
-                          speed_sd = unlist(lapply(list(speed, 
-                                                          error_speed, 
-                                                          clean_speed),
-                                                     sd, na.rm = TRUE)),
-                          scale = scale,
-                          p_outlier = p_outlier,
-                          ext_outlier = ext_outlier
-                          )
-  results <- list(land_props = land_props,
-                  speed_res = speed_res)
-
-  return(results)
-}, 
-param_c$p_outlier, 
-param_c$ext_outlier,
-param_c$land_scale,
-SIMPLIFY = FALSE)
-```
-
-```{r outlier_removal_land_prop}
-# get bound list
-land_prop_data <- rbindlist(lapply(sim_data, function(l) { l$land_props }))
-# get differences between outliers and cleaned wrt ref
-#land_prop_data[, `:=`(d_cleaned = abs(cleaned - reference),
- #               d_outliers = abs(outliers - reference))]
-
-land_prop_data_long <- copy(land_prop_data)
-
-land_prop_data_long <- melt(land_prop_data_long, 
-                            id.vars = c("land_values", "ext_outlier", 
-                                        "p_outlier", "scale"))
-
-ggplot(land_prop_data_long)+
-  # geom_boxplot(aes(x = factor(land_values),
-  #                  y = value,
-  #                  group = interaction(variable, scale, land_values),
-  #                  fill = variable))+
-  geom_line(aes(x = (land_values),
-                y = value,
-                group = interaction(variable, scale),
-                col = variable,
-                lty = factor(scale)),
-            stat = "summary")+
-  geom_point(aes(x = (land_values),
-                 y = value,
-                 group = interaction(variable, scale),
-                 col = variable,
-                 shape = factor(scale)),
-             size = 2,
-             stat = "summary")+
-  facet_grid( ~ ext_outlier, 
-              scales = "free_y",
-              labeller = label_both)+
-  theme_classic()+
-  theme(strip.background = element_rect(fill = "grey90",
-                                        colour = NA))
-```
-
-```{r outlier_removal_speed}
-# get speed data
-speed_data <- rbindlist(lapply(sim_data, function(l) { l$speed_res }))
-
-# melt to
-speed_data_long <- copy(speed_data)
-
-speed_data_long <- melt(speed_data_long, 
-                            id.vars = c("ext_outlier", "type",
-                                        "p_outlier", "scale"))
-
-ggplot(speed_data_long)+
-  # geom_boxplot(aes(x = factor(land_values),
-  #                  y = value,
-  #                  group = interaction(variable, scale, land_values),
-  #                  fill = variable))+
-  geom_boxplot(aes(x = factor(ext_outlier),
-                   y = value,
-                   col = type))+
-  facet_grid(~ variable, 
-              scales = "free_y",
-              labeller = label_both)+
-  theme_classic()+
-  theme(strip.background = element_rect(fill = "grey90",
-                                        colour = NA))
+param_c <- CJ(p_outlier, ext_outlier, replicate)
 ```
@@
 # make a copy of data with spikes removed
-data_with_reflection <- copy(data_non_move)
+data_with_reflection <- copy(data_filtered)
 
@@
 Removing reflections involves using the phenomenon’s properties to first identify its bounds and then removing all positions between these bounds.
+
+In some instances, the estimated location is instantaneously displaced from the real movement track for multiple consecutive timesteps in the sampling process, followed by a return to the real position.
+Removing such prolonged spikes (which we call `reflections') using Bjorneraas and colleagues' approach (2010) would be ineffective --- the bounds of the reflection have either extreme incoming \textit{or} outgoing speeds, and there are no positions with both.
+When encountering such data, users are advised to identify the bounds of the reflection as first step to removing them: the inner bound (where the position estimate first shifts away from the track) will have a high incoming speed ($v_i$), while the outer bound (after which the position estimate returns to the real track) will have a high outgoing speed ($v_o$).
+We suggest locating the first reflected point $p_1$ by comparing the incoming speeds of the track positions ($v_i$) against the speed cutoff $S$.
+An optional turning angle cutoff $A$ may be used if the track includes plausible fast transit segments.
+The first position to satisfy this comparison $v_i > S$ AND $\theta > A$ can be considered the inner bound of the reflection and set as an `anchor point'.
+The position after the anchor point with the highest outgoing speed $v_o$ (such that $v_o > S$) can be considered to be the outer bound of the reflection.
+The positions between these bounds can be identified as a trajectory reflection and removed.
+For a system-specific reasons and the pre-processing steps thus far, a reflection may not end before the last position in the movement track.
+In such cases, the entirety of the track following the first anchor point should be removed, as it is better to exclude moderate chunks of data than to make inferences from unsound data.
+Removing reflections is implemented in a convenient form in \texttt{atlastools} in the \texttt{{atl\_remove\_reflections}} function.
 
@@
 # attempt to remove reflections
-data_no_reflection <- copy(data_with_reflection)
-
 data_no_reflection <- atl_remove_reflections(data_with_reflection,
                                               point_angle_cutoff = 10,
-                                              reflection_speed_cutoff = 0.0015)
+                                              reflection_speed_cutoff = 0.03)
 ```
@@
 ```{r}
+# data to plot
+plot_data <- list(data_with_reflection, data_no_reflection)
+
+# get plots
+plots <- lapply(plot_data, function(df) {
   ggplot()+
-  geom_path(data = data_with_reflection,
-            aes(x, y))+
-  geom_path(data = data_no_reflection,
+    geom_path(data = df,
               aes(x, y),
-            colour = "red")
+              col = "red", 
+              alpha = 1, lwd = 0.2)+
+    geom_point(data = df,
+               aes(x, y),
+               alpha = 0.1,
+               colour = "darkblue",
+               shape = 16,
+               show.legend = F)+
+    scale_colour_manual(values = c("darkblue", "red"))+
+    theme_test()+
+    coord_equal()+
+    theme(axis.title = element_blank(),
+          axis.text = element_blank())
+})
+
+# wrap figures
+plot_figure <- wrap_plots(plots) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+plot_figure
+
+ggsave(plot_figure, filename = "figures/fig_filter_reflection.png", 
+       width = 185, height = 90, units = "mm")
 ```
 
-## Assessing whether filtering helps
+## Export data
+
+```{r}
+fwrite(data_no_reflection, file = "data/data_no_reflection.csv")
+```

### 2020-09-18 03_smoothing_tracks.Rmd

@@
 
-# Smoothing positioning errors
+# Smoothing tracks and aggregating dat
 
@@
 
-## Introducing 'reasonable' psoitioning errors {-}
+## Reading in data {-}
 
@@
 # read in the data
-data <- fread("data/data_sim.csv")
-
-# add small position errors to either X or Y
-data_with_errors <- copy(data)
-data_with_errors[, `:=`(x = x + rnorm(nrow(data), mean = 0, sd = 0.005),
-                        y = y + rnorm(nrow(data), mean = 0, sd = 0.005))]
-
-# plot for sanity
-ggplot()+
-  geom_point(data = data_with_errors,
-             aes(x, y),
-             shape = 16,
-             size = 0.2)+
-  theme_void()
+data <- fread("data/data_sim.csv")[5000:10000, ]
+# data with small scale errors but no reflections or outliers
+data_errors <- fread("data/data_no_reflection.csv")
 ```
@@
 ```{r}
-# make a copy
-data_median_smooth <- copy(data_with_errors)
-data_median_smooth <- atl_median_smooth(data_median_smooth,
+# smooth the data
+list_of_smooths <- lapply(c(3, 5, 11, 21), function(z) {
+  data_copy <- copy(data_errors)
+  atl_median_smooth(data = data_copy,
                     x = "x", 
@@
                     time = "time",
-                                        moving_window = 5)
+                    moving_window = z)
+})
+```
 
-# plot for sanity
-ggplot()+
-  geom_path(data = data,
-            aes(x, y),
-            colour = "blue",
-            size = 2)+
-  geom_point(data = data_with_errors,
-             aes(x, y),
-             shape = 16,
-             size = 0.2)+
-  geom_path(data = data_median_smooth,
-            aes(x, y),
-            colour = "red",
-            size = 1)+
-  theme_void()
+```{r}
+# prepare data to plot
+# make list of data to plot
+data_plot <- append(list(data, data_errors), list_of_smooths)
+
+# make plot list
+plots <- lapply(data_plot, function(df) {
+  ggplot(data = df)+
+
+    geom_point(aes(x, y),
+              col = "grey", 
+              alpha = 0.1,
+              shape = 16)+
+    geom_path(aes(x, y),
+              col = "red", 
+              alpha = 0.5)+
+    coord_equal(xlim = c(0.5, 1),
+                ylim = c(0.5, 1)) +
+    theme_test()+
+    theme(axis.text = element_blank(),
+          axis.title = element_blank())
+})
+
+# wrap plots
+plot_figure <- wrap_plots(plots) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+plot_figure
+
+# save figure
+ggsave(plot_figure, filename = "figures/fig_median_smooth.png",
+       width = 185, height = 185 * 2 / 3, units = "mm")
 ```
+
+## Aggregating data
+
+```{r}
+# choose the data
+data_agg <- copy(data)
+
+list_of_agg <- lapply(c(10, 30, 120), function(z) {
+  atl_thin_data(data = data_agg,
+                interval = z,
+                method = "aggregate")
+})
+```
+
+```{r}
+# plot
+data_plot <- append(list(data[, SD := 0]), list_of_agg)
+
+# make plot list
+plots <- lapply(data_plot, function(df) {
+  ggplot(data = df)+
+
+    geom_point(aes(x, y, size = SD), 
+              alpha = 0.5,
+              shape = 16, show.legend = F)+
+    geom_path(aes(x, y),
+              col = "red", 
+              alpha = 1, lwd = 0.2)+
+    coord_equal()+
+    theme_test()+
+    theme(axis.text = element_blank(),
+          axis.title = element_blank())
+})
+
+# wrap plots
+plot_figure <- wrap_plots(plots, ncol = 4) +
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+plot_figure
+
+# save figure
+ggsave(plot_figure, filename = "figures/fig_aggregate_data.png",
+       width = 185, height = 90, units = "mm")
+```

### 2020-09-18 11_residence_patch.Rmd

@@
 Further environmental covariates for these patches are easily extracted from raster or vector spatial data.
+
+```{r}
+# read in the data
+data <- fread("data/data_sim.csv")[5000:10000, ]
+data[, id := "a"]
+
+# make residence patch
+patch <- atl_res_patch(data, buffer_radius = 0.01, 
+                       lim_spat_indep = 0.05, lim_time_indep = 5)
+
+patch_sf <- atl_patch_summary(patch_data = patch, which_data = "summary", buffer_radius = 0.01)
+```

### 2020-09-18 03_smoothing_tracks.Rmd
new figures

--- 2020-08-25 03_smoothing_tracks.Rmd
@@
 # read in the data
-data <- fread("data/data_sim.csv")
-
-# add small position errors to either X or Y
-data_with_errors <- copy(data)
-data_with_errors[, `:=`(x = x + rnorm(nrow(data), mean = 0, sd = 0.005),
-                        y = y + rnorm(nrow(data), mean = 0, sd = 0.005))]
-
-# plot for sanity
-ggplot()+
-  geom_point(data = data_with_errors,
-             aes(x, y),
-             shape = 16,
-             size = 0.2)+
-  theme_void()
+data <- fread("data/data_sim.csv")[5000:10000, ]
+data[, window_size := NA]
+# data with small scale errors but no reflections or outliers
+data_errors <- fread("data/data_no_reflection.csv")
+data_errors[, window_size := 0]
 ```
@@
 ```{r}
-# make a copy
-data_median_smooth <- copy(data_with_errors)
-data_median_smooth <- atl_median_smooth(data_median_smooth,
+# smooth the data
+list_of_smooths <- lapply(c(3, 5, 11, 21), function(z) {
+  data_copy <- copy(data_errors)
+  data_copy <- atl_median_smooth(data = data_copy,
                     x = "x", 
@@
                     time = "time",
-                                        moving_window = 5)
+                    moving_window = z)
+  data_copy[, window_size := z]
+})
+```
 
-# plot for sanity
-ggplot()+
-  geom_path(data = data,
-            aes(x, y),
-            colour = "blue",
-            size = 2)+
-  geom_point(data = data_with_errors,
-             aes(x, y),
-             shape = 16,
-             size = 0.2)+
-  geom_path(data = data_median_smooth,
-            aes(x, y),
-            colour = "red",
-            size = 1)+
-  theme_void()
+```{r}
+# prepare data to plot
+# make list of data to plot
+data_plot <- append(list(data, data_errors), list_of_smooths)
+
+# add offset
+data_plot <- mapply(FUN = function(df, offset) {
+  df[, x := x + offset]
+}, data_plot, seq(from = 0, to = 1, length.out = length(data_plot)),
+SIMPLIFY = FALSE)
+
+# bind rows
+data_plot <- rbindlist(data_plot, fill = TRUE)
+
+plot_figure <- ggplot()+
+  # geom_point(data = data_plot,
+  #            aes(x, y),
+  #            col = "darkgrey", 
+  #            alpha = 1,
+  #            shape = 1)+
+  geom_path(data = data_plot,
+            aes(x, y, 
+                col = window_size,
+                group = window_size),
+            alpha = 1,
+            show.legend = FALSE)+
+  annotate(x = 0.6 + seq(0, 1, length.out = 6),
+           y = 1.05,
+           label = glue::glue('({letters[seq(6)]})'),
+           geom = "text",
+           fontface = "bold")+
+  scale_colour_viridis_c(option = "D",
+                         na.value = "red")+
+  coord_equal() +
+  theme_test()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())
+
+plot_figure
+
+# save figure
+ggsave(plot_figure, filename = "figures/fig_median_smooth.png",
+       width = 185, height = 185 * 2 / 3, units = "mm")
 ```
+
+## Aggregating data
+
+```{r}
+# choose the data
+data_agg <- copy(data)
+
+list_of_agg <- lapply(c(10, 30, 120), function(z) {
+  data_agg <- atl_thin_data(data = data_agg,
+                            interval = z,
+                            method = "aggregate")
+  data_agg[, interval := z]
+
+})
+```
+
+```{r}
+# plot
+data_plot <- append(list(data[, `:=`(SD = 0, interval = NA)]), list_of_agg)
+
+# add offset
+data_plot <- mapply(FUN = function(df, offset) {
+  df[, x := x + offset]
+}, data_plot, seq(from = 0, to = 1, length.out = length(data_plot)),
+SIMPLIFY = FALSE)
+
+data_plot <- rbindlist(data_plot, fill = TRUE)
+
+plot_figure <- ggplot(data = data_plot)+
+  geom_point(aes(x, y, size = SD),
+             colour = "darkblue",
+             alpha = 0.5,
+             shape = 16, show.legend = F)+
+  geom_path(aes(x, y,
+                group = interval,
+                col = interval),
+            col = "red",
+            show.legend = FALSE)+
+  annotate(x = 0.6 + seq(0, 1, length.out = 4),
+           y = 1.05,
+           label = glue::glue('({letters[seq(4)]})'),
+           geom = "text",
+           fontface = "bold")+
+  coord_equal()+
+  theme_test()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())
+
+plot_figure
+
+# save figure
+ggsave(plot_figure, filename = "figures/fig_aggregate_data.png",
+       width = 185, height = 90, units = "mm")
+```

### 2020-09-22 11_residence_patch.Rmd
add res patch code

--- 2020-08-04 11_residence_patch.Rmd
@@
+---
+output: html_document
+editor_options: 
+  chunk_output_type: console
+---
 ## Residence patches and their construction
@@
 Further environmental covariates for these patches are easily extracted from raster or vector spatial data.
+
+## Prepare libraries {-}
+
+```{r prep_libs_02_01}
+library(data.table)
+library(atlastools)
+library(ggplot2)
+```
+
+## Read data and classify
+
+```{r}
+# read in the data
+data <- fread("data/data_sim.csv")[5000:10000, ]
+data[, id := "a"]
+
+# make residence patch
+patch <- atl_res_patch(data, 
+                       buffer_radius = 0.0001, 
+                       lim_spat_indep = 0.02, 
+                       lim_time_indep = 3)
+# get spatial representation
+patch_sf <- atl_patch_summary(patch_data = patch, 
+                              which_data = "spatial", 
+                              buffer_radius = 0.01)
+
+# get summary data
+patch_summary <- atl_patch_summary(patch_data = patch,
+                                   which_data = "summary")
+```
+
+## Plot classified residence patches
+
+```{r}
+plot_patches <- ggplot()+
+  geom_sf(data = patch_sf,
+          colour = "NA",
+          aes(fill = duration),
+          show.legend = FALSE)+
+  geom_point(data = data,
+             aes(x, y),
+             size = 0.1,
+             alpha = 0.1)+
+  geom_point(data = patch_summary,
+             aes(x_median + 0.5, y_median),
+             size = 5,
+             shape = 21,
+             fill = "grey")+
+  annotate(geom = "text",
+           x = patch_summary$x_median + 0.5, 
+           y = patch_summary$y_median + 0.02,
+           label = patch_summary$patch,
+           col = "red")+
+  annotate(geom = "text",
+           x = patch_summary$x_median, 
+           y = patch_summary$y_median + 0.02,
+           label = patch_summary$patch,
+           col = "red")+
+  annotate(geom = "text",
+           x = c(0.5, 1.1),
+           y = 1,
+           label = sprintf("(%s)", c("a", "b")),
+           fontface = "bold")+
+  geom_path(data = patch_summary,
+            aes(x_median + 0.5, y_median),
+            colour = "darkblue",
+            arrow = arrow(angle = 10, 
+                          type = "closed"))+
+  scale_fill_viridis_c()+
+  theme_void()
+
+plot_patches
+
+# save figure
+ggsave(plot_patches, 
+       filename = "figures/fig_residence_patch.png",
+       width = 185, height = 185 * 2 / 3, units = "mm")
+```

### 2020-09-29 11_residence_patch.Rmd
add res patches on calibration data

@@
 ---
-## Residence patches and their construction
 
-We propose a simpler alternative which is to synthesise high frequency trajectories into ‘residence patches’ using an intuitive multi-step, multi-scale segmentation and clustering procedure based on first principles/prior knowledge of the species’ biology.
-The residence patch is defined as a spatio-temporally proximate sequence of consistent positions where the individual was stationary (as in Oudman et al.
-2019).
-While there are sophisticated ways of identifying stationary points in a track (e.g.
-recursion analysis; Bracis et al.
-2018), the simplest method is to subset data with low speeds.
-The residence patch method uses only three parameters to determine the appropriate clustering.
-
-First, the distance between consecutive positions is compared against the buffer radius, which is the maximum distance between two temporally consecutive positions p1 and p2 for them to be considered to be part of the same sequence.
-Sequences of positions meeting this criterion are clustered into ‘proto-patches’.
-Next, recognising that proto-patches might represent discontinuous sequences of a larger residence patch (e.g.
-due to missing data), the time and distance between consecutive proto-patches is compared against the temporal independence limit and the spatial independence limit respectively.
-Both time difference and distance are calculated from the last position of proto-patch q1 to the first position of the next proto-patch q2.
-If either of these criteria for independence are met, i.e., proto-patch q2 is located at a distance greater than the spatial independence limit from q1, or begins more than temporal independence limit time units after q1 ends, the two are considered independent.
-Only q1 takes its final form as a residence patch, while q2 and q3 are then assessed for independence; a proto-patch is always merged with the proto-patch before it if they are found to be non-independent.
-
-While this algorithm considers only spatio-temporal differences between proto-patches as criteria for being independent residence patches, it can be easily extended to compare differences in any variable against an appropriate parameter.
-This comparison may be between variables at the end and beginning of consecutive proto-patches, or it may be between the median values of the proto-patches as a whole.
-For example, the median residence time (Bracis et al.
-2018) of consecutive proto-patches can be very different when individuals switch foraging modes, and this difference can be used to separate them into independent patches despite being spatially and temporally proximate.
-
-Having classified the track into residence patches, it is useful to extract summary data which may be passed on to a statistical analysis.
-Simple metrics of interest to animal ecologists include the patch duration, the distance travelled within the patch, the displacement within the patch, the tortuosity of movement within the patch (the distance – displacement ratio), the displacement between consecutive patches, as well as the median coordinates of the patch.
-These may be supplemented by more advanced metrics, such as the total patch area (assuming the previously defined buffer radius around each position), and the patch circularity which is a measure of roundness (e.g.
-Polsby and Popper 1991).
-
-The latter spatial metrics require the patch positions to be converted into spatial objects and merged to correct for overlapping buffers.
-These multi-polygon spatial objects representing each patch can be combined into a convenient spatial object representing the individual’s entire trajectory as a string of residence patches with summary statistics as covariates.
-Further environmental covariates for these patches are easily extracted from raster or vector spatial data.
+# Residence patches and their construction
 
@@
 
-## Read data and classify
+## An example with simulated data
+
+### Read data and classify
 
@@
 
-## Plot classified residence patches
+### Plot classified residence patches
 

### 2020-09-29 05_calibration_data.Rmd

---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Processing calibration data

## Prepare libraries

```{r}
# load libs
library(data.table)
library(atlastools)
library(ggplot2)
```

## Preliminary visualisation

```{r}
# read and plot example data
data <- fread("data/atlas1060_allTrials_annotated.csv")

# plot data
fig_data_raw <- ggplot(data)+
  geom_path(aes(x, y),
             col = "grey", alpha = 0.5, size = 0.3)+
  geom_point(aes(x, y),
             col = "steelblue", alpha = 0.2, size = 2)+
  ggthemes::theme_map()+
  coord_sf(crs = 32631)

# save figure
ggsave(fig_data_raw, filename = "figures/fig_calibration_raw.png",
       width = 185 / 25)
```

## Filter by bounding box

Save a raw copy.

```{r}
data_raw <- copy(data)
```

Filter by a bounding box.

```{r}
# remove inside must be set to falses
data <- atl_filter_bounds(data = data, 
                          x = "x", y = "y", 
                          x_range = c(645000, max(data$x)), 
                          remove_inside = FALSE)
```

Plot the result.

```{r}
# plot data
fig_data_bbox <-
  ggplot()+
  geom_path(data = data_raw,
            aes(x, y),
             col = "grey", alpha = 0.5, size = 0.3)+
  geom_point(data = data_raw[!data, on = c("x", "y")],
             aes(x, y),
             col = "red", alpha = 1, size = 2)+
  geom_point(data = data,
             aes(x, y),
             col = "steelblue", alpha = 0.2, size = 2)+
  geom_vline(xintercept = 645000,
             col = "grey", lty = 2)+
  ggthemes::theme_map()+
  coord_sf(crs = 32631)

# save result
ggsave(fig_data_bbox, filename = "figures/fig_calib_bbox.png",
       width = 185 / 25)
```

### Remove low base station positions

```{r}
# save a copy
data_raw <- copy(data)

# remove NBS <= 3
data <- atl_filter_covariates(data = data,
                              filters = "NBS > 3")
```

```{r}
fig_low_nbs <- ggplot()+
  geom_path(data = data,
            aes(x, y),
             col = "grey", alpha = 0.5, size = 0.3)+
  geom_point(data = data_raw[!data, on = c("x", "y")],
            aes(x, y),
            shape = 4,
            col = "red", alpha = 0.5, size = 0.3)+
  geom_point(data = data,
             aes(x, y, col = NBS),
             alpha = 1, size = 0.2)+
  scico::scale_colour_scico(palette = "davos",
                         labels = scales::comma, end = 0.9)+
  ggthemes::theme_map()+
  coord_sf(crs = 32631)+
  labs(colour = "NBS")

# save figure
ggsave(fig_low_nbs, filename = "figures/fig_low_nbs.png",
       width = 185 / 25, height = 185 / 25)
```

## Filter trajectories

### Handle time

Time in ATLAS tracking is counted in milliseconds and is represented by a 64-bit integer (type `long`), which is not natively supported in R; it will instead be converted to a `numeric`, or `double`.

This is not what is intended, but it works. The `bit64` package can help handle 64-bit integers if you want to keep to intended type.

A further issue is that 64-bit integers (whether represented as `bit64` or `double`) do not yield meaninful results when you try to convert them to a date-time object, such as of the class `POSIXct`.

This is because `as.POSIXct` fails when trying to work with 64-bit integers (it cannot interpret this type), and returns a date many thousands of years in the future (approx. 52,000 CE) if the time column is converted to `numeric`.

There are two possible solutions. The parsimonious one is to convert the 64-bit number to a 32-bit short integer (dividing by 1000), or to use the `nanotime` package.

The conversion method loses an imperceptible amount of precision. The `nanotime` requires installing another package. The first method is shown here.

In the spirit of not destroying data, we create a second lower-case column called `time`.

```{r}
# divide by 1000, convert to integer, then convert to POSIXct
data[, time := as.POSIXct(as.integer(TIME / 1000), 
                          origin = "1970-01-01")]
```

### Add speed and turning angle

```{r}
# add incoming and outgoing speed
data[, `:=` (speed_in = atl_get_speed(data, 
                                      x = "x", 
                                      y = "y", 
                                      time = "time"),
             speed_out = atl_get_speed(data, type = "out"))]

# add turning angle
data[, angle := atl_turning_angle(data = data)]
```

### Get 95th percentile of speed and angle

```{r}
# use sapply
speed_angle_thresholds <- 
  sapply(data[, list(speed_in, speed_out, angle)], 
       quantile, probs = 0.9, na.rm = T)
```

### Plot to see speeds

```{r}
# plot filtered data
fig_speed_outliers <-
  ggplot()+
  geom_path(data = data,
            aes(x, y),
             col = "grey", alpha = 0.5, size = 0.3)+
  geom_point(data = data,
             aes(x, y, col = speed_in),
             size = 0.2)+
  scico::scale_colour_scico(palette = "davos", 
                         direction = 1, 
                         limits = c(NA, speed_angle_thresholds[1]),
                         na.value = "red",
                         labels = scales::comma,
                         end = 0.9,
                         trans = "sqrt")+
  ggthemes::theme_map()+
  theme(legend.position = c(0, 0.8))+
  coord_sf(crs = 32631)+
  labs(colour = "speed (m/s)")

# save
ggsave(fig_speed_outliers, filename = "figures/fig_speed_outlier.png",
       width = 185 / 25, height = 185 / 25)
```

### Filter on speed

Here we use a speed threshold of 20 m/s, 4 m/s lower than the 95th percentile.

```{r}
# make a copy
data_raw <- copy(data)

# remove speed outliers
data <- atl_filter_covariates(data = data,
            filters = c("(speed_in < 20 & speed_out < 20)"))

# recalculate speed and angle
data[, `:=` (speed_in = atl_get_speed(data, 
                                      x = "x", 
                                      y = "y", 
                                      time = "time"),
             speed_out = atl_get_speed(data, type = "out"))]

# add turning angle
data[, angle := atl_turning_angle(data = data)]
```

## Smoothing the trajectory

```{r}
# apply a 5 point median smooth, first make a copy
data_raw <- copy(data)

# now apply the smooth
data <- atl_median_smooth(data = data,
                          x = "x", y = "y", time = "time",
                          moving_window = 5)
```

```{r}
# make figure
fig_speed_outliers <-
  ggplot()+
  geom_path(data = data_raw,
            aes(x, y, col = "red"), 
            alpha = 0.5, size = 4)+
  geom_path(data = data,
             aes(x, y, col = "black"))+
  scico::scale_colour_scico_d(palette = "roma",
                         direction = -1, begin = 0.1,
                         labels = c("smoothed", "raw"))+
  ggthemes::theme_map()+
  theme(legend.position = c(0, 0.8))+
  coord_sf(crs = 32631)+
  labs(colour = "track")

# save figure
ggsave(fig_speed_outliers, 
       filename = "figures/fig_calib_median_smooth.png",
       width = 185 / 25, height = 185 / 25)
```

## Making residence patches

### Prepare data

An indicator of individual residence at or near a position can be useful when attempting to identify residence patches. Positions can be filtered on a metric such as residence time (see Bracis et al. 2018).

In this dataset, residence positions are marked in the `tID` column as beginning with `WP`. These can be extracted and converted into residence patches.

```{r}
# make a data copy
data_raw <- copy(data)

# assign residence positions
data[, residence := stringi::stri_detect(tID, regex = "WP")]

# make a copy of only residence points
data_res <- data[residence == T, ]

# use only residence or walking points
data <- data[residence == T | tID == "walking", ]
```

### Run residence patch method

```{r}
# on known residence points
patch_res_known <- atl_res_patch(data_res, 
                                buffer_radius = 10,
                                lim_spat_indep = 100,
                                lim_time_indep = 30,
                                min_fixes = 3)

# on data where the residence is unknown
patch_res_unknown <- atl_res_patch(data, 
                                buffer_radius = 10,
                                lim_spat_indep = 100,
                                lim_time_indep = 30,
                                min_fixes = 3)
```

### Get spatial and summary objects

```{r}
# for the known and unkniwn patches
patch_sf_data <- lapply(list(patch_res_known, patch_res_unknown),
                        atl_patch_summary, which_data = "spatial")

patch_summary_data <- lapply(list(patch_res_known, patch_res_unknown),
                        atl_patch_summary)
```

### Plot the outcome

```{r}
# all patch figures
fig_patches <- mapply(function(raw_data, patch_sf, patch_summary) {
  ggplot()+
    geom_point(data = raw_data,
               aes(x, y, col = residence),
               size = 0.1)+
    geom_sf(data = patch_sf,
            fill = "yellow", col = "black",
            alpha = 0.5)+
    geom_segment(data = patch_summary,
              aes(x = shift(x_end, type = "lead"), 
                  y = shift(y_end, type = "lead"),
                  xend = shift(x_start), yend = shift(y_start)),
              col = "blue", size = 1)+
    scale_colour_manual(values = c("red", "blue"))+
    ggthemes::theme_map()
}, list(data_res, data), patch_sf_data, patch_summary_data,
SIMPLIFY = FALSE)

# patch with residence points and walking patches
fig_basic_residence <-
  ggplot()+
  geom_point(data = data,
             aes(x, y,
                 col = residence),
            size = ifelse(data$residence, 4, 0.2),
            shape = ifelse(data$residence, 2, 1),
            show.legend = FALSE)+
  scico::scale_color_scico_d(palette = "berlin",
                             direction = -1,
                             begin = 0.2, end = 0.8)+
  ggthemes::theme_map()

# figure of residence patches known
fig_patch_known <- 
  ggplot()+
  geom_sf(data = patch_sf_data[[1]],
          fill = "steelblue", col = NA)+
  geom_path(data = patch_summary_data[[1]],
            aes(x_median, y_median),
            col = "red", size = 0.2)+
  ggthemes::theme_map()

# figure of unknown patches
fig_patch_unknown <- 
  ggplot()+
  geom_sf(data = patch_sf_data[[2]],
          fill = "steelblue", col = NA)+
  geom_path(data = patch_summary_data[[2]],
            aes(x_median, y_median),
            col = "red", size = 0.2)+
    ggthemes::theme_map()

# arrange plots
fig_res_patch <- 
  wrap_plots(list(fig_basic_residence, fig_patch_known, fig_patch_unknown),
           design = "AAAA\nAAAA\nBBCC\nBBCC")+
  plot_annotation(tag_levels = "a", 
                  tag_prefix = "(", tag_suffix = ")")

# save figure
ggsave(fig_res_patch, filename = "figures/fig_calib_respatch.png",
       height = 185/25, width = 185/25)
```

### 2020-09-30 05_calibration_data.Rmd
add some figures

@@
 data <- atl_filter_covariates(data = data,
-            filters = c("(speed_in < 20 & speed_out < 20)"))
+            filters = c("(speed_in < 15.7 & speed_out < 15.7)"))
 
@@
 
-# use only residence or walking points
-data <- data[residence == T | tID == "walking", ]
+# get a simple inverse of speed as residence time classification
+data_slow <- copy(data_raw)
+data_slow <- data_slow[speed_in < 1, ]
 ```
@@
 # on data where the residence is unknown
-patch_res_unknown <- atl_res_patch(data, 
+patch_res_unknown <- atl_res_patch(data_slow, 
                                 buffer_radius = 10,
@@
                                 lim_time_indep = 30,
-                                min_fixes = 3)
+                                min_fixes = 10)
 ```
@@
 ```{r}
-# all patch figures
-fig_patches <- mapply(function(raw_data, patch_sf, patch_summary) {
-  ggplot()+
-    geom_point(data = raw_data,
-               aes(x, y, col = residence),
-               size = 0.1)+
-    geom_sf(data = patch_sf,
-            fill = "yellow", col = "black",
-            alpha = 0.5)+
-    geom_segment(data = patch_summary,
-              aes(x = shift(x_end, type = "lead"), 
-                  y = shift(y_end, type = "lead"),
-                  xend = shift(x_start), yend = shift(y_start)),
-              col = "blue", size = 1)+
-    scale_colour_manual(values = c("red", "blue"))+
-    ggthemes::theme_map()
-}, list(data_res, data), patch_sf_data, patch_summary_data,
-SIMPLIFY = FALSE)
-
 # patch with residence points and walking patches
@@
             aes(x_median, y_median),
-            col = "red", size = 0.2)+
+            col = "red")+
   ggthemes::theme_map()
@@
   ggplot()+
+  geom_path(data = patch_summary_data[[1]],
+            aes(x_median, y_median),
+            col = "black", size = 0.2, lty = 2)+
   geom_sf(data = patch_sf_data[[2]],
           fill = "steelblue", col = NA)+
-  geom_path(data = patch_summary_data[[2]],
-            aes(x_median, y_median),
-            col = "red", size = 0.2)+
+  geom_segment(data = patch_summary_data[[2]],
+               aes(x = x_end,
+                   xend = shift(x_start, type = "lead"),
+                   y = y_end,
+                   yend = shift(y_start, type = "lead")),
+               col = "red"
+  )+
   ggthemes::theme_map()
@@
 ggsave(fig_res_patch, filename = "figures/fig_calib_respatch.png",
-       height = 185/25, width = 185/25)
+       height = 200/25, width = 185/25)
 ```
+
+## Compare patch metrics
+
+### Compare known patches
+
+```{r}
+# get known patch summary
+patch_summary_real <- data_res[, list(nfixes_real = .N,
+                                      x_median = round(median(x), digits = -2),
+                                      y_median = round(median(y), digits = -2),
+                                      duration_real = as.numeric(
+                                        max(time) - min(time))), 
+                               by = "tID"]
+
+# round median coordinate for inferred patches
+patch_summary_inferred <- 
+        patch_summary_data[[1]][, 
+                                c("x_median", "y_median", 
+                                  "nfixes", "duration")
+                                ][, `:=`(x_median = round(x_median, digits = -2),
+                                         y_median = round(y_median, digits = -2))]
+
+# join with respatch summary
+patch_summary_compare <- 
+  merge(patch_summary_real,
+        patch_summary_inferred, 
+        on = c("x_median", "y_median"),
+        all.x = TRUE, all.y = TRUE)
+```
+
+### Plot known and unknown patches
+
+```{r}
+# make figure comparing different methods
+ggplot()+
+  geom_point(data = patch_summary_compare,
+             aes(nfixes_real, 
+                 nfixes),
+             show.legend = F,
+            size = 4,
+            shape = 1)+
+  geom_smooth(data = patch_summary_compare,
+             aes(nfixes_real, 
+                 nfixes),
+             show.legend = F)+
+  scale_x_log10()+
+  scale_y_log10()+
+  coord_cartesian(xlim = c(10, NA))+
+  ggthemes::theme_few()+
+  labs(x = "# positions (real)",
+       y = "# positions (inferred)")
+```

### 2020-10-02 05_calibration_data.Rmd
some new figures

@@
 data <- fread("data/atlas1060_allTrials_annotated.csv")
-
+data_raw <- copy(data)
 # plot data
@@
 
-Save a raw copy.
+Save an unprocessed copy.
 
 ```{r}
-data_raw <- copy(data)
+data_unproc <- copy(data)
 ```
@@
   ggplot()+
-  geom_path(data = data_raw,
+  geom_path(data = data_unproc,
             aes(x, y),
              col = "grey", alpha = 0.5, size = 0.3)+
-  geom_point(data = data_raw[!data, on = c("x", "y")],
-             aes(x, y),
-             col = "red", alpha = 1, size = 2)+
-  geom_point(data = data,
-             aes(x, y),
-             col = "steelblue", alpha = 0.2, size = 2)+
+  geom_point(data = data_unproc,
+             aes(x, y,
+                 col = x > 645000),
+             alpha = ifelse(data_unproc$x > 645000, 0.5, 1),
+             size = ifelse(data_unproc$x > 645000, 0.3, 2),
+             show.legend = F)+
+  scale_color_brewer(palette = "Set1")+
   geom_vline(xintercept = 645000,
              col = "grey", lty = 2)+
+  ggspatial::annotation_scale(location = "br")+
   ggthemes::theme_map()+
@@
 # save a copy
-data_raw <- copy(data)
+data_unproc <- copy(data)
 
@@
 ```{r}
-fig_low_nbs <- ggplot()+
+fig_low_nbs <-
+  ggplot()+
   geom_path(data = data,
@@
              col = "grey", alpha = 0.5, size = 0.3)+
-  geom_point(data = data_raw[!data, on = c("x", "y")],
-            aes(x, y),
-            shape = 4,
-            col = "red", alpha = 0.5, size = 0.3)+
-  geom_point(data = data,
-             aes(x, y, col = NBS),
-             alpha = 1, size = 0.2)+
-  scico::scale_colour_scico(palette = "davos",
-                         labels = scales::comma, end = 0.9)+
+  geom_point(data = data_unproc,
+            aes(x, y,
+                col = as.factor(NBS > 3)),
+            shape = 1,
+            alpha = 0.5, size = 0.3, show.legend = F)+
+  scale_color_brewer(palette = "Set1")+
   ggthemes::theme_map()+
+  ggspatial::annotation_scale(location = "tl")+
   coord_sf(crs = 32631)+
@@
   geom_point(data = data,
-             aes(x, y, col = speed_in),
-             size = 0.2)+
-  scico::scale_colour_scico(palette = "davos", 
-                         direction = 1, 
-                         limits = c(NA, speed_angle_thresholds[1]),
-                         na.value = "red",
-                         labels = scales::comma,
-                         end = 0.9,
-                         trans = "sqrt")+
+             aes(x, y, 
+                 col = speed_in < 15.7),
+             size = 0.3, alpha = 0.5,
+             show.legend = F)+
+  scale_color_brewer(palette = "Set1")+
   ggthemes::theme_map()+
-  theme(legend.position = c(0, 0.8))+
+  ggspatial::annotation_scale(location = "tl")+
   coord_sf(crs = 32631)+
@@
 # make a copy
-data_raw <- copy(data)
+data_unproc <- copy(data)
 
@@
 # apply a 5 point median smooth, first make a copy
-data_raw <- copy(data)
+data_unproc <- copy(data)
 
@@
 # make figure
-fig_speed_outliers <-
+fig_smooth <-
   ggplot()+
-  geom_path(data = data_raw,
-            aes(x, y, col = "red"), 
-            alpha = 0.5, size = 4)+
   geom_path(data = data,
-             aes(x, y, col = "black"))+
-  scico::scale_colour_scico_d(palette = "roma",
-                         direction = -1, begin = 0.1,
-                         labels = c("smoothed", "raw"))+
-  ggthemes::theme_map()+
-  theme(legend.position = c(0, 0.8))+
+             aes(x, y),
+            col = RColorBrewer::brewer.pal(4, "Set1")[4])+
+  annotate(geom = "rect",
+           xmin = c(650785, 650045),
+           ymin = c(5904450, 5902700),
+           xmax = c(652500, 650682),
+           ymax = c(5906133, 5903080),
+           fill = "grey90",
+           alpha = 0.5,
+           col = NA)+
+    annotate(geom = "text",
+           x = c(650785, 650045) + 400,
+           y = c(5904450, 5902700) + 300,
+           col = "black",
+           label = c("(a)", "(b)"))+
+  theme_void()+
   coord_sf(crs = 32631)+
@@
 
+# make zoomed in figures
+fig_inset <- mapply(function(xl, yl) {
+  ggplot()+
+    geom_point(data = data_raw[!data_unproc, on = c("x", "y")],
+               aes(x, y),
+               col = "grey",
+               shape = 4,
+               size = 0.4)+
+    geom_point(data = data_unproc,
+               aes(x, y),
+               col = RColorBrewer::brewer.pal(3, "Set1")[3],
+               shape = 1, 
+               alpha = 0.5)+
+    geom_path(data = data,
+              aes(x, y),
+              col = RColorBrewer::brewer.pal(4, "Set1")[4])+
+    ggthemes::theme_few()+
+    theme(axis.title = element_blank(),
+          axis.text = element_blank())+
+    ggspatial::annotation_scale(location = "br")+
+    coord_sf(crs = 32631,
+             xlim = xl,
+             ylim = yl, expand = F
+             )+
+    labs(colour = "track")
+}, list(xl1 = c(650785, 652500),
+        xl2 = c(650045, 650682)),
+   list(yl1 = c(5904450, 5906133),
+        yl2 = c(5902700, 5903080)),
+SIMPLIFY = FALSE)
+
+# add global figure to second inset
+fig_inset[[2]] <- fig_inset[[2]] +
+  annotation_custom(grob = ggplotGrob(fig_smooth),
+                    xmin = 649990,
+                    xmax = 650045 + (650682 - 650045)/2,
+                    ymin = 5903080 - (5903080 - 5902700)/2.5,
+                    ymax = 5903080
+                    )
+
+wrap_plots(fig_inset)+
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")")
 # save figure
-ggsave(fig_speed_outliers, 
-       filename = "figures/fig_calib_median_smooth.png",
-       width = 185 / 25, height = 185 / 25)
+ggsave(filename = "figures/fig_calib_median_smooth.png",
+       width = 185 / 25, height = 90 / 25)
 ```
@@
 # make a data copy
-data_raw <- copy(data)
+data_unproc <- copy(data)
 
@@
 # get a simple inverse of speed as residence time classification
-data_slow <- copy(data_raw)
+data_slow <- copy(data_unproc)
 data_slow <- data_slow[speed_in < 1, ]
@@
 
+# assign a crs
+patch_sf_data <- lapply(patch_sf_data, function(df) {
+  sf::st_crs(df) <- 32631
+  return(df)
+})
+
 patch_summary_data <- lapply(list(patch_res_known, patch_res_unknown),
@@
             show.legend = FALSE)+
-  scico::scale_color_scico_d(palette = "berlin",
-                             direction = -1,
-                             begin = 0.2, end = 0.8)+
-  ggthemes::theme_map()
+  scale_color_brewer(palette = "Set1")+
+  ggthemes::theme_map()+
+  coord_sf(crs = 32631)
 
 # figure of residence patches known
-fig_patch_known <- 
+# fig_patch_known <- 
   ggplot()+
@@
   ggthemes::theme_map()
+
+pla = mapply(function(pol) {
+
+  bbox = st_bbox(pol)
+
+  ggplot()+
+    geom_path(data = data_unproc,
+               aes(x, y),
+               col = RColorBrewer::brewer.pal(4, "Set1")[4])+
+    geom_sf(data = pol,
+            alpha = 0.2)+
+    ggthemes::theme_map()+
+    coord_sf(
+      expand = F,
+      xlim = c(bbox[["xmin"]], bbox[["xmax"]]),
+      ylim = c(bbox[["ymin"]], bbox[["ymax"]]))
+}, patch_sf_data[[1]]$polygons, SIMPLIFY = F)
+
+fig_patches_small <- wrap_plots(pla)+
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")")
 
@@
   geom_point(data = patch_summary_compare,
-             aes(nfixes_real, 
-                 nfixes),
+             aes(duration_real, 
+                 duration),
              show.legend = F,
@@
   geom_smooth(data = patch_summary_compare,
-             aes(nfixes_real, 
-                 nfixes),
+             aes(duration_real, 
+                 duration),
+             se = F,
+             method = "glm",
              show.legend = F)+
-  scale_x_log10()+
-  scale_y_log10()+
-  coord_cartesian(xlim = c(10, NA))+
+  # scale_x_log10()+
+  # scale_y_log10()+
+  # coord_cartesian(xlim = c(10, NA))+
   ggthemes::theme_few()+

### 2020-10-02 02_filtering_data.Rmd
update figures'

@@
 ---
-
-This section covers:
-
-1. Simple spatio-temporal filtering of point data to remove gross positioning errors,
-
-2. Filtering positions on covariates that may be associated with increased positioning error,
-
-3. Filtering movement tracks using the 'non-movement' approach to remove implausible movement segments, and
-
-4. The challenge of reflected positions, and how it can be addressed.
 
@@
 
-Here we read the simulated data in.
-The limits of the coordinates are 0 and 1, so we add a few extraneous coordinates within a plausible distance to demonstrate the basic forms of filtering.
-We also assign a fictional _NBS_ value ranging between 2 and 6 to demonstrate filtering on covariates.
-Finally, we introduce a 'reflection', a phenomenon that results when a TOA system's position likelihood estimator shifts the real position by a large distance for an extended duration.
-
 ```{r read_sim_data}
@@
 data_copy <- copy(data)
-
+# add a reflection
+data_copy[500:800, `:=`(x = x + 0.25,
+                        y = y + 0.25)]
 # add normal error
@@
 
+```{r}
+# define red and blue
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+```
+
 ## Show data with errors
@@
 ```{r}
-# make list of data to plot
-data_plot <- list(data, data_copy)
-
-# make plot list
-plots <- lapply(data_plot, function(df) {
-  ggplot(data = df)+
-
-    geom_path(aes(x, y),
-              col = "red", 
-              alpha = 1, lwd = 0.2)+
-    geom_point(aes(x, y),
-              col = "darkblue", 
-              alpha = 0.1,
-              shape = 16)+
-    # coord_cartesian(xlim = c(0.5, 1),
-    #                 ylim = c(0.5, 1)) +
+# make figure
+figure_raw <-
+  ggplot()+
+  geom_point(data = data_copy,
+            aes(x, y),
+            col = "grey",
+            alpha = 1,
+            size = 0.2)+
+  geom_path(data = data,
+            aes(x, y),
+            col = "grey20",
+            alpha = 1)+
+  ggthemes::theme_map()+
   coord_equal()+
-    theme_test()+
   theme(axis.text = element_blank(),
-          axis.title = element_blank())
-})
-
-# wrap plots
-plot_figure <- wrap_plots(plots) +
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")") &
-  theme(plot.tag = element_text(face = "bold"))
-
-plot_figure
-
-# save figure
-ggsave(plot_figure, filename = "figures/fig_add_errors.png",
-       width = 185, height = 90, units = "mm")
+        axis.title = element_blank(),
+        legend.position = "none")+
+    theme(plot.background = element_rect(fill = NA))+
+  labs(colour = NULL)
 ```
@@
 # NB: set remove_inside to FALSE
-data_inside_bbox <- atlastools::atl_filter_bounds(data = data_copy,
+data_inside_bbox <- atl_filter_bounds(data = data_copy,
                                       y_range = c(0.5, 1),
@@
 ```{r}
-# data to plot
-data_plot <- list(data_copy, data_inside_bbox)
-
-# make plots
-plots <- lapply(data_plot, function(df) {
+# plot data inside and outside bbox
+fig_filter_bounds <-
   ggplot()+
-
-    geom_path(data = df,
+  geom_point(data = data_inside_bbox,
              aes(x, y),
-              col = "red", 
-              alpha = 1, lwd = 0.2)+
-    geom_point(data = df,
+             col = pal[2],
+             alpha = 1, size = 0.2)+
+  geom_point(data = data_copy[!data_inside_bbox,
+                              on = c("x", "y")],
              aes(x, y),
-              col = "darkblue", 
-              alpha = 0.1,
-              pch = 16)+
-    theme_test()+
-    coord_equal() +
-    theme(axis.title = element_blank(),
-          axis.text = element_blank())
-})
-
-# add elements
-plots[[1]] <- plots[[1]] + 
-  annotate(xmin = 0.5, xmax = 1,
-           ymin = 0.5, ymax = 1,
-           fill = NA, col = "black", 
-           lty = 2, geom = "rect")+
-  annotate(x = 0.5, y = 1.1, label = "(b)", 
-           fontface = "bold", geom = "text", size = 4)
+             col = pal[1],
+             alpha = 0.5, size = 0.2)+
+  geom_path(data = data,
+            aes(x, y),
+            col = "grey20",
+            alpha = 1)+
+  geom_hline(yintercept = c(0.5, 1),
+             col = "grey",
+             lty = 2)+
+  ggthemes::theme_map()+
+  theme(plot.background = element_rect(fill = NA))+
+  coord_equal()
 
 # wrap plots
-plot_figure <- wrap_plots(plots) +
+plot_figure <-
+  wrap_plots(list(figure_raw, fig_filter_bounds)) +
   plot_annotation(tag_levels = "a",
@@
 
-ggsave(plot_figure, filename = "figures/fig_filter_bounds.png", 
-       width = 185, height = 90, units = "mm")
+ggsave(plot_figure, filename = "figures/fig_raw_bounds.png", 
+       width = 170, height = 170, units = "mm")
 ```
@@
 ## Removing point outliers: The non-movement approach
-
-Bjorneraas et al.
-(2010) describe intuitive filtering steps for trajectories that are based in the biology of the study species (moose _Alces alces_ in their case).
-Briefly, this 'non-movement method' identifies implausible segments of a trajectory, and removes them.
-Here, we adopt elements of their method to identify and remove 'spikes' or point outliers in the movement track, i.e., positions which indicate an improbably fast and short excursion away from the real trajectory.
-
-We begin by calculating the speed and turning angle along the track.
-For both metrics, the first value is set at `NA`, indicating that we do not know the position of the individual prior to the start of the trajectory.
-The turning angle is calculated using the cosine rule and requires three positions to yield a values; thus it has an `NA` at the very last position as well.
 
@@
 This is more general than identifying the limits of implausibility for each individual (since there may be inter-individual differences), let alone each species in a large dataset.
+
 An alternative approach is to define a speed cutoff based on expert knowledge, and this is best suited to well studied species.
@@
 data_filtered <- atl_filter_covariates(data_copy,
-            filters = c("(in_speed < 0.03 & out_speed < 0.03) | angle < 40"))
+    filters = c("(in_speed < 0.024 & out_speed < 0.024) | angle < 40"))
 ```
@@
 # data plot
-data_plot <- list(data_copy, data_filtered)
-
-plots <- lapply(data_plot, function(df) {
+fig_outlier_remove <-
   ggplot()+
-     geom_path(data = df,
+  geom_path(data = data_copy,
             aes(x, y),
-              col = "red", 
-              alpha = 1, lwd = 0.2)+
-    geom_point(data = df,
-               alpha = 0.1,
+            col = "grey",
+            lwd = 0.2)+
+  geom_point(data = data_copy[500:800, ],
+             aes(x, y),
+             shape = 2, col = pal[4])+
+  geom_point(data = data_copy[!data_copy[500:800, ],
+                              on = c("x", "y")],
              aes(x, y, 
-                   col = (in_speed >= 0.03 & out_speed >= 0.03)),
-               shape = ifelse((df$in_speed >= 0.03 & df$out_speed >= 0.03), 
-                              16, 1),
-               show.legend = F)+
-    scale_colour_manual(values = c("darkblue", "red"))+
-    theme_test()+
+                 col = (in_speed >= 0.03 & out_speed >= 0.03),
+                 shape = (in_speed >= 0.03 & out_speed >= 0.03)),
+             show.legend = F,
+             alpha = 0.5)+
+  geom_path(data = data,
+            aes(x, y),
+            col = "grey20",
+            alpha = 1)+
+  geom_path(data = data,
+            aes(x, y),
+            col = "grey20",
+            alpha = 1)+
+  ggthemes::theme_map()+
+  scale_color_brewer(palette = "Set1", direction = -1)+
+  scale_shape_manual(values = c(1, 19))+
   coord_equal()+
-    theme(axis.title = element_blank(),
-          axis.text = element_blank())
-})
-
-plot_figure <- wrap_plots(plots) +
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")") &
-  theme(plot.tag = element_text(face = "bold"))
-
-plot_figure
-ggsave(plot_figure, filename = "figures/fig_filter_spikes.png", 
-       width = 185, height = 90, units = "mm")
-```
-
-## Does removing point outliers help?
-
-### Making error combinations
-
-Point outliers in a trajectory can be boiled down to two parameters: (1) how many positions have been converted into point outliers or spikes, and (2) how large the spikes are relative to the scale of the movement track.
-
-Here we create combinations of these parameters and create 25 replicates of each combination using the error generator on the imported simulated data.
-
-```{r outlier_error_combinations}
-# define unique values
-p_outlier <- c(0.05)
-ext_outlier <- c(0.01, 0.05, 0.1)
-replicate <- seq_len(10)
-
-# make combinations
-param_c <- CJ(p_outlier, ext_outlier, replicate)
+  theme(plot.background = element_rect(fill = NA))
 ```
@@
 ## Reflections: Challenges to the non-movement approach
-
-A peculiar issue affecting TOA tracking systems is the phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.
-
-These reflections cannot be resolved in the same way as spikes, since they consist of several points with very realistic speeds bookended by a pair with either an extreme incoming or outgoing speed.
-
-A reflection looks like this:
-
-```{r}
-# make a copy of data with spikes removed
-data_with_reflection <- copy(data_filtered)
-
-# add a reflection
-data_with_reflection[500:600, `:=`(x = runif(101, 1, 1.01),
-                                   y = runif(101, 1, 1.01))]
-```
-
-Removing reflections involves using the phenomenon’s properties to first identify its bounds and then removing all positions between these bounds.
-
-In some instances, the estimated location is instantaneously displaced from the real movement track for multiple consecutive timesteps in the sampling process, followed by a return to the real position.
-Removing such prolonged spikes (which we call `reflections') using Bjorneraas and colleagues' approach (2010) would be ineffective --- the bounds of the reflection have either extreme incoming \textit{or} outgoing speeds, and there are no positions with both.
-When encountering such data, users are advised to identify the bounds of the reflection as first step to removing them: the inner bound (where the position estimate first shifts away from the track) will have a high incoming speed ($v_i$), while the outer bound (after which the position estimate returns to the real track) will have a high outgoing speed ($v_o$).
-We suggest locating the first reflected point $p_1$ by comparing the incoming speeds of the track positions ($v_i$) against the speed cutoff $S$.
-An optional turning angle cutoff $A$ may be used if the track includes plausible fast transit segments.
-The first position to satisfy this comparison $v_i > S$ AND $\theta > A$ can be considered the inner bound of the reflection and set as an `anchor point'.
-The position after the anchor point with the highest outgoing speed $v_o$ (such that $v_o > S$) can be considered to be the outer bound of the reflection.
-The positions between these bounds can be identified as a trajectory reflection and removed.
-For a system-specific reasons and the pre-processing steps thus far, a reflection may not end before the last position in the movement track.
-In such cases, the entirety of the track following the first anchor point should be removed, as it is better to exclude moderate chunks of data than to make inferences from unsound data.
-Removing reflections is implemented in a convenient form in \texttt{atlastools} in the \texttt{{atl\_remove\_reflections}} function.
-
-The procedure is as follows:
-
-1. Identify positions with anomalous (usually extremely high) speeds and turning angles, as with spikes, and
-
-2. Split the trajectory into pre- and post-anomaly segments, and the first position of the post-anomaly segment, which is the inner bound of the reflection, can be set as an anchor point.
-
-3. The first position after the anchor point with an instantaneous outgoing speed >= the speed cutoff is set as the outer bound of the reflection.
-
-4. The positions between these bounds can be identified as a trajectory reflection and removed, along with one position before and after the bounds for safety's sake.
-
-This method is robust to variable displacement of the reflections from the real positions, and also to displacement in the real position of the individual itself.
 
@@
                           point_angle_cutoff = 10,
-                                              reflection_speed_cutoff = 0.03)
+                          reflection_speed_cutoff = 0.024)
+# get reflections
+reflection <- data_with_reflection[!data_no_reflection,
+                                         on = c("x", "y")]
+reflection <- na.omit(reflection)
 ```
@@
 ```{r}
-# data to plot
-plot_data <- list(data_with_reflection, data_no_reflection)
-
 # get plots
-plots <- lapply(plot_data, function(df) {
+fig_reflection <-
   ggplot()+
-    geom_path(data = df,
+    geom_path(data = reflection,
              aes(x, y),
-              col = "red", 
-              alpha = 1, lwd = 0.2)+
-    geom_point(data = df,
+             alpha = 1,
+             col = "grey")+
+  geom_point(data = reflection,
              aes(x, y),
-               alpha = 0.1,
-               colour = "darkblue",
-               shape = 16,
+             alpha = 0.5,
+             col = pal[1],
+             shape = 2)+
+  geom_point(data = data_no_reflection,
+             aes(x, y),
+             alpha = 0.5,
+             colour = pal[2],
+             shape = 1,
              show.legend = F)+
-    scale_colour_manual(values = c("darkblue", "red"))+
-    theme_test()+
+  geom_path(data = data,
+            aes(x, y),
+            col = "grey20",
+            alpha = 1)+
+  ggthemes::theme_map()+
   coord_equal()+
-    theme(axis.title = element_blank(),
-          axis.text = element_blank())
-})
+  theme(plot.background = element_rect(fill = NA))
 
 # wrap figures
-plot_figure <- wrap_plots(plots) +
+plot_figure <- wrap_plots(list(fig_outlier_remove, 
+                               fig_reflection)) +
   plot_annotation(tag_levels = "a",
@@
 
-ggsave(plot_figure, filename = "figures/fig_filter_reflection.png", 
-       width = 185, height = 90, units = "mm")
+ggsave(plot_figure, filename = "figures/fig_correct_tracks.png", 
+       width = 170, height = 170, units = "mm")
 ```

### 2020-10-02 03_smoothing_tracks.Rmd
update figures'

@@
 ```{r prep_libs_02_01}
+# prep libs
 library(data.table)
@@
 library(ggplot2)
+library(patchwork)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
 ```
@@
 # make list of data to plot
-data_plot <- append(list(data, data_errors), list_of_smooths)
+data_plot <- append(list(data_errors), list_of_smooths)
 
-# add offset
-data_plot <- mapply(FUN = function(df, offset) {
-  df[, x := x + offset]
-}, data_plot, seq(from = 0, to = 1, length.out = length(data_plot)),
-SIMPLIFY = FALSE)
+figure_median_smooth <- 
+  mapply(function(df, col) {
+    ggplot()+
+      geom_point(data = df,
+                 aes(x, y),
+                 size = 0.5,
+                 show.legend = FALSE,
+                 col = col,
+                 shape = 16)+
+      geom_path(data = data,
+                aes(x, y),
+                lwd = 0.2,
+                col = pal[4])+
+      scale_colour_distiller(
+        palette = "YlGnBu",
+        na.value = "grey20")+
+      coord_cartesian(expand = T,
+                  ylim = c(0.6, 0.8)) +
+      ggthemes::theme_map()+
+      theme(plot.background = element_rect(fill = NA))
+}, data_plot, c("grey", RColorBrewer::brewer.pal(7, "Greens")[2:5]),
+SIMPLIFY = F)
 
-# bind rows
-data_plot <- rbindlist(data_plot, fill = TRUE)
+# make basic plot with rectangle
+fig_bounds <- ggplot()+
+  geom_path(data = data,
+            aes(x, y),
+            col = pal[4])+
+  geom_hline(yintercept = c(0.6, 0.8),
+             col = "grey20",
+             lty = 2, lwd = 0.2)+
+  coord_equal() +
+  ggthemes::theme_map()
 
-plot_figure <- ggplot()+
-  # geom_point(data = data_plot,
-  #            aes(x, y),
-  #            col = "darkgrey", 
-  #            alpha = 1,
-  #            shape = 1)+
-  geom_path(data = data_plot,
-            aes(x, y, 
-                col = window_size,
-                group = window_size),
-            alpha = 1,
-            show.legend = FALSE)+
-  annotate(x = 0.6 + seq(0, 1, length.out = 6),
-           y = 1.05,
-           label = glue::glue('({letters[seq(6)]})'),
-           geom = "text",
-           fontface = "bold")+
-  scale_colour_viridis_c(option = "D",
-                         na.value = "red")+
-  coord_equal() +
-  theme_test()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())
+# append to first figure
+figure_median_smooth[[1]] <-
+  figure_median_smooth[[1]]+
+  annotation_custom(grob = ggplotGrob(fig_bounds),
+                    ymin = 0.6, ymax = 0.8,
+                    xmin = 0.55, xmax = 0.7)
 
-plot_figure
+# wrap plots
+fig_median_smooth <- wrap_plots(figure_median_smooth, ncol = 3)+
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
 
+fig_median_smooth
 # save figure
-ggsave(plot_figure, filename = "figures/fig_median_smooth.png",
-       width = 185, height = 185 * 2 / 3, units = "mm")
+ggsave(fig_median_smooth, filename = "figures/fig_median_smooth.png",
+       width = 170, height = 90, units = "mm")
 ```

### 2020-10-05 03_smoothing_tracks.Rmd
some code edits

--- 2020-09-18 03_smoothing_tracks.Rmd
@@
 
+### Robustness to large gaps
+
+```{r}
+# make a single list
+data_plot <- append(list(data_errors), list_of_smooths)
+
+# now prep gap
+fig_smooth_1d <- 
+  ggplot()+
+  geom_path(data = data[1:2000,],
+            aes(time, y),
+            size = 0.2)+
+  geom_point(data = rbindlist(data_plot),
+             aes(time, y,
+                 col = factor(window_size)),
+             alpha = 0.5,
+             size = 1,
+             show.legend = FALSE,
+             shape = 16)+
+  coord_cartesian(xlim = c(5000, 7000),
+                  ylim = c(0.7, NA),
+                  expand = F)+
+  scale_colour_brewer(palette = "Spectral")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank())+
+  labs(x = "position index", y = "Y coordinate")
+```
+
+### Plot results
+
 ```{r}
@@
 # make list of data to plot
-data_plot <- append(list(data, data_errors), list_of_smooths)
+figure_median_smooth <- 
+  mapply(function(df, col) {
+    ggplot()+
+      geom_point(data = df,
+                 aes(x, y),
+                 size = 0.5,
+                 alpha = 0.5,
+                 show.legend = FALSE,
+                 col = col,
+                 shape = 1)+
+      geom_path(data = data,
+                aes(x, y),
+                col = "grey20")+
+      coord_cartesian(expand = T,
+                  ylim = c(0.6, 0.8)) +
+      ggthemes::theme_map()
+}, data_plot, c(pal[2], rep(
+  RColorBrewer::brewer.pal(3, "Greens")[2],
+  4)),
+SIMPLIFY = F)
 
-# add offset
-data_plot <- mapply(FUN = function(df, offset) {
-  df[, x := x + offset]
-}, data_plot, seq(from = 0, to = 1, length.out = length(data_plot)),
-SIMPLIFY = FALSE)
+# make basic plot with rectangle
+fig_bounds <- 
+  ggplot()+
+  geom_path(data = data,
+            aes(x, y))+
+  geom_hline(yintercept = c(0.6, 0.8),
+             col = "grey20",
+             lty = 2, lwd = 0.2)+
+  coord_equal(expand = F) +
+  ggthemes::theme_map()
 
-# bind rows
-data_plot <- rbindlist(data_plot, fill = TRUE)
+# append to first figure
+figure_median_smooth[[1]] <-
+  figure_median_smooth[[1]]+
+  annotation_custom(grob = ggplotGrob(fig_bounds),
+                    ymin = 0.6, ymax = 0.8,
+                    xmin = 0.5, xmax = 0.7)
 
-plot_figure <- ggplot()+
-  # geom_point(data = data_plot,
-  #            aes(x, y),
-  #            col = "darkgrey", 
-  #            alpha = 1,
-  #            shape = 1)+
-  geom_path(data = data_plot,
-            aes(x, y, 
-                col = window_size,
-                group = window_size),
-            alpha = 1,
-            show.legend = FALSE)+
-  annotate(x = 0.6 + seq(0, 1, length.out = 6),
-           y = 1.05,
-           label = glue::glue('({letters[seq(6)]})'),
-           geom = "text",
-           fontface = "bold")+
-  scale_colour_viridis_c(option = "D",
-                         na.value = "red")+
-  coord_equal() +
-  theme_test()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())
+# wrap plots
+fig_median_smooth <- wrap_plots(append(figure_median_smooth, 
+                                       list(fig_smooth_1d)), 
+                                       ncol = 3)+
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
 
-plot_figure
+fig_median_smooth
 
 # save figure
-ggsave(plot_figure, filename = "figures/fig_median_smooth.png",
-       width = 185, height = 185 * 2 / 3, units = "mm")
+ggsave(fig_median_smooth, filename = "figures/fig_median_smooth.png",
+       width = 225, height = 225 / 1.77, units = "mm")
 ```
@@
 # choose the data
-data_agg <- copy(data)
+data_agg <- copy(list_of_smooths[[3]])
 
-list_of_agg <- lapply(c(10, 30, 120), function(z) {
-  data_agg <- atl_thin_data(data = data_agg,
+# get list of aggregated data
+list_of_agg <- lapply(c(3, 10, 30, 120), function(z) {
+  data_return <- atl_thin_data(data = data_agg,
                             interval = z,
                             method = "aggregate")
-  data_agg[, interval := z]
-
+  data_return[, interval := z]
+  return(data_return)
 })
@@
 ```{r}
-# plot
-data_plot <- append(list(data[, `:=`(SD = 0, interval = NA)]), list_of_agg)
-
-# add offset
-data_plot <- mapply(FUN = function(df, offset) {
-  df[, x := x + offset]
-}, data_plot, seq(from = 0, to = 1, length.out = length(data_plot)),
+# make figure
+figure_aggregate <- 
+  mapply(FUN = function(df) {
+  ggplot(data = df)+
+      geom_point(data = list_of_smooths[[3]],
+                 aes(x, y),
+                 col = pal[3],
+                 size = 0.3, alpha = 0.5)+
+      geom_point(aes(x, y, size = SD),
+                 colour = pal[4],
+                 alpha = 1,
+                 stroke = 1,
+                 shape = 4, show.legend = F)+
+      geom_path(aes(x, y),
+                 colour = pal[4],
+                 alpha = 1)+
+      geom_path(data = data,
+                lwd = 0.2,
+                aes(x, y),
+                col = "grey20")+
+      coord_equal(ylim = c(0.6, 0.8))+
+      ggthemes::theme_few()+
+      theme(axis.title = element_blank(),
+            axis.text = element_blank())
+}, list_of_agg[2:3],
 SIMPLIFY = FALSE)
 
-data_plot <- rbindlist(data_plot, fill = TRUE)
+figure_aggregate <- wrap_plots(figure_aggregate)
 
-plot_figure <- ggplot(data = data_plot)+
-  geom_point(aes(x, y, size = SD),
-             colour = "darkblue",
-             alpha = 0.5,
-             shape = 16, show.legend = F)+
-  geom_path(aes(x, y,
-                group = interval,
-                col = interval),
-            col = "red",
-            show.legend = FALSE)+
-  annotate(x = 0.6 + seq(0, 1, length.out = 4),
-           y = 1.05,
-           label = glue::glue('({letters[seq(4)]})'),
-           geom = "text",
-           fontface = "bold")+
-  coord_equal()+
-  theme_test()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())
-
-plot_figure
-
+figure_aggregate
 # save figure
@@
 ```
+
+### Aggregation before correction, a warning
+
+```{r}
+# read data with errors
+data_errors <- fread("data/data_errors.csv")
+
+# aggregate before correction
+list_of_agg <- lapply(c(3, 10, 30, 120), function(z) {
+  data_return <- atl_thin_data(data = data_errors,
+                            interval = z,
+                            method = "aggregate")
+  data_return[, interval := z]
+  data_return[, speed := atl_get_speed(data_return)]
+  return(data_return)
+})
+
+# get real speed
+data[, speed := atl_get_speed(data)]
+
+### plot figures
+fig_agg_data <- 
+  lapply(list_of_agg, function(df) {
+    ggplot(df)+
+      geom_path(aes(x,y), 
+                # col = factor(interval), 
+                # group = interval),
+                lwd = 0.2)+
+      geom_point(aes(x,y, 
+                     size = SD,
+                     shape = speed < quantile(data$speed, 0.95, na.rm = T),
+                     col = speed < quantile(data$speed, 0.95, na.rm = T),
+                     group = interval),
+                 stroke = ifelse(df$speed < quantile(data$speed, 0.95, na.rm = T),
+                                 1, 1),
+                 show.legend = F)+
+      geom_path(data = data,
+                aes(x, y),
+                lwd = 0.1)+
+      scale_shape_manual(values = c(2, 4))+
+      scale_colour_manual(values = pal[c(1, 2)])+
+      coord_cartesian(ylim = c(0.6, NA))+
+      ggthemes::theme_map()  
+  })
+```
+
+```{r}
+# now plot hist of speed
+data_agg <- rbindlist(list_of_agg)
+
+# show boxplot of speed
+fig_agg_speed <- ggplot(data_agg)+
+  geom_jitter(aes(factor(interval), speed),
+              col = "grey",
+              shape = 4, alpha = 0.5)+
+  geom_boxplot(aes(factor(interval), speed),
+               fill = NA,
+               show.legend = F, 
+               width = 0.3, 
+               outlier.size = 0.2)+
+  geom_hline(yintercept =
+               c(mean(data$speed, na.rm = T),
+                 quantile(data$speed, na.rm = T,
+                          probs = c(0.95))
+               ),
+             lty = c(1, 2))+
+  scale_y_log10(label = scales::comma)+
+  ggthemes::theme_few()+
+  theme(axis.text.y = element_blank())+
+  labs(x = "interval (s)",
+       y = "speed")
+```
+
+### Combine with aggregations
+
+```{r}
+# make combined figure
+fig_aggregate <- 
+  wrap_plots(append(fig_agg_data[2:3], list(fig_agg_speed)),
+           design = "AAABBBCC")+
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+# save figure
+ggsave(fig_aggregate,
+       filename = "figures/fig_aggregate_data.png",
+       width = 170, height = 170 / 1.67, units = "mm")
+```

### 2020-10-05 11_residence_patch.Rmd
some code edits

@@
 library(ggplot2)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
 ```
@@
                        lim_time_indep = 3)
+
 # get spatial representation
@@
 ```{r}
-plot_patches <- ggplot()+
+plot_patches <-
+  ggplot()+
   geom_sf(data = patch_sf,
-          colour = "NA",
-          aes(fill = duration),
+          colour = "grey20",
+          fill = "grey90",
           show.legend = FALSE)+
-  geom_point(data = data,
+  geom_point(data = patch_summary,
+             aes(x_median + 0.5, 
+                 y_median,
+                 size = duration),
+             # size = 8,
+             shape = 2,
+             col = pal[2],
+             stroke = 2,
+             show.legend = F)+
+  geom_path(data = data,
              aes(x, y),
-             size = 0.1,
-             alpha = 0.1)+
-  geom_point(data = patch_summary,
-             aes(x_median + 0.5, y_median),
-             size = 5,
-             shape = 21,
-             fill = "grey")+
-  annotate(geom = "text",
-           x = patch_summary$x_median + 0.5, 
-           y = patch_summary$y_median + 0.02,
-           label = patch_summary$patch,
-           col = "red")+
-  annotate(geom = "text",
-           x = patch_summary$x_median, 
-           y = patch_summary$y_median + 0.02,
-           label = patch_summary$patch,
-           col = "red")+
+            size = 0.3)+
+  geom_curve(data = patch_summary,
+            aes(x = x_end + 0.5, y = y_end,
+                xend = shift(patch_summary$x_start, 
+                             type = "lag") + 0.5,
+                yend = shift(patch_summary$y_start, 
+                             type = "lag")),
+            col = "grey30",
+            size = 0.3,
+            ncp = 2,
+            curvature = -0.4,
+            arrow = arrow(angle = 20,
+                          length = unit(2, units = "mm"), 
+                          type = "closed", ends = "first"))+
   annotate(geom = "text",
@@
            fontface = "bold")+
-  geom_path(data = patch_summary,
-            aes(x_median + 0.5, y_median),
-            colour = "darkblue",
-            arrow = arrow(angle = 10, 
-                          type = "closed"))+
-  scale_fill_viridis_c()+
-  theme_void()
+  scale_colour_distiller(palette = "YlGnBu",
+                         trans = "log10")+
+  coord_sf()+
+  ggthemes::theme_map()+
+  theme(legend.position = "right")
 
@@
        filename = "figures/fig_residence_patch.png",
-       width = 185, height = 185 * 2 / 3, units = "mm")
+       width = 170, height = 100, units = "mm")
 ```

### 2020-10-06 03_smoothing_tracks.Rmd
some more changes

@@
   ggplot()+
-  geom_path(data = data[1:2000,],
-            aes(time, y),
-            size = 0.2)+
   geom_point(data = rbindlist(data_plot),
@@
   ggplot()+
+  geom_point(data = data_errors,
+            aes(x, y),
+            size = 0.1, alpha = 0.2, 
+            col = "grey")+
   geom_path(data = data,
-            aes(x, y))+
+            aes(x, y),
+            size = 0.1)+
   geom_hline(yintercept = c(0.6, 0.8),

### 2020-10-06 05_calibration_data.Rmd
some more changes

@@
 # plot data
-fig_data_raw <- ggplot(data)+
+fig_data_raw <-
+  ggplot(data)+
   geom_path(aes(x, y),
@@
   ggplot()+
-  geom_path(data = data_unproc,
+  geom_path(data = data_raw,
             aes(x, y),
              col = "grey", alpha = 0.5, size = 0.3)+
-  geom_point(data = data_unproc,
+  geom_point(data = data_raw,
              aes(x, y,
                  col = x > 645000),
-             alpha = ifelse(data_unproc$x > 645000, 0.5, 1),
-             size = ifelse(data_unproc$x > 645000, 0.3, 2),
+             alpha = ifelse(data_raw$x > 645000, 0.5, 1),
+             size = ifelse(data_raw$x > 645000, 0.3, 2),
              show.legend = F)+
@@
            col = "black",
-           label = c("(a)", "(b)"))+
+           label = c("d", "e"))+
   theme_void()+
@@
     ggspatial::annotation_scale(location = "br")+
-    coord_sf(crs = 32631,
+    coord_cartesian(
              xlim = xl,
@@
 
+fig_median_smooth <- 
   wrap_plots(fig_inset)+
@@
        width = 185 / 25, height = 90 / 25)
+```
+
+### Plot pre-processing steps
+
+```{r}
+# make combined walkthrough figure
+figure_walkthrough <- 
+  wrap_plots(
+  append(
+    list(
+      fig_data_bbox, fig_low_nbs, fig_speed_outliers), 
+    fig_inset),
+design = "ABD\nACE")+
+  plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")")
+
+# save combined figure
+ggsave(figure_walkthrough,
+       filename = "figures/fig_walkthrough.png",
+       height = 130, width = 225, units = "mm")
 ```
@@
 data_res <- data[residence == T, ]
-
-# get a simple inverse of speed as residence time classification
-data_slow <- copy(data_unproc)
-data_slow <- data_slow[speed_in < 1, ]
 ```
@@
                                 min_fixes = 3)
-
-# on data where the residence is unknown
-patch_res_unknown <- atl_res_patch(data_slow, 
-                                buffer_radius = 10,
-                                lim_spat_indep = 100,
-                                lim_time_indep = 30,
-                                min_fixes = 10)
 ```
@@
 # for the known and unkniwn patches
-patch_sf_data <- lapply(list(patch_res_known, patch_res_unknown),
-                        atl_patch_summary, which_data = "spatial")
+patch_sf_data <- atl_patch_summary(patch_res_known, which_data = "spatial")
 
-# assign a crs
-patch_sf_data <- lapply(patch_sf_data, function(df) {
-  sf::st_crs(df) <- 32631
-  return(df)
-})
+# assign crs
+sf::st_crs(patch_sf_data) <- 32631
 
-patch_summary_data <- lapply(list(patch_res_known, patch_res_unknown),
-                        atl_patch_summary)
+# get summary data
+patch_summary_data <- atl_patch_summary(patch_res_known, which_data = "summary")
 ```
@@
 ```{r}
-# patch with residence points and walking patches
+# patch with residence points and all patches
 fig_basic_residence <-
   ggplot()+
-  geom_point(data = data,
-             aes(x, y,
-                 col = residence),
-            size = ifelse(data$residence, 4, 0.2),
-            shape = ifelse(data$residence, 2, 1),
-            show.legend = FALSE)+
-  scale_color_brewer(palette = "Set1")+
+  geom_path(data = data,
+            aes(x, y),
+            lwd = 0.2,
+            col = "grey20")+
+  geom_point(data = data_res,
+             aes(x, y),
+             shape = 2,
+             size = 2,
+             col = pal[2])+
+  geom_path(data = patch_summary_data,
+            aes(x_median + 200, y_median),
+            col = "orange", lty = 1)+
+  geom_point(data = patch_summary_data,
+             aes(x_median + 200, y_median,
+                 fill = duration / 60),
+             size = 5,
+             shape = 21)+
+
+  scale_fill_distiller(palette = "YlOrRd",
+                       direction = 1,
+                       trans = "log10",
+                       limits = c(1, 60))+
   ggthemes::theme_map()+
-  coord_sf(crs = 32631)
-
-# figure of residence patches known
-# fig_patch_known <- 
-  ggplot()+
-  geom_sf(data = patch_sf_data[[1]],
-          fill = "steelblue", col = NA)+
-  geom_path(data = patch_summary_data[[1]],
-            aes(x_median, y_median),
-            col = "red")+
-  ggthemes::theme_map()
-
-pla = mapply(function(pol) {
-
-  bbox = st_bbox(pol)
-
-  ggplot()+
-    geom_path(data = data_unproc,
-               aes(x, y),
-               col = RColorBrewer::brewer.pal(4, "Set1")[4])+
-    geom_sf(data = pol,
-            alpha = 0.2)+
-    ggthemes::theme_map()+
-    coord_sf(
-      expand = F,
-      xlim = c(bbox[["xmin"]], bbox[["xmax"]]),
-      ylim = c(bbox[["ymin"]], bbox[["ymax"]]))
-}, patch_sf_data[[1]]$polygons, SIMPLIFY = F)
-
-fig_patches_small <- wrap_plots(pla)+
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")")
-
-# figure of unknown patches
-fig_patch_unknown <-
-  ggplot()+
-  geom_path(data = patch_summary_data[[1]],
-            aes(x_median, y_median),
-            col = "black", size = 0.2, lty = 2)+
-  geom_sf(data = patch_sf_data[[2]],
-          fill = "steelblue", col = NA)+
-  geom_segment(data = patch_summary_data[[2]],
-               aes(x = x_end,
-                   xend = shift(x_start, type = "lead"),
-                   y = y_end,
-                   yend = shift(y_start, type = "lead")),
-               col = "red"
-  )+
-  ggthemes::theme_map()
-
-# arrange plots
-fig_res_patch <- 
-  wrap_plots(list(fig_basic_residence, fig_patch_known, fig_patch_unknown),
-           design = "AAAA\nAAAA\nBBCC\nBBCC")+
-  plot_annotation(tag_levels = "a", 
-                  tag_prefix = "(", tag_suffix = ")")
-
-# save figure
-ggsave(fig_res_patch, filename = "figures/fig_calib_respatch.png",
-       height = 200/25, width = 185/25)
+  theme(legend.position = c(0, 0.3),
+        legend.key.width = unit(3, units = "mm"),
+        legend.background = element_blank())+
+  ggspatial::annotation_scale(location = "tl")+
+  coord_sf(crs = 32631)+
+  labs(fill = "inferred\nduration (min)")
 ```
@@
 patch_summary_inferred <- 
-        patch_summary_data[[1]][, 
+  patch_summary_data[, 
                      c("x_median", "y_median", 
@@
 
-### Plot known and unknown patches
+### Plot durations comparisons
+
+```{r}
+# get linear model
+model_duration <- lm(duration ~ duration_real,
+                     data = patch_summary_compare)
+
+# get R2
+summary(model_duration)
+```
 
@@
 # make figure comparing different methods
+figure_patch_duration <-
   ggplot()+
+  geom_segment(data = patch_summary_compare,
+               aes(x = 100,
+                   xend = duration_real,
+                   y = duration,
+                   yend = duration),
+               col = "grey")+
+  geom_segment(data = patch_summary_compare,
+               aes(x = duration_real,
+                   xend = duration_real,
+                   y = 100,
+                   yend = duration),
+               col = "grey")+
   geom_point(data = patch_summary_compare,
@@
              show.legend = F,
-            size = 4,
-            shape = 1)+
+            size = 3,
+            stroke = 1,
+            shape = 21,
+            fill = pal[2],
+            col = "white")+
   geom_smooth(data = patch_summary_compare,
@@
              se = F,
+             size = 0.5,
+             col = pal[1],
+             alpha = 0.2,
              method = "glm",
              show.legend = F)+
-  # scale_x_log10()+
-  # scale_y_log10()+
-  # coord_cartesian(xlim = c(10, NA))+
+  annotate(geom = "text",
+           x = 300, y = 900,
+           size = 5,
+           label = "R^2 == 0.845",
+           parse = T)+
+  scale_x_log10(breaks = c(300, 600, 900),
+                labels = as.integer(c(300, 600, 900) / 60))+
+  scale_y_log10(breaks = c(300, 600, 900),
+                labels = as.integer(c(300, 600, 900) / 60))+
+  coord_equal(expand = F,
+              xlim = c(200, 1200),
+              ylim = c(200, 1200))+
   ggthemes::theme_few()+
-  labs(x = "# positions (real)",
-       y = "# positions (inferred)")
+  labs(x = "duration (min)",
+       y = "inferred duration (min)")
 ```
+
+### Join patch map and duration comparison figure
+
+```{r}
+# wrap together
+figure_res_patch <-
+  wrap_plots(list(fig_basic_residence, figure_patch_duration))+
+    plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+  theme(plot.tag = element_text(face = "bold"))
+
+# save figure
+ggsave(figure_res_patch,
+       filename = "figures/fig_calib_residence_patch.png",
+       height = 130, width = 225, units = "mm")
+```

### 2020-10-07 01_background.Rmd
remove useless sections

@@
 This paper refers extensively to the `R` package `atlastools` (Gupte 2020), which can be installed from [Github.](https://github.com/pratikunterwegs/atlastools).
-See the repository for information on the build status.
+
+Releases of the package from Github can be found on Zenodo: [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4033155.svg)](https://doi.org/10.5281/zenodo.4033155)
+
+The build status of the 'Master' branch is shown here.
+
+<!-- badges: start -->
+  [![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/pratikunterwegs/atlastools?branch=master&svg=true)](https://ci.appveyor.com/project/pratikunterwegs/atlastools) 
+  [![Build Status](https://travis-ci.org/pratikunterwegs/atlastools.svg?branch=master)](https://travis-ci.org/pratikunterwegs/atlastools) 
+  [![codecov.io](https://codecov.io/github/pratikunterwegs/atlastools/coverage.svg?branch=master)](https://codecov.io/github/pratikunterwegs/atlastools/branch/master)
+<!-- badges: end -->
+
 The code chunk below shows how to install `atlastools`.
@@
 
-```{r prep_libs_01_02}
+First, we load `smoove` and `data.table`, as well as some helper functions that make use of them to simulate data for use.
+
+```{r prep_libs_01_02, eval=FALSE}
 # load smoove and datatable
@@
 library(data.table)
+
+# source helper functions
+source("R/helper_functions.R")
 ```
@@
 data <- do_smoove_data()
-
 # save simulated data
@@
 ```
-
-Here, we write functions to add normally distributed errors to the position coordinates of the simulated data.
-
-Example of introduced error.
-
-```{r}
-# load data
-data <- fread("data/data_sim.csv")
-
-# introduce errors to data
-data[, `:=`(x_error = do_add_error(x, std_dev = 0.01),
-            y_error = do_add_error(y, std_dev = 0.01))]
-
-# plot real and introduced error
-ggplot()+
-  geom_path(data = data, aes(x, y),
-            col = "blue")
-
-atlastools::atl_compare_tracks(d2, data)
-```

### 2020-10-07 02_filtering_data.Rmd
some code edits
add some text

@@
 
-# Filtering data
+# Reducing Large-Scale Positioning Error by Filtering
 
 ## Prepare libraries {-}
+
+Here we load some useful libraries, and the helper functions.
 
@@
 library(atlastools)
-
-# to simulate movement
-library(smoove)
-
-# to generate a landscape
-library(raster)
-library(NLMR)
-
-# strings
-library(glue)
 
@@
 
-## Introducing errors {-}
+## Introducing Errors to the Data
+
+Here we introduce three kinds of errors to the data:
+
+1. Small-scale normally distributed errors at each position;
+
+2. Large-scale error at a random 0.5% of positions;
+
+3. A large-scale displacement of a sequence of 300 positions.
+
+While the data are 10,000 positions at 1-second interval, we shall use only 5,000 of these.
 
@@
 ```{r add_outlier}
-# make a dopy
+# make a copy
 data_copy <- copy(data)
-# add a reflection
+
+# add a prolonged spike or reflection to 300 positions
 data_copy[500:800, `:=`(x = x + 0.25,
                         y = y + 0.25)]
+
 # add normal error
@@
                  y = do_add_error(y, std_dev = 0.005))]
+
 # add 100 outliers
@@
 
+Save the data to which errors have been added.
+
 ```{r}
-# define red and blue
+fwrite(data_copy, file = "data/data_errors.csv")
+```
+
+Define a palette with 4 colours for convenience.
+
+```{r}
+# define a four colour palette
 pal <- RColorBrewer::brewer.pal(4, "Set1")
@@
 
-## Show data with errors
+## Make Figure of Simulated Data
+
+We make a figure of the canonical data (grey line) along with the artificially added error (grey points).
 
 ```{r}
-# make figure
+# make figure of canonical data with added errors
 figure_raw <-
@@
 
-First we remove gross positioning errors using a bounding box filter, i.e., knowing or assuming the real bounds of the movement track (in this case, 0 -- 1), we remove all positions outside those bounds.
+First we remove gross positioning errors using a bounding box filter using the function `atl_filter_bounds`. In this example, we show filtering only on the Y coordinate.
+`atl_filter_bounds` takes coordinate ranges as two-element vectors of the lower and higher bound.
+It is possible to pass one the bounds as `NA`, in which case, only the other bound is used for filtering, i.e., `y_range = c(NA, 1)` is equivalent to selecting all Y coordinates < 1.
+
+`atl_filter_bounds` was initially designed to remove positions _inside_ a specific range, hence the argument `remove_inside`.
+The default value of the argument is `FALSE`, and `atl_filter_bounds` is thus a bounding box filter.
 
@@
 ```
+
+`atl_filter_bounds` is not vectorised, and if there are two or more bounds per coordinate (for instance, $x_1 \ldots x_2$, and $x_3_ \ldots x_4$), they must be passed in two different function calls.
 

### 2020-10-07 02_filtering_data.Rmd
add text to spikes and reflections sections

@@
 
-# Reducing Large-Scale Positioning Error by Filtering
+# Reducing Large-Scale Positioning Error by Filtering Data
 
-## Prepare libraries {-}
+## Prepare Libraries
 
@@
 
-```{r prep_libs_02_01}
+```{r}
 # to handle movement data
@@
 
-## Make Figure of Simulated Data
-
 We make a figure of the canonical data (grey line) along with the artificially added error (grey points).
@@
 
-## Filtering by spatial bounds
+## Filtering by Spatial Bounds
 
-First we remove gross positioning errors using a bounding box filter using the function `atl_filter_bounds`. In this example, we show filtering only on the Y coordinate.
+Filtering by spatial bounds is a good way to begin reducing gross, large-scale positioning errors.
+There are two main ways of doing this:
+
+1. Filtering by a bounding box: Compare the coordinates of observations against a range of acceptable coordinates, and retain those which fall within the range, or
+
+2. Filtering by a spatial polygon: An explicit geometric intersection between the positions and a polygon representing an area of interest.
+
+Here, we remove gross positioning errors using a bounding box filter using the function `atl_filter_bounds`. In this example, we show filtering only on the Y coordinate.
+
 `atl_filter_bounds` takes coordinate ranges as two-element vectors of the lower and higher bound.
-It is possible to pass one the bounds as `NA`, in which case, only the other bound is used for filtering, i.e., `y_range = c(NA, 1)` is equivalent to selecting all Y coordinates < 1.
+It is possible to pass one of the bounds as `NA`, in which case, only the other bound is used for filtering, i.e., `y_range = c(NA, 1)` is equivalent to selecting all Y coordinates < 1.
 
@@
 
-`atl_filter_bounds` is not vectorised, and if there are two or more bounds per coordinate (for instance, $x_1 \ldots x_2$, and $x_3_ \ldots x_4$), they must be passed in two different function calls.
+`atl_filter_bounds` is not vectorised, and if there are two or more bounds per coordinate (for instance, $X_1 \ldots X_2$, and $X_3 \ldots X_4$), they must be passed in two different function calls.
+The same is true for filtering by an `sf` polygon. `atl_filter_bounds` also supports filtering by multi-polygon objects.
+
+Having filtered the data we shall prepare it for plotting.
 
@@
 
-plot_figure
-
 ggsave(plot_figure, filename = "figures/fig_raw_bounds.png", 
@@
 
-## Removing point outliers: The non-movement approach
+![](figures/fig_raw_bounds.png)
+
+**(a)** A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the `smoove` package (Gurarie et al.
+2017).
+Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track.
+The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (solid black line) as closely as possible.
+**(b)** Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the `atlastools` function  `atl_filter_bounds`.
+Setting the function argument `remove_inside = FALSE` retains positions within user supplied bounds (blue points), and excludes those outside (red points).
+
+### A Note on Filtering by Bounds
+
+The filter on spatial bounds is only for demonstration, and is not applied to the data.
+
+## Filtering Unrealistic Movement
+
+Large-scale positioning error can affect both point locations as well as entire subsets of a track, leading to a track appearing to show unrealistic movement for the study species [@bjorneraas2010].
+
+The best way to examine whether a track contains such positioning errors is to plot the data, and especially to join the dots, i.e., to connect the positions with lines rather than simply plot points.
+
+Users with extensive experience of their study system will readily recognise segments of a track where the movement appears to be unrealistic.
+
+Briefly, the two main kinds of errors are mentioned here, and named based on their appearance in a tack: those that affect single positions lend a spiky appearance to a track, and are referred to as _point outliers_ or _spikes_,  while errors affecting continuous subsets of a track cause it to appear as though it has been reflected along a plane, and are hence called _reflections_, or _prolonged spikes_.
+
+### Filtering Point Outliers or Spikes
+
+We begin by removing unrealistic movement in the form of point outliers, or spikes.
+
+The first step is to determine how spikes should be identified.
+The non-movement approach prescribes determining whether movement metrics associated with each position are realistic or not, and targeting those positions where the movement metrics are unrealistic for the study species.
+
+In this case we shall calculate only two metrics, speed and turning angle.
+These are conveniently implemented in `atlastools` using the functions `atl_get_speed` and `atl_turning_angle`.
 
@@
                  angle = atl_turning_angle(data_copy))]
-
 ```
 
-Next we idenfify the 90th and 95th percentile of speed and turning angle.
-This is more general than identifying the limits of implausibility for each individual (since there may be inter-individual differences), let alone each species in a large dataset.
+Having calculated speed and turning angle, the next step is to remove positions with extremely high incoming and outgoing speeds.
+This means using the `atl_filter_covariates` function to remove positions with speed $\geq$ a plausible speed threshold.
+The use of a turning angle filter is optional, and not necessary in this case.
 
-An alternative approach is to define a speed cutoff based on expert knowledge, and this is best suited to well studied species.
+One approach is to define a speed cutoff based on expert knowledge, and this is best suited to well studied species.
+
+For simulated data, there is no plausible speed threshold from prior knowledge --- a reasonable choice here is to use the 90^th or 95^th percentile of speed and turning angle (when this metric is used).
+
+This is also more general than identifying the limits of implausibility for each individual (since there may be inter-individual differences), let alone each species in a large dataset.
 
@@
 
-Finally, we remove positions whose incoming and outgoing speeds are both greater than the 95th speed percentile.
+Finally, we shall remove positions whose incoming and outgoing speeds are both greater than the 95th speed percentile using `atl_filter_covariates`.
 
@@
 ```
+
+We prepare the data for plotting.
 
@@
 
-## Reflections: Challenges to the non-movement approach
+### Filtering Reflections or Track Subsets
+
+When entire track subsets are affected by positioning error, they are more difficult to remove.
+This is the case when positions are _reflected_ along a plane in the coordinate axis ---  the simplest way in which this can be explained is to examine the way in which we simulated the reflection earlier in this text, by simply adding an offset to the X and Y coordinates of each of 300 consecutive positions (indices 500 -- 800).
+
+An explanation of why these reflections occur is found in WATLAS PAPER: Bijleveld et al.
+in prep.
+In short, they are results of an error in the ATLAS localisation algorithm, and are usually removed in the localisation step's quality control procedures.
+However, some reflections may remain to confront users.
+
+Reflections and other issues affecting track subsets cannot be resolved by targeting single positions, since there are rarely any position-specific covariates that can be used to identify a position as part of a larger subset that should be removed.
+
+However, it is often possible to identify the _bounds_ of problematic subsets, such as reflections, and remove positions between them.
+The way this is done is conceptually similar to the point outlier algorithm.
+
+The `atl_remove_reflection` function implements one method remove reflected subsets of a track. The working is described:
+
+1. Remove point outliers,
+
+2. Re-calculate speed and turning angles
+
+3. Identify the first unrealistic movement position (fast speed and high turning angle)
+
+4. Setting this point as an anchor, identify the next position with unrealistic movement (as above)
+
+5. Remove all positions between these two points,
+
+6. Search for the next unrealistic movement, and repeat the process.
+
+This is a minimal algorithm which can be developed further, and iterated multiple times to comprehensively remove reflections.
+
+One important function argument is the estimated reflection length, i.e., how many positions are estimated to be reflected.
+This argument controls how many positions after the anchor are candidates for the reflection's end.
+
+If the reflection does not end within this number of positions, the algorithm goes awry, and considers part of the reflection to be the valid data.
+Fortunately, this can be avoided by setting the length to the number of observations in the data.
+
+However, if the reflection does not end at all, the algorithm will simply delete all positions from the anchor point onwards.
+
+Thus, the algorithm is less of a double-edged sword and more a sword with no hilt --- there are few safe ways to use it.
+
+The difficulty in implementing a reasonable algorithm for dealing with this kind of track-subset wide positioning error reveals the general problem that algorithms are **not** easy to conceptualise or efficiently implement.
+
+Users are cautioned to use this and simi
 
@@
 
-## Export data
+![](figures/figure_correct_tracks.png)
+
+Finally, we export the data without spikes and reflections to be used later.
 

### 2020-10-08 02_filtering_data.Rmd
add fig caption

@@
 
-Users are cautioned to use this and simi
-
 ```{r}
@@
 
-![](figures/figure_correct_tracks.png)
+![](figures/fig_correct_tracks.png)
+
+Reducing large-scale positioning error in a movement track.
+**(a)** Positioning error may affect single observations resulting in point outliers or 'spikes' (red points), but it may also affect continuous susbets of a track, which we call a `prolonged spike' (purple triangles).
+While the former may be targeted by filtering on appropriate covariates such as speed and turning angle using the `atlastools` function `atl_filter_covariates`, the latter cannot be effectively corrected by targeting each coordinate pair in isolation.
+**(b)** The `atlastools` function `atl_remove_reflections` to identify prolonged spikes (red triangles) in tracking data is an illustrative example of targeting positioning errors that affect track subsets.
+While this method returns the canonical track without the prolonged spike (blue points) in this example, users are cautioned to frequently check this and similar semi-supervised algorithms' results.
 

### 2020-10-08 03_smoothing_tracks.Rmd
add text to spikes and reflections sections
add text

@@
 
-# Smoothing tracks and aggregating dat
+# Reducing Small-Scale Positioning Error and Aggregating Data
 
-## Prepare libraries {-}
+Once large-scale positioning errors have been removed using filters on bounds or on unrealistic movement, small-scale positioning errors may remain. Why is this the case?
 
-```{r prep_libs_02_01}
+The simple reason is that high-frequency tracking has a large component of error relative to the real movement of an individual, and the error / movement ratio increases with the frequency of tracking.
+[@noonan2019] has a better explanation, which users are strongly recommended to read.
+
+The effect of these small-scale errors for high-frequency data is to cause overestimations of straight-line displacement (distance between two positions) and speed.
+If users were to apply a very strict speed filter, many of these positions would be lost.
+This loss of information would certainly reduce the sampling frequency, which would lead to better estimates of total distance and mean speed.
+However, while _total_ distance and _mean_ speed estimates would improve, the instantaneous distance and speed estimates would remain biased, which will most likely lead to erroneous inferences from the data.
+
+However, users can turn the high-frequency of their data against itself by applying a simple smooth. The median smooth is recommended because it is robust to outliers, and is very well understood.
+
+## Preparing Libraries and Data
+
+We first load some useful packages and define a colour palette.
+
+```{r}
 # prep libs
@@
 
-## Reading in data {-}
+We load both the simulated, canonical data (which has no errors), as well as the data with errors which has been pre-processed to remove large-scale errors.
 
-Here we read the simulated data in.
-The limits of the coordinates are 0 and 1, so we add a few extraneous coordinates within a plausible distance to demonstrate the basic forms of filtering.
-We also assign a fictional _NBS_ value ranging between 2 and 6 to demonstrate filtering on covariates.
-Finally, we introduce a 'reflection', a phenomenon that results when a TOA system's position likelihood estimator shifts the real position by a large distance for an extended duration.
+We also assign a variable, `window_size`, which will help identify the datasets when comparing median smooth moving window ($K$) sizes.
 
 ```{r read_sim_data_2}
-# read in the data
+# read in the data and set the window size variable
 data <- fread("data/data_sim.csv")[5000:10000, ]
 data[, window_size := NA]
+
 # data with small scale errors but no reflections or outliers
@@
 
-## Applying a median smooth
+## Median Smoothing
+
+We apply a median smooth with four different values of $K$, the moving window size --- 3, 5, 11, and 21 positions.
+$K$ must be an odd number, but apart from that there is no correct magnitude of $K$.
+Very large $K$ will lead to unrealistic tracks, while small $K$ will not result in much reduction of error.
+Users are encouraged to plot their data before and after smoothing to examine the effect of different window sizes.
+
+The `atlastools` function `atl_median_smooth` is quite fast and users can readily try multiple $K$ values as shown in the example below.
 
 ```{r}
-# smooth the data
-list_of_smooths <- lapply(c(3, 5, 11, 21), function(z) {
+# smooth the data over four K values
+list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
+
   data_copy <- copy(data_errors)
+
   data_copy <- atl_median_smooth(data = data_copy,
@@
                     time = "time",
-                    moving_window = z)
-  data_copy[, window_size := z]
+                    moving_window = K)
+
+  data_copy[, window_size := K]
 })
@@
 
-### Robustness to large gaps
+### Robustness to Large Gaps
+
+The median smooth is robust to large gaps in the data, and tracks do not need to be split when a large chunk of data are missing.
+The easiest way to illustrate this is with a figure.
 
 ```{r}
-# make a single list
+# make a single list of data
 data_plot <- append(list(data_errors), list_of_smooths)
 
-# now prep gap
+# now prep smooth over Y coordinate
 fig_smooth_1d <-
@@
 
-### Plot results
+We prepare a plot of the smoothed data as well as the demonstration of the method relative to gaps.
 
@@
 
-fig_median_smooth
-
 # save figure
@@
 
-## Aggregating data
+## Thinning Data by Aggregation
+
+Evenly thinning data is a good idea if statistical methods require even sampling, or if the volume of data is too large for statistical packages to efficiently handle it.
+In `R`, both may be true at once.
+
+Here, we demonstrate thinning by aggregation on data that has been median smoothed using a $K$ of 11.
 
 ```{r}
-# choose the data
+# choose the 11 point median smooth data
 data_agg <- copy(list_of_smooths[[3]])
@@
 list_of_agg <- lapply(c(3, 10, 30, 120), function(z) {
+
   data_return <- atl_thin_data(data = data_agg,
@@
                             method = "aggregate")
+
   data_return[, interval := z]
+
   return(data_return)
@@
 
+We shall prepare a figure showing the aggregation steps.
+
 ```{r}
-# make figure
+# make figure of aggregation
 figure_aggregate <- 
@@
 
-figure_aggregate <- wrap_plots(figure_aggregate)
+figure_aggregate <- wrap_plots(figure_aggregate)+
+    plot_annotation(tag_levels = "a",
+                  tag_prefix = "(",
+                  tag_suffix = ")") &
+    theme(plot.tag = element_text(face = "bold"))
 
-figure_aggregate
 # save figure
-ggsave(plot_figure, filename = "figures/fig_aggregate_data.png",
+ggsave(figure_aggregate, 
+       filename = "figures/fig_aggregate_data.png",
        width = 185, height = 90, units = "mm")
@@
 
-### Aggregation before correction, a warning
+## Aggregation Before Reducing Positioning Errors
+
+Users may rightly wonder whether they can get away with aggregating their data, using a median aggregation function, and reduce data volumes, correct uneven sampling frequency, and reduce large-scale errors all in one go.
+
+The answer to most questions in ecology is, "It depends".
+Median aggregation before correcting positioning errors can indeed be advantageous; for instance for faster visualisation while preserving the broad structure of a track.
+
+However, there are drawbacks. The main one is that information is lost in the aggregation process, lending less power to steps such as smoothing applied after aggregation.
+Further, aggregation (and indeed any kind of thinning) results in significantly different estimates of speed and distance from the real speed.
+
+We show the effect of aggregating before any error correction here.
 
@@
 data[, speed := atl_get_speed(data)]
+```
 
+We prepare the figures.
+
+```{r}
 ### plot figures
@@
 ```{r}
-# now plot hist of speed
+# now plot distribution of speed
 data_agg <- rbindlist(list_of_agg)
@@
 
-### Combine with aggregations
-
 ```{r}
@@
 ggsave(fig_aggregate,
-       filename = "figures/fig_aggregate_data.png",
+       filename = "figures/fig_aggregate_errors.png",
        width = 170, height = 170 / 1.67, units = "mm")
 ```
+
+![](figures/fig_aggregate_errors.png)
+
+Thinning a movement track using median aggregation preserves track structure, but affects essential track metrics such as speed.
+**(a, b)** Movement tracks with a canonical interval of 1s aggregated over intervals of **(a)** 10 and **(b)** 30 seconds, without removing large- or small-scale positioning errors.
+All symbols represent positions in the aggregated track, with the size of the symbol representing the standard deviation at each position.
+Blue crosses represent positions with speed $\leq$ the 95^th percentile of canonical speeds, while red triangles represent positions with speed $\geq$ 95^th percentile of canonical speeds.
+**(c)** Boxplot of instantaneous speeds after median aggregation of a 1s interval track over intervals of 3, 10, 30, and 120 seconds, but without the removal of positioning errors.
+The mean and 95^th percentile of speed in the canonical track are shown as solid and dashed lines, respectively.
+Aggregation without reducing positioning errors can result in speed estimates that are substantially different from the true speed.

### 2020-10-08 index.rmd
add text to spikes and reflections sections
add to index page

--- 2020-07-23 index.rmd
@@
 knit: "bookdown::render_book"
-title: "Pre-processing high frequency animal tracking data"
+title: "A Rough Guide to Pre-processing High-Frequency Animal Tracking Data"
 author: "Pratik R. Gupte and others"
@@
 github-repo: pratikunterwegs/atlas-best-practices
-description: "TRES Tidyverse Tutorial"
+description: "Source code for a guide to pre-processing high-frequency tracking data."
 ---
 
-# Outline {-}
+# Abstract
 
-An outline here.
+Data cleaning is a ubiquitous pre-processing step in analysis pipelines, and its automation is essential for large data volumes such as those generated in animal tracking studies using high-frequency Time-of-Arrival (TOA) systems.
+Users of systems such as ATLAS must contend with two intertwined data cleaning challenges: (1) reducing positioning errors, and (2) the high volume of data itself.
+Making biological inferences from data with positioning errors is not straightforward, and processing large data volumes is computationally intensive.
+Though reducing positioning error is widely recommended, users are without uniform guidance on how to go about this, and lack a common set of computationally efficient tools.
+Further, many methods that synthesize movement tracks for ecological inference are either (1) not suited to very large datasets, or (2) not intuitive to understand in terms of the tracked animal's biology.
+
+In this article we introduce a pipeline to pre-process high-frequency animal tracking data in order to prepare it for subsequent analysis.
+We demonstrate this pipeline on simulated movement data to which we have randomly added positioning errors.
+This pipeline is suited to any tracking study in which the high data volume combined with knowledge of the tracked individuals’ biology can be used to smooth out positioning errors. 
+We further suggest how large volumes of cleaned data may be synthesized into biologically meaningful ‘residence patches’, and demonstrate how this accurately captures animal space-use.
+Finally, we introduce the `R` package `atlastools` which provides fast implementations of the methods we describe. 
+Though aimed at ATLAS systems, `atlastools` can be used with any time-series animal movement data, and we demonstrate its usage with both simulated and empirical examples.
+
+![](figures/fig_recipe.png)

### 2020-10-08 index.rmd
add bibliography

@@
 documentclass: book
-# bibliography: [refs.bib]
+bibliography: [references.bib]
 biblio-style: apalike
+link-citations: yes
 link-citations: yes

### 2020-10-08 01_background.Rmd

@@
 
-This paper refers extensively to the `R` package `atlastools` (Gupte 2020), which can be installed from [Github.](https://github.com/pratikunterwegs/atlastools).
+This paper refers extensively to the `R` package `atlastools` (Gupte 2020), which can be installed from [Github](https://github.com/pratikunterwegs/atlastools).
 
@@
 
-The build status of the 'Master' branch is shown here.
+The build status of the 'Master' branch is shown here:
 

### 2020-10-08 05_calibration_data.Rmd

@@
 data_raw <- copy(data)
+
 # plot data
@@
 
+### Calculate residence time
+
+```{r}
+# load recurse
+library(recurse)
+
+# get recurse data for a 10m radius
+data_recurse <- getRecursions(data_unproc[, list(x, y, time, TAG)],
+                              radius = 200)
+```
+
 ### Run residence patch method
@@
 # get linear model
-model_duration <- lm(duration ~ duration_real,
+model_duration <- lm(duration_real ~ duration,
                      data = patch_summary_compare)
@@
                aes(x = 100,
-                   xend = duration_real,
-                   y = duration,
-                   yend = duration),
+                   xend = duration,
+                   y = duration_real,
+                   yend = duration_real),
                col = "grey")+
   geom_segment(data = patch_summary_compare,
-               aes(x = duration_real,
-                   xend = duration_real,
+               aes(x = duration,
+                   xend = duration,
                    y = 100,
-                   yend = duration),
+                   yend = duration_real),
                col = "grey")+
   geom_point(data = patch_summary_compare,
-             aes(duration_real, 
-                 duration),
+             aes(duration, 
+                 duration_real),
              show.legend = F,
@@
   geom_smooth(data = patch_summary_compare,
-             aes(duration_real, 
-                 duration),
+             aes(duration, 
+                 duration_real),
              se = F,
@@
            size = 5,
-           label = "R^2 == 0.845",
+           label = "R^2 == 0.849",
            parse = T)+
@@
   ggthemes::theme_few()+
-  labs(x = "duration (min)",
-       y = "inferred duration (min)")
+  labs(x = "inferred duration (min)",
+       y = "real duration (min)")
 ```

### 2020-10-08 index.rmd

--- 2020-10-08 index.rmd
@@
 documentclass: book
-# bibliography: [refs.bib]
+bibliography: [references.bib]
 biblio-style: apalike

### 2020-10-08 02_filtering_data.Rmd
make edits to colours etc

@@
 # define a four colour palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
+pal <- RColorBrewer::brewer.pal(3, "PRGn")
 ```
@@
             alpha = 1)+
-  ggthemes::theme_map()+
-  coord_equal()+
+  ggthemes::theme_few()+
   theme(axis.text = element_blank(),
         axis.title = element_blank(),
-        legend.position = "none")+
-    theme(plot.background = element_rect(fill = NA))+
+        panel.background = element_rect(fill = "grey99"))+
+  coord_equal()+
   labs(colour = NULL)
@@
              aes(x, y),
-             col = pal[2],
+             col = pal[3],
              alpha = 1, size = 0.2)+
@@
              aes(x, y),
-             col = pal[1],
+             col = "grey",
              alpha = 0.5, size = 0.2)+
@@
              lty = 2)+
-  ggthemes::theme_map()+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))+
   theme(plot.background = element_rect(fill = NA))+
-  coord_equal()
+  coord_equal(expand = T)
 
@@
              aes(x, y),
-             shape = 2, col = pal[4])+
+             shape = 2, col = pal[1])+
   geom_point(data = data_copy[!data_copy[500:800, ],
@@
             alpha = 1)+
-  geom_path(data = data,
-            aes(x, y),
-            col = "grey20",
-            alpha = 1)+
-  ggthemes::theme_map()+
-  scale_color_brewer(palette = "Set1", direction = -1)+
-  scale_shape_manual(values = c(1, 19))+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))+
+  scale_color_manual(values = c(pal[3], "black"))+
+  scale_shape_manual(values = c(16, 21))+
   coord_equal()+
@@
 # attempt to remove reflections
-data_no_reflection <- atl_remove_reflections(data_with_reflection,
+data_no_reflection <- atl_remove_reflections(data_filtered,
                           point_angle_cutoff = 10,
@@
 # get reflections
-reflection <- data_with_reflection[!data_no_reflection,
+reflection <- data_filtered[!data_no_reflection,
                                          on = c("x", "y")]
@@
              alpha = 1,
-             col = "grey")+
+             col = "grey",
+             lwd = 0.2)+
   geom_point(data = reflection,
@@
              alpha = 0.5,
-             col = pal[1],
-             shape = 2)+
+             col = "grey",
+             fill = pal[2],
+             shape = 21)+
   geom_point(data = data_no_reflection,
@@
              alpha = 0.5,
-             colour = pal[2],
-             shape = 1,
+             colour = pal[3],
+             shape = 16,
              show.legend = F)+
@@
             alpha = 1)+
-  ggthemes::theme_map()+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))+
   coord_equal()+
@@
   theme(plot.tag = element_text(face = "bold"))
-
-plot_figure
 

### 2020-10-08 03_smoothing_tracks.Rmd
make edits to colours etc

@@
 # prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
+pal <- RColorBrewer::brewer.pal(3, "PRGn")
 ```
@@
 ```{r}
-# make a single list of data
-data_plot <- append(list(data_errors), list_of_smooths)
+# bind list after offset
+data_plot <- mapply(function(df, offset) {
+  df[, x := x + offset]
+}, list_of_smooths, seq(0.4, 1.25, length.out = 4),
+SIMPLIFY = F)
 
-# now prep smooth over Y coordinate
-fig_smooth_1d <-
-  ggplot()+
-  geom_point(data = rbindlist(data_plot),
-             aes(time, y,
-                 col = factor(window_size)),
-             alpha = 0.5,
-             size = 1,
-             show.legend = FALSE,
-             shape = 16)+
-  coord_cartesian(xlim = c(5000, 7000),
-                  ylim = c(0.7, NA),
-                  expand = F)+
-  scale_colour_brewer(palette = "Spectral")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank())+
-  labs(x = "position index", y = "Y coordinate")
+data_plot <- rbindlist(data_plot)
 ```
 
-We prepare a plot of the smoothed data as well as the demonstration of the method relative to gaps.
+We prepare a plot of the smoothed data.
 
@@
 figure_median_smooth <-
-  mapply(function(df, col) {
-    ggplot()+
-      geom_point(data = df,
-                 aes(x, y),
-                 size = 0.5,
-                 alpha = 0.5,
-                 show.legend = FALSE,
-                 col = col,
-                 shape = 1)+
-      geom_path(data = data,
-                aes(x, y),
-                col = "grey20")+
-      coord_cartesian(expand = T,
-                  ylim = c(0.6, 0.8)) +
-      ggthemes::theme_map()
-}, data_plot, c(pal[2], rep(
-  RColorBrewer::brewer.pal(3, "Greens")[2],
-  4)),
-SIMPLIFY = F)
-
-# make basic plot with rectangle
-fig_bounds <-
   ggplot()+
@@
              aes(x, y),
-            size = 0.1, alpha = 0.2, 
-            col = "grey")+
+             col = pal[3],
+             size = 0.2, alpha = 0.5)+
   geom_path(data = data,
             aes(x, y),
-            size = 0.1)+
-  geom_hline(yintercept = c(0.6, 0.8),
             col = "grey20",
-             lty = 2, lwd = 0.2)+
-  coord_equal(expand = F) +
-  ggthemes::theme_map()
+            lwd = 0.5)+          
+  geom_path(data = data_plot,
+            aes(x, y,
+                col = window_size,
+                group = window_size),
+            show.legend = F,
+            lwd = 0.5)+
+  coord_equal(expand = F,
+              ylim = c(0.6, 0.85),
+              xlim = c(NA, 2.3),
+              ratio = 1.75)+
+  annotate(geom = "text",
+           x = c(0.75, seq(1.1, 2, length.out = 4)),
+           y = 0.82,
+           label = sprintf("(%s)", letters[seq(5)]),
+           fontface = "bold")+
+  scale_colour_distiller(palette = "BuPu", direction = 1,
+                         values = c(-0.5, 1))+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))
 
-# append to first figure
-figure_median_smooth[[1]] <-
-  figure_median_smooth[[1]]+
-  annotation_custom(grob = ggplotGrob(fig_bounds),
-                    ymin = 0.6, ymax = 0.8,
-                    xmin = 0.5, xmax = 0.7)
+# # make basic plot with rectangle
+# fig_bounds <-
+#   ggplot()+
+#   geom_point(data = data_errors,
+#             aes(x, y),
+#             size = 0.1, alpha = 0.2, 
+#             col = "grey")+
+#   geom_path(data = data,
+#             aes(x, y),
+#             size = 0.1)+
+#   geom_hline(yintercept = c(0.6, 0.8),
+#              col = "grey20",
+#              lty = 2, lwd = 0.2)+
+#   coord_equal(expand = F) +
+#   ggthemes::theme_map()
+
+# # append to first figure
+# figure_median_smooth[[1]] <-
+#   figure_median_smooth[[1]]+
+#   annotation_custom(grob = ggplotGrob(fig_bounds),
+#                     ymin = 0.6, ymax = 0.8,
+#                     xmin = 0.5, xmax = 0.7)
 
 # wrap plots
-fig_median_smooth <- wrap_plots(append(figure_median_smooth, 
-                                       list(fig_smooth_1d)), 
-                                       ncol = 3)+
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")") &
-  theme(plot.tag = element_text(face = "bold"))
+# fig_median_smooth <- 
+  # wrap_plots(figure_median_smooth, 
+  #            ncol = 3)+
+  # plot_annotation(tag_levels = "a",
+  #                 tag_prefix = "(",
+  #                 tag_suffix = ")") &
+  # theme(plot.tag = element_text(face = "bold"))
 
 # save figure
-ggsave(fig_median_smooth, filename = "figures/fig_median_smooth.png",
-       width = 225, height = 225 / 1.77, units = "mm")
+ggsave(figure_median_smooth, filename = "figures/fig_median_smooth.png",
+       width = 170, height = 170 / 3, units = "mm")
 ```
@@
 })
-```
-
-We shall prepare a figure showing the aggregation steps.
-
-```{r}
-# make figure of aggregation
-figure_aggregate <- 
-  mapply(FUN = function(df) {
-  ggplot(data = df)+
-      geom_point(data = list_of_smooths[[3]],
-                 aes(x, y),
-                 col = pal[3],
-                 size = 0.3, alpha = 0.5)+
-      geom_point(aes(x, y, size = SD),
-                 colour = pal[4],
-                 alpha = 1,
-                 stroke = 1,
-                 shape = 4, show.legend = F)+
-      geom_path(aes(x, y),
-                 colour = pal[4],
-                 alpha = 1)+
-      geom_path(data = data,
-                lwd = 0.2,
-                aes(x, y),
-                col = "grey20")+
-      coord_equal(ylim = c(0.6, 0.8))+
-      ggthemes::theme_few()+
-      theme(axis.title = element_blank(),
-            axis.text = element_blank())
-}, list_of_agg[2:3],
-SIMPLIFY = FALSE)
-
-figure_aggregate <- wrap_plots(figure_aggregate)+
-    plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")") &
-    theme(plot.tag = element_text(face = "bold"))
-
-# save figure
-ggsave(figure_aggregate, 
-       filename = "figures/fig_aggregate_data.png",
-       width = 185, height = 90, units = "mm")
 ```
@@
     ggplot(df)+
+      geom_point(data = data_errors,
+                 aes(x, y),
+                 size = 0.2,
+                 col = pal[3])+
+            geom_path(data = data,
+                aes(x, y),
+                lwd = 0.3)+
       geom_path(aes(x,y), 
-                # col = factor(interval), 
-                # group = interval),
+                col = pal[1],
                 lwd = 0.2)+
@@
                      size = SD,
-                     shape = speed < quantile(data$speed, 0.95, na.rm = T),
-                     col = speed < quantile(data$speed, 0.95, na.rm = T),
+                     fill = speed,
                      group = interval),
-                 stroke = ifelse(df$speed < quantile(data$speed, 0.95, na.rm = T),
-                                 1, 1),
+                 shape = 16,
                  show.legend = F)+
-      geom_path(data = data,
-                aes(x, y),
-                lwd = 0.1)+
-      scale_shape_manual(values = c(2, 4))+
-      scale_colour_manual(values = pal[c(1, 2)])+
-      coord_cartesian(ylim = c(0.6, NA))+
-      ggthemes::theme_map()  
+      scale_fill_distiller(palette = "Purples", direction = 1,
+                             values = c(-0.25, 1))+
+        ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))+
+      coord_cartesian(ylim = c(0.6, NA))
   })

### 2020-10-08 03_smoothing_tracks.Rmd
better aggregate track

@@
       geom_path(aes(x,y), 
-                col = pal[1],
-                lwd = 0.2)+
+                col = pal[1])+
       geom_point(aes(x,y, 
                      size = SD,
-                     fill = speed,
+                     fill = SD,
                      group = interval),
-                 shape = 16,
+                 shape = 21,
+                 alpha = 0.8,
                  show.legend = F)+
-      scale_fill_distiller(palette = "Purples", direction = 1,
+      scale_fill_distiller(palette = "RdPu", direction = 1,
                            values = c(-0.25, 1))+
+      scale_size(range = c(2, 4))+
       ggthemes::theme_few()+
@@
 # show boxplot of speed
-fig_agg_speed <- ggplot(data_agg)+
+fig_agg_speed <-
+  ggplot(data_agg)+
   geom_jitter(aes(factor(interval), speed),
-              col = "grey",
+              col = pal[1],
               shape = 4, alpha = 0.5)+
@@
 fig_aggregate <-
-  wrap_plots(append(fig_agg_data[2:3], list(fig_agg_speed)),
-           design = "AAABBBCC")+
+  wrap_plots(append(fig_agg_data[3:4], list(fig_agg_speed)),
+           design = "AABBC")+
   plot_annotation(tag_levels = "a",
@@
        filename = "figures/fig_aggregate_errors.png",
-       width = 170, height = 170 / 1.67, units = "mm")
+       width = 170, height = 85, units = "mm")
 ```

### 2020-10-08 03_smoothing_tracks.Rmd
minor edit

@@
 
-### Robustness to Large Gaps
+We save the 11 point smoothed data.
 
-The median smooth is robust to large gaps in the data, and tracks do not need to be split when a large chunk of data are missing.
-The easiest way to illustrate this is with a figure.
+```{r}
+fwrite(list_of_smooths[[3]], file = "data/data_smooth.csv")
+```
+
+We prepare the data for plotting.
 

### 2020-10-09 05_calibration_data.Rmd
remove filter NBS, prep for recurse

@@
              show.legend = F)+
-  scale_color_brewer(palette = "Set1")+
+  scale_colour_manual(values = c("grey", pal[3]))+
   geom_vline(xintercept = 645000,
@@
   ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_map()+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))+
   coord_sf(crs = 32631)
@@
        width = 185 / 25)
-```
-
-### Remove low base station positions
-
-```{r}
-# save a copy
-data_unproc <- copy(data)
-
-# remove NBS <= 3
-data <- atl_filter_covariates(data = data,
-                              filters = "NBS > 3")
-```
-
-```{r}
-fig_low_nbs <-
-  ggplot()+
-  geom_path(data = data,
-            aes(x, y),
-             col = "grey", alpha = 0.5, size = 0.3)+
-  geom_point(data = data_unproc,
-            aes(x, y,
-                col = as.factor(NBS > 3)),
-            shape = 1,
-            alpha = 0.5, size = 0.3, show.legend = F)+
-  scale_color_brewer(palette = "Set1")+
-  ggthemes::theme_map()+
-  ggspatial::annotation_scale(location = "tl")+
-  coord_sf(crs = 32631)+
-  labs(colour = "NBS")
-
-# save figure
-ggsave(fig_low_nbs, filename = "figures/fig_low_nbs.png",
-       width = 185 / 25, height = 185 / 25)
 ```
@@
   ggplot()+
-  geom_path(data = data,
-            aes(x, y),
-             col = "grey", alpha = 0.5, size = 0.3)+
   geom_point(data = data,
              aes(x, y, 
-                 col = speed_in < 15.7),
+                 col = speed_in < 24),
              size = 0.3, alpha = 0.5,
              show.legend = F)+
-  scale_color_brewer(palette = "Set1")+
-  ggthemes::theme_map()+
+  scale_color_manual(values = c("grey", pal[3]))+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))+
   ggspatial::annotation_scale(location = "tl")+
@@
 # make figure
-fig_smooth <-
-  ggplot()+
-  geom_path(data = data,
-             aes(x, y),
-            col = RColorBrewer::brewer.pal(4, "Set1")[4])+
-  annotate(geom = "rect",
-           xmin = c(650785, 650045),
-           ymin = c(5904450, 5902700),
-           xmax = c(652500, 650682),
-           ymax = c(5906133, 5903080),
-           fill = "grey90",
-           alpha = 0.5,
-           col = NA)+
-    annotate(geom = "text",
-           x = c(650785, 650045) + 400,
-           y = c(5904450, 5902700) + 300,
-           col = "black",
-           label = c("d", "e"))+
-  theme_void()+
-  coord_sf(crs = 32631)+
-  labs(colour = "track")
+# fig_smooth <-
+#   ggplot()+
+#   geom_path(data = data,
+#              aes(x, y),
+#             col = RColorBrewer::brewer.pal(4, "Set1")[4])+
+#   annotate(geom = "rect",
+#            xmin = c(650785, 650045),
+#            ymin = c(5904450, 5902700),
+#            xmax = c(652500, 650682),
+#            ymax = c(5906133, 5903080),
+#            fill = "grey90",
+#            alpha = 0.5,
+#            col = NA)+
+#     annotate(geom = "text",
+#            x = c(650785, 650045) + 400,
+#            y = c(5904450, 5902700) + 300,
+#            col = "black",
+#            label = c("d", "e"))+
+#   theme_void()+
+#   coord_sf(crs = 32631)+
+#   labs(colour = "track")
 
@@
                aes(x, y),
-               col = RColorBrewer::brewer.pal(3, "Set1")[3],
-               shape = 1, 
+               col = pal[3],
+               shape = 19, 
                alpha = 0.5)+
@@
     labs(colour = "track")
-}, list(xl1 = c(650785, 652500),
+}, list(xl1 = c(650785, 653250),
         xl2 = c(650045, 650682)),
@@
 
-# add global figure to second inset
-fig_inset[[2]] <- fig_inset[[2]] +
-  annotation_custom(grob = ggplotGrob(fig_smooth),
-                    xmin = 649990,
-                    xmax = 650045 + (650682 - 650045)/2,
-                    ymin = 5903080 - (5903080 - 5902700)/2.5,
-                    ymax = 5903080
-                    )
+# # add global figure to second inset
+# fig_inset[[2]] <- fig_inset[[2]] +
+#   annotation_custom(grob = ggplotGrob(fig_smooth),
+#                     xmin = 649990,
+#                     xmax = 650045 + (650682 - 650045)/2,
+#                     ymin = 5903080 - (5903080 - 5902700)/2.5,
+#                     ymax = 5903080
+#                     )
 
-fig_median_smooth <- 
-  wrap_plots(fig_inset)+
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")")
+fig_median_smooth <- fig_inset[[1]]
+  # wrap_plots(fig_inset)+
+  # plot_annotation(tag_levels = "a",
+  #                 tag_prefix = "(",
+  #                 tag_suffix = ")")
 # save figure
-ggsave(filename = "figures/fig_calib_median_smooth.png",
-       width = 185 / 25, height = 90 / 25)
+ggsave(fig_median_smooth,
+       filename = "figures/fig_calib_median_smooth.png",
+       width = 90 / 25, height = 90 / 25)
 ```
@@
 wrap_plots(
-  append(
   list(
-      fig_data_bbox, fig_low_nbs, fig_speed_outliers), 
-    fig_inset),
-design = "ABD\nACE")+
+    fig_data_bbox, fig_speed_outliers, fig_median_smooth),
+  design = "AB\nAC")+
   plot_annotation(tag_levels = "a",
@@
        filename = "figures/fig_walkthrough.png",
-       height = 130, width = 225, units = "mm")
+       height = 130, width = 170, units = "mm")
 ```
@@
 data_recurse <- getRecursions(data_unproc[, list(x, y, time, TAG)],
-                              radius = 200)
+                              radius = 100)
 ```

### 2020-10-12 05_calibration_data.Rmd
try recurse on boating data

@@
   geom_point(aes(x, y),
-             col = "steelblue", alpha = 0.2, size = 2)+
+             col = pal[3], alpha = 0.2, size = 2)+
   ggthemes::theme_map()+
@@
 # divide by 1000, convert to integer, then convert to POSIXct
-data[, time := as.POSIXct(as.integer(TIME / 1000), 
-                          origin = "1970-01-01")]
+data[, time := as.integer(TIME / 1000)]
 ```
@@
              aes(x, y, 
-                 col = speed_in < 24),
+                 col = speed_in < 25.6),
              size = 0.3, alpha = 0.5,
@@
 ggsave(fig_speed_outliers, filename = "figures/fig_speed_outlier.png",
-       width = 185 / 25, height = 185 / 25)
+       width = 170 / 25, height = 170 / 25)
 ```
@@
 
-Here we use a speed threshold of 20 m/s, 4 m/s lower than the 95th percentile.
+Here we use a speed threshold of 25.6 m/s, a little lower than the 95th percentile.
 
@@
 data <- atl_filter_covariates(data = data,
-            filters = c("(speed_in < 15.7 & speed_out < 15.7)"))
+            filters = c("(speed_in < 15 & speed_out < 15)"))
 
@@
 data_unproc <- copy(data)
-
-# assign residence positions
-data[, residence := stringi::stri_detect(tID, regex = "WP")]
-
-# make a copy of only residence points
-data_res <- data[residence == T, ]
 ```
@@
 
+Subset for boating segments: 2020-08-24 11:13:00 -- 2020-08-24 12:36:11.
+
 ```{r}
@@
 library(recurse)
+data_recurse <- data_unproc[, list(x, y, time, TAG, UTCtime)]
+
+# filter for boating segment
+data_recurse <- atl_filter_covariates(data = data_recurse,
+  filters = c("between(UTCtime, '2020-08-24 11:13:00', '2020-08-24 12:36:11')"))
+
+# get 4 column data
+data_recurse <- data_recurse[, list(x, y, time, TAG)]
 
 # get recurse data for a 10m radius
-data_recurse <- getRecursions(data_unproc[, list(x, y, time, TAG)],
-                              radius = 100)
+recurse_stats <- getRecursions(data_recurse,
+                               radius = 500, timeunits = "mins")
+
+# assign to recurse data
+data_recurse[, res_time := recurse_stats$residenceTime]
+```
+
+### Plot residence time
+
+```{r}
+fig_calib_residence <- 
+  ggplot()+
+  geom_point(data = data_recurse,
+             aes(x, y, col = res_time),
+             shape = 19, alpha = 0.5,
+             show.legend = F)+
+  scale_colour_distiller(palette = "Greens", direction = 1,
+                         trans = "log10",
+                         labels = scales::comma,
+                         limits = c(5, NA),
+                         na.value = "grey")+
+  ggspatial::annotation_scale(location = "br")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))+
+  coord_sf(crs = 32631)
 ```
@@
 ```{r}
+# assign id as tag
+data_recurse[, id := as.character(TAG)]
+
+# subset on 5 minute residence time
+data_recurse_subset <- data_recurse[res_time >= 5, ]
+
 # on known residence points
-patch_res_known <- atl_res_patch(data_res, 
-                                buffer_radius = 10,
+patch_res_known <- atl_res_patch(data_recurse_subset, 
+                                buffer_radius = 5,
                                 lim_spat_indep = 100,
@@
   ggplot()+
-  geom_path(data = data,
-            aes(x, y),
-            lwd = 0.2,
-            col = "grey20")+
-  geom_point(data = data_res,
-             aes(x, y),
-             shape = 2,
-             size = 2,
-             col = pal[2])+
+  geom_point(data = data_recurse,
+            aes(x, y,
+                col = res_time > 5),
+            show.legend = F,
+            size = 1)+
   geom_path(data = patch_summary_data,
-            aes(x_median + 200, y_median),
-            col = "orange", lty = 1)+
+            aes(x_median, y_median),
+            col = pal[4], lty = 1)+
   geom_point(data = patch_summary_data,
-             aes(x_median + 200, y_median,
-                 fill = duration / 60),
-             size = 5,
-             shape = 21)+
-
-  scale_fill_distiller(palette = "YlOrRd",
-                       direction = 1,
-                       trans = "log10",
-                       limits = c(1, 60))+
-  ggthemes::theme_map()+
-  theme(legend.position = c(0, 0.3),
-        legend.key.width = unit(3, units = "mm"),
-        legend.background = element_blank())+
+             aes(x_median, y_median),
+             size = 10,
+             stroke = 2,
+             col = pal[4],
+             shape = 1,
+             show.legend = F)+
+  scale_colour_manual(values = c("grey", pal[3]))+
+  scale_fill_distiller(palette = "RdPu")+
   ggspatial::annotation_scale(location = "tl")+
-  coord_sf(crs = 32631)+
-  labs(fill = "inferred\nduration (min)")
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "grey99"))+
+  coord_sf(crs = 32631,
+           xlim = c(NA, 650000))
 ```
@@
 # get known patch summary
+data_res <- data_unproc[stringi::stri_detect(tID, regex = "(WP)"), ]
+# filter for boating
+data_res <- data_res[between(UTCtime, '2020-08-24 11:13:00', '2020-08-24 12:36:11'), ]
+
+# get waypoint summary
 patch_summary_real <- data_res[, list(nfixes_real = .N,
@@
 # make figure comparing different methods
-figure_patch_duration <-
+# figure_patch_duration <-
   ggplot()+
-  geom_segment(data = patch_summary_compare,
-               aes(x = 100,
-                   xend = duration,
-                   y = duration_real,
-                   yend = duration_real),
-               col = "grey")+
-  geom_segment(data = patch_summary_compare,
-               aes(x = duration,
-                   xend = duration,
-                   y = 100,
-                   yend = duration_real),
-               col = "grey")+
   geom_point(data = patch_summary_compare,
@@
   annotate(geom = "text",
-           x = 300, y = 900,
+           x = 500, y = 320,
            size = 5,
@@
            parse = T)+
-  scale_x_log10(breaks = c(300, 600, 900),
-                labels = as.integer(c(300, 600, 900) / 60))+
-  scale_y_log10(breaks = c(300, 600, 900),
-                labels = as.integer(c(300, 600, 900) / 60))+
-  coord_equal(expand = F,
-              xlim = c(200, 1200),
-              ylim = c(200, 1200))+
+  # scale_x_log10(breaks = c(300, 600, 900),
+  #               labels = as.integer(c(300, 600, 900) / 60))+
+  # scale_y_log10(breaks = c(300, 600, 900),
+  #               labels = as.integer(c(300, 600, 900) / 60))+
+  # coord_equal(expand = F)+
   ggthemes::theme_few()+

### 2020-10-12 03_smoothing_tracks.Rmd

@@
 
-# # make basic plot with rectangle
-# fig_bounds <-
-#   ggplot()+
-#   geom_point(data = data_errors,
-#             aes(x, y),
-#             size = 0.1, alpha = 0.2, 
-#             col = "grey")+
-#   geom_path(data = data,
-#             aes(x, y),
-#             size = 0.1)+
-#   geom_hline(yintercept = c(0.6, 0.8),
-#              col = "grey20",
-#              lty = 2, lwd = 0.2)+
-#   coord_equal(expand = F) +
-#   ggthemes::theme_map()
-
-# # append to first figure
-# figure_median_smooth[[1]] <-
-#   figure_median_smooth[[1]]+
-#   annotation_custom(grob = ggplotGrob(fig_bounds),
-#                     ymin = 0.6, ymax = 0.8,
-#                     xmin = 0.5, xmax = 0.7)
-
-# wrap plots
-# fig_median_smooth <- 
-  # wrap_plots(figure_median_smooth, 
-  #            ncol = 3)+
-  # plot_annotation(tag_levels = "a",
-  #                 tag_prefix = "(",
-  #                 tag_suffix = ")") &
-  # theme(plot.tag = element_text(face = "bold"))
-
 # save figure
@@
 })
+
+# get mean speed estimate and sd
+speed_agg_smooth <- 
+  lapply(list_of_agg, function(df) {
+    na.omit(df)
+    df[, speed := atl_get_speed(df)]
+    df[, list(median = median(speed, na.rm = T),
+              sd = sd(speed, na.rm = T),
+              interval = first(interval))]
+  })
+
+# bind
+speed_agg_smooth <- rbindlist(speed_agg_smooth)
 ```
@@
   ggplot(data_agg)+
-  geom_jitter(aes(factor(interval), speed),
+  geom_jitter(aes(factor(interval), 1 + speed),
               col = pal[1],
               shape = 4, alpha = 0.5)+
-  geom_boxplot(aes(factor(interval), speed),
-               fill = NA,
+
+  geom_hline(yintercept =
+               1 + quantile(data$speed, na.rm = T,
+                        probs = c(0.5, 0.95)),
+             lty = c(1, 2))+
+  geom_errorbar(data = speed_agg_smooth,
+                aes(x = factor(interval),
+                    ymin = 1 + median - sd,
+                    ymax = 1 + median + sd),
+                width = 0.2,
+                position = position_nudge(x = 0.25)) +
+  geom_point(data = speed_agg_smooth,
+             aes(x = factor(interval),
+                 y = 1 + median),
+             shape = 21,
+             size = 3,
+             fill = pal[3],
+             position = position_nudge(x = 0.25)) +
+  geom_boxplot(aes(factor(interval), 1 + speed),
+               position = position_nudge(x = -0.25,),
+               fill = pal[2],
+               alpha = 0.5,
                show.legend = F, 
-               width = 0.3, 
+               width = 0.25, 
                outlier.size = 0.2)+
-  geom_hline(yintercept =
-               c(mean(data$speed, na.rm = T),
-                 quantile(data$speed, na.rm = T,
-                          probs = c(0.95))
-               ),
-             lty = c(1, 2))+
-  scale_y_log10(label = scales::comma)+
+  scale_y_log10(label = scales::comma,
+                limits = c(NA, 1.005))+
   ggthemes::theme_few()+

### 2020-10-13 05_calibration_data.Rmd
resample data

@@
 # filter for boating segment
-data_recurse <- atl_filter_covariates(data = data_recurse,
-  filters = c("between(UTCtime, '2020-08-24 11:13:00', '2020-08-24 12:36:11')"))
+# data_recurse <- atl_filter_covariates(data = data_recurse,
+#  filters = c("between(UTCtime, '2020-08-24 11:13:00', '2020-08-24 12:36:11')"))
+
+# thin the data
+data_recurse <- atl_thin_data(data = data_recurse,
+                              interval = 30, id_columns = "TAG",
+                              method = "resample")
 
@@
 recurse_stats <- getRecursions(data_recurse,
-                               radius = 500, timeunits = "mins")
+                               radius = 100, timeunits = "mins")
 
@@
 data_recurse[, res_time := recurse_stats$residenceTime]
+
+# save recurse data
+fwrite(data_recurse, file = "data/data_calib_recurse.csv")
 ```

### 2020-10-13 05_calibration_data.Rmd
redo residence patch

@@
 recurse_stats <- getRecursions(data_recurse,
-                               radius = 100, timeunits = "mins")
+                               radius = 50, timeunits = "mins")
 
@@
 
+Read in the data.
+
+```{r}
+data_recurse <- fread("data/data_calib_recurse.csv")
+```
+
+Subset waypoint data.
+
+```{r}
+library(stringi)
+data_res <- data_unproc[stri_detect(tID, regex = "(WP)")]
+```
+
 ### Plot residence time
@@
   ggplot()+
+  geom_point(data = data_res, 
+             aes(x, y),
+             size = 5, shape = 2,
+             alpha = 0.1)+
   geom_point(data = data_recurse,
-             aes(x, y, col = res_time),
-             shape = 19, alpha = 0.5,
+             aes(x, y, 
+                 col = res_time > 5,
+                 size = res_time > 5),
+             shape = 19, alpha = 0.4,
              show.legend = F)+
-  scale_colour_distiller(palette = "Greens", direction = 1,
-                         trans = "log10",
-                         labels = scales::comma,
-                         limits = c(5, NA),
-                         na.value = "grey")+
+  geom_path(data = data_recurse,
+             aes(x, y),
+            col = "grey", lwd = 0.2)+
+  scale_colour_manual(values = c("grey40", pal[3]))+
+  scale_size_manual(values = c(0.5, 2))+
   ggspatial::annotation_scale(location = "br")+
@@
 patch_res_known <- atl_res_patch(data_recurse_subset, 
-                                buffer_radius = 5,
+                                buffer_radius = 10,
                                 lim_spat_indep = 100,
-                                lim_time_indep = 30,
+                                lim_time_indep = 5,
                                 min_fixes = 3)
@@
 # for the known and unkniwn patches
-patch_sf_data <- atl_patch_summary(patch_res_known, which_data = "spatial")
+patch_sf_data <- atl_patch_summary(patch_res_known, which_data = "spatial",
+                                   buffer_radius = 20)
 
@@
 
+### Get waypoint centroids
+
+```{r}
+# get centroid
+data_res_summary <- data_res[, list(x_median = median(x),
+                                    y_median = median(y),
+                                    t_median = median(time)),
+                             by = "tID"]
+
+# now get times 10 mins before and after
+data_res_summary[, `:=`(t_min = t_median - (10 * 60),
+                        t_max = t_median + (10 * 60))]
+
+# make a list of positions 10min before and after
+wp_data <- mapply(function(l, u, mx, my) {
+  tmp_data <- data_unproc[inrange(time, l, u)]
+  tmp_data[, distance := sqrt((mx - x)^2 + (my - y)^2)]
+  # keep within 50
+  tmp_data <- tmp_data[distance <= 50, ]
+  # get duration
+  return(diff(range(tmp_data$time)))
+}, data_res_summary$t_min, data_res_summary$t_max,
+   data_res_summary$x_median, data_res_summary$y_median,
+SIMPLIFY = FALSE)
+```
+
 ### Plot the outcome
@@
   ggplot()+
+  geom_sf(data = griend)+
+  geom_point(data = data_res_summary, 
+             aes(x_median, y_median),
+             size = 10, shape = 7,
+             # stroke = ,
+             col = "grey20",
+             alpha = 1)+
+    # geom_path(data = data_recurse,
+    #          aes(x, y),
+    #         col = "grey", lwd = 0.2)+
+  # geom_text(data = data_res_summary,
+  #           aes(x_median, y_median,
+  #               label = tID),
+  #           # size = 10, shape = 7,
+  #           # stroke = ,
+  #           nudge_x = 100,
+  #           fontface = "bold",
+  #           col = "grey20",
+  #           alpha = 1)+
+  # geom_text(data = patch_summary_data, 
+  #           aes(x_median, y_median,
+  #               label = patch),
+  #           # size = 10, shape = 7,
+  #           # stroke = ,
+  #           nudge_x = -50,
+  #           fontface = "bold",
+  #           col = pal[4],
+  #           alpha = 1)+
   geom_point(data = data_recurse,
             aes(x, y,
-                col = res_time > 5),
+                col = res_time > 5,
+                shape = res_time > 5),
             show.legend = F,
-            size = 1)+
-  geom_path(data = patch_summary_data,
-            aes(x_median, y_median),
-            col = pal[4], lty = 1)+
-  geom_point(data = patch_summary_data,
-             aes(x_median, y_median),
-             size = 10,
-             stroke = 2,
+            size = 2)+
+  geom_sf(data = patch_sf_data,
+          fill = pal[4],
           col = pal[4],
-             shape = 1,
-             show.legend = F)+
-  scale_colour_manual(values = c("grey", pal[3]))+
+          alpha = 0.3)+
+  geom_segment(data = patch_summary_data,
+            aes(x = x_end, y = y_end,
+                xend = shift(x_start, type = "lag"),
+                yend = shift(y_start, type = "lag")),
+            col = pal[4], lty = 1,
+            lwd = 0.5)+
+  # geom_point(data = patch_summary_data,
+  #            aes(x_median, y_median),
+  #            size = 5,
+  #            stroke = 1,
+  #            col = pal[4],
+  #            shape = 1,
+  #            show.legend = F)+
+  scale_colour_manual(values = c("grey40", pal[3]))+
+  scale_shape_manual(values = c(1, 19))+
   scale_fill_distiller(palette = "RdPu")+
@@
   coord_sf(crs = 32631,
-           xlim = c(NA, 650000))
+           expand = F,
+           ylim = c(5901700, 5903150),
+           xlim = c(649300, 650700))
 ```
@@
 data_res <- data_unproc[stringi::stri_detect(tID, regex = "(WP)"), ]
-# filter for boating
-data_res <- data_res[between(UTCtime, '2020-08-24 11:13:00', '2020-08-24 12:36:11'), ]
 
@@
                                       x_median = round(median(x), digits = -2),
-                                      y_median = round(median(y), digits = -2),
-                                      duration_real = as.numeric(
-                                        max(time) - min(time))), 
+                                      y_median = round(median(y), digits = -2)), 
                                by = "tID"]
 
+# add real duration
+patch_summary_real[, duration_real := unlist(wp_data)]
 # round median coordinate for inferred patches
@@
                      c("x_median", "y_median", 
-                       "nfixes", "duration")
+                       "nfixes", "duration", "patch")
                      ][, `:=`(x_median = round(x_median, digits = -2),
@@
         all.x = TRUE, all.y = TRUE)
+
+patch_summary_compare[!is.na(tID)]
+
+# drop nas
+patch_summary_compare <- na.omit(patch_summary_compare)
+
+# drop patch around WP080
+patch_summary_compare <- patch_summary_compare[tID != "WP080", ]
 ```
+
+12 patches are identified where there are no waypoints, while 2 waypoints are not identified as patches.
+These waypoints consisted of 6 and 15 (WP098 and WP092) positions respectively, and were lost when the data were aggregated to 30 second intervals.
 
@@
 # make figure comparing different methods
-# figure_patch_duration <-
+figure_patch_duration <-
   ggplot()+
@@
             size = 3,
-            stroke = 1,
-            shape = 21,
-            fill = pal[2],
-            col = "white")+
+            stroke = 2,
+            shape = 1,
+            col = pal[2])+
   geom_smooth(data = patch_summary_compare,
@@
              se = F,
-             size = 0.5,
              col = pal[1],
-             alpha = 0.2,
-             method = "glm",
+             lwd = 1,
+             method = "lm",
              show.legend = F)+
   annotate(geom = "text",
-           x = 500, y = 320,
+           x = 600, y = 1020,
            size = 5,
-           label = "R^2 == 0.849",
+           label = "R^2 == 0.98",
            parse = T)+
-  # scale_x_log10(breaks = c(300, 600, 900),
-  #               labels = as.integer(c(300, 600, 900) / 60))+
-  # scale_y_log10(breaks = c(300, 600, 900),
-  #               labels = as.integer(c(300, 600, 900) / 60))+
-  # coord_equal(expand = F)+
+  scale_x_continuous(
+    breaks = c(400, 600, 800, 1000),
+    labels = as.integer(c(400, 600, 800, 1000) / 60)
+    )+
+  scale_y_continuous(
+    breaks = c(400, 600, 800, 1000),
+    labels = as.integer(c(400, 600, 800, 1000) / 60)
+  )+
+  coord_cartesian(expand = T,
+  #             # xlim = c(400, 1050),
+  #             ylim = c(250, 1050)
+  )+
+  geom_abline(slope = 1, lty = 2,
+              col = "black")+
   ggthemes::theme_few()+
+  theme(panel.background = element_rect(fill = "grey99"),
+        panel.grid.minor = element_line(colour = "grey"))+
   labs(x = "inferred duration (min)",
@@
 figure_res_patch <-
-  wrap_plots(list(fig_basic_residence, figure_patch_duration))+
+wrap_plots(list(fig_basic_residence, figure_patch_duration),
+           ncol = 1,
+           design = "AA\nAA\nAA\nAA\nBB")+
   plot_annotation(tag_levels = "a",
@@
        filename = "figures/fig_calib_residence_patch.png",
-       height = 130, width = 225, units = "mm")
+       height = 225, width = 170, units = "mm")
 ```

### 2020-10-13 05_calibration_data.Rmd
minor changes

@@
 
-Here we use a speed threshold of 25.6 m/s, a little lower than the 95th percentile.
+Here we use a speed threshold of 15 m/s, the fastest known boat speed.
 
@@
 data_recurse <- data_unproc[, list(x, y, time, TAG, UTCtime)]
-
-# filter for boating segment
-# data_recurse <- atl_filter_covariates(data = data_recurse,
-#  filters = c("between(UTCtime, '2020-08-24 11:13:00', '2020-08-24 12:36:11')"))
 

### 2020-10-14 05_calibration_data.Rmd
add griend remove segments

@@
 
+### Read Griend
+
+```{r}
+# read griend
+griend <- sf::st_read("data/griend_polygon/griend_polygon.shp")
+```
+
 ### Plot the outcome
@@
           alpha = 0.3)+
-  geom_segment(data = patch_summary_data,
-            aes(x = x_end, y = y_end,
-                xend = shift(x_start, type = "lag"),
-                yend = shift(y_start, type = "lag")),
-            col = pal[4], lty = 1,
-            lwd = 0.5)+
+  # geom_segment(data = patch_summary_data,
+  #           aes(x = x_end, y = y_end,
+  #               xend = shift(x_start, type = "lag"),
+  #               yend = shift(y_start, type = "lag")),
+  #           col = pal[4], lty = 1,
+  #           lwd = 0.5)+
   # geom_point(data = patch_summary_data,

### 2020-10-14 01_background.Rmd

@@
 ```
+
+## Simualte for residence patches
+
+```{r}
+# do smoove data using a RACVM
+data <- smoove::simulateRACVM(dt = 0.1, Tmax = 500, omega = 5, v0 = 1, mu = 0.1)
+data <- as.data.table(data$XY)
+
+# assign id
+data[, id := "test"]
+data[, time := seq_len(nrow(data))]
+# rename x and y
+setnames(data, old = c("x", "y"), new = c("y", "x"))
+
+# save data
+fwrite(data, "data/data_for_res_patch.csv")
+```

### 2020-10-14 03_smoothing_tracks.Rmd

@@
 
+```{r}
+### plot figures
+fig_agg_data <- 
+  lapply(list_of_agg, function(df) {
+    ggplot(df)+
+      geom_point(data = data_agg,
+                 aes(x, y),
+                 size = 0.2,
+                 col = pal[3])+
+      geom_path(aes(x,y), 
+                col = pal[1])+
+      geom_point(aes(x,y, 
+                     size = SD,
+                     fill = SD,
+                     group = interval),
+                 shape = 21,
+                 alpha = 0.8,
+                 show.legend = F)+
+      scale_fill_distiller(palette = "Purples", direction = 1,
+                           values = c(-0.25, 1))+
+      scale_size(range = c(2, 4))+
+      ggthemes::theme_few()+
+      theme(axis.text = element_blank(),
+            axis.title = element_blank(),
+            panel.background = element_rect(fill = "grey99"))+
+      coord_cartesian(ylim = c(0.6, NA))
+  })
+```
+
 ## Aggregation Before Reducing Positioning Errors
@@
 
-We prepare the figures.
-
-```{r}
-### plot figures
-fig_agg_data <- 
-  lapply(list_of_agg, function(df) {
-    ggplot(df)+
-      geom_point(data = data_errors,
-                 aes(x, y),
-                 size = 0.2,
-                 col = pal[3])+
-      geom_path(data = data,
-                aes(x, y),
-                lwd = 0.3)+
-      geom_path(aes(x,y), 
-                col = pal[1])+
-      geom_point(aes(x,y, 
-                     size = SD,
-                     fill = SD,
-                     group = interval),
-                 shape = 21,
-                 alpha = 0.8,
-                 show.legend = F)+
-      scale_fill_distiller(palette = "RdPu", direction = 1,
-                           values = c(-0.25, 1))+
-      scale_size(range = c(2, 4))+
-      ggthemes::theme_few()+
-      theme(axis.text = element_blank(),
-            axis.title = element_blank(),
-            panel.background = element_rect(fill = "grey99"))+
-      coord_cartesian(ylim = c(0.6, NA))
-  })
-```
+### Comparing median aggregation with and without smoothing
 
@@
   ggplot(data_agg)+
-  geom_jitter(aes(factor(interval), 1 + speed),
-              col = pal[1],
-              shape = 4, alpha = 0.5)+
-
   geom_hline(yintercept =

### 2020-10-16 02_filtering_data.Rmd
change figure background

@@
         axis.title = element_blank(),
-        panel.background = element_rect(fill = "grey99"))+
+        panel.background = element_rect(fill = "white"))+
   coord_equal()+
@@
         axis.title = element_blank(),
-        panel.background = element_rect(fill = "grey99"))+
+        panel.background = element_rect(fill = "white"))+
   theme(plot.background = element_rect(fill = NA))+
@@
         axis.title = element_blank(),
-        panel.background = element_rect(fill = "grey99"))+
+        panel.background = element_rect(fill = "white"))+
   scale_color_manual(values = c(pal[3], "black"))+

### 2020-10-16 03_smoothing_tracks.Rmd
change figure background

@@
         axis.title = element_blank(),
-        panel.background = element_rect(fill = "grey99"))
+        panel.background = element_rect(fill = "white"))
 
@@
             axis.title = element_blank(),
-            panel.background = element_rect(fill = "grey99"))+
+            panel.background = element_rect(fill = "white"))+
       coord_cartesian(ylim = c(0.6, NA))

### 2020-10-16 05_calibration_data.Rmd
change figure background

@@
         axis.title = element_blank(),
-        panel.background = element_rect(fill = "grey99"))+
+        panel.background = element_rect(fill = "white"))+
   coord_sf(crs = 32631)
@@
         axis.title = element_blank(),
-        panel.background = element_rect(fill = "grey99"))+
+        panel.background = element_rect(fill = "white"))+
   ggspatial::annotation_scale(location = "tl")+
@@
         axis.title = element_blank(),
-        panel.background = element_rect(fill = "grey99"))+
+        panel.background = element_rect(fill = "white"))+
   coord_sf(crs = 32631,
@@
   ggthemes::theme_few()+
-  theme(panel.background = element_rect(fill = "grey99"),
+  theme(panel.background = element_rect(fill = "white"),
         panel.grid.minor = element_line(colour = "grey"))+

### 2020-10-16 02_filtering_data.Rmd
clean up figures

@@
 
+```{r}
+data_copy <- fread("data/data_errors.csv")
+```
+
 Define a palette with 4 colours for convenience.
@@
 # define a four colour palette
-pal <- RColorBrewer::brewer.pal(3, "PRGn")
+pal <- RColorBrewer::brewer.pal(4, "Set1")
 ```
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        panel.background = element_rect(fill = "white"))+
+        axis.title = element_blank())+
   coord_equal()+
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        panel.background = element_rect(fill = "white"))+
+        axis.title = element_blank())+
   theme(plot.background = element_rect(fill = NA))+
@@
              aes(x, y),
-             shape = 2, col = pal[1])+
+             size = 0.2,
+             shape = 2, col = pal[4])+
   geom_point(data = data_copy[!data_copy[500:800, ],
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        panel.background = element_rect(fill = "white"))+
+        axis.title = element_blank())+
   scale_color_manual(values = c(pal[3], "black"))+
@@
              alpha = 0.5,
-             col = "grey",
-             fill = pal[2],
+             col = "grey40",
+             fill = "grey",
              shape = 21)+

### 2020-10-16 05_calibration_data.Rmd
clean up figures

@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        panel.background = element_rect(fill = "white"))+
+        axis.title = element_blank())+
   coord_sf(crs = 32631)
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        panel.background = element_rect(fill = "white"))+
+        axis.title = element_blank())+
   ggspatial::annotation_scale(location = "tl")+
@@
                   tag_prefix = "(",
-                  tag_suffix = ")")
+                  tag_suffix = ")")&
+  theme(plot.tag = element_text(face = "bold"))
 
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        panel.background = element_rect(fill = "white"))+
+        axis.title = element_blank())+
   coord_sf(crs = 32631,

### 2020-10-28 06_bat_data.Rmd
add prelim bat data code

---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Processing calibration data

## Prepare libraries

```{r}
# load libs
library(data.table)
library(RSQLite)
library(atlastools)
library(ggplot2)
library(patchwork)

# prepare a palette
pal <- RColorBrewer::brewer.pal(4, "Set1")
```

## Read bat data

Read the bat data and convert to a csv file.

```{r read_bat_data}
# prepare the connection
con <- dbConnect(drv = SQLite(), dbname = "data/Three_example_bats.sql")

# list the tables
table_name <- dbListTables(con)

# prepare to query all tables
query <- sprintf('select * from \"%s\"', table_name)

# query the database
data <- dbGetQuery(conn = con, statement = query)

# disconnect from database
dbDisconnect(con)
```

Convert data to csv.

```{r}
# convert data to datatable
setDT(data)

# write data for QGIS
fwrite(data, file = "data/bat_data.csv")
```

## WIP

### 2020-10-28 06_bat_data.Rmd
add bat processing script

@@
 
-# Processing calibration data
+# Processing Egyptian fruit bat tracks
+
+We show the pre-processing pipeline at work on the tracks of three Egyptian fruit bats (_Rousettus aegyptiacus_), and construct residence patches.
 
@@
 
-## WIP
+## Sanity check: Plot bat data
+
+```{r}
+# plot bat data to examine for errors
+fig_bat_raw <-
+  ggplot(data)+
+  geom_path(aes(X, Y,
+                group = TAG,
+                colour = TAG),
+            show.legend = FALSE)+
+  scale_colour_brewer(palette = "Dark2")+
+  ggspatial::annotation_scale(location = "br")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  coord_sf(crs = 2039)
+```
+
+Save the raw data plot.
+
+```{r}
+# save to figures
+ggsave(fig_bat_raw,
+       filename = "figures/fig_bat_raw.png",
+       dpi = 150,
+       width = 6, height = 4)
+```
+
+Show the plot.
+
+![](figures/fig_bat_raw.png)
+
+## Prepare data for filtering
+
+### Prepare data per individual
+
+```{r}
+# split bat data by tag
+# first make a copy using the data.table function copy
+# this prevents the orignal data from being modified by atlastools
+# functions which DO MODIFY BY REFERENCE!
+data_split <- copy(data)
+
+# now split
+data_split <- split(data_split, data$TAG)
+```
+
+## Filter by covariates
+
+No natural bounds suggest themselves, and we proceed to filter by covariates, since point outliers are obviously visible.
+
+We use filter out positions with `SD > 20` and positions calculated using only 3 base stations.
+
+We use the function `atl_filter_covariates`.
+
+```{r}
+# get SD.
+# since the data are data.tables, no assignment is necessary
+invisible(
+  lapply(data_split, function(dt) {
+    dt[, SD := sqrt(VARX + VARY + (2 * COVXY))]
+  })
+)
+
+# filter for SD <= 20
+# here, reassignment is necessary as rows are being removed
+# the atl_filter_covariates function could have been used here
+data_split <- lapply(data_split, function(dt) {
+  dt <- atl_filter_covariates(
+    data = dt,
+    filters = c("SD <= 20",
+                "NBS > 3")
+  )
+})
+
+# check whether the filter has worked
+invisible(
+  lapply(data_split, function(dt) {
+    assertthat::assert_that(min(dt$SD) <= 20,
+                            msg = "some SDs above 20 remain")
+    assertthat::assert_that(min(dt$NBS) > 3,
+                            msg = "some NBS below 3 remain")
+  })
+)
+```
+
+### Sanity check: Plot filtered data
+
+```{r}
+# bind all individuals together
+data_split <- rbindlist(data_split)
+```
+
+```{r}
+# now plot over raw data
+fig_bat_filter_cov <- 
+  ggplot()+
+  geom_point(data = data[!data_split, 
+                         on = c("TAG", "X", "Y", "TIME")],
+             aes(X, Y),
+             col = "grey",
+             shape = 4)+
+  geom_path(data = data_split,
+             aes(X, Y,
+                 col = TAG),
+             size = 0.3,
+             show.legend = FALSE)+
+  scale_colour_brewer(palette = "Dark2")+
+  ggspatial::annotation_scale(location = "br")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  coord_sf(crs = 2039,
+           xlim = range(data_split$X),
+           ylim = range(data_split$Y))
+```
+
+Save the filtered data plot.
+
+```{r}
+# save to figures
+ggsave(fig_bat_filter_cov,
+       filename = "figures/fig_bat_filter_cov.png",
+       dpi = 150,
+       width = 4, height = 6)
+```
+
+Show the plot.
+
+![](figures/fig_bat_filter_cov.png)
+## Filter by speed
+
+Some point outliers remain, and should be removed using a speed filter.
+
+First we calculate speeds.
+
+```{r}
+# calculate speed on split data once more
+data_split <- split(data_split, data_split$TAG)
+
+# get speeds as with SD, no reassignment required for columns
+invisible(
+  lapply(data_split, function(dt) {
+
+    # first process time to seconds
+    # assign to a new column
+    dt[, time := floor(TIME / 1000)]
+
+    dt[, `:=`(speed_in = atl_get_speed(dt, 
+                                       x = "X", y = "Y", 
+                                       time = "time",
+                                       type = "in"),
+              speed_out = atl_get_speed(dt, 
+                                       x = "X", y = "Y", 
+                                       time = "time",
+                                       type = "out"))]
+  })
+)
+```
+
+Now filter for speeds > 20 m/s (around 70 km/h).
+
+```{r}
+# filter speeds
+# reassignment is required here
+data_split <- lapply(data_split, function(dt) {
+  dt <- na.omit(dt, cols = c("speed_in", "speed_out"))
+
+  dt <- atl_filter_covariates(data = dt,
+                              filters = c("speed_in <= 20",
+                                          "speed_out <= 20"))
+})
+```
+
+### Sanity check: Plot speed filtered data
+
+```{r}
+# bind all individuals together
+data_split <- rbindlist(data_split)
+```
+
+```{r}
+# now plot over raw data
+fig_bat_filter_speed <-
+  ggplot()+
+  geom_point(data = data[!data_split, 
+                         on = c("TAG", "X", "Y", "TIME")],
+             aes(X, Y),
+             col = "grey",
+             shape = 4)+
+  geom_path(data = data_split,
+             aes(X, Y,
+                 col = TAG),
+             size = 0.3,
+             show.legend = FALSE)+
+  scale_colour_brewer(palette = "Dark2")+
+  ggspatial::annotation_scale(location = "br")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  coord_sf(crs = 2039,
+           xlim = range(data_split$X),
+           ylim = range(data_split$Y))
+```
+
+Save the speed filtered data plot.
+
+```{r}
+# save to figures
+ggsave(fig_bat_filter_speed,
+       filename = "figures/fig_bat_filter_speed.png",
+       dpi = 150,
+       width = 4, height = 6)
+```
+
+Show the plot.
+
+![](figures/fig_bat_filter_speed.png)
+
+## Median smoothing
+
+Apply a 5 point median smooth to the data.
+
+```{r}
+# since the function modifies in place, we shall make a copy
+data_smooth <- copy(data_split)
+
+# split the data again
+data_smooth <- split(data_smooth, data_smooth$TAG)
+```
+
+Remember, **`atl_median_smooth` MODIFIES IN PLACE**.
+
+```{r}
+# apply the median smooth to each list element
+# no reassignment is required as THE FUNCTION MODIFIES IN PLACE!
+invisible(
+  # the function arguments to atl_median_smooth
+  # can be passed directly in lapply
+  lapply(data_smooth, atl_median_smooth,
+         time = "time", moving_window = 5)
+)
+```
+
+### Sanity check: Plot smoothed data
+
+```{r}
+# recombine split up data that has been smoothed
+data_smooth <- rbindlist(data_smooth)
+```
+
+```{r}
+# plot the raw, filtered, and smoothed data
+fig_bat_smooth <- 
+  ggplot()+
+  geom_point(data = data[!data_split, 
+                         on = c("TAG", "X", "Y", "TIME")],
+             aes(X, Y),
+             col = "grey",
+             shape = 4)+
+  geom_point(data = data_split,
+             aes(X, Y),
+             col = pal[3],
+             alpha = 0.5,
+             shape = 1,
+             show.legend = FALSE)+
+  geom_path(data = data_smooth,
+            aes(X, Y,
+                col = as.numeric(TAG)),
+            lwd = 0.3,
+            show.legend = FALSE)+
+  scale_fill_distiller(palette = "Greens",
+                       direction = 1)+
+  scale_colour_distiller(palette = "Purples",
+                         direction = 1,
+                         values = c(-0.2, 1))+
+  ggspatial::annotation_scale(location = "br")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  coord_sf(crs = 2039,
+           xlim = range(data_smooth$X),
+           ylim = range(data_smooth$Y))
+```
+
+Save the speed filtered data plot.
+
+```{r}
+# save to figures
+ggsave(fig_bat_smooth,
+       filename = "figures/fig_bat_smooth.png",
+       dpi = 150,
+       width = 4, height = 6)
+```
+
+Show the plot.
+
+![](figures/fig_bat_smooth.png)
+
+## Making residence patches
+
+### Calculating residence time
+
+First, the data is put through the `recurse` package to get residence time.
+
+```{r}
+# load recurse
+library(recurse)
+
+# split the data
+data_smooth <- split(data_smooth, data_smooth$TAG)
+```
+
+Get residence time. Since bats may revisit the same features, we want to prevent confusion between frequent revisits and prolonged residence.
+
+For this, we stop summing residence times within Z metres of a location if the animal exited the area for one hour or more.
+
+```{r}
+# get residence times
+
+data_residence <- lapply(data_smooth, function(dt) {
+  # do basic recurse
+  dt_recurse <- getRecursions(
+    x = dt[, c("X", "Y", "time", "TAG")],
+    radius = 50,
+    timeunits = "mins"
+  )
+
+  # get revisit stats
+  dt_recurse <- setDT(
+    dt_recurse[["revisitStats"]]
+  )
+
+  # count long absences from the area
+  dt_recurse[, timeSinceLastVisit :=
+          ifelse(is.na(timeSinceLastVisit), -Inf, timeSinceLastVisit)]
+  dt_recurse[, longAbsenceCounter := cumsum(timeSinceLastVisit > 60),
+             by = .(coordIdx)
+             ]
+  # get data before the first long absence of 60 mins
+  dt_recurse <- dt_recurse[longAbsenceCounter < 1, ]
+
+  dt_recurse <- dt_recurse[, list(
+    resTime = sum(timeInside),
+    fpt = first(timeInside),
+    revisits = max(visitIdx)
+  ),
+  by = .(coordIdx, x, y)
+  ]
+
+  # prepare and merge existing data with recursion data
+  dt[, coordIdx := seq(nrow(dt))]
+
+  dt <- merge(dt, 
+              dt_recurse[, c("coordIdx", "resTime")], 
+              by = c("coordIdx"))
+
+  setorderv(dt, "time")
+})
+```
+
+### Sanity check: Residence-time time-series
+
+```{r}
+# bind the list
+data_residence <- rbindlist(data_residence)
+
+# get time as human readable
+data_residence[, ts := as.POSIXct(time, origin = "1970-01-01")]
+```
+
+```{r}
+# fig_bat_res_timeseries <- 
+  ggplot()+
+  geom_path(data = data_residence,
+            aes(ts, resTime,
+                colour = TAG),
+            show.legend = F)+
+  scale_color_brewer(palette = "Dark2")+
+  scale_x_datetime()+
+  ggthemes::theme_few()+
+  coord_cartesian(ylim = c(0, 50))+
+  labs(x = NULL,
+       y = "residence time (mins)")
+```
+
+Save the speed filtered data plot.
+
+```{r}
+# save to figures
+ggsave(fig_bat_res_timeseries,
+       filename = "figures/fig_bat_res_timeseries.png",
+       dpi = 150,
+       width = 6, height = 3)
+```
+
+Show the plot.
+
+![](figures/fig_bat_res_timeseries.png)
+
+### Constructing residence patches
+
+Split the data and construct residence patches.
+
+Some preparation is required. First, the function requires columns `x`, `y`,
+`time`, and `id`.
+
+```{r}
+# add an id column
+data_residence[, `:=`(id = TAG,
+                      x = X, y = Y)]
+
+# filter for residence time > 5 minutes
+data_residence <- data_residence[resTime > 5, ]
+
+# split the data
+data_residence <- split(data_residence, data_residence$TAG)
+```
+
+Now segment-cluster into residence patches.
+
+```{r}
+# segment into residence patches
+data_patches <- lapply(data_residence, atl_res_patch,
+                       buffer_radius = 25)
+```
+
+### Getting residence patch data
+
+We get the residence patch data as spatial `sf-MULTIPOLYGON` objects.
+
+```{r}
+# get data spatials
+data_spatials <- lapply(data_patches, atl_patch_summary,
+                        which_data = "spatial",
+                        buffer_radius = 25)
+
+# bind list
+data_spatials <- rbindlist(data_spatials)
+
+# convert to sf
+library(sf)
+data_spatials <- st_sf(data_spatials, sf_column_name = "polygons")
+
+# assign a crs
+st_crs(data_spatials) <- st_crs(2039)
+```
+
+### Write patch spatial representations
+
+```{r}
+st_write(data_spatials,
+         dsn = "data/data_bat_residence_patches.gpkg")
+```
+
+Write cleaned bat data.
+
+```{r}
+data_clean <- fwrite(rbindlist(data_smooth),
+                     file = "data/data_bat_smooth.csv")
+```
+
+Write patch summary.
+
+```{r}
+# get summary
+patch_summary <- lapply(data_patches, atl_patch_summary)
+
+# bind summary
+patch_summary <- rbindlist(patch_summary)
+
+# write
+fwrite(patch_summary,
+       "data/data_bat_patch_summary.csv")
+```

### 2020-10-28 06_bat_data.Rmd
some minor section issues

@@
 ![](figures/fig_bat_filter_cov.png)
+
 ## Filter by speed
@@
   ggplot()+
-  geom_point(data = data[!data_split, 
-                         on = c("TAG", "X", "Y", "TIME")],
-             aes(X, Y),
-             col = "grey",
-             shape = 4)+
   geom_point(data = data_split,
@@
              col = pal[3],
-             alpha = 0.5,
-             shape = 1,
+             alpha = 0.2,
+             shape = 16,
              show.legend = FALSE)+
@@
 
-Save the speed filtered data plot.
+Save the smoothed data plot.
 
@@
 ```
+
+## Processed bat patches
+
+![](figures/fig_bat_patches.png)

### 2020-10-28 03_smoothing_tracks.Rmd

@@
 # prepare a palette
-pal <- RColorBrewer::brewer.pal(3, "PRGn")
+pal <- RColorBrewer::brewer.pal(4, "Set1")
 ```
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        panel.background = element_rect(fill = "white"))
+        axis.title = element_blank())
 
@@
                position = position_nudge(x = -0.25,),
-               fill = pal[2],
+               fill = "grey",
                alpha = 0.5,

### 2020-10-28 06_bat_data.Rmd
fix figures

@@
        filename = "figures/fig_bat_raw.png",
-       dpi = 150,
+       dpi = 300,
        width = 6, height = 4)
@@
        filename = "figures/fig_bat_filter_cov.png",
-       dpi = 150,
+       dpi = 300,
        width = 4, height = 6)
@@
        filename = "figures/fig_bat_filter_speed.png",
-       dpi = 150,
+       dpi = 300,
        width = 4, height = 6)
@@
              aes(X, Y),
-             col = pal[3],
-             alpha = 0.2,
-             shape = 16,
+             col = "grey",
+             shape = 4,
              show.legend = FALSE)+
@@
             aes(X, Y,
-                col = as.numeric(TAG)),
+                col = TAG),
             lwd = 0.3,
             show.legend = FALSE)+
-  scale_fill_distiller(palette = "Greens",
-                       direction = 1)+
-  scale_colour_distiller(palette = "Purples",
-                         direction = 1,
-                         values = c(-0.2, 1))+
+  scale_colour_brewer(palette = "Dark2")+
   ggspatial::annotation_scale(location = "br")+
@@
        filename = "figures/fig_bat_smooth.png",
-       dpi = 150,
+       dpi = 300,
        width = 4, height = 6)
@@
 ```{r}
-# fig_bat_res_timeseries <- 
+fig_bat_res_timeseries <-
   ggplot()+
@@
        filename = "figures/fig_bat_res_timeseries.png",
-       dpi = 150,
+       dpi = 300,
        width = 6, height = 3)

### 2020-10-28 render_books.sh
pdf bookdown

@@ -1,0 +1,3 @@
+#!/bin/bash
+Rscript --slave -e 'bookdown::render_book("index.rmd")'
+Rscript --slave -e 'bookdown::render_book("index.rmd", "rmarkdown::pdf_document"); warnings()'

### 2020-10-28 01_background.Rmd

@@
 
-Releases of the package from Github can be found on Zenodo: [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4033155.svg)](https://doi.org/10.5281/zenodo.4033155)
-
-The build status of the 'Master' branch is shown here:
-
-<!-- badges: start -->
-  [![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/pratikunterwegs/atlastools?branch=master&svg=true)](https://ci.appveyor.com/project/pratikunterwegs/atlastools) 
-  [![Build Status](https://travis-ci.org/pratikunterwegs/atlastools.svg?branch=master)](https://travis-ci.org/pratikunterwegs/atlastools) 
-  [![codecov.io](https://codecov.io/github/pratikunterwegs/atlastools/coverage.svg?branch=master)](https://codecov.io/github/pratikunterwegs/atlastools/branch/master)
-<!-- badges: end -->
+Releases of the package from Github can be found on Zenodo: (https://doi.org/10.5281/zenodo.4033155)
 

### 2020-10-28 06_bat_data.Rmd

@@
   ggthemes::theme_few()+
-  coord_cartesian(ylim = c(0, 50))+
   labs(x = NULL,

### 2020-10-29 02_filtering_data.Rmd
hide ggplot in bookdown

@@
 
-```{r}
+```{r echo=FALSE}
 # make figure of canonical data with added errors
@@
 
-```{r}
+```{r echo=FALSE}
 # plot data inside and outside bbox
@@
 
-```{r}
+```{r echo=FALSE}
 # data plot
@@
 
-```{r}
+```{r echo=FALSE}
 # get plots

### 2020-10-29 index.rmd
new options for bookdown

--- 2020-10-08 index.rmd
@@
 site: bookdown::bookdown_site
-output: bookdown::pdf_book
-documentclass: book
-# bibliography: [refs.bib]
-biblio-style: apalike
+output: bookdown::pdf_document2
+toc-depth: 1
+secnumdepth: 2
+documentclass: scrreprt
+geometry: "left=4cm, right=3cm, top=2.5cm, bottom=2.5cm"
+bibliography: [references.bib]
 link-citations: yes

### 2020-10-29 docs/Pre-processing-high-frequency-tracking-data.tex
add new bookdown figures and pdf'

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{scrreprt}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A Rough Guide to Pre-processing High-Frequency Animal Tracking Data},
  pdfauthor={Pratik R. Gupte and others},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[left=4cm, right=3cm, top=2.5cm, bottom=2.5cm]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{2}

\usepackage{fontspec}
% use nice fonts if available else use boring defaults
\IfFontExistsTF{IBM Plex Mono}{\setmonofont[]{IBM Plex Mono}}{}

\usepackage{lineno}
\usepackage{tgtermes}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{A Rough Guide to Pre-processing High-Frequency Animal Tracking Data}
\author{Pratik R. Gupte and others}
\date{2020-10-29}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{abstract}{%
\chapter{Abstract}\label{abstract}}

Data cleaning is a ubiquitous pre-processing step in analysis pipelines, and its automation is essential for large data volumes such as those generated in animal tracking studies using high-frequency Time-of-Arrival (TOA) systems.
Users of systems such as ATLAS must contend with two intertwined data cleaning challenges: (1) reducing positioning errors, and (2) the high volume of data itself.
Making biological inferences from data with positioning errors is not straightforward, and processing large data volumes is computationally intensive.
Though reducing positioning error is widely recommended, users are without uniform guidance on how to go about this, and lack a common set of computationally efficient tools.
Further, many methods that synthesize movement tracks for ecological inference are either (1) not suited to very large datasets, or (2) not intuitive to understand in terms of the tracked animal's biology.

In this article we introduce a pipeline to pre-process high-frequency animal tracking data in order to prepare it for subsequent analysis.
We demonstrate this pipeline on simulated movement data to which we have randomly added positioning errors.
This pipeline is suited to any tracking study in which the high data volume combined with knowledge of the tracked individuals' biology can be used to smooth out positioning errors.
We further suggest how large volumes of cleaned data may be synthesized into biologically meaningful `residence patches', and demonstrate how this accurately captures animal space-use.
Finally, we introduce the \texttt{R} package \texttt{atlastools} which provides fast implementations of the methods we describe.
Though aimed at ATLAS systems, \texttt{atlastools} can be used with any time-series animal movement data, and we demonstrate its usage with both simulated and empirical examples.

\includegraphics{figures/fig_recipe.png}

\hypertarget{getting-started}{%
\chapter{Getting started}\label{getting-started}}

This section covers:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Installing the \texttt{R} package \texttt{atlastools},
\item
  Simulating some realistic looking movement data using the \texttt{R} package \texttt{smoove} from (Gurarie et al. \protect\hyperlink{ref-gurarie2017}{2017}), and
\item
  Introducing positioning errors into the simulated movement data.
\end{enumerate}

\hypertarget{installing-atlastools}{%
\section{\texorpdfstring{Installing \texttt{atlastools}}{Installing atlastools}}\label{installing-atlastools}}

This paper refers extensively to the \texttt{R} package \texttt{atlastools} (Gupte 2020), which can be installed from \href{https://github.com/pratikunterwegs/atlastools}{Github}.

Releases of the package from Github can be found on Zenodo: (\url{https://doi.org/10.5281/zenodo.4033155})

The code chunk below shows how to install \texttt{atlastools}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# use either devtools or remotes to install}
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}

\CommentTok{\# installation using devtools}
\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{simulating-movement-data}{%
\section{Simulating movement data}\label{simulating-movement-data}}

Here, we simulate some movement data using the \texttt{smoove} \texttt{R} package from (Gurarie et al. \protect\hyperlink{ref-gurarie2017}{2017}).

First, we load \texttt{smoove} and \texttt{data.table}, as well as some helper functions that make use of them to simulate data for use.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# load smoove and datatable}
\KeywordTok{library}\NormalTok{(smoove)}
\KeywordTok{library}\NormalTok{(data.table)}

\CommentTok{\# source helper functions}
\KeywordTok{source}\NormalTok{(}\StringTok{"R/helper\_functions.R"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{do\_smoove\_data}\NormalTok{()}
\CommentTok{\# save simulated data}
\KeywordTok{fwrite}\NormalTok{(data, }\StringTok{"data/data\_sim.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{simualte-for-residence-patches}{%
\section{Simualte for residence patches}\label{simualte-for-residence-patches}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# do smoove data using a RACVM}
\NormalTok{data <{-}}\StringTok{ }\NormalTok{smoove}\OperatorTok{::}\KeywordTok{simulateRACVM}\NormalTok{(}\DataTypeTok{dt =} \FloatTok{0.1}\NormalTok{, }\DataTypeTok{Tmax =} \DecValTok{500}\NormalTok{, }\DataTypeTok{omega =} \DecValTok{5}\NormalTok{, }\DataTypeTok{v0 =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mu =} \FloatTok{0.1}\NormalTok{)}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{as.data.table}\NormalTok{(data}\OperatorTok{$}\NormalTok{XY)}

\CommentTok{\# assign id}
\NormalTok{data[, id }\OperatorTok{:}\ErrorTok{=}\StringTok{ "test"}\NormalTok{]}
\NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{seq\_len}\NormalTok{(}\KeywordTok{nrow}\NormalTok{(data))]}
\CommentTok{\# rename x and y}
\KeywordTok{setnames}\NormalTok{(data, }\DataTypeTok{old =} \KeywordTok{c}\NormalTok{(}\StringTok{"x"}\NormalTok{, }\StringTok{"y"}\NormalTok{), }\DataTypeTok{new =} \KeywordTok{c}\NormalTok{(}\StringTok{"y"}\NormalTok{, }\StringTok{"x"}\NormalTok{))}

\CommentTok{\# save data}
\KeywordTok{fwrite}\NormalTok{(data, }\StringTok{"data/data\_for\_res\_patch.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{reducing-large-scale-positioning-error-by-filtering-data}{%
\chapter{Reducing Large-Scale Positioning Error by Filtering Data}\label{reducing-large-scale-positioning-error-by-filtering-data}}

\hypertarget{prepare-libraries}{%
\section{Prepare Libraries}\label{prepare-libraries}}

Here we load some useful libraries, and the helper functions.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# to handle movement data}
\KeywordTok{library}\NormalTok{(data.table)}
\KeywordTok{library}\NormalTok{(atlastools)}

\CommentTok{\# to plot}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{library}\NormalTok{(patchwork)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# source helper functions}
\KeywordTok{source}\NormalTok{(}\StringTok{"R/helper\_functions.R"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{introducing-errors-to-the-data}{%
\section{Introducing Errors to the Data}\label{introducing-errors-to-the-data}}

Here we introduce three kinds of errors to the data:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Small-scale normally distributed errors at each position;
\item
  Large-scale error at a random 0.5\% of positions;
\item
  A large-scale displacement of a sequence of 300 positions.
\end{enumerate}

While the data are 10,000 positions at 1-second interval, we shall use only 5,000 of these.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# read in the data}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_sim.csv"}\NormalTok{)[}\DecValTok{5000}\OperatorTok{:}\DecValTok{10000}\NormalTok{, ]}
\end{Highlighting}
\end{Shaded}

We add outliers at random to the data to demonstrate their removal.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# make a copy}
\NormalTok{data\_copy <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}

\CommentTok{\# add a prolonged spike or reflection to 300 positions}
\NormalTok{data\_copy[}\DecValTok{500}\OperatorTok{:}\DecValTok{800}\NormalTok{, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ x }\OperatorTok{+}\StringTok{ }\FloatTok{0.25}\NormalTok{,}
                        \DataTypeTok{y =}\NormalTok{ y }\OperatorTok{+}\StringTok{ }\FloatTok{0.25}\NormalTok{)]}

\CommentTok{\# add normal error}
\NormalTok{data\_copy[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{x =} \KeywordTok{do\_add\_error}\NormalTok{(x, }\DataTypeTok{std\_dev =} \FloatTok{0.01}\NormalTok{),}
                 \DataTypeTok{y =} \KeywordTok{do\_add\_error}\NormalTok{(y, }\DataTypeTok{std\_dev =} \FloatTok{0.005}\NormalTok{))]}

\CommentTok{\# add 100 outliers}
\NormalTok{data\_copy <{-}}\StringTok{ }\KeywordTok{do\_add\_outliers}\NormalTok{(data\_copy, }\DataTypeTok{p\_data =} \FloatTok{0.005}\NormalTok{, }\DataTypeTok{std\_dev =} \FloatTok{0.1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Save the data to which errors have been added.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{fwrite}\NormalTok{(data\_copy, }\DataTypeTok{file =} \StringTok{"data/data\_errors.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{data\_copy <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_errors.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Define a palette with 4 colours for convenience.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# define a four colour palette}
\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We make a figure of the canonical data (grey line) along with the artificially added error (grey points).

\hypertarget{filtering-by-spatial-bounds}{%
\section{Filtering by Spatial Bounds}\label{filtering-by-spatial-bounds}}

Filtering by spatial bounds is a good way to begin reducing gross, large-scale positioning errors.
There are two main ways of doing this:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Filtering by a bounding box: Compare the coordinates of observations against a range of acceptable coordinates, and retain those which fall within the range, or
\item
  Filtering by a spatial polygon: An explicit geometric intersection between the positions and a polygon representing an area of interest.
\end{enumerate}

Here, we remove gross positioning errors using a bounding box filter using the function \texttt{atl\_filter\_bounds}. In this example, we show filtering only on the Y coordinate.

\texttt{atl\_filter\_bounds} takes coordinate ranges as two-element vectors of the lower and higher bound.
It is possible to pass one of the bounds as \texttt{NA}, in which case, only the other bound is used for filtering, i.e., \texttt{y\_range\ =\ c(NA,\ 1)} is equivalent to selecting all Y coordinates \textless{} 1.

\texttt{atl\_filter\_bounds} was initially designed to remove positions \emph{inside} a specific range, hence the argument \texttt{remove\_inside}.
The default value of the argument is \texttt{FALSE}, and \texttt{atl\_filter\_bounds} is thus a bounding box filter.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# remove positions outside a bounding box}
\CommentTok{\# NB: set remove\_inside to FALSE}
\NormalTok{data\_inside\_bbox <{-}}\StringTok{ }\KeywordTok{atl\_filter\_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_copy,}
                                      \DataTypeTok{y\_range =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{1}\NormalTok{),}
                                      \DataTypeTok{remove\_inside =} \OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\texttt{atl\_filter\_bounds} is not vectorised, and if there are two or more bounds per coordinate (for instance, \(X_1 \ldots X_2\), and \(X_3 \ldots X_4\)), they must be passed in two different function calls.
The same is true for filtering by an \texttt{sf} polygon. \texttt{atl\_filter\_bounds} also supports filtering by multi-polygon objects.

Having filtered the data we shall prepare it for plotting.

\includegraphics{figures/fig_raw_bounds.png}

\textbf{(a)} A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the \texttt{smoove} package (Gurarie et al.~2017). Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track.
The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (solid black line) as closely as possible.
\textbf{(b)} Tracks can be quickly filtered by spatial bounds (\(0.5 \leq Y \leq 1.0\); dashed grey lines) using the \texttt{atlastools} function \texttt{atl\_filter\_bounds}. Setting the function argument \texttt{remove\_inside\ =\ FALSE} retains positions within user supplied bounds (blue points), and excludes those outside (red points).

\hypertarget{a-note-on-filtering-by-bounds}{%
\subsection{A Note on Filtering by Bounds}\label{a-note-on-filtering-by-bounds}}

The filter on spatial bounds is only for demonstration, and is not applied to the data.

\hypertarget{filtering-unrealistic-movement}{%
\section{Filtering Unrealistic Movement}\label{filtering-unrealistic-movement}}

Large-scale positioning error can affect both point locations as well as entire subsets of a track, leading to a track appearing to show unrealistic movement for the study species (Bjørneraas et al. \protect\hyperlink{ref-bjorneraas2010}{2010}).

The best way to examine whether a track contains such positioning errors is to plot the data, and especially to join the dots, i.e., to connect the positions with lines rather than simply plot points.

Users with extensive experience of their study system will readily recognise segments of a track where the movement appears to be unrealistic.

Briefly, the two main kinds of errors are mentioned here, and named based on their appearance in a tack: those that affect single positions lend a spiky appearance to a track, and are referred to as \emph{point outliers} or \emph{spikes}, while errors affecting continuous subsets of a track cause it to appear as though it has been reflected along a plane, and are hence called \emph{reflections}, or \emph{prolonged spikes}.

\hypertarget{filtering-point-outliers-or-spikes}{%
\subsection{Filtering Point Outliers or Spikes}\label{filtering-point-outliers-or-spikes}}

We begin by removing unrealistic movement in the form of point outliers, or spikes.

The first step is to determine how spikes should be identified. The non-movement approach prescribes determining whether movement metrics associated with each position are realistic or not, and targeting those positions where the movement metrics are unrealistic for the study species.

In this case we shall calculate only two metrics, speed and turning angle.
These are conveniently implemented in \texttt{atlastools} using the functions \texttt{atl\_get\_speed} and \texttt{atl\_turning\_angle}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get speed and turning angle}
\NormalTok{data\_copy[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{in\_speed =} \KeywordTok{atl\_get\_speed}\NormalTok{(data\_copy,}
                                          \DataTypeTok{type =} \StringTok{"in"}\NormalTok{),}
                 \DataTypeTok{out\_speed =} \KeywordTok{atl\_get\_speed}\NormalTok{(data\_copy,}
                                           \DataTypeTok{type =} \StringTok{"out"}\NormalTok{),}
                 \DataTypeTok{angle =} \KeywordTok{atl\_turning\_angle}\NormalTok{(data\_copy))]}
\end{Highlighting}
\end{Shaded}

Having calculated speed and turning angle, the next step is to remove positions with extremely high incoming and outgoing speeds. This means using the \texttt{atl\_filter\_covariates} function to remove positions with speed \(\geq\) a plausible speed threshold. The use of a turning angle filter is optional, and not necessary in this case.

One approach is to define a speed cutoff based on expert knowledge, and this is best suited to well studied species.

For simulated data, there is no plausible speed threshold from prior knowledge --- a reasonable choice here is to use the 90\^{}th or 95\^{}th percentile of speed and turning angle (when this metric is used).

This is also more general than identifying the limits of implausibility for each individual (since there may be inter-individual differences), let alone each species in a large dataset.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get 90 and 95 percentile of speed and turning angle}
\KeywordTok{sapply}\NormalTok{(data\_copy[, }\KeywordTok{c}\NormalTok{(}\StringTok{"in\_speed"}\NormalTok{, }\StringTok{"angle"}\NormalTok{)], }\ControlFlowTok{function}\NormalTok{(z) \{}
  \KeywordTok{quantile}\NormalTok{(z, }\DataTypeTok{probs =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.9}\NormalTok{, }\FloatTok{0.95}\NormalTok{), }\DataTypeTok{na.rm =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{\})}
\end{Highlighting}
\end{Shaded}

Finally, we shall remove positions whose incoming and outgoing speeds are both greater than the 95th speed percentile using \texttt{atl\_filter\_covariates}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# filter the copy by the 95th percentile}
\NormalTok{data\_filtered <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(data\_copy,}
    \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(in\_speed < 0.024 \& out\_speed < 0.024) | angle < 40"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

We prepare the data for plotting.

\hypertarget{filtering-reflections-or-track-subsets}{%
\subsection{Filtering Reflections or Track Subsets}\label{filtering-reflections-or-track-subsets}}

When entire track subsets are affected by positioning error, they are more difficult to remove. This is the case when positions are \emph{reflected} along a plane in the coordinate axis --- the simplest way in which this can be explained is to examine the way in which we simulated the reflection earlier in this text, by simply adding an offset to the X and Y coordinates of each of 300 consecutive positions (indices 500 -- 800).

An explanation of why these reflections occur is found in WATLAS PAPER: Bijleveld et al.~in prep. In short, they are results of an error in the ATLAS localisation algorithm, and are usually removed in the localisation step's quality control procedures. However, some reflections may remain to confront users.

Reflections and other issues affecting track subsets cannot be resolved by targeting single positions, since there are rarely any position-specific covariates that can be used to identify a position as part of a larger subset that should be removed.

However, it is often possible to identify the \emph{bounds} of problematic subsets, such as reflections, and remove positions between them. The way this is done is conceptually similar to the point outlier algorithm.

The \texttt{atl\_remove\_reflection} function implements one method remove reflected subsets of a track. The working is described:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Remove point outliers,
\item
  Re-calculate speed and turning angles
\item
  Identify the first unrealistic movement position (fast speed and high turning angle)
\item
  Setting this point as an anchor, identify the next position with unrealistic movement (as above)
\item
  Remove all positions between these two points,
\item
  Search for the next unrealistic movement, and repeat the process.
\end{enumerate}

This is a minimal algorithm which can be developed further, and iterated multiple times to comprehensively remove reflections.

One important function argument is the estimated reflection length, i.e., how many positions are estimated to be reflected. This argument controls how many positions after the anchor are candidates for the reflection's end.

If the reflection does not end within this number of positions, the algorithm goes awry, and considers part of the reflection to be the valid data.
Fortunately, this can be avoided by setting the length to the number of observations in the data.

However, if the reflection does not end at all, the algorithm will simply delete all positions from the anchor point onwards.

Thus, the algorithm is less of a double-edged sword and more a sword with no hilt --- there are few safe ways to use it.

The difficulty in implementing a reasonable algorithm for dealing with this kind of track-subset wide positioning error reveals the general problem that algorithms are \textbf{not} easy to conceptualise or efficiently implement.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# attempt to remove reflections}
\NormalTok{data\_no\_reflection <{-}}\StringTok{ }\KeywordTok{atl\_remove\_reflections}\NormalTok{(data\_filtered,}
                          \DataTypeTok{point\_angle\_cutoff =} \DecValTok{10}\NormalTok{,}
                          \DataTypeTok{reflection\_speed\_cutoff =} \FloatTok{0.024}\NormalTok{)}
\CommentTok{\# get reflections}
\NormalTok{reflection <{-}}\StringTok{ }\NormalTok{data\_filtered[}\OperatorTok{!}\NormalTok{data\_no\_reflection,}
\NormalTok{                                         on =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"x"}\NormalTok{, }\StringTok{"y"}\NormalTok{)]}
\NormalTok{reflection <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(reflection)}
\end{Highlighting}
\end{Shaded}

\includegraphics{figures/fig_correct_tracks.png}

Reducing large-scale positioning error in a movement track. \textbf{(a)} Positioning error may affect single observations resulting in point outliers or `spikes' (red points), but it may also affect continuous susbets of a track, which we call a \texttt{prolonged\ spike\textquotesingle{}\ (purple\ triangles).\ While\ the\ former\ may\ be\ targeted\ by\ filtering\ on\ appropriate\ covariates\ such\ as\ speed\ and\ turning\ angle\ using\ the}atlastools\texttt{function}atl\_filter\_covariates\texttt{,\ the\ latter\ cannot\ be\ effectively\ corrected\ by\ targeting\ each\ coordinate\ pair\ in\ isolation.\ **(b)**\ The}atlastools\texttt{function}atl\_remove\_reflections` to identify prolonged spikes (red triangles) in tracking data is an illustrative example of targeting positioning errors that affect track subsets. While this method returns the canonical track without the prolonged spike (blue points) in this example, users are cautioned to frequently check this and similar semi-supervised algorithms' results.

Finally, we export the data without spikes and reflections to be used later.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{fwrite}\NormalTok{(data\_no\_reflection, }\DataTypeTok{file =} \StringTok{"data/data\_no\_reflection.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{reducing-small-scale-positioning-error-and-aggregating-data}{%
\chapter{Reducing Small-Scale Positioning Error and Aggregating Data}\label{reducing-small-scale-positioning-error-and-aggregating-data}}

Once large-scale positioning errors have been removed using filters on bounds or on unrealistic movement, small-scale positioning errors may remain. Why is this the case?

The simple reason is that high-frequency tracking has a large component of error relative to the real movement of an individual, and the error / movement ratio increases with the frequency of tracking. (Noonan et al. \protect\hyperlink{ref-noonan2019}{2019}) has a better explanation, which users are strongly recommended to read.

The effect of these small-scale errors for high-frequency data is to cause overestimations of straight-line displacement (distance between two positions) and speed. If users were to apply a very strict speed filter, many of these positions would be lost. This loss of information would certainly reduce the sampling frequency, which would lead to better estimates of total distance and mean speed. However, while \emph{total} distance and \emph{mean} speed estimates would improve, the instantaneous distance and speed estimates would remain biased, which will most likely lead to erroneous inferences from the data.

However, users can turn the high-frequency of their data against itself by applying a simple smooth. The median smooth is recommended because it is robust to outliers, and is very well understood.

\hypertarget{preparing-libraries-and-data}{%
\section{Preparing Libraries and Data}\label{preparing-libraries-and-data}}

We first load some useful packages and define a colour palette.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# prep libs}
\KeywordTok{library}\NormalTok{(data.table)}
\KeywordTok{library}\NormalTok{(atlastools)}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{library}\NormalTok{(patchwork)}

\CommentTok{\# prepare a palette}
\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We load both the simulated, canonical data (which has no errors), as well as the data with errors which has been pre-processed to remove large-scale errors.

We also assign a variable, \texttt{window\_size}, which will help identify the datasets when comparing median smooth moving window (\(K\)) sizes.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# read in the data and set the window size variable}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_sim.csv"}\NormalTok{)[}\DecValTok{5000}\OperatorTok{:}\DecValTok{10000}\NormalTok{, ]}
\NormalTok{data[, window\_size }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\OtherTok{NA}\NormalTok{]}

\CommentTok{\# data with small scale errors but no reflections or outliers}
\NormalTok{data\_errors <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_no\_reflection.csv"}\NormalTok{)}
\NormalTok{data\_errors[, window\_size }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\DecValTok{0}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\hypertarget{median-smoothing}{%
\section{Median Smoothing}\label{median-smoothing}}

We apply a median smooth with four different values of \(K\), the moving window size --- 3, 5, 11, and 21 positions.
\(K\) must be an odd number, but apart from that there is no correct magnitude of \(K\).
Very large \(K\) will lead to unrealistic tracks, while small \(K\) will not result in much reduction of error.
Users are encouraged to plot their data before and after smoothing to examine the effect of different window sizes.

The \texttt{atlastools} function \texttt{atl\_median\_smooth} is quite fast and users can readily try multiple \(K\) values as shown in the example below.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# smooth the data over four K values}
\NormalTok{list\_of\_smooths <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{11}\NormalTok{, }\DecValTok{21}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(K) \{}

\NormalTok{  data\_copy <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data\_errors)}

\NormalTok{  data\_copy <{-}}\StringTok{ }\KeywordTok{atl\_median\_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_copy,}
                    \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
                    \DataTypeTok{y =} \StringTok{"y"}\NormalTok{,}
                    \DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
                    \DataTypeTok{moving\_window =}\NormalTok{ K)}

\NormalTok{  data\_copy[, window\_size }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{K]}
\NormalTok{\})}
\end{Highlighting}
\end{Shaded}

We save the 11 point smoothed data.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{fwrite}\NormalTok{(list\_of\_smooths[[}\DecValTok{3}\NormalTok{]], }\DataTypeTok{file =} \StringTok{"data/data\_smooth.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We prepare the data for plotting.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# bind list after offset}
\NormalTok{data\_plot <{-}}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(df, offset) \{}
\NormalTok{  df[, x }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{offset]}
\NormalTok{\}, list\_of\_smooths, }\KeywordTok{seq}\NormalTok{(}\FloatTok{0.4}\NormalTok{, }\FloatTok{1.25}\NormalTok{, }\DataTypeTok{length.out =} \DecValTok{4}\NormalTok{),}
\DataTypeTok{SIMPLIFY =}\NormalTok{ F)}

\NormalTok{data\_plot <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_plot)}
\end{Highlighting}
\end{Shaded}

We prepare a plot of the smoothed data.

\hypertarget{thinning-data-by-aggregation}{%
\section{Thinning Data by Aggregation}\label{thinning-data-by-aggregation}}

Evenly thinning data is a good idea if statistical methods require even sampling, or if the volume of data is too large for statistical packages to efficiently handle it. In \texttt{R}, both may be true at once.

Here, we demonstrate thinning by aggregation on data that has been median smoothed using a \(K\) of 11.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# choose the 11 point median smooth data}
\NormalTok{data\_agg <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(list\_of\_smooths[[}\DecValTok{3}\NormalTok{]])}

\CommentTok{\# get list of aggregated data}
\NormalTok{list\_of\_agg <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{120}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(z) \{}

\NormalTok{  data\_return <{-}}\StringTok{ }\KeywordTok{atl\_thin\_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_agg,}
                            \DataTypeTok{interval =}\NormalTok{ z,}
                            \DataTypeTok{method =} \StringTok{"aggregate"}\NormalTok{)}

\NormalTok{  data\_return[, interval }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{z]}

  \KeywordTok{return}\NormalTok{(data\_return)}
\NormalTok{\})}

\CommentTok{\# get mean speed estimate and sd}
\NormalTok{speed\_agg\_smooth <{-}}\StringTok{ }
\StringTok{  }\KeywordTok{lapply}\NormalTok{(list\_of\_agg, }\ControlFlowTok{function}\NormalTok{(df) \{}
    \KeywordTok{na.omit}\NormalTok{(df)}
\NormalTok{    df[, speed }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_get\_speed}\NormalTok{(df)]}
\NormalTok{    df[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{median =} \KeywordTok{median}\NormalTok{(speed, }\DataTypeTok{na.rm =}\NormalTok{ T),}
              \DataTypeTok{sd =} \KeywordTok{sd}\NormalTok{(speed, }\DataTypeTok{na.rm =}\NormalTok{ T),}
              \DataTypeTok{interval =} \KeywordTok{first}\NormalTok{(interval))]}
\NormalTok{  \})}

\CommentTok{\# bind}
\NormalTok{speed\_agg\_smooth <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(speed\_agg\_smooth)}
\end{Highlighting}
\end{Shaded}

\hypertarget{aggregation-before-reducing-positioning-errors}{%
\section{Aggregation Before Reducing Positioning Errors}\label{aggregation-before-reducing-positioning-errors}}

Users may rightly wonder whether they can get away with aggregating their data, using a median aggregation function, and reduce data volumes, correct uneven sampling frequency, and reduce large-scale errors all in one go.

The answer to most questions in ecology is, ``It depends''. Median aggregation before correcting positioning errors can indeed be advantageous; for instance for faster visualisation while preserving the broad structure of a track.

However, there are drawbacks. The main one is that information is lost in the aggregation process, lending less power to steps such as smoothing applied after aggregation.
Further, aggregation (and indeed any kind of thinning) results in significantly different estimates of speed and distance from the real speed.

We show the effect of aggregating before any error correction here.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# read data with errors}
\NormalTok{data\_errors <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_errors.csv"}\NormalTok{)}

\CommentTok{\# aggregate before correction}
\NormalTok{list\_of\_agg <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{120}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(z) \{}
\NormalTok{  data\_return <{-}}\StringTok{ }\KeywordTok{atl\_thin\_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_errors,}
                            \DataTypeTok{interval =}\NormalTok{ z,}
                            \DataTypeTok{method =} \StringTok{"aggregate"}\NormalTok{)}
\NormalTok{  data\_return[, interval }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{z]}
\NormalTok{  data\_return[, speed }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_get\_speed}\NormalTok{(data\_return)]}
  \KeywordTok{return}\NormalTok{(data\_return)}
\NormalTok{\})}

\CommentTok{\# get real speed}
\NormalTok{data[, speed }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_get\_speed}\NormalTok{(data)]}
\end{Highlighting}
\end{Shaded}

\hypertarget{comparing-median-aggregation-with-and-without-smoothing}{%
\subsection{Comparing median aggregation with and without smoothing}\label{comparing-median-aggregation-with-and-without-smoothing}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# now plot distribution of speed}
\NormalTok{data\_agg <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(list\_of\_agg)}
\end{Highlighting}
\end{Shaded}

\includegraphics{figures/fig_aggregate_errors.png}

Thinning a movement track using median aggregation preserves track structure, but affects essential track metrics such as speed.
\textbf{(a, b)} Movement tracks with a canonical interval of 1s aggregated over intervals of \textbf{(a)} 10 and \textbf{(b)} 30 seconds, without removing large- or small-scale positioning errors. All symbols represent positions in the aggregated track, with the size of the symbol representing the standard deviation at each position.
Blue crosses represent positions with speed \(\leq\) the 95\^{}th percentile of canonical speeds, while red triangles represent positions with speed \(\geq\) 95\^{}th percentile of canonical speeds.
\textbf{(c)} Boxplot of instantaneous speeds after median aggregation of a 1s interval track over intervals of 3, 10, 30, and 120 seconds, but without the removal of positioning errors. The mean and 95\^{}th percentile of speed in the canonical track are shown as solid and dashed lines, respectively.
Aggregation without reducing positioning errors can result in speed estimates that are substantially different from the true speed.

\hypertarget{residence-patches-and-their-construction}{%
\chapter{Residence patches and their construction}\label{residence-patches-and-their-construction}}

\hypertarget{prepare-libraries}{%
\section*{Prepare libraries}\label{prepare-libraries}}
\addcontentsline{toc}{section}{Prepare libraries}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(data.table)}
\KeywordTok{library}\NormalTok{(atlastools)}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{library}\NormalTok{(patchwork)}

\CommentTok{\# for residence time}
\KeywordTok{library}\NormalTok{(recurse)}

\CommentTok{\# prepare a palette}
\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{an-example-with-simulated-data}{%
\section{An example with simulated data}\label{an-example-with-simulated-data}}

\hypertarget{read-data-and-classify}{%
\subsection{Read data and classify}\label{read-data-and-classify}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# read patch data}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_for\_res\_patch.csv"}\NormalTok{)}

\CommentTok{\# filter}
\NormalTok{data <{-}}\StringTok{ }\NormalTok{data[y }\OperatorTok{<}\StringTok{ }\DecValTok{21}\NormalTok{, ]}

\CommentTok{\# do recurse}
\NormalTok{data\_recurse <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(x, y, time, id)],}
                              \DataTypeTok{radius =} \DecValTok{1}\NormalTok{)}

\CommentTok{\# assign residence time}
\NormalTok{data[, residence\_time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{data\_recurse}\OperatorTok{$}\NormalTok{residenceTime]}
\end{Highlighting}
\end{Shaded}

We first plot a figure of residence time per positions, and residence time per timestamp.

We construct residence patches from data where residence time is \textgreater{} 0.15.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# make residence patch}
\NormalTok{patch <{-}}\StringTok{ }\KeywordTok{atl\_res\_patch}\NormalTok{(data[residence\_time }\OperatorTok{>}\StringTok{ }\FloatTok{0.04}\NormalTok{, ], }
                       \DataTypeTok{buffer\_radius =} \FloatTok{0.1}\NormalTok{, }
                       \DataTypeTok{lim\_spat\_indep =} \DecValTok{1}\NormalTok{, }
                       \DataTypeTok{lim\_time\_indep =} \DecValTok{30}\NormalTok{)}

\CommentTok{\# get spatial representation}
\NormalTok{patch\_sf <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(}\DataTypeTok{patch\_data =}\NormalTok{ patch,}
                              \DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{, }
                              \DataTypeTok{buffer\_radius =} \FloatTok{0.15}\NormalTok{)}

\CommentTok{\# get summary data}
\NormalTok{patch\_summary <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(}\DataTypeTok{patch\_data =}\NormalTok{ patch,}
                                   \DataTypeTok{which\_data =} \StringTok{"summary"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{plot-classified-residence-patches}{%
\subsection{Plot classified residence patches}\label{plot-classified-residence-patches}}

We arrange the figures together.

\includegraphics{figures/fig_residence.png}

\hypertarget{processing-calibration-data}{%
\chapter{Processing calibration data}\label{processing-calibration-data}}

\hypertarget{prepare-libraries}{%
\section{Prepare libraries}\label{prepare-libraries}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# load libs}
\KeywordTok{library}\NormalTok{(data.table)}
\KeywordTok{library}\NormalTok{(atlastools)}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{library}\NormalTok{(patchwork)}

\CommentTok{\# prepare a palette}
\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{preliminary-visualisation}{%
\section{Preliminary visualisation}\label{preliminary-visualisation}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# read and plot example data}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/atlas1060\_allTrials\_annotated.csv"}\NormalTok{)}
\NormalTok{data\_raw <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
\end{Highlighting}
\end{Shaded}

\hypertarget{filter-by-bounding-box}{%
\section{Filter by bounding box}\label{filter-by-bounding-box}}

Save an unprocessed copy.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
\end{Highlighting}
\end{Shaded}

Filter by a bounding box.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# remove inside must be set to falses}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_filter\_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data, }
                          \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
                          \DataTypeTok{x\_range =} \KeywordTok{c}\NormalTok{(}\DecValTok{645000}\NormalTok{, }\KeywordTok{max}\NormalTok{(data}\OperatorTok{$}\NormalTok{x)), }
                          \DataTypeTok{remove\_inside =} \OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{filter-trajectories}{%
\section{Filter trajectories}\label{filter-trajectories}}

\hypertarget{handle-time}{%
\subsection{Handle time}\label{handle-time}}

Time in ATLAS tracking is counted in milliseconds and is represented by a 64-bit integer (type \texttt{long}), which is not natively supported in R; it will instead be converted to a \texttt{numeric}, or \texttt{double}.

This is not what is intended, but it works. The \texttt{bit64} package can help handle 64-bit integers if you want to keep to intended type.

A further issue is that 64-bit integers (whether represented as \texttt{bit64} or \texttt{double}) do not yield meaninful results when you try to convert them to a date-time object, such as of the class \texttt{POSIXct}.

This is because \texttt{as.POSIXct} fails when trying to work with 64-bit integers (it cannot interpret this type), and returns a date many thousands of years in the future (approx. 52,000 CE) if the time column is converted to \texttt{numeric}.

There are two possible solutions. The parsimonious one is to convert the 64-bit number to a 32-bit short integer (dividing by 1000), or to use the \texttt{nanotime} package.

The conversion method loses an imperceptible amount of precision. The \texttt{nanotime} requires installing another package. The first method is shown here.

In the spirit of not destroying data, we create a second lower-case column called \texttt{time}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# divide by 1000, convert to integer, then convert to POSIXct}
\NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

\hypertarget{add-speed-and-turning-angle}{%
\subsection{Add speed and turning angle}\label{add-speed-and-turning-angle}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# add incoming and outgoing speed}
\NormalTok{data[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{ (}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }
                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
             \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}

\CommentTok{\# add turning angle}
\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_turning\_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
\end{Highlighting}
\end{Shaded}

\hypertarget{get-95th-percentile-of-speed-and-angle}{%
\subsection{Get 95th percentile of speed and angle}\label{get-95th-percentile-of-speed-and-angle}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# use sapply}
\NormalTok{speed\_angle\_thresholds <{-}}\StringTok{ }
\StringTok{  }\KeywordTok{sapply}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(speed\_in, speed\_out, angle)], }
\NormalTok{       quantile, }\DataTypeTok{probs =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{na.rm =}\NormalTok{ T)}
\end{Highlighting}
\end{Shaded}

\hypertarget{plot-to-see-speeds}{%
\subsection{Plot to see speeds}\label{plot-to-see-speeds}}

\includegraphics{figures/fig_speed_outlier.png}

\hypertarget{filter-on-speed}{%
\subsection{Filter on speed}\label{filter-on-speed}}

Here we use a speed threshold of 15 m/s, the fastest known boat speed.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# make a copy}
\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}

\CommentTok{\# remove speed outliers}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
            \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(speed\_in < 15 \& speed\_out < 15)"}\NormalTok{))}

\CommentTok{\# recalculate speed and angle}
\NormalTok{data[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{ (}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }
                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
             \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}

\CommentTok{\# add turning angle}
\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_turning\_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
\end{Highlighting}
\end{Shaded}

\hypertarget{smoothing-the-trajectory}{%
\section{Smoothing the trajectory}\label{smoothing-the-trajectory}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# apply a 5 point median smooth, first make a copy}
\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}

\CommentTok{\# now apply the smooth}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_median\_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
                          \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }\DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
                          \DataTypeTok{moving\_window =} \DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{plot-pre-processing-steps}{%
\subsection{Plot pre-processing steps}\label{plot-pre-processing-steps}}

\includegraphics{figures/fig_walkthrough.png}

\hypertarget{making-residence-patches}{%
\section{Making residence patches}\label{making-residence-patches}}

\hypertarget{prepare-data}{%
\subsection{Prepare data}\label{prepare-data}}

An indicator of individual residence at or near a position can be useful when attempting to identify residence patches. Positions can be filtered on a metric such as residence time (see Bracis et al.~2018).

In this dataset, residence positions are marked in the \texttt{tID} column as beginning with \texttt{WP}. These can be extracted and converted into residence patches.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# make a data copy}
\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
\end{Highlighting}
\end{Shaded}

\hypertarget{calculate-residence-time}{%
\subsection{Calculate residence time}\label{calculate-residence-time}}

Subset for boating segments: 2020-08-24 11:13:00 -- 2020-08-24 12:36:11.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# load recurse}
\KeywordTok{library}\NormalTok{(recurse)}
\NormalTok{data\_recurse <{-}}\StringTok{ }\NormalTok{data\_unproc[, }\KeywordTok{list}\NormalTok{(x, y, time, TAG, UTCtime)]}

\CommentTok{\# thin the data}
\NormalTok{data\_recurse <{-}}\StringTok{ }\KeywordTok{atl\_thin\_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_recurse,}
                              \DataTypeTok{interval =} \DecValTok{30}\NormalTok{, }\DataTypeTok{id\_columns =} \StringTok{"TAG"}\NormalTok{,}
                              \DataTypeTok{method =} \StringTok{"resample"}\NormalTok{)}

\CommentTok{\# get 4 column data}
\NormalTok{data\_recurse <{-}}\StringTok{ }\NormalTok{data\_recurse[, }\KeywordTok{list}\NormalTok{(x, y, time, TAG)]}

\CommentTok{\# get recurse data for a 10m radius}
\NormalTok{recurse\_stats <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data\_recurse,}
                               \DataTypeTok{radius =} \DecValTok{50}\NormalTok{, }\DataTypeTok{timeunits =} \StringTok{"mins"}\NormalTok{)}

\CommentTok{\# assign to recurse data}
\NormalTok{data\_recurse[, res\_time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{recurse\_stats}\OperatorTok{$}\NormalTok{residenceTime]}

\CommentTok{\# save recurse data}
\KeywordTok{fwrite}\NormalTok{(data\_recurse, }\DataTypeTok{file =} \StringTok{"data/data\_calib\_recurse.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Read in the data.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{data\_recurse <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_calib\_recurse.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Subset waypoint data.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(stringi)}
\NormalTok{data\_res <{-}}\StringTok{ }\NormalTok{data\_unproc[}\KeywordTok{stri\_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

\hypertarget{plot-residence-time}{%
\subsection{Plot residence time}\label{plot-residence-time}}

\hypertarget{run-residence-patch-method}{%
\subsection{Run residence patch method}\label{run-residence-patch-method}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# assign id as tag}
\NormalTok{data\_recurse[, id }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.character}\NormalTok{(TAG)]}

\CommentTok{\# subset on 5 minute residence time}
\NormalTok{data\_recurse\_subset <{-}}\StringTok{ }\NormalTok{data\_recurse[res\_time }\OperatorTok{>=}\StringTok{ }\DecValTok{5}\NormalTok{, ]}

\CommentTok{\# on known residence points}
\NormalTok{patch\_res\_known <{-}}\StringTok{ }\KeywordTok{atl\_res\_patch}\NormalTok{(data\_recurse\_subset, }
                                \DataTypeTok{buffer\_radius =} \DecValTok{10}\NormalTok{,}
                                \DataTypeTok{lim\_spat\_indep =} \DecValTok{100}\NormalTok{,}
                                \DataTypeTok{lim\_time\_indep =} \DecValTok{5}\NormalTok{,}
                                \DataTypeTok{min\_fixes =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{get-spatial-and-summary-objects}{%
\subsection{Get spatial and summary objects}\label{get-spatial-and-summary-objects}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# for the known and unkniwn patches}
\NormalTok{patch\_sf\_data <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(patch\_res\_known, }\DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{,}
                                   \DataTypeTok{buffer\_radius =} \DecValTok{20}\NormalTok{)}

\CommentTok{\# assign crs}
\NormalTok{sf}\OperatorTok{::}\KeywordTok{st\_crs}\NormalTok{(patch\_sf\_data) <{-}}\StringTok{ }\DecValTok{32631}

\CommentTok{\# get summary data}
\NormalTok{patch\_summary\_data <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(patch\_res\_known, }\DataTypeTok{which\_data =} \StringTok{"summary"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{get-waypoint-centroids}{%
\subsection{Get waypoint centroids}\label{get-waypoint-centroids}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get centroid}
\NormalTok{data\_res\_summary <{-}}\StringTok{ }\NormalTok{data\_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{x\_median =} \KeywordTok{median}\NormalTok{(x),}
                                    \DataTypeTok{y\_median =} \KeywordTok{median}\NormalTok{(y),}
                                    \DataTypeTok{t\_median =} \KeywordTok{median}\NormalTok{(time)),}
\NormalTok{                             by =}\StringTok{ "tID"}\NormalTok{]}

\CommentTok{\# now get times 10 mins before and after}
\NormalTok{data\_res\_summary[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{t\_min =}\NormalTok{ t\_median }\OperatorTok{{-}}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{),}
                        \DataTypeTok{t\_max =}\NormalTok{ t\_median }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{))]}

\CommentTok{\# make a list of positions 10min before and after}
\NormalTok{wp\_data <{-}}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(l, u, mx, my) \{}
\NormalTok{  tmp\_data <{-}}\StringTok{ }\NormalTok{data\_unproc[}\KeywordTok{inrange}\NormalTok{(time, l, u)]}
\NormalTok{  tmp\_data[, distance }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{sqrt}\NormalTok{((mx }\OperatorTok{{-}}\StringTok{ }\NormalTok{x)}\OperatorTok{\^{}}\DecValTok{2} \OperatorTok{+}\StringTok{ }\NormalTok{(my }\OperatorTok{{-}}\StringTok{ }\NormalTok{y)}\OperatorTok{\^{}}\DecValTok{2}\NormalTok{)]}
  \CommentTok{\# keep within 50}
\NormalTok{  tmp\_data <{-}}\StringTok{ }\NormalTok{tmp\_data[distance }\OperatorTok{<=}\StringTok{ }\DecValTok{50}\NormalTok{, ]}
  \CommentTok{\# get duration}
  \KeywordTok{return}\NormalTok{(}\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(tmp\_data}\OperatorTok{$}\NormalTok{time)))}
\NormalTok{\}, data\_res\_summary}\OperatorTok{$}\NormalTok{t\_min, data\_res\_summary}\OperatorTok{$}\NormalTok{t\_max,}
\NormalTok{   data\_res\_summary}\OperatorTok{$}\NormalTok{x\_median, data\_res\_summary}\OperatorTok{$}\NormalTok{y\_median,}
\DataTypeTok{SIMPLIFY =} \OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{read-griend}{%
\subsection{Read Griend}\label{read-griend}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# read griend}
\NormalTok{griend <{-}}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st\_read}\NormalTok{(}\StringTok{"data/griend\_polygon/griend\_polygon.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We prepare a plot of the residence patches.

\hypertarget{compare-patch-metrics}{%
\section{Compare patch metrics}\label{compare-patch-metrics}}

\hypertarget{compare-known-patches}{%
\subsection{Compare known patches}\label{compare-known-patches}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get known patch summary}
\NormalTok{data\_res <{-}}\StringTok{ }\NormalTok{data\_unproc[stringi}\OperatorTok{::}\KeywordTok{stri\_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{), ]}

\CommentTok{\# get waypoint summary}
\NormalTok{patch\_summary\_real <{-}}\StringTok{ }\NormalTok{data\_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{nfixes\_real =}\NormalTok{ .N,}
                                      \DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
                                      \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{)), }
\NormalTok{                               by =}\StringTok{ "tID"}\NormalTok{]}

\CommentTok{\# add real duration}
\NormalTok{patch\_summary\_real[, duration\_real }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{unlist}\NormalTok{(wp\_data)]}
\CommentTok{\# round median coordinate for inferred patches}
\NormalTok{patch\_summary\_inferred <{-}}\StringTok{ }
\StringTok{  }\NormalTok{patch\_summary\_data[, }
                     \KeywordTok{c}\NormalTok{(}\StringTok{"x\_median"}\NormalTok{, }\StringTok{"y\_median"}\NormalTok{, }
                       \StringTok{"nfixes"}\NormalTok{, }\StringTok{"duration"}\NormalTok{, }\StringTok{"patch"}\NormalTok{)}
\NormalTok{                     ][, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(x\_median, }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
                              \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(y\_median, }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{))]}

\CommentTok{\# join with respatch summary}
\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }
\StringTok{  }\KeywordTok{merge}\NormalTok{(patch\_summary\_real,}
\NormalTok{        patch\_summary\_inferred, }
        \DataTypeTok{on =} \KeywordTok{c}\NormalTok{(}\StringTok{"x\_median"}\NormalTok{, }\StringTok{"y\_median"}\NormalTok{),}
        \DataTypeTok{all.x =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{all.y =} \OtherTok{TRUE}\NormalTok{)}

\NormalTok{patch\_summary\_compare[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(tID)]}

\CommentTok{\# drop nas}
\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(patch\_summary\_compare)}

\CommentTok{\# drop patch around WP080}
\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }\NormalTok{patch\_summary\_compare[tID }\OperatorTok{!=}\StringTok{ "WP080"}\NormalTok{, ]}
\end{Highlighting}
\end{Shaded}

12 patches are identified where there are no waypoints, while 2 waypoints are not identified as patches. These waypoints consisted of 6 and 15 (WP098 and WP092) positions respectively, and were lost when the data were aggregated to 30 second intervals.

\hypertarget{plot-durations-comparisons}{%
\subsection{Plot durations comparisons}\label{plot-durations-comparisons}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get linear model}
\NormalTok{model\_duration <{-}}\StringTok{ }\KeywordTok{lm}\NormalTok{(duration\_real }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{duration,}
                     \DataTypeTok{data =}\NormalTok{ patch\_summary\_compare)}

\CommentTok{\# get R2}
\KeywordTok{summary}\NormalTok{(model\_duration)}
\end{Highlighting}
\end{Shaded}

\hypertarget{join-patch-map-and-duration-comparison-figure}{%
\subsection{Join patch map and duration comparison figure}\label{join-patch-map-and-duration-comparison-figure}}

\includegraphics{figures/fig_calib_residence_patch.png}

\hypertarget{processing-egyptian-fruit-bat-tracks}{%
\chapter{Processing Egyptian fruit bat tracks}\label{processing-egyptian-fruit-bat-tracks}}

We show the pre-processing pipeline at work on the tracks of three Egyptian fruit bats (\emph{Rousettus aegyptiacus}), and construct residence patches.

\hypertarget{prepare-libraries}{%
\section{Prepare libraries}\label{prepare-libraries}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# load libs}
\KeywordTok{library}\NormalTok{(data.table)}
\KeywordTok{library}\NormalTok{(RSQLite)}
\KeywordTok{library}\NormalTok{(atlastools)}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{library}\NormalTok{(patchwork)}

\CommentTok{\# prepare a palette}
\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{read-bat-data}{%
\section{Read bat data}\label{read-bat-data}}

Read the bat data and convert to a csv file.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# prepare the connection}
\NormalTok{con <{-}}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}\DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(), }\DataTypeTok{dbname =} \StringTok{"data/Three\_example\_bats.sql"}\NormalTok{)}

\CommentTok{\# list the tables}
\NormalTok{table\_name <{-}}\StringTok{ }\KeywordTok{dbListTables}\NormalTok{(con)}

\CommentTok{\# prepare to query all tables}
\NormalTok{query <{-}}\StringTok{ }\KeywordTok{sprintf}\NormalTok{(}\StringTok{\textquotesingle{}select * from }\CharTok{\textbackslash{}"}\StringTok{\%s}\CharTok{\textbackslash{}"}\StringTok{\textquotesingle{}}\NormalTok{, table\_name)}

\CommentTok{\# query the database}
\NormalTok{data <{-}}\StringTok{ }\KeywordTok{dbGetQuery}\NormalTok{(}\DataTypeTok{conn =}\NormalTok{ con, }\DataTypeTok{statement =}\NormalTok{ query)}

\CommentTok{\# disconnect from database}
\KeywordTok{dbDisconnect}\NormalTok{(con)}
\end{Highlighting}
\end{Shaded}

Convert data to csv.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# convert data to datatable}
\KeywordTok{setDT}\NormalTok{(data)}

\CommentTok{\# write data for QGIS}
\KeywordTok{fwrite}\NormalTok{(data, }\DataTypeTok{file =} \StringTok{"data/bat\_data.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{sanity-check-plot-bat-data}{%
\section{Sanity check: Plot bat data}\label{sanity-check-plot-bat-data}}

Save the raw data plot.

Show the plot.

\includegraphics{figures/fig_bat_raw.png}

\hypertarget{prepare-data-for-filtering}{%
\section{Prepare data for filtering}\label{prepare-data-for-filtering}}

\hypertarget{prepare-data-per-individual}{%
\subsection{Prepare data per individual}\label{prepare-data-per-individual}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# split bat data by tag}
\CommentTok{\# first make a copy using the data.table function copy}
\CommentTok{\# this prevents the orignal data from being modified by atlastools}
\CommentTok{\# functions which DO MODIFY BY REFERENCE!}
\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}

\CommentTok{\# now split}
\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_split, data}\OperatorTok{$}\NormalTok{TAG)}
\end{Highlighting}
\end{Shaded}

\hypertarget{filter-by-covariates}{%
\section{Filter by covariates}\label{filter-by-covariates}}

No natural bounds suggest themselves, and we proceed to filter by covariates, since point outliers are obviously visible.

We use filter out positions with \texttt{SD\ \textgreater{}\ 20} and positions calculated using only 3 base stations.

We use the function \texttt{atl\_filter\_covariates}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get SD.}
\CommentTok{\# since the data are data.tables, no assignment is necessary}
\KeywordTok{invisible}\NormalTok{(}
  \KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
\NormalTok{    dt[, SD }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(VARX }\OperatorTok{+}\StringTok{ }\NormalTok{VARY }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{COVXY))]}
\NormalTok{  \})}
\NormalTok{)}

\CommentTok{\# filter for SD <= 20}
\CommentTok{\# here, reassignment is necessary as rows are being removed}
\CommentTok{\# the atl\_filter\_covariates function could have been used here}
\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}
    \DataTypeTok{data =}\NormalTok{ dt,}
    \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"SD <= 20"}\NormalTok{,}
                \StringTok{"NBS > 3"}\NormalTok{)}
\NormalTok{  )}
\NormalTok{\})}

\CommentTok{\# check whether the filter has worked}
\KeywordTok{invisible}\NormalTok{(}
  \KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
\NormalTok{    assertthat}\OperatorTok{::}\KeywordTok{assert\_that}\NormalTok{(}\KeywordTok{min}\NormalTok{(dt}\OperatorTok{$}\NormalTok{SD) }\OperatorTok{<=}\StringTok{ }\DecValTok{20}\NormalTok{,}
                            \DataTypeTok{msg =} \StringTok{"some SDs above 20 remain"}\NormalTok{)}
\NormalTok{    assertthat}\OperatorTok{::}\KeywordTok{assert\_that}\NormalTok{(}\KeywordTok{min}\NormalTok{(dt}\OperatorTok{$}\NormalTok{NBS) }\OperatorTok{>}\StringTok{ }\DecValTok{3}\NormalTok{,}
                            \DataTypeTok{msg =} \StringTok{"some NBS below 3 remain"}\NormalTok{)}
\NormalTok{  \})}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{sanity-check-plot-filtered-data}{%
\subsection{Sanity check: Plot filtered data}\label{sanity-check-plot-filtered-data}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# bind all individuals together}
\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_split)}
\end{Highlighting}
\end{Shaded}

Save the filtered data plot.

Show the plot.

\includegraphics{figures/fig_bat_filter_cov.png}

\hypertarget{filter-by-speed}{%
\section{Filter by speed}\label{filter-by-speed}}

Some point outliers remain, and should be removed using a speed filter.

First we calculate speeds.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# calculate speed on split data once more}
\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_split, data\_split}\OperatorTok{$}\NormalTok{TAG)}

\CommentTok{\# get speeds as with SD, no reassignment required for columns}
\KeywordTok{invisible}\NormalTok{(}
  \KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}

    \CommentTok{\# first process time to seconds}
    \CommentTok{\# assign to a new column}
\NormalTok{    dt[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{floor}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}

\NormalTok{    dt[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(dt, }
                                       \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{, }
                                       \DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
                                       \DataTypeTok{type =} \StringTok{"in"}\NormalTok{),}
              \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(dt, }
                                       \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{, }
                                       \DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
                                       \DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
\NormalTok{  \})}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Now filter for speeds \textgreater{} 20 m/s (around 70 km/h).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# filter speeds}
\CommentTok{\# reassignment is required here}
\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(dt, }\DataTypeTok{cols =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed\_in"}\NormalTok{, }\StringTok{"speed\_out"}\NormalTok{))}

\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dt,}
                              \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed\_in <= 20"}\NormalTok{,}
                                          \StringTok{"speed\_out <= 20"}\NormalTok{))}
\NormalTok{\})}
\end{Highlighting}
\end{Shaded}

\hypertarget{sanity-check-plot-speed-filtered-data}{%
\subsection{Sanity check: Plot speed filtered data}\label{sanity-check-plot-speed-filtered-data}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# bind all individuals together}
\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_split)}
\end{Highlighting}
\end{Shaded}

Save the speed filtered data plot.

Show the plot.

\includegraphics{figures/fig_bat_filter_speed.png}

\hypertarget{median-smoothing}{%
\section{Median smoothing}\label{median-smoothing}}

Apply a 5 point median smooth to the data.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# since the function modifies in place, we shall make a copy}
\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data\_split)}

\CommentTok{\# split the data again}
\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_smooth, data\_smooth}\OperatorTok{$}\NormalTok{TAG)}
\end{Highlighting}
\end{Shaded}

Remember, \textbf{\texttt{atl\_median\_smooth} MODIFIES IN PLACE}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# apply the median smooth to each list element}
\CommentTok{\# no reassignment is required as THE FUNCTION MODIFIES IN PLACE!}
\KeywordTok{invisible}\NormalTok{(}
  \CommentTok{\# the function arguments to atl\_median\_smooth}
  \CommentTok{\# can be passed directly in lapply}
  \KeywordTok{lapply}\NormalTok{(data\_smooth, atl\_median\_smooth,}
         \DataTypeTok{time =} \StringTok{"time"}\NormalTok{, }\DataTypeTok{moving\_window =} \DecValTok{5}\NormalTok{)}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{sanity-check-plot-smoothed-data}{%
\subsection{Sanity check: Plot smoothed data}\label{sanity-check-plot-smoothed-data}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# recombine split up data that has been smoothed}
\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_smooth)}
\end{Highlighting}
\end{Shaded}

Save the smoothed data plot.

Show the plot.

\includegraphics{figures/fig_bat_smooth.png}

\hypertarget{making-residence-patches}{%
\section{Making residence patches}\label{making-residence-patches}}

\hypertarget{calculating-residence-time}{%
\subsection{Calculating residence time}\label{calculating-residence-time}}

First, the data is put through the \texttt{recurse} package to get residence time.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# load recurse}
\KeywordTok{library}\NormalTok{(recurse)}

\CommentTok{\# split the data}
\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_smooth, data\_smooth}\OperatorTok{$}\NormalTok{TAG)}
\end{Highlighting}
\end{Shaded}

Get residence time. Since bats may revisit the same features, we want to prevent confusion between frequent revisits and prolonged residence.

For this, we stop summing residence times within Z metres of a location if the animal exited the area for one hour or more.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get residence times}

\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_smooth, }\ControlFlowTok{function}\NormalTok{(dt) \{}
  \CommentTok{\# do basic recurse}
\NormalTok{  dt\_recurse <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(}
    \DataTypeTok{x =}\NormalTok{ dt[, }\KeywordTok{c}\NormalTok{(}\StringTok{"X"}\NormalTok{, }\StringTok{"Y"}\NormalTok{, }\StringTok{"time"}\NormalTok{, }\StringTok{"TAG"}\NormalTok{)],}
    \DataTypeTok{radius =} \DecValTok{50}\NormalTok{,}
    \DataTypeTok{timeunits =} \StringTok{"mins"}
\NormalTok{  )}

  \CommentTok{\# get revisit stats}
\NormalTok{  dt\_recurse <{-}}\StringTok{ }\KeywordTok{setDT}\NormalTok{(}
\NormalTok{    dt\_recurse[[}\StringTok{"revisitStats"}\NormalTok{]]}
\NormalTok{  )}

  \CommentTok{\# count long absences from the area}
\NormalTok{  dt\_recurse[, timeSinceLastVisit }\OperatorTok{:}\ErrorTok{=}
\StringTok{          }\KeywordTok{ifelse}\NormalTok{(}\KeywordTok{is.na}\NormalTok{(timeSinceLastVisit), }\OperatorTok{{-}}\OtherTok{Inf}\NormalTok{, timeSinceLastVisit)]}
\NormalTok{  dt\_recurse[, longAbsenceCounter }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{cumsum}\NormalTok{(timeSinceLastVisit }\OperatorTok{>}\StringTok{ }\DecValTok{60}\NormalTok{),}
\NormalTok{             by =}\StringTok{ }\NormalTok{.(coordIdx)}
\NormalTok{             ]}
  \CommentTok{\# get data before the first long absence of 60 mins}
\NormalTok{  dt\_recurse <{-}}\StringTok{ }\NormalTok{dt\_recurse[longAbsenceCounter }\OperatorTok{<}\StringTok{ }\DecValTok{1}\NormalTok{, ]}

\NormalTok{  dt\_recurse <{-}}\StringTok{ }\NormalTok{dt\_recurse[, }\KeywordTok{list}\NormalTok{(}
    \DataTypeTok{resTime =} \KeywordTok{sum}\NormalTok{(timeInside),}
    \DataTypeTok{fpt =} \KeywordTok{first}\NormalTok{(timeInside),}
    \DataTypeTok{revisits =} \KeywordTok{max}\NormalTok{(visitIdx)}
\NormalTok{  ),}
\NormalTok{  by =}\StringTok{ }\NormalTok{.(coordIdx, x, y)}
\NormalTok{  ]}

  \CommentTok{\# prepare and merge existing data with recursion data}
\NormalTok{  dt[, coordIdx }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\KeywordTok{nrow}\NormalTok{(dt))]}

\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{merge}\NormalTok{(dt, }
\NormalTok{              dt\_recurse[, }\KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{, }\StringTok{"resTime"}\NormalTok{)], }
              \DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{))}

  \KeywordTok{setorderv}\NormalTok{(dt, }\StringTok{"time"}\NormalTok{)}
\NormalTok{\})}
\end{Highlighting}
\end{Shaded}

\hypertarget{sanity-check-residence-time-time-series}{%
\subsection{Sanity check: Residence-time time-series}\label{sanity-check-residence-time-time-series}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# bind the list}
\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_residence)}

\CommentTok{\# get time as human readable}
\NormalTok{data\_residence[, ts }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.POSIXct}\NormalTok{(time, }\DataTypeTok{origin =} \StringTok{"1970{-}01{-}01"}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

Save the speed filtered data plot.

Show the plot.

\includegraphics{figures/fig_bat_res_timeseries.png}

\hypertarget{constructing-residence-patches}{%
\subsection{Constructing residence patches}\label{constructing-residence-patches}}

Split the data and construct residence patches.

Some preparation is required. First, the function requires columns \texttt{x}, \texttt{y},
\texttt{time}, and \texttt{id}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# add an id column}
\NormalTok{data\_residence[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{id =}\NormalTok{ TAG,}
                      \DataTypeTok{x =}\NormalTok{ X, }\DataTypeTok{y =}\NormalTok{ Y)]}

\CommentTok{\# filter for residence time > 5 minutes}
\NormalTok{data\_residence <{-}}\StringTok{ }\NormalTok{data\_residence[resTime }\OperatorTok{>}\StringTok{ }\DecValTok{5}\NormalTok{, ]}

\CommentTok{\# split the data}
\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_residence, data\_residence}\OperatorTok{$}\NormalTok{TAG)}
\end{Highlighting}
\end{Shaded}

Now segment-cluster into residence patches.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# segment into residence patches}
\NormalTok{data\_patches <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_residence, atl\_res\_patch,}
                       \DataTypeTok{buffer\_radius =} \DecValTok{25}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{getting-residence-patch-data}{%
\subsection{Getting residence patch data}\label{getting-residence-patch-data}}

We get the residence patch data as spatial \texttt{sf-MULTIPOLYGON} objects.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get data spatials}
\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_patches, atl\_patch\_summary,}
                        \DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{,}
                        \DataTypeTok{buffer\_radius =} \DecValTok{25}\NormalTok{)}

\CommentTok{\# bind list}
\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_spatials)}

\CommentTok{\# convert to sf}
\KeywordTok{library}\NormalTok{(sf)}
\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{st\_sf}\NormalTok{(data\_spatials, }\DataTypeTok{sf\_column\_name =} \StringTok{"polygons"}\NormalTok{)}

\CommentTok{\# assign a crs}
\KeywordTok{st\_crs}\NormalTok{(data\_spatials) <{-}}\StringTok{ }\KeywordTok{st\_crs}\NormalTok{(}\DecValTok{2039}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{write-patch-spatial-representations}{%
\subsection{Write patch spatial representations}\label{write-patch-spatial-representations}}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st\_write}\NormalTok{(data\_spatials,}
         \DataTypeTok{dsn =} \StringTok{"data/data\_bat\_residence\_patches.gpkg"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Write cleaned bat data.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{data\_clean <{-}}\StringTok{ }\KeywordTok{fwrite}\NormalTok{(}\KeywordTok{rbindlist}\NormalTok{(data\_smooth),}
                     \DataTypeTok{file =} \StringTok{"data/data\_bat\_smooth.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Write patch summary.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# get summary}
\NormalTok{patch\_summary <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_patches, atl\_patch\_summary)}

\CommentTok{\# bind summary}
\NormalTok{patch\_summary <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(patch\_summary)}

\CommentTok{\# write}
\KeywordTok{fwrite}\NormalTok{(patch\_summary,}
       \StringTok{"data/data\_bat\_patch\_summary.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{processed-bat-patches}{%
\section{Processed bat patches}\label{processed-bat-patches}}

This figure made in QGIS.

\includegraphics{figures/fig_bat_patches.png}

\hypertarget{references}{%
\chapter{References}\label{references}}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-bjorneraas2010}{}%
Bjørneraas, Kari, Bram Van Moorter, Christer Moe Rolandsen, and Ivar Herfindal. 2010. ``Screening Global Positioning System Location Data for Errors Using Animal Movement Characteristics.'' \emph{The Journal of Wildlife Management} 74 (6): 1361--6. \url{https://doi.org/10.1111/j.1937-2817.2010.tb01258.x}.

\leavevmode\hypertarget{ref-gurarie2017}{}%
Gurarie, Eliezer, Christen H. Fleming, William F. Fagan, Kristin L. Laidre, Jesús Hernández-Pliego, and Otso Ovaskainen. 2017. ``Correlated Velocity Models as a Fundamental Unit of Animal Movement: Synthesis and Applications.'' \emph{Movement Ecology} 5 (1): 13. \url{https://doi.org/10.1186/s40462-017-0103-3}.

\leavevmode\hypertarget{ref-noonan2019}{}%
Noonan, Michael J., Christen H. Fleming, Thomas S. Akre, Jonathan Drescher-Lehman, Eliezer Gurarie, Autumn-Lynn Harrison, Roland Kays, and Justin M. Calabrese. 2019. ``Scale-Insensitive Estimation of Speed and Distance Traveled from Animal Tracking Data.'' \emph{Movement Ecology} 7 (1): 35. \url{https://doi.org/10.1186/s40462-019-0177-1}.
\end{cslreferences}

\end{document}

### 2020-10-29 03_smoothing_tracks.Rmd

@@
 
-```{r}
+```{r echo=FALSE}
 # prepare data to plot
@@
 
-```{r}
+```{r echo=FALSE}
 ### plot figures
@@
 data_agg <- rbindlist(list_of_agg)
+```
 
+```{r echo=FALSE}
 # show boxplot of speed
@@
 
-```{r}
+```{r echo=FALSE}
 # make combined figure

### 2020-10-29 05_calibration_data.Rmd

@@
 data_raw <- copy(data)
+```
 
+```{r echo=FALSE}
 # plot data
@@
 
-Plot the result.
-
-```{r}
+```{r echo=FALSE}
 # plot data
@@
 
-```{r}
+```{r echo=FALSE}
 # plot filtered data
@@
 ```
+
+![](figures/fig_speed_outlier.png)
 
@@
 
-```{r}
-# make figure
-# fig_smooth <-
-#   ggplot()+
-#   geom_path(data = data,
-#              aes(x, y),
-#             col = RColorBrewer::brewer.pal(4, "Set1")[4])+
-#   annotate(geom = "rect",
-#            xmin = c(650785, 650045),
-#            ymin = c(5904450, 5902700),
-#            xmax = c(652500, 650682),
-#            ymax = c(5906133, 5903080),
-#            fill = "grey90",
-#            alpha = 0.5,
-#            col = NA)+
-#     annotate(geom = "text",
-#            x = c(650785, 650045) + 400,
-#            y = c(5904450, 5902700) + 300,
-#            col = "black",
-#            label = c("d", "e"))+
-#   theme_void()+
-#   coord_sf(crs = 32631)+
-#   labs(colour = "track")
-
+```{r echo=FALSE}
 # make zoomed in figures
@@
 
-```{r}
+```{r echo=FALSE}
 # make combined walkthrough figure
@@
 ```
+
+![](figures/fig_walkthrough.png)
 
@@
 
-```{r}
+```{r echo=FALSE}
 fig_calib_residence <-
@@
 
-### Plot the outcome
+We prepare a plot of the residence patches.
 
-```{r}
+```{r echo=FALSE}
 # patch with residence points and all patches
@@
 
-```{r}
+```{r echo=FALSE}
 # make figure comparing different methods
@@
 
-```{r}
+```{r echo=FALSE}
 # wrap together
@@
 ```
+
+![](figures/fig_calib_residence_patch.png)

### 2020-10-29 06_bat_data.Rmd

@@
 ---
-output: html_document
 editor_options:
   chunk_output_type: console
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 
-```{r}
+```{r echo=FALSE}
 # plot bat data to examine for errors
@@
 
-```{r}
+```{r echo=FALSE}
 # save to figures
@@
 
-```{r}
+```{r echo=FALSE}
 # now plot over raw data
@@
 
-```{r}
+```{r echo=FALSE}
 # plot the raw, filtered, and smoothed data
@@
 
-```{r}
+```{r echo=FALSE}
 fig_bat_res_timeseries <-
@@
 
+This figure made in QGIS.
+
 ![](figures/fig_bat_patches.png)

### 2020-11-11 02_filtering_data.Rmd
take care to not eval

@@
 
-```{r}
+```{r eval=FALSE}
 fwrite(data_copy, file = "data/data_errors.csv")
@@
 sapply(data_copy[, c("in_speed", "angle")], function(z) {
-  quantile(z, probs = c(0.9, 0.95), na.rm = TRUE)
+  quantile(z, probs = c(0.1, 0.9, 0.95), na.rm = TRUE)
 })
@@
 data_filtered <- atl_filter_covariates(data_copy,
-    filters = c("(in_speed < 0.024 & out_speed < 0.024) | angle < 40"))
+    filters = c("(in_speed < 0.025 & out_speed < 0.025) | angle < 35"))
 ```

### 2020-11-11 01_background.Rmd
add transit segments

@@
 library(data.table)
+library(scales)
 
@@
 
+Here we construct a movement track with two clear modes, one representing putative thermalling behaviour (animal moves in circles over a small area), and one representing transit between thermals (animal moves in a relatively straight line between areas).
+
+### Simulate thermalling
+
 ```{r}
-# do smoove data using a RACVM
-data <- smoove::simulateRACVM(dt = 0.1, Tmax = 500, omega = 5, v0 = 1, mu = 0.1)
-data <- as.data.table(data$XY)
+# do smoove data using a RACVM for three different rotational speed
+data_thermals <- Map(function(rot_speed, patch) {
+ dt <- smoove::simulateRACVM(dt = 0.1, Tmax = 50, omega = rot_speed, v0 = 0.1) 
+ dt <- as.data.table(dt$XY)
+ dt[, patch := patch]
+}, c(7, 7, 7), sprintf('patch_%i', seq(3)))
 
-# assign id
-data[, id := "test"]
-data[, time := seq_len(nrow(data))]
-# rename x and y
-setnames(data, old = c("x", "y"), new = c("y", "x"))
+# rescale x and y
+data_thermals <- Map(function(df, lim1, lim2, t1, t2) {
+  df[, x := x + lim1]
+  df[, y := y + lim2]
+  df[, time := seq(t1, t2)]
+}, data_thermals, c(1, 5, 9), c(0, 2.5, 5),
+   c(100, 800, 1500), c(600, 1300, 2000))
+
+# get thermals data
+data_thermals <- rbindlist(data_thermals)
+```
+
+### Simulate movement between patches
+
+```{r}
+# get limits
+starts <- data_thermals[, lapply(.SD, first), by = "patch"]
+ends <- data_thermals[, lapply(.SD, last), by = "patch"]
+
+# bind
+gaps <- rbind(starts, ends)
+setorder(gaps, "time")
+
+# get transition phases
+# THIS IS HARDCODED BECAUSE THERE DOESN'T APPEAR TO BE A BETTER WAY
+# FOR THE SAME TIME
+transit_1 <- gaps[c(2, 3), ]
+transit_2 <- gaps[c(4, 5), ]
+
+# simulate movement
+transits <- Map(function(df){
+  time_lim <- as.vector(df$time)
+
+  track <- as.data.table(approx(df$x, df$y, n = 10))
+
+  # add some error
+  track[, `:=`(y = do_add_error(y, 0.01))]
+
+  track[, `:=`(time = seq(time_lim[1], time_lim[2], length.out = 10),
+               patch = "transit")]
+
+  return(track)
+
+}, list(transit_1, transit_2))
+
+# bind
+transits <- rbindlist(transits)
+```
+
+```{r}
+# bind transits to thermals
+data_move <- merge(data_thermals, transits, on = "time", all = T)
+setorder(data_move, time)
 
 # save data
-fwrite(data, "data/data_for_res_patch.csv")
+fwrite(data_move, "data/data_for_res_patch.csv")
 ```

### 2020-11-11 05_calibration_data.Rmd
add box around inset region

@@
              show.legend = F)+
+  annotate(geom = "rect",
+           xmin = 650785, xmax = 653250,
+           ymin = 5904450, ymax = 5906133,
+           fill = NA, col = "black")+
   scale_color_manual(values = c("grey", pal[3]))+

### 2020-11-11 05_calibration_data.Rmd
make model plot square

@@
              size = 10, shape = 7,
-             # stroke = ,
              col = "grey20",
              alpha = 1)+
-    # geom_path(data = data_recurse,
-    #          aes(x, y),
-    #         col = "grey", lwd = 0.2)+
-  # geom_text(data = data_res_summary,
-  #           aes(x_median, y_median,
-  #               label = tID),
-  #           # size = 10, shape = 7,
-  #           # stroke = ,
-  #           nudge_x = 100,
-  #           fontface = "bold",
-  #           col = "grey20",
-  #           alpha = 1)+
-  # geom_text(data = patch_summary_data, 
-  #           aes(x_median, y_median,
-  #               label = patch),
-  #           # size = 10, shape = 7,
-  #           # stroke = ,
-  #           nudge_x = -50,
-  #           fontface = "bold",
-  #           col = pal[4],
-  #           alpha = 1)+
   geom_point(data = data_recurse,
@@
           alpha = 0.3)+
-  # geom_segment(data = patch_summary_data,
-  #           aes(x = x_end, y = y_end,
-  #               xend = shift(x_start, type = "lag"),
-  #               yend = shift(y_start, type = "lag")),
-  #           col = pal[4], lty = 1,
-  #           lwd = 0.5)+
-  # geom_point(data = patch_summary_data,
-  #            aes(x_median, y_median),
-  #            size = 5,
-  #            stroke = 1,
-  #            col = pal[4],
-  #            shape = 1,
-  #            show.legend = F)+
   scale_colour_manual(values = c("grey40", pal[3]))+
@@
 patch_summary_real[, duration_real := unlist(wp_data)]
+
 # round median coordinate for inferred patches
@@
              show.legend = F,
-            size = 3,
-            stroke = 2,
-            shape = 1,
-            col = pal[2])+
+            size = 2, col = "grey")+
   geom_smooth(data = patch_summary_compare,
@@
              se = F,
-             col = pal[1],
+             col = "grey20",
              lwd = 1,
@@
   scale_x_continuous(
-    breaks = c(400, 600, 800, 1000),
-    labels = as.integer(c(400, 600, 800, 1000) / 60)
+    breaks = 60 * seq(3, 18, 2),
+    labels = seq(3, 18, 2)
     )+
   scale_y_continuous(
-    breaks = c(400, 600, 800, 1000),
-    labels = as.integer(c(400, 600, 800, 1000) / 60)
+    breaks = 60 * seq(5, 20, 2),
+    labels = seq(5, 20, 2)
   )+
@@
   ggthemes::theme_few()+
-  theme(panel.background = element_rect(fill = "white"),
-        panel.grid.minor = element_line(colour = "grey"))+
   labs(x = "inferred duration (min)",
@@
   wrap_plots(list(fig_basic_residence, figure_patch_duration),
-           ncol = 1,
-           design = "AA\nAA\nAA\nAA\nBB")+
+             design = "AA\nAA\nB#")+
   plot_annotation(tag_levels = "a",
@@
        filename = "figures/fig_calib_residence_patch.png",
-       height = 225, width = 170, units = "mm")
+       height = 225, width = 150, units = "mm")
 ```

### 2020-11-12 05_calibration_data.Rmd
combine smooth and thin

@@
 # now apply the smooth
-data <- atl_median_smooth(data = data,
+atl_median_smooth(data = data,
                   x = "x", y = "y", time = "time",
@@
 
+### Plot smoothed data
+
+```{r}
+# make zoomed in figures
+fig_smooth <-
+  ggplot()+
+  geom_point(data = data_raw[!data_unproc, on = c("x", "y")],
+             aes(x, y),
+             col = "grey",
+             shape = 4,
+             size = 0.4)+
+  # smoothed data
+  geom_path(data = data_unproc,
+             aes(x, y),
+             col = pal[3]) +
+  geom_point(data = data_unproc,
+             aes(x, y),
+             col = pal[3],
+             shape = 19,
+             alpha = 0.5) +
+  coord_cartesian(
+    xlim = c(650785, 653250),
+    ylim = c(5904450, 5906133), 
+    expand = F
+  )+
+  ggthemes::theme_few()+
+  theme(axis.title = element_blank(),
+        axis.text = element_blank())+
+  ggspatial::annotation_scale(location = "br")
+
+ggsave(fig_smooth,
+       filename = "figures/fig_calib_median_smooth.png",
+       width = 90 / 25, height = 90 / 25)
+```
+
+## Thinning the data
+
+```{r}
+# save a copy
+data_unproc <- copy(data)
+
+# remove columns we don't need
+data <- data[, setdiff(colnames(data), 
+                       c("tID", "Timestamp", "id", "TIME", "UTCtime")), 
+             with = FALSE]
+
+# thin to a 30s interval
+data_thin <- atl_thin_data(data = data,
+                           interval = 30,
+                           method = "aggregate",
+                           id_columns = "TAG")
+```
+
+### Plot thinned data
+
 ```{r echo=FALSE}
 # make zoomed in figures
-fig_inset <- mapply(function(xl, yl) {
+fig_smooth_thin <-
   ggplot()+
+  geom_errorbar(data = data_thin,
+                aes(x, y,
+                    ymin = y - VARY,
+                    ymax = y + VARY),
+                col = pal[4],
+                alpha = 0.2,
+                lwd = 0.1)+
+  geom_errorbarh(data = data_thin,
+                 aes(x, y,
+                     xmin = x - VARX,
+                     xmax = x + VARX),
+                 col = pal[4],
+                 alpha = 0.2,
+                 lwd = 0.1)+
   geom_point(data = data_raw[!data_unproc, on = c("x", "y")],
@@
              shape = 19,
-               alpha = 0.5)+
-    geom_path(data = data,
+             alpha = 0.5) +
+  geom_point(data = data_thin,
              aes(x, y),
-              col = RColorBrewer::brewer.pal(4, "Set1")[4])+
+             col = pal[4],
+             shape = 1)+
+  geom_path(data = data_thin,
+            aes(x, y),
+            col = pal[4], size = 0.2)+
+  coord_cartesian(
+    xlim = c(650785, 653250),
+    ylim = c(5904450, 5906133), 
+    expand = F
+  )+
   ggthemes::theme_few()+
@@
         axis.text = element_blank())+
-    ggspatial::annotation_scale(location = "br")+
-    coord_cartesian(
-             xlim = xl,
-             ylim = yl, expand = F
-             )+
-    labs(colour = "track")
-}, list(xl1 = c(650785, 653250),
-        xl2 = c(650045, 650682)),
-   list(yl1 = c(5904450, 5906133),
-        yl2 = c(5902700, 5903080)),
-SIMPLIFY = FALSE)
+  ggspatial::annotation_scale(location = "br")
 
-# # add global figure to second inset
-# fig_inset[[2]] <- fig_inset[[2]] +
-#   annotation_custom(grob = ggplotGrob(fig_smooth),
-#                     xmin = 649990,
-#                     xmax = 650045 + (650682 - 650045)/2,
-#                     ymin = 5903080 - (5903080 - 5902700)/2.5,
-#                     ymax = 5903080
-#                     )
-
-fig_median_smooth <- fig_inset[[1]]
-  # wrap_plots(fig_inset)+
-  # plot_annotation(tag_levels = "a",
-  #                 tag_prefix = "(",
-  #                 tag_suffix = ")")
-# save figure
-ggsave(fig_median_smooth,
-       filename = "figures/fig_calib_median_smooth.png",
-       width = 90 / 25, height = 90 / 25)
+ggsave(fig_smooth_thin,
+       filename = "figures/fig_calib_smooth_thin.png",
+       width = 90 / 25, height = 90 / 25, dpi = 300)
 ```
@@
   list(
-    fig_data_bbox, fig_speed_outliers, fig_median_smooth),
-  design = "AB\nAC")+
+    fig_speed_outliers, fig_smooth_thin),
+  design = "A\nB")+
   plot_annotation(tag_levels = "a",
@@
        filename = "figures/fig_walkthrough.png",
-       height = 130, width = 170, units = "mm")
+       height = 170, width = 90, units = "mm")
 ```
@@
 
-## Making residence patches
+### Calculate residence time
+
+Subset for boating segments: 2020-08-24 11:13:00 -- 2020-08-24 12:36:11.
+
+```{r}
+# load recurse
+library(recurse)
+data_recurse <- data[, list(x, y, time, TAG)]
+
+# thin the data
+data_recurse <- atl_thin_data(data = data_recurse,
+                              interval = 30, id_columns = "TAG",
+                              method = "resample")
+```
+
+## Residence patches
 
@@
 
-### Calculate residence time
-
-Subset for boating segments: 2020-08-24 11:13:00 -- 2020-08-24 12:36:11.
-
 ```{r}
-# load recurse
-library(recurse)
-data_recurse <- data_unproc[, list(x, y, time, TAG, UTCtime)]
-
-# thin the data
-data_recurse <- atl_thin_data(data = data_recurse,
-                              interval = 30, id_columns = "TAG",
-                              method = "resample")
-
 # get 4 column data

### 2020-11-12 05_calibration_data.Rmd
minor changes to figure 8

@@
 library(recurse)
-data_recurse <- data[, list(x, y, time, TAG)]
-
-# thin the data
-data_recurse <- atl_thin_data(data = data_recurse,
-                              interval = 30, id_columns = "TAG",
-                              method = "resample")
 ```
@@
 ```{r}
-# make a data copy
-data_unproc <- copy(data)
-```
-
-```{r}
 # get 4 column data
-data_recurse <- data_recurse[, list(x, y, time, TAG)]
+data_recurse <- data_thin[, list(x, y, time, TAG)]
 
@@
 
-### Plot residence time
-
-```{r echo=FALSE}
-fig_calib_residence <-
-  ggplot()+
-  geom_point(data = data_res, 
-             aes(x, y),
-             size = 5, shape = 2,
-             alpha = 0.1)+
-  geom_point(data = data_recurse,
-             aes(x, y, 
-                 col = res_time > 5,
-                 size = res_time > 5),
-             shape = 19, alpha = 0.4,
-             show.legend = F)+
-  geom_path(data = data_recurse,
-             aes(x, y),
-            col = "grey", lwd = 0.2)+
-  scale_colour_manual(values = c("grey40", pal[3]))+
-  scale_size_manual(values = c(0.5, 2))+
-  ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  coord_sf(crs = 32631)
-```
-
 ### Run residence patch method
@@
 # for the known and unkniwn patches
-patch_sf_data <- atl_patch_summary(patch_res_known, which_data = "spatial",
+patch_sf_data <- atl_patch_summary(patch_res_known, 
+                                   which_data = "spatial",
                                    buffer_radius = 20)
@@
 # get summary data
-patch_summary_data <- atl_patch_summary(patch_res_known, which_data = "summary")
+patch_summary_data <- atl_patch_summary(patch_res_known, 
+                                        which_data = "summary")
 ```
@@
           alpha = 0.3)+
-  scale_colour_manual(values = c("grey40", pal[3]))+
-  scale_shape_manual(values = c(1, 19))+
-  scale_fill_distiller(palette = "RdPu")+
+  scale_colour_manual(values = c("grey50", pal[3]))+
+  scale_shape_manual(values = c(4, 19))+
   ggspatial::annotation_scale(location = "tl")+
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank())+
+        axis.title = element_blank(),
+        panel.background = element_rect(fill = "aliceblue"))+
   coord_sf(crs = 32631,
@@
              se = F,
-             col = "grey20",
-             lwd = 1,
+             col = pal[4],
              method = "lm",

### 2020-11-12 05_calibration_data.Rmd
update fig 8

@@
   ggplot()+
-  geom_sf(data = griend)+
+  geom_sf(data = griend, fill = "antiquewhite")+
   geom_point(data = data_res_summary, 

### 2020-11-12 05_calibration_data.Rmd
update fig 8

@@
              se = F,
-             col = pal[4],
+             col = pal[1],
              method = "lm",

### 2020-11-16 render_books.sh
add code to render source code

@@ -1,3 +1,6 @@
 #!/bin/bash
 Rscript --slave -e 'bookdown::render_book("index.rmd")'
 Rscript --slave -e 'bookdown::render_book("index.rmd", "rmarkdown::pdf_document"); warnings()'
+
+# to render source files
+Rscript --slave -a 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'

### 2020-11-16 R/01_background.R
add source code as R scripts

@@ -1,0 +1,87 @@
+## ----install_package-------------------------------------------------------------------------------------------------------------------------------
+# use either devtools or remotes to install
+install.packages("devtools")
+
+# installation using devtools
+devtools::install_github("pratikunterwegs/atlastools")
+
+
+## ----prep_libs_01_02-------------------------------------------------------------------------------------------------------------------------------
+# load smoove and datatable
+library(smoove)
+library(data.table)
+library(scales)
+
+# source helper functions
+source("R/helper_functions.R")
+
+
+## ----save_data-------------------------------------------------------------------------------------------------------------------------------------
+data <- do_smoove_data()
+# save simulated data
+fwrite(data, "data/data_sim.csv")
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# do smoove data using a RACVM for three different rotational speed
+data_thermals <- Map(function(rot_speed, patch) {
+ dt <- smoove::simulateRACVM(dt = 0.1, Tmax = 50, omega = rot_speed, v0 = 0.1) 
+ dt <- as.data.table(dt$XY)
+ dt[, patch := patch]
+}, c(7, 7, 7), sprintf('patch_%i', seq(3)))
+
+# rescale x and y
+data_thermals <- Map(function(df, lim1, lim2, t1, t2) {
+  df[, x := x + lim1]
+  df[, y := y + lim2]
+  df[, time := seq(t1, t2)]
+}, data_thermals, c(1, 5, 9), c(0, 2.5, 5),
+   c(100, 800, 1500), c(600, 1300, 2000))
+
+# get thermals data
+data_thermals <- rbindlist(data_thermals)
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# get limits
+starts <- data_thermals[, lapply(.SD, first), by = "patch"]
+ends <- data_thermals[, lapply(.SD, last), by = "patch"]
+
+# bind
+gaps <- rbind(starts, ends)
+setorder(gaps, "time")
+
+# get transition phases
+# THIS IS HARDCODED BECAUSE THERE DOESN'T APPEAR TO BE A BETTER WAY
+# FOR THE SAME TIME
+transit_1 <- gaps[c(2, 3), ]
+transit_2 <- gaps[c(4, 5), ]
+
+# simulate movement
+transits <- Map(function(df){
+  time_lim <- as.vector(df$time)
+  
+  track <- as.data.table(approx(df$x, df$y, n = 10))
+  
[…]

### 2020-11-16 R/02_filtering_data.R

@@ -1,0 +1,226 @@
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# to handle movement data
+library(data.table)
+library(atlastools)
+
+# to plot
+library(ggplot2)
+library(patchwork)
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# source helper functions
+source("R/helper_functions.R")
+
+
+## ----read_sim_data---------------------------------------------------------------------------------------------------------------------------------
+# read in the data
+data <- fread("data/data_sim.csv")[5000:10000, ]
+
+
+## ----add_outlier-----------------------------------------------------------------------------------------------------------------------------------
+# make a copy
+data_copy <- copy(data)
+
+# add a prolonged spike or reflection to 300 positions
+data_copy[500:800, `:=`(x = x + 0.25,
+                        y = y + 0.25)]
+
+# add normal error
+data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
+                 y = do_add_error(y, std_dev = 0.005))]
+
+# add 100 outliers
+data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
+
+
+## ----eval=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## fwrite(data_copy, file = "data/data_errors.csv")
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+data_copy <- fread("data/data_errors.csv")
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# define a four colour palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+
+
+## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+# make figure of canonical data with added errors
+figure_raw <-
+  ggplot()+
+  geom_point(data = data_copy,
+            aes(x, y),
+            col = "grey",
+            alpha = 1,
+            size = 0.2)+
+  geom_path(data = data,
+            aes(x, y),
+            col = "grey20",
+            alpha = 1)+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  coord_equal()+
+  labs(colour = NULL)
+
+
[…]

### 2020-11-16 R/03_smoothing_tracks.R

@@ -1,0 +1,224 @@
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# prep libs
+library(data.table)
+library(atlastools)
+library(ggplot2)
+library(patchwork)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+
+
+## ----read_sim_data_2-------------------------------------------------------------------------------------------------------------------------------
+# read in the data and set the window size variable
+data <- fread("data/data_sim.csv")[5000:10000, ]
+data[, window_size := NA]
+
+# data with small scale errors but no reflections or outliers
+data_errors <- fread("data/data_no_reflection.csv")
+data_errors[, window_size := 0]
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# smooth the data over four K values
+list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
+  
+  data_copy <- copy(data_errors)
+  
+  data_copy <- atl_median_smooth(data = data_copy,
+                    x = "x", 
+                    y = "y",
+                    time = "time",
+                    moving_window = K)
+  
+  data_copy[, window_size := K]
+})
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+fwrite(list_of_smooths[[3]], file = "data/data_smooth.csv")
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# bind list after offset
+data_plot <- mapply(function(df, offset) {
+  df[, x := x + offset]
+}, list_of_smooths, seq(0.4, 1.25, length.out = 4),
+SIMPLIFY = F)
+
+data_plot <- rbindlist(data_plot)
+
+
+## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+# prepare data to plot
+# make list of data to plot
+figure_median_smooth <-
+  ggplot()+
+  geom_point(data = data_errors,
+             aes(x, y),
+             col = pal[3],
+             size = 0.2, alpha = 0.5)+
+  geom_path(data = data,
+            aes(x, y),
+            col = "grey20",
+            lwd = 0.5)+          
+  geom_path(data = data_plot,
+            aes(x, y,
+                col = window_size,
+                group = window_size),
+            show.legend = F,
[…]

### 2020-11-16 R/04_residence_patch.R

@@ -1,0 +1,157 @@
+## ----prep_libs_02_01-------------------------------------------------------------------------------------------------------------------------------
+library(data.table)
+library(atlastools)
+library(ggplot2)
+library(patchwork)
+
+# for residence time
+library(recurse)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# read patch data
+data <- fread("data/data_for_res_patch.csv")
+
+# an id is required
+data[, id := "test"]
+
+# flip lengthwise
+setnames(data, old = c("x", "y"), new = c("y", "x"))
+
+# do recurse
+data_recurse <- getRecursions(data[, list(x, y, time, id)],
+                              radius = 1)
+
+# assign residence time
+data[, residence_time := data_recurse$residenceTime]
+
+
+## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+# restime by position
+fig_res_a <-
+  ggplot()+
+  # geom_path(data = data,
+  #           aes(x, y),
+  #           lwd = 0.2,
+  #           col = "grey20")+
+  geom_path(data = data,
+             aes(x, y,
+                 group = NA,
+                 col = residence_time > 0.04),
+             show.legend = F,
+             alpha = 1)+
+  # geom_point(data = data,
+  #            aes(x, y,
+  #                col = residence_time),
+  #            size = 0.7,
+  #            show.legend = F,
+  #            alpha = 1)+
+  coord_equal()+
+  scale_colour_manual(values = c("grey", pal[3]))+
+  # scale_shape_manual(values = c(, 1))+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# make residence patch
+patch <- atl_res_patch(data[residence_time > 0.05, ], 
+                       buffer_radius = 0.1, 
+                       lim_spat_indep = 1, 
+                       lim_time_indep = 5)
+
+
+
+# get spatial representation
+patch_sf <- atl_patch_summary(patch_data = patch,
+                              which_data = "spatial", 
+                              buffer_radius = 0.15)
+
+# get summary data
[…]

### 2020-11-16 R/05_calibration_data.R

@@ -1,0 +1,488 @@
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# load libs
+library(data.table)
+library(atlastools)
+library(ggplot2)
+library(patchwork)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# read and plot example data
+data <- fread("data/atlas1060_allTrials_annotated.csv")
+data_raw <- copy(data)
+
+
+## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+# plot data
+fig_data_raw <-
+  ggplot(data)+
+  geom_path(aes(x, y),
+             col = "grey", alpha = 0.5, size = 0.3)+
+  geom_point(aes(x, y),
+             col = pal[3], alpha = 0.2, size = 2)+
+  ggthemes::theme_map()+
+  coord_sf(crs = 32631)
+
+# save figure
+ggsave(fig_data_raw, filename = "figures/fig_calibration_raw.png",
+       width = 185 / 25)
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+data_unproc <- copy(data)
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# remove inside must be set to falses
+data <- atl_filter_bounds(data = data, 
+                          x = "x", y = "y", 
+                          x_range = c(645000, max(data$x)), 
+                          remove_inside = FALSE)
+
+
+## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+# plot data
+fig_data_bbox <-
+  ggplot()+
+  geom_path(data = data_raw,
+            aes(x, y),
+             col = "grey", alpha = 0.5, size = 0.3)+
+  geom_point(data = data_raw,
+             aes(x, y,
+                 col = x > 645000),
+             alpha = ifelse(data_raw$x > 645000, 0.5, 1),
+             size = ifelse(data_raw$x > 645000, 0.3, 2),
+             show.legend = F)+
+  scale_colour_manual(values = c("grey", pal[3]))+
+  geom_vline(xintercept = 645000,
+             col = "grey", lty = 2)+
+  ggspatial::annotation_scale(location = "br")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  coord_sf(crs = 32631)
+
+# save result
[…]

### 2020-11-16 R/06_bat_data.R

@@ -1,0 +1,407 @@
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# load libs
+library(data.table)
+library(RSQLite)
+library(atlastools)
+library(ggplot2)
+library(patchwork)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+
+
+## ----read_bat_data---------------------------------------------------------------------------------------------------------------------------------
+# prepare the connection
+con <- dbConnect(drv = SQLite(), dbname = "data/Three_example_bats.sql")
+
+# list the tables
+table_name <- dbListTables(con)
+
+# prepare to query all tables
+query <- sprintf('select * from \"%s\"', table_name)
+
+# query the database
+data <- dbGetQuery(conn = con, statement = query)
+
+# disconnect from database
+dbDisconnect(con)
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# convert data to datatable
+setDT(data)
+
+# write data for QGIS
+fwrite(data, file = "data/bat_data.csv")
+
+
+## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+# plot bat data to examine for errors
+fig_bat_raw <-
+  ggplot(data)+
+  geom_path(aes(X, Y,
+                group = TAG,
+                colour = TAG),
+            show.legend = FALSE)+
+  scale_colour_brewer(palette = "Dark2")+
+  ggspatial::annotation_scale(location = "br")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  coord_sf(crs = 2039)
+
+
+## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+# save to figures
+ggsave(fig_bat_raw,
+       filename = "figures/fig_bat_raw.png",
+       dpi = 300,
+       width = 6, height = 4)
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
+# split bat data by tag
+# first make a copy using the data.table function copy
+# this prevents the orignal data from being modified by atlastools
+# functions which DO MODIFY BY REFERENCE!
+data_split <- copy(data)
+
+# now split
+data_split <- split(data_split, data$TAG)
+
+
+## --------------------------------------------------------------------------------------------------------------------------------------------------
[…]

### 2020-11-16 R/06_bat_data.R
rename example code to be ignored by bookdown

@@ -1,407 +1,0 @@
-## --------------------------------------------------------------------------------------------------------------------------------------------------
-# load libs
-library(data.table)
-library(RSQLite)
-library(atlastools)
-library(ggplot2)
-library(patchwork)
-
-# prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
-
-
-## ----read_bat_data---------------------------------------------------------------------------------------------------------------------------------
-# prepare the connection
-con <- dbConnect(drv = SQLite(), dbname = "data/Three_example_bats.sql")
-
-# list the tables
-table_name <- dbListTables(con)
-
-# prepare to query all tables
-query <- sprintf('select * from \"%s\"', table_name)
-
-# query the database
-data <- dbGetQuery(conn = con, statement = query)
-
-# disconnect from database
-dbDisconnect(con)
-
-
-## --------------------------------------------------------------------------------------------------------------------------------------------------
-# convert data to datatable
-setDT(data)
-
-# write data for QGIS
-fwrite(data, file = "data/bat_data.csv")
-
-
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
-# plot bat data to examine for errors
-fig_bat_raw <-
-  ggplot(data)+
-  geom_path(aes(X, Y,
-                group = TAG,
-                colour = TAG),
-            show.legend = FALSE)+
-  scale_colour_brewer(palette = "Dark2")+
-  ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  coord_sf(crs = 2039)
-
-
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
-# save to figures
-ggsave(fig_bat_raw,
-       filename = "figures/fig_bat_raw.png",
-       dpi = 300,
-       width = 6, height = 4)
-
-
-## --------------------------------------------------------------------------------------------------------------------------------------------------
-# split bat data by tag
-# first make a copy using the data.table function copy
-# this prevents the orignal data from being modified by atlastools
-# functions which DO MODIFY BY REFERENCE!
-data_split <- copy(data)
-
-# now split
-data_split <- split(data_split, data$TAG)
-
-
-## --------------------------------------------------------------------------------------------------------------------------------------------------
[…]

### 2020-11-16 R/_01_background.R

@@ -1,4 +1,4 @@
-## ----install_package-------------------------------------------------------------------------------------------------------------------------------
+## ----install_package--------------------------------------------------
 # use either devtools or remotes to install
 install.packages("devtools")
 
@@ -6,7 +6,7 @@ install.packages("devtools")
 devtools::install_github("pratikunterwegs/atlastools")
 
 
-## ----prep_libs_01_02-------------------------------------------------------------------------------------------------------------------------------
+## ----prep_libs_01_02--------------------------------------------------
 # load smoove and datatable
 library(smoove)
 library(data.table)
@@ -16,13 +16,13 @@ library(scales)
 source("R/helper_functions.R")
 
 
-## ----save_data-------------------------------------------------------------------------------------------------------------------------------------
+## ----save_data--------------------------------------------------------
 data <- do_smoove_data()
 # save simulated data
 fwrite(data, "data/data_sim.csv")
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # do smoove data using a RACVM for three different rotational speed
 data_thermals <- Map(function(rot_speed, patch) {
  dt <- smoove::simulateRACVM(dt = 0.1, Tmax = 50, omega = rot_speed, v0 = 0.1) 
@@ -42,7 +42,7 @@ data_thermals <- Map(function(df, lim1, lim2, t1, t2) {
 data_thermals <- rbindlist(data_thermals)
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # get limits
 starts <- data_thermals[, lapply(.SD, first), by = "patch"]
 ends <- data_thermals[, lapply(.SD, last), by = "patch"]
@@ -77,7 +77,7 @@ transits <- Map(function(df){
 transits <- rbindlist(transits)
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # bind transits to thermals
 data_move <- merge(data_thermals, transits, on = "time", all = T)
 setorder(data_move, time)

### 2020-11-16 R/_02_filtering_data.R

@@ -1,4 +1,4 @@
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # to handle movement data
 library(data.table)
 library(atlastools)
@@ -8,17 +8,17 @@ library(ggplot2)
 library(patchwork)
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # source helper functions
 source("R/helper_functions.R")
 
 
-## ----read_sim_data---------------------------------------------------------------------------------------------------------------------------------
+## ----read_sim_data----------------------------------------------------
 # read in the data
 data <- fread("data/data_sim.csv")[5000:10000, ]
 
 
-## ----add_outlier-----------------------------------------------------------------------------------------------------------------------------------
+## ----add_outlier------------------------------------------------------
 # make a copy
 data_copy <- copy(data)
 
@@ -34,20 +34,20 @@ data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
 data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
 
 
-## ----eval=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## ----eval=FALSE-------------------------------------------------------
 ## fwrite(data_copy, file = "data/data_errors.csv")
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 data_copy <- fread("data/data_errors.csv")
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # define a four colour palette
 pal <- RColorBrewer::brewer.pal(4, "Set1")
 
 
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## ----echo=FALSE-------------------------------------------------------
 # make figure of canonical data with added errors
 figure_raw <-
   ggplot()+
@@ -67,7 +67,7 @@ figure_raw <-
   labs(colour = NULL)
 
 
-## ----remove_outside_bbox---------------------------------------------------------------------------------------------------------------------------
+## ----remove_outside_bbox----------------------------------------------
 # remove positions outside a bounding box
 # NB: set remove_inside to FALSE
[…]

### 2020-11-16 R/_03_smoothing_tracks.R

@@ -1,4 +1,4 @@
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # prep libs
 library(data.table)
 library(atlastools)
@@ -9,7 +9,7 @@ library(patchwork)
 pal <- RColorBrewer::brewer.pal(4, "Set1")
 
 
-## ----read_sim_data_2-------------------------------------------------------------------------------------------------------------------------------
+## ----read_sim_data_2--------------------------------------------------
 # read in the data and set the window size variable
 data <- fread("data/data_sim.csv")[5000:10000, ]
 data[, window_size := NA]
@@ -19,7 +19,7 @@ data_errors <- fread("data/data_no_reflection.csv")
 data_errors[, window_size := 0]
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # smooth the data over four K values
 list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
   
@@ -35,11 +35,11 @@ list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
 })
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 fwrite(list_of_smooths[[3]], file = "data/data_smooth.csv")
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # bind list after offset
 data_plot <- mapply(function(df, offset) {
   df[, x := x + offset]
@@ -49,7 +49,7 @@ SIMPLIFY = F)
 data_plot <- rbindlist(data_plot)
 
 
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## ----echo=FALSE-------------------------------------------------------
 # prepare data to plot
 # make list of data to plot
 figure_median_smooth <-
@@ -88,7 +88,7 @@ ggsave(figure_median_smooth, filename = "figures/fig_median_smooth.png",
        width = 170, height = 170 / 3, units = "mm")
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # choose the 11 point median smooth data
 data_agg <- copy(list_of_smooths[[3]])
 
@@ -118,7 +118,7 @@ speed_agg_smooth <-
 speed_agg_smooth <- rbindlist(speed_agg_smooth)
 
 
[…]

### 2020-11-16 R/_04_residence_patch.R

@@ -1,4 +1,4 @@
-## ----prep_libs_02_01-------------------------------------------------------------------------------------------------------------------------------
+## ----prep_libs_02_01--------------------------------------------------
 library(data.table)
 library(atlastools)
 library(ggplot2)
@@ -11,7 +11,7 @@ library(recurse)
 pal <- RColorBrewer::brewer.pal(4, "Set1")
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # read patch data
 data <- fread("data/data_for_res_patch.csv")
 
@@ -29,7 +29,7 @@ data_recurse <- getRecursions(data[, list(x, y, time, id)],
 data[, residence_time := data_recurse$residenceTime]
 
 
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## ----echo=FALSE-------------------------------------------------------
 # restime by position
 fig_res_a <-
   ggplot()+
@@ -57,7 +57,7 @@ fig_res_a <-
         axis.title = element_blank())
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # make residence patch
 patch <- atl_res_patch(data[residence_time > 0.05, ], 
                        buffer_radius = 0.1, 
@@ -76,7 +76,7 @@ patch_summary <- atl_patch_summary(patch_data = patch,
                                    which_data = "summary")
 
 
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## ----echo=FALSE-------------------------------------------------------
 # plot_patches <-
 fig_res_b <-
   ggplot()+
@@ -141,7 +141,7 @@ fig_res_c <-
        y = "res. time")
 
 
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## ----echo=FALSE-------------------------------------------------------
 fig_residence <-
   wrap_plots(list(fig_res_a, fig_res_c, fig_res_b),
            design = "ACC\nBCC")+

### 2020-11-16 R/_05_calibration_data.R

@@ -1,4 +1,4 @@
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # load libs
 library(data.table)
 library(atlastools)
@@ -9,13 +9,13 @@ library(patchwork)
 pal <- RColorBrewer::brewer.pal(4, "Set1")
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # read and plot example data
 data <- fread("data/atlas1060_allTrials_annotated.csv")
 data_raw <- copy(data)
 
 
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## ----echo=FALSE-------------------------------------------------------
 # plot data
 fig_data_raw <-
   ggplot(data)+
@@ -31,11 +31,11 @@ ggsave(fig_data_raw, filename = "figures/fig_calibration_raw.png",
        width = 185 / 25)
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 data_unproc <- copy(data)
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # remove inside must be set to falses
 data <- atl_filter_bounds(data = data, 
                           x = "x", y = "y", 
@@ -43,7 +43,7 @@ data <- atl_filter_bounds(data = data,
                           remove_inside = FALSE)
 
 
-## ----echo=FALSE------------------------------------------------------------------------------------------------------------------------------------
+## ----echo=FALSE-------------------------------------------------------
 # plot data
 fig_data_bbox <-
   ggplot()+
@@ -70,12 +70,12 @@ ggsave(fig_data_bbox, filename = "figures/fig_calib_bbox.png",
        width = 185 / 25)
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # divide by 1000, convert to integer, then convert to POSIXct
 data[, time := as.integer(TIME / 1000)]
 
 
-## --------------------------------------------------------------------------------------------------------------------------------------------------
+## ---------------------------------------------------------------------
 # add incoming and outgoing speed
 data[, `:=` (speed_in = atl_get_speed(data, 
                                       x = "x", 
[…]

### 2020-11-16 _01_background.Rmd
rename files'

--- 2020-11-11 01_background.Rmd
@@
 
-```{r install_package, eval=FALSE}
+```{r install_package}
 # use either devtools or remotes to install
@@
 
-```{r prep_libs_01_02, eval=FALSE}
+```{r prep_libs_01_02}
 # load smoove and datatable

### 2020-11-16 docs/supplementary_material.tex
bookdown things

--- 2020-10-29 docs/Pre-processing-high-frequency-tracking-data.tex
@@
 \documentclass[
-]{scrreprt}
+]{article}
 \usepackage{lmodern}
@@
 \hypersetup{
-  pdftitle={A Rough Guide to Pre-processing High-Frequency Animal Tracking Data},
-  pdfauthor={Pratik R. Gupte and others},
+  pdftitle={Supplementary Material for: A Rough Guide to Pre-processing High-Frequency Animal Tracking Data},
+  pdfauthor={Pratik R. Gupte; Christine E. Beardsworth; Orr Spiegel; Emmanuel Lourie; Allert Bijleveld},
   hidelinks,
@@
 % use nice fonts if available else use boring defaults
-\IfFontExistsTF{IBM Plex Mono}{\setmonofont[]{IBM Plex Mono}}{}
 
 \usepackage{lineno}
-\usepackage{tgtermes}
+\setmainfont{Times New Roman}
+
+\linenumbers
 \newlength{\cslhangindent}
@@
 
-\title{A Rough Guide to Pre-processing High-Frequency Animal Tracking Data}
-\author{Pratik R. Gupte and others}
-\date{2020-10-29}
+\title{Supplementary Material for: A Rough Guide to Pre-processing High-Frequency Animal Tracking Data}
+\author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Allert Bijleveld}
+\date{2020-11-16}
 
@@
 {
-\setcounter{tocdepth}{1}
+\setcounter{tocdepth}{2}
 \tableofcontents
 }
-\hypertarget{abstract}{%
-\chapter{Abstract}\label{abstract}}
-
-Data cleaning is a ubiquitous pre-processing step in analysis pipelines, and its automation is essential for large data volumes such as those generated in animal tracking studies using high-frequency Time-of-Arrival (TOA) systems.
-Users of systems such as ATLAS must contend with two intertwined data cleaning challenges: (1) reducing positioning errors, and (2) the high volume of data itself.
-Making biological inferences from data with positioning errors is not straightforward, and processing large data volumes is computationally intensive.
-Though reducing positioning error is widely recommended, users are without uniform guidance on how to go about this, and lack a common set of computationally efficient tools.
-Further, many methods that synthesize movement tracks for ecological inference are either (1) not suited to very large datasets, or (2) not intuitive to understand in terms of the tracked animal's biology.
-
-In this article we introduce a pipeline to pre-process high-frequency animal tracking data in order to prepare it for subsequent analysis.
-We demonstrate this pipeline on simulated movement data to which we have randomly added positioning errors.
-This pipeline is suited to any tracking study in which the high data volume combined with knowledge of the tracked individuals' biology can be used to smooth out positioning errors.
-We further suggest how large volumes of cleaned data may be synthesized into biologically meaningful `residence patches', and demonstrate how this accurately captures animal space-use.
-Finally, we introduce the \texttt{R} package \texttt{atlastools} which provides fast implementations of the methods we describe.
-Though aimed at ATLAS systems, \texttt{atlastools} can be used with any time-series animal movement data, and we demonstrate its usage with both simulated and empirical examples.
-
-\includegraphics{figures/fig_recipe.png}
-
-\hypertarget{getting-started}{%
-\chapter{Getting started}\label{getting-started}}
-
-This section covers:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Installing the \texttt{R} package \texttt{atlastools},
-\item
-  Simulating some realistic looking movement data using the \texttt{R} package \texttt{smoove} from (Gurarie et al. \protect\hyperlink{ref-gurarie2017}{2017}), and
-\item
-  Introducing positioning errors into the simulated movement data.
-\end{enumerate}
-
-\hypertarget{installing-atlastools}{%
-\section{\texorpdfstring{Installing \texttt{atlastools}}{Installing atlastools}}\label{installing-atlastools}}
-
-This paper refers extensively to the \texttt{R} package \texttt{atlastools} (Gupte 2020), which can be installed from \href{https://github.com/pratikunterwegs/atlastools}{Github}.
-
-Releases of the package from Github can be found on Zenodo: (\url{https://doi.org/10.5281/zenodo.4033155})
-
-The code chunk below shows how to install \texttt{atlastools}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# use either devtools or remotes to install}
-\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
-
-\CommentTok{\# installation using devtools}
-\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{simulating-movement-data}{%
-\section{Simulating movement data}\label{simulating-movement-data}}
-
-Here, we simulate some movement data using the \texttt{smoove} \texttt{R} package from (Gurarie et al. \protect\hyperlink{ref-gurarie2017}{2017}).
-
-First, we load \texttt{smoove} and \texttt{data.table}, as well as some helper functions that make use of them to simulate data for use.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# load smoove and datatable}
-\KeywordTok{library}\NormalTok{(smoove)}
-\KeywordTok{library}\NormalTok{(data.table)}
-
-\CommentTok{\# source helper functions}
-\KeywordTok{source}\NormalTok{(}\StringTok{"R/helper\_functions.R"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{do\_smoove\_data}\NormalTok{()}
-\CommentTok{\# save simulated data}
-\KeywordTok{fwrite}\NormalTok{(data, }\StringTok{"data/data\_sim.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{simualte-for-residence-patches}{%
-\section{Simualte for residence patches}\label{simualte-for-residence-patches}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# do smoove data using a RACVM}
-\NormalTok{data <{-}}\StringTok{ }\NormalTok{smoove}\OperatorTok{::}\KeywordTok{simulateRACVM}\NormalTok{(}\DataTypeTok{dt =} \FloatTok{0.1}\NormalTok{, }\DataTypeTok{Tmax =} \DecValTok{500}\NormalTok{, }\DataTypeTok{omega =} \DecValTok{5}\NormalTok{, }\DataTypeTok{v0 =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mu =} \FloatTok{0.1}\NormalTok{)}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{as.data.table}\NormalTok{(data}\OperatorTok{$}\NormalTok{XY)}
-
-\CommentTok{\# assign id}
-\NormalTok{data[, id }\OperatorTok{:}\ErrorTok{=}\StringTok{ "test"}\NormalTok{]}
-\NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{seq\_len}\NormalTok{(}\KeywordTok{nrow}\NormalTok{(data))]}
-\CommentTok{\# rename x and y}
-\KeywordTok{setnames}\NormalTok{(data, }\DataTypeTok{old =} \KeywordTok{c}\NormalTok{(}\StringTok{"x"}\NormalTok{, }\StringTok{"y"}\NormalTok{), }\DataTypeTok{new =} \KeywordTok{c}\NormalTok{(}\StringTok{"y"}\NormalTok{, }\StringTok{"x"}\NormalTok{))}
-
-\CommentTok{\# save data}
-\KeywordTok{fwrite}\NormalTok{(data, }\StringTok{"data/data\_for\_res\_patch.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{reducing-large-scale-positioning-error-by-filtering-data}{%
-\chapter{Reducing Large-Scale Positioning Error by Filtering Data}\label{reducing-large-scale-positioning-error-by-filtering-data}}
+\hypertarget{processing-egyptian-fruit-bat-tracks}{%
+\section{Processing Egyptian fruit bat tracks}\label{processing-egyptian-fruit-bat-tracks}}
+
+We show the pre-processing pipeline at work on the tracks of three Egyptian fruit bats (\emph{Rousettus aegyptiacus}), and construct residence patches.
 
 \hypertarget{prepare-libraries}{%
-\section{Prepare Libraries}\label{prepare-libraries}}
-
-Here we load some useful libraries, and the helper functions.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# to handle movement data}
-\KeywordTok{library}\NormalTok{(data.table)}
-\KeywordTok{library}\NormalTok{(atlastools)}
-
-\CommentTok{\# to plot}
-\KeywordTok{library}\NormalTok{(ggplot2)}
-\KeywordTok{library}\NormalTok{(patchwork)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# source helper functions}
-\KeywordTok{source}\NormalTok{(}\StringTok{"R/helper\_functions.R"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{introducing-errors-to-the-data}{%
-\section{Introducing Errors to the Data}\label{introducing-errors-to-the-data}}
-
-Here we introduce three kinds of errors to the data:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Small-scale normally distributed errors at each position;
-\item
-  Large-scale error at a random 0.5\% of positions;
-\item
-  A large-scale displacement of a sequence of 300 positions.
-\end{enumerate}
-
-While the data are 10,000 positions at 1-second interval, we shall use only 5,000 of these.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# read in the data}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_sim.csv"}\NormalTok{)[}\DecValTok{5000}\OperatorTok{:}\DecValTok{10000}\NormalTok{, ]}
-\end{Highlighting}
-\end{Shaded}
-
-We add outliers at random to the data to demonstrate their removal.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# make a copy}
-\NormalTok{data\_copy <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-
-\CommentTok{\# add a prolonged spike or reflection to 300 positions}
-\NormalTok{data\_copy[}\DecValTok{500}\OperatorTok{:}\DecValTok{800}\NormalTok{, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ x }\OperatorTok{+}\StringTok{ }\FloatTok{0.25}\NormalTok{,}
-                        \DataTypeTok{y =}\NormalTok{ y }\OperatorTok{+}\StringTok{ }\FloatTok{0.25}\NormalTok{)]}
-
-\CommentTok{\# add normal error}
-\NormalTok{data\_copy[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{x =} \KeywordTok{do\_add\_error}\NormalTok{(x, }\DataTypeTok{std\_dev =} \FloatTok{0.01}\NormalTok{),}
-                 \DataTypeTok{y =} \KeywordTok{do\_add\_error}\NormalTok{(y, }\DataTypeTok{std\_dev =} \FloatTok{0.005}\NormalTok{))]}
-
-\CommentTok{\# add 100 outliers}
-\NormalTok{data\_copy <{-}}\StringTok{ }\KeywordTok{do\_add\_outliers}\NormalTok{(data\_copy, }\DataTypeTok{p\_data =} \FloatTok{0.005}\NormalTok{, }\DataTypeTok{std\_dev =} \FloatTok{0.1}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Save the data to which errors have been added.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{fwrite}\NormalTok{(data\_copy, }\DataTypeTok{file =} \StringTok{"data/data\_errors.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{data\_copy <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_errors.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Define a palette with 4 colours for convenience.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# define a four colour palette}
-\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-We make a figure of the canonical data (grey line) along with the artificially added error (grey points).
-
-\hypertarget{filtering-by-spatial-bounds}{%
-\section{Filtering by Spatial Bounds}\label{filtering-by-spatial-bounds}}
-
-Filtering by spatial bounds is a good way to begin reducing gross, large-scale positioning errors.
-There are two main ways of doing this:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Filtering by a bounding box: Compare the coordinates of observations against a range of acceptable coordinates, and retain those which fall within the range, or
-\item
-  Filtering by a spatial polygon: An explicit geometric intersection between the positions and a polygon representing an area of interest.
-\end{enumerate}
-
-Here, we remove gross positioning errors using a bounding box filter using the function \texttt{atl\_filter\_bounds}. In this example, we show filtering only on the Y coordinate.
-
-\texttt{atl\_filter\_bounds} takes coordinate ranges as two-element vectors of the lower and higher bound.
-It is possible to pass one of the bounds as \texttt{NA}, in which case, only the other bound is used for filtering, i.e., \texttt{y\_range\ =\ c(NA,\ 1)} is equivalent to selecting all Y coordinates \textless{} 1.
-
-\texttt{atl\_filter\_bounds} was initially designed to remove positions \emph{inside} a specific range, hence the argument \texttt{remove\_inside}.
-The default value of the argument is \texttt{FALSE}, and \texttt{atl\_filter\_bounds} is thus a bounding box filter.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# remove positions outside a bounding box}
-\CommentTok{\# NB: set remove\_inside to FALSE}
-\NormalTok{data\_inside\_bbox <{-}}\StringTok{ }\KeywordTok{atl\_filter\_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_copy,}
-                                      \DataTypeTok{y\_range =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{1}\NormalTok{),}
-                                      \DataTypeTok{remove\_inside =} \OtherTok{FALSE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\texttt{atl\_filter\_bounds} is not vectorised, and if there are two or more bounds per coordinate (for instance, \(X_1 \ldots X_2\), and \(X_3 \ldots X_4\)), they must be passed in two different function calls.
-The same is true for filtering by an \texttt{sf} polygon. \texttt{atl\_filter\_bounds} also supports filtering by multi-polygon objects.
-
-Having filtered the data we shall prepare it for plotting.
-
-\includegraphics{figures/fig_raw_bounds.png}
-
-\textbf{(a)} A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the \texttt{smoove} package (Gurarie et al.~2017).
-Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track.
-The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (solid black line) as closely as possible.
-\textbf{(b)} Tracks can be quickly filtered by spatial bounds (\(0.5 \leq Y \leq 1.0\); dashed grey lines) using the \texttt{atlastools} function \texttt{atl\_filter\_bounds}.
-Setting the function argument \texttt{remove\_inside\ =\ FALSE} retains positions within user supplied bounds (blue points), and excludes those outside (red points).
-
-\hypertarget{a-note-on-filtering-by-bounds}{%
-\subsection{A Note on Filtering by Bounds}\label{a-note-on-filtering-by-bounds}}
-
-The filter on spatial bounds is only for demonstration, and is not applied to the data.
-
-\hypertarget{filtering-unrealistic-movement}{%
-\section{Filtering Unrealistic Movement}\label{filtering-unrealistic-movement}}
-
-Large-scale positioning error can affect both point locations as well as entire subsets of a track, leading to a track appearing to show unrealistic movement for the study species (Bjørneraas et al.
-\protect\hyperlink{ref-bjorneraas2010}{2010}).
-
-The best way to examine whether a track contains such positioning errors is to plot the data, and especially to join the dots, i.e., to connect the positions with lines rather than simply plot points.
-
-Users with extensive experience of their study system will readily recognise segments of a track where the movement appears to be unrealistic.
-
-Briefly, the two main kinds of errors are mentioned here, and named based on their appearance in a tack: those that affect single positions lend a spiky appearance to a track, and are referred to as \emph{point outliers} or \emph{spikes}, while errors affecting continuous subsets of a track cause it to appear as though it has been reflected along a plane, and are hence called \emph{reflections}, or \emph{prolonged spikes}.
-
-\hypertarget{filtering-point-outliers-or-spikes}{%
-\subsection{Filtering Point Outliers or Spikes}\label{filtering-point-outliers-or-spikes}}
-
-We begin by removing unrealistic movement in the form of point outliers, or spikes.
-
-The first step is to determine how spikes should be identified.
-The non-movement approach prescribes determining whether movement metrics associated with each position are realistic or not, and targeting those positions where the movement metrics are unrealistic for the study species.
-
-In this case we shall calculate only two metrics, speed and turning angle.
-These are conveniently implemented in \texttt{atlastools} using the functions \texttt{atl\_get\_speed} and \texttt{atl\_turning\_angle}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# get speed and turning angle}
-\NormalTok{data\_copy[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{in\_speed =} \KeywordTok{atl\_get\_speed}\NormalTok{(data\_copy,}
-                                          \DataTypeTok{type =} \StringTok{"in"}\NormalTok{),}
-                 \DataTypeTok{out\_speed =} \KeywordTok{atl\_get\_speed}\NormalTok{(data\_copy,}
-                                           \DataTypeTok{type =} \StringTok{"out"}\NormalTok{),}
-                 \DataTypeTok{angle =} \KeywordTok{atl\_turning\_angle}\NormalTok{(data\_copy))]}
-\end{Highlighting}
-\end{Shaded}
-
-Having calculated speed and turning angle, the next step is to remove positions with extremely high incoming and outgoing speeds.
-This means using the \texttt{atl\_filter\_covariates} function to remove positions with speed \(\geq\) a plausible speed threshold.
-The use of a turning angle filter is optional, and not necessary in this case.
-
-One approach is to define a speed cutoff based on expert knowledge, and this is best suited to well studied species.
-
-For simulated data, there is no plausible speed threshold from prior knowledge --- a reasonable choice here is to use the 90\^{}th or 95\^{}th percentile of speed and turning angle (when this metric is used).
-
-This is also more general than identifying the limits of implausibility for each individual (since there may be inter-individual differences), let alone each species in a large dataset.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# get 90 and 95 percentile of speed and turning angle}
-\KeywordTok{sapply}\NormalTok{(data\_copy[, }\KeywordTok{c}\NormalTok{(}\StringTok{"in\_speed"}\NormalTok{, }\StringTok{"angle"}\NormalTok{)], }\ControlFlowTok{function}\NormalTok{(z) \{}
-  \KeywordTok{quantile}\NormalTok{(z, }\DataTypeTok{probs =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.9}\NormalTok{, }\FloatTok{0.95}\NormalTok{), }\DataTypeTok{na.rm =} \OtherTok{TRUE}\NormalTok{)}
-\NormalTok{\})}
-\end{Highlighting}
-\end{Shaded}
-
-Finally, we shall remove positions whose incoming and outgoing speeds are both greater than the 95th speed percentile using \texttt{atl\_filter\_covariates}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# filter the copy by the 95th percentile}
-\NormalTok{data\_filtered <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(data\_copy,}
-    \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(in\_speed < 0.024 \& out\_speed < 0.024) | angle < 40"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-We prepare the data for plotting.
-
-\hypertarget{filtering-reflections-or-track-subsets}{%
-\subsection{Filtering Reflections or Track Subsets}\label{filtering-reflections-or-track-subsets}}
-
-When entire track subsets are affected by positioning error, they are more difficult to remove.
-This is the case when positions are \emph{reflected} along a plane in the coordinate axis --- the simplest way in which this can be explained is to examine the way in which we simulated the reflection earlier in this text, by simply adding an offset to the X and Y coordinates of each of 300 consecutive positions (indices 500 -- 800).
-
-An explanation of why these reflections occur is found in WATLAS PAPER: Bijleveld et al.~in prep.
-In short, they are results of an error in the ATLAS localisation algorithm, and are usually removed in the localisation step's quality control procedures.
-However, some reflections may remain to confront users.
-
-Reflections and other issues affecting track subsets cannot be resolved by targeting single positions, since there are rarely any position-specific covariates that can be used to identify a position as part of a larger subset that should be removed.
-
-However, it is often possible to identify the \emph{bounds} of problematic subsets, such as reflections, and remove positions between them.
-The way this is done is conceptually similar to the point outlier algorithm.
-
-The \texttt{atl\_remove\_reflection} function implements one method remove reflected subsets of a track. The working is described:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Remove point outliers,
-\item
-  Re-calculate speed and turning angles
-\item
-  Identify the first unrealistic movement position (fast speed and high turning angle)
-\item
-  Setting this point as an anchor, identify the next position with unrealistic movement (as above)
-\item
-  Remove all positions between these two points,
-\item
-  Search for the next unrealistic movement, and repeat the process.
-\end{enumerate}
-
-This is a minimal algorithm which can be developed further, and iterated multiple times to comprehensively remove reflections.
-
-One important function argument is the estimated reflection length, i.e., how many positions are estimated to be reflected.
-This argument controls how many positions after the anchor are candidates for the reflection's end.
-
-If the reflection does not end within this number of positions, the algorithm goes awry, and considers part of the reflection to be the valid data.
-Fortunately, this can be avoided by setting the length to the number of observations in the data.
-
-However, if the reflection does not end at all, the algorithm will simply delete all positions from the anchor point onwards.
-
-Thus, the algorithm is less of a double-edged sword and more a sword with no hilt --- there are few safe ways to use it.
-
-The difficulty in implementing a reasonable algorithm for dealing with this kind of track-subset wide positioning error reveals the general problem that algorithms are \textbf{not} easy to conceptualise or efficiently implement.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# attempt to remove reflections}
-\NormalTok{data\_no\_reflection <{-}}\StringTok{ }\KeywordTok{atl\_remove\_reflections}\NormalTok{(data\_filtered,}
-                          \DataTypeTok{point\_angle\_cutoff =} \DecValTok{10}\NormalTok{,}
-                          \DataTypeTok{reflection\_speed\_cutoff =} \FloatTok{0.024}\NormalTok{)}
-\CommentTok{\# get reflections}
-\NormalTok{reflection <{-}}\StringTok{ }\NormalTok{data\_filtered[}\OperatorTok{!}\NormalTok{data\_no\_reflection,}
-\NormalTok{                                         on =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"x"}\NormalTok{, }\StringTok{"y"}\NormalTok{)]}
-\NormalTok{reflection <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(reflection)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{figures/fig_correct_tracks.png}
-
-Reducing large-scale positioning error in a movement track.
-\textbf{(a)} Positioning error may affect single observations resulting in point outliers or `spikes' (red points), but it may also affect continuous susbets of a track, which we call a \texttt{prolonged\ spike\textquotesingle{}\ (purple\ triangles).\ While\ the\ former\ may\ be\ targeted\ by\ filtering\ on\ appropriate\ covariates\ such\ as\ speed\ and\ turning\ angle\ using\ the}atlastools\texttt{function}atl\_filter\_covariates\texttt{,\ the\ latter\ cannot\ be\ effectively\ corrected\ by\ targeting\ each\ coordinate\ pair\ in\ isolation.\ **(b)**\ The}atlastools\texttt{function}atl\_remove\_reflections` to identify prolonged spikes (red triangles) in tracking data is an illustrative example of targeting positioning errors that affect track subsets.
-While this method returns the canonical track without the prolonged spike (blue points) in this example, users are cautioned to frequently check this and similar semi-supervised algorithms' results.
-
-Finally, we export the data without spikes and reflections to be used later.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{fwrite}\NormalTok{(data\_no\_reflection, }\DataTypeTok{file =} \StringTok{"data/data\_no\_reflection.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{reducing-small-scale-positioning-error-and-aggregating-data}{%
-\chapter{Reducing Small-Scale Positioning Error and Aggregating Data}\label{reducing-small-scale-positioning-error-and-aggregating-data}}
-
-Once large-scale positioning errors have been removed using filters on bounds or on unrealistic movement, small-scale positioning errors may remain. Why is this the case?
-
-The simple reason is that high-frequency tracking has a large component of error relative to the real movement of an individual, and the error / movement ratio increases with the frequency of tracking.
-(Noonan et al.
-\protect\hyperlink{ref-noonan2019}{2019}) has a better explanation, which users are strongly recommended to read.
-
-The effect of these small-scale errors for high-frequency data is to cause overestimations of straight-line displacement (distance between two positions) and speed.
-If users were to apply a very strict speed filter, many of these positions would be lost.
-This loss of information would certainly reduce the sampling frequency, which would lead to better estimates of total distance and mean speed.
-However, while \emph{total} distance and \emph{mean} speed estimates would improve, the instantaneous distance and speed estimates would remain biased, which will most likely lead to erroneous inferences from the data.
-
-However, users can turn the high-frequency of their data against itself by applying a simple smooth. The median smooth is recommended because it is robust to outliers, and is very well understood.
-
-\hypertarget{preparing-libraries-and-data}{%
-\section{Preparing Libraries and Data}\label{preparing-libraries-and-data}}
-
-We first load some useful packages and define a colour palette.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# prep libs}
-\KeywordTok{library}\NormalTok{(data.table)}
-\KeywordTok{library}\NormalTok{(atlastools)}
-\KeywordTok{library}\NormalTok{(ggplot2)}
-\KeywordTok{library}\NormalTok{(patchwork)}
-
-\CommentTok{\# prepare a palette}
-\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-We load both the simulated, canonical data (which has no errors), as well as the data with errors which has been pre-processed to remove large-scale errors.
-
-We also assign a variable, \texttt{window\_size}, which will help identify the datasets when comparing median smooth moving window (\(K\)) sizes.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# read in the data and set the window size variable}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_sim.csv"}\NormalTok{)[}\DecValTok{5000}\OperatorTok{:}\DecValTok{10000}\NormalTok{, ]}
-\NormalTok{data[, window\_size }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\OtherTok{NA}\NormalTok{]}
-
-\CommentTok{\# data with small scale errors but no reflections or outliers}
-\NormalTok{data\_errors <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_no\_reflection.csv"}\NormalTok{)}
-\NormalTok{data\_errors[, window\_size }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\DecValTok{0}\NormalTok{]}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{median-smoothing}{%
-\section{Median Smoothing}\label{median-smoothing}}
-
-We apply a median smooth with four different values of \(K\), the moving window size --- 3, 5, 11, and 21 positions.
-\(K\) must be an odd number, but apart from that there is no correct magnitude of \(K\).
-Very large \(K\) will lead to unrealistic tracks, while small \(K\) will not result in much reduction of error.
-Users are encouraged to plot their data before and after smoothing to examine the effect of different window sizes.
-
-The \texttt{atlastools} function \texttt{atl\_median\_smooth} is quite fast and users can readily try multiple \(K\) values as shown in the example below.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# smooth the data over four K values}
-\NormalTok{list\_of\_smooths <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{11}\NormalTok{, }\DecValTok{21}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(K) \{}
-
-\NormalTok{  data\_copy <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data\_errors)}
-
-\NormalTok{  data\_copy <{-}}\StringTok{ }\KeywordTok{atl\_median\_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_copy,}
-                    \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
-                    \DataTypeTok{y =} \StringTok{"y"}\NormalTok{,}
-                    \DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
-                    \DataTypeTok{moving\_window =}\NormalTok{ K)}
-
-\NormalTok{  data\_copy[, window\_size }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{K]}
-\NormalTok{\})}
-\end{Highlighting}
-\end{Shaded}
-
-We save the 11 point smoothed data.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{fwrite}\NormalTok{(list\_of\_smooths[[}\DecValTok{3}\NormalTok{]], }\DataTypeTok{file =} \StringTok{"data/data\_smooth.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-We prepare the data for plotting.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# bind list after offset}
-\NormalTok{data\_plot <{-}}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(df, offset) \{}
-\NormalTok{  df[, x }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{offset]}
-\NormalTok{\}, list\_of\_smooths, }\KeywordTok{seq}\NormalTok{(}\FloatTok{0.4}\NormalTok{, }\FloatTok{1.25}\NormalTok{, }\DataTypeTok{length.out =} \DecValTok{4}\NormalTok{),}
-\DataTypeTok{SIMPLIFY =}\NormalTok{ F)}
-
-\NormalTok{data\_plot <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_plot)}
-\end{Highlighting}
-\end{Shaded}
-
-We prepare a plot of the smoothed data.
-
-\hypertarget{thinning-data-by-aggregation}{%
-\section{Thinning Data by Aggregation}\label{thinning-data-by-aggregation}}
-
-Evenly thinning data is a good idea if statistical methods require even sampling, or if the volume of data is too large for statistical packages to efficiently handle it.
-In \texttt{R}, both may be true at once.
-
-Here, we demonstrate thinning by aggregation on data that has been median smoothed using a \(K\) of 11.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# choose the 11 point median smooth data}
-\NormalTok{data\_agg <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(list\_of\_smooths[[}\DecValTok{3}\NormalTok{]])}
-
-\CommentTok{\# get list of aggregated data}
-\NormalTok{list\_of\_agg <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{120}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(z) \{}
-
-\NormalTok{  data\_return <{-}}\StringTok{ }\KeywordTok{atl\_thin\_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_agg,}
-                            \DataTypeTok{interval =}\NormalTok{ z,}
-                            \DataTypeTok{method =} \StringTok{"aggregate"}\NormalTok{)}
-
-\NormalTok{  data\_return[, interval }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{z]}
-
-  \KeywordTok{return}\NormalTok{(data\_return)}
-\NormalTok{\})}
-
-\CommentTok{\# get mean speed estimate and sd}
-\NormalTok{speed\_agg\_smooth <{-}}\StringTok{ }
-\StringTok{  }\KeywordTok{lapply}\NormalTok{(list\_of\_agg, }\ControlFlowTok{function}\NormalTok{(df) \{}
-    \KeywordTok{na.omit}\NormalTok{(df)}
-\NormalTok{    df[, speed }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_get\_speed}\NormalTok{(df)]}
-\NormalTok{    df[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{median =} \KeywordTok{median}\NormalTok{(speed, }\DataTypeTok{na.rm =}\NormalTok{ T),}
-              \DataTypeTok{sd =} \KeywordTok{sd}\NormalTok{(speed, }\DataTypeTok{na.rm =}\NormalTok{ T),}
-              \DataTypeTok{interval =} \KeywordTok{first}\NormalTok{(interval))]}
-\NormalTok{  \})}
-
-\CommentTok{\# bind}
-\NormalTok{speed\_agg\_smooth <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(speed\_agg\_smooth)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{aggregation-before-reducing-positioning-errors}{%
-\section{Aggregation Before Reducing Positioning Errors}\label{aggregation-before-reducing-positioning-errors}}
-
-Users may rightly wonder whether they can get away with aggregating their data, using a median aggregation function, and reduce data volumes, correct uneven sampling frequency, and reduce large-scale errors all in one go.
-
-The answer to most questions in ecology is, ``It depends''.
-Median aggregation before correcting positioning errors can indeed be advantageous; for instance for faster visualisation while preserving the broad structure of a track.
-
-However, there are drawbacks. The main one is that information is lost in the aggregation process, lending less power to steps such as smoothing applied after aggregation.
-Further, aggregation (and indeed any kind of thinning) results in significantly different estimates of speed and distance from the real speed.
-
-We show the effect of aggregating before any error correction here.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# read data with errors}
-\NormalTok{data\_errors <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_errors.csv"}\NormalTok{)}
-
-\CommentTok{\# aggregate before correction}
-\NormalTok{list\_of\_agg <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{120}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(z) \{}
-\NormalTok{  data\_return <{-}}\StringTok{ }\KeywordTok{atl\_thin\_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_errors,}
-                            \DataTypeTok{interval =}\NormalTok{ z,}
-                            \DataTypeTok{method =} \StringTok{"aggregate"}\NormalTok{)}
-\NormalTok{  data\_return[, interval }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{z]}
-\NormalTok{  data\_return[, speed }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_get\_speed}\NormalTok{(data\_return)]}
-  \KeywordTok{return}\NormalTok{(data\_return)}
-\NormalTok{\})}
-
-\CommentTok{\# get real speed}
-\NormalTok{data[, speed }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_get\_speed}\NormalTok{(data)]}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{comparing-median-aggregation-with-and-without-smoothing}{%
-\subsection{Comparing median aggregation with and without smoothing}\label{comparing-median-aggregation-with-and-without-smoothing}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# now plot distribution of speed}
-\NormalTok{data\_agg <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(list\_of\_agg)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{figures/fig_aggregate_errors.png}
-
-Thinning a movement track using median aggregation preserves track structure, but affects essential track metrics such as speed.
-\textbf{(a, b)} Movement tracks with a canonical interval of 1s aggregated over intervals of \textbf{(a)} 10 and \textbf{(b)} 30 seconds, without removing large- or small-scale positioning errors.
-All symbols represent positions in the aggregated track, with the size of the symbol representing the standard deviation at each position.
-Blue crosses represent positions with speed \(\leq\) the 95\^{}th percentile of canonical speeds, while red triangles represent positions with speed \(\geq\) 95\^{}th percentile of canonical speeds.
-\textbf{(c)} Boxplot of instantaneous speeds after median aggregation of a 1s interval track over intervals of 3, 10, 30, and 120 seconds, but without the removal of positioning errors.
-The mean and 95\^{}th percentile of speed in the canonical track are shown as solid and dashed lines, respectively.
-Aggregation without reducing positioning errors can result in speed estimates that are substantially different from the true speed.
-
-\hypertarget{residence-patches-and-their-construction}{%
-\chapter{Residence patches and their construction}\label{residence-patches-and-their-construction}}
-
-\hypertarget{prepare-libraries}{%
-\section*{Prepare libraries}\label{prepare-libraries}}
-\addcontentsline{toc}{section}{Prepare libraries}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(data.table)}
-\KeywordTok{library}\NormalTok{(atlastools)}
-\KeywordTok{library}\NormalTok{(ggplot2)}
-\KeywordTok{library}\NormalTok{(patchwork)}
-
-\CommentTok{\# for residence time}
-\KeywordTok{library}\NormalTok{(recurse)}
-
-\CommentTok{\# prepare a palette}
-\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{an-example-with-simulated-data}{%
-\section{An example with simulated data}\label{an-example-with-simulated-data}}
-
-\hypertarget{read-data-and-classify}{%
-\subsection{Read data and classify}\label{read-data-and-classify}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# read patch data}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_for\_res\_patch.csv"}\NormalTok{)}
-
-\CommentTok{\# filter}
-\NormalTok{data <{-}}\StringTok{ }\NormalTok{data[y }\OperatorTok{<}\StringTok{ }\DecValTok{21}\NormalTok{, ]}
-
-\CommentTok{\# do recurse}
-\NormalTok{data\_recurse <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(x, y, time, id)],}
-                              \DataTypeTok{radius =} \DecValTok{1}\NormalTok{)}
-
-\CommentTok{\# assign residence time}
-\NormalTok{data[, residence\_time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{data\_recurse}\OperatorTok{$}\NormalTok{residenceTime]}
-\end{Highlighting}
-\end{Shaded}
-
-We first plot a figure of residence time per positions, and residence time per timestamp.
-
-We construct residence patches from data where residence time is \textgreater{} 0.15.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# make residence patch}
-\NormalTok{patch <{-}}\StringTok{ }\KeywordTok{atl\_res\_patch}\NormalTok{(data[residence\_time }\OperatorTok{>}\StringTok{ }\FloatTok{0.04}\NormalTok{, ], }
-                       \DataTypeTok{buffer\_radius =} \FloatTok{0.1}\NormalTok{, }
-                       \DataTypeTok{lim\_spat\_indep =} \DecValTok{1}\NormalTok{, }
-                       \DataTypeTok{lim\_time\_indep =} \DecValTok{30}\NormalTok{)}
-
-\CommentTok{\# get spatial representation}
-\NormalTok{patch\_sf <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(}\DataTypeTok{patch\_data =}\NormalTok{ patch,}
-                              \DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{, }
-                              \DataTypeTok{buffer\_radius =} \FloatTok{0.15}\NormalTok{)}
-
-\CommentTok{\# get summary data}
-\NormalTok{patch\_summary <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(}\DataTypeTok{patch\_data =}\NormalTok{ patch,}
-                                   \DataTypeTok{which\_data =} \StringTok{"summary"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{plot-classified-residence-patches}{%
-\subsection{Plot classified residence patches}\label{plot-classified-residence-patches}}
-
-We arrange the figures together.
-
-\includegraphics{figures/fig_residence.png}
-
-\hypertarget{processing-calibration-data}{%
-\chapter{Processing calibration data}\label{processing-calibration-data}}
-
-\hypertarget{prepare-libraries}{%
-\section{Prepare libraries}\label{prepare-libraries}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# load libs}
-\KeywordTok{library}\NormalTok{(data.table)}
-\KeywordTok{library}\NormalTok{(atlastools)}
-\KeywordTok{library}\NormalTok{(ggplot2)}
-\KeywordTok{library}\NormalTok{(patchwork)}
-
-\CommentTok{\# prepare a palette}
-\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{preliminary-visualisation}{%
-\section{Preliminary visualisation}\label{preliminary-visualisation}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# read and plot example data}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/atlas1060\_allTrials\_annotated.csv"}\NormalTok{)}
-\NormalTok{data\_raw <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{filter-by-bounding-box}{%
-\section{Filter by bounding box}\label{filter-by-bounding-box}}
-
-Save an unprocessed copy.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-\end{Highlighting}
-\end{Shaded}
-
-Filter by a bounding box.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# remove inside must be set to falses}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_filter\_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data, }
-                          \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
-                          \DataTypeTok{x\_range =} \KeywordTok{c}\NormalTok{(}\DecValTok{645000}\NormalTok{, }\KeywordTok{max}\NormalTok{(data}\OperatorTok{$}\NormalTok{x)), }
-                          \DataTypeTok{remove\_inside =} \OtherTok{FALSE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{filter-trajectories}{%
-\section{Filter trajectories}\label{filter-trajectories}}
-
-\hypertarget{handle-time}{%
-\subsection{Handle time}\label{handle-time}}
-
-Time in ATLAS tracking is counted in milliseconds and is represented by a 64-bit integer (type \texttt{long}), which is not natively supported in R; it will instead be converted to a \texttt{numeric}, or \texttt{double}.
-
-This is not what is intended, but it works. The \texttt{bit64} package can help handle 64-bit integers if you want to keep to intended type.
-
-A further issue is that 64-bit integers (whether represented as \texttt{bit64} or \texttt{double}) do not yield meaninful results when you try to convert them to a date-time object, such as of the class \texttt{POSIXct}.
-
-This is because \texttt{as.POSIXct} fails when trying to work with 64-bit integers (it cannot interpret this type), and returns a date many thousands of years in the future (approx.
-52,000 CE) if the time column is converted to \texttt{numeric}.
-
-There are two possible solutions. The parsimonious one is to convert the 64-bit number to a 32-bit short integer (dividing by 1000), or to use the \texttt{nanotime} package.
-
-The conversion method loses an imperceptible amount of precision. The \texttt{nanotime} requires installing another package. The first method is shown here.
-
-In the spirit of not destroying data, we create a second lower-case column called \texttt{time}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# divide by 1000, convert to integer, then convert to POSIXct}
-\NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{add-speed-and-turning-angle}{%
-\subsection{Add speed and turning angle}\label{add-speed-and-turning-angle}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# add incoming and outgoing speed}
-\NormalTok{data[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{ (}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }
-                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
-                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
-                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
-             \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
-
-\CommentTok{\# add turning angle}
-\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_turning\_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{get-95th-percentile-of-speed-and-angle}{%
-\subsection{Get 95th percentile of speed and angle}\label{get-95th-percentile-of-speed-and-angle}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# use sapply}
-\NormalTok{speed\_angle\_thresholds <{-}}\StringTok{ }
-\StringTok{  }\KeywordTok{sapply}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(speed\_in, speed\_out, angle)], }
-\NormalTok{       quantile, }\DataTypeTok{probs =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{na.rm =}\NormalTok{ T)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{plot-to-see-speeds}{%
-\subsection{Plot to see speeds}\label{plot-to-see-speeds}}
-
-\includegraphics{figures/fig_speed_outlier.png}
-
-\hypertarget{filter-on-speed}{%
-\subsection{Filter on speed}\label{filter-on-speed}}
-
-Here we use a speed threshold of 15 m/s, the fastest known boat speed.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# make a copy}
-\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-
-\CommentTok{\# remove speed outliers}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
-            \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(speed\_in < 15 \& speed\_out < 15)"}\NormalTok{))}
-
-\CommentTok{\# recalculate speed and angle}
-\NormalTok{data[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{ (}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }
-                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
-                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
-                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
-             \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
-
-\CommentTok{\# add turning angle}
-\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_turning\_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{smoothing-the-trajectory}{%
-\section{Smoothing the trajectory}\label{smoothing-the-trajectory}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# apply a 5 point median smooth, first make a copy}
-\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-
-\CommentTok{\# now apply the smooth}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_median\_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
-                          \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }\DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
-                          \DataTypeTok{moving\_window =} \DecValTok{5}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{plot-pre-processing-steps}{%
-\subsection{Plot pre-processing steps}\label{plot-pre-processing-steps}}
-
-\includegraphics{figures/fig_walkthrough.png}
-
-\hypertarget{making-residence-patches}{%
-\section{Making residence patches}\label{making-residence-patches}}
-
-\hypertarget{prepare-data}{%
-\subsection{Prepare data}\label{prepare-data}}
-
-An indicator of individual residence at or near a position can be useful when attempting to identify residence patches.
-Positions can be filtered on a metric such as residence time (see Bracis et al.~2018).
-
-In this dataset, residence positions are marked in the \texttt{tID} column as beginning with \texttt{WP}. These can be extracted and converted into residence patches.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# make a data copy}
-\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{calculate-residence-time}{%
-\subsection{Calculate residence time}\label{calculate-residence-time}}
-
-Subset for boating segments: 2020-08-24 11:13:00 -- 2020-08-24 12:36:11.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# load recurse}
-\KeywordTok{library}\NormalTok{(recurse)}
-\NormalTok{data\_recurse <{-}}\StringTok{ }\NormalTok{data\_unproc[, }\KeywordTok{list}\NormalTok{(x, y, time, TAG, UTCtime)]}
-
-\CommentTok{\# thin the data}
-\NormalTok{data\_recurse <{-}}\StringTok{ }\KeywordTok{atl\_thin\_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data\_recurse,}
-                              \DataTypeTok{interval =} \DecValTok{30}\NormalTok{, }\DataTypeTok{id\_columns =} \StringTok{"TAG"}\NormalTok{,}
-                              \DataTypeTok{method =} \StringTok{"resample"}\NormalTok{)}
-
-\CommentTok{\# get 4 column data}
-\NormalTok{data\_recurse <{-}}\StringTok{ }\NormalTok{data\_recurse[, }\KeywordTok{list}\NormalTok{(x, y, time, TAG)]}
-
-\CommentTok{\# get recurse data for a 10m radius}
-\NormalTok{recurse\_stats <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data\_recurse,}
-                               \DataTypeTok{radius =} \DecValTok{50}\NormalTok{, }\DataTypeTok{timeunits =} \StringTok{"mins"}\NormalTok{)}
-
-\CommentTok{\# assign to recurse data}
-\NormalTok{data\_recurse[, res\_time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{recurse\_stats}\OperatorTok{$}\NormalTok{residenceTime]}
-
-\CommentTok{\# save recurse data}
-\KeywordTok{fwrite}\NormalTok{(data\_recurse, }\DataTypeTok{file =} \StringTok{"data/data\_calib\_recurse.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Read in the data.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{data\_recurse <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/data\_calib\_recurse.csv"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Subset waypoint data.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(stringi)}
-\NormalTok{data\_res <{-}}\StringTok{ }\NormalTok{data\_unproc[}\KeywordTok{stri\_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{)]}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{plot-residence-time}{%
-\subsection{Plot residence time}\label{plot-residence-time}}
-
-\hypertarget{run-residence-patch-method}{%
-\subsection{Run residence patch method}\label{run-residence-patch-method}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# assign id as tag}
-\NormalTok{data\_recurse[, id }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.character}\NormalTok{(TAG)]}
-
-\CommentTok{\# subset on 5 minute residence time}
-\NormalTok{data\_recurse\_subset <{-}}\StringTok{ }\NormalTok{data\_recurse[res\_time }\OperatorTok{>=}\StringTok{ }\DecValTok{5}\NormalTok{, ]}
-
-\CommentTok{\# on known residence points}
-\NormalTok{patch\_res\_known <{-}}\StringTok{ }\KeywordTok{atl\_res\_patch}\NormalTok{(data\_recurse\_subset, }
-                                \DataTypeTok{buffer\_radius =} \DecValTok{10}\NormalTok{,}
-                                \DataTypeTok{lim\_spat\_indep =} \DecValTok{100}\NormalTok{,}
-                                \DataTypeTok{lim\_time\_indep =} \DecValTok{5}\NormalTok{,}
-                                \DataTypeTok{min\_fixes =} \DecValTok{3}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{get-spatial-and-summary-objects}{%
-\subsection{Get spatial and summary objects}\label{get-spatial-and-summary-objects}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# for the known and unkniwn patches}
-\NormalTok{patch\_sf\_data <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(patch\_res\_known, }\DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{,}
-                                   \DataTypeTok{buffer\_radius =} \DecValTok{20}\NormalTok{)}
-
-\CommentTok{\# assign crs}
-\NormalTok{sf}\OperatorTok{::}\KeywordTok{st\_crs}\NormalTok{(patch\_sf\_data) <{-}}\StringTok{ }\DecValTok{32631}
-
-\CommentTok{\# get summary data}
-\NormalTok{patch\_summary\_data <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(patch\_res\_known, }\DataTypeTok{which\_data =} \StringTok{"summary"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{get-waypoint-centroids}{%
-\subsection{Get waypoint centroids}\label{get-waypoint-centroids}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# get centroid}
-\NormalTok{data\_res\_summary <{-}}\StringTok{ }\NormalTok{data\_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{x\_median =} \KeywordTok{median}\NormalTok{(x),}
-                                    \DataTypeTok{y\_median =} \KeywordTok{median}\NormalTok{(y),}
-                                    \DataTypeTok{t\_median =} \KeywordTok{median}\NormalTok{(time)),}
-\NormalTok{                             by =}\StringTok{ "tID"}\NormalTok{]}
-
-\CommentTok{\# now get times 10 mins before and after}
-\NormalTok{data\_res\_summary[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{t\_min =}\NormalTok{ t\_median }\OperatorTok{{-}}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{),}
-                        \DataTypeTok{t\_max =}\NormalTok{ t\_median }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{))]}
-
-\CommentTok{\# make a list of positions 10min before and after}
-\NormalTok{wp\_data <{-}}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(l, u, mx, my) \{}
-\NormalTok{  tmp\_data <{-}}\StringTok{ }\NormalTok{data\_unproc[}\KeywordTok{inrange}\NormalTok{(time, l, u)]}
-\NormalTok{  tmp\_data[, distance }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{sqrt}\NormalTok{((mx }\OperatorTok{{-}}\StringTok{ }\NormalTok{x)}\OperatorTok{\^{}}\DecValTok{2} \OperatorTok{+}\StringTok{ }\NormalTok{(my }\OperatorTok{{-}}\StringTok{ }\NormalTok{y)}\OperatorTok{\^{}}\DecValTok{2}\NormalTok{)]}
-  \CommentTok{\# keep within 50}
-\NormalTok{  tmp\_data <{-}}\StringTok{ }\NormalTok{tmp\_data[distance }\OperatorTok{<=}\StringTok{ }\DecValTok{50}\NormalTok{, ]}
-  \CommentTok{\# get duration}
-  \KeywordTok{return}\NormalTok{(}\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(tmp\_data}\OperatorTok{$}\NormalTok{time)))}
-\NormalTok{\}, data\_res\_summary}\OperatorTok{$}\NormalTok{t\_min, data\_res\_summary}\OperatorTok{$}\NormalTok{t\_max,}
-\NormalTok{   data\_res\_summary}\OperatorTok{$}\NormalTok{x\_median, data\_res\_summary}\OperatorTok{$}\NormalTok{y\_median,}
-\DataTypeTok{SIMPLIFY =} \OtherTok{FALSE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{read-griend}{%
-\subsection{Read Griend}\label{read-griend}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# read griend}
-\NormalTok{griend <{-}}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st\_read}\NormalTok{(}\StringTok{"data/griend\_polygon/griend\_polygon.shp"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-We prepare a plot of the residence patches.
-
-\hypertarget{compare-patch-metrics}{%
-\section{Compare patch metrics}\label{compare-patch-metrics}}
-
-\hypertarget{compare-known-patches}{%
-\subsection{Compare known patches}\label{compare-known-patches}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# get known patch summary}
-\NormalTok{data\_res <{-}}\StringTok{ }\NormalTok{data\_unproc[stringi}\OperatorTok{::}\KeywordTok{stri\_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{), ]}
-
-\CommentTok{\# get waypoint summary}
-\NormalTok{patch\_summary\_real <{-}}\StringTok{ }\NormalTok{data\_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{nfixes\_real =}\NormalTok{ .N,}
-                                      \DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
-                                      \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{)), }
-\NormalTok{                               by =}\StringTok{ "tID"}\NormalTok{]}
-
-\CommentTok{\# add real duration}
-\NormalTok{patch\_summary\_real[, duration\_real }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{unlist}\NormalTok{(wp\_data)]}
-\CommentTok{\# round median coordinate for inferred patches}
-\NormalTok{patch\_summary\_inferred <{-}}\StringTok{ }
-\StringTok{  }\NormalTok{patch\_summary\_data[, }
-                     \KeywordTok{c}\NormalTok{(}\StringTok{"x\_median"}\NormalTok{, }\StringTok{"y\_median"}\NormalTok{, }
-                       \StringTok{"nfixes"}\NormalTok{, }\StringTok{"duration"}\NormalTok{, }\StringTok{"patch"}\NormalTok{)}
-\NormalTok{                     ][, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(x\_median, }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
-                              \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(y\_median, }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{))]}
-
-\CommentTok{\# join with respatch summary}
-\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }
-\StringTok{  }\KeywordTok{merge}\NormalTok{(patch\_summary\_real,}
-\NormalTok{        patch\_summary\_inferred, }
-        \DataTypeTok{on =} \KeywordTok{c}\NormalTok{(}\StringTok{"x\_median"}\NormalTok{, }\StringTok{"y\_median"}\NormalTok{),}
-        \DataTypeTok{all.x =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{all.y =} \OtherTok{TRUE}\NormalTok{)}
-
-\NormalTok{patch\_summary\_compare[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(tID)]}
-
-\CommentTok{\# drop nas}
-\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(patch\_summary\_compare)}
-
-\CommentTok{\# drop patch around WP080}
-\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }\NormalTok{patch\_summary\_compare[tID }\OperatorTok{!=}\StringTok{ "WP080"}\NormalTok{, ]}
-\end{Highlighting}
-\end{Shaded}
-
-12 patches are identified where there are no waypoints, while 2 waypoints are not identified as patches.
-These waypoints consisted of 6 and 15 (WP098 and WP092) positions respectively, and were lost when the data were aggregated to 30 second intervals.
-
-\hypertarget{plot-durations-comparisons}{%
-\subsection{Plot durations comparisons}\label{plot-durations-comparisons}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# get linear model}
-\NormalTok{model\_duration <{-}}\StringTok{ }\KeywordTok{lm}\NormalTok{(duration\_real }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{duration,}
-                     \DataTypeTok{data =}\NormalTok{ patch\_summary\_compare)}
-
-\CommentTok{\# get R2}
-\KeywordTok{summary}\NormalTok{(model\_duration)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{join-patch-map-and-duration-comparison-figure}{%
-\subsection{Join patch map and duration comparison figure}\label{join-patch-map-and-duration-comparison-figure}}
-
-\includegraphics{figures/fig_calib_residence_patch.png}
-
-\hypertarget{processing-egyptian-fruit-bat-tracks}{%
-\chapter{Processing Egyptian fruit bat tracks}\label{processing-egyptian-fruit-bat-tracks}}
-
-We show the pre-processing pipeline at work on the tracks of three Egyptian fruit bats (\emph{Rousettus aegyptiacus}), and construct residence patches.
-
-\hypertarget{prepare-libraries}{%
-\section{Prepare libraries}\label{prepare-libraries}}
+\subsection{Prepare libraries}\label{prepare-libraries}}
+
+Install the required \texttt{R} libraries that are available on CRAN.
 
@@
 \KeywordTok{library}\NormalTok{(RSQLite)}
-\KeywordTok{library}\NormalTok{(atlastools)}
 \KeywordTok{library}\NormalTok{(ggplot2)}
@@
 
+\hypertarget{install-atlastools-from-github.}{%
+\subsection{\texorpdfstring{Install \texttt{atlastools} from Github.}{Install atlastools from Github.}}\label{install-atlastools-from-github.}}
+
+\texttt{atlastools} is available from Github and is archived on Zenodo (Gupte \protect\hyperlink{ref-gupte2020a}{2020}).
+It can be installed using \texttt{remotes} or \texttt{devtools}. Here we use the \texttt{remotes} function \texttt{install\_github}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{install.packages}\NormalTok{(}\StringTok{"remotes"}\NormalTok{)}
+
+\CommentTok{\# installation using remotes}
+\NormalTok{remotes}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
 \hypertarget{read-bat-data}{%
-\section{Read bat data}\label{read-bat-data}}
-
-Read the bat data and convert to a csv file.
+\subsection{Read bat data}\label{read-bat-data}}
+
+Read the bat data from an \texttt{SQLite} database local file and convert to a plain text csv file.
+This data can be found in the ``data'' folder.
 
@@
 \CommentTok{\# prepare the connection}
-\NormalTok{con <{-}}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}\DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(), }\DataTypeTok{dbname =} \StringTok{"data/Three\_example\_bats.sql"}\NormalTok{)}
+\NormalTok{con <{-}}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}\DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(), }
+                 \DataTypeTok{dbname =} \StringTok{"data/Three\_example\_bats.sql"}\NormalTok{)}
 
@@
 
-Convert data to csv.
+Convert data to csv, and save a local copy in the folder ``data''.
 
@@
 
-\hypertarget{sanity-check-plot-bat-data}{%
-\section{Sanity check: Plot bat data}\label{sanity-check-plot-bat-data}}
-
-Save the raw data plot.
-
-Show the plot.
-
+\hypertarget{a-first-visual-inspection}{%
+\subsection{A First Visual Inspection}\label{a-first-visual-inspection}}
+
+Plot the bat data as a sanity check, and inspect it visually for errors (Figure 1).
+The plot code is hidden in the rendered copy (PDF) of this supplementary material, but is available in the \texttt{Rmarkdown} file ``06\_bat\_data.Rmd''.
+The saved plot is shown below as Figure 1.
+
+\begin{figure}
+\centering
 \includegraphics{figures/fig_bat_raw.png}
+\caption{Movement data from three Egyptian fruit bats tracked using the ATLAS system (\emph{Rousettus aegyptiacus}; (Toledo et al.
+\protect\hyperlink{ref-toledo2020}{2020}; Shohami and Nathan \protect\hyperlink{ref-shohami2020}{2020})).
+The bats were tracked in the Hula Valley, Israel (33.1\(^{\circ}\)N, 35.6\(^{\circ}\)E), and we use three nights of tracking (5\textsuperscript{th}, 6\textsuperscript{th}, and 7\textsuperscript{th} May, 2018), for our demonstration, with an average of 13,370 positions (SD = 2,173; range = 11,195 -- 15,542; interval = 8 seconds) per individual.
+After first plotting the individual tracks, we notice severe distortions, making pre-processing necesary}
+\end{figure}
 
 \hypertarget{prepare-data-for-filtering}{%
-\section{Prepare data for filtering}\label{prepare-data-for-filtering}}
+\subsection{Prepare data for filtering}\label{prepare-data-for-filtering}}
+
+Here we apply a series of simple filters.
+It is always safer to deal with one individual at a time, so we split the data.table
+into a list of data.tables to avoid mixups among individuals.
 
 \hypertarget{prepare-data-per-individual}{%
-\subsection{Prepare data per individual}\label{prepare-data-per-individual}}
+\subsubsection{Prepare data per individual}\label{prepare-data-per-individual}}
 
@@
 \CommentTok{\# now split}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_split, data}\OperatorTok{$}\NormalTok{TAG)}
+\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_split, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
 \end{Highlighting}
@@
 \hypertarget{filter-by-covariates}{%
-\section{Filter by covariates}\label{filter-by-covariates}}
-
-No natural bounds suggest themselves, and we proceed to filter by covariates, since point outliers are obviously visible.
-
-We use filter out positions with \texttt{SD\ \textgreater{}\ 20} and positions calculated using only 3 base stations.
-
-We use the function \texttt{atl\_filter\_covariates}.
+\subsection{Filter by covariates}\label{filter-by-covariates}}
+
+No natural bounds suggest themselves, so instead we proceed to filter by covariates, since point outliers are obviously visible.
+
+We use filter out positions with \texttt{SD\ \textgreater{}\ 20} and positions calculated using only 3 base stations, using the function \texttt{atl\_filter\_covariates}.
+
+First we calculate the variable \texttt{SD}.
 
@@
 \NormalTok{)}
-
+\end{Highlighting}
+\end{Shaded}
+
+Then we pass the filters to \texttt{atl\_filter\_covariates}.
+We apply the filter to each individual's data using an \texttt{lapply}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \CommentTok{\# filter for SD <= 20}
@@
 \NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+
 \NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}
@@
                 \StringTok{"NBS > 3"}\NormalTok{)}
+
 \NormalTok{  )}
 \NormalTok{\})}
-
-\CommentTok{\# check whether the filter has worked}
-\KeywordTok{invisible}\NormalTok{(}
-  \KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
-\NormalTok{    assertthat}\OperatorTok{::}\KeywordTok{assert\_that}\NormalTok{(}\KeywordTok{min}\NormalTok{(dt}\OperatorTok{$}\NormalTok{SD) }\OperatorTok{<=}\StringTok{ }\DecValTok{20}\NormalTok{,}
-                            \DataTypeTok{msg =} \StringTok{"some SDs above 20 remain"}\NormalTok{)}
-\NormalTok{    assertthat}\OperatorTok{::}\KeywordTok{assert\_that}\NormalTok{(}\KeywordTok{min}\NormalTok{(dt}\OperatorTok{$}\NormalTok{NBS) }\OperatorTok{>}\StringTok{ }\DecValTok{3}\NormalTok{,}
-                            \DataTypeTok{msg =} \StringTok{"some NBS below 3 remain"}\NormalTok{)}
-\NormalTok{  \})}
-\NormalTok{)}
 \end{Highlighting}
@@
 \hypertarget{sanity-check-plot-filtered-data}{%
-\subsection{Sanity check: Plot filtered data}\label{sanity-check-plot-filtered-data}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# bind all individuals together}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_split)}
-\end{Highlighting}
-\end{Shaded}
-
-Save the filtered data plot.
-
-Show the plot.
-
+\subsubsection{Sanity check: Plot filtered data}\label{sanity-check-plot-filtered-data}}
+
+We plot the data to check whether the filtering has improved the data (Figure 2).
+The plot code is once again hidden in this rendering, but is available in the source code file.
+
+\begin{figure}
+\centering
 \includegraphics{figures/fig_bat_filter_cov.png}
+\caption{Bat data filtered for large location errors, removing observations with standard deviation \(>\) 20.
+Grey crosses show data that were removed.
+Since the number of base stations used in the location process is a good indicator of error (Weiser et al.
+\protect\hyperlink{ref-weiser2016}{2016}), we also removed observations calculated using fewer than four base stations.
+Both steps used the function \texttt{atl\_filter\_covariates}.
+This filtering reduced the data to an average of 10,447 positions per individual (78\% of the raw data on average). However, some point outliers remain.}
+\end{figure}
 
 \hypertarget{filter-by-speed}{%
-\section{Filter by speed}\label{filter-by-speed}}
-
-Some point outliers remain, and should be removed using a speed filter.
-
-First we calculate speeds.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# calculate speed on split data once more}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_split, data\_split}\OperatorTok{$}\NormalTok{TAG)}
-
+\subsection{Filter by speed}\label{filter-by-speed}}
+
+Some point outliers remain (Figure 2), and could be removed using a speed filter.
+
+First we calculate speeds, using \texttt{atl\_get\_speed}.
+We must assign the speed output to a new column in the data.table, which has a special syntax which modifies in place, and is shown below.
+This syntax is a feature of the \texttt{data.table} package, not strictly of \texttt{atlastools} (Dowle and Srinivasan \protect\hyperlink{ref-dowle2020}{2020}).
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \CommentTok{\# get speeds as with SD, no reassignment required for columns}
@@
 
-Now filter for speeds \textgreater{} 20 m/s (around 70 km/h).
+Now filter for speeds \textgreater{} 20 m/s (around 70 km/h), passing the predicate (a statement return TRUE or FALSE) to \texttt{atl\_filter\_covariates}.
+First, we remove positions which have \texttt{NA} for their \texttt{speed\_in} (the first position) and their \texttt{speed\_out} (last position).
 
@@
 \hypertarget{sanity-check-plot-speed-filtered-data}{%
-\subsection{Sanity check: Plot speed filtered data}\label{sanity-check-plot-speed-filtered-data}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# bind all individuals together}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_split)}
-\end{Highlighting}
-\end{Shaded}
-
-Save the speed filtered data plot.
-
-Show the plot.
-
+\subsubsection{Sanity check: Plot speed filtered data}\label{sanity-check-plot-speed-filtered-data}}
+
+The speed filtered data is now inspected for errors (Figure 3). The plot code is once again hidden.
+
+\begin{figure}
+\centering
 \includegraphics{figures/fig_bat_filter_speed.png}
+\caption{Bat data with unrealistic speeds removed.
+Grey crosses show data that were removed.
+We calculated the incoming and outgoing speed of each position using \texttt{atl\_get\_speed}, and filtered out positions with speeds \textgreater{} 20 m/s using \texttt{atl\_filter\_covariates}, leaving 10,337 positions per individual on average (98\% from the previous step).}
+\end{figure}
 
 \hypertarget{median-smoothing}{%
-\section{Median smoothing}\label{median-smoothing}}
-
-Apply a 5 point median smooth to the data.
+\subsection{Median smoothing}\label{median-smoothing}}
+
+The quality of this data (Figure 3) is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless.
+
+Since the median smoothing function \texttt{atl\_median\_smooth} modifies in place, we first make a copy of the data, using \texttt{data.table}'s \texttt{copy} function.
+No reassignment is required, in this case. The \texttt{lapply} function allows arguments to \texttt{atl\_median\_smooth} to be passed within \texttt{lapply} itself.
+
+In this case, the same moving window \(K\) is applied to all individuals, but modifying this code to use the multivariate version \texttt{Map} allows different \(K\) to be used for different individuals.
+This is a programming matter, and is not covered here further.
 
@@
 \CommentTok{\# split the data again}
-\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_smooth, data\_smooth}\OperatorTok{$}\NormalTok{TAG)}
-\end{Highlighting}
-\end{Shaded}
-
-Remember, \textbf{\texttt{atl\_median\_smooth} MODIFIES IN PLACE}.
+\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_smooth, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 \KeywordTok{invisible}\NormalTok{(}
+
   \CommentTok{\# the function arguments to atl\_median\_smooth}
   \CommentTok{\# can be passed directly in lapply}
-  \KeywordTok{lapply}\NormalTok{(data\_smooth, atl\_median\_smooth,}
+
+  \KeywordTok{lapply}\NormalTok{(}\DataTypeTok{X =}\NormalTok{ data\_smooth, }
+         \DataTypeTok{FUN =}\NormalTok{ atl\_median\_smooth,}
          \DataTypeTok{time =} \StringTok{"time"}\NormalTok{, }\DataTypeTok{moving\_window =} \DecValTok{5}\NormalTok{)}
@@
 \hypertarget{sanity-check-plot-smoothed-data}{%
-\subsection{Sanity check: Plot smoothed data}\label{sanity-check-plot-smoothed-data}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# recombine split up data that has been smoothed}
-\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_smooth)}
-\end{Highlighting}
-\end{Shaded}
-
-Save the smoothed data plot.
-
-Show the plot.
-
+\subsubsection{Sanity check: Plot smoothed data}\label{sanity-check-plot-smoothed-data}}
+
+\begin{figure}
+\centering
 \includegraphics{figures/fig_bat_smooth.png}
+\caption{Bat data after applying a median smooth with a moving window \(K\) = 5. Grey crosses show data prior to smoothing. The smoothing step did not discard any data.}
+\end{figure}
 
 \hypertarget{making-residence-patches}{%
-\section{Making residence patches}\label{making-residence-patches}}
+\subsection{Making residence patches}\label{making-residence-patches}}
 
 \hypertarget{calculating-residence-time}{%
-\subsection{Calculating residence time}\label{calculating-residence-time}}
-
-First, the data is put through the \texttt{recurse} package to get residence time.
+\subsubsection{Calculating residence time}\label{calculating-residence-time}}
+
+First, the data is put through the \texttt{recurse} package to get residence time (Bracis, Bildstein, and Mueller \protect\hyperlink{ref-bracis2018}{2018}).
 
@@
 
-Get residence time. Since bats may revisit the same features, we want to prevent confusion between frequent revisits and prolonged residence.
-
-For this, we stop summing residence times within Z metres of a location if the animal exited the area for one hour or more.
+We calculated residence time, but since bats may revisit the same features, we want to prevent confusion between frequent revisits and prolonged residence.
+
+For this, we stop summing residence times within \(Z\) metres of a location if the animal exited the area for one hour or more.
+The value of \(Z\) (radius, in \texttt{recurse} parameter terms) was chosen as 50m.
+
+This step is relatively complicated and is only required for individuals which frequently return to the same location, or pass over the same areas repeatedly, and for which revisits (cumulative time spent) may be confused for residence time in a single visit.
+
+While a simpler implementation using total residence time divided by the number of revisits is also possible, this does assume that each revisit had the same residence time.
 
@@
 
-\hypertarget{sanity-check-residence-time-time-series}{%
-\subsection{Sanity check: Residence-time time-series}\label{sanity-check-residence-time-time-series}}
+We bind the data together and assign a human readable timestamp column.
 
@@
 
-Save the speed filtered data plot.
-
-Show the plot.
-
-\includegraphics{figures/fig_bat_res_timeseries.png}
-
 \hypertarget{constructing-residence-patches}{%
-\subsection{Constructing residence patches}\label{constructing-residence-patches}}
-
-Split the data and construct residence patches.
+\subsubsection{Constructing residence patches}\label{constructing-residence-patches}}
 
 Some preparation is required. First, the function requires columns \texttt{x}, \texttt{y},
-\texttt{time}, and \texttt{id}.
+\texttt{time}, and \texttt{id}, which we assign using the \texttt{data.table} syntax.
+Then we subset the data to only work with positions where the individual had a residence time of more than 5 minutes.
 
@@
 
-Now segment-cluster into residence patches.
+We apply the residence patch method, using the default argument values (\texttt{lim\_spat\_indep\ =\ 100} (metres), \texttt{lim\_time\_indep\ =\ 30} (minutes), and \texttt{min\_fixes\ =\ 3}).
+We change the \texttt{buffer\_radius} to 25 metres (twice the buffer radius is used, so points must be separated by 50m to be independent bouts).
 
@@
 \hypertarget{getting-residence-patch-data}{%
-\subsection{Getting residence patch data}\label{getting-residence-patch-data}}
-
-We get the residence patch data as spatial \texttt{sf-MULTIPOLYGON} objects.
+\subsubsection{Getting residence patch data}\label{getting-residence-patch-data}}
+
+We extract the residence patch data as spatial \texttt{sf-MULTIPOLYGON} objects.
+These are returned as a list and must be converted into a single \texttt{sf} object.
+These objects and the raw movement data are shown in Figure 5.
 
@@
 \hypertarget{write-patch-spatial-representations}{%
-\subsection{Write patch spatial representations}\label{write-patch-spatial-representations}}
+\subsubsection{Write patch spatial representations}\label{write-patch-spatial-representations}}
 
@@
 
-\hypertarget{processed-bat-patches}{%
-\section{Processed bat patches}\label{processed-bat-patches}}
-
-This figure made in QGIS.
-
+\begin{figure}
+\centering
 \includegraphics{figures/fig_bat_patches.png}
+\caption{A visual examination of plots of the bats' residence patches and linear approximations of paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site over the three nights \textbf{(a)}.
+Bats tended to be resident near fruit trees, which are their main food source, travelling repeatedly between previously visited areas \textbf{(b, c)}.
+However, bats also appeared to spend some time at locations where no fruit trees were recorded, prompting questions about their use of other food sources \textbf{(b, c)}.
+When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar \textbf{(c)}.
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap \textbf{(b, c)}.}
+\end{figure}
 
 \hypertarget{references}{%
-\chapter{References}\label{references}}
+\section{References}\label{references}}
 
@@
 \begin{cslreferences}
-\leavevmode\hypertarget{ref-bjorneraas2010}{}%
-Bjørneraas, Kari, Bram Van Moorter, Christer Moe Rolandsen, and Ivar Herfindal.
-2010.
-``Screening Global Positioning System Location Data for Errors Using Animal Movement Characteristics.'' \emph{The Journal of Wildlife Management} 74 (6): 1361--6.
-\url{https://doi.org/10.1111/j.1937-2817.2010.tb01258.x}.
-
-\leavevmode\hypertarget{ref-gurarie2017}{}%
-Gurarie, Eliezer, Christen H.
-Fleming, William F.
-Fagan, Kristin L.
-Laidre, Jesús Hernández-Pliego, and Otso Ovaskainen.
-2017.
-``Correlated Velocity Models as a Fundamental Unit of Animal Movement: Synthesis and Applications.'' \emph{Movement Ecology} 5 (1): 13.
-\url{https://doi.org/10.1186/s40462-017-0103-3}.
-
-\leavevmode\hypertarget{ref-noonan2019}{}%
-Noonan, Michael J., Christen H.
-Fleming, Thomas S.
-Akre, Jonathan Drescher-Lehman, Eliezer Gurarie, Autumn-Lynn Harrison, Roland Kays, and Justin M.
-Calabrese.
-2019.
-``Scale-Insensitive Estimation of Speed and Distance Traveled from Animal Tracking Data.'' \emph{Movement Ecology} 7 (1): 35.
-\url{https://doi.org/10.1186/s40462-019-0177-1}.
+\leavevmode\hypertarget{ref-bracis2018}{}%
+Bracis, Chloe, Keith L.
+Bildstein, and Thomas Mueller.
+2018.
+``Revisitation Analysis Uncovers Spatio-Temporal Patterns in Animal Movement Data.'' \emph{Ecography} 41 (11): 1801--11.
+\url{https://doi.org/10.1111/ecog.03618}.
+
+\leavevmode\hypertarget{ref-dowle2020}{}%
+Dowle, Matt, and Arun Srinivasan. 2020. \emph{Data.Table: Extension of `data.Frame`}. Manual.
+
+\leavevmode\hypertarget{ref-gupte2020a}{}%
+Gupte, Pratik Rajan. 2020. ``Atlastools: Pre-Processing Tools for High Frequency Tracking Data.'' Zenodo. \url{https://doi.org/10.5281/ZENODO.4033154}.
+
+\leavevmode\hypertarget{ref-shohami2020}{}%
+Shohami, David, and Ran Nathan. 2020. ``Cognitive Map-Based Navigation in Wild Bats Revealed by a New High-Throughput Tracking System.'' Dryad. \url{https://doi.org/10.5061/DRYAD.G4F4QRFN2}.
+
+\leavevmode\hypertarget{ref-toledo2020}{}%
+Toledo, Sivan, David Shohami, Ingo Schiffner, Emmanuel Lourie, Yotam Orchan, Yoav Bartan, and Ran Nathan.
+2020.
+``Cognitive MapBased Navigation in Wild Bats Revealed by a New High-Throughput Tracking System.'' \emph{Science} 369 (6500): 188--93.
+\url{https://doi.org/10.1126/science.aax6904}.
+
+\leavevmode\hypertarget{ref-weiser2016}{}%
+Weiser, Adi Weller, Yotam Orchan, Ran Nathan, Motti Charter, Anthony J.
+Weiss, and Sivan Toledo.
+2016.
+``Characterizing the Accuracy of a Self-Synchronized Reverse-GPS Wildlife Localization System.'' In \emph{2016 15th ACM/IEEE International Conference on Information Processing in Sensor Networks (IPSN)}, 1--12.
+\url{https://doi.org/10.1109/IPSN.2016.7460662}.
 \end{cslreferences}

### 2020-11-16 06_bat_data.Rmd
minor changes to prep for supplement

@@
   html_document: default
+  bibliography: references.bib
 ---
@@
 
-We show the pre-processing pipeline at work on the tracks of three Egyptian fruit bats (_Rousettus aegyptiacus_), and construct residence patches.
+We show the pre-processing pipeline at work on the tracks of three Egyptian fruit bats (*Rousettus aegyptiacus*), and construct residence patches.
 
 ## Prepare libraries
+
+Install the required `R` libraries that are available on CRAN.
 
@@
 library(RSQLite)
-library(atlastools)
 library(ggplot2)
@@
 
+## Install `atlastools` from Github.
+
+`atlastools` is available from Github and is archived on Zenodo [@gupte2020a].
+It can be installed using `remotes` or `devtools`. Here we use the `remotes` function `install_github`.
+
+```{r install_atlastools}
+install.packages("remotes")
+
+# installation using remotes
+remotes::install_github("pratikunterwegs/atlastools")
+```
+
 ## Read bat data
 
-Read the bat data and convert to a csv file.
+Read the bat data from an `SQLite` database local file and convert to a plain text csv file.
+This data can be found in the "data" folder.
 
@@
 # prepare the connection
-con <- dbConnect(drv = SQLite(), dbname = "data/Three_example_bats.sql")
+con <- dbConnect(drv = SQLite(), 
+                 dbname = "data/Three_example_bats.sql")
 
@@
 
-Convert data to csv.
+Convert data to csv, and save a local copy in the folder "data".
 
@@
 
-## Sanity check: Plot bat data
+## A First Visual Inspection
+
+Plot the bat data as a sanity check, and inspect it visually for errors (Figure 1).
+The plot code is hidden in the rendered copy (PDF) of this supplementary material, but is available in the `Rmarkdown` file "06_bat_data.Rmd".
+The saved plot is shown below as Figure 1.
 
@@
 
-Save the raw data plot.
-
 ```{r echo=FALSE}
@@
 
-Show the plot.
-
-![](figures/fig_bat_raw.png)
+![Movement data from three Egyptian fruit bats tracked using the ATLAS system (*Rousettus aegyptiacus*; [@toledo2020; @shohami2020]).
+The bats were tracked in the Hula Valley, Israel (33.1$^{\circ}$N, 35.6$^{\circ}$E), and we use three nights of tracking (5\textsuperscript{th}, 6\textsuperscript{th}, and 7\textsuperscript{th} May, 2018), for our demonstration, with an average of 13,370 positions (SD = 2,173; range = 11,195 -- 15,542; interval = 8 seconds) per individual.
+After first plotting the individual tracks, we notice severe distortions, making pre-processing necesary](figures/fig_bat_raw.png)
 
 ## Prepare data for filtering
+
+Here we apply a series of simple filters.
+It is always safer to deal with one individual at a time, so we split the data.table
+into a list of data.tables to avoid mixups among individuals.
 
@@
 # now split
-data_split <- split(data_split, data$TAG)
+data_split <- split(data_split, by = "TAG")
 ```
@@
 
-No natural bounds suggest themselves, and we proceed to filter by covariates, since point outliers are obviously visible.
+No natural bounds suggest themselves, so instead we proceed to filter by covariates, since point outliers are obviously visible.
 
-We use filter out positions with `SD > 20` and positions calculated using only 3 base stations.
+We use filter out positions with `SD > 20` and positions calculated using only 3 base stations, using the function `atl_filter_covariates`.
 
-We use the function `atl_filter_covariates`.
+First we calculate the variable `SD`.
 
@@
 )
+```
 
+Then we pass the filters to `atl_filter_covariates`.
+We apply the filter to each individual's data using an `lapply`.
+
+```{r}
 # filter for SD <= 20
@@
 data_split <- lapply(data_split, function(dt) {
+
   dt <- atl_filter_covariates(
@@
                 "NBS > 3")
+
   )
 })
-
-# check whether the filter has worked
-invisible(
-  lapply(data_split, function(dt) {
-    assertthat::assert_that(min(dt$SD) <= 20,
-                            msg = "some SDs above 20 remain")
-    assertthat::assert_that(min(dt$NBS) > 3,
-                            msg = "some NBS below 3 remain")
-  })
-)
 ```
@@
 
-```{r}
+We plot the data to check whether the filtering has improved the data (Figure 2).
+The plot code is once again hidden in this rendering, but is available in the source code file.
+
+```{r echo=FALSE}
 # bind all individuals together
@@
 
-Save the filtered data plot.
-
 ```{r echo=FALSE}
@@
 
-Show the plot.
-
-![](figures/fig_bat_filter_cov.png)
+![Bat data filtered for large location errors, removing observations with standard deviation $>$ 20.
+Grey crosses show data that were removed.
+Since the number of base stations used in the location process is a good indicator of error [@weiser2016], we also removed observations calculated using fewer than four base stations.
+Both steps used the function \texttt{atl\_filter\_covariates}.
+This filtering reduced the data to an average of 10,447 positions per individual (78% of the raw data on average). However, some point outliers remain.](figures/fig_bat_filter_cov.png)
 
@@
 
-Some point outliers remain, and should be removed using a speed filter.
+Some point outliers remain (Figure 2), and could be removed using a speed filter.
 
-First we calculate speeds.
+First we calculate speeds, using `atl_get_speed`.
+We must assign the speed output to a new column in the data.table, which has a special syntax which modifies in place, and is shown below.
+This syntax is a feature of the `data.table` package, not strictly of `atlastools` [@dowle2020].
+
+```{r echo=FALSE}
+# calculate speed on split data once more
+data_split <- split(data_split, data_split$TAG)
+```
 
 ```{r}
-# calculate speed on split data once more
-data_split <- split(data_split, data_split$TAG)
-
 # get speeds as with SD, no reassignment required for columns
@@
 
-Now filter for speeds > 20 m/s (around 70 km/h).
+Now filter for speeds > 20 m/s (around 70 km/h), passing the predicate (a statement return TRUE or FALSE) to `atl_filter_covariates`.
+First, we remove positions which have `NA` for their `speed_in` (the first position) and their `speed_out` (last position).
 
@@
 
-```{r}
+The speed filtered data is now inspected for errors (Figure 3). The plot code is once again hidden.
+
+```{r echo=FALSE}
 # bind all individuals together
@@
 
-Save the speed filtered data plot.
-
 ```{r echo=FALSE}
@@
 
-Show the plot.
-
-![](figures/fig_bat_filter_speed.png)
+![Bat data with unrealistic speeds removed.
+Grey crosses show data that were removed.
+We calculated the incoming and outgoing speed of each position using `atl_get_speed`, and filtered out positions with speeds > 20 m/s using `atl_filter_covariates`, leaving 10,337 positions per individual on average (98% from the previous step).](figures/fig_bat_filter_speed.png)
 
@@
 
-Apply a 5 point median smooth to the data.
+The quality of this data (Figure 3) is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless.
+
+Since the median smoothing function `atl_median_smooth` modifies in place, we first make a copy of the data, using `data.table`'s `copy` function.
+No reassignment is required, in this case. The `lapply` function allows arguments to `atl_median_smooth` to be passed within `lapply` itself.
+
+In this case, the same moving window $K$ is applied to all individuals, but modifying this code to use the multivariate version `Map` allows different $K$ to be used for different individuals.
+This is a programming matter, and is not covered here further.
 
@@
 # split the data again
-data_smooth <- split(data_smooth, data_smooth$TAG)
+data_smooth <- split(data_smooth, by = "TAG")
 ```
-
-Remember, **`atl_median_smooth` MODIFIES IN PLACE**.
 
@@
 invisible(
+
   # the function arguments to atl_median_smooth
   # can be passed directly in lapply
-  lapply(data_smooth, atl_median_smooth,
+
+  lapply(X = data_smooth, 
+         FUN = atl_median_smooth,
          time = "time", moving_window = 5)
@@
 
-```{r}
+```{r echo=FALSE}
 # recombine split up data that has been smoothed
@@
 
-Save the smoothed data plot.
-
 ```{r echo=FALSE}
@@
 
-Show the plot.
-
-![](figures/fig_bat_smooth.png)
+![Bat data after applying a median smooth with a moving window $K$ = 5. Grey crosses show data prior to smoothing. The smoothing step did not discard any data.](figures/fig_bat_smooth.png)
 
@@
 
-First, the data is put through the `recurse` package to get residence time.
+First, the data is put through the `recurse` package to get residence time [@bracis2018].
 
@@
 
-Get residence time. Since bats may revisit the same features, we want to prevent confusion between frequent revisits and prolonged residence.
+We calculated residence time, but since bats may revisit the same features, we want to prevent confusion between frequent revisits and prolonged residence.
 
-For this, we stop summing residence times within Z metres of a location if the animal exited the area for one hour or more.
+For this, we stop summing residence times within $Z$ metres of a location if the animal exited the area for one hour or more. The value of $Z$ (radius, in `recurse` parameter terms) was chosen as 50m.
+
+This step is relatively complicated and is only required for individuals which frequently return to the same location, or pass over the same areas repeatedly, and for which revisits (cumulative time spent) may be confused for residence time in a single visit.
+
+While a simpler implementation using total residence time divided by the number of revisits is also possible, this does assume that each revisit had the same residence time.
 
@@
 
-### Sanity check: Residence-time time-series
+We bind the data together and assign a human readable timestamp column.
 
@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, eval=FALSE}
+# not shown or mentioned
 fig_bat_res_timeseries <-
@@
 
-Save the speed filtered data plot.
-
 ```{r echo=FALSE}
+# not shown or mentioned
 # save to figures
@@
 
-Show the plot.
-
-![](figures/fig_bat_res_timeseries.png)
-
 ### Constructing residence patches
 
-Split the data and construct residence patches.
-
 Some preparation is required. First, the function requires columns `x`, `y`,
-`time`, and `id`.
+`time`, and `id`, which we assign using the `data.table` syntax.
+Then we subset the data to only work with positions where the individual had a residence time of more than 5 minutes.
 
@@
 
-Now segment-cluster into residence patches.
+We apply the residence patch method, using the default argument values (`lim_spat_indep = 100` (metres), `lim_time_indep = 30` (minutes), and `min_fixes = 3`).
+We change the `buffer_radius` to 25 metres (twice the buffer radius is used, so points must be separated by 50m to be independent bouts).
 
@@
 
-We get the residence patch data as spatial `sf-MULTIPOLYGON` objects.
+We extract the residence patch data as spatial `sf-MULTIPOLYGON` objects.
+These are returned as a list and must be converted into a single `sf` object.
+These objects and the raw movement data are shown in Figure 5.
 
@@
 
-## Processed bat patches
-
-This figure made in QGIS.
-
-![](figures/fig_bat_patches.png)
+![A visual examination of plots of the bats' residence patches and linear approximations of paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site over the three nights **(a)**.
+Bats tended to be resident near fruit trees, which are their main food source, travelling repeatedly between previously visited areas **(b, c)**.
+However, bats also appeared to spend some time at locations where no fruit trees were recorded, prompting questions about their use of other food sources **(b, c)**.
+When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar **(c)**.
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_bat_patches.png)

### 2020-11-16 index.rmd
update index

--- 2020-10-08 index.rmd
@@
 knit: "bookdown::render_book"
-title: "A Rough Guide to Pre-processing High-Frequency Animal Tracking Data"
-author: "Pratik R. Gupte and others"
+title: "Supplementary Material for: A Rough Guide to Pre-processing High-Frequency Animal Tracking Data"
+author:
+    - Pratik R. Gupte
+    - Christine E. Beardsworth
+    - Orr Spiegel
+    - Emmanuel Lourie
+    - Allert Bijleveld
 date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-output: bookdown::pdf_book
-documentclass: book
+output: bookdown::pdf_document2
+documentclass: article
+toc-depth: 2
+secnumdepth: 2
+geometry: "left=4cm, right=3cm, top=2.5cm, bottom=2.5cm"
 bibliography: [references.bib]
-biblio-style: apalike
 link-citations: yes
 github-repo: pratikunterwegs/atlas-best-practices
-description: "Source code for a guide to pre-processing high-frequency tracking data."
+description: "Supplementary Material for: A Rough Guide to Pre-processing High-Frequency Animal Tracking Data"
 ---
-
-# Abstract
-
-Data cleaning is a ubiquitous pre-processing step in analysis pipelines, and its automation is essential for large data volumes such as those generated in animal tracking studies using high-frequency Time-of-Arrival (TOA) systems.
-Users of systems such as ATLAS must contend with two intertwined data cleaning challenges: (1) reducing positioning errors, and (2) the high volume of data itself.
-Making biological inferences from data with positioning errors is not straightforward, and processing large data volumes is computationally intensive.
-Though reducing positioning error is widely recommended, users are without uniform guidance on how to go about this, and lack a common set of computationally efficient tools.
-Further, many methods that synthesize movement tracks for ecological inference are either (1) not suited to very large datasets, or (2) not intuitive to understand in terms of the tracked animal's biology.
-
-In this article we introduce a pipeline to pre-process high-frequency animal tracking data in order to prepare it for subsequent analysis.
-We demonstrate this pipeline on simulated movement data to which we have randomly added positioning errors.
-This pipeline is suited to any tracking study in which the high data volume combined with knowledge of the tracked individuals’ biology can be used to smooth out positioning errors. 
-We further suggest how large volumes of cleaned data may be synthesized into biologically meaningful ‘residence patches’, and demonstrate how this accurately captures animal space-use.
-Finally, we introduce the `R` package `atlastools` which provides fast implementations of the methods we describe. 
-Though aimed at ATLAS systems, `atlastools` can be used with any time-series animal movement data, and we demonstrate its usage with both simulated and empirical examples.
-
-![](figures/fig_recipe.png)

### 2020-11-16 render_books.sh
no website

@@ -1,5 +1,5 @@
 #!/bin/bash
-Rscript --slave -e 'bookdown::render_book("index.rmd")'
+
 Rscript --slave -e 'bookdown::render_book("index.rmd", "rmarkdown::pdf_document"); warnings()'
 
 # to render source files

### 2020-11-16 _03_smoothing_tracks.Rmd

--- 2020-10-29 03_smoothing_tracks.Rmd
@@
 
-# Reducing Small-Scale Positioning Error and Aggregating Data
-
-Once large-scale positioning errors have been removed using filters on bounds or on unrealistic movement, small-scale positioning errors may remain. Why is this the case?
-
-The simple reason is that high-frequency tracking has a large component of error relative to the real movement of an individual, and the error / movement ratio increases with the frequency of tracking.
-[@noonan2019] has a better explanation, which users are strongly recommended to read.
-
-The effect of these small-scale errors for high-frequency data is to cause overestimations of straight-line displacement (distance between two positions) and speed.
-If users were to apply a very strict speed filter, many of these positions would be lost.
-This loss of information would certainly reduce the sampling frequency, which would lead to better estimates of total distance and mean speed.
-However, while _total_ distance and _mean_ speed estimates would improve, the instantaneous distance and speed estimates would remain biased, which will most likely lead to erroneous inferences from the data.
-
-However, users can turn the high-frequency of their data against itself by applying a simple smooth. The median smooth is recommended because it is robust to outliers, and is very well understood.
+# Reducing Small-Scale Error and Aggregating Data
 
@@
     ggplot(df)+
+      geom_path(data = data,
+                aes(x, y),
+                col = "grey20",
+                size = 0.2)+
       geom_point(data = data_agg,
@@
       geom_point(aes(x,y,
-                     size = SD,
-                     fill = SD,
                      group = interval),
-                 shape = 21,
-                 alpha = 0.8,
+                 shape = 19,
+                 col = pal[4],
+                 size = 2,
+                 alpha = 0.6,
                  show.legend = F)+
-      scale_fill_distiller(palette = "Purples", direction = 1,
-                           values = c(-0.25, 1))+
-      scale_size(range = c(2, 4))+
       ggthemes::theme_few()+

### 2020-11-16 _05_calibration_data.Rmd

--- 2020-11-12 05_calibration_data.Rmd
@@
   ggplot()+
-  geom_errorbar(data = data_thin,
-                aes(x, y,
-                    ymin = y - VARY,
-                    ymax = y + VARY),
-                col = pal[4],
-                alpha = 0.2,
-                lwd = 0.1)+
-  geom_errorbarh(data = data_thin,
-                 aes(x, y,
-                     xmin = x - VARX,
-                     xmax = x + VARX),
-                 col = pal[4],
-                 alpha = 0.2,
-                 lwd = 0.1)+
   geom_point(data = data_raw[!data_unproc, on = c("x", "y")],
@@
              size = 0.4)+
+  geom_errorbar(data = data_thin,
+                aes(x, y,
+                    ymin = y - SD,
+                    ymax = y + SD),
+                col = pal[4],
+                alpha = 0.5)+
+  geom_errorbarh(data = data_thin,
+                 aes(x, y,
+                     xmin = x - SD,
+                     xmax = x + SD),
+                 col = pal[4],
+                 alpha = 0.5)+
   geom_point(data = data_unproc,
              aes(x, y),
+             size = 0.4,
              col = pal[3],
-             shape = 19,
-             alpha = 0.5) +
-  geom_point(data = data_thin,
-             aes(x, y),
-             col = pal[4],
-             shape = 1)+
+             shape = 16) +
+  # geom_point(data = data_thin,
+  #            aes(x, y),
+  #            col = pal[4],
+  #            alpha = 0.5,
+  #            shape = 0, show.legend = F)+
   geom_path(data = data_thin,

### 2020-11-20 _01_background.Rmd

@@
 ```{r}
+# set a seed
+set.seed(0)
+
 # do smoove data using a RACVM for three different rotational speed
@@
  dt[, patch := patch]
+ dt[, time := seq(.N)]
 }, c(7, 7, 7), sprintf('patch_%i', seq(3)))
@@
 # rescale x and y
-data_thermals <- Map(function(df, lim1, lim2, t1, t2) {
+data_thermals <- Map(function(df, lim1, lim2, offset) {
   df[, x := x + lim1]
   df[, y := y + lim2]
-  df[, time := seq(t1, t2)]
+  df[, time := time + offset]
 }, data_thermals, c(1, 5, 9), c(0, 2.5, 5),
-   c(100, 800, 1500), c(600, 1300, 2000))
+   c(100, 800, 1500))
 
@@
 setorder(gaps, "time")
+gaps[, event := rep(c("start", "end"), 3)]
 
-# get transition phases
-# THIS IS HARDCODED BECAUSE THERE DOESN'T APPEAR TO BE A BETTER WAY
-# FOR THE SAME TIME
-transit_1 <- gaps[c(2, 3), ]
-transit_2 <- gaps[c(4, 5), ]
+# cast
+gaps <- dcast(gaps, patch ~ event, value.var = c("x", "y", "time"))
 
-# simulate movement
-transits <- Map(function(df){
-  time_lim <- as.vector(df$time)
+# make a trajectory section
+transit <- data.table(
+  x1 = shift(gaps$x_end),
+  x2 = gaps$x_start,
+  y1 = shift(gaps$y_end),
+  y2 = gaps$y_start,
+  t1 = shift(gaps$time_end),
+  t2 = gaps$time_start,
+  transit = sprintf("transit_%i", seq(nrow(gaps))),
+  pseudo_time = seq(nrow(gaps))
+)
 
-  track <- as.data.table(approx(df$x, df$y, n = 10))
+# nafill
+transit[, c("x1", "y1", "t1") := 
+          lapply(list(x1, y1, t1), nafill, fill = 0)]
 
-  # add some error
-  track[, `:=`(y = do_add_error(y, 0.01))]
+# interpolate
+transit[, d := Map(function(x1, x2, y1, y2) {
+  as.data.table(approx(x = c(x1, x2), 
+                       y = c(y1, y2), 
+                       n = 20))
+  }, x1, x2, y1, y2)]
 
-  track[, `:=`(time = seq(time_lim[1], time_lim[2], length.out = 10),
-               patch = "transit")]
+# add errors
+transit[, d := Map(function(df) {
+  df[, lapply(.SD, do_add_error, std_dev = 0.05)]
+}, d)]
 
-  return(track)
+# get transit data
+transit_data <- Map(function(df, t1, t2) {
+  df <- copy(df)
+  df[, time := seq(t1, t2, length.out = nrow(df))]
+}, transit$d, transit$t1, transit$t2)
 
-}, list(transit_1, transit_2))
-
-# bind
-transits <- rbindlist(transits)
+# merge with thermals
+transit_data <- rbindlist(transit_data)
 ```
@@
 # bind transits to thermals
-data_move <- merge(data_thermals, transits, on = "time", all = T)
+data_move <- merge(data_thermals, transit_data, 
+                   on = "time", all = T)
 setorder(data_move, time)

### 2020-11-20 _02_filtering_data.Rmd

--- 2020-11-11 02_filtering_data.Rmd
@@
 
-```{r remove_outside_bbox, eval=FALSE}
+```{r remove_outside_bbox}
 # remove positions outside a bounding box
@@
 
-```{r example_remove_outliers, eval=FALSE}
+```{r example_remove_outliers}
 # get speed and turning angle

### 2020-11-21 _02_filtering_data.Rmd
add rmd files

@@
 
-ggsave(plot_figure, filename = "figures/fig_raw_bounds.png", 
+ggsave(plot_figure, filename = "figures/fig_02_bounds.png", 
        width = 170, height = 170, units = "mm")
@@
                           point_angle_cutoff = 10,
-                          reflection_speed_cutoff = 0.024)
+                          reflection_speed_cutoff = 0.025)
 # get reflections
@@
              alpha = 0.5,
-             col = "grey40",
-             fill = "grey",
-             shape = 21)+
+             col = "grey50",
+             shape = 16)+
   geom_point(data = data_no_reflection,
@@
 
-ggsave(plot_figure, filename = "figures/fig_correct_tracks.png", 
+ggsave(plot_figure, filename = "figures/fig_03_filter_speed.png", 
        width = 170, height = 170, units = "mm")
@@
 
-![](figures/fig_correct_tracks.png)
+![](figures/fig_03_filter_speed.png)
 

### 2020-11-21 _03_smoothing_tracks.Rmd
add rmd files

@@
 
+![](figures/fig_04_median_smooth.png)
+
 ```{r}
@@
 # save figure
-ggsave(figure_median_smooth, filename = "figures/fig_median_smooth.png",
+ggsave(figure_median_smooth, filename = "figures/fig_04_median_smooth.png",
        width = 170, height = 170 / 3, units = "mm")
@@
 ggsave(fig_aggregate,
-       filename = "figures/fig_aggregate_errors.png",
+       filename = "figures/fig_05_thinning.png",
        width = 170, height = 85, units = "mm")
@@
 
-![](figures/fig_aggregate_errors.png)
+![](figures/fig_05_thinning.png)
 

### 2020-11-21 _05_calibration_data.Rmd
add rmd files

--- 2020-11-12 05_calibration_data.Rmd
@@
            ymin = 5904450, ymax = 5906133,
-           fill = NA, col = "black")+
+           fill = NA, col = "grey20")+
+  annotate(geom = "rect",
+           ymin = 5901650, ymax = 5903100,
+           xmin = 650000, xmax = 650600,
+           fill = NA, col = "grey20"
+           )+
   scale_color_manual(values = c("grey", pal[3]))+
@@
   # smoothed data
-  geom_path(data = data_unproc,
+  geom_path(data = data,
              aes(x, y),
              col = pal[3]) +
-  geom_point(data = data_unproc,
+  geom_point(data = data,
              aes(x, y),
@@
              size = 0.4)+
+
+  geom_point(data = data_thin,
+             aes(x, y, size = count),
+             col = pal[4],
+             alpha = 0.8,
+             shape = 0, show.legend = F)+
   geom_point(data = data_unproc,
              aes(x, y),
+             alpha = 0.6,
              col = pal[3],
-             shape = 19,
-             alpha = 0.5) +
-  geom_point(data = data_thin,
-             aes(x, y),
-             col = pal[4],
-             shape = 1)+
-  geom_path(data = data_thin,
-            aes(x, y),
-            col = pal[4], size = 0.2)+
+             shape = 16) +
+  scale_size(range = c(2, 6))+
   coord_cartesian(
@@
              size = 10, shape = 7,
-             col = "grey20",
+             stroke = 1,
+             col = pal[4],
              alpha = 1)+
+  geom_path(data = data_recurse,
+            aes(x, y),
+            show.legend = F,
+            size = 0.3,
+            col = "grey")+
   geom_point(data = data_recurse,
-            aes(x, y,
-                col = res_time > 5,
-                shape = res_time > 5),
+            aes(x, y),
             show.legend = F,
-            size = 2)+
+            shape = 1, col = "grey50")+
   geom_sf(data = patch_sf_data,
-          fill = pal[4],
-          col = pal[4],
+          fill = pal[3],
+          col = pal[3],
           alpha = 0.3)+
@@
            expand = F,
-           ylim = c(5901700, 5903150),
-           xlim = c(649300, 650700))
+           ylim = c(5901650, 5903100),
+           xlim = c(650000, 650600))
 ```
@@
 patch_summary_compare <- patch_summary_compare[tID != "WP080", ]
+
+# compare real and inferred duration
+patch_summary_compare[, duration_diff := (duration - duration_real) / duration_real]
+
+# get mean and sd of duration difference
+patch_summary_compare[, .(mean_diff = mean(duration_diff),
+                          sd_diff = sd(duration_diff))]
 ```
@@
 summary(model_duration)
+
+# write to file
+writeLines(
+  text = capture.output(
+    summary(model_duration)
+  ),
+  con = "data/model_output_residence_patch.txt"
+)
 ```
@@
              col = pal[1],
+             lwd = 0.5,
              method = "lm",
@@
            size = 5,
-           label = "R^2 == 0.98",
-           parse = T)+
+           label = "R^2 == 0.95",
+           parse = T, family = "Arial")+
   scale_x_continuous(
@@
   )+
-  coord_cartesian(expand = T,
+  coord_fixed(expand = T,
+              ratio = 1.115
   #             # xlim = c(400, 1050),
@@
 
-### Join patch map and duration comparison figure
+## Make Figure 07
 
@@
 figure_res_patch <-
-  wrap_plots(list(fig_basic_residence, figure_patch_duration),
-             design = "AA\nAA\nB#")+
+  wrap_plots(list(fig_speed_outliers,
+                  fig_smooth_thin,
+                  fig_basic_residence),
+             design = "AC\nAC\nBC\nBC"
+             )+
   plot_annotation(tag_levels = "a",
@@
 ggsave(figure_res_patch,
-       filename = "figures/fig_calib_residence_patch.png",
-       height = 225, width = 150, units = "mm")
+       filename = "figures/fig_07_calib_residence_patch.png",
+       height = 170, width = 170, units = "mm")
 ```

### 2020-11-22 _05_calibration_data.Rmd
clean fig 07

@@
 
-# save combined figure
-ggsave(figure_walkthrough,
-       filename = "figures/fig_walkthrough.png",
-       height = 170, width = 90, units = "mm")
+# # save combined figure
+# ggsave(figure_walkthrough,
+#        filename = "figures/fig_walkthrough.png",
+#        height = 170, width = 90, units = "mm")
 ```
@@
 
-# subset on 5 minute residence time
-data_recurse_subset <- data_recurse[res_time >= 5, ]
-
 # on known residence points
-patch_res_known <- atl_res_patch(data_recurse_subset, 
-                                buffer_radius = 10,
-                                lim_spat_indep = 100,
+patch_res_known <- atl_res_patch(data_recurse, 
+                                buffer_radius = 5,
+                                lim_spat_indep = 50,
                                 lim_time_indep = 5,
@@
 # patch with residence points and all patches
-fig_basic_residence <-
+# fig_basic_residence <-
   ggplot()+
@@
             shape = 1, col = "grey50")+
-  geom_sf(data = patch_sf_data,
+  geom_sf(data = patch_sf_data[patch_sf_data$duration >= 5 * 60, ],
           fill = pal[3],

### 2020-11-22 _05_calibration_data.Rmd
remove recurse fix figure

@@
 
-### Calculate residence time
-
-Subset for boating segments: 2020-08-24 11:13:00 -- 2020-08-24 12:36:11.
-
-```{r}
-# load recurse
-library(recurse)
-```
-
 ## Residence patches
@@
 # get 4 column data
-data_recurse <- data_thin[, list(x, y, time, TAG)]
-
-# get recurse data for a 10m radius
-recurse_stats <- getRecursions(data_recurse,
-                               radius = 50, timeunits = "mins")
-
-# assign to recurse data
-data_recurse[, res_time := recurse_stats$residenceTime]
+data_for_patch <- data_thin[, list(x, y, time, TAG)]
 
 # save recurse data
-fwrite(data_recurse, file = "data/data_calib_recurse.csv")
-```
-
-Read in the data.
-
-```{r}
-data_recurse <- fread("data/data_calib_recurse.csv")
+fwrite(data_for_patch, file = "data/data_calib_for_patch.csv")
 ```
@@
 # assign id as tag
-data_recurse[, id := as.character(TAG)]
+data_for_patch[, id := as.character(TAG)]
 
 # on known residence points
-patch_res_known <- atl_res_patch(data_recurse, 
+patch_res_known <- atl_res_patch(data_for_patch, 
                                 buffer_radius = 5,
@@
 # patch with residence points and all patches
-# fig_basic_residence <-
+fig_basic_residence <-
   ggplot()+
@@
              alpha = 1)+
-  geom_path(data = data_recurse,
+  geom_path(data = data_for_patch,
             aes(x, y),
@@
             col = "grey")+
-  geom_point(data = data_recurse,
+  geom_point(data = data_for_patch,
             aes(x, y),
@@
 patch_summary_compare <- patch_summary_compare[tID != "WP080", ]
-
-# compare real and inferred duration
-patch_summary_compare[, duration_diff := (duration - duration_real) / duration_real]
-
-# get mean and sd of duration difference
-patch_summary_compare[, .(mean_diff = mean(duration_diff),
-                          sd_diff = sd(duration_diff))]
 ```

### 2020-11-22 _05_calibration_data.Rmd
re add res time

@@
 
-In this dataset, residence positions are marked in the `tID` column as beginning with `WP`. These can be extracted and converted into residence patches.
+### Calculate residence time
+
+```{r}
+# load recurse
+library(recurse)
+```
 
@@
 data_for_patch <- data_thin[, list(x, y, time, TAG)]
+
+# get recurse data for a 10m radius
+recurse_stats <- getRecursions(data_for_patch,
+                               radius = 50, timeunits = "mins")
+
+# assign to recurse data
+data_for_patch[, res_time := recurse_stats$residenceTime]
 
@@
 
+We subset data with a residence time > 5 minutes.
+
 ```{r}
@@
 # on known residence points
-patch_res_known <- atl_res_patch(data_for_patch, 
+patch_res_known <- atl_res_patch(data_for_patch[res_time >= 5, ], 
                                 buffer_radius = 5,

### 2020-11-22 _02_filtering_data.Rmd

@@
 
-```{r add_outlier}
+```{r add_outlier, eval=FALSE}
 # make a copy
@@
              aes(x, y),
-             size = 0.2,
+             size = 0.5,
              shape = 2, col = pal[4])+
@@
              show.legend = F,
-             alpha = 0.5)+
+             alpha = 1,
+             size = 0.5)+
   geom_path(data = data,
@@
              alpha = 0.5,
-             col = "grey50",
+             col = "grey",
+             size = 0.5,
              shape = 16)+
@@
              aes(x, y),
-             alpha = 0.5,
+             # alpha = 0.5,
+             size = 0.5,
              colour = pal[3],

### 2020-11-22 05_calibration_data.Rmd
make calibration supplement

--- 2020-11-12 05_calibration_data.Rmd
@@
 
+Here we show how the residence patch method [@barraquand2008; @bijleveld2016; @oudman2019] accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
+
 ## Prepare libraries
+
+First we prepare the libraries we need. Libraries can be installed from CRAN if necessary.
 
@@
 
-## Preliminary visualisation
+## Access data and preliminary visualisation
+
+First we access the data from a local file using the `data.table` package [@dowle2020].
+We then visualise the raw data.
 
@@
   geom_path(aes(x, y),
-             col = "grey", alpha = 0.5, size = 0.3)+
+             col = "grey", alpha = 1, size = 0.2)+
   geom_point(aes(x, y),
-             col = pal[3], alpha = 0.2, size = 2)+
-  ggthemes::theme_map()+
+             col = "grey", alpha = 0.2, size = 0.2)+
+  ggthemes::theme_few()+
+  theme(axis.title = element_blank(),
+        axis.text = element_blank())+
   coord_sf(crs = 32631)
@@
 
+![](figures/fig_calibration_raw.png)
+
 ## Filter by bounding box
 
-Save an unprocessed copy.
+We first save a copy of the data, so that we can plot the raw data with the cleaned data plotted over it for comparison.
 
 ```{r}
+# make a copy using the data.table copy function
 data_unproc <- copy(data)
@@
 
-Filter by a bounding box.
+We then filter by a bounding box in order to remove the point outlier to the far south east of the main track.
+We use the `atl_filter_bounds` functions using the `x_range` argument, to which we pass the limit in the UTM 31N coordinate reference system.
+This limit is used to exclude all points with an X coordinate < 645,000.
+
+We then plot the result of filtering, with the excluded point in black, and the points that are retained in green.
 
@@
                  col = x > 645000),
-             alpha = ifelse(data_raw$x > 645000, 0.5, 1),
+             alpha = ifelse(data_raw$x > 645000, 1, 1),
              size = ifelse(data_raw$x > 645000, 0.3, 2),
              show.legend = F)+
-  scale_colour_manual(values = c("grey", pal[3]))+
+  scale_colour_manual(values = c("black", pal[3]))+
   geom_vline(xintercept = 645000,
@@
 
-### Plot to see speeds
-
 ```{r echo=FALSE}
@@
 
-![](figures/fig_speed_outlier.png)
-
 ### Filter on speed
@@
 Here we use a speed threshold of 15 m/s, the fastest known boat speed.
+We then plot the data with the extreme speeds shown in grey, and the positions retained shown in green.
 
@@
 
+![](figures/fig_speed_outlier.png)
+
 ## Smoothing the trajectory
+
+We then apply a median smooth over a moving window ($K$ = 5).
+This function modifies in place, and does not need to be assigned to a new variable.
+We create a copy of the data before applying the smooth so that we can compare the data before and after smoothin.
 
@@
 
-### Plot smoothed data
-
 ```{r}
@@
              size = 0.4)+
-  # smoothed data
-  geom_path(data = data_unproc,
-             aes(x, y),
-             col = pal[3]) +
-  geom_point(data = data_unproc,
+  geom_point(data = data,
              aes(x, y),
@@
 
+![](figures/fig_calib_median_smooth.png)
+
 ## Thinning the data
+
+Next we thin the data to demonstrate thinning by median smoothing.
+Following this, we plot the median smooth and thinning by aggregation.
 
@@
 
-### Plot thinned data
-
 ```{r echo=FALSE}
@@
 
-### Plot pre-processing steps
-
 ```{r echo=FALSE}
@@
     fig_speed_outliers, fig_smooth_thin),
-  design = "A\nB")+
+  design = "AB")+
   plot_annotation(tag_levels = "a",
@@
 
-# save combined figure
+# # save combined figure
 ggsave(figure_walkthrough,
        filename = "figures/fig_walkthrough.png",
-       height = 170, width = 90, units = "mm")
+       height = 90, width = 170, units = "mm")
 ```
 
-![](figures/fig_walkthrough.png)
+![](figures/fig_calib_smooth_thin.png)
+
+## Residence patches
+
+### Get waypoint centroids
+
+We subset the annotated calibration data to select the waypoints and the positions around them which are supposed to be the locations of known stops.
+Since each stop was supposed to be 5 minutes long, there are multiple points in each known stop.
+
+```{r}
+library(stringi)
+data_res <- data_unproc[stri_detect(tID, regex = "(WP)")]
+```
+
+From this data, we get the centroid of known stops, and determine the time difference between the first and last point within 50 metres, and within 10 minutes of the waypoint positions' median time.
+
+Essentially, this means that the maximum duration of a stop can be 20 minutes, and stops above this duration are not expected.
+
+```{r}
+# get centroid
+data_res_summary <- data_res[, list(x_median = round(median(x), digits = -2),
+                                    y_median = round(median(y), digits = -2)),
+                                    t_median = median(time)),
+                             by = "tID"]
+
+# now get times 10 mins before and after
+data_res_summary[, `:=`(t_min = t_median - (10 * 60),
+                        t_max = t_median + (10 * 60))]
+
+# make a list of positions 10min before and after
+wp_data <- mapply(function(l, u, mx, my) {
+  tmp_data <- data_unproc[inrange(time, l, u)]
+  tmp_data[, distance := sqrt((mx - x)^2 + (my - y)^2)]
+
+  # keep within 50
+  tmp_data <- tmp_data[distance <= 50, ]
+
+  # get duration
+  return(diff(range(tmp_data$time)))
+
+}, data_res_summary$t_min, data_res_summary$t_max,
+   data_res_summary$x_median, data_res_summary$y_median,
+SIMPLIFY = TRUE)
+```
+
+### Prepare data
+
+An indicator of individual residence at or near a position can be useful when attempting to identify residence patches. Positions can be filtered on a metric such as residence time [@bracis2018].
 
@@
 
-Subset for boating segments: 2020-08-24 11:13:00 -- 2020-08-24 12:36:11.
+First we calculate the residence time with a radius of 50 metres.
+For this, we need a dataframe with coordinates, the timestamp, and the animal id.
+We save this data to file for later use.
 
@@
 
-## Residence patches
-
-### Prepare data
-
-An indicator of individual residence at or near a position can be useful when attempting to identify residence patches.
-Positions can be filtered on a metric such as residence time (see Bracis et al.
-2018).
-
-In this dataset, residence positions are marked in the `tID` column as beginning with `WP`. These can be extracted and converted into residence patches.
-
 ```{r}
 # get 4 column data
-data_recurse <- data_thin[, list(x, y, time, TAG)]
+data_for_patch <- data_thin[, list(x, y, time, TAG)]
 
 # get recurse data for a 10m radius
-recurse_stats <- getRecursions(data_recurse,
+recurse_stats <- getRecursions(data_for_patch,
                                radius = 50, timeunits = "mins")
@@
 # assign to recurse data
-data_recurse[, res_time := recurse_stats$residenceTime]
+data_for_patch[, res_time := recurse_stats$residenceTime]
 
 # save recurse data
-fwrite(data_recurse, file = "data/data_calib_recurse.csv")
-```
-
-Read in the data.
-
-```{r}
-data_recurse <- fread("data/data_calib_recurse.csv")
-```
-
-Subset waypoint data.
-
-```{r}
-library(stringi)
-data_res <- data_unproc[stri_detect(tID, regex = "(WP)")]
+fwrite(data_for_patch, file = "data/data_calib_for_patch.csv")
 ```
@@
 
+We subset data with a residence time > 5 minutes in order to construct residence patches.
+From this subset, we construct residence patches using the parameters: `buffer_radius` = 5 metres, `lim_spat_indep` = 50 metres, `lim_time_indep` = 5 minutes, and `min_fixes` = 3.
+
 ```{r}
 # assign id as tag
-data_recurse[, id := as.character(TAG)]
-
-# subset on 5 minute residence time
-data_recurse_subset <- data_recurse[res_time >= 5, ]
+data_for_patch[, id := as.character(TAG)]
 
 # on known residence points
-patch_res_known <- atl_res_patch(data_recurse_subset, 
-                                buffer_radius = 10,
-                                lim_spat_indep = 100,
+patch_res_known <- atl_res_patch(data_for_patch[res_time >= 5, ], 
+                                buffer_radius = 5,
+                                lim_spat_indep = 50,
                                 lim_time_indep = 5,
@@
 ### Get spatial and summary objects
+
+We get spatial and summary ouput of the residence patch method using the `atl_patch_summary` function using the options `which_data` = "spatial" and `which_data` = "summary.
+We use a buffer radius here of 20 metres for the spatial buffer, despite using a buffer radius of 5 metres earlier, simply because it is easier to visualise in the output figure.
 
@@
 
-### Get waypoint centroids
+### Prepare to plot data
 
-```{r}
-# get centroid
-data_res_summary <- data_res[, list(x_median = median(x),
-                                    y_median = median(y),
-                                    t_median = median(time)),
-                             by = "tID"]
-
-# now get times 10 mins before and after
-data_res_summary[, `:=`(t_min = t_median - (10 * 60),
-                        t_max = t_median + (10 * 60))]
-
-# make a list of positions 10min before and after
-wp_data <- mapply(function(l, u, mx, my) {
-  tmp_data <- data_unproc[inrange(time, l, u)]
-  tmp_data[, distance := sqrt((mx - x)^2 + (my - y)^2)]
-  # keep within 50
-  tmp_data <- tmp_data[distance <= 50, ]
-  # get duration
-  return(diff(range(tmp_data$time)))
-}, data_res_summary$t_min, data_res_summary$t_max,
-   data_res_summary$x_median, data_res_summary$y_median,
-SIMPLIFY = FALSE)
-```
-
-### Read Griend
+We read in the island's shapefile to plot it as a background for the residence patch figure.
 
@@
 
-We prepare a plot of the residence patches.
-
 ```{r echo=FALSE}
@@
              size = 10, shape = 7,
-             col = "grey20",
+             stroke = 1,
+             col = pal[4],
              alpha = 1)+
-  geom_point(data = data_recurse,
-            aes(x, y,
-                col = res_time > 5,
-                shape = res_time > 5),
+  geom_path(data = data_for_patch,
+            aes(x, y),
             show.legend = F,
-            size = 2)+
-  geom_sf(data = patch_sf_data,
-          fill = pal[4],
-          col = pal[4],
+            size = 0.3,
+            col = "grey")+
+  geom_point(data = data_for_patch,
+            aes(x, y),
+            show.legend = F,
+            shape = 1, col = "grey50")+
+  geom_sf(data = patch_sf_data[patch_sf_data$duration >= 5 * 60, ],
+          fill = pal[3],
+          col = pal[3],
           alpha = 0.3)+
@@
 
-### Compare known patches
+We then merge the annoated, known stop data with the calculated patch duration.
+We filter this data to exclude one exceedingly long outlier of about an hour (WP080), which how
 
@@
 # add real duration
-patch_summary_real[, duration_real := unlist(wp_data)]
+patch_summary_real[, duration_real := wp_data]
 
@@
 
-patch_summary_compare[!is.na(tID)]
-
 # drop nas
@@
 
-12 patches are identified where there are no waypoints, while 2 waypoints are not identified as patches.
+7 patches are identified where there are no waypoints, while 2 waypoints are not identified as patches.
 These waypoints consisted of 6 and 15 (WP098 and WP092) positions respectively, and were lost when the data were aggregated to 30 second intervals.
 
-### Plot durations comparisons
+### Linear model durations
+
+We run a simple linear model.
 
@@
 # make figure comparing different methods
-figure_patch_duration <-
+figure_lm_duration <-
   ggplot()+
@@
              se = F,
-             col = pal[4],
+             col = pal[1],
+             lwd = 0.5,
              method = "lm",
@@
            size = 5,
-           label = "R^2 == 0.98",
-           parse = T)+
+           label = "R^2 == 0.908",
+           parse = T, family = "Arial")+
   scale_x_continuous(
@@
        y = "real duration (min)")
+
+ggsave(figure_lm_duration, filename = "figures/fig_calib_lm_duration.png",
+       width = 80 / 25, height = 80 / 25)
 ```
 
-### Join patch map and duration comparison figure
+![](figures/fig_calib_lm_duration.png)
 
@@
 ggsave(figure_res_patch,
-       filename = "figures/fig_calib_residence_patch.png",
-       height = 225, width = 150, units = "mm")
+       filename = "figures/fig_07_calib_residence_patch.png",
+       height = 170, width = 170, units = "mm")
 ```
-
-![](figures/fig_calib_residence_patch.png)

### 2020-11-22 05_calibration_data.Rmd
no plotting

@@
 
-Here we show how the residence patch method [@barraquand2008; @bijleveld2016; @oudman2019] accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
+Here we show how the residence patch method [@barraquand2008; @bijleveld2016; @oudman2018] accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
 
@@
 
-```{r}
+```{r, echo=FALSE}
 # make zoomed in figures

### 2020-11-22 render_books.sh
book rendering corrections

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-Rscript --slave -e 'bookdown::render_book("index.rmd", "rmarkdown::pdf_document"); warnings()'
+Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2"); warnings()'
 
 # to render source files
-Rscript --slave -a 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
+Rscript --slave -e 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'

### 2020-11-22 R/05_calibration_data.R
add raw r scripts

@@ -1,0 +1,493 @@
+## -----------------------------------------------------------------------------
+# load libs
+library(data.table)
+library(atlastools)
+library(ggplot2)
+library(patchwork)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+
+
+## -----------------------------------------------------------------------------
+# read and plot example data
+data <- fread("data/atlas1060_allTrials_annotated.csv")
+data_raw <- copy(data)
+
+
+## ----echo=FALSE---------------------------------------------------------------
+# plot data
+fig_data_raw <-
+  ggplot(data)+
+  geom_path(aes(x, y),
+             col = "grey", alpha = 1, size = 0.2)+
+  geom_point(aes(x, y),
+             col = "grey", alpha = 0.2, size = 0.2)+
+  ggthemes::theme_few()+
+  theme(axis.title = element_blank(),
+        axis.text = element_blank())+
+  coord_sf(crs = 32631)
+
+# save figure
+ggsave(fig_data_raw, filename = "figures/fig_calibration_raw.png",
+       width = 185 / 25)
+
+
+## -----------------------------------------------------------------------------
+# make a copy using the data.table copy function
+data_unproc <- copy(data)
+
+
+## -----------------------------------------------------------------------------
+# remove inside must be set to falses
+data <- atl_filter_bounds(data = data, 
+                          x = "x", y = "y", 
+                          x_range = c(645000, max(data$x)), 
+                          remove_inside = FALSE)
+
+
+## ----echo=FALSE---------------------------------------------------------------
+# plot data
+fig_data_bbox <-
+  ggplot()+
+  geom_path(data = data_raw,
+            aes(x, y),
+             col = "grey", alpha = 0.5, size = 0.3)+
+  geom_point(data = data_raw,
+             aes(x, y,
+                 col = x > 645000),
+             alpha = ifelse(data_raw$x > 645000, 1, 1),
+             size = ifelse(data_raw$x > 645000, 0.3, 2),
+             show.legend = F)+
+  scale_colour_manual(values = c("black", pal[3]))+
+  geom_vline(xintercept = 645000,
+             col = "grey", lty = 2)+
+  ggspatial::annotation_scale(location = "br")+
+  ggthemes::theme_few()+
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())+
+  coord_sf(crs = 32631)
+
+# save result
[…]

### 2020-11-22 R/_02_filtering_data.R

@@ -1,4 +1,4 @@
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 # to handle movement data
 library(data.table)
 library(atlastools)
@@ -8,46 +8,46 @@ library(ggplot2)
 library(patchwork)
 
 
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 # source helper functions
 source("R/helper_functions.R")
 
 
-## ----read_sim_data----------------------------------------------------
+## ----read_sim_data------------------------------------------------------------
 # read in the data
 data <- fread("data/data_sim.csv")[5000:10000, ]
 
 
-## ----add_outlier------------------------------------------------------
-# make a copy
-data_copy <- copy(data)
+## ----add_outlier, eval=FALSE--------------------------------------------------
+## # make a copy
+## data_copy <- copy(data)
+## 
+## # add a prolonged spike or reflection to 300 positions
+## data_copy[500:800, `:=`(x = x + 0.25,
+##                         y = y + 0.25)]
+## 
+## # add normal error
+## data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
+##                  y = do_add_error(y, std_dev = 0.005))]
+## 
+## # add 100 outliers
+## data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
 
-# add a prolonged spike or reflection to 300 positions
-data_copy[500:800, `:=`(x = x + 0.25,
-                        y = y + 0.25)]
 
-# add normal error
-data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
-                 y = do_add_error(y, std_dev = 0.005))]
-
-# add 100 outliers
-data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
-
-
-## ----eval=FALSE-------------------------------------------------------
+## ----eval=FALSE---------------------------------------------------------------
 ## fwrite(data_copy, file = "data/data_errors.csv")
 
 
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 data_copy <- fread("data/data_errors.csv")
 
 
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 # define a four colour palette
 pal <- RColorBrewer::brewer.pal(4, "Set1")
 
 
-## ----echo=FALSE-------------------------------------------------------
[…]

### 2020-11-22 R/_03_smoothing_tracks.R

@@ -1,4 +1,4 @@
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 # prep libs
 library(data.table)
 library(atlastools)
@@ -9,7 +9,7 @@ library(patchwork)
 pal <- RColorBrewer::brewer.pal(4, "Set1")
 
 
-## ----read_sim_data_2--------------------------------------------------
+## ----read_sim_data_2----------------------------------------------------------
 # read in the data and set the window size variable
 data <- fread("data/data_sim.csv")[5000:10000, ]
 data[, window_size := NA]
@@ -19,7 +19,7 @@ data_errors <- fread("data/data_no_reflection.csv")
 data_errors[, window_size := 0]
 
 
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 # smooth the data over four K values
 list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
   
@@ -35,11 +35,11 @@ list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
 })
 
 
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 fwrite(list_of_smooths[[3]], file = "data/data_smooth.csv")
 
 
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 # bind list after offset
 data_plot <- mapply(function(df, offset) {
   df[, x := x + offset]
@@ -49,7 +49,7 @@ SIMPLIFY = F)
 data_plot <- rbindlist(data_plot)
 
 
-## ----echo=FALSE-------------------------------------------------------
+## ----echo=FALSE---------------------------------------------------------------
 # prepare data to plot
 # make list of data to plot
 figure_median_smooth <-
@@ -84,13 +84,13 @@ figure_median_smooth <-
         axis.title = element_blank())
 
 # save figure
-ggsave(figure_median_smooth, filename = "figures/fig_median_smooth.png",
+ggsave(figure_median_smooth, filename = "figures/fig_04_median_smooth.png",
        width = 170, height = 170 / 3, units = "mm")
 
 
-## ---------------------------------------------------------------------
+## -----------------------------------------------------------------------------
 # choose the 11 point median smooth data
-data_agg <- copy(list_of_smooths[[3]])
+data_agg <- fread("data/data_smooth.csv")
 
[…]

### 2020-11-22 06_bat_data.Rmd
minor note

@@
 
-Install the required `R` libraries that are available on CRAN.
+Install the required `R` libraries that are required from CRAN if not already installed.
 
@@
 When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar **(c)**.
-Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_bat_patches.png)
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_08_bats.png)

### 2020-11-22 docs/supplementary_material.tex

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
 \PassOptionsToPackage{hyphens}{url}
 %
-\documentclass[
-]{article}
+\documentclass[]{scrartcl}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
+  \usepackage{textcomp} % provides euro and other symbols
+\else % if luatex or xelatex
   \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
 \fi
-% Use upquote if available, for straight quotes in verbatim environments
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
+% use microtype if available
+\IfFileExists{microtype.sty}{%
 \usepackage[]{microtype}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
 \IfFileExists{parskip.sty}{%
@@
 \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\usepackage{hyperref}
 \hypersetup{
@@
             pdfauthor={Pratik R. Gupte; Christine E. Beardsworth; Orr Spiegel; Emmanuel Lourie; Allert Bijleveld},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage[left=4cm, right=3cm, top=2.5cm, bottom=2.5cm]{geometry}
@@
 \usepackage{longtable,booktabs}
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
+% Fix footnotes in tables (requires footnote package)
+\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 \setcounter{secnumdepth}{2}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
@@
 \linenumbers
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{cslreferences}%
-  {\setlength{\parindent}{0pt}%
-  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
-  {\par}
 
@@
 \author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Allert Bijleveld}
-\date{2020-11-16}
+\date{2020-11-22}
 
@@
 }
+\hypertarget{processing-calibration-data}{%
+\section{Processing calibration data}\label{processing-calibration-data}}
+
+Here we show how the residence patch method (Barraquand and Benhamou \protect\hyperlink{ref-barraquand2008}{2008}; Bijleveld et al.
+\protect\hyperlink{ref-bijleveld2016}{2016}; Oudman et al.
+\protect\hyperlink{ref-oudman2018}{2018}) accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
+
+\hypertarget{prepare-libraries}{%
+\subsection{Prepare libraries}\label{prepare-libraries}}
+
+First we prepare the libraries we need. Libraries can be installed from CRAN if necessary.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# load libs}
+\KeywordTok{library}\NormalTok{(data.table)}
+\KeywordTok{library}\NormalTok{(atlastools)}
+\KeywordTok{library}\NormalTok{(ggplot2)}
+\KeywordTok{library}\NormalTok{(patchwork)}
+
+\CommentTok{# prepare a palette}
+\NormalTok{pal <-}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{access-data-and-preliminary-visualisation}{%
+\subsection{Access data and preliminary visualisation}\label{access-data-and-preliminary-visualisation}}
+
+First we access the data from a local file using the \texttt{data.table} package (Dowle and Srinivasan \protect\hyperlink{ref-dowle2020}{2020}).
+We then visualise the raw data.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# read and plot example data}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/atlas1060_allTrials_annotated.csv"}\NormalTok{)}
+\NormalTok{data_raw <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_calibration_raw.png}
+
+\hypertarget{filter-by-bounding-box}{%
+\subsection{Filter by bounding box}\label{filter-by-bounding-box}}
+
+We first save a copy of the data, so that we can plot the raw data with the cleaned data plotted over it for comparison.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# make a copy using the data.table copy function}
+\NormalTok{data_unproc <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+\end{Highlighting}
+\end{Shaded}
+
+We then filter by a bounding box in order to remove the point outlier to the far south east of the main track.
+We use the \texttt{atl\_filter\_bounds} functions using the \texttt{x\_range} argument, to which we pass the limit in the UTM 31N coordinate reference system.
+This limit is used to exclude all points with an X coordinate \textless{} 645,000.
+
+We then plot the result of filtering, with the excluded point in black, and the points that are retained in green.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# remove inside must be set to falses}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{atl_filter_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data, }
+                          \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
+                          \DataTypeTok{x_range =} \KeywordTok{c}\NormalTok{(}\DecValTok{645000}\NormalTok{, }\KeywordTok{max}\NormalTok{(data}\OperatorTok{$}\NormalTok{x)), }
+                          \DataTypeTok{remove_inside =} \OtherTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{filter-trajectories}{%
+\subsection{Filter trajectories}\label{filter-trajectories}}
+
+\hypertarget{handle-time}{%
+\subsubsection{Handle time}\label{handle-time}}
+
+Time in ATLAS tracking is counted in milliseconds and is represented by a 64-bit integer (type \texttt{long}), which is not natively supported in R; it will instead be converted to a \texttt{numeric}, or \texttt{double}.
+
+This is not what is intended, but it works. The \texttt{bit64} package can help handle 64-bit integers if you want to keep to intended type.
+
+A further issue is that 64-bit integers (whether represented as \texttt{bit64} or \texttt{double}) do not yield meaninful results when you try to convert them to a date-time object, such as of the class \texttt{POSIXct}.
+
+This is because \texttt{as.POSIXct} fails when trying to work with 64-bit integers (it cannot interpret this type), and returns a date many thousands of years in the future (approx.
+52,000 CE) if the time column is converted to \texttt{numeric}.
+
+There are two possible solutions. The parsimonious one is to convert the 64-bit number to a 32-bit short integer (dividing by 1000), or to use the \texttt{nanotime} package.
+
+The conversion method loses an imperceptible amount of precision. The \texttt{nanotime} requires installing another package. The first method is shown here.
+
+In the spirit of not destroying data, we create a second lower-case column called \texttt{time}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# divide by 1000, convert to integer, then convert to POSIXct}
+\NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{add-speed-and-turning-angle}{%
+\subsubsection{Add speed and turning angle}\label{add-speed-and-turning-angle}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# add incoming and outgoing speed}
+\NormalTok{data[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{ (}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(data, }
+                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
+                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
+                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
+             \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
+
+\CommentTok{# add turning angle}
+\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl_turning_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{get-95th-percentile-of-speed-and-angle}{%
+\subsubsection{Get 95th percentile of speed and angle}\label{get-95th-percentile-of-speed-and-angle}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# use sapply}
+\NormalTok{speed_angle_thresholds <-}\StringTok{ }
+\StringTok{  }\KeywordTok{sapply}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(speed_in, speed_out, angle)], }
+\NormalTok{       quantile, }\DataTypeTok{probs =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{na.rm =}\NormalTok{ T)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{filter-on-speed}{%
+\subsubsection{Filter on speed}\label{filter-on-speed}}
+
+Here we use a speed threshold of 15 m/s, the fastest known boat speed.
+We then plot the data with the extreme speeds shown in grey, and the positions retained shown in green.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# make a copy}
+\NormalTok{data_unproc <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{# remove speed outliers}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+            \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(speed_in < 15 & speed_out < 15)"}\NormalTok{))}
+
+\CommentTok{# recalculate speed and angle}
+\NormalTok{data[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{ (}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(data, }
+                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
+                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
+                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
+             \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
+
+\CommentTok{# add turning angle}
+\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl_turning_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_speed_outlier.png}
+
+\hypertarget{smoothing-the-trajectory}{%
+\subsection{Smoothing the trajectory}\label{smoothing-the-trajectory}}
+
+We then apply a median smooth over a moving window (\(K\) = 5).
+This function modifies in place, and does not need to be assigned to a new variable.
+We create a copy of the data before applying the smooth so that we can compare the data before and after smoothin.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# apply a 5 point median smooth, first make a copy}
+\NormalTok{data_unproc <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{# now apply the smooth}
+\KeywordTok{atl_median_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+                  \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }\DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
+                  \DataTypeTok{moving_window =} \DecValTok{5}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_calib_median_smooth.png}
+
+\hypertarget{thinning-the-data}{%
+\subsection{Thinning the data}\label{thinning-the-data}}
+
+Next we thin the data to demonstrate thinning by median smoothing.
+Following this, we plot the median smooth and thinning by aggregation.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# save a copy}
+\NormalTok{data_unproc <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{# remove columns we don't need}
+\NormalTok{data <-}\StringTok{ }\NormalTok{data[, }\KeywordTok{setdiff}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(data), }
+                       \KeywordTok{c}\NormalTok{(}\StringTok{"tID"}\NormalTok{, }\StringTok{"Timestamp"}\NormalTok{, }\StringTok{"id"}\NormalTok{, }\StringTok{"TIME"}\NormalTok{, }\StringTok{"UTCtime"}\NormalTok{)), }
+\NormalTok{             with =}\StringTok{ }\OtherTok{FALSE}\NormalTok{]}
+
+\CommentTok{# thin to a 30s interval}
+\NormalTok{data_thin <-}\StringTok{ }\KeywordTok{atl_thin_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+                           \DataTypeTok{interval =} \DecValTok{30}\NormalTok{,}
+                           \DataTypeTok{method =} \StringTok{"aggregate"}\NormalTok{,}
+                           \DataTypeTok{id_columns =} \StringTok{"TAG"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_calib_smooth_thin.png}
+
+\hypertarget{residence-patches}{%
+\subsection{Residence patches}\label{residence-patches}}
+
+\hypertarget{get-waypoint-centroids}{%
+\subsubsection{Get waypoint centroids}\label{get-waypoint-centroids}}
+
+We subset the annotated calibration data to select the waypoints and the positions around them which are supposed to be the locations of known stops.
+Since each stop was supposed to be 5 minutes long, there are multiple points in each known stop.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(stringi)}
+\NormalTok{data_res <-}\StringTok{ }\NormalTok{data_unproc[}\KeywordTok{stri_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{)]}
+\end{Highlighting}
+\end{Shaded}
+
+From this data, we get the centroid of known stops, and determine the time difference between the first and last point within 50 metres, and within 10 minutes of the waypoint positions' median time.
+
+Essentially, this means that the maximum duration of a stop can be 20 minutes, and stops above this duration are not expected.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# get centroid}
+\NormalTok{data_res_summary <-}\StringTok{ }\NormalTok{data_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{x_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{),}
+                                    \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{)),}
+\NormalTok{                                    t_median =}\StringTok{ }\KeywordTok{median}\NormalTok{(time)}\ErrorTok{)}\NormalTok{,}
+\NormalTok{                             by =}\StringTok{ "tID"}\NormalTok{]}
+
+\CommentTok{# now get times 10 mins before and after}
+\NormalTok{data_res_summary[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{t_min =}\NormalTok{ t_median }\OperatorTok{-}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{),}
+                        \DataTypeTok{t_max =}\NormalTok{ t_median }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{))]}
+
+\CommentTok{# make a list of positions 10min before and after}
+\NormalTok{wp_data <-}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(l, u, mx, my) \{}
+\NormalTok{  tmp_data <-}\StringTok{ }\NormalTok{data_unproc[}\KeywordTok{inrange}\NormalTok{(time, l, u)]}
+\NormalTok{  tmp_data[, distance }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{sqrt}\NormalTok{((mx }\OperatorTok{-}\StringTok{ }\NormalTok{x)}\OperatorTok{^}\DecValTok{2} \OperatorTok{+}\StringTok{ }\NormalTok{(my }\OperatorTok{-}\StringTok{ }\NormalTok{y)}\OperatorTok{^}\DecValTok{2}\NormalTok{)]}
+
+  \CommentTok{# keep within 50}
+\NormalTok{  tmp_data <-}\StringTok{ }\NormalTok{tmp_data[distance }\OperatorTok{<=}\StringTok{ }\DecValTok{50}\NormalTok{, ]}
+
+  \CommentTok{# get duration}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(tmp_data}\OperatorTok{$}\NormalTok{time)))}
+
+\NormalTok{\}, data_res_summary}\OperatorTok{$}\NormalTok{t_min, data_res_summary}\OperatorTok{$}\NormalTok{t_max,}
+\NormalTok{   data_res_summary}\OperatorTok{$}\NormalTok{x_median, data_res_summary}\OperatorTok{$}\NormalTok{y_median,}
+\DataTypeTok{SIMPLIFY =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{prepare-data}{%
+\subsubsection{Prepare data}\label{prepare-data}}
+
+An indicator of individual residence at or near a position can be useful when attempting to identify residence patches.
+Positions can be filtered on a metric such as residence time (Bracis, Bildstein, and Mueller \protect\hyperlink{ref-bracis2018}{2018}).
+
+\hypertarget{calculate-residence-time}{%
+\subsubsection{Calculate residence time}\label{calculate-residence-time}}
+
+First we calculate the residence time with a radius of 50 metres.
+For this, we need a dataframe with coordinates, the timestamp, and the animal id.
+We save this data to file for later use.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# load recurse}
+\KeywordTok{library}\NormalTok{(recurse)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# get 4 column data}
+\NormalTok{data_for_patch <-}\StringTok{ }\NormalTok{data_thin[, }\KeywordTok{list}\NormalTok{(x, y, time, TAG)]}
+
+\CommentTok{# get recurse data for a 10m radius}
+\NormalTok{recurse_stats <-}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data_for_patch,}
+                               \DataTypeTok{radius =} \DecValTok{50}\NormalTok{, }\DataTypeTok{timeunits =} \StringTok{"mins"}\NormalTok{)}
+
+\CommentTok{# assign to recurse data}
+\NormalTok{data_for_patch[, res_time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{recurse_stats}\OperatorTok{$}\NormalTok{residenceTime]}
+
+\CommentTok{# save recurse data}
+\KeywordTok{fwrite}\NormalTok{(data_for_patch, }\DataTypeTok{file =} \StringTok{"data/data_calib_for_patch.csv"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{run-residence-patch-method}{%
+\subsubsection{Run residence patch method}\label{run-residence-patch-method}}
+
+We subset data with a residence time \textgreater{} 5 minutes in order to construct residence patches.
+From this subset, we construct residence patches using the parameters: \texttt{buffer\_radius} = 5 metres, \texttt{lim\_spat\_indep} = 50 metres, \texttt{lim\_time\_indep} = 5 minutes, and \texttt{min\_fixes} = 3.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# assign id as tag}
+\NormalTok{data_for_patch[, id }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.character}\NormalTok{(TAG)]}
+
+\CommentTok{# on known residence points}
+\NormalTok{patch_res_known <-}\StringTok{ }\KeywordTok{atl_res_patch}\NormalTok{(data_for_patch[res_time }\OperatorTok{>=}\StringTok{ }\DecValTok{5}\NormalTok{, ], }
+                                \DataTypeTok{buffer_radius =} \DecValTok{5}\NormalTok{,}
+                                \DataTypeTok{lim_spat_indep =} \DecValTok{50}\NormalTok{,}
+                                \DataTypeTok{lim_time_indep =} \DecValTok{5}\NormalTok{,}
+                                \DataTypeTok{min_fixes =} \DecValTok{3}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{get-spatial-and-summary-objects}{%
+\subsubsection{Get spatial and summary objects}\label{get-spatial-and-summary-objects}}
+
+We get spatial and summary ouput of the residence patch method using the \texttt{atl\_patch\_summary} function using the options \texttt{which\_data} = ``spatial'' and \texttt{which\_data} = "summary.
+We use a buffer radius here of 20 metres for the spatial buffer, despite using a buffer radius of 5 metres earlier, simply because it is easier to visualise in the output figure.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# for the known and unkniwn patches}
+\NormalTok{patch_sf_data <-}\StringTok{ }\KeywordTok{atl_patch_summary}\NormalTok{(patch_res_known, }
+                                   \DataTypeTok{which_data =} \StringTok{"spatial"}\NormalTok{,}
+                                   \DataTypeTok{buffer_radius =} \DecValTok{20}\NormalTok{)}
+
+\CommentTok{# assign crs}
+\NormalTok{sf}\OperatorTok{::}\KeywordTok{st_crs}\NormalTok{(patch_sf_data) <-}\StringTok{ }\DecValTok{32631}
+
+\CommentTok{# get summary data}
+\NormalTok{patch_summary_data <-}\StringTok{ }\KeywordTok{atl_patch_summary}\NormalTok{(patch_res_known, }
+                                        \DataTypeTok{which_data =} \StringTok{"summary"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{prepare-to-plot-data}{%
+\subsubsection{Prepare to plot data}\label{prepare-to-plot-data}}
+
+We read in the island's shapefile to plot it as a background for the residence patch figure.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# read griend}
+\NormalTok{griend <-}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st_read}\NormalTok{(}\StringTok{"data/griend_polygon/griend_polygon.shp"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{compare-patch-metrics}{%
+\subsection{Compare patch metrics}\label{compare-patch-metrics}}
+
+We then merge the annoated, known stop data with the calculated patch duration.
+We filter this data to exclude one exceedingly long outlier of about an hour (WP080), which how
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# get known patch summary}
+\NormalTok{data_res <-}\StringTok{ }\NormalTok{data_unproc[stringi}\OperatorTok{::}\KeywordTok{stri_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{), ]}
+
+\CommentTok{# get waypoint summary}
+\NormalTok{patch_summary_real <-}\StringTok{ }\NormalTok{data_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{nfixes_real =}\NormalTok{ .N,}
+                                      \DataTypeTok{x_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{),}
+                                      \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{)), }
+\NormalTok{                               by =}\StringTok{ "tID"}\NormalTok{]}
+
+\CommentTok{# add real duration}
+\NormalTok{patch_summary_real[, duration_real }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{wp_data]}
+
+\CommentTok{# round median coordinate for inferred patches}
+\NormalTok{patch_summary_inferred <-}\StringTok{ }
+\StringTok{  }\NormalTok{patch_summary_data[, }
+                     \KeywordTok{c}\NormalTok{(}\StringTok{"x_median"}\NormalTok{, }\StringTok{"y_median"}\NormalTok{, }
+                       \StringTok{"nfixes"}\NormalTok{, }\StringTok{"duration"}\NormalTok{, }\StringTok{"patch"}\NormalTok{)}
+\NormalTok{                     ][, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{x_median =} \KeywordTok{round}\NormalTok{(x_median, }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{),}
+                              \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(y_median, }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{))]}
+
+\CommentTok{# join with respatch summary}
+\NormalTok{patch_summary_compare <-}\StringTok{ }
+\StringTok{  }\KeywordTok{merge}\NormalTok{(patch_summary_real,}
+\NormalTok{        patch_summary_inferred, }
+        \DataTypeTok{on =} \KeywordTok{c}\NormalTok{(}\StringTok{"x_median"}\NormalTok{, }\StringTok{"y_median"}\NormalTok{),}
+        \DataTypeTok{all.x =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{all.y =} \OtherTok{TRUE}\NormalTok{)}
+
+\CommentTok{# drop nas}
+\NormalTok{patch_summary_compare <-}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(patch_summary_compare)}
+
+\CommentTok{# drop patch around WP080}
+\NormalTok{patch_summary_compare <-}\StringTok{ }\NormalTok{patch_summary_compare[tID }\OperatorTok{!=}\StringTok{ "WP080"}\NormalTok{, ]}
+\end{Highlighting}
+\end{Shaded}
+
+7 patches are identified where there are no waypoints, while 2 waypoints are not identified as patches.
+These waypoints consisted of 6 and 15 (WP098 and WP092) positions respectively, and were lost when the data were aggregated to 30 second intervals.
+
+\hypertarget{linear-model-durations}{%
+\subsubsection{Linear model durations}\label{linear-model-durations}}
+
+We run a simple linear model.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# get linear model}
+\NormalTok{model_duration <-}\StringTok{ }\KeywordTok{lm}\NormalTok{(duration_real }\OperatorTok{~}\StringTok{ }\NormalTok{duration,}
+                     \DataTypeTok{data =}\NormalTok{ patch_summary_compare)}
+
+\CommentTok{# get R2}
+\KeywordTok{summary}\NormalTok{(model_duration)}
+
+\CommentTok{# write to file}
+\KeywordTok{writeLines}\NormalTok{(}
+  \DataTypeTok{text =} \KeywordTok{capture.output}\NormalTok{(}
+    \KeywordTok{summary}\NormalTok{(model_duration)}
+\NormalTok{  ),}
+  \DataTypeTok{con =} \StringTok{"data/model_output_residence_patch.txt"}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_calib_lm_duration.png}
+
 \hypertarget{processing-egyptian-fruit-bat-tracks}{%
@@
 
-\hypertarget{prepare-libraries}{%
-\subsection{Prepare libraries}\label{prepare-libraries}}
+\hypertarget{prepare-libraries-1}{%
+\subsection{Prepare libraries}\label{prepare-libraries-1}}
 
-Install the required \texttt{R} libraries that are available on CRAN.
+Install the required \texttt{R} libraries that are required from CRAN if not already installed.
 
@@
 \begin{Highlighting}[]
-\CommentTok{\# load libs}
+\CommentTok{# load libs}
 \KeywordTok{library}\NormalTok{(data.table)}
@@
 
-\CommentTok{\# prepare a palette}
-\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
+\CommentTok{# prepare a palette}
+\NormalTok{pal <-}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
 \end{Highlighting}
@@
 
-\CommentTok{\# installation using remotes}
-\NormalTok{remotes}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
+\CommentTok{# installation using remotes}
+\NormalTok{remotes}\OperatorTok{::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# prepare the connection}
-\NormalTok{con <{-}}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}\DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(), }
-                 \DataTypeTok{dbname =} \StringTok{"data/Three\_example\_bats.sql"}\NormalTok{)}
+\CommentTok{# prepare the connection}
+\NormalTok{con <-}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}\DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(), }
+                 \DataTypeTok{dbname =} \StringTok{"data/Three_example_bats.sql"}\NormalTok{)}
 
-\CommentTok{\# list the tables}
-\NormalTok{table\_name <{-}}\StringTok{ }\KeywordTok{dbListTables}\NormalTok{(con)}
+\CommentTok{# list the tables}
+\NormalTok{table_name <-}\StringTok{ }\KeywordTok{dbListTables}\NormalTok{(con)}
 
-\CommentTok{\# prepare to query all tables}
-\NormalTok{query <{-}}\StringTok{ }\KeywordTok{sprintf}\NormalTok{(}\StringTok{\textquotesingle{}select * from }\CharTok{\textbackslash{}"}\StringTok{\%s}\CharTok{\textbackslash{}"}\StringTok{\textquotesingle{}}\NormalTok{, table\_name)}
+\CommentTok{# prepare to query all tables}
+\NormalTok{query <-}\StringTok{ }\KeywordTok{sprintf}\NormalTok{(}\StringTok{'select * from }\CharTok{\textbackslash{}"}\StringTok{%s}\CharTok{\textbackslash{}"}\StringTok{'}\NormalTok{, table_name)}
 
-\CommentTok{\# query the database}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{dbGetQuery}\NormalTok{(}\DataTypeTok{conn =}\NormalTok{ con, }\DataTypeTok{statement =}\NormalTok{ query)}
+\CommentTok{# query the database}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{dbGetQuery}\NormalTok{(}\DataTypeTok{conn =}\NormalTok{ con, }\DataTypeTok{statement =}\NormalTok{ query)}
 
-\CommentTok{\# disconnect from database}
+\CommentTok{# disconnect from database}
 \KeywordTok{dbDisconnect}\NormalTok{(con)}
@@
 \begin{Highlighting}[]
-\CommentTok{\# convert data to datatable}
+\CommentTok{# convert data to datatable}
 \KeywordTok{setDT}\NormalTok{(data)}
 
-\CommentTok{\# write data for QGIS}
-\KeywordTok{fwrite}\NormalTok{(data, }\DataTypeTok{file =} \StringTok{"data/bat\_data.csv"}\NormalTok{)}
+\CommentTok{# write data for QGIS}
+\KeywordTok{fwrite}\NormalTok{(data, }\DataTypeTok{file =} \StringTok{"data/bat_data.csv"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# split bat data by tag}
-\CommentTok{\# first make a copy using the data.table function copy}
-\CommentTok{\# this prevents the orignal data from being modified by atlastools}
-\CommentTok{\# functions which DO MODIFY BY REFERENCE!}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+\CommentTok{# split bat data by tag}
+\CommentTok{# first make a copy using the data.table function copy}
+\CommentTok{# this prevents the orignal data from being modified by atlastools}
+\CommentTok{# functions which DO MODIFY BY REFERENCE!}
+\NormalTok{data_split <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
 
-\CommentTok{\# now split}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_split, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
+\CommentTok{# now split}
+\NormalTok{data_split <-}\StringTok{ }\KeywordTok{split}\NormalTok{(data_split, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get SD.}
-\CommentTok{\# since the data are data.tables, no assignment is necessary}
+\CommentTok{# get SD.}
+\CommentTok{# since the data are data.tables, no assignment is necessary}
 \KeywordTok{invisible}\NormalTok{(}
-  \KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+  \KeywordTok{lapply}\NormalTok{(data_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
 \NormalTok{    dt[, SD }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(VARX }\OperatorTok{+}\StringTok{ }\NormalTok{VARY }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{COVXY))]}
@@
 \begin{Highlighting}[]
-\CommentTok{\# filter for SD <= 20}
-\CommentTok{\# here, reassignment is necessary as rows are being removed}
-\CommentTok{\# the atl\_filter\_covariates function could have been used here}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+\CommentTok{# filter for SD <= 20}
+\CommentTok{# here, reassignment is necessary as rows are being removed}
+\CommentTok{# the atl_filter_covariates function could have been used here}
+\NormalTok{data_split <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
 
-\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}
     \DataTypeTok{data =}\NormalTok{ dt,}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get speeds as with SD, no reassignment required for columns}
+\CommentTok{# get speeds as with SD, no reassignment required for columns}
 \KeywordTok{invisible}\NormalTok{(}
-  \KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+  \KeywordTok{lapply}\NormalTok{(data_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
 
-    \CommentTok{\# first process time to seconds}
-    \CommentTok{\# assign to a new column}
+    \CommentTok{# first process time to seconds}
+    \CommentTok{# assign to a new column}
 \NormalTok{    dt[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{floor}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}
 
-\NormalTok{    dt[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(dt, }
+\NormalTok{    dt[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(dt, }
                                        \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{, }
@@
                                        \DataTypeTok{type =} \StringTok{"in"}\NormalTok{),}
-              \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(dt, }
+              \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(dt, }
                                        \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{, }
@@
 \begin{Highlighting}[]
-\CommentTok{\# filter speeds}
-\CommentTok{\# reassignment is required here}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
-\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(dt, }\DataTypeTok{cols =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed\_in"}\NormalTok{, }\StringTok{"speed\_out"}\NormalTok{))}
+\CommentTok{# filter speeds}
+\CommentTok{# reassignment is required here}
+\NormalTok{data_split <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(dt, }\DataTypeTok{cols =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed_in"}\NormalTok{, }\StringTok{"speed_out"}\NormalTok{))}
 
-\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dt,}
-                              \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed\_in <= 20"}\NormalTok{,}
-                                          \StringTok{"speed\_out <= 20"}\NormalTok{))}
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dt,}
+                              \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed_in <= 20"}\NormalTok{,}
+                                          \StringTok{"speed_out <= 20"}\NormalTok{))}
 \NormalTok{\})}
@@
 \begin{Highlighting}[]
-\CommentTok{\# since the function modifies in place, we shall make a copy}
-\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data\_split)}
+\CommentTok{# since the function modifies in place, we shall make a copy}
+\NormalTok{data_smooth <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data_split)}
 
-\CommentTok{\# split the data again}
-\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_smooth, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
+\CommentTok{# split the data again}
+\NormalTok{data_smooth <-}\StringTok{ }\KeywordTok{split}\NormalTok{(data_smooth, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# apply the median smooth to each list element}
-\CommentTok{\# no reassignment is required as THE FUNCTION MODIFIES IN PLACE!}
+\CommentTok{# apply the median smooth to each list element}
+\CommentTok{# no reassignment is required as THE FUNCTION MODIFIES IN PLACE!}
 \KeywordTok{invisible}\NormalTok{(}
 
-  \CommentTok{\# the function arguments to atl\_median\_smooth}
-  \CommentTok{\# can be passed directly in lapply}
+  \CommentTok{# the function arguments to atl_median_smooth}
+  \CommentTok{# can be passed directly in lapply}
 
-  \KeywordTok{lapply}\NormalTok{(}\DataTypeTok{X =}\NormalTok{ data\_smooth, }
-         \DataTypeTok{FUN =}\NormalTok{ atl\_median\_smooth,}
-         \DataTypeTok{time =} \StringTok{"time"}\NormalTok{, }\DataTypeTok{moving\_window =} \DecValTok{5}\NormalTok{)}
+  \KeywordTok{lapply}\NormalTok{(}\DataTypeTok{X =}\NormalTok{ data_smooth, }
+         \DataTypeTok{FUN =}\NormalTok{ atl_median_smooth,}
+         \DataTypeTok{time =} \StringTok{"time"}\NormalTok{, }\DataTypeTok{moving_window =} \DecValTok{5}\NormalTok{)}
 \NormalTok{)}
@@
 \begin{Highlighting}[]
-\CommentTok{\# load recurse}
+\CommentTok{# load recurse}
 \KeywordTok{library}\NormalTok{(recurse)}
 
-\CommentTok{\# split the data}
-\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_smooth, data\_smooth}\OperatorTok{$}\NormalTok{TAG)}
+\CommentTok{# split the data}
+\NormalTok{data_smooth <-}\StringTok{ }\KeywordTok{split}\NormalTok{(data_smooth, data_smooth}\OperatorTok{$}\NormalTok{TAG)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get residence times}
+\CommentTok{# get residence times}
 
-\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_smooth, }\ControlFlowTok{function}\NormalTok{(dt) \{}
-  \CommentTok{\# do basic recurse}
-\NormalTok{  dt\_recurse <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(}
+\NormalTok{data_residence <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_smooth, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+  \CommentTok{# do basic recurse}
+\NormalTok{  dt_recurse <-}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(}
     \DataTypeTok{x =}\NormalTok{ dt[, }\KeywordTok{c}\NormalTok{(}\StringTok{"X"}\NormalTok{, }\StringTok{"Y"}\NormalTok{, }\StringTok{"time"}\NormalTok{, }\StringTok{"TAG"}\NormalTok{)],}
@@
 
-  \CommentTok{\# get revisit stats}
-\NormalTok{  dt\_recurse <{-}}\StringTok{ }\KeywordTok{setDT}\NormalTok{(}
-\NormalTok{    dt\_recurse[[}\StringTok{"revisitStats"}\NormalTok{]]}
+  \CommentTok{# get revisit stats}
+\NormalTok{  dt_recurse <-}\StringTok{ }\KeywordTok{setDT}\NormalTok{(}
+\NormalTok{    dt_recurse[[}\StringTok{"revisitStats"}\NormalTok{]]}
 \NormalTok{  )}
 
-  \CommentTok{\# count long absences from the area}
-\NormalTok{  dt\_recurse[, timeSinceLastVisit }\OperatorTok{:}\ErrorTok{=}
-\StringTok{          }\KeywordTok{ifelse}\NormalTok{(}\KeywordTok{is.na}\NormalTok{(timeSinceLastVisit), }\OperatorTok{{-}}\OtherTok{Inf}\NormalTok{, timeSinceLastVisit)]}
-\NormalTok{  dt\_recurse[, longAbsenceCounter }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{cumsum}\NormalTok{(timeSinceLastVisit }\OperatorTok{>}\StringTok{ }\DecValTok{60}\NormalTok{),}
+  \CommentTok{# count long absences from the area}
+\NormalTok{  dt_recurse[, timeSinceLastVisit }\OperatorTok{:}\ErrorTok{=}
+\StringTok{          }\KeywordTok{ifelse}\NormalTok{(}\KeywordTok{is.na}\NormalTok{(timeSinceLastVisit), }\OperatorTok{-}\OtherTok{Inf}\NormalTok{, timeSinceLastVisit)]}
+\NormalTok{  dt_recurse[, longAbsenceCounter }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{cumsum}\NormalTok{(timeSinceLastVisit }\OperatorTok{>}\StringTok{ }\DecValTok{60}\NormalTok{),}
 \NormalTok{             by =}\StringTok{ }\NormalTok{.(coordIdx)}
 \NormalTok{             ]}
-  \CommentTok{\# get data before the first long absence of 60 mins}
-\NormalTok{  dt\_recurse <{-}}\StringTok{ }\NormalTok{dt\_recurse[longAbsenceCounter }\OperatorTok{<}\StringTok{ }\DecValTok{1}\NormalTok{, ]}
+  \CommentTok{# get data before the first long absence of 60 mins}
+\NormalTok{  dt_recurse <-}\StringTok{ }\NormalTok{dt_recurse[longAbsenceCounter }\OperatorTok{<}\StringTok{ }\DecValTok{1}\NormalTok{, ]}
 
-\NormalTok{  dt\_recurse <{-}}\StringTok{ }\NormalTok{dt\_recurse[, }\KeywordTok{list}\NormalTok{(}
+\NormalTok{  dt_recurse <-}\StringTok{ }\NormalTok{dt_recurse[, }\KeywordTok{list}\NormalTok{(}
     \DataTypeTok{resTime =} \KeywordTok{sum}\NormalTok{(timeInside),}
@@
 
-  \CommentTok{\# prepare and merge existing data with recursion data}
+  \CommentTok{# prepare and merge existing data with recursion data}
 \NormalTok{  dt[, coordIdx }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\KeywordTok{nrow}\NormalTok{(dt))]}
 
-\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{merge}\NormalTok{(dt, }
-\NormalTok{              dt\_recurse[, }\KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{, }\StringTok{"resTime"}\NormalTok{)], }
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{merge}\NormalTok{(dt, }
+\NormalTok{              dt_recurse[, }\KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{, }\StringTok{"resTime"}\NormalTok{)], }
               \DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{))}
@@
 \begin{Highlighting}[]
-\CommentTok{\# bind the list}
-\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_residence)}
+\CommentTok{# bind the list}
+\NormalTok{data_residence <-}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data_residence)}
 
-\CommentTok{\# get time as human readable}
-\NormalTok{data\_residence[, ts }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.POSIXct}\NormalTok{(time, }\DataTypeTok{origin =} \StringTok{"1970{-}01{-}01"}\NormalTok{)]}
+\CommentTok{# get time as human readable}
+\NormalTok{data_residence[, ts }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.POSIXct}\NormalTok{(time, }\DataTypeTok{origin =} \StringTok{"1970-01-01"}\NormalTok{)]}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# add an id column}
-\NormalTok{data\_residence[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{id =}\NormalTok{ TAG,}
+\CommentTok{# add an id column}
+\NormalTok{data_residence[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{id =}\NormalTok{ TAG,}
                       \DataTypeTok{x =}\NormalTok{ X, }\DataTypeTok{y =}\NormalTok{ Y)]}
 
-\CommentTok{\# filter for residence time > 5 minutes}
-\NormalTok{data\_residence <{-}}\StringTok{ }\NormalTok{data\_residence[resTime }\OperatorTok{>}\StringTok{ }\DecValTok{5}\NormalTok{, ]}
+\CommentTok{# filter for residence time > 5 minutes}
+\NormalTok{data_residence <-}\StringTok{ }\NormalTok{data_residence[resTime }\OperatorTok{>}\StringTok{ }\DecValTok{5}\NormalTok{, ]}
 
-\CommentTok{\# split the data}
-\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_residence, data\_residence}\OperatorTok{$}\NormalTok{TAG)}
+\CommentTok{# split the data}
+\NormalTok{data_residence <-}\StringTok{ }\KeywordTok{split}\NormalTok{(data_residence, data_residence}\OperatorTok{$}\NormalTok{TAG)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# segment into residence patches}
-\NormalTok{data\_patches <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_residence, atl\_res\_patch,}
-                       \DataTypeTok{buffer\_radius =} \DecValTok{25}\NormalTok{)}
+\CommentTok{# segment into residence patches}
+\NormalTok{data_patches <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_residence, atl_res_patch,}
+                       \DataTypeTok{buffer_radius =} \DecValTok{25}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get data spatials}
-\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_patches, atl\_patch\_summary,}
-                        \DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{,}
-                        \DataTypeTok{buffer\_radius =} \DecValTok{25}\NormalTok{)}
+\CommentTok{# get data spatials}
+\NormalTok{data_spatials <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_patches, atl_patch_summary,}
+                        \DataTypeTok{which_data =} \StringTok{"spatial"}\NormalTok{,}
+                        \DataTypeTok{buffer_radius =} \DecValTok{25}\NormalTok{)}
 
-\CommentTok{\# bind list}
-\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_spatials)}
+\CommentTok{# bind list}
+\NormalTok{data_spatials <-}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data_spatials)}
 
-\CommentTok{\# convert to sf}
+\CommentTok{# convert to sf}
 \KeywordTok{library}\NormalTok{(sf)}
-\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{st\_sf}\NormalTok{(data\_spatials, }\DataTypeTok{sf\_column\_name =} \StringTok{"polygons"}\NormalTok{)}
+\NormalTok{data_spatials <-}\StringTok{ }\KeywordTok{st_sf}\NormalTok{(data_spatials, }\DataTypeTok{sf_column_name =} \StringTok{"polygons"}\NormalTok{)}
 
-\CommentTok{\# assign a crs}
-\KeywordTok{st\_crs}\NormalTok{(data\_spatials) <{-}}\StringTok{ }\KeywordTok{st\_crs}\NormalTok{(}\DecValTok{2039}\NormalTok{)}
+\CommentTok{# assign a crs}
+\KeywordTok{st_crs}\NormalTok{(data_spatials) <-}\StringTok{ }\KeywordTok{st_crs}\NormalTok{(}\DecValTok{2039}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\KeywordTok{st\_write}\NormalTok{(data\_spatials,}
-         \DataTypeTok{dsn =} \StringTok{"data/data\_bat\_residence\_patches.gpkg"}\NormalTok{)}
+\KeywordTok{st_write}\NormalTok{(data_spatials,}
+         \DataTypeTok{dsn =} \StringTok{"data/data_bat_residence_patches.gpkg"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{data\_clean <{-}}\StringTok{ }\KeywordTok{fwrite}\NormalTok{(}\KeywordTok{rbindlist}\NormalTok{(data\_smooth),}
-                     \DataTypeTok{file =} \StringTok{"data/data\_bat\_smooth.csv"}\NormalTok{)}
+\NormalTok{data_clean <-}\StringTok{ }\KeywordTok{fwrite}\NormalTok{(}\KeywordTok{rbindlist}\NormalTok{(data_smooth),}
+                     \DataTypeTok{file =} \StringTok{"data/data_bat_smooth.csv"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get summary}
-\NormalTok{patch\_summary <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_patches, atl\_patch\_summary)}
+\CommentTok{# get summary}
+\NormalTok{patch_summary <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_patches, atl_patch_summary)}
 
-\CommentTok{\# bind summary}
-\NormalTok{patch\_summary <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(patch\_summary)}
+\CommentTok{# bind summary}
+\NormalTok{patch_summary <-}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(patch_summary)}
 
-\CommentTok{\# write}
-\KeywordTok{fwrite}\NormalTok{(patch\_summary,}
-       \StringTok{"data/data\_bat\_patch\_summary.csv"}\NormalTok{)}
+\CommentTok{# write}
+\KeywordTok{fwrite}\NormalTok{(patch_summary,}
+       \StringTok{"data/data_bat_patch_summary.csv"}\NormalTok{)}
 \end{Highlighting}
@@
 \centering
-\includegraphics{figures/fig_bat_patches.png}
+\includegraphics{figures/fig_08_bats.png}
 \caption{A visual examination of plots of the bats' residence patches and linear approximations of paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site over the three nights \textbf{(a)}.
@@
 \hypertarget{refs}{}
-\begin{cslreferences}
+\leavevmode\hypertarget{ref-barraquand2008}{}%
+Barraquand, Frédéric, and Simon Benhamou.
+2008.
+``Animal Movements in Heterogeneous Landscapes: Identifying Profitable Places and Homogeneous Movement Bouts.'' \emph{Ecology} 89 (12): 3336--48.
+\url{https://doi.org/10.1890/08-0162.1}.
+
+\leavevmode\hypertarget{ref-bijleveld2016}{}%
+Bijleveld, Allert Imre, Robert B MacCurdy, Ying-Chi Chan, Emma Penning, Richard M.
+Gabrielson, John Cluderay, Erik L.
+Spaulding, et al.
+2016.
+``Understanding Spatial Distributions: Negative Density-Dependence in Prey Causes Predators to Trade-Off Prey Quantity with Quality.'' \emph{Proceedings of the Royal Society B: Biological Sciences} 283 (1828): 20151557.
+\url{https://doi.org/10.1098/rspb.2015.1557}.
+
 \leavevmode\hypertarget{ref-bracis2018}{}%
@@
 Gupte, Pratik Rajan. 2020. ``Atlastools: Pre-Processing Tools for High Frequency Tracking Data.'' Zenodo. \url{https://doi.org/10.5281/ZENODO.4033154}.
+
+\leavevmode\hypertarget{ref-oudman2018}{}%
+Oudman, Thomas, Theunis Piersma, Mohamed V.
+Ahmedou Salem, Marieke E.
+Feis, Anne Dekinga, Sander Holthuijsen, Job ten Horn, Jan A.
+van Gils, and Allert I.
+Bijleveld.
+2018.
+``Resource Landscapes Explain Contrasting Patterns of Aggregation and Site Fidelity by Red Knots at Two Wintering Sites.'' \emph{Movement Ecology} 6 (1): 24--24.
+\url{https://doi.org/10.1186/s40462-018-0142-4}.
 
@@
 \url{https://doi.org/10.1109/IPSN.2016.7460662}.
-\end{cslreferences}
 

### 2020-11-22 index.rmd

@@
 output: bookdown::pdf_document2
-documentclass: article
+documentclass: scrartcl
 toc-depth: 2

### 2020-11-23 05_calibration_data.Rmd
add captions and model out

@@
 
-![](figures/fig_calibration_raw.png)
+![The raw data from a calibration exercise conducted around the island of Griend in the Dutch Wadden Sea.
+A handheld WATLAS tag was used to examine how ATLAS data compared to GPS tracks, and we use the WATLAS data here to demonstrate the basics of the pre-processing pipeline, as well as validate the residence patch method.
+It is immediately clear from the figure that the track shows location errors, both in the form of point outliers as well as small-scale errors around the true location.](figures/fig_calibration_raw.png)
 
@@
 ```
+
+![Removal of a point outlier using the function `atl_filter_bounds`.
+The point outlier (black point) is removed based on its X coordinate value, with the data filtered to exclude positions with an X coordinate < 645,000 in the UTM 31N system.
+Positions that are retained are shown in green.](figures/fig_calib_bbox.png)
 
@@
 
-![](figures/fig_speed_outlier.png)
+![Improving data quality by filtering out positions that would require unrealistic movement.
+We removed positions with speeds $\geq$ 15 m/s, which is the fastest possible speed in this calibration data, part of which was collected in a moving boat around Griend.
+Grey positions are removed, while green positions are retained.
+Rectangles indicate areas expanded for visualisation in following figures.](figures/fig_speed_outlier.png)
 
@@
 
-![](figures/fig_calib_median_smooth.png)
+![Reducing small-scale location error using a median smooth with a moving window $K$ = 5.
+Median smoothed positions are shown in green, while raw, unfiltered data is shown in grey.
+Median smoothing successfully recovers the likely path of the track without a loss of data.
+The area shown is the upper rectangle from Figure 3.](figures/fig_calib_median_smooth.png)
 
@@
 
-![](figures/fig_calib_smooth_thin.png)
+![Thinning by aggregation over a 30 second interval (down from 1 second) preserves track structure while reducing the data volume for computation.
+Here, thinned positions are shown as purple squares, with the size of the square indicating the number of positions within the 30 second bin used to obtain the average position.
+Green points show the median smoothed data from Figure 4, while the raw data are shown in grey.](figures/fig_calib_smooth_thin.png)
 
@@
 
+```{r echo=FALSE}
+ggsave(fig_basic_residence,
+       filename = "figures/fig_calib_residence.png",
+       height = 170, width = 80, units = "mm")
+```
+
+![Classifying thinned data into residence patches yields robust estimates of the duration of known stops.
+The island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) is shown in beige.
+Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
+However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without crossed-squares).
+The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).](figures/fig_calib_residence.png)
+
 ## Compare patch metrics
@@
 
-![](figures/fig_calib_lm_duration.png)
+![The inferred duration of residence patches corresponds very closely to the real duration (grey circles, red line shows linear model fit), with an underestimation of the true duration of around 2%.
+The dashed black line represents $y = x$ for reference.](figures/fig_calib_lm_duration.png)
+
+### Linear model summary
+
+```{r}
+cat(
+  readLines(
+    con = "data/model_output_residence_patch.txt"
+  ), sep = "\n"
+)
+```
 

### 2020-11-23 docs/supplementary_material.tex
add supplement tex

--- 2020-11-16 docs/supplementary_material.tex
@@
 \documentclass[
-]{article}
+]{scrartcl}
 \usepackage{lmodern}
@@
 \setmainfont{Times New Roman}
+\setsansfont{Arial}
 
@@
 \author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Allert Bijleveld}
-\date{2020-11-16}
+\date{2020-11-23}
 
@@
 }
+\hypertarget{processing-calibration-data}{%
+\section{Processing calibration data}\label{processing-calibration-data}}
+
+Here we show how the residence patch method (Barraquand and Benhamou \protect\hyperlink{ref-barraquand2008}{2008}; Bijleveld et al.
+\protect\hyperlink{ref-bijleveld2016}{2016}; Oudman et al.
+\protect\hyperlink{ref-oudman2018}{2018}) accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
+
+\hypertarget{prepare-libraries}{%
+\subsection{Prepare libraries}\label{prepare-libraries}}
+
+First we prepare the libraries we need. Libraries can be installed from CRAN if necessary.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# load libs}
+\KeywordTok{library}\NormalTok{(data.table)}
+\KeywordTok{library}\NormalTok{(atlastools)}
+\KeywordTok{library}\NormalTok{(ggplot2)}
+\KeywordTok{library}\NormalTok{(patchwork)}
+
+\CommentTok{\# prepare a palette}
+\NormalTok{pal <{-}}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{access-data-and-preliminary-visualisation}{%
+\subsection{Access data and preliminary visualisation}\label{access-data-and-preliminary-visualisation}}
+
+First we access the data from a local file using the \texttt{data.table} package (Dowle and Srinivasan \protect\hyperlink{ref-dowle2020}{2020}).
+We then visualise the raw data.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# read and plot example data}
+\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/atlas1060\_allTrials\_annotated.csv"}\NormalTok{)}
+\NormalTok{data\_raw <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_calibration_raw.png}
+
+\hypertarget{filter-by-bounding-box}{%
+\subsection{Filter by bounding box}\label{filter-by-bounding-box}}
+
+We first save a copy of the data, so that we can plot the raw data with the cleaned data plotted over it for comparison.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# make a copy using the data.table copy function}
+\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+\end{Highlighting}
+\end{Shaded}
+
+We then filter by a bounding box in order to remove the point outlier to the far south east of the main track.
+We use the \texttt{atl\_filter\_bounds} functions using the \texttt{x\_range} argument, to which we pass the limit in the UTM 31N coordinate reference system.
+This limit is used to exclude all points with an X coordinate \textless{} 645,000.
+
+We then plot the result of filtering, with the excluded point in black, and the points that are retained in green.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# remove inside must be set to falses}
+\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_filter\_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data, }
+                          \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
+                          \DataTypeTok{x\_range =} \KeywordTok{c}\NormalTok{(}\DecValTok{645000}\NormalTok{, }\KeywordTok{max}\NormalTok{(data}\OperatorTok{$}\NormalTok{x)), }
+                          \DataTypeTok{remove\_inside =} \OtherTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{filter-trajectories}{%
+\subsection{Filter trajectories}\label{filter-trajectories}}
+
+\hypertarget{handle-time}{%
+\subsubsection{Handle time}\label{handle-time}}
+
+Time in ATLAS tracking is counted in milliseconds and is represented by a 64-bit integer (type \texttt{long}), which is not natively supported in R; it will instead be converted to a \texttt{numeric}, or \texttt{double}.
+
+This is not what is intended, but it works. The \texttt{bit64} package can help handle 64-bit integers if you want to keep to intended type.
+
+A further issue is that 64-bit integers (whether represented as \texttt{bit64} or \texttt{double}) do not yield meaninful results when you try to convert them to a date-time object, such as of the class \texttt{POSIXct}.
+
+This is because \texttt{as.POSIXct} fails when trying to work with 64-bit integers (it cannot interpret this type), and returns a date many thousands of years in the future (approx.
+52,000 CE) if the time column is converted to \texttt{numeric}.
+
+There are two possible solutions. The parsimonious one is to convert the 64-bit number to a 32-bit short integer (dividing by 1000), or to use the \texttt{nanotime} package.
+
+The conversion method loses an imperceptible amount of precision. The \texttt{nanotime} requires installing another package. The first method is shown here.
+
+In the spirit of not destroying data, we create a second lower-case column called \texttt{time}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# divide by 1000, convert to integer, then convert to POSIXct}
+\NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{add-speed-and-turning-angle}{%
+\subsubsection{Add speed and turning angle}\label{add-speed-and-turning-angle}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# add incoming and outgoing speed}
+\NormalTok{data[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{ (}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }
+                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
+                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
+                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
+             \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
+
+\CommentTok{\# add turning angle}
+\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_turning\_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{get-95th-percentile-of-speed-and-angle}{%
+\subsubsection{Get 95th percentile of speed and angle}\label{get-95th-percentile-of-speed-and-angle}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# use sapply}
+\NormalTok{speed\_angle\_thresholds <{-}}\StringTok{ }
+\StringTok{  }\KeywordTok{sapply}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(speed\_in, speed\_out, angle)], }
+\NormalTok{       quantile, }\DataTypeTok{probs =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{na.rm =}\NormalTok{ T)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{filter-on-speed}{%
+\subsubsection{Filter on speed}\label{filter-on-speed}}
+
+Here we use a speed threshold of 15 m/s, the fastest known boat speed.
+We then plot the data with the extreme speeds shown in grey, and the positions retained shown in green.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# make a copy}
+\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{\# remove speed outliers}
+\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+            \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(speed\_in < 15 \& speed\_out < 15)"}\NormalTok{))}
+
+\CommentTok{\# recalculate speed and angle}
+\NormalTok{data[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{ (}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }
+                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
+                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
+                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
+             \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
+
+\CommentTok{\# add turning angle}
+\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_turning\_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_speed_outlier.png}
+
+\hypertarget{smoothing-the-trajectory}{%
+\subsection{Smoothing the trajectory}\label{smoothing-the-trajectory}}
+
+We then apply a median smooth over a moving window (\(K\) = 5).
+This function modifies in place, and does not need to be assigned to a new variable.
+We create a copy of the data before applying the smooth so that we can compare the data before and after smoothin.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# apply a 5 point median smooth, first make a copy}
+\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{\# now apply the smooth}
+\KeywordTok{atl\_median\_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+                  \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }\DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
+                  \DataTypeTok{moving\_window =} \DecValTok{5}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_calib_median_smooth.png}
+
+\hypertarget{thinning-the-data}{%
+\subsection{Thinning the data}\label{thinning-the-data}}
+
+Next we thin the data to demonstrate thinning by median smoothing.
+Following this, we plot the median smooth and thinning by aggregation.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# save a copy}
+\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{\# remove columns we don\textquotesingle{}t need}
+\NormalTok{data <{-}}\StringTok{ }\NormalTok{data[, }\KeywordTok{setdiff}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(data), }
+                       \KeywordTok{c}\NormalTok{(}\StringTok{"tID"}\NormalTok{, }\StringTok{"Timestamp"}\NormalTok{, }\StringTok{"id"}\NormalTok{, }\StringTok{"TIME"}\NormalTok{, }\StringTok{"UTCtime"}\NormalTok{)), }
+\NormalTok{             with =}\StringTok{ }\OtherTok{FALSE}\NormalTok{]}
+
+\CommentTok{\# thin to a 30s interval}
+\NormalTok{data\_thin <{-}}\StringTok{ }\KeywordTok{atl\_thin\_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+                           \DataTypeTok{interval =} \DecValTok{30}\NormalTok{,}
+                           \DataTypeTok{method =} \StringTok{"aggregate"}\NormalTok{,}
+                           \DataTypeTok{id\_columns =} \StringTok{"TAG"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_calib_smooth_thin.png}
+
+\hypertarget{residence-patches}{%
+\subsection{Residence patches}\label{residence-patches}}
+
+\hypertarget{get-waypoint-centroids}{%
+\subsubsection{Get waypoint centroids}\label{get-waypoint-centroids}}
+
+We subset the annotated calibration data to select the waypoints and the positions around them which are supposed to be the locations of known stops.
+Since each stop was supposed to be 5 minutes long, there are multiple points in each known stop.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(stringi)}
+\NormalTok{data\_res <{-}}\StringTok{ }\NormalTok{data\_unproc[}\KeywordTok{stri\_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{)]}
+\end{Highlighting}
+\end{Shaded}
+
+From this data, we get the centroid of known stops, and determine the time difference between the first and last point within 50 metres, and within 10 minutes of the waypoint positions' median time.
+
+Essentially, this means that the maximum duration of a stop can be 20 minutes, and stops above this duration are not expected.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# get centroid}
+\NormalTok{data\_res\_summary <{-}}\StringTok{ }\NormalTok{data\_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
+                                    \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{)),}
+\NormalTok{                                    t\_median =}\StringTok{ }\KeywordTok{median}\NormalTok{(time)}\ErrorTok{)}\NormalTok{,}
+\NormalTok{                             by =}\StringTok{ "tID"}\NormalTok{]}
+
+\CommentTok{\# now get times 10 mins before and after}
+\NormalTok{data\_res\_summary[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{t\_min =}\NormalTok{ t\_median }\OperatorTok{{-}}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{),}
+                        \DataTypeTok{t\_max =}\NormalTok{ t\_median }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{))]}
+
+\CommentTok{\# make a list of positions 10min before and after}
+\NormalTok{wp\_data <{-}}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(l, u, mx, my) \{}
+\NormalTok{  tmp\_data <{-}}\StringTok{ }\NormalTok{data\_unproc[}\KeywordTok{inrange}\NormalTok{(time, l, u)]}
+\NormalTok{  tmp\_data[, distance }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{sqrt}\NormalTok{((mx }\OperatorTok{{-}}\StringTok{ }\NormalTok{x)}\OperatorTok{\^{}}\DecValTok{2} \OperatorTok{+}\StringTok{ }\NormalTok{(my }\OperatorTok{{-}}\StringTok{ }\NormalTok{y)}\OperatorTok{\^{}}\DecValTok{2}\NormalTok{)]}
+
+  \CommentTok{\# keep within 50}
+\NormalTok{  tmp\_data <{-}}\StringTok{ }\NormalTok{tmp\_data[distance }\OperatorTok{<=}\StringTok{ }\DecValTok{50}\NormalTok{, ]}
+
+  \CommentTok{\# get duration}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(tmp\_data}\OperatorTok{$}\NormalTok{time)))}
+
+\NormalTok{\}, data\_res\_summary}\OperatorTok{$}\NormalTok{t\_min, data\_res\_summary}\OperatorTok{$}\NormalTok{t\_max,}
+\NormalTok{   data\_res\_summary}\OperatorTok{$}\NormalTok{x\_median, data\_res\_summary}\OperatorTok{$}\NormalTok{y\_median,}
+\DataTypeTok{SIMPLIFY =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{prepare-data}{%
+\subsubsection{Prepare data}\label{prepare-data}}
+
+An indicator of individual residence at or near a position can be useful when attempting to identify residence patches.
+Positions can be filtered on a metric such as residence time (Bracis, Bildstein, and Mueller \protect\hyperlink{ref-bracis2018}{2018}).
+
+\hypertarget{calculate-residence-time}{%
+\subsubsection{Calculate residence time}\label{calculate-residence-time}}
+
+First we calculate the residence time with a radius of 50 metres.
+For this, we need a dataframe with coordinates, the timestamp, and the animal id.
+We save this data to file for later use.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# load recurse}
+\KeywordTok{library}\NormalTok{(recurse)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# get 4 column data}
+\NormalTok{data\_for\_patch <{-}}\StringTok{ }\NormalTok{data\_thin[, }\KeywordTok{list}\NormalTok{(x, y, time, TAG)]}
+
+\CommentTok{\# get recurse data for a 10m radius}
+\NormalTok{recurse\_stats <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data\_for\_patch,}
+                               \DataTypeTok{radius =} \DecValTok{50}\NormalTok{, }\DataTypeTok{timeunits =} \StringTok{"mins"}\NormalTok{)}
+
+\CommentTok{\# assign to recurse data}
+\NormalTok{data\_for\_patch[, res\_time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{recurse\_stats}\OperatorTok{$}\NormalTok{residenceTime]}
+
+\CommentTok{\# save recurse data}
+\KeywordTok{fwrite}\NormalTok{(data\_for\_patch, }\DataTypeTok{file =} \StringTok{"data/data\_calib\_for\_patch.csv"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{run-residence-patch-method}{%
+\subsubsection{Run residence patch method}\label{run-residence-patch-method}}
+
+We subset data with a residence time \textgreater{} 5 minutes in order to construct residence patches.
+From this subset, we construct residence patches using the parameters: \texttt{buffer\_radius} = 5 metres, \texttt{lim\_spat\_indep} = 50 metres, \texttt{lim\_time\_indep} = 5 minutes, and \texttt{min\_fixes} = 3.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# assign id as tag}
+\NormalTok{data\_for\_patch[, id }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.character}\NormalTok{(TAG)]}
+
+\CommentTok{\# on known residence points}
+\NormalTok{patch\_res\_known <{-}}\StringTok{ }\KeywordTok{atl\_res\_patch}\NormalTok{(data\_for\_patch[res\_time }\OperatorTok{>=}\StringTok{ }\DecValTok{5}\NormalTok{, ], }
+                                \DataTypeTok{buffer\_radius =} \DecValTok{5}\NormalTok{,}
+                                \DataTypeTok{lim\_spat\_indep =} \DecValTok{50}\NormalTok{,}
+                                \DataTypeTok{lim\_time\_indep =} \DecValTok{5}\NormalTok{,}
+                                \DataTypeTok{min\_fixes =} \DecValTok{3}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{get-spatial-and-summary-objects}{%
+\subsubsection{Get spatial and summary objects}\label{get-spatial-and-summary-objects}}
+
+We get spatial and summary ouput of the residence patch method using the \texttt{atl\_patch\_summary} function using the options \texttt{which\_data} = ``spatial'' and \texttt{which\_data} = "summary.
+We use a buffer radius here of 20 metres for the spatial buffer, despite using a buffer radius of 5 metres earlier, simply because it is easier to visualise in the output figure.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# for the known and unkniwn patches}
+\NormalTok{patch\_sf\_data <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(patch\_res\_known, }
+                                   \DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{,}
+                                   \DataTypeTok{buffer\_radius =} \DecValTok{20}\NormalTok{)}
+
+\CommentTok{\# assign crs}
+\NormalTok{sf}\OperatorTok{::}\KeywordTok{st\_crs}\NormalTok{(patch\_sf\_data) <{-}}\StringTok{ }\DecValTok{32631}
+
+\CommentTok{\# get summary data}
+\NormalTok{patch\_summary\_data <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(patch\_res\_known, }
+                                        \DataTypeTok{which\_data =} \StringTok{"summary"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{prepare-to-plot-data}{%
+\subsubsection{Prepare to plot data}\label{prepare-to-plot-data}}
+
+We read in the island's shapefile to plot it as a background for the residence patch figure.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# read griend}
+\NormalTok{griend <{-}}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st\_read}\NormalTok{(}\StringTok{"data/griend\_polygon/griend\_polygon.shp"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{compare-patch-metrics}{%
+\subsection{Compare patch metrics}\label{compare-patch-metrics}}
+
+We then merge the annoated, known stop data with the calculated patch duration.
+We filter this data to exclude one exceedingly long outlier of about an hour (WP080), which how
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# get known patch summary}
+\NormalTok{data\_res <{-}}\StringTok{ }\NormalTok{data\_unproc[stringi}\OperatorTok{::}\KeywordTok{stri\_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{), ]}
+
+\CommentTok{\# get waypoint summary}
+\NormalTok{patch\_summary\_real <{-}}\StringTok{ }\NormalTok{data\_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{nfixes\_real =}\NormalTok{ .N,}
+                                      \DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
+                                      \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{)), }
+\NormalTok{                               by =}\StringTok{ "tID"}\NormalTok{]}
+
+\CommentTok{\# add real duration}
+\NormalTok{patch\_summary\_real[, duration\_real }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{wp\_data]}
+
+\CommentTok{\# round median coordinate for inferred patches}
+\NormalTok{patch\_summary\_inferred <{-}}\StringTok{ }
+\StringTok{  }\NormalTok{patch\_summary\_data[, }
+                     \KeywordTok{c}\NormalTok{(}\StringTok{"x\_median"}\NormalTok{, }\StringTok{"y\_median"}\NormalTok{, }
+                       \StringTok{"nfixes"}\NormalTok{, }\StringTok{"duration"}\NormalTok{, }\StringTok{"patch"}\NormalTok{)}
+\NormalTok{                     ][, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(x\_median, }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
+                              \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(y\_median, }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{))]}
+
+\CommentTok{\# join with respatch summary}
+\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }
+\StringTok{  }\KeywordTok{merge}\NormalTok{(patch\_summary\_real,}
+\NormalTok{        patch\_summary\_inferred, }
+        \DataTypeTok{on =} \KeywordTok{c}\NormalTok{(}\StringTok{"x\_median"}\NormalTok{, }\StringTok{"y\_median"}\NormalTok{),}
+        \DataTypeTok{all.x =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{all.y =} \OtherTok{TRUE}\NormalTok{)}
+
+\CommentTok{\# drop nas}
+\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(patch\_summary\_compare)}
+
+\CommentTok{\# drop patch around WP080}
+\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }\NormalTok{patch\_summary\_compare[tID }\OperatorTok{!=}\StringTok{ "WP080"}\NormalTok{, ]}
+\end{Highlighting}
+\end{Shaded}
+
+7 patches are identified where there are no waypoints, while 2 waypoints are not identified as patches.
+These waypoints consisted of 6 and 15 (WP098 and WP092) positions respectively, and were lost when the data were aggregated to 30 second intervals.
+
+\hypertarget{linear-model-durations}{%
+\subsubsection{Linear model durations}\label{linear-model-durations}}
+
+We run a simple linear model.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# get linear model}
+\NormalTok{model\_duration <{-}}\StringTok{ }\KeywordTok{lm}\NormalTok{(duration\_real }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{duration,}
+                     \DataTypeTok{data =}\NormalTok{ patch\_summary\_compare)}
+
+\CommentTok{\# get R2}
+\KeywordTok{summary}\NormalTok{(model\_duration)}
+
+\CommentTok{\# write to file}
+\KeywordTok{writeLines}\NormalTok{(}
+  \DataTypeTok{text =} \KeywordTok{capture.output}\NormalTok{(}
+    \KeywordTok{summary}\NormalTok{(model\_duration)}
+\NormalTok{  ),}
+  \DataTypeTok{con =} \StringTok{"data/model\_output\_residence\_patch.txt"}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{figures/fig_calib_lm_duration.png}
+
 \hypertarget{processing-egyptian-fruit-bat-tracks}{%
@@
 
-Install the required \texttt{R} libraries that are available on CRAN.
+Install the required \texttt{R} libraries that are required from CRAN if not already installed.
 
@@
 \begin{cslreferences}
+\leavevmode\hypertarget{ref-barraquand2008}{}%
+Barraquand, Frédéric, and Simon Benhamou.
+2008.
+``Animal Movements in Heterogeneous Landscapes: Identifying Profitable Places and Homogeneous Movement Bouts.'' \emph{Ecology} 89 (12): 3336--48.
+\url{https://doi.org/10.1890/08-0162.1}.
+
+\leavevmode\hypertarget{ref-bijleveld2016}{}%
+Bijleveld, Allert Imre, Robert B MacCurdy, Ying-Chi Chan, Emma Penning, Richard M.
+Gabrielson, John Cluderay, Erik L.
+Spaulding, et al.
+2016.
+``Understanding Spatial Distributions: Negative Density-Dependence in Prey Causes Predators to Trade-Off Prey Quantity with Quality.'' \emph{Proceedings of the Royal Society B: Biological Sciences} 283 (1828): 20151557.
+\url{https://doi.org/10.1098/rspb.2015.1557}.
+
 \leavevmode\hypertarget{ref-bracis2018}{}%

### 2020-11-23 docs/supplementary_material.tex
add supplement pdf and tex

@@
   pdftitle={Supplementary Material for: A Rough Guide to Pre-processing High-Frequency Animal Tracking Data},
-  pdfauthor={Pratik R. Gupte; Christine E. Beardsworth; Orr Spiegel; Emmanuel Lourie; Allert Bijleveld},
+  pdfauthor={Pratik R. Gupte; Christine E. Beardsworth; Orr Spiegel; Emmanuel Lourie; Allert I. Bijleveld},
   hidelinks,
@@
 \title{Supplementary Material for: A Rough Guide to Pre-processing High-Frequency Animal Tracking Data}
-\author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Allert Bijleveld}
+\author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Allert I. Bijleveld}
 \date{2020-11-23}
@@
 
+\begin{figure}
+\centering
 \includegraphics{figures/fig_calibration_raw.png}
+\caption{The raw data from a calibration exercise conducted around the island of Griend in the Dutch Wadden Sea.
+A handheld WATLAS tag was used to examine how ATLAS data compared to GPS tracks, and we use the WATLAS data here to demonstrate the basics of the pre-processing pipeline, as well as validate the residence patch method.
+It is immediately clear from the figure that the track shows location errors, both in the form of point outliers as well as small-scale errors around the true location.}
+\end{figure}
 
@@
 \end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{figures/fig_calib_bbox.png}
+\caption{Removal of a point outlier using the function \texttt{atl\_filter\_bounds}.
+The point outlier (black point) is removed based on its X coordinate value, with the data filtered to exclude positions with an X coordinate \textless{} 645,000 in the UTM 31N system.
+Positions that are retained are shown in green.}
+\end{figure}
 
@@
 
+\begin{figure}
+\centering
 \includegraphics{figures/fig_speed_outlier.png}
+\caption{Improving data quality by filtering out positions that would require unrealistic movement.
+We removed positions with speeds \(\geq\) 15 m/s, which is the fastest possible speed in this calibration data, part of which was collected in a moving boat around Griend.
+Grey positions are removed, while green positions are retained.
+Rectangles indicate areas expanded for visualisation in following figures.}
+\end{figure}
 
@@
 
+\begin{figure}
+\centering
 \includegraphics{figures/fig_calib_median_smooth.png}
+\caption{Reducing small-scale location error using a median smooth with a moving window \(K\) = 5.
+Median smoothed positions are shown in green, while raw, unfiltered data is shown in grey.
+Median smoothing successfully recovers the likely path of the track without a loss of data.
+The area shown is the upper rectangle from Figure 3.}
+\end{figure}
 
@@
 
+\begin{figure}
+\centering
 \includegraphics{figures/fig_calib_smooth_thin.png}
+\caption{Thinning by aggregation over a 30 second interval (down from 1 second) preserves track structure while reducing the data volume for computation.
+Here, thinned positions are shown as purple squares, with the size of the square indicating the number of positions within the 30 second bin used to obtain the average position.
+Green points show the median smoothed data from Figure 4, while the raw data are shown in grey.}
+\end{figure}
 
@@
 \end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{figures/fig_calib_residence.png}
+\caption{Classifying thinned data into residence patches yields robust estimates of the duration of known stops.
+The island of Griend (53.25\(^{\circ}\)N, 5.25\(^{\circ}\)E) is shown in beige.
+Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
+However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without crossed-squares).
+The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).}
+\end{figure}
 
@@
 
+\begin{figure}
+\centering
 \includegraphics{figures/fig_calib_lm_duration.png}
+\caption{The inferred duration of residence patches corresponds very closely to the real duration (grey circles, red line shows linear model fit), with an underestimation of the true duration of around 2\%.
+The dashed black line represents \(y = x\) for reference.}
+\end{figure}
+
+\hypertarget{linear-model-summary}{%
+\subsubsection{Linear model summary}\label{linear-model-summary}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{cat}\NormalTok{(}
+  \KeywordTok{readLines}\NormalTok{(}
+    \DataTypeTok{con =} \StringTok{"data/model\_output\_residence\_patch.txt"}
+\NormalTok{  ), }\DataTypeTok{sep =} \StringTok{"}\CharTok{\textbackslash{}n}\StringTok{"}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 

### 2020-11-23 docs/supplementary_material.tex
minor changes to supplement

@@
 Here, thinned positions are shown as purple squares, with the size of the square indicating the number of positions within the 30 second bin used to obtain the average position.
-Green points show the median smoothed data from Figure 4, while the raw data are shown in grey.}
+Green points show the median smoothed data from Figure 4, while the raw data are shown in grey.
+The area shown is the upper rectangle in Figure 3.}
 \end{figure}
@@
 However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without crossed-squares).
-The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).}
+The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).
+The area shown is the lower rectangle in Figure 3.}
 \end{figure}

### 2020-11-23 05_calibration_data.Rmd

@@
 Here, thinned positions are shown as purple squares, with the size of the square indicating the number of positions within the 30 second bin used to obtain the average position.
-Green points show the median smoothed data from Figure 4, while the raw data are shown in grey.](figures/fig_calib_smooth_thin.png)
+Green points show the median smoothed data from Figure 4, while the raw data are shown in grey.
+The area shown is the upper rectangle in Figure 3.](figures/fig_calib_smooth_thin.png)
 
@@
 However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without crossed-squares).
-The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).](figures/fig_calib_residence.png)
+The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).
+The area shown is the lower rectangle in Figure 3.](figures/fig_calib_residence.png)
 

### 2020-11-23 index.rmd

--- 2020-11-16 index.rmd
@@
     - Emmanuel Lourie
-    - Allert Bijleveld
+    - Allert I. Bijleveld
 date: "`r Sys.Date()`"
 output: bookdown::pdf_document2
-documentclass: article
+documentclass: scrartcl
 toc-depth: 2

### 2020-11-23 docs/supplementary_material.tex
conditional fonts

@@
 \usepackage{lineno}
-\setmainfont{Times New Roman}
-\setsansfont{Arial}
+
+\IfFontExistsTF{Times New Roman}{\setmainfont[]{Times New Roman}}{} 
+\IfFontExistsTF{Arial}{\setsansfont[]{Arial}}{}
+\IfFontExistsTF{Fira Code}{\setmonofont{Fira Code}}
 
@@
   \KeywordTok{readLines}\NormalTok{(}
-    \DataTypeTok{con =} \StringTok{"data/model\_output\_residence\_patch.txt"}
+    \DataTypeTok{con =} \StringTok{"data/model\_output\_residence\_patch.txt"}\NormalTok{,}
+    \DataTypeTok{encoding =} \StringTok{"UTF{-}8"}
 \NormalTok{  ), }\DataTypeTok{sep =} \StringTok{"}\CharTok{\textbackslash{}n}\StringTok{"}
 \NormalTok{)}
+\CommentTok{\#> }
+\CommentTok{\#> Call:}
+\CommentTok{\#> lm(formula = duration\_real \textasciitilde{} duration, data = patch\_summary\_compare)}
+\CommentTok{\#> }
+\CommentTok{\#> Residuals:}
+\CommentTok{\#>      Min       1Q   Median       3Q      Max }
+\CommentTok{\#> {-}103.237  {-}19.277   {-}2.917    7.003   93.431 }
+\CommentTok{\#> }
+\CommentTok{\#> Coefficients:}
+\CommentTok{\#>              Estimate Std. Error t value Pr(>|t|)    }
+\CommentTok{\#> (Intercept) 101.42061   47.66936   2.128   0.0493 *  }
+\CommentTok{\#> duration      1.02108    0.07876  12.965 6.66e{-}10 ***}
+\CommentTok{\#> {-}{-}{-}}
+\CommentTok{\#> Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1}
+\CommentTok{\#> }
+\CommentTok{\#> Residual standard error: 50.35 on 16 degrees of freedom}
+\CommentTok{\#> Multiple R{-}squared:  0.9131,	Adjusted R{-}squared:  0.9077 }
+\CommentTok{\#> F{-}statistic: 168.1 on 1 and 16 DF,  p{-}value: 6.655e{-}10}
 \end{Highlighting}

### 2020-11-23 docs/supplementary_material.tex
minor changes

--- 2020-11-23 docs/supplementary_material.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
 \PassOptionsToPackage{hyphens}{url}
 %
-\documentclass[
-]{scrartcl}
+\documentclass[]{scrartcl}
 \usepackage{lmodern}
@@
             pdfauthor={Pratik R. Gupte; Christine E. Beardsworth; Orr Spiegel; Emmanuel Lourie; Allert I. Bijleveld},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage[left=4cm, right=3cm, top=2.5cm, bottom=2.5cm]{geometry}
@@
 \usepackage{lineno}
-\setmainfont{Times New Roman}
-\setsansfont{Arial}
+
+\IfFontExistsTF{Times New Roman}{\setmainfont[]{Times New Roman}}{} 
+\IfFontExistsTF{Arial}{\setsansfont[]{Arial}}{}
+\IfFontExistsTF{Fira Code}{\setmonofont{Fira Code}}
 
 \linenumbers
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{cslreferences}%
-  {\setlength{\parindent}{0pt}%
-  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
-  {\par}
 
@@
 \begin{Highlighting}[]
-\CommentTok{\# read and plot example data}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/atlas1060\_allTrials\_annotated.csv"}\NormalTok{)}
-\NormalTok{data\_raw <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+\CommentTok{# read and plot example data}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{fread}\NormalTok{(}\StringTok{"data/atlas1060_allTrials_annotated.csv"}\NormalTok{)}
+\NormalTok{data_raw <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# make a copy using the data.table copy function}
-\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+\CommentTok{# make a copy using the data.table copy function}
+\NormalTok{data_unproc <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# remove inside must be set to falses}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_filter\_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data, }
+\CommentTok{# remove inside must be set to falses}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{atl_filter_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data, }
                           \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
-                          \DataTypeTok{x\_range =} \KeywordTok{c}\NormalTok{(}\DecValTok{645000}\NormalTok{, }\KeywordTok{max}\NormalTok{(data}\OperatorTok{$}\NormalTok{x)), }
-                          \DataTypeTok{remove\_inside =} \OtherTok{FALSE}\NormalTok{)}
+                          \DataTypeTok{x_range =} \KeywordTok{c}\NormalTok{(}\DecValTok{645000}\NormalTok{, }\KeywordTok{max}\NormalTok{(data}\OperatorTok{$}\NormalTok{x)), }
+                          \DataTypeTok{remove_inside =} \OtherTok{FALSE}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# divide by 1000, convert to integer, then convert to POSIXct}
+\CommentTok{# divide by 1000, convert to integer, then convert to POSIXct}
 \NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}
@@
 \begin{Highlighting}[]
-\CommentTok{\# add incoming and outgoing speed}
-\NormalTok{data[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{ (}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }
+\CommentTok{# add incoming and outgoing speed}
+\NormalTok{data[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{ (}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(data, }
                                       \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
@@
                                       \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
-             \DataTypeTok{speed\_out =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
-
-\CommentTok{\# add turning angle}
-\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl\_turning\_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
+             \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
+
+\CommentTok{# add turning angle}
+\NormalTok{data[, angle }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{atl_turning_angle}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data)]}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# use sapply}
-\NormalTok{speed\_angle\_thresholds <{-}}\StringTok{ }
-\StringTok{  }\KeywordTok{sapply}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(speed\_in, speed\_out, angle)], }
+\CommentTok{# use sapply}
+\NormalTok{speed_angle_thresholds <-}\StringTok{ }
+\StringTok{  }\KeywordTok{sapply}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(speed_in, speed_out, angle)], }
 \NormalTok{       quantile, }\DataTypeTok{probs =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{na.rm =}\NormalTok{ T)}
@@
 \begin{Highlighting}[]
-\CommentTok{\# make a copy}
-\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-
-\CommentTok{\# remove speed outliers}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
-            \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(speed\_in < 15 \& speed\_out < 15)"}\NormalTok{))}
-
-\CommentTok{\# recalculate speed and angle}
-\NormalTok{data[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{ (}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(data, }
+\CommentTok{# make a copy}
+\NormalTok{data_unproc <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{# remove speed outliers}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+            \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(speed_in < 15 & speed_out < 15)"}\NormalTok{))}
+
+\CommentTok{# recalculate speed and angle}
+\NormalTok{data[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{ (}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(data, }
                                       \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
@@
 \begin{Highlighting}[]
-\CommentTok{\# apply a 5 point median smooth, first make a copy}
-\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-
-\CommentTok{\# now apply the smooth}
-\KeywordTok{atl\_median\_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+\CommentTok{# apply a 5 point median smooth, first make a copy}
+\NormalTok{data_unproc <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{# now apply the smooth}
+\KeywordTok{atl_median_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
                   \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }\DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
-                  \DataTypeTok{moving\_window =} \DecValTok{5}\NormalTok{)}
+                  \DataTypeTok{moving_window =} \DecValTok{5}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# save a copy}
-\NormalTok{data\_unproc <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-
-\CommentTok{\# remove columns we don\textquotesingle{}t need}
-\NormalTok{data <{-}}\StringTok{ }\NormalTok{data[, }\KeywordTok{setdiff}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(data), }
+\CommentTok{# save a copy}
+\NormalTok{data_unproc <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{# remove columns we don't need}
+\NormalTok{data <-}\StringTok{ }\NormalTok{data[, }\KeywordTok{setdiff}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(data), }
                        \KeywordTok{c}\NormalTok{(}\StringTok{"tID"}\NormalTok{, }\StringTok{"Timestamp"}\NormalTok{, }\StringTok{"id"}\NormalTok{, }\StringTok{"TIME"}\NormalTok{, }\StringTok{"UTCtime"}\NormalTok{)), }
@@
 
-\CommentTok{\# thin to a 30s interval}
-\NormalTok{data\_thin <{-}}\StringTok{ }\KeywordTok{atl\_thin\_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+\CommentTok{# thin to a 30s interval}
+\NormalTok{data_thin <-}\StringTok{ }\KeywordTok{atl_thin_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
                            \DataTypeTok{interval =} \DecValTok{30}\NormalTok{,}
                            \DataTypeTok{method =} \StringTok{"aggregate"}\NormalTok{,}
-                           \DataTypeTok{id\_columns =} \StringTok{"TAG"}\NormalTok{)}
+                           \DataTypeTok{id_columns =} \StringTok{"TAG"}\NormalTok{)}
 \end{Highlighting}
@@
 \KeywordTok{library}\NormalTok{(stringi)}
-\NormalTok{data\_res <{-}}\StringTok{ }\NormalTok{data\_unproc[}\KeywordTok{stri\_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{)]}
+\NormalTok{data_res <-}\StringTok{ }\NormalTok{data_unproc[}\KeywordTok{stri_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{)]}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get centroid}
-\NormalTok{data\_res\_summary <{-}}\StringTok{ }\NormalTok{data\_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
-                                    \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{)),}
-\NormalTok{                                    t\_median =}\StringTok{ }\KeywordTok{median}\NormalTok{(time)}\ErrorTok{)}\NormalTok{,}
+\CommentTok{# get centroid}
+\NormalTok{data_res_summary <-}\StringTok{ }\NormalTok{data_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{x_median =} \KeywordTok{median}\NormalTok{(x),}
+                                    \DataTypeTok{y_median =} \KeywordTok{median}\NormalTok{(y),}
+                                    \DataTypeTok{t_median =} \KeywordTok{median}\NormalTok{(time)),}
 \NormalTok{                             by =}\StringTok{ "tID"}\NormalTok{]}
 
-\CommentTok{\# now get times 10 mins before and after}
-\NormalTok{data\_res\_summary[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{t\_min =}\NormalTok{ t\_median }\OperatorTok{{-}}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{),}
-                        \DataTypeTok{t\_max =}\NormalTok{ t\_median }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{))]}
-
-\CommentTok{\# make a list of positions 10min before and after}
-\NormalTok{wp\_data <{-}}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(l, u, mx, my) \{}
-\NormalTok{  tmp\_data <{-}}\StringTok{ }\NormalTok{data\_unproc[}\KeywordTok{inrange}\NormalTok{(time, l, u)]}
-\NormalTok{  tmp\_data[, distance }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{sqrt}\NormalTok{((mx }\OperatorTok{{-}}\StringTok{ }\NormalTok{x)}\OperatorTok{\^{}}\DecValTok{2} \OperatorTok{+}\StringTok{ }\NormalTok{(my }\OperatorTok{{-}}\StringTok{ }\NormalTok{y)}\OperatorTok{\^{}}\DecValTok{2}\NormalTok{)]}
-
-  \CommentTok{\# keep within 50}
-\NormalTok{  tmp\_data <{-}}\StringTok{ }\NormalTok{tmp\_data[distance }\OperatorTok{<=}\StringTok{ }\DecValTok{50}\NormalTok{, ]}
-
-  \CommentTok{\# get duration}
-  \KeywordTok{return}\NormalTok{(}\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(tmp\_data}\OperatorTok{$}\NormalTok{time)))}
-
-\NormalTok{\}, data\_res\_summary}\OperatorTok{$}\NormalTok{t\_min, data\_res\_summary}\OperatorTok{$}\NormalTok{t\_max,}
-\NormalTok{   data\_res\_summary}\OperatorTok{$}\NormalTok{x\_median, data\_res\_summary}\OperatorTok{$}\NormalTok{y\_median,}
+\CommentTok{# now get times 10 mins before and after}
+\NormalTok{data_res_summary[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{t_min =}\NormalTok{ t_median }\OperatorTok{-}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{),}
+                        \DataTypeTok{t_max =}\NormalTok{ t_median }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{))]}
+
+\CommentTok{# make a list of positions 10min before and after}
+\NormalTok{wp_data <-}\StringTok{ }\KeywordTok{mapply}\NormalTok{(}\ControlFlowTok{function}\NormalTok{(l, u, mx, my) \{}
+\NormalTok{  tmp_data <-}\StringTok{ }\NormalTok{data_unproc[}\KeywordTok{inrange}\NormalTok{(time, l, u)]}
+\NormalTok{  tmp_data[, distance }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{sqrt}\NormalTok{((mx }\OperatorTok{-}\StringTok{ }\NormalTok{x)}\OperatorTok{^}\DecValTok{2} \OperatorTok{+}\StringTok{ }\NormalTok{(my }\OperatorTok{-}\StringTok{ }\NormalTok{y)}\OperatorTok{^}\DecValTok{2}\NormalTok{)]}
+
+  \CommentTok{# keep within 50}
+\NormalTok{  tmp_data <-}\StringTok{ }\NormalTok{tmp_data[distance }\OperatorTok{<=}\StringTok{ }\DecValTok{50}\NormalTok{, ]}
+
+  \CommentTok{# get duration}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(tmp_data}\OperatorTok{$}\NormalTok{time)))}
+
+\NormalTok{\}, data_res_summary}\OperatorTok{$}\NormalTok{t_min, data_res_summary}\OperatorTok{$}\NormalTok{t_max,}
+\NormalTok{   data_res_summary}\OperatorTok{$}\NormalTok{x_median, data_res_summary}\OperatorTok{$}\NormalTok{y_median,}
 \DataTypeTok{SIMPLIFY =} \OtherTok{TRUE}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\CommentTok{\# load recurse}
+\CommentTok{# load recurse}
 \KeywordTok{library}\NormalTok{(recurse)}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get 4 column data}
-\NormalTok{data\_for\_patch <{-}}\StringTok{ }\NormalTok{data\_thin[, }\KeywordTok{list}\NormalTok{(x, y, time, TAG)]}
-
-\CommentTok{\# get recurse data for a 10m radius}
-\NormalTok{recurse\_stats <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data\_for\_patch,}
+\CommentTok{# get 4 column data}
+\NormalTok{data_for_patch <-}\StringTok{ }\NormalTok{data_thin[, }\KeywordTok{list}\NormalTok{(x, y, time, TAG)]}
+
+\CommentTok{# get recurse data for a 10m radius}
+\NormalTok{recurse_stats <-}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data_for_patch,}
                                \DataTypeTok{radius =} \DecValTok{50}\NormalTok{, }\DataTypeTok{timeunits =} \StringTok{"mins"}\NormalTok{)}
 
-\CommentTok{\# assign to recurse data}
-\NormalTok{data\_for\_patch[, res\_time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{recurse\_stats}\OperatorTok{$}\NormalTok{residenceTime]}
-
-\CommentTok{\# save recurse data}
-\KeywordTok{fwrite}\NormalTok{(data\_for\_patch, }\DataTypeTok{file =} \StringTok{"data/data\_calib\_for\_patch.csv"}\NormalTok{)}
+\CommentTok{# assign to recurse data}
+\NormalTok{data_for_patch[, res_time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{recurse_stats}\OperatorTok{$}\NormalTok{residenceTime]}
+
+\CommentTok{# save recurse data}
+\KeywordTok{fwrite}\NormalTok{(data_for_patch, }\DataTypeTok{file =} \StringTok{"data/data_calib_for_patch.csv"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# assign id as tag}
-\NormalTok{data\_for\_patch[, id }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.character}\NormalTok{(TAG)]}
-
-\CommentTok{\# on known residence points}
-\NormalTok{patch\_res\_known <{-}}\StringTok{ }\KeywordTok{atl\_res\_patch}\NormalTok{(data\_for\_patch[res\_time }\OperatorTok{>=}\StringTok{ }\DecValTok{5}\NormalTok{, ], }
-                                \DataTypeTok{buffer\_radius =} \DecValTok{5}\NormalTok{,}
-                                \DataTypeTok{lim\_spat\_indep =} \DecValTok{50}\NormalTok{,}
-                                \DataTypeTok{lim\_time\_indep =} \DecValTok{5}\NormalTok{,}
-                                \DataTypeTok{min\_fixes =} \DecValTok{3}\NormalTok{)}
+\CommentTok{# assign id as tag}
+\NormalTok{data_for_patch[, id }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.character}\NormalTok{(TAG)]}
+
+\CommentTok{# on known residence points}
+\NormalTok{patch_res_known <-}\StringTok{ }\KeywordTok{atl_res_patch}\NormalTok{(data_for_patch[res_time }\OperatorTok{>=}\StringTok{ }\DecValTok{5}\NormalTok{, ], }
+                                \DataTypeTok{buffer_radius =} \DecValTok{5}\NormalTok{,}
+                                \DataTypeTok{lim_spat_indep =} \DecValTok{50}\NormalTok{,}
+                                \DataTypeTok{lim_time_indep =} \DecValTok{5}\NormalTok{,}
+                                \DataTypeTok{min_fixes =} \DecValTok{3}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# for the known and unkniwn patches}
-\NormalTok{patch\_sf\_data <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(patch\_res\_known, }
-                                   \DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{,}
-                                   \DataTypeTok{buffer\_radius =} \DecValTok{20}\NormalTok{)}
-
-\CommentTok{\# assign crs}
-\NormalTok{sf}\OperatorTok{::}\KeywordTok{st\_crs}\NormalTok{(patch\_sf\_data) <{-}}\StringTok{ }\DecValTok{32631}
-
-\CommentTok{\# get summary data}
-\NormalTok{patch\_summary\_data <{-}}\StringTok{ }\KeywordTok{atl\_patch\_summary}\NormalTok{(patch\_res\_known, }
-                                        \DataTypeTok{which\_data =} \StringTok{"summary"}\NormalTok{)}
+\CommentTok{# for the known and unkniwn patches}
+\NormalTok{patch_sf_data <-}\StringTok{ }\KeywordTok{atl_patch_summary}\NormalTok{(patch_res_known, }
+                                   \DataTypeTok{which_data =} \StringTok{"spatial"}\NormalTok{,}
+                                   \DataTypeTok{buffer_radius =} \DecValTok{20}\NormalTok{)}
+
+\CommentTok{# assign crs}
+\NormalTok{sf}\OperatorTok{::}\KeywordTok{st_crs}\NormalTok{(patch_sf_data) <-}\StringTok{ }\DecValTok{32631}
+
+\CommentTok{# get summary data}
+\NormalTok{patch_summary_data <-}\StringTok{ }\KeywordTok{atl_patch_summary}\NormalTok{(patch_res_known, }
+                                        \DataTypeTok{which_data =} \StringTok{"summary"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# read griend}
-\NormalTok{griend <{-}}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st\_read}\NormalTok{(}\StringTok{"data/griend\_polygon/griend\_polygon.shp"}\NormalTok{)}
+\CommentTok{# read griend and hut}
+\NormalTok{griend <-}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st_read}\NormalTok{(}\StringTok{"data/griend_polygon/griend_polygon.shp"}\NormalTok{)}
+\NormalTok{hut <-}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st_read}\NormalTok{(}\StringTok{"data/griend_hut.gpkg"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get known patch summary}
-\NormalTok{data\_res <{-}}\StringTok{ }\NormalTok{data\_unproc[stringi}\OperatorTok{::}\KeywordTok{stri\_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{), ]}
-
-\CommentTok{\# get waypoint summary}
-\NormalTok{patch\_summary\_real <{-}}\StringTok{ }\NormalTok{data\_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{nfixes\_real =}\NormalTok{ .N,}
-                                      \DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
-                                      \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{)), }
+\CommentTok{# get known patch summary}
+\NormalTok{data_res <-}\StringTok{ }\NormalTok{data_unproc[stringi}\OperatorTok{::}\KeywordTok{stri_detect}\NormalTok{(tID, }\DataTypeTok{regex =} \StringTok{"(WP)"}\NormalTok{), ]}
+
+\CommentTok{# get waypoint summary}
+\NormalTok{patch_summary_real <-}\StringTok{ }\NormalTok{data_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{nfixes_real =}\NormalTok{ .N,}
+                                      \DataTypeTok{x_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{),}
+                                      \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{)), }
 \NormalTok{                               by =}\StringTok{ "tID"}\NormalTok{]}
 
-\CommentTok{\# add real duration}
-\NormalTok{patch\_summary\_real[, duration\_real }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{wp\_data]}
-
-\CommentTok{\# round median coordinate for inferred patches}
-\NormalTok{patch\_summary\_inferred <{-}}\StringTok{ }
-\StringTok{  }\NormalTok{patch\_summary\_data[, }
-                     \KeywordTok{c}\NormalTok{(}\StringTok{"x\_median"}\NormalTok{, }\StringTok{"y\_median"}\NormalTok{, }
+\CommentTok{# add real duration}
+\NormalTok{patch_summary_real[, duration_real }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{wp_data]}
+
+\CommentTok{# round median coordinate for inferred patches}
+\NormalTok{patch_summary_inferred <-}\StringTok{ }
+\StringTok{  }\NormalTok{patch_summary_data[, }
+                     \KeywordTok{c}\NormalTok{(}\StringTok{"x_median"}\NormalTok{, }\StringTok{"y_median"}\NormalTok{, }
                        \StringTok{"nfixes"}\NormalTok{, }\StringTok{"duration"}\NormalTok{, }\StringTok{"patch"}\NormalTok{)}
-\NormalTok{                     ][, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{x\_median =} \KeywordTok{round}\NormalTok{(x\_median, }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{),}
-                              \DataTypeTok{y\_median =} \KeywordTok{round}\NormalTok{(y\_median, }\DataTypeTok{digits =} \DecValTok{{-}2}\NormalTok{))]}
-
-\CommentTok{\# join with respatch summary}
-\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }
-\StringTok{  }\KeywordTok{merge}\NormalTok{(patch\_summary\_real,}
-\NormalTok{        patch\_summary\_inferred, }
-        \DataTypeTok{on =} \KeywordTok{c}\NormalTok{(}\StringTok{"x\_median"}\NormalTok{, }\StringTok{"y\_median"}\NormalTok{),}
+\NormalTok{                     ][, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{x_median =} \KeywordTok{round}\NormalTok{(x_median, }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{),}
+                              \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(y_median, }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{))]}
+
+\CommentTok{# join with respatch summary}
+\NormalTok{patch_summary_compare <-}\StringTok{ }
+\StringTok{  }\KeywordTok{merge}\NormalTok{(patch_summary_real,}
+\NormalTok{        patch_summary_inferred, }
+        \DataTypeTok{on =} \KeywordTok{c}\NormalTok{(}\StringTok{"x_median"}\NormalTok{, }\StringTok{"y_median"}\NormalTok{),}
         \DataTypeTok{all.x =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{all.y =} \OtherTok{TRUE}\NormalTok{)}
 
-\CommentTok{\# drop nas}
-\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(patch\_summary\_compare)}
-
-\CommentTok{\# drop patch around WP080}
-\NormalTok{patch\_summary\_compare <{-}}\StringTok{ }\NormalTok{patch\_summary\_compare[tID }\OperatorTok{!=}\StringTok{ "WP080"}\NormalTok{, ]}
+\CommentTok{# drop nas}
+\NormalTok{patch_summary_compare <-}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(patch_summary_compare)}
+
+\CommentTok{# drop patch around WP080}
+\NormalTok{patch_summary_compare <-}\StringTok{ }\NormalTok{patch_summary_compare[tID }\OperatorTok{!=}\StringTok{ "WP080"}\NormalTok{, ]}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get linear model}
-\NormalTok{model\_duration <{-}}\StringTok{ }\KeywordTok{lm}\NormalTok{(duration\_real }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{duration,}
-                     \DataTypeTok{data =}\NormalTok{ patch\_summary\_compare)}
-
-\CommentTok{\# get R2}
-\KeywordTok{summary}\NormalTok{(model\_duration)}
-
-\CommentTok{\# write to file}
+\CommentTok{# get linear model}
+\NormalTok{model_duration <-}\StringTok{ }\KeywordTok{lm}\NormalTok{(duration_real }\OperatorTok{~}\StringTok{ }\NormalTok{duration,}
+                     \DataTypeTok{data =}\NormalTok{ patch_summary_compare)}
+
+\CommentTok{# get R2}
+\KeywordTok{summary}\NormalTok{(model_duration)}
+
+\CommentTok{# write to file}
 \KeywordTok{writeLines}\NormalTok{(}
   \DataTypeTok{text =} \KeywordTok{capture.output}\NormalTok{(}
-    \KeywordTok{summary}\NormalTok{(model\_duration)}
+    \KeywordTok{summary}\NormalTok{(model_duration)}
 \NormalTok{  ),}
-  \DataTypeTok{con =} \StringTok{"data/model\_output\_residence\_patch.txt"}
+  \DataTypeTok{con =} \StringTok{"data/model_output_residence_patch.txt"}
 \NormalTok{)}
@@
   \KeywordTok{readLines}\NormalTok{(}
-    \DataTypeTok{con =} \StringTok{"data/model\_output\_residence\_patch.txt"}
+    \DataTypeTok{con =} \StringTok{"data/model_output_residence_patch.txt"}\NormalTok{,}
+    \DataTypeTok{encoding =} \StringTok{"UTF-8"}
 \NormalTok{  ), }\DataTypeTok{sep =} \StringTok{"}\CharTok{\textbackslash{}n}\StringTok{"}
 \NormalTok{)}
+\CommentTok{#> }
+\CommentTok{#> Call:}
+\CommentTok{#> lm(formula = duration_real ~ duration, data = patch_summary_compare)}
+\CommentTok{#> }
+\CommentTok{#> Residuals:}
+\CommentTok{#>      Min       1Q   Median       3Q      Max }
+\CommentTok{#> -103.237  -19.277   -2.917    7.003   93.431 }
+\CommentTok{#> }
+\CommentTok{#> Coefficients:}
+\CommentTok{#>              Estimate Std. Error t value Pr(>|t|)    }
+\CommentTok{#> (Intercept) 101.42061   47.66936   2.128   0.0493 *  }
+\CommentTok{#> duration      1.02108    0.07876  12.965 6.66e-10 ***}
+\CommentTok{#> ---}
+\CommentTok{#> Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1}
+\CommentTok{#> }
+\CommentTok{#> Residual standard error: 50.35 on 16 degrees of freedom}
+\CommentTok{#> Multiple R-squared:  0.9131, Adjusted R-squared:  0.9077 }
+\CommentTok{#> F-statistic: 168.1 on 1 and 16 DF,  p-value: 6.655e-10}
 \end{Highlighting}
@@
 
-\hypertarget{prepare-libraries}{%
-\subsection{Prepare libraries}\label{prepare-libraries}}
+\hypertarget{prepare-libraries-1}{%
+\subsection{Prepare libraries}\label{prepare-libraries-1}}
 
@@
 \begin{Highlighting}[]
-\CommentTok{\# prepare the connection}
-\NormalTok{con <{-}}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}\DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(), }
-                 \DataTypeTok{dbname =} \StringTok{"data/Three\_example\_bats.sql"}\NormalTok{)}
-
-\CommentTok{\# list the tables}
-\NormalTok{table\_name <{-}}\StringTok{ }\KeywordTok{dbListTables}\NormalTok{(con)}
-
-\CommentTok{\# prepare to query all tables}
-\NormalTok{query <{-}}\StringTok{ }\KeywordTok{sprintf}\NormalTok{(}\StringTok{\textquotesingle{}select * from }\CharTok{\textbackslash{}"}\StringTok{\%s}\CharTok{\textbackslash{}"}\StringTok{\textquotesingle{}}\NormalTok{, table\_name)}
-
-\CommentTok{\# query the database}
-\NormalTok{data <{-}}\StringTok{ }\KeywordTok{dbGetQuery}\NormalTok{(}\DataTypeTok{conn =}\NormalTok{ con, }\DataTypeTok{statement =}\NormalTok{ query)}
-
-\CommentTok{\# disconnect from database}
+\CommentTok{# prepare the connection}
+\NormalTok{con <-}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}\DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(), }
+                 \DataTypeTok{dbname =} \StringTok{"data/Three_example_bats.sql"}\NormalTok{)}
+
+\CommentTok{# list the tables}
+\NormalTok{table_name <-}\StringTok{ }\KeywordTok{dbListTables}\NormalTok{(con)}
+
+\CommentTok{# prepare to query all tables}
+\NormalTok{query <-}\StringTok{ }\KeywordTok{sprintf}\NormalTok{(}\StringTok{'select * from }\CharTok{\textbackslash{}"}\StringTok{%s}\CharTok{\textbackslash{}"}\StringTok{'}\NormalTok{, table_name)}
+
+\CommentTok{# query the database}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{dbGetQuery}\NormalTok{(}\DataTypeTok{conn =}\NormalTok{ con, }\DataTypeTok{statement =}\NormalTok{ query)}
+
+\CommentTok{# disconnect from database}
 \KeywordTok{dbDisconnect}\NormalTok{(con)}
@@
 \begin{Highlighting}[]
-\CommentTok{\# split bat data by tag}
-\CommentTok{\# first make a copy using the data.table function copy}
-\CommentTok{\# this prevents the orignal data from being modified by atlastools}
-\CommentTok{\# functions which DO MODIFY BY REFERENCE!}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
-
-\CommentTok{\# now split}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_split, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
+\CommentTok{# split bat data by tag}
+\CommentTok{# first make a copy using the data.table function copy}
+\CommentTok{# this prevents the orignal data from being modified by atlastools}
+\CommentTok{# functions which DO MODIFY BY REFERENCE!}
+\NormalTok{data_split <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{# now split}
+\NormalTok{data_split <-}\StringTok{ }\KeywordTok{split}\NormalTok{(data_split, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# filter for SD <= 20}
-\CommentTok{\# here, reassignment is necessary as rows are being removed}
-\CommentTok{\# the atl\_filter\_covariates function could have been used here}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
-
-\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}
+\CommentTok{# filter for SD <= 20}
+\CommentTok{# here, reassignment is necessary as rows are being removed}
+\CommentTok{# the atl_filter_covariates function could have been used here}
+\NormalTok{data_split <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}
     \DataTypeTok{data =}\NormalTok{ dt,}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get speeds as with SD, no reassignment required for columns}
+\CommentTok{# get speeds as with SD, no reassignment required for columns}
 \KeywordTok{invisible}\NormalTok{(}
-  \KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
-
-    \CommentTok{\# first process time to seconds}
-    \CommentTok{\# assign to a new column}
+  \KeywordTok{lapply}\NormalTok{(data_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+
+    \CommentTok{# first process time to seconds}
+    \CommentTok{# assign to a new column}
 \NormalTok{    dt[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{floor}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}
 
-\NormalTok{    dt[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{speed\_in =} \KeywordTok{atl\_get\_speed}\NormalTok{(dt, }
+\NormalTok{    dt[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(dt, }
                                        \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{, }
@@
 \begin{Highlighting}[]
-\CommentTok{\# filter speeds}
-\CommentTok{\# reassignment is required here}
-\NormalTok{data\_split <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
-\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(dt, }\DataTypeTok{cols =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed\_in"}\NormalTok{, }\StringTok{"speed\_out"}\NormalTok{))}
-
-\NormalTok{  dt <{-}}\StringTok{ }\KeywordTok{atl\_filter\_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dt,}
-                              \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed\_in <= 20"}\NormalTok{,}
-                                          \StringTok{"speed\_out <= 20"}\NormalTok{))}
+\CommentTok{# filter speeds}
+\CommentTok{# reassignment is required here}
+\NormalTok{data_split <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{na.omit}\NormalTok{(dt, }\DataTypeTok{cols =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed_in"}\NormalTok{, }\StringTok{"speed_out"}\NormalTok{))}
+
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dt,}
+                              \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed_in <= 20"}\NormalTok{,}
+                                          \StringTok{"speed_out <= 20"}\NormalTok{))}
 \NormalTok{\})}
@@
 \begin{Highlighting}[]
-\CommentTok{\# since the function modifies in place, we shall make a copy}
-\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{copy}\NormalTok{(data\_split)}
-
-\CommentTok{\# split the data again}
-\NormalTok{data\_smooth <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_smooth, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# apply the median smooth to each list element}
-\CommentTok{\# no reassignment is required as THE FUNCTION MODIFIES IN PLACE!}
+\CommentTok{# since the function modifies in place, we shall make a copy}
+\NormalTok{data_smooth <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data_split)}
+
+\CommentTok{# split the data again}
+\NormalTok{data_smooth <-}\StringTok{ }\KeywordTok{split}\NormalTok{(data_smooth, }\DataTypeTok{by =} \StringTok{"TAG"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# apply the median smooth to each list element}
+\CommentTok{# no reassignment is required as THE FUNCTION MODIFIES IN PLACE!}
 \KeywordTok{invisible}\NormalTok{(}
 
-  \CommentTok{\# the function arguments to atl\_median\_smooth}
-  \CommentTok{\# can be passed directly in lapply}
-
-  \KeywordTok{lapply}\NormalTok{(}\DataTypeTok{X =}\NormalTok{ data\_smooth, }
-         \DataTypeTok{FUN =}\NormalTok{ atl\_median\_smooth,}
-         \DataTypeTok{time =} \StringTok{"time"}\NormalTok{, }\DataTypeTok{moving\_window =} \DecValTok{5}\NormalTok{)}
+  \CommentTok{# the function arguments to atl_median_smooth}
+  \CommentTok{# can be passed directly in lapply}
+
+  \KeywordTok{lapply}\NormalTok{(}\DataTypeTok{X =}\NormalTok{ data_smooth, }
+         \DataTypeTok{FUN =}\NormalTok{ atl_median_smooth,}
+         \DataTypeTok{time =} \StringTok{"time"}\NormalTok{, }\DataTypeTok{moving_window =} \DecValTok{5}\NormalTok{)}
 \NormalTok{)}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get residence times}
-
-\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_smooth, }\ControlFlowTok{function}\NormalTok{(dt) \{}
-  \CommentTok{\# do basic recurse}
-\NormalTok{  dt\_recurse <{-}}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(}
+\CommentTok{# get residence times}
+
+\NormalTok{data_residence <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_smooth, }\ControlFlowTok{function}\NormalTok{(dt) \{}
+  \CommentTok{# do basic recurse}
+\NormalTok{  dt_recurse <-}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(}
     \DataTypeTok{x =}\NormalTok{ dt[, }\KeywordTok{c}\NormalTok{(}\StringTok{"X"}\NormalTok{, }\StringTok{"Y"}\NormalTok{, }\StringTok{"time"}\NormalTok{, }\StringTok{"TAG"}\NormalTok{)],}
@@
 
-  \CommentTok{\# get revisit stats}
-\NormalTok{  dt\_recurse <{-}}\StringTok{ }\KeywordTok{setDT}\NormalTok{(}
-\NormalTok{    dt\_recurse[[}\StringTok{"revisitStats"}\NormalTok{]]}
+  \CommentTok{# get revisit stats}
+\NormalTok{  dt_recurse <-}\StringTok{ }\KeywordTok{setDT}\NormalTok{(}
+\NormalTok{    dt_recurse[[}\StringTok{"revisitStats"}\NormalTok{]]}
 \NormalTok{  )}
 
-  \CommentTok{\# count long absences from the area}
-\NormalTok{  dt\_recurse[, timeSinceLastVisit }\OperatorTok{:}\ErrorTok{=}
-\StringTok{          }\KeywordTok{ifelse}\NormalTok{(}\KeywordTok{is.na}\NormalTok{(timeSinceLastVisit), }\OperatorTok{{-}}\OtherTok{Inf}\NormalTok{, timeSinceLastVisit)]}
-\NormalTok{  dt\_recurse[, longAbsenceCounter }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{cumsum}\NormalTok{(timeSinceLastVisit }\OperatorTok{>}\StringTok{ }\DecValTok{60}\NormalTok{),}
+  \CommentTok{# count long absences from the area}
+\NormalTok{  dt_recurse[, timeSinceLastVisit }\OperatorTok{:}\ErrorTok{=}
+\StringTok{          }\KeywordTok{ifelse}\NormalTok{(}\KeywordTok{is.na}\NormalTok{(timeSinceLastVisit), }\OperatorTok{-}\OtherTok{Inf}\NormalTok{, timeSinceLastVisit)]}
+\NormalTok{  dt_recurse[, longAbsenceCounter }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{cumsum}\NormalTok{(timeSinceLastVisit }\OperatorTok{>}\StringTok{ }\DecValTok{60}\NormalTok{),}
 \NormalTok{             by =}\StringTok{ }\NormalTok{.(coordIdx)}
 \NormalTok{             ]}
-  \CommentTok{\# get data before the first long absence of 60 mins}
-\NormalTok{  dt\_recurse <{-}}\StringTok{ }\NormalTok{dt\_recurse[longAbsenceCounter }\OperatorTok{<}\StringTok{ }\DecValTok{1}\NormalTok{, ]}
-
-\NormalTok{  dt\_recurse <{-}}\StringTok{ }\NormalTok{dt\_recurse[, }\KeywordTok{list}\NormalTok{(}
+  \CommentTok{# get data before the first long absence of 60 mins}
+\NormalTok{  dt_recurse <-}\StringTok{ }\NormalTok{dt_recurse[longAbsenceCounter }\OperatorTok{<}\StringTok{ }\DecValTok{1}\NormalTok{, ]}
+
+\NormalTok{  dt_recurse <-}\StringTok{ }\NormalTok{dt_recurse[, }\KeywordTok{list}\NormalTok{(}
     \DataTypeTok{resTime =} \KeywordTok{sum}\NormalTok{(timeInside),}
@@
 \begin{Highlighting}[]
-\CommentTok{\# bind the list}
-\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_residence)}
-
-\CommentTok{\# get time as human readable}
-\NormalTok{data\_residence[, ts }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.POSIXct}\NormalTok{(time, }\DataTypeTok{origin =} \StringTok{"1970{-}01{-}01"}\NormalTok{)]}
+\CommentTok{# bind the list}
+\NormalTok{data_residence <-}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data_residence)}
+
+\CommentTok{# get time as human readable}
+\NormalTok{data_residence[, ts }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.POSIXct}\NormalTok{(time, }\DataTypeTok{origin =} \StringTok{"1970-01-01"}\NormalTok{)]}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# add an id column}
-\NormalTok{data\_residence[, }\StringTok{\textasciigrave{}}\DataTypeTok{:=}\StringTok{\textasciigrave{}}\NormalTok{(}\DataTypeTok{id =}\NormalTok{ TAG,}
+\CommentTok{# add an id column}
+\NormalTok{data_residence[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{id =}\NormalTok{ TAG,}
                       \DataTypeTok{x =}\NormalTok{ X, }\DataTypeTok{y =}\NormalTok{ Y)]}
 
-\CommentTok{\# filter for residence time > 5 minutes}
-\NormalTok{data\_residence <{-}}\StringTok{ }\NormalTok{data\_residence[resTime }\OperatorTok{>}\StringTok{ }\DecValTok{5}\NormalTok{, ]}
-
-\CommentTok{\# split the data}
-\NormalTok{data\_residence <{-}}\StringTok{ }\KeywordTok{split}\NormalTok{(data\_residence, data\_residence}\OperatorTok{$}\NormalTok{TAG)}
+\CommentTok{# filter for residence time > 5 minutes}
+\NormalTok{data_residence <-}\StringTok{ }\NormalTok{data_residence[resTime }\OperatorTok{>}\StringTok{ }\DecValTok{5}\NormalTok{, ]}
+
+\CommentTok{# split the data}
+\NormalTok{data_residence <-}\StringTok{ }\KeywordTok{split}\NormalTok{(data_residence, data_residence}\OperatorTok{$}\NormalTok{TAG)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get data spatials}
-\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_patches, atl\_patch\_summary,}
-                        \DataTypeTok{which\_data =} \StringTok{"spatial"}\NormalTok{,}
-                        \DataTypeTok{buffer\_radius =} \DecValTok{25}\NormalTok{)}
-
-\CommentTok{\# bind list}
-\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data\_spatials)}
-
-\CommentTok{\# convert to sf}
+\CommentTok{# get data spatials}
+\NormalTok{data_spatials <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_patches, atl_patch_summary,}
+                        \DataTypeTok{which_data =} \StringTok{"spatial"}\NormalTok{,}
+                        \DataTypeTok{buffer_radius =} \DecValTok{25}\NormalTok{)}
+
+\CommentTok{# bind list}
+\NormalTok{data_spatials <-}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(data_spatials)}
+
+\CommentTok{# convert to sf}
 \KeywordTok{library}\NormalTok{(sf)}
-\NormalTok{data\_spatials <{-}}\StringTok{ }\KeywordTok{st\_sf}\NormalTok{(data\_spatials, }\DataTypeTok{sf\_column\_name =} \StringTok{"polygons"}\NormalTok{)}
-
-\CommentTok{\# assign a crs}
-\KeywordTok{st\_crs}\NormalTok{(data\_spatials) <{-}}\StringTok{ }\KeywordTok{st\_crs}\NormalTok{(}\DecValTok{2039}\NormalTok{)}
+\NormalTok{data_spatials <-}\StringTok{ }\KeywordTok{st_sf}\NormalTok{(data_spatials, }\DataTypeTok{sf_column_name =} \StringTok{"polygons"}\NormalTok{)}
+
+\CommentTok{# assign a crs}
+\KeywordTok{st_crs}\NormalTok{(data_spatials) <-}\StringTok{ }\KeywordTok{st_crs}\NormalTok{(}\DecValTok{2039}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{\# get summary}
-\NormalTok{patch\_summary <{-}}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data\_patches, atl\_patch\_summary)}
-
-\CommentTok{\# bind summary}
-\NormalTok{patch\_summary <{-}}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(patch\_summary)}
-
-\CommentTok{\# write}
-\KeywordTok{fwrite}\NormalTok{(patch\_summary,}
-       \StringTok{"data/data\_bat\_patch\_summary.csv"}\NormalTok{)}
+\CommentTok{# get summary}
+\NormalTok{patch_summary <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_patches, atl_patch_summary)}
+
+\CommentTok{# bind summary}
+\NormalTok{patch_summary <-}\StringTok{ }\KeywordTok{rbindlist}\NormalTok{(patch_summary)}
+
+\CommentTok{# write}
+\KeywordTok{fwrite}\NormalTok{(patch_summary,}
+       \StringTok{"data/data_bat_patch_summary.csv"}\NormalTok{)}
 \end{Highlighting}
@@
 \hypertarget{refs}{}
-\begin{cslreferences}
 \leavevmode\hypertarget{ref-barraquand2008}{}%

### 2020-11-23 R/05_calibration_data.R

@@ -263,8 +263,8 @@ data_res <- data_unproc[stri_detect(tID, regex = "(WP)")]
 
 ## -----------------------------------------------------------------------------
 # get centroid
-data_res_summary <- data_res[, list(x_median = round(median(x), digits = -2),
-                                    y_median = round(median(y), digits = -2)),
+data_res_summary <- data_res[, list(x_median = median(x),
+                                    y_median = median(y),
                                     t_median = median(time)),
                              by = "tID"]
 
@@ -335,8 +335,9 @@ patch_summary_data <- atl_patch_summary(patch_res_known,
 
 
 ## -----------------------------------------------------------------------------
-# read griend
+# read griend and hut
 griend <- sf::st_read("data/griend_polygon/griend_polygon.shp")
+hut <- sf::st_read("data/griend_hut.gpkg")
 
 
 ## ----echo=FALSE---------------------------------------------------------------
@@ -350,6 +351,10 @@ fig_basic_residence <-
              stroke = 1,
              col = pal[4],
              alpha = 1)+
+  geom_sf(data = hut,
+          size = 10, shape = 7,
+          stroke = 1,
+          col = pal[1])+
   geom_path(data = data_for_patch,
             aes(x, y),
             show.legend = F,
@@ -374,6 +379,12 @@ fig_basic_residence <-
            expand = F,
            ylim = c(5901650, 5903100),
            xlim = c(650000, 650600))
+
+
+## ----echo=FALSE---------------------------------------------------------------
+ggsave(fig_basic_residence,
+       filename = "figures/fig_calib_residence.png",
+       height = 170, width = 80, units = "mm")
 
 
 ## -----------------------------------------------------------------------------
@@ -471,6 +482,15 @@ figure_lm_duration <-
 
 ggsave(figure_lm_duration, filename = "figures/fig_calib_lm_duration.png",
        width = 80 / 25, height = 80 / 25)
+
+
+## ---- eval=TRUE---------------------------------------------------------------
+cat(
+  readLines(
+    con = "data/model_output_residence_patch.txt",
+    encoding = "UTF-8"
+  ), sep = "\n"
+)
 
 
 ## ----echo=FALSE---------------------------------------------------------------

### 2020-11-23 05_calibration_data.Rmd
allow model output
conditional fonts

@@
 # get centroid
-data_res_summary <- data_res[, list(x_median = round(median(x), digits = -2),
-                                    y_median = round(median(y), digits = -2)),
+data_res_summary <- data_res[, list(x_median = median(x),
+                                    y_median = median(y),
                                     t_median = median(time)),
@@
 ```{r}
-# read griend
+# read griend and hut
 griend <- sf::st_read("data/griend_polygon/griend_polygon.shp")
+hut <- sf::st_read("data/griend_hut.gpkg")
 ```
@@
              alpha = 1)+
+  geom_sf(data = hut,
+          size = 10, shape = 7,
+          stroke = 1,
+          col = pal[1])+
   geom_path(data = data_for_patch,
@@
 
-```{r}
+```{r, eval=TRUE}
 cat(
   readLines(
-    con = "data/model_output_residence_patch.txt"
+    con = "data/model_output_residence_patch.txt",
+    encoding = "UTF-8"
   ), sep = "\n"

### 2020-11-23 docs/supplementary_material.tex
add sup mat text

@@
 %
-\documentclass[]{scrartcl}
+\documentclass[]{scrreprt}
 \usepackage{lmodern}
@@
 \newenvironment{Shaded}{}{}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
+\newcommand{\AttributeTok}[1]{#1}
+\newcommand{\BaseNTok}[1]{#1}
 \newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.00,0.50,0.50}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
+\newcommand{\ConstantTok}[1]{#1}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.00,1.00}{#1}}
+\newcommand{\DataTypeTok}[1]{#1}
+\newcommand{\DecValTok}[1]{#1}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
 \newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
 \newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
+\newcommand{\FloatTok}[1]{#1}
+\newcommand{\FunctionTok}[1]{#1}
 \newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.00,1.00}{#1}}
 \newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
+\newcommand{\OperatorTok}[1]{#1}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{1.00,0.25,0.00}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{1.00,0.25,0.00}{#1}}
 \newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.50,0.50}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.00,0.50,0.50}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.00,0.50,0.50}{#1}}
+\newcommand{\VariableTok}[1]{#1}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.00,0.50,0.50}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{\textbf{#1}}}
 \usepackage{longtable,booktabs}
@@
 \usepackage{lineno}
+\KOMAoption{fontsize}{10pt}
 
@@
 {
-\setcounter{tocdepth}{2}
+\setcounter{tocdepth}{1}
 \tableofcontents
@@
 \hypertarget{processing-calibration-data}{%
-\section{Processing calibration data}\label{processing-calibration-data}}
+\chapter{Processing calibration data}\label{processing-calibration-data}}
 
@@
 \hypertarget{prepare-libraries}{%
-\subsection{Prepare libraries}\label{prepare-libraries}}
+\section{Prepare libraries}\label{prepare-libraries}}
 
@@
 \hypertarget{access-data-and-preliminary-visualisation}{%
-\subsection{Access data and preliminary visualisation}\label{access-data-and-preliminary-visualisation}}
+\section{Access data and preliminary visualisation}\label{access-data-and-preliminary-visualisation}}
 
@@
 \hypertarget{filter-by-bounding-box}{%
-\subsection{Filter by bounding box}\label{filter-by-bounding-box}}
+\section{Filter by bounding box}\label{filter-by-bounding-box}}
 
@@
 \hypertarget{filter-trajectories}{%
-\subsection{Filter trajectories}\label{filter-trajectories}}
+\section{Filter trajectories}\label{filter-trajectories}}
 
 \hypertarget{handle-time}{%
-\subsubsection{Handle time}\label{handle-time}}
+\subsection{Handle time}\label{handle-time}}
 
@@
 \hypertarget{add-speed-and-turning-angle}{%
-\subsubsection{Add speed and turning angle}\label{add-speed-and-turning-angle}}
+\subsection{Add speed and turning angle}\label{add-speed-and-turning-angle}}
 
@@
 \hypertarget{get-95th-percentile-of-speed-and-angle}{%
-\subsubsection{Get 95th percentile of speed and angle}\label{get-95th-percentile-of-speed-and-angle}}
+\subsection{Get 95th percentile of speed and angle}\label{get-95th-percentile-of-speed-and-angle}}
 
@@
 \hypertarget{filter-on-speed}{%
-\subsubsection{Filter on speed}\label{filter-on-speed}}
+\subsection{Filter on speed}\label{filter-on-speed}}
 
@@
 \hypertarget{smoothing-the-trajectory}{%
-\subsection{Smoothing the trajectory}\label{smoothing-the-trajectory}}
+\section{Smoothing the trajectory}\label{smoothing-the-trajectory}}
 
@@
 \hypertarget{thinning-the-data}{%
-\subsection{Thinning the data}\label{thinning-the-data}}
+\section{Thinning the data}\label{thinning-the-data}}
 
@@
 \hypertarget{residence-patches}{%
-\subsection{Residence patches}\label{residence-patches}}
+\section{Residence patches}\label{residence-patches}}
 
 \hypertarget{get-waypoint-centroids}{%
-\subsubsection{Get waypoint centroids}\label{get-waypoint-centroids}}
+\subsection{Get waypoint centroids}\label{get-waypoint-centroids}}
 
@@
 \hypertarget{prepare-data}{%
-\subsubsection{Prepare data}\label{prepare-data}}
+\subsection{Prepare data}\label{prepare-data}}
 
@@
 \hypertarget{calculate-residence-time}{%
-\subsubsection{Calculate residence time}\label{calculate-residence-time}}
+\subsection{Calculate residence time}\label{calculate-residence-time}}
 
@@
 \hypertarget{run-residence-patch-method}{%
-\subsubsection{Run residence patch method}\label{run-residence-patch-method}}
+\subsection{Run residence patch method}\label{run-residence-patch-method}}
 
@@
 \hypertarget{get-spatial-and-summary-objects}{%
-\subsubsection{Get spatial and summary objects}\label{get-spatial-and-summary-objects}}
+\subsection{Get spatial and summary objects}\label{get-spatial-and-summary-objects}}
 
@@
 \hypertarget{prepare-to-plot-data}{%
-\subsubsection{Prepare to plot data}\label{prepare-to-plot-data}}
+\subsection{Prepare to plot data}\label{prepare-to-plot-data}}
 
@@
 \hypertarget{compare-patch-metrics}{%
-\subsection{Compare patch metrics}\label{compare-patch-metrics}}
+\section{Compare patch metrics}\label{compare-patch-metrics}}
 
@@
 \hypertarget{linear-model-durations}{%
-\subsubsection{Linear model durations}\label{linear-model-durations}}
+\subsection{Linear model durations}\label{linear-model-durations}}
 
@@
 \hypertarget{linear-model-summary}{%
-\subsubsection{Linear model summary}\label{linear-model-summary}}
+\subsection{Linear model summary}\label{linear-model-summary}}
 
@@
 \hypertarget{processing-egyptian-fruit-bat-tracks}{%
-\section{Processing Egyptian fruit bat tracks}\label{processing-egyptian-fruit-bat-tracks}}
+\chapter{Processing Egyptian fruit bat tracks}\label{processing-egyptian-fruit-bat-tracks}}
 
@@
 \hypertarget{prepare-libraries-1}{%
-\subsection{Prepare libraries}\label{prepare-libraries-1}}
+\section{Prepare libraries}\label{prepare-libraries-1}}
 
@@
 \hypertarget{install-atlastools-from-github.}{%
-\subsection{\texorpdfstring{Install \texttt{atlastools} from Github.}{Install atlastools from Github.}}\label{install-atlastools-from-github.}}
+\section{\texorpdfstring{Install \texttt{atlastools} from Github.}{Install atlastools from Github.}}\label{install-atlastools-from-github.}}
 
@@
 \hypertarget{read-bat-data}{%
-\subsection{Read bat data}\label{read-bat-data}}
+\section{Read bat data}\label{read-bat-data}}
 
@@
 \hypertarget{a-first-visual-inspection}{%
-\subsection{A First Visual Inspection}\label{a-first-visual-inspection}}
+\section{A First Visual Inspection}\label{a-first-visual-inspection}}
 
@@
 \hypertarget{prepare-data-for-filtering}{%
-\subsection{Prepare data for filtering}\label{prepare-data-for-filtering}}
+\section{Prepare data for filtering}\label{prepare-data-for-filtering}}
 
@@
 \hypertarget{prepare-data-per-individual}{%
-\subsubsection{Prepare data per individual}\label{prepare-data-per-individual}}
+\subsection{Prepare data per individual}\label{prepare-data-per-individual}}
 
@@
 \hypertarget{filter-by-covariates}{%
-\subsection{Filter by covariates}\label{filter-by-covariates}}
+\section{Filter by covariates}\label{filter-by-covariates}}
 
@@
 \hypertarget{sanity-check-plot-filtered-data}{%
-\subsubsection{Sanity check: Plot filtered data}\label{sanity-check-plot-filtered-data}}
-
-We plot the data to check whether the filtering has improved the data (Figure 2).
+\subsection{Sanity check: Plot filtered data}\label{sanity-check-plot-filtered-data}}
+
+We plot the data to check whether the filtering has improved the data (Figure 9).
 The plot code is once again hidden in this rendering, but is available in the source code file.
@@
 \hypertarget{filter-by-speed}{%
-\subsection{Filter by speed}\label{filter-by-speed}}
+\section{Filter by speed}\label{filter-by-speed}}
 
@@
 \hypertarget{sanity-check-plot-speed-filtered-data}{%
-\subsubsection{Sanity check: Plot speed filtered data}\label{sanity-check-plot-speed-filtered-data}}
-
-The speed filtered data is now inspected for errors (Figure 3). The plot code is once again hidden.
+\subsection{Sanity check: Plot speed filtered data}\label{sanity-check-plot-speed-filtered-data}}
+
+The speed filtered data is now inspected for errors (Figure 10). The plot code is once again hidden.
 
@@
 \hypertarget{median-smoothing}{%
-\subsection{Median smoothing}\label{median-smoothing}}
-
-The quality of this data (Figure 3) is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless.
+\section{Median smoothing}\label{median-smoothing}}
+
+The quality of this data (Figure 10) is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless.
 
@@
 \hypertarget{sanity-check-plot-smoothed-data}{%
-\subsubsection{Sanity check: Plot smoothed data}\label{sanity-check-plot-smoothed-data}}
+\subsection{Sanity check: Plot smoothed data}\label{sanity-check-plot-smoothed-data}}
 
@@
 \hypertarget{making-residence-patches}{%
-\subsection{Making residence patches}\label{making-residence-patches}}
+\section{Making residence patches}\label{making-residence-patches}}
 
 \hypertarget{calculating-residence-time}{%
-\subsubsection{Calculating residence time}\label{calculating-residence-time}}
+\subsection{Calculating residence time}\label{calculating-residence-time}}
 
@@
 \hypertarget{constructing-residence-patches}{%
-\subsubsection{Constructing residence patches}\label{constructing-residence-patches}}
+\subsection{Constructing residence patches}\label{constructing-residence-patches}}
 
@@
 \hypertarget{getting-residence-patch-data}{%
-\subsubsection{Getting residence patch data}\label{getting-residence-patch-data}}
+\subsection{Getting residence patch data}\label{getting-residence-patch-data}}
 
@@
 These are returned as a list and must be converted into a single \texttt{sf} object.
-These objects and the raw movement data are shown in Figure 5.
+These objects and the raw movement data are shown in Figure 12.
 
@@
 \hypertarget{write-patch-spatial-representations}{%
-\subsubsection{Write patch spatial representations}\label{write-patch-spatial-representations}}
+\subsection{Write patch spatial representations}\label{write-patch-spatial-representations}}
 
@@
 \hypertarget{references}{%
-\section{References}\label{references}}
+\chapter{References}\label{references}}
 

### 2020-11-24 06_bat_data.Rmd

@@
 
-We plot the data to check whether the filtering has improved the data (Figure 2).
+We plot the data to check whether the filtering has improved the data (Figure 9).
 The plot code is once again hidden in this rendering, but is available in the source code file.
@@
 
-The speed filtered data is now inspected for errors (Figure 3). The plot code is once again hidden.
+The speed filtered data is now inspected for errors (Figure 10). The plot code is once again hidden.
 
@@
 
-The quality of this data (Figure 3) is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless.
+The quality of this data (Figure 10) is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless.
 
@@
 These are returned as a list and must be converted into a single `sf` object.
-These objects and the raw movement data are shown in Figure 5.
+These objects and the raw movement data are shown in Figure 12.
 

### 2020-11-24 index.rmd

@@
 output: bookdown::pdf_document2
-documentclass: scrartcl
-toc-depth: 2
+documentclass: scrreprt
+toc-depth: 1
 secnumdepth: 2

### 2020-12-03 05_calibration_data.Rmd
minor changes
add outline

@@
 
-# Processing calibration data
+# Validating the Residence Patch Method with Calibration Data
 
 Here we show how the residence patch method [@barraquand2008; @bijleveld2016; @oudman2018] accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
+
+## Outline of Cleaning Steps
+
+We began by visualising the data to check for location errors, and found a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2; main text Fig. 7.a).
+This outlier was removed by filtering data by the X coordinate bounds using the function `atl_filter\_bounds`; X coordinate bounds $\leq$ 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
+1.2).
+We then calculated the incoming and outgoing speed, as well as the turning angle at each position using the functions `atl_get_speed` and `atl_turning_angle` respectively, as a precursor to targeting large-scale location errors in the form of point outliers.
+We used the function `atl_filter_covariates` to remove positions with incoming and outgoing speeds $\geq$ the speed threshold of 15 m/s (n = 13,491, 26.5%; remaining positions = 37,324, 73.5%; Fig.
+1.3; main text Fig.
+7.b).
+This speed threshold was chosen as 5 m/s faster than the known boat speed, 10 m/s.
+Finally, we targeted small-scale location errors by applying a median smooth with a moving window size $K$ = 5 using the function `atl_median_smooth` (Fig. 1.4; main text Fig. 7.c).
+Smoothing did not reduce the number of positions.
+We thinned the data to a 30 second interval leaving 1,803 positions (4.8\% positions of the smoothed track)
 
@@
 This function modifies in place, and does not need to be assigned to a new variable.
-We create a copy of the data before applying the smooth so that we can compare the data before and after smoothin.
+We create a copy of the data before applying the smooth so that we can compare the data before and after smoothing.
 

### 2020-12-03 05_calibration_data.Rmd

@@
 Median smoothing successfully recovers the likely path of the track without a loss of data.
-The area shown is the upper rectangle from Figure 3.](figures/fig_calib_median_smooth.png)
+The area shown is the upper rectangle from Fig.
+1.3.](figures/fig_calib_median_smooth.png)
 
@@
 Here, thinned positions are shown as purple squares, with the size of the square indicating the number of positions within the 30 second bin used to obtain the average position.
-Green points show the median smoothed data from Figure 4, while the raw data are shown in grey.
-The area shown is the upper rectangle in Figure 3.](figures/fig_calib_smooth_thin.png)
+Green points show the median smoothed data from Fig.
+1.4, while the raw data are shown in grey.
+The area shown is the upper rectangle in Fig.
+1.3.](figures/fig_calib_smooth_thin.png)
 
@@
 The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).
-The area shown is the lower rectangle in Figure 3.](figures/fig_calib_residence.png)
+The area shown is the lower rectangle in Fig.
+1.3.](figures/fig_calib_residence.png)
 

### 2020-12-03 06_bat_data.Rmd

--- 2020-11-22 06_bat_data.Rmd
@@
 
-Plot the bat data as a sanity check, and inspect it visually for errors (Figure 1).
+Plot the bat data as a sanity check, and inspect it visually for errors (Fig. 2.1).
 The plot code is hidden in the rendered copy (PDF) of this supplementary material, but is available in the `Rmarkdown` file "06_bat_data.Rmd".
-The saved plot is shown below as Figure 1.
+The saved plot is shown below as Fig. 2.1.
 
@@
 
-We plot the data to check whether the filtering has improved the data (Figure 2).
+We plot the data to check whether the filtering has improved the data (Fig. 2.2).
 The plot code is once again hidden in this rendering, but is available in the source code file.
@@
 
-Some point outliers remain (Figure 2), and could be removed using a speed filter.
+Some point outliers remain, and could be removed using a speed filter.
 
@@
 
-The speed filtered data is now inspected for errors (Figure 3). The plot code is once again hidden.
+The speed filtered data is now inspected for errors (Fig. 2.3). The plot code is once again hidden.
 
@@
 
-The quality of this data (Figure 3) is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless.
+The quality of the data is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless (Fig. 2.4).
 
@@
 These are returned as a list and must be converted into a single `sf` object.
-These objects and the raw movement data are shown in Figure 5.
+These objects and the raw movement data are shown in Fig. 2.5.
 

### 2020-12-03 docs/supplementary_material.tex

@@
 \author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Allert I. Bijleveld}
-\date{2020-11-23}
+\date{2020-12-03}
 
@@
 }
-\hypertarget{processing-calibration-data}{%
-\chapter{Processing calibration data}\label{processing-calibration-data}}
+\hypertarget{validating-the-residence-patch-method-with-calibration-data}{%
+\chapter{Validating the Residence Patch Method with Calibration Data}\label{validating-the-residence-patch-method-with-calibration-data}}
 
@@
 \protect\hyperlink{ref-oudman2018}{2018}) accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
+
+\hypertarget{outline-of-cleaning-steps}{%
+\section{Outline of Cleaning Steps}\label{outline-of-cleaning-steps}}
+
+We began by visualising the data to check for location errors, and found a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2; main text Fig. 7.a).
+This outlier was removed by filtering data by the X coordinate bounds using the function \texttt{atl\_filter\textbackslash{}\_bounds}; X coordinate bounds \(\leq\) 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
+1.2).
+We then calculated the incoming and outgoing speed, as well as the turning angle at each position using the functions \texttt{atl\_get\_speed} and \texttt{atl\_turning\_angle} respectively, as a precursor to targeting large-scale location errors in the form of point outliers.
+We used the function \texttt{atl\_filter\_covariates} to remove positions with incoming and outgoing speeds \(\geq\) the speed threshold of 15 m/s (n = 13,491, 26.5\%; remaining positions = 37,324, 73.5\%; Fig.
+1.3; main text Fig.
+7.b).
+This speed threshold was chosen as 5 m/s faster than the known boat speed, 10 m/s.
+Finally, we targeted small-scale location errors by applying a median smooth with a moving window size \(K\) = 5 using the function \texttt{atl\_median\_smooth} (Fig. 1.4; main text Fig. 7.c).
+Smoothing did not reduce the number of positions.
+We thinned the data to a 30 second interval leaving 1,803 positions (4.8\% positions of the smoothed track)
 
@@
 Median smoothing successfully recovers the likely path of the track without a loss of data.
-The area shown is the upper rectangle from Figure 3.}
+The area shown is the upper rectangle from Fig.
+1.3.}
 \end{figure}
@@
 Here, thinned positions are shown as purple squares, with the size of the square indicating the number of positions within the 30 second bin used to obtain the average position.
-Green points show the median smoothed data from Figure 4, while the raw data are shown in grey.
-The area shown is the upper rectangle in Figure 3.}
+Green points show the median smoothed data from Fig.
+1.4, while the raw data are shown in grey.
+The area shown is the upper rectangle in Fig.
+1.3.}
 \end{figure}
@@
 The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).
-The area shown is the lower rectangle in Figure 3.}
+The area shown is the lower rectangle in Fig.
+1.3.}
 \end{figure}
@@
 
-Plot the bat data as a sanity check, and inspect it visually for errors (Figure 1).
+Plot the bat data as a sanity check, and inspect it visually for errors (Fig. 2.1).
 The plot code is hidden in the rendered copy (PDF) of this supplementary material, but is available in the \texttt{Rmarkdown} file ``06\_bat\_data.Rmd''.
-The saved plot is shown below as Figure 1.
+The saved plot is shown below as Fig. 2.1.
 
@@
 
-We plot the data to check whether the filtering has improved the data (Figure 9).
+We plot the data to check whether the filtering has improved the data (Fig. 2.2).
 The plot code is once again hidden in this rendering, but is available in the source code file.
@@
 
-The speed filtered data is now inspected for errors (Figure 10). The plot code is once again hidden.
+The speed filtered data is now inspected for errors (Fig. 2.3). The plot code is once again hidden.
 
@@
 
-The quality of this data (Figure 10) is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless.
+The quality of the data is relatively high, and a median smooth is not strictly necessary. We demonstrate the application of a 5 point median smooth to the data nonetheless (Fig. 2.4).
 
@@
 These are returned as a list and must be converted into a single \texttt{sf} object.
-These objects and the raw movement data are shown in Figure 12.
+These objects and the raw movement data are shown in Fig. 2.5.
 

### 2020-12-09 render_books.sh
build and name supplement correctly

@@ -4,3 +4,11 @@ Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2"
 
 # to render source files
 Rscript --slave -e 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
+
+# copy pdfs and name correctly
+cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_`date -I`.pdf
+cp -p docs/supplementary_material.pdf docs/supplementary_material_`date -I`.pdf
+rm docs/supplementary_material.pdf
+
+# build atlastools manual
+Rscript --slave -e 'devtools::build_manual(pkg = "../atlastools", path = "docs/")'

### 2020-12-09 render_books.sh
better naming

@@ -6,8 +6,8 @@ Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2"
 Rscript --slave -e 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
 
 # copy pdfs and name correctly
-cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_`date -I`.pdf
-cp -p docs/supplementary_material.pdf docs/supplementary_material_`date -I`.pdf
+cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text`date -I`.pdf
+cp -p docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_`date -I`.pdf
 rm docs/supplementary_material.pdf
 
 # build atlastools manual

### 2020-12-10 R/tmp_code.R
remove old code and refs

@@ -1,100 +1,0 @@
-### Simulate movement replicates
-
-```{r}
-sim_data <- mapply(function(p_outlier, ext_outlier, replicate) {
-  
-  # make simulated data
-  ref_data <- do_smoove_data()
-  
-  # extract 'true' movement metrics
-  speed <- atl_get_speed(ref_data, type = "in")
-  
-  # now add error
-  error_data <- do_add_outliers(ref_data, 
-                                p_data = p_outlier,
-                                std_dev = ext_outlier)
-  
-  # get movement metrics with error
-  error_speed <- atl_get_speed(error_data, type = "in")
-  
-  # now clean data
-  clean_data <- copy(error_data)
-  clean_data[, `:=`(in_speed = atl_get_speed(clean_data, type = "in"),
-                    out_speed = atl_get_speed(clean_data, type = "out"),
-                    angle = atl_turning_angle(clean_data))]
-  upper_speed_cutoff <- stats::quantile(clean_data$in_speed, 
-                                        probs = 0.9,
-                                        na.rm = TRUE)
-  angle_cutoff <- 40
-  clean_data <- atl_filter_covariates(clean_data,
-                                      filters = c(glue('{c("in_speed", 
-                                                           "out_speed")} < 
-                                                         {upper_speed_cutoff}'),
-                                                  glue('angle < {angle_cutoff}')
-                                                  )
-                                      )
-  
-  # get speed and land proportion
-  clean_speed <- atl_get_speed(clean_data, type = "in")
-  
-  # make speed results
-  speed_res <- data.table(type = c("reference", "outliers", "cleaned"),
-                          speed_mean = unlist(lapply(list(speed, 
-                                                          error_speed, 
-                                                          clean_speed),
-                                                     mean, na.rm = TRUE)),
-                          speed_sd = unlist(lapply(list(speed, 
-                                                          error_speed, 
-                                                          clean_speed),
-                                                     sd, na.rm = TRUE)),
-                          p_outlier = p_outlier,
-                          ext_outlier = ext_outlier,
-                          replicate = replicate
-                          )
-  results <- list(speed_res = speed_res)
-  
-  return(results)
-}, 
-param_c$p_outlier, 
-param_c$ext_outlier,
-param_c$replicate,
-SIMPLIFY = FALSE)
-```
-
-
[…]

### 2020-12-10 R/helper_functions.R
remove old funs and add header

@@ -1,4 +1,8 @@
+#### Functions to simulate and add errors to movement data ####
 
+# functions are not documented and are not intended for general use
+
+# a function to use smoove to simulate movement data
 do_smoove_data <- function(t_max = 100, dt = dt) {
   # define parameters
   nu <- 2
@@ -37,20 +41,4 @@ do_add_outliers <- function(data, p_data = 0.1, std_dev = 0.05) {
                                   }),
             .SDcols = c("x", "y")]
   return(tmp_data)
-}
-
-# function for extracting and tallying landscape values
-do_count_landcover <- function(land, track) {
-  land_values <- raster::extract(land, track[, c("x", "y")])
-  land_values <- data.table::as.data.table(table(land_values))
-  land_values[, p_land := N / sum(N)]
-  return(land_values)
-}
-
-# a function to scale the landscape
-do_scale_landscape <- function(land_, scale_factor) {
-  bounds <- c(0, 1, 0, 1)
-  bounds <- bounds * scale_factor
-  raster::extent(land_) <- bounds
-  return(land_)
 }

### 2020-12-15 docs/supplementary_material.tex
add edit to tex file
add even more final pdfs

@@
 \hypersetup{
-            pdftitle={Supplementary Material for: A Rough Guide to Pre-processing High-Frequency Animal Tracking Data},
-            pdfauthor={Pratik R. Gupte; Christine E. Beardsworth; Orr Spiegel; Emmanuel Lourie; Allert I. Bijleveld},
+            pdftitle={Supplementary Material for A Guide to Pre-processing High-Frequency Animal Tracking Data},
+            pdfauthor={Pratik R. Gupte; Christine E. Beardsworth; Orr Spiegel; Emmanuel Lourie; Sivan Toledo; Ran Nathan; Allert I. Bijleveld},
             pdfborder={0 0 0},
@@
 
-\title{Supplementary Material for: A Rough Guide to Pre-processing High-Frequency Animal Tracking Data}
-\author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Allert I. Bijleveld}
-\date{2020-12-03}
+\title{Supplementary Material for \emph{A Guide to Pre-processing High-Frequency Animal Tracking Data}}
+\author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Sivan Toledo \and Ran Nathan \and Allert I. Bijleveld}
+\date{2020-12-15}
 
@@
 \hypertarget{processing-egyptian-fruit-bat-tracks}{%
-\chapter{Processing Egyptian fruit bat tracks}\label{processing-egyptian-fruit-bat-tracks}}
+\chapter{Processing Egyptian Fruit Bat Tracks}\label{processing-egyptian-fruit-bat-tracks}}
 

### 2021-02-01 render_books.sh
clean up tex copy code

@@ -1,14 +1,25 @@
 #!/bin/bash
 
+# remove old tex files and pdfs
+rm docs/*.pdf
+rm docs/*.tex
+
 Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2")'
 
 # to render source files
 Rscript --slave -e 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
 
 # copy pdfs and name correctly
+# remove old pdfs
+
+
 cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text`date -I`.pdf
-cp -p docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_`date -I`.pdf
-rm docs/supplementary_material.pdf
+mv docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_`date -I`.pdf
+
+# copy tex and replace old tex
+cp -p atlas-manuscript/arxiv_article.tex docs/ms_atlas_preproc_main_text_`date -I`.tex
+
+mv docs/supplementary_material.tex docs/ms_atlas_preproc_supplement_`date -I`.tex
 
 # build atlastools manual
 Rscript --slave -e 'devtools::build_manual(pkg = "../atlastools", path = "docs/")'

### 2021-02-01 docs/ms_atlas_preproc_main_text_2021-02-01.tex
main tex before reformat

\documentclass[10pt,paper=a4]{scrartcl}
\usepackage{tgtermes}
\usepackage{helvet}
\usepackage{amsmath}
\usepackage[total={6.75in, 9.5in}]{geometry}
\setlength{\parindent}{2em}
% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
\usepackage[utf8]{inputenc}

% line numbers
\usepackage[right]{lineno}

% text layout - change as needed
% \raggedright

% Remove % for double line spacing
\usepackage{setspace} 
\doublespacing

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

% graphiccs and colours
\usepackage{graphicx,color}

% \RequirePackage{natbib}
\RequirePackage[authoryear,sort]{natbib}% uncomment this for author-year bibliography
% \RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
\usepackage{listings}
\definecolor{stringstyleCol}{rgb}{0.192,0.494,0.8}
\definecolor{NumCol}{rgb}{0.686,0.059,0.569}
\definecolor{basicstyleCol}{rgb}{0.345, 0.345, 0.345}
\lstdefinestyle{customR}{
  belowcaptionskip=1\baselineskip,
  captionpos=b,
  breaklines=true,
  numbers=left,
  frame=L,
  xleftmargin=\parindent,
  language=R,
  showstringspaces=false,
  basicstyle=\small\ttfamily\color{black},
%   keywordstyle=\bfseries\color{blue!40!black},
%   commentstyle=\itshape\color{purple!40!black},
%   identifierstyle=\color{green!40!black},
%   stringstyle=\color{red}
%   keywordstyle=\color{keywordstyleCol},      % keyword style
  commentstyle=\color{commentstyleCol},   % comment style
  stringstyle=\color{stringstyleCol},      % string literal style
  literate=%
   *{0}{{{\color{NumCol}0}}}1
    {1}{{{\color{NumCol}1}}}1
    {2}{{{\color{NumCol}2}}}1
    {3}{{{\color{NumCol}3}}}1
    {4}{{{\color{NumCol}4}}}1
    {5}{{{\color{NumCol}5}}}1
    {6}{{{\color{NumCol}6}}}1
    {7}{{{\color{NumCol}7}}}1
    {8}{{{\color{NumCol}8}}}1
    {9}{{{\color{NumCol}9}}}1
}
\usepackage{ulem}
\graphicspath{ {./figures/} }
\usepackage{wrapfig}
\setlength\linenumbersep{1cm}
\renewcommand\linenumberfont{\normalfont\footnotesize\sffamily}

% \usepackage{fancyhdr}
% \pagestyle{fancy}
% \fancyhf{}
% \fancyhead[RE,LO]{Pre-Processing ATLAS Data}

% %command for custom margin size
% \def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
% \let\endchangemargin=\endlist

% document begins here
\begin{document}
\vspace*{0.35in}

% title goes here:
\begin{flushleft}
{\Large
{\sffamily \textbf\newline{A Guide to Pre-Processing High-Throughput Animal Tracking Data}}
}
\newline
% authors go here:
\\
Pratik Rajan Gupte\textsuperscript{1, 2},
Christine E. Beardsworth\textsuperscript{2},
Orr Spiegel\textsuperscript{3},
Emmanuel Lourie\textsuperscript{4},
Sivan Toledo\textsuperscript{5},
Ran Nathan\textsuperscript{4},
Allert I. Bijleveld\textsuperscript{2}
% Author 4\textsuperscript{x},
% Author 5\textsuperscript{x},
% Author 6\textsuperscript{x},
% Author 7\textsuperscript{x}
\\
\bigskip
\textbf{1} Groningen Institute for Evolutionary Life Sciences, University of Groningen, The Netherlands.
\\
\textbf{2} Deptartment of Coastal Systems, NIOZ Royal Netherlands Institute for Sea Research, The Netherlands.
\\
\textbf{3} School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv 69978, Israel.
\\
\textbf{4} Movement Ecology Lab, Department of Ecology, Evolution, and Behavior, Alexander Silberman Institute of Life Sciences, The Hebrew University of Jerusalem, 91904 Jerusalem, Israel.
\\
\textbf{5} Blavatnik School of Computer Science, Tel Aviv University, Israel.
\\
\bigskip

\textbf{Correspondence Author}
\\
Pratik R. Gupte
\\
\textbf{Address}
\\
Groningen Institute for Evolutionary Life Sciences, 
\\
Nijenborgh 7/5172.0581 9747 AG Groningen
\\
The Netherlands.
\\
\textbf{Email}
\\
pratikgupte16@gmail.com OR p.r.gupte@rug.nl
\\
\textbf{ORCID}
\\
https://orcid.org/0000-0001-5294-7819
\\
\textbf{Running Headline}
\\
Cleaning high-throughput animal tracking data

\end{flushleft}

\newpage

% now start line numbers
\linenumbers

\section*{Abstract} % abstract
\noindent
\begin{enumerate}
    \item Modern, high-throughput animal tracking studies collect increasingly large volumes of data at very fine temporal scales.
    At these scales, location error can exceed the animal's step size, confounding inferences from tracking data.
    `Cleaning' the data to exclude positions with large location errors prior to analyses is one of the main ways movement ecologists deal with location errors.
    Cleaning data to reduce location error before making biological inferences is widely recommended, and ecologists routinely consider cleaned data to be the ground-truth.
    Nonetheless, uniform guidance on this crucial step is scarce.

    \item Cleaning high-throughput data must strike a balance between rejecting location errors without discarding valid animal movements.
    Additionally, users of high-throughput systems face challenges resulting from the high volume of data itself, since processing large data volumes is computationally intensive and difficult without a common set of efficient tools.
    Furthermore, many methods that cluster movement tracks for ecological inference are based on statistical phenomena, and may not be intuitive to understand in terms of the tracked animal's biology.

    \item In this article we introduce a pipeline to pre-process high-throughput animal tracking data in order to prepare it for subsequent analysis. 
    We demonstrate this pipeline on simulated movement data to which we have randomly added location errors.
    We further suggest how large volumes of cleaned data may be synthesized into biologically meaningful ‘residence patches’.
    We then use calibration data to show how the pipeline improves its quality, and to verify that the residence patch synthesis accurately captures animal space-use.
    Finally, turning to real tracking data from Egyptian fruit bats (\textit{Rousettus aegyptiacus}), we demonstrate the pre-processing pipeline and residence patch method in a fully worked out example.

    \item To help with fast implementations of our pipeline, and to help standardise methods, we developed the \texttt{R} package \texttt{atlastools}, which we introduce here.
    Our pre-processing pipeline and \texttt{atlastools} can be used with any high-throughput animal movement data in which the high data volume combined with knowledge of the tracked individuals’ biology can be used to reduce location errors.
    The use of common pre-processing steps that are simple yet robust promotes standardised methods in the field of movement ecology and better inferences from data.
\end{enumerate}

\textbf{Keywords} ATLAS, data cleaning, movement ecology, high-throughput tracking, \textit{R} package \textit{atlastools}, residence patch

\linenumbers

\section{Introduction}

Tracking individual animals is the methodological mainstay of movement ecology, which seeks to link animal movement with internal causes, environmental factors, and resulting consequences \citep{nathan2008a, holyoak2008}. 
Investigating fine-scale environmental and social drivers and the consequences of movement requires position data from many individuals at high temporal and spatial resolution.
Such high-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}, yet ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015} routinely produce high-throughput tracking data at much lower costs.
Although high-resolution tracking inherently provides more detailed information about the true path of the tracked animal, high-throughput data presents a two-fold challenge to ecologists.
First, the location error of each position may approach or exceed the true step size of the animal compared to low-resolution tracking with the same measurement error.
This biases derived metrics such as speed and tortuosity \citep[see][]{ranacher2016, noonan2019, hurford2009, calenge2009}.
Additionally, the absolute location errors around a position introduces uncertainty when studying the location of an animal relative to fixed landscape features (e.g. roads) or mobile elements (e.g. other tracked individuals).
Users have two main options to improve data quality: making inferences after modelling the system-specific location error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
The first approach may be limited by the animal movement models it can fit to the data \citep{fleming2014a, noonan2019, fleming2020}, and may be beyond the computational capacity of common hardware, leading users to prefer data cleaning instead.
When attempting data cleaning, the second challenge of high-throughput tracking reveals itself: the large number of observations themselves \citep{weiser2016, toledo2020}.
Having large volumes of data to clean makes manual identification and removal of errors from individual tracks prohibitively time consuming, incentivising automation based on a protocol.

Pre-processing steps must justifiably discard large location errors, also called outliers, from tracks (analogous to reducing false positives) while avoiding the overzealous rejection of valid animal movements (analogous to reducing false negatives).
How well researchers balance these imperatives has consequences for downstream analyses \citep{stine2001}.
For instance, small-scale resource selection functions can easily infer spurious preference and avoidance effects when there is uncertainty about an animal's true position and movement \citep{visscher2006}.
Ecologists recognise that tracking data are imperfect observations of the underlying process of animal movement, yet they implicitly consider cleaned data equivalent to be the ground-truth.
This assumption is reflected by popular statistical methods in movement ecology such as Hidden Markov Models (HMMs) \citep{langrock2012}, stationary-phase identification methods \citep{patin2020a}, or step-selection functions (SSFs) \citep{barnett2008, signer2017, avgar2016}, which expect low location errors relative to real animal movement (i.e., a high signal-to-noise ratio).
This makes the reproducible, standardised removal of location errors crucial to any animal tracking study.
While gross errors are often removed by positioning-system algorithms, ‘reasonable’ errors often remain to confront end users \citep{fischler1981, weiser2016, ranacher2016}.
Location errors may be exacerbated by the conditions of deployment, with systematic differences in location error among different landscape types  \citep[see][]{lewis2007, deon2002, frair2004}. 
% For instance, the physical features of an animal’s habitat such as hills or woodland may impede the transmissions of radio-telemetry tags (Beardsworth et al. \textit{in prep.}), while canopy cover can lead to location errors in GPS tracking \citep{lewis2007}.
Standardised pre-processing steps should thus be general enough to tackle animal movement data recovered from a range of environments.

Despite the importance and ubiquity of reducing location errors in tracking data, movement ecologists lack formal guidance on this crucial step.
Pre-processing protocols are often system-specific, or have limited availability, or may not be easily tractable for mainstream computing hardware and software.
Furthermore, filtering out positions on their location error estimates may not be sufficient to recover a good track estimate when location error does not correspond to biological realism in movement \citep{weiser2016, ranacher2016}.
This makes identifying and removing implausible movements from a track an important component of recovering true animal movement \citep{bjorneraas2010}.
Even after removing large location errors and evidently unrealistic movement, a track may comprise of positions that are distributed around the true animal location.
Discarding these positions as `untrue' would lead to excessive data loss, but treating them as `real' locations would lead to unrealistic estimates of metrics such as speed \citep{noonan2019}.
The large data-volumes of high-throughput tracking allow for a neat solution: tracks can be `smoothed' to reduce small location errors that have remained undetected.
This large data volume becomes a challenge when users seek to examine animal space-use in relation to relevant environmental covariates, such as the predictors of prolonged residence in an area \citep[see][]{bracis2018}.
First, statistical modelling using high-throughput data may suffer from pseudo-replication since both the response (e.g. step-length) and the predictors (e.g. resource landscape values) are spatio-temporally auto-correlated and hence non-independent \citep{aarts2008, fleming2014a,bijleveld2016, oudman2018, harel2016}.
Second, fitting statistical models to datasets with many millions of observations may require significant innovation in mainstream statistical software \citep[e.g. ][]{wood2015}, and users may instead want to reduce data volumes by thinning or clustering.

Here, we present a hands-on guide to pre-processing high-throughput tracking data, and demonstrate a relatively simple data cleaning pipeline to prepare these data for subsequent analyses (see Fig. 1). 
We take two important considerations into account, (1) that methods must be computationally efficient, and (2) that the pre-processing steps should be easily understood and reproduced.
With these in mind we have developed the \texttt{R} package \texttt{atlastools} \citep{gupte2020a, rcoreteam2020}, which implements important steps of the pre-processing pipeline.
\texttt{R} is the computational environment of choice in movement ecology \citep{joo2020} and formalising tools as an \texttt{R} package improves portability and reproducibility \citep{marwick2018}.
\texttt{atlastools} functions are easy to use and understand, and rely on the efficiency of the \texttt{R} package \texttt{data.table} \citep{dowle2020}.
We begin by showing how the pipeline can clean simulated tracks to which we have artificially introduced location error \citep{gurarie2017}.
Beginning with basic spatio-temporal filtering of positions, we cover filtering movement tracks, and reducing the effect of location error with a median smooth.
Then, we suggest one solution to issues of computational tractability, which is to synthesise residence patches \citep[\textit{sensu}][]{bijleveld2016, oudman2018, barraquand2008} from clusters of spatio-temporally proximate positions, and so reveal important aspects of animal space use.
Using calibration data from a manually transported ATLAS tag, we demonstrate how the residence patch method in \texttt{atlastools} accurately identifies areas of prolonged residence under real field conditions.
Finally, we turn to data from Egyptian fruit bats (\textit{Rousettus aegyptiacus}) tracked in the Hula Valley, Israel, to show a fully worked out example of the pre-processing pipeline and the residence patch method.

\begin{figure}
    \centering
    \includegraphics[width=0.99\textwidth]{figures/fig_01_recipe.png}
    \caption{A general, modular pipeline for pre-processing high-throughput tracking data from raw localisations to cleaned data, and optionally into residence patches. 
    Users should begin by plotting their data to determine its status, for instance whether it needs to be susbet within a certain time interval, or whether there are obvious location errors.
    Users can then pre-process their data using the appropriate module, following the pre-processing steps corresponding to each module.
    The \texttt{atlastools} function that may be used to implement each pre-processing step is shown in the grey boxes underneath each step.
    Users may use the pipeline in full or just the module most appropriate for their data.
    In all cases, users are strongly encouraged to visualise their data and scan it for location errors as a first step, and to visualise it along the way.
    }
    \label{fig:figure_pipeline}
\end{figure}

\section{Pipeline Overview, Usage, and Simulating Data}

\subsection{The Pipeline and \texttt{atlastools}}

We lay out a modular pipeline for pre-processing raw high-throughput tracking data using the \texttt{R} package \texttt{atlastools} (Fig. 1).
While the pipeline and package were designed with ATLAS systems in mind, the principles and functions can be used with any high-throughput tracking data.
Users may follow the pipeline in full, or implement the module most suitable for their data.

\begin{enumerate}

    \item After data access, we first encourage users to visualise their data to check for evident location errors \citep{slingsby2016}.
    \item Users can install \texttt{atlastools} using the \texttt{install\_github} function from the \texttt{R} package \texttt{remotes}.
    The package can be installed using the \texttt{R} command

    \texttt{remotes::install\_github("pratikunterwegs/atlastools")}.

    Package functions are prefixed `\texttt{atl\_}', and rely heavily on the \texttt{data.table} package (Dowle and Srinivasan 2020).
    Certain functions modify the input data "in place", and users must work on a copy of their data to preserve both the original and cleaned data (details in the Supplementary Material).
    Since ATLAS systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} assumes a coordinate system in metres, and users must convert their geographic coordinates to metres.
    \item Users with a specific area or time of interest can use simple spatio-temporal range filters to select positions (see \textsc{Spatio-Temporal Filtering}).
    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by an error measure, or by the plausibility of associated movement metrics (see \textsc{Filtering to Reduce Location Errors}).
    \item Users should then reduce small-scale location errors by applying a median smooth (see \textsc{Smoothing and Thinning Data}).
    \item Users who need uniformly thinned data can then thin by either aggregating or resampling (see \textsc{Smoothing and Thinning Data}).
    At this stage, the data are ready for a number of popular statistical treatments such as Hidden Markov Model-based classification \citep{michelot2016}.
    \item Finally, users wishing to study prolonged animal residence in an area can classify their data into residence patches based on the movement ecology of their study system, after filtering out non-stationary positions (see \textsc{Synthesising Movement Tracks into Residence Patches}).
    % Residence patches are spatio-temporally independent clusters of positions that represent prolonged residence of a tracked individual in an area, and which can indicate aspects of space use such resource exploitation \citep{barraquand2008, bijleveld2016, oudman2018}.

\end{enumerate}

\subsection{Simulating Movement Tracks}

To demonstrate the pipeline, we simulated a realistic movement track of 5,000 positions (unbiased correlated velocity model; UCVM) using the \texttt{R} package \texttt{smoove} \citep[][see Fig. 2.a]{gurarie2017}.
We added three kinds of error to the simulated track: (1) normally distributed small-scale offsets to the X and Y coordinates independently, (2) normally distributed large-scale offsets to a random subset (0.5 \%) of the positions, and (3) large-scale displacement of a continuous sequence of 300 of the 5,000 positions (indices 500 -- 800) (Fig. 2.a).
To demonstrate the residence patch method, we chose to simulate three independent rotational/advective correlated velocity movement (RACVM) tracks of 500 positions each ($\omega$ = 7, initial velocity = 0.1, $\mu$ = 0; see Gurarie et al. 2017), and connected them together with a roughly linear path (see Fig. 6.a).
RACVM models approximate the tracks of soaring birds which circle on thermals over a relatively small area, and move between thermals \citep[`thermalling'; ][]{gurarie2017, harel2016}.
This complex track structure provides a suitable challenge for the residence patch method and helps to demonstrate its generality.

\section{Spatio-Temporal Filtering}

\subsection{Spatial Filtering Using Bounding Boxes and Polygons}

First, data can be filtered to exclude positions outside the spatial bounds of a study area.
This simply involves comparing position coordinates with the range of acceptable coordinates (the bounding box), and removing those positions outside them (Fig. 2.b; Listing 1). 
A bounding box filter retains data from within a subset of the tracking range without the need for a geospatial representation such as a shapefile.
This can help remove unreliable data from a tracking system that is less accurate beyond a certain range (e.g. ATLAS; Beardsworth et al. \textit{in prep.}).
In some special cases, users may wish to \textit{remove} positions inside a bounding box, either because movement behaviour within an area is not the focus of a study, or because positions recorded within an area are known to be erroneous.
An example of the former is studies seeking to study transit behaviour between features which can be approximated by their bounding boxes. 
Instances of the latter are likely to be system specific, but are known from ATLAS systems (Bijleveld et al. \textit{in prep.}). 
One drawback to bounding box filters is that they are restricted to rectangular areas.
Users seeking to filter for areas with other geometries, such as a circular or irregular study area, need a geometric intersection between their data and a spatial representation of the area of interest (e.g. shapefile, geopackage, or \texttt{sf}-object in \texttt{R}).
The \texttt{atlastools} function \texttt{atl\_filter\_bounds} implements bounding box and explicit spatial filters and accepts X and Y coordinate ranges, an \texttt{sf}-polygon object \citep{pebesma2018}, or any combination of the three to filter the data (Listing 1).
Multipolygon objects are supported, allowing data from different areas to be selected.
When both coordinate ranges and a polygon are provided, the data is first filtered by the ranges and then the polygon.
The boolean function argument \texttt{remove\_inside} determines whether positions inside the bounds are retained (\texttt{FALSE}) or removed (\texttt{TRUE}).

\begin{lstlisting}[float,floatplacement=h!,language=R, style=customR, caption = {
    The \texttt{atl\_filter\_bounds} function removes positions outside an area defined by coordinate ranges, a polygon, or all three (\texttt{remove\_inside = FALSE}), or positions inside the area (\texttt{remove\_inside = TRUE}).
    The arguments \texttt{x} and \texttt{y} determine which columns are considered the X and Y coordinates, the arguments \texttt{x\_range} and \texttt{y\_range} determine the acceptable range of coordinates in a reference system based in metres, while the \texttt{sf\_polygon} argument allows the data to be filtered by the user specified \texttt{sf-(MULTI)POLYGON} object. 
    \texttt{atl\_filter\_bounds} returns a filtered \texttt{data.table}, which must be saved as an object (here, \texttt{filtered\_data}).}]
filtered_data <- atl_filter_bounds(data = data,
                    x = "X", y = "Y",
                    x_range = c(x_min, x_max),
                    y_range = c(y_min, y_max),
                    sf_polygon = your_polygon,
                    remove_inside = FALSE)
\end{lstlisting}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_02_bounds.png}
    \caption{\textbf{(a)} A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the \texttt{smoove} package (Gurarie et al. 2017). Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track. 
    The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (black line) as closely as possible.  
    \textbf{(b)} Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the \texttt{atlastools} function  \texttt{atl\_filter\_bounds}.
    Setting the function argument \texttt{remove\_inside = FALSE} retains positions within user supplied bounds (dashed grey lines; green points), and excludes those outside (grey points).}
    \label{fig:figure_simulated_data}
\end{figure}

\subsection{Temporal and Spatio-temporal Filters}

Temporal filtering can exclude positions from periods during tracking when data are expected to be unreliable, either due to abnormal movement behaviour or poor tracking quality. 
An animal’s movement may be non-representative when it has been fitted with a tracker but has not been released, leading to artificial stationary positions, or in the time shortly after release when its movement may be influenced by the stress of capture and handling (see example implementation in \texttt{amt}; \citet{signer2019}). 
In these cases, all positions from a short time after release (e.g. 24 hours) can be excluded. 
Periods of poor tracking quality may result from system malfunctions, and temporal filters can be applied to exclude such chunks from the datastream. 
Temporal filters can be combined with spatial filters to exclude time-location combinations which would be unrealistic for the tracked animal. 
For example, red knots in the Dutch Wadden Sea congregate at communal roosts during high-tide, when their foraging grounds on the inter-tidal mudflats are inaccessible \citep{vangils2006}.
A spatio-temporal filter that retains positions inside the bounds of roosts during high-tide can quickly return data that reveals individual presence at high-tide roosts, while excluding positions that erroneously place individuals on the inundated mudflats.
% Howevever, in this example, such a filter would also exclude potentially exploratory movements over the flooded mudflats.
% We caution users to be certain that filters do not exclude positions that reveal interesting aspects of their animals' movement.
Users should apply filters in sequence rather than all at once, and visualise the output after each filtering step (`sanity checks').

The atlastools function \texttt{atl\_filter\_covariates} allows convenient filtering of a dataset by any number of logical statements (Listing 2).
This function can be used to easily filter timestamps in a range, as well as combine simple spatial and temporal filters.
It accepts a character vector of \texttt{R} expressions that each return a logical vector (i.e., \texttt{TRUE} or \texttt{FALSE}; Listing 2).
The function returns only those data which satisfy each of the filter conditions. 
Users must ensure that the filtering variables exist in their dataset in order to avoid errors.

\begin{lstlisting}[float, language=R, style=customR, caption = {
    The \texttt{atl\_filter\_covariates} function can be used both as a simple temporal filter, and also as a combined spatio-temporal filter. 
    Filter predicates are passed to the \texttt{filters} argument as a character vector, each of which is then evaluated as an \texttt{R} expression in the context of the data supplied. 
    Only rows in the data satisfying all the conditions passed as filters are retained. 
    Users must make sure the filter variables exist in their dataset.
    Here, the first example shows how nighttime data can be retained using a predicate (\texttt{inrange} from \texttt{data.table}) that determine whether the value of `hour' is between 6 and 18. 
    The \texttt{!} sign indicates that the negative of the predicate should be returned, i.e., \texttt{TRUE} when the hour is between 6 PM and 6 AM, but \texttt{FALSE} when the hour is between 6 AM and 6 PM.
    The second example shows the use of multiple filter statements; this data will be filtered to be within the range of times bounded by \texttt{t\_min} and \texttt{t\_max}, and with X coordinates between \texttt{x\_min} and \texttt{x\_max}.
    The \texttt{between} function is from \texttt{data.table}.}]
night_data <- atl_filter_covariates(data = dataset,
                filters = c("!inrange(hour, 6, 18)"))

data_in_area <- atl_filter_covariates(data = dataset,
                    filters = c("between(time, t_min, t_max)",
                                "between(x, x_min, x_max)"))
\end{lstlisting}

\section{Filtering to Reduce Location Errors}

\subsection{Filtering on Quality Covariates}

Tracking data covariates can be good indicators of the reliability of calculated positions (Beardsworth et al. \textit{in prep.}).
For intance, in ATLAS systems the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
GPS systems' location error may be similarly indicated indirectly by the number of satellites involved in the localisation, or directly by an error measure such as the Horizontal Dilution of Precision (HDOP).
ATLAS and other TOA systems also calculate direct measures of location error during localisation: \texttt{VARX}, \texttt{VARY}, and \texttt{COVXY}, which are the variance of the X and Y coordinates, and the covariance of the X and Y coordinates, respectively \citep{maccurdy2009, maccurdy2019, weiser2016}.
A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
\begin{linenomath*}
    \begin{equation*}
        SD = \sqrt{{VARX} + {VARY} + 2 \times {COVXY}}
     \end{equation*}
\end{linenomath*}
Filtering on the position-specific standard deviation allows removing unreliable positions, and the filter can be applied using \texttt{atl\_filter\_covariates}.

\begin{lstlisting}[float, language=R, style=customR, caption = {
    Filtering ATLAS data on position covariates. 
    The \texttt{filters} argument accepts a character vector with the logical statements. 
    The function only retains data for which \textit{all} the conditions are satisfied; here that is positions calculated using $>$ 3 base stations (\texttt{NBS}), with location error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.}]
filtered_data <- atl_filter_covariates(data = data,
                    filters = c("NBS > 3",
                                "SD < 100",
                                "between(day, 5, 8)"))
\end{lstlisting}

\subsection{Filtering Unrealistic Movement}

Filtering on system-generated measures of error may not result in the removal of all erroneous positions, and data may remain which would require biologically implausible movement.
Users are encouraged to visualise their tracks before and after filtering point locations, and especially to `join the dots' and connect consecutive positions with lines.
Whether the resulting track looks realistic is ultimately a subjective human judgement, but one which implicitly integrates prior knowledge of the movement ecology of the study species to ask, `Does the animal move this way?'.
Segments which appear to represent unrealistic animal movement are often obvious to researchers with extensive experience of the study system \citep[the non-movement approach; see][]{bjorneraas2010}.
Since it is both difficult and prohibitively time consuming to exactly reproduce expert judgement when dealing with large volumes of tracking data from multiple individuals, some automation is necessary.
Users should first manually examine a representative subset of tracks and attempt to visually identify problems --- either with individual positions, or with subsets of the track --- that persist after basic filtering.
Once such problems are identified, users can conceptualise algorithms that can be applied to their data to resolve them.

An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 3.a).
Point outliers are characterised by artificially high speeds between the outlier and the positions before and after \citep[called incoming and outgoing speed, respectively][]{bjorneraas2010}, lending a `spiky' appearance to the track.
Removing spikes is simple: remove positions with extreme incoming and outgoing speeds.
Users must first define plausible upper limits for speed and turning angle for the study species \citep{calenge2009, seidel2018}.
Here, it is important to remember that speed estimates are scale-dependent; high-throughput tracking typically overestimates the speed between positions where the animal is stationary or moving slowly due to small-scale location errors \citep{ranacher2016, noonan2019}. 
Even after data with large location errors have been removed by filters, it is advisable to begin with a liberal (high) speed threshold that excludes only the most unlikely of speeds.
Estimates of maximum speed may not always be readily obtained for all species, and an alternative is to use a data-driven threshold such as the 95\textsuperscript{th} percentile of speeds from the track.
Once a speed threshold $S$ has been chosen, positions with incoming \textit{and} outgoing speeds $\geq S$ may be identified as spikes and removed.
Some species can realistically achieve speeds $\geq S$ in fast transit segments when assisted by environmental factors, such as birds with tailwinds, and a simple filter on incoming and outgoing speeds would exclude this valid data.
To avoid removing real, fast transit segments while still excluding spikes we suggest combining the speed filter with a filter on the turning angles of each position \citep[see][]{calenge2009}.
This combined filter assumes that positions in high-throughput tracking with both high speeds and large turning angles are likely to be due to location errors, since most species are unable to turn sharply at high speed.
Users can then retain only those positions whose incoming and outgoing speeds are both $< S$ which or satisfy the condition $\theta < A$, where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
The removal of spikes is implemented using the \texttt{atl\_filter\_covariates} function (Listing 4).
Many other track metrics may be used to identify implausible movement and on which data may be filtered \citep{seidel2018}.

\begin{lstlisting}[float, language=R, style=customR, caption = {
    Filtering a movement track on incoming and outgoing speeds, and on turning angle to remove unrealistic movement.
    The functions \texttt{atl\_get\_speed} and \texttt{atl\_turning\_angle} are used to get the speeds and turning angles before filtering, and assigned to a column in the data (assignment of \texttt{speed\_out} is not shown).
    The filter step only retains positions with speeds below the speed threshold $S$ \textit{or} angles above the turning angle threshold $\theta$, i.e., positions where the animal is slow but makes sharp turns, and data where the animal moves quickly in a relatively straight line.}]
data$speed_in <- atl_get_speed(data,
                x = "x", y = "y",
                time = "time", type = c("in"))

data$angle <- atl_turning_angle(data,
                  x = "x", y = "y", time = "time")

filtered_data <- atl_filter_covariates(data = data,
        filters = c("(speed_in < S & speed_out < S) | angle < A"))
\end{lstlisting}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_03_filter_speed.png}
    \caption{Reducing large-scale outliers in a movement track. The goal is to remove positions far away from the canonical track (black line), and retain those with only reasonable displacements from the canonical track (green points). 
    \textbf{(a)} location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles). 
    The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
    Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
    Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
    \textbf{(b)} The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
    While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult. 
    Users are cautioned to frequently check this and similar methods' results.}
    \label{fig:figure_filter_bounds}
\end{figure}

Sometimes entire susbets of the track may be affected by the same large-scale location error.
For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 3.a, b).
These cannot be corrected by targeted removal of individual positions, as in Bjorneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
Since filtering individual positions will not suffice, algorithms to correct such errors must take a track-level view, and target the displaced sequence overall.
Track-subset algorithms are likely to be system-specific, and may be challenging to conceptualise or implement.
In the case of prolonged spikes, one relatively simple solution is identifying the bounds and removing positions between them.
We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
Users are strongly encouraged to visualise their data before and after applying this method, and we caution against relying on this method if data are heavily distorted by errors affecting entire track-subsets.

\begin{lstlisting}[float, language=R, style=customR, caption = {
    Removing prolonged spikes from a movement track. 
    The important function arguments here are \texttt{point\_angle\_cutoff} ($A$), \texttt{reflection\_speed\_cutoff} ($S$), and \texttt{est\_ref\_len}, the maximum number of positions after the inner bound that are candidates for the end of the prolonged spike, i.e., the outer bound. 
    If the prolonged spike ends after less than $N$ positions, the true end point is used as the outer bound of the spike.
    However, the algorithm behind this function fails when the prolonged spike ends after more than $N$ positions. 
    Users are advised to use a liberally large value of N in the \texttt{est\_ref\_len} argument; 1,000 may be appropriate for 3s interval data.
    Further, users are cautioned against relying on such algorithms for severely distorted data.}]
filtered_data <- atl_remove_reflections(data = track_data,
                       x = "x", y = "y", time = "time",
                       point_angle_cutoff = A,
                       reflection_speed_cutoff = S,
                       est_ref_len = N)
\end{lstlisting}

\section{Smoothing and Thinning Data}

\subsection{Median Smoothing}

After filtering out large location errors may, the track may still look ‘spiky’ at small scales, and this is due to smaller location errors.
These smaller errors are challenging to remove since their covariates (such as speed and turning angles) are within the expected range of movement behaviour for the study species. 
The large data volumes of high-throughput tracking allow users to resolve this problem by smoothing the positions. 
A ‘smooth’ works by approximating the value of an observation based on neighbouring values.
For a one-dimensional series of observations, the neighbouring values are the $K$ observations centred on each index value $i$.
The range $ {i - (K-1)/2} \ldots {i + (K-1)/2} $ is referred to as the moving window as it shifts with $i$, and $K$ is called the moving window size.
A common smooth is nearest neighbour averaging, in which the value of an observation $x_i$ is the average of the moving window $K$.
The median smooth is a variant of nearest neighbour averaging which uses the median rather than the mean, and it is robust to outliers (\citeauthor{tukey1977} 1977).
The median smoothed value of the X coordinate, for instance, is
\begin{linenomath*}
    \begin{equation*}
        X_i = Median(X_{i - (K-1)/2} \ldots X_{i + (K-1)/2})
     \end{equation*}
\end{linenomath*}
Users can apply a median smooth with an appropriate $K$ independently to the X and Y coordinates of a movement track to smooth it (see Fig. 4.a -- e). 
% The first and last $(K - 1)/2$ values for a moving window size of $K$ are assigned a value using Tukey's (1977) end-point smoothing rule as implemented in \texttt{R} \citep{hardle1995}.
The median smooth is robust to even very large temporal and spatial gaps, and does not interpolate between positions when data are missing. 
Thus it is not necessary to split the data into segments separated by periods of missing observations when applying the filter (see Fig. 4).

Smoothing does not change the number of observations, but does decouple the coordinates from some of their covariates.
For instance, smoothing breaks the relationship between a coordinate and the location error estimate around it (\texttt{VARX}, \texttt{VARY}, and \texttt{SD} in ATLAS systems, or HDOP in GPS tracking).
This makes subsequent filtering on covariates of data quality unreliable, and smoothed data are unsuitable for use with methods that model location uncertainty \citep{noonan2019, fleming2014a, fleming2020, calabrese2016}.
Furthermore, while larger $K$ may result in smoother tracks (Fig. 4.b -- e), one drawback of using a large $K$ is that short, quick forays away from the main track are likely to be smoothed away, leading to a loss in detail of the individual's small-scale movement.
Users must themselves judge how best to trade large-scale and small-scale accuracy, and choose $K$ accordingly.
One empirical way to compare $K$ is by calculating the root mean squared error (RMSE) for different $K$ on the same data.
% In general, large $K$s lead to unrealistic tracks, both because the first and last $(K - 1) / 2$ positions are handled as special cases, and because the local median approaches the global median at larger $K$.
Median smoothing is provided by the \texttt{atlastools} function \texttt{atl\_median\_smooth}, with the only option being the moving window size, which must be an odd integer (Listing 6).

\begin{lstlisting}[float, language=R, style=customR, caption = {
    Median smoothing a movement track using the function \texttt{atl\_median\_smooth} function with a moving window \textit{K = 5}. 
    Larger values of $K$ yield smoother tracks, but $K$ should always be some orders of magnitude lower than the number of observations.}]
atl_median_smooth(data = track_data,
                  x = "x", y = "y",
                  time = "time",
                  moving_window = 5)
\end{lstlisting}

% median smooth figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_04_median_smooth.png}
    \caption{Median smoothing position coordinates reduces small-scale location error in tracking data. 
    \textbf{(a)} The goal of this step is to approximate the simulated canonical track (black line), given positions with small-scale error remaining from previous steps (green points). 
    The canonical track is shown `zoomed in' for better representation, and the missing data at the top of the track (line with no points) is from where the prolonged spike was earlier removed.
    Median smoothing the given position coordinates ((a) green points) over a moving window ($K$) of \textbf{(b)} 3, \textbf{(c)} 5, \textbf{(d)} 11, and \textbf{(e)} 21 positions respectively, yields approximations (purple lines) of the canonical track.
    Users are cautioned that there is no correct $K$, and they must subjectively choose a $K$ which most usefully trades small-scale details of the track for large-scale accuracy.}
    \label{fig:figure_median_smooth}
\end{figure}

\subsection{Thinning Movement Tracks}

Most data at this stage is technically ‘clean’, yet its volume alone may pose challenges for lower-specification or older hardware and software if these are not optimised for efficient computation.
Thinning data need not compromise researchers' ability to answer scientific questions with them; for instance, social interactions lasting 1 -- 2 minutes would still be detected on thinning from a sampling interval of 1 second to 1 minute.
Indeed, temporal auto-correlation may hinder some methods such as the estimation of home-ranges or step-selection functions \citep{fleming2014a, dupke2017}.
Added to the requirement of uniform sampling intervals by many methods in the field, evenly reducing data volumes is a worthwhile endeavour \citep[e.g.][]{fleming2014a, michelot2016, avgar2016}.
Two plausible approaches here are resampling and aggregation, and both approaches begin with identifying time-interval groups (e.g. of 1 minute).
Resampling picks one position from each time-interval group.
Aggregation involves computing the mean values of all covariates for positions within a time-interval group.
Both approaches yield one position per time-interval group.
Categorical variables, such as the habitat type associated with each position can be aggregated using a suitable measure such as the mode.

The aggregation method is less sensitive to selecting point outliers by chance than resampling.
When users want to account for location error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location error when thinning is important.
In ATLAS systems the location error (\texttt{SD}; see \textsc{Filtering on Position Covariates}) is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
The error around each coordinate (\texttt{VARX} or \texttt{VARY} in ATLAS systems) can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
\begin{linenomath*}
    \begin{equation*}
        Var(X)_{agg} = \left( \sum_{i=1}^{i=N} Var(X)_i \right) / N ^ 2
    \end{equation*}
\end{linenomath*}
Similarly, the overall location error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance ($SD ^ 2$):
\begin{linenomath*}
    \begin{equation*}
        SD_{agg} = \sqrt{ \left( \sum_{i=1}^{i=N} SD_i^2 \right) / N ^ 2  }
    \end{equation*}    
\end{linenomath*}

Users may question why thinning should be implemented after cleaning steps, when aggregation can obtain consensus positions over an interval, correctly propagate the location error, and also reduce data volumes.
We caution users that thinning causes an extensive loss of small-scale detail in the data.
Additionally, thinning prior to excluding unrealistic movement and smoothing can lead to estimates of essential metrics --- such as speed --- that are substantially different from the true value (Noonan et al. 2019; see Fig. 5.c).
The mis-estimation of track metrics could have knock-on consequences for the implementation of subsequent filters based on detecting unrealistic movement.
However, thinning before data-cleaning may have its place as a useful step before exploratory visualisation of the movement track, since reduced data volumes are easier to handle for plotting software.

Thinning is implemented in \texttt{atlastools} using the \texttt{atl\_thin\_data} function, with either aggregation or resampling (specified by the \texttt{method} argument) over an interval using the \texttt{interval} argument.
The column of timestamps must be named `time' and column classes except the identity columns (\textit{see below}) should allow averaging.
The `resample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
Using `resample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
In both cases, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
Due to its development for ATLAS systems, the `aggregate' assumes the variance around coordinates is named \texttt{VARX} and \texttt{VARY}, and standard deviation around each position is named \texttt{SD}.
These columns must be present together for the function to correctly handle the error.
If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
Grouping variables' names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 7).

\begin{lstlisting}[float, language=R, style=customR, caption = {Code to thin data by aggregation in \texttt{atlastools}. The method can be either "aggregate" or "resample". 
The time interval is specified in seconds, while the \texttt{id\_columns} allows a character vector of column names to be passed to the function, with these columns used as identity variables.
Both methods return a dataset with as many rows as there are time-intervals.
While the resampling method retains all columns, the aggregation method drops the ATLAS specific columns specifying covariance in X and Y coordinates (COVXY), and location error (SD).}]
thinned_data <- atl_thin_data(data,
                    interval = 60,
                    id_columns = c("animal_id"),
                    method = "aggregate")
\end{lstlisting}

% figure for aggregation thinning
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_05_thinning.png}
    \caption{Thinning a movement track using aggregation preserves track structure while reducing data volume, but thinning without removing gross location errors and unrealistic movement affects track metrics such as speed.
    (a, b) Movement track positions with an interval of 1 second (green points) aggregated over intervals of (a) 10 and (b) 30 seconds (purple circles), after removing large location errors or filtering unrealistic movement.
    The canonical track is shown as a black line in both panels, while the red line shows the track connecting thinned positions.
    (c) Boxplots show the median and interquartile ranges for speed estimates of tracks aggregated over intervals of 3, 10, 30, and 120 seconds, but without the removal of location errors.
    For comparison, the median and 95\textsuperscript{th} percentile of speed of the canonical track are shown as solid and dashed horizontal lines, respectively.
    Green circles and associated error bars show, in contrast, the median ($\pm$ standard deviation) speed for the same aggregation interval, but after removing large- and small-scale location errors (median smooth $K$ = 11).
    Removing location errors and unrealistic movement before aggregation substantially improves speed estimates, especially at smaller aggregation intervals, as it removes the apparent spikes that contribute to over estimation of speed if not filtered.}
    \label{fig:figure_thinning}
\end{figure}

\section{Synthesising Movement Tracks into Residence Patches}

\subsection{The Residence Patch Algorithm}

Tracking data that are still too large for statistical packages, or have strong autocorrelation, can benefit from segmentation-clustering into `residence patches' \citep{bijleveld2016, oudman2018, barraquand2008}.
Making the patch the unit of observation conveniently sidesteps pseudo-replication while and reduces computational requirements.
Furthermore, metrics such as the distance travelled within and between patches can help compare broad-scale individual movement strategies.
% Here, high-throughput tracking every 1s -- 30s has its advantages: users may be permitted the assumption that an individual is truly absent from a location where it was not reported, and that at least during the tracking period, such absences are not simply artefacts of the sampling interval.

First, users should identify positions representing bouts of stationary behaviour, for instance on their speed or residence time \citep{bracis2018}.
Patch identification assesses whether consecutive positions and the bouts they comprise are spatio-temporally independent, and clusters them together if they are not.
The splitting and lumping of positions into bouts, and bouts into patches is based on simple user specified thresholds --- the distance and the time interval between positions (and bouts) beyond which they should be considered independent.
% Patches are considered independent if either the distance or time between them exceeds their respective thresholds.
Users are encouraged to base these thresholds in the movement habits of their study species.
For example, residence patch classification of red knot movement tracks considers consecutive stationary positions independent if they are 20m apart and considers consecutive bouts independent if the distance between them $\geq$ 100m, or the time difference between them $\geq$ 30 minutes (Listing 8).
The 20m distance represents a maximum speed of 6.667 m/s between positions, above which the individual is more likely in transit.
The 100m and 30 minute thresholds are chosen to account for potentially missing data between bouts; if a track ends abruptly and then reappears $\geq$ 100m away or $\geq$ 30 minutes later, this is more safely considered a new residence patch.

A cleaned movement track can be classified into residence patches using the function \texttt{atl\_res\_patch} (see Fig. 6.c).
\texttt{atl\_res\_patch} requires three parameters: (1) the distance threshold between positions (called \texttt{buffer\_size}), (2) the distance threshold between clusters of positions (called \texttt{lim\_spat\_indep}), and (3) the time interval between clusters (called \texttt{lim\_time\_indep}).
Clusters formed of fewer than a minimum number of positions can be excluded.
% The exclusion of clusters with few positions can help in removing bias due to short stops, but if such short stops are also of interest, they can be included by reducing the \texttt{min\_fixes} argument.
Our residence patch algorithm is capable of correctly identifying clusters of related residence points from a movement track (Fig. 7.a, 7.b).
This includes clusters where the animal is relatively stationary (orange and green patches, Fig. 7.c), as well as clusters where the animal is moving slowly (blue patch, Fig. 7.c).
This flexibility is especially useful when studying movements that may represent two different modes of the same behaviour, for instance, area-restricted search, as well as slow, searching behaviour with a directional component.

The function \texttt{atl\_patch\_summary} can be used to extract patch-specific summary data such as the median coordinates, the patch duration, the distance travelled within the patch, and the patch area.
Position covariates such as speed may also be summarised patch-wise by passing covariate names and  summary functions as character vectors to the \texttt{summary\_variables} and \texttt{summary\_functions} arguments, respectively.
Setting the \texttt{which\_data} argument to \texttt{"spatial"}, returns \texttt{sf\ MULTIPOLYGON} objects, and setting \texttt{which\_data = "points"} returns the positions in each patch, with patch-specific covariates.
% The advantage of this latter option is that it can be matched to the larger dataset of positions to more accurately calculate distances between patches.

\begin{lstlisting}[float, language=R, style=customR, caption ={The \texttt{atl\_res\_patch} function can be used to classify a track into residence patches. The arguments \texttt{buffer\_radius} and \texttt{lim\_spat\_indep} are specified in metres, while the \texttt{lim\_time\_indep} is provided in minutes. In this example, specifying \texttt{summary\_variables = c("speed")}, and \texttt{summary\_functions = c("mean", "sd")} will provide the mean and standard deviation of instantaneous speed in each residence patch. The \texttt{atl\_patch\_summary} function is used to access the classified patch in one of three ways, here using the \texttt{summary} option which returns a table of patch-wise summary statistics.}]
patches <- atl_res_patch(data = track_data,
                buffer_radius = 10,
                lim_spat_indep = 100,
                lim_time_indep = 30,
                min_fixes = 3,
                summary_variables = c("speed"),
                summary_functions = c("mean", "sd"))

patch_summary <- atl_patch_summary(patch_data = patches,
                    which_data = "summary",
                    buffer_radius = 10)
\end{lstlisting}

% first residence patch figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_06_residence.png}
    \caption{A rotation-advection movement track, similar to the movement of a soaring bird (Gurarie et al. 2017) classified into residence patches with the \texttt{atlastools} function.
    (a) The goal is to segment-cluster the track into areas of prolonged residence (clusters of points) while leaving out the transit between them.
    (b) A plot of residence time against time (solid line; Bracis et al. 2018) shows how the residence patch algorithm segments and clusters positions of prolonged residence. 
    Regions are shaded by the temporal bounds of each residence patch.
    The arguments passed to \texttt{atl\_res\_patch} determine the clustering, and can be adjusted to get a result that fits the study system.
    Users are cautioned that there are no `correct' arguments.
    (c) The residence patch method correctly identifies clusters of positions where the individual is relatively stationary (orange and green patches), as well as positions where it is slowly moving (blue patch).
    This is especially useful when studying more complex behaviour such as area-restricted search, which may have a directional component.
    The function \texttt{atl\_patch\_summary} conveniently presents the classified data for further use, either as an \texttt{sf} object (left: coloured polygons around clustered points), or as summary data of each patch (right: coloured circles).
    While the \texttt{sf} object retains the patch geometry as well as the patch attributes (such as duration), the summary data discards the patch geometry.
    Both spatial and summary objects lose the details of movement between patches (compare black lines of the canonical track [left], and lines connecting patch centroids [right]), but are helpul for questions where the general structure is more relevant.}
    \label{fig:figure_residence_patch}
\end{figure}

\subsection{Validating the Residence Patch Method}

We applied the pre-processing pipeline using \texttt{atlastools} functions described above to a calibration dataset to verify that the residence patch method could correctly identify known stopping points (see Fig. 7).
We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 (Beardsworth et al. \textit{in prep.}; Bijleveld et al. \textit{in prep.}).
Stops in the calibration track were recorded as waypoints using a handheld GPS device (Garmin Dakota 10) at each stop.
We estimated the real duration of each stop as the time difference between the first and last position recorded within 50m of each waypoint, within a 10 minute window before and after the waypoint timestamp (to avoid biased durations from revisits).
Stops had a median duration of 10.28 minutes (range: 1.75 minutes -- 20 minutes; see Supplementary Material).
We cleaned the data before constructing residence patches by (1) removing a single outlier ($>$ 15 km away), removing unrealistic movement ($\geq$ 15 m/s), smoothing the data ($K$ = 5), and (4) thinning the data by resampling over a 30 second interval.
The cleaning steps retained 37,324 positions (74.45\%), while thinning reduced these to 1,803 positions (4.8\% positions of the smoothed track).
Details and code are provided in the Supplementary Material (see \textsc{Validating the Residence Patch Method with Calibration Data}).

% We began by visualising the data to check for location errors, and found a single outlier position approx. 15km away from the study area (Fig. 7.a).
% This outlier was removed by filtering data by the X coordinate bounds using the function \texttt{atl\_filter\_bounds}; X coordinate bounds $\leq$ 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815).
% We then calculated the incoming and outgoing speed, as well as the turning angle at each position using the functions \texttt{atl\_get\_speed} and \texttt{atl\_turning\_angle} respectively, as a precursor to targeting large-scale location errors in the form of point outliers.
% We used the function \texttt{atl\_filter\_covariates} to remove positions with incoming and outgoing speeds $\geq$ the speed threshold of 15 m/s (n = 13,491, 26.5\%; remaining positions = 37,324, 73.5\%; Fig. 7.b).
% This speed threshold was chosen as 5 m/s faster than the known boat speed, 10 m/s.
% Finally, we targeted small-scale location errors by applying a median smooth with a moving window size $K$ = 5 using the function \texttt{atl\_median\_smooth} (Fig. 7.c).
% This step does not reduce the number of positions.

We identified stationary positions (residence time $\geq$ 5 minutes) using the \texttt{recurse} package \citep[n = 837, 46.42 \%; radius = 50m][]{bracis2018}.
We clustered these positions into residence patches with a buffer radius of 5m, spatial independence limit of 50m, temporal independence limit of 5 minutes, and a minimum of 3 positions per patch.
Inferred residence patches corresponded well to the locations of stops (see Fig. 7.c).
However, the residence patch algorithm detected more stops than were logged as waypoints (n = 28, n waypoints = 21).
One of these was the field station on Griend where the tag was stored between trips (red crossed-square, Fig. 7.c).
The method also did not detect two stops of 105 and 563 seconds (1.75 and 9.4 minutes) since they were data poor and aggregated away in the thinning step (n positions = 6, 15).
To determine whether the residence patch method correctly identified the duration of stops in the calibration track, we first extracted the patch attributes using the function \texttt{atl\_patch\_summary}.
We then matched the patches to the waypoints by their median coordinates (rounded to 100 metres).
We assigned the inferred duration of the stop as the duration of the spatially matched residence patch.
We compared the inferred duration with the real duration using a linear model with the inferred duration as the only predictor of the real duration.
% We excluded a single waypoint (WP080) since we had accidentally stopped there for $>$ 60 minutes.
Inferred duration was a good predictor of the real duration of a stop (linear model estimate = 1.021, t-value = 12.965, $p <$ 0.0001, $R^2$ = 0.908; see Supplementary Material Fig. 1.7).
This translates to a 2\% underestimation of the stop duration at a tracking interval of 30 seconds.

% calibration figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_07_calib_residence_patch.png}
    \caption{Pre-processing steps for WATLAS calibration data showing filtering filtering on speed, median smoothing and thinning by aggregation, and making residence patches.
    \textbf{(a)} Positions with incoming and outgoing speed $\geq$ the 15 m/s are removed (N removed = 13,491, 26.6\%; grey crosses = removed, green points = retained).
    Rectangles show the region expanded in subfigures (b) and (c).
    \textbf{(b)} Expanded view of upper rectangle in (a) showing the raw data (grey crosses), the median smoothed positions (green circles; moving window $K$ = 5), and the smoothed track thinned by aggregation to a 30 second interval (purple squares).
    Square size corresponds to the number of positions used to calculate the averaged position during thinning, with no significant differences.
    Median smoothing retains all 37,324 observations (though changing their coordinates). Thinning aggregates these positions into 1,803 aggregates (4.8\% of the smoothed data volume).
    Median smoothing recovers a good estimate of the true track from significantly over-dispersed raw data, while thinning reduces the data volume but preserves track structure.
    \textbf{(c)} Classifying thinned data into residence patches yields robust estimates of the duration of known stops. The island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) is shown in beige.
    Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
    However, the algorithm identified all areas with prolonged residence, including those which were not intended stops, such as stops at the field station (n = 12; green polygon over red crossed-squares).
    The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).}
    \label{fig:figure_calibration}
\end{figure}

\section{Worked-Out Example on Animal Tracking Data}

We present a fully worked-out example of our pre-processing pipeline and residence patch method using movement data from three Egyptian fruit bats tracked using the ATLAS system (\textit{Rousettus aegyptiacus}; \citet{toledo2020}).
Code can be found in the Supplementary Material (see \textsc{Processing Egyptian Fruit Bat Tracks}).
Bats were tracked over three nights (5\textsuperscript{th}, 6\textsuperscript{th}, and 7\textsuperscript{th} May, 2018) in the Hula Valley, Israel (33.1$^{\circ}$N, 35.6$^{\circ}$E), with an average of 13,370 positions (SD = 2,173; range = 11,195 -- 15,542; interval = 8 seconds) per individual.
Plotting the tracks showed severe distortions (see Supplementary Material Fig. 2.1).
We first reduced location errors by removing observations with ATLAS SD $>$ 20, and observations calculated using fewer than four base stations (mean positions remaining = 10,447 / individual; 78\% of the raw data on average).
% We converted the ATLAS time format (milliseconds since the UNIX Epoch) to the more common UNIX format (seconds since the Epoch) by taking the floored value of the time divided by 1000.
We removed unrealistic movement represented by positions with incoming and outgoing speeds $>$ 20 m/s leaving 10,337 positions per individual on average (98\% of previous step).
We median smoothed the data with a moving window $K$ size = 5, and no observations were lost.

We began the construction of residence patches by finding the residence time within 50 metres of each position \citep{bracis2018}.
Bats may repeatedly traverse the same routes, and this could artificially inflate the residence time of positions along these routes.
To avoid confusing revisits with residence, we limited the summation of residence times at each position to the period until the first departure of 60 minutes or more.
Thus, two nearby locations ($\leq$ 50m apart) each visited for one minute at a time, but separated by an interval of some hours would not have a residence time of two minutes each, but only one minute each.
Bats had a mean residence time at locations of 100.54 minutes (SD = 114.7); this measure was strongly biased by time spent at the roost.
We opted for a first-principles approach and selected as residence positions any locations with a residence time $>$ 5 minutes, reasoning that a flying animal stopping for $>$ 5 minutes at a location should plausibly indicate resource use or another interesting behaviour.
This step retained 7,819 positions per bat on average (75.6\%) of the smoothed data; suggesting the extent to which roosting biases the data.

We constructed residence patches with a buffer distance of 25m, a spatial independence limit of 100m, a temporal independence limit of 30 minutes, and rejected patches with fewer than three positions.
We extracted summary data and spatial polygons from the constructed residence patches.
Plotting the bats' residence patches and the linear paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site (Fig. 9.a).
Bats tended to show prolonged residence near known food sources (fruit trees), travelling repeatedly between previously visited areas (Fig. 9.b).
However, bats also appeared to spend some time at locations where no fruit trees were recorded, prompting questions about their use of other food sources, or another behaviour entirely (Fig. 9.b, 9.c).
Bats occurring close together did not have strongly overlapping residence patches, and their paths to and from area of co-occurrence were different (Fig. 9.a, 9.c).
Constructing residence patches for multiple individuals over multiple nights suggests interesting dynamics of within- and between-individual overlap.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_08_bats.png}
    \caption{Synthesising animal tracks into residence patches can reveal how individuals in a population move in relation to landscape features, prior exploration, and other individuals.
    \textbf{(a)} Linear approximations of the paths (coloured lines) between residence patches (circles) show the movement of three Egyptian fruit bats (\textit{Rousettus aegyptiacus}), tracked over three nights in the Hula Valley, Israel.
    Bat tracks are are shown as thin lines below the linear approximations.
    Colours show bat identity. The grey hexagon represents a roost site.
    Dark green points represent known fruit trees.
    Background is shaded by elevation at 30 metre resolution (SRTM).    
    \textbf{(b)} Spatial representations of an individual bat's residence patches (green polygons) can be used to study site-fidelity by examining overlaps between patches, or to study resource selection by inspecting overlaps with known resources (here, black circles show the location of clusters of fruit trees).
    In addition, the linear approximation of movement between patches (blue paths in panel (a)) can be contrasted with the estimated real path between patches (green lines).
    \textbf{(c)} Residence patch polygons can also show how individuals partition space and resources (black circles are fruit trees), and the combined influence of resource and potential social cues on movement between patches (coloured paths).
    Patches and paths are coloured by bat identity.}
    \label{fig:figure_bats}
\end{figure}

\section{Discussion}

Our guide anticipates high-throughput animal tracking data becoming increasingly common.
A uniform pipeline and toolset for data cleaning promotes reproducibility and standardisation across studies, making comparative inferences more robust.
The open-source \texttt{R} package \texttt{atlastools} serves as a starting point for methodological collaboration among movement ecologists.
Efficient location error modelling approaches \citep{fleming2020} may eventually make data-cleaning optional.
Yet cleaning tracking data even partially before modelling location error will be faster than error-modelling on the full data, and the removal of large location errors may improve model fits.
Thus we see our pipeline as complementary to these approaches \citep{fleming2014a,fleming2020}.
Finally, we recognise that the diversity and complexity of animal movement and data collection techniques often requires bespoke pre-processing solutions.
Though the principles outlined here are readily generalised, users' requirements will eventually exceed the particular tools we provide.
We see this as an incentive for more users to be involved in developing methods for their systems.
We offer our pipeline and package as a foundation for system specific tools in the belief that simple, robust concepts are key to methods development that balances system-specificity and broad applicability.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Backmatter}

\subsection{Competing Interests}

The authors declare that they have no competing interests.

\subsection{Acknowledgements}

PRG would like to thank Pedro M. Santos Neves for introducing PRG to \texttt{R} package development, for help with setting up \texttt{atlastools}, and for help with archiving it on Zenodo; 
A. Ramesh and F.J. Weissing for discussions on error propagation;
J.R.L. Gismann and E. Gobbens for feedback that improved the manuscript; 
members of the Modelling Adaptive Response Mechanisms Group (Weissing Lab), and the Theoretical Biology department at the University of Groningen for helpful discussions on \texttt{atlastools} and the manuscript.
All authors thank the attendees of ATLAS workshops held in May and June 2020 at the Hebrew University of Jerusalem for helpful comments on the pipeline and \texttt{atlastools}.

\subsection{Authors' Contributions}

PRG wrote the manuscript and inline code snippets, performed the analyses, prepared the figures, and developed the \texttt{R} package \texttt{atlastools}.
CEB and AIB collected the calibration track, and EL collected the bat movement data and fruit tree locations.
RN conceived the idea of writing this manuscript, and PRG, AIB, OS, CEB, ST, and EL contributed to its design, and the design of \texttt{atlastools}.
All authors contributed to the writing of the manuscript, and the design of figures.

\subsection{Data Availability}

The data and source code to reproduce the figures and analyses in this article and in the Supplementary Material can be found in the Zenodo repository at https://doi.org/10.5281/zenodo.4287462.

\subsection{Supplementary Material}
\begin{enumerate}
    \item Supplementary Material 1: Code for worked out examples on calibration data from the Dutch Wadden Sea, and bat tracking data from the Hula Valley, Israel.
    \item Supplementary Material 2: Manual for the \texttt{R} package \texttt{atlastools}.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\nolinenumbers
\singlespacing

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{amnatnat} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{references.bib}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

%%
%% Do not use \listoffigures as most will included as separate files

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Figures here                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2021-02-01 render_books.sh
minor edit

@@ -10,9 +10,6 @@ Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2"
 Rscript --slave -e 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
 
 # copy pdfs and name correctly
-# remove old pdfs
-
-
 cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text`date -I`.pdf
 mv docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_`date -I`.pdf
 

### 2021-02-01 render_books.sh
edit doc name

@@ -10,7 +10,7 @@ Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2"
 Rscript --slave -e 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
 
 # copy pdfs and name correctly
-cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text`date -I`.pdf
+cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text_`date -I`.pdf
 mv docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_`date -I`.pdf
 
 # copy tex and replace old tex

### 2021-02-01 docs/ms_atlas_preproc_main_text_2021-02-01.tex
final edits before reformat

@@
-\documentclass[10pt,paper=a4]{scrartcl}
-\usepackage{tgtermes}
+\documentclass[10pt,paper=a4,headings=standardclasses
+]{scrartcl}
+\usepackage{mathpazo}
 \usepackage{helvet}
@@
 
+\title{A Guide to Pre-Processing High-Throughput Animal Tracking Data}
+\date{}
+\author{Pratik Rajan Gupte\textsuperscript{1, 2} \and 
+Christine E. Beardsworth\textsuperscript{2} \and
+Orr Spiegel\textsuperscript{3} \and
+Emmanuel Lourie\textsuperscript{4} \and
+Sivan Toledo\textsuperscript{5} \and
+Ran Nathan\textsuperscript{4} \and
+Allert I. Bijleveld\textsuperscript{2}}
+
 % document begins here
 \begin{document}
-\vspace*{0.35in}
+% \vspace*{0.35in}
 
 % title goes here:
+% {\Large
+% {\sffamily \textbf\newline{A Guide to Pre-Processing High-Throughput Animal Tracking Data}}
+% }
+\maketitle
+
+% \newline
+% authors go here:
+
 \begin{flushleft}
-{\Large
-{\sffamily \textbf\newline{A Guide to Pre-Processing High-Throughput Animal Tracking Data}}
-}
-\newline
-% authors go here:
-\\
-Pratik Rajan Gupte\textsuperscript{1, 2},
-Christine E. Beardsworth\textsuperscript{2},
-Orr Spiegel\textsuperscript{3},
-Emmanuel Lourie\textsuperscript{4},
-Sivan Toledo\textsuperscript{5},
-Ran Nathan\textsuperscript{4},
-Allert I. Bijleveld\textsuperscript{2}
 % Author 4\textsuperscript{x},
@@
 % Author 7\textsuperscript{x}
-\\
 \bigskip
@@
 PRG would like to thank Pedro M. Santos Neves for introducing PRG to \texttt{R} package development, for help with setting up \texttt{atlastools}, and for help with archiving it on Zenodo; 
-A. Ramesh and F.J. Weissing for discussions on error propagation;
-J.R.L. Gismann and E. Gobbens for feedback that improved the manuscript; 
+Aparajitha Ramesh and Franjo Weissing for discussions on error propagation;
+Geert Aarts, Jacob RL Gismann, Evy Gobbens for feedback that improved the manuscript; 
 members of the Modelling Adaptive Response Mechanisms Group (Weissing Lab), and the Theoretical Biology department at the University of Groningen for helpful discussions on \texttt{atlastools} and the manuscript.

### 2021-02-03 _02_filtering_data.Rmd
redo figure 02 code

@@
             alpha = 1,
+            shape = 4,
             size = 0.2)+
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank())+
+        axis.title = element_blank(),
+        plot.title = element_text(
+          face = "bold",
+          margin = margin(t = 30, b = -30),
+          hjust = 0.1
+        ))+
   coord_equal()+
-  labs(colour = NULL)
+  labs(colour = NULL)+
+  labs(
+    title = "(a)"
+  )
 ```
@@
              col = "grey",
-             alpha = 0.5, size = 0.2)+
+             shape = 4,
+             size = 0.2)+
   geom_path(data = data,
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank())+
+        axis.title = element_blank(),
+        plot.title = element_text(
+          face = "bold",
+          margin = margin(t = 30, b = -30),
+          hjust = 0.1
+        ))+
   theme(plot.background = element_rect(fill = NA))+
-  coord_equal(expand = T)
+  coord_equal(expand = T)+
+  labs(
+    title = "(b)"
+  )
 
@@
 plot_figure <-
-  wrap_plots(list(figure_raw, fig_filter_bounds)) +
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")") &
-  theme(plot.tag = element_text(face = "bold"))
+  wrap_plots(list(figure_raw, fig_filter_bounds))
 
 ggsave(plot_figure, filename = "figures/fig_02_bounds.png", 
-       width = 170, height = 170, units = "mm")
+       width = 170, height = 150, units = "mm")
 ```

### 2021-02-03 _02_filtering_data.Rmd
redo figure 3

@@
              aes(x, y),
-             size = 0.5,
-             shape = 2, col = pal[4])+
+             # size = 0.5,
+             shape = 4, col = pal[4])+
   geom_point(data = data_copy[!data_copy[500:800, ],
@@
                  col = (in_speed >= 0.03 & out_speed >= 0.03),
-                 shape = (in_speed >= 0.03 & out_speed >= 0.03)),
+                 shape = (in_speed >= 0.03 & out_speed >= 0.03),
+                 size = (in_speed >= 0.03 & out_speed >= 0.03)),
              show.legend = F,
-             alpha = 1,
-             size = 0.5)+
+             alpha = 1)+
   geom_path(data = data,
@@
             alpha = 1)+
+  scale_color_manual(values = c(pal[3], pal[1]))+
+  scale_shape_manual(values = c(1, 4))+
+  scale_size_manual(
+    values = c(0.5, 2)
+  )+
   ggthemes::theme_few()+
@@
         axis.title = element_blank())+
-  scale_color_manual(values = c(pal[3], "black"))+
-  scale_shape_manual(values = c(16, 21))+
   coord_equal()+
@@
 ```{r echo=FALSE}
-# get plots
-fig_reflection <-
-  ggplot()+
-    geom_path(data = reflection,
-             aes(x, y),
-             alpha = 1,
-             col = "grey",
-             lwd = 0.2)+
-  geom_point(data = reflection,
-             aes(x, y),
-             alpha = 0.5,
-             col = "grey",
-             size = 0.5,
-             shape = 16)+
-  geom_point(data = data_no_reflection,
-             aes(x, y),
-             # alpha = 0.5,
-             size = 0.5,
-             colour = pal[3],
-             shape = 16,
-             show.legend = F)+
-  geom_path(data = data,
-            aes(x, y),
-            col = "grey20",
-            alpha = 1)+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  coord_equal()+
-  theme(plot.background = element_rect(fill = NA))
-
-# wrap figures
-plot_figure <- wrap_plots(list(fig_outlier_remove, 
-                               fig_reflection)) +
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")") &
-  theme(plot.tag = element_text(face = "bold"))
-
-ggsave(plot_figure, filename = "figures/fig_03_filter_speed.png", 
-       width = 170, height = 170, units = "mm")
+# save figure
+ggsave(fig_outlier_remove, 
+       filename = "figures/fig_03_filter_speed.png", 
+       width = 80, height = 150, units = "mm")
 ```

### 2021-02-03 _03_smoothing_tracks.Rmd
fix thinning plot
minor edits

@@
              col = pal[3],
-             size = 0.2, alpha = 0.5)+
+             size = 0.2)+
   geom_path(data = data,
@@
            fontface = "bold")+
-  scale_colour_distiller(palette = "BuPu", direction = 1,
+  scale_colour_distiller(palette = "Blues", direction = 1,
                          values = c(-0.5, 1))+
@@
 # choose the 11 point median smooth data
-data_agg <- copy(list_of_smooths[[3]])
+data_agg <- fread("data/data_smooth.csv")
 

### 2021-02-03 _03_smoothing_tracks.Rmd
prepare edit to fig 5

@@
     ggplot(df)+
-      geom_path(data = data,
+      # geom_path(data = data,
+      #           aes(x, y),
+      #           col = "grey20",
+      #           size = 0.2)+
+      geom_path(data = data_agg,
                  aes(x, y),
-                col = "grey20",
-                size = 0.2)+
-      geom_point(data = data_agg,
-                 aes(x, y),
-                 size = 0.2,
-                 col = pal[3])+
+                 size = 0.3,
+                 col = pal[2])+
       geom_path(aes(x,y), 
-                col = pal[1])+
+                col = pal[4])+
       geom_point(aes(x,y,
                      group = interval),
-                 shape = 19,
+                 shape = 0,
                  col = pal[4],
-                 size = 2,
-                 alpha = 0.6,
+                 alpha = 1,
                  show.legend = F)+
@@
              size = 3,
-             fill = pal[3],
+             fill = pal[2],
              position = position_nudge(x = 0.25)) +
@@
 # make combined figure
-fig_aggregate <-
+# fig_aggregate <-
   wrap_plots(append(fig_agg_data[3:4], list(fig_agg_speed)),

### 2021-02-03 _02_filtering_data.Rmd
no more separate raw data

@@
 # make figure of canonical data with added errors
-figure_raw <-
-  ggplot()+
-  geom_point(data = data_copy,
-            aes(x, y),
-            col = "grey",
-            alpha = 1,
-            shape = 4,
-            size = 0.2)+
-  geom_path(data = data,
-            aes(x, y),
-            col = "grey20",
-            alpha = 1)+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        plot.title = element_text(
-          face = "bold",
-          margin = margin(t = 30, b = -30),
-          hjust = 0.1
-        ))+
-  coord_equal()+
-  labs(colour = NULL)+
-  labs(
-    title = "(a)"
-  )
+# figure_raw <-
+#   ggplot()+
+#   geom_point(data = data_copy,
+#             aes(x, y),
+#             col = "grey",
+#             alpha = 1,
+#             shape = 4,
+#             size = 0.2)+
+#   geom_path(data = data,
+#             aes(x, y),
+#             col = "grey20",
+#             alpha = 1)+
+#   ggthemes::theme_few()+
+#   theme(axis.text = element_blank(),
+#         axis.title = element_blank(),
+#         plot.title = element_text(
+#           face = "bold",
+#           margin = margin(t = 30, b = -30),
+#           hjust = 0.1
+#         ))+
+#   coord_equal()+
+#   labs(colour = NULL)+
+#   labs(
+#     title = "(a)"
+#   )
 ```
@@
   labs(
-    title = "(b)"
+    title = "(a)"
   )
-
-# wrap plots
-plot_figure <-
-  wrap_plots(list(figure_raw, fig_filter_bounds))
-
-ggsave(plot_figure, filename = "figures/fig_02_bounds.png", 
-       width = 170, height = 150, units = "mm")
 ```
@@
              aes(x, y),
-             # size = 0.5,
+             size = 0.5,
+             alpha = 0.6,
              shape = 4, col = pal[4])+
@@
   scale_size_manual(
-    values = c(0.5, 2)
+    values = c(0.2, 2)
   )+
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank())+
+        axis.title = element_blank(),
+        plot.title = element_text(
+          face = "bold",
+          margin = margin(t = 30, b = -30),
+          hjust = 0.1
+        ))+
   coord_equal()+
-  theme(plot.background = element_rect(fill = NA))
+  theme(plot.background = element_rect(fill = NA))+
+  labs(
+    title = "(b)"
+  )
 ```
@@
 ```{r echo=FALSE}
+# wrap plot
+figure_01 =
+  wrap_plots(
+    fig_filter_bounds,
+    fig_outlier_remove
+  )
+
 # save figure
-ggsave(fig_outlier_remove, 
-       filename = "figures/fig_03_filter_speed.png", 
-       width = 80, height = 150, units = "mm")
+ggsave(figure_01, 
+       filename = "figures/fig_02_filtering_data.png", 
+       width = 170, height = 150, units = "mm")
 ```

### 2021-02-03 _03_smoothing_tracks.Rmd
rename fig 04 to 03
redo aggregation code

@@
 # save figure
-ggsave(figure_median_smooth, filename = "figures/fig_04_median_smooth.png",
+ggsave(figure_median_smooth, 
+       filename = "figures/fig_03_median_smooth.png",
        width = 170, height = 170 / 3, units = "mm")
@@
 fig_agg_data <-
-  lapply(list_of_agg, function(df) {
+  Map(function(df, title) {
     ggplot(df)+
@@
       #           size = 0.2)+
-      geom_path(data = data_agg,
+      geom_point(data = data_agg,
                  aes(x, y),
                  size = 0.3,
-                 col = pal[2])+
+                 col = pal[3])+
       geom_path(aes(x,y), 
@@
             axis.title = element_blank(),
-            panel.background = element_rect(fill = "white"))+
-      coord_cartesian(ylim = c(0.6, NA))
-  })
+            plot.title = element_text(
+              face = "bold",
+              margin = margin(t = 30, b = -30),
+              hjust = 0.8
+            ))+
+      coord_cartesian(ylim = c(0.6, NA))+
+      labs(
+        title = sprintf("(%s)", title)
+      )
+  },
+  list_of_agg, 
+  c(NA, NA, "a", "b")
+  )
 ```
@@
              size = 3,
-             fill = pal[2],
+             fill = pal[3],
              position = position_nudge(x = 0.25)) +
@@
   ggthemes::theme_few()+
-  theme(axis.text.y = element_blank())+
+  theme(
+    axis.text.y = element_blank(),
+    axis.text = element_blank(),
+        axis.title = element_blank(),
+        plot.title = element_text(
+          face = "bold",
+          margin = margin(t = 30, b = -30),
+          hjust = 0.8
+        ))+
   labs(x = "interval (s)",
-       y = "speed")
+       y = "speed",
+       title = "(c)")
 ```
@@
 # make combined figure
-# fig_aggregate <-
+fig_aggregate <-
   wrap_plots(append(fig_agg_data[3:4], list(fig_agg_speed)),
-             design = "AABBC")+
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")") &
-  theme(plot.tag = element_text(face = "bold"))
+             design = "AABBC")
 
@@
 ggsave(fig_aggregate,
-       filename = "figures/fig_05_thinning.png",
+       filename = "figures/fig_04_thinning.png",
        width = 170, height = 85, units = "mm")

### 2021-02-03 render_books.sh
copy figures to overleaf fodler

@@ -4,6 +4,8 @@
 rm docs/*.pdf
 rm docs/*.tex
 
+cp figures/fig_0*.png atlas-manuscript/figures
+
 Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2")'
 
 # to render source files

### 2021-02-03 06_bat_data.Rmd
change title
style file

@@
 
-# Processing Egyptian fruit bat tracks
+# Processing Egyptian Fruit Bat Tracks
 
@@
 # prepare the connection
-con <- dbConnect(drv = SQLite(), 
-                 dbname = "data/Three_example_bats.sql")
+con <- dbConnect(
+  drv = SQLite(),
+  dbname = "data/Three_example_bats.sql"
+)
 
@@
 fig_bat_raw <-
-  ggplot(data)+
+  ggplot(data) +
   geom_path(aes(X, Y,
     group = TAG,
-                colour = TAG),
-            show.legend = FALSE)+
-  scale_colour_brewer(palette = "Dark2")+
-  ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
+    colour = TAG
+  ),
+  show.legend = FALSE
+  ) +
+  scale_colour_brewer(palette = "Dark2") +
+  ggspatial::annotation_scale(location = "br") +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  ) +
   coord_sf(crs = 2039)
@@
   dpi = 300,
-       width = 6, height = 4)
+  width = 6, height = 4
+)
 ```
@@
 data_split <- lapply(data_split, function(dt) {
-
   dt <- atl_filter_covariates(
     data = dt,
-    filters = c("SD <= 20",
-                "NBS > 3")
-
+    filters = c(
+      "SD <= 20",
+      "NBS > 3"
+    )
   )
@@
 fig_bat_filter_cov <-
-  ggplot()+
-  geom_point(data = data[!data_split, 
-                         on = c("TAG", "X", "Y", "TIME")],
+  ggplot() +
+  geom_point(
+    data = data[!data_split,
+      on = c("TAG", "X", "Y", "TIME")
+    ],
     aes(X, Y),
     col = "grey",
-             shape = 4)+
-  geom_path(data = data_split,
+    shape = 4
+  ) +
+  geom_path(
+    data = data_split,
     aes(X, Y,
-                 col = TAG),
+      col = TAG
+    ),
     size = 0.3,
-             show.legend = FALSE)+
-  scale_colour_brewer(palette = "Dark2")+
-  ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  coord_sf(crs = 2039,
+    show.legend = FALSE
+  ) +
+  scale_colour_brewer(palette = "Dark2") +
+  ggspatial::annotation_scale(location = "br") +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  ) +
+  coord_sf(
+    crs = 2039,
     xlim = range(data_split$X),
-           ylim = range(data_split$Y))
+    ylim = range(data_split$Y)
+  )
 ```
@@
   dpi = 300,
-       width = 4, height = 6)
+  width = 4, height = 6
+)
 ```
@@
 
-    dt[, `:=`(speed_in = atl_get_speed(dt, 
+    dt[, `:=`(
+      speed_in = atl_get_speed(dt,
         x = "X", y = "Y",
         time = "time",
-                                       type = "in"),
+        type = "in"
+      ),
       speed_out = atl_get_speed(dt,
@@
         time = "time",
-                                       type = "out"))]
+        type = "out"
+      )
+    )]
   })
@@
 
-  dt <- atl_filter_covariates(data = dt,
-                              filters = c("speed_in <= 20",
-                                          "speed_out <= 20"))
+  dt <- atl_filter_covariates(
+    data = dt,
+    filters = c(
+      "speed_in <= 20",
+      "speed_out <= 20"
+    )
+  )
 })
@@
 fig_bat_filter_speed <-
-  ggplot()+
-  geom_point(data = data[!data_split, 
-                         on = c("TAG", "X", "Y", "TIME")],
+  ggplot() +
+  geom_point(
+    data = data[!data_split,
+      on = c("TAG", "X", "Y", "TIME")
+    ],
     aes(X, Y),
     col = "grey",
-             shape = 4)+
-  geom_path(data = data_split,
+    shape = 4
+  ) +
+  geom_path(
+    data = data_split,
     aes(X, Y,
-                 col = TAG),
+      col = TAG
+    ),
     size = 0.3,
-             show.legend = FALSE)+
-  scale_colour_brewer(palette = "Dark2")+
-  ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  coord_sf(crs = 2039,
+    show.legend = FALSE
+  ) +
+  scale_colour_brewer(palette = "Dark2") +
+  ggspatial::annotation_scale(location = "br") +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  ) +
+  coord_sf(
+    crs = 2039,
     xlim = range(data_split$X),
-           ylim = range(data_split$Y))
+    ylim = range(data_split$Y)
+  )
 ```
@@
 
-  lapply(X = data_smooth, 
+  lapply(
+    X = data_smooth,
     FUN = atl_median_smooth,
-         time = "time", moving_window = 5)
+    time = "time", moving_window = 5
+  )
 )
@@
 fig_bat_smooth <-
-  ggplot()+
-  geom_point(data = data_split,
+  ggplot() +
+  geom_point(
+    data = data_split,
     aes(X, Y),
@@
     shape = 4,
-             show.legend = FALSE)+
-  geom_path(data = data_smooth,
+    show.legend = FALSE
+  ) +
+  geom_path(
+    data = data_smooth,
     aes(X, Y,
-                col = TAG),
+      col = TAG
+    ),
     lwd = 0.3,
-            show.legend = FALSE)+
-  scale_colour_brewer(palette = "Dark2")+
-  ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  coord_sf(crs = 2039,
+    show.legend = FALSE
+  ) +
+  scale_colour_brewer(palette = "Dark2") +
+  ggspatial::annotation_scale(location = "br") +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  ) +
+  coord_sf(
+    crs = 2039,
     xlim = range(data_smooth$X),
-           ylim = range(data_smooth$Y))
+    ylim = range(data_smooth$Y)
+  )
 ```
@@
     dt_recurse[, c("coordIdx", "resTime")],
-              by = c("coordIdx"))
+    by = c("coordIdx")
+  )
 
@@
 fig_bat_res_timeseries <-
-  ggplot()+
-  geom_path(data = data_residence,
+  ggplot() +
+  geom_path(
+    data = data_residence,
     aes(ts, resTime,
-                colour = TAG),
-            show.legend = F)+
-  scale_color_brewer(palette = "Dark2")+
-  scale_x_datetime()+
-  ggthemes::theme_few()+
-  labs(x = NULL,
-       y = "residence time (mins)")
+      colour = TAG
+    ),
+    show.legend = F
+  ) +
+  scale_color_brewer(palette = "Dark2") +
+  scale_x_datetime() +
+  ggthemes::theme_few() +
+  labs(
+    x = NULL,
+    y = "residence time (mins)"
+  )
 ```
@@
   dpi = 300,
-       width = 6, height = 3)
+  width = 6, height = 3
+)
 ```
@@
 # add an id column
-data_residence[, `:=`(id = TAG,
-                      x = X, y = Y)]
+data_residence[, `:=`(
+  id = TAG,
+  x = X, y = Y
+)]
 
@@
 data_patches <- lapply(data_residence, atl_res_patch,
-                       buffer_radius = 25)
+  buffer_radius = 25
+)
 ```
@@
   which_data = "spatial",
-                        buffer_radius = 25)
+  buffer_radius = 25
+)
 
@@
 st_write(data_spatials,
-         dsn = "data/data_bat_residence_patches.gpkg")
+  dsn = "data/data_bat_residence_patches.gpkg"
+)
 ```
@@
 data_clean <- fwrite(rbindlist(data_smooth),
-                     file = "data/data_bat_smooth.csv")
+  file = "data/data_bat_smooth.csv"
+)
 ```
@@
 # write
-fwrite(patch_summary,
-       "data/data_bat_patch_summary.csv")
+fwrite(
+  patch_summary,
+  "data/data_bat_patch_summary.csv"
+)
 ```

### 2021-02-03 render_books.sh
add styler code

@@ -6,6 +6,9 @@ rm docs/*.tex
 
 cp figures/fig_0*.png atlas-manuscript/figures
 
+# style files
+Rscript --vanilla --slave -e 'styler::style_dir(".", filetype = "Rmd", recursive = FALSE)'
+
 Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2")'
 
 # to render source files

### 2021-02-03 R/05_calibration_data.R
add r scripts

@@ -508,6 +508,6 @@ figure_res_patch <-
 
 # save figure
 ggsave(figure_res_patch,
-       filename = "figures/fig_07_calib_residence_patch.png",
+       filename = "figures/fig_06_calib_residence_patch.png",
        height = 170, width = 170, units = "mm")
 

### 2021-02-03 R/_02_filtering_data.R

@@ -49,22 +49,31 @@ pal <- RColorBrewer::brewer.pal(4, "Set1")
 
 ## ----echo=FALSE---------------------------------------------------------------
 # make figure of canonical data with added errors
-figure_raw <-
-  ggplot()+
-  geom_point(data = data_copy,
-            aes(x, y),
-            col = "grey",
-            alpha = 1,
-            size = 0.2)+
-  geom_path(data = data,
-            aes(x, y),
-            col = "grey20",
-            alpha = 1)+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  coord_equal()+
-  labs(colour = NULL)
+# figure_raw <-
+#   ggplot()+
+#   geom_point(data = data_copy,
+#             aes(x, y),
+#             col = "grey",
+#             alpha = 1,
+#             shape = 4,
+#             size = 0.2)+
+#   geom_path(data = data,
+#             aes(x, y),
+#             col = "grey20",
+#             alpha = 1)+
+#   ggthemes::theme_few()+
+#   theme(axis.text = element_blank(),
+#         axis.title = element_blank(),
+#         plot.title = element_text(
+#           face = "bold",
+#           margin = margin(t = 30, b = -30),
+#           hjust = 0.1
+#         ))+
+#   coord_equal()+
+#   labs(colour = NULL)+
+#   labs(
+#     title = "(a)"
+#   )
 
 
 ## ----remove_outside_bbox------------------------------------------------------
@@ -87,7 +96,8 @@ fig_filter_bounds <-
                               on = c("x", "y")],
              aes(x, y),
              col = "grey",
-             alpha = 0.5, size = 0.2)+
+             shape = 4,
+             size = 0.2)+
   geom_path(data = data,
             aes(x, y),
             col = "grey20",
@@ -97,20 +107,17 @@ fig_filter_bounds <-
              lty = 2)+
   ggthemes::theme_few()+
   theme(axis.text = element_blank(),
-        axis.title = element_blank())+
+        axis.title = element_blank(),
+        plot.title = element_text(
+          face = "bold",
+          margin = margin(t = 30, b = -30),
+          hjust = 0.1
+        ))+
   theme(plot.background = element_rect(fill = NA))+
-  coord_equal(expand = T)
-
-# wrap plots
-plot_figure <-
-  wrap_plots(list(figure_raw, fig_filter_bounds)) +
-  plot_annotation(tag_levels = "a",
[…]

### 2021-02-03 R/_03_smoothing_tracks.R

@@ -57,7 +57,7 @@ figure_median_smooth <-
   geom_point(data = data_errors,
              aes(x, y),
              col = pal[3],
-             size = 0.2, alpha = 0.5)+
+             size = 0.2)+
   geom_path(data = data,
             aes(x, y),
             col = "grey20",
@@ -77,14 +77,15 @@ figure_median_smooth <-
            y = 0.82,
            label = sprintf("(%s)", letters[seq(5)]),
            fontface = "bold")+
-  scale_colour_distiller(palette = "BuPu", direction = 1,
+  scale_colour_distiller(palette = "Blues", direction = 1,
                          values = c(-0.5, 1))+
   ggthemes::theme_few()+
   theme(axis.text = element_blank(),
         axis.title = element_blank())
 
 # save figure
-ggsave(figure_median_smooth, filename = "figures/fig_04_median_smooth.png",
+ggsave(figure_median_smooth, 
+       filename = "figures/fig_03_median_smooth.png",
        width = 170, height = 170 / 3, units = "mm")
 
 
@@ -121,31 +122,40 @@ speed_agg_smooth <- rbindlist(speed_agg_smooth)
 ## ----echo=FALSE---------------------------------------------------------------
 ### plot figures
 fig_agg_data <-
-  lapply(list_of_agg, function(df) {
+  Map(function(df, title) {
     ggplot(df)+
-      geom_path(data = data,
-                aes(x, y),
-                col = "grey20",
-                size = 0.2)+
+      # geom_path(data = data,
+      #           aes(x, y),
+      #           col = "grey20",
+      #           size = 0.2)+
       geom_point(data = data_agg,
                  aes(x, y),
-                 size = 0.2,
+                 size = 0.3,
                  col = pal[3])+
       geom_path(aes(x,y), 
-                col = pal[1])+
+                col = pal[4])+
       geom_point(aes(x,y,
                      group = interval),
-                 shape = 19,
+                 shape = 0,
                  col = pal[4],
-                 size = 2,
-                 alpha = 0.6,
+                 alpha = 1,
                  show.legend = F)+
       ggthemes::theme_few()+
       theme(axis.text = element_blank(),
             axis.title = element_blank(),
-            panel.background = element_rect(fill = "white"))+
-      coord_cartesian(ylim = c(0.6, NA))
-  })
+            plot.title = element_text(
+              face = "bold",
[…]

### 2021-02-03 R/_04_residence_patch.R

@@ -31,25 +31,25 @@ data[, residence_time := data_recurse$residenceTime]
 
 ## ----echo=FALSE---------------------------------------------------------------
 # restime by position
-fig_res_a <-
-  ggplot()+
-  geom_path(data = data,
-            aes(x, y),
-            lwd = 0.2,
-            col = "grey")+
-  geom_point(data = data,
-             aes(x, y,
-                 group = NA
-                 ),
-             size = 0.3,
-             show.legend = F,
-             alpha = 0.2,
-             col = "grey20")+
-  coord_equal()+
-  scale_colour_manual(values = c("grey", pal[3]))+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())
+# fig_res_a <-
+#   ggplot()+
+#   geom_path(data = data,
+#             aes(x, y),
+#             lwd = 0.2,
+#             col = "grey")+
+#   geom_point(data = data,
+#              aes(x, y,
+#                  group = NA
+#                  ),
+#              size = 0.3,
+#              show.legend = F,
+#              alpha = 0.2,
+#              col = "grey20")+
+#   coord_equal()+
+#   scale_colour_manual(values = c("grey", pal[3]))+
+#   ggthemes::theme_few()+
+#   theme(axis.text = element_blank(),
+#         axis.title = element_blank())
 
 
 ## -----------------------------------------------------------------------------
@@ -73,7 +73,7 @@ patch_summary <- atl_patch_summary(patch_data = patch,
 
 ## ----echo=FALSE---------------------------------------------------------------
 # plot_patches <-
-fig_res_b <-
+fig_res_patches <-
   ggplot()+
   geom_sf(data = patch_sf,
           aes(fill = patch),
@@ -106,13 +106,24 @@ fig_res_b <-
              shape = 21,
              show.legend = F)+
   scale_fill_distiller(palette = "Paired")+
-  coord_sf()+
+  coord_sf(
+    xlim = c(-3, NA)
+  )+
   ggthemes::theme_few()+
   theme(axis.text = element_blank(),
-        axis.title = element_blank())
+        axis.title = element_blank(),
+        plot.title = element_text(
+              face = "bold",
+              margin = margin(t = 30, b = -30),
+              hjust = 0.925
+            )
+  )+
+  labs(
+    title = "(a)"
+  )
 
[…]

### 2021-02-03 06_bat_data.Rmd
correct image file

@@
 When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar **(c)**.
-Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_08_bats.png)
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_07_bats.png)

### 2021-02-03 render_books.sh
correct remove pdf and tex

@@ -1,8 +1,11 @@
 #!/bin/bash
 
 # remove old tex files and pdfs
-rm docs/*.pdf
-rm docs/*.tex
+cd docs
+rm *.pdf
+rm *.tex
+
+cd ..
 
 cp figures/fig_0*.png atlas-manuscript/figures
 

### 2021-02-03 R/05_calibration_data.R
style files

@@ -18,19 +18,25 @@ data_raw <- copy(data)
 ## ----echo=FALSE---------------------------------------------------------------
 # plot data
 fig_data_raw <-
-  ggplot(data)+
+  ggplot(data) +
   geom_path(aes(x, y),
-             col = "grey", alpha = 1, size = 0.2)+
+    col = "grey", alpha = 1, size = 0.2
+  ) +
   geom_point(aes(x, y),
-             col = "grey", alpha = 0.2, size = 0.2)+
-  ggthemes::theme_few()+
-  theme(axis.title = element_blank(),
-        axis.text = element_blank())+
+    col = "grey", alpha = 0.2, size = 0.2
+  ) +
+  ggthemes::theme_few() +
+  theme(
+    axis.title = element_blank(),
+    axis.text = element_blank()
+  ) +
   coord_sf(crs = 32631)
 
 # save figure
-ggsave(fig_data_raw, filename = "figures/fig_calibration_raw.png",
-       width = 185 / 25)
+ggsave(fig_data_raw,
+  filename = "figures/fig_calibration_raw.png",
+  width = 185 / 25
+)
 
 
 ## -----------------------------------------------------------------------------
@@ -40,37 +46,50 @@ data_unproc <- copy(data)
 
 ## -----------------------------------------------------------------------------
 # remove inside must be set to falses
-data <- atl_filter_bounds(data = data, 
-                          x = "x", y = "y", 
-                          x_range = c(645000, max(data$x)), 
-                          remove_inside = FALSE)
+data <- atl_filter_bounds(
+  data = data,
+  x = "x", y = "y",
+  x_range = c(645000, max(data$x)),
+  remove_inside = FALSE
+)
 
 
 ## ----echo=FALSE---------------------------------------------------------------
 # plot data
 fig_data_bbox <-
-  ggplot()+
-  geom_path(data = data_raw,
-            aes(x, y),
-             col = "grey", alpha = 0.5, size = 0.3)+
-  geom_point(data = data_raw,
-             aes(x, y,
-                 col = x > 645000),
-             alpha = ifelse(data_raw$x > 645000, 1, 1),
-             size = ifelse(data_raw$x > 645000, 0.3, 2),
-             show.legend = F)+
-  scale_colour_manual(values = c("black", pal[3]))+
-  geom_vline(xintercept = 645000,
-             col = "grey", lty = 2)+
[…]

### 2021-02-03 R/_01_background.R

@@ -28,19 +28,21 @@ set.seed(0)
 
 # do smoove data using a RACVM for three different rotational speed
 data_thermals <- Map(function(rot_speed, patch) {
- dt <- smoove::simulateRACVM(dt = 0.1, Tmax = 50, omega = rot_speed, v0 = 0.1) 
- dt <- as.data.table(dt$XY)
- dt[, patch := patch]
- dt[, time := seq(.N)]
-}, c(7, 7, 7), sprintf('patch_%i', seq(3)))
+  dt <- smoove::simulateRACVM(dt = 0.1, Tmax = 50, omega = rot_speed, v0 = 0.1)
+  dt <- as.data.table(dt$XY)
+  dt[, patch := patch]
+  dt[, time := seq(.N)]
+}, c(7, 7, 7), sprintf("patch_%i", seq(3)))
 
 # rescale x and y
-data_thermals <- Map(function(df, lim1, lim2, offset) {
-  df[, x := x + lim1]
-  df[, y := y + lim2]
-  df[, time := time + offset]
-}, data_thermals, c(1, 5, 9), c(0, 2.5, 5),
-   c(100, 800, 1500))
+data_thermals <- Map(
+  function(df, lim1, lim2, offset) {
+    df[, x := x + lim1]
+    df[, y := y + lim2]
+    df[, time := time + offset]
+  }, data_thermals, c(1, 5, 9), c(0, 2.5, 5),
+  c(100, 800, 1500)
+)
 
 # get thermals data
 data_thermals <- rbindlist(data_thermals)
@@ -72,15 +74,17 @@ transit <- data.table(
 )
 
 # nafill
-transit[, c("x1", "y1", "t1") := 
-          lapply(list(x1, y1, t1), nafill, fill = 0)]
+transit[, c("x1", "y1", "t1") :=
+  lapply(list(x1, y1, t1), nafill, fill = 0)]
 
 # interpolate
 transit[, d := Map(function(x1, x2, y1, y2) {
-  as.data.table(approx(x = c(x1, x2), 
-                       y = c(y1, y2), 
-                       n = 20))
-  }, x1, x2, y1, y2)]
+  as.data.table(approx(
+    x = c(x1, x2),
+    y = c(y1, y2),
+    n = 20
+  ))
+}, x1, x2, y1, y2)]
 
 # add errors
 transit[, d := Map(function(df) {
@@ -99,8 +103,9 @@ transit_data <- rbindlist(transit_data)
 
 ## -----------------------------------------------------------------------------
[…]

### 2021-02-03 R/_02_filtering_data.R

@@ -23,12 +23,16 @@ data <- fread("data/data_sim.csv")[5000:10000, ]
 ## data_copy <- copy(data)
 ## 
 ## # add a prolonged spike or reflection to 300 positions
-## data_copy[500:800, `:=`(x = x + 0.25,
-##                         y = y + 0.25)]
+## data_copy[500:800, `:=`(
+##   x = x + 0.25,
+##   y = y + 0.25
+## )]
 ## 
 ## # add normal error
-## data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
-##                  y = do_add_error(y, std_dev = 0.005))]
+## data_copy[, `:=`(
+##   x = do_add_error(x, std_dev = 0.01),
+##   y = do_add_error(y, std_dev = 0.005)
+## )]
 ## 
 ## # add 100 outliers
 ## data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
@@ -79,42 +83,55 @@ pal <- RColorBrewer::brewer.pal(4, "Set1")
 ## ----remove_outside_bbox------------------------------------------------------
 # remove positions outside a bounding box
 # NB: set remove_inside to FALSE
-data_inside_bbox <- atl_filter_bounds(data = data_copy,
-                                      y_range = c(0.5, 1),
-                                      remove_inside = FALSE)
+data_inside_bbox <- atl_filter_bounds(
+  data = data_copy,
+  y_range = c(0.5, 1),
+  remove_inside = FALSE
+)
 
 
 ## ----echo=FALSE---------------------------------------------------------------
 # plot data inside and outside bbox
 fig_filter_bounds <-
-  ggplot()+
-  geom_point(data = data_inside_bbox,
-             aes(x, y),
-             col = pal[3],
-             alpha = 1, size = 0.2)+
-  geom_point(data = data_copy[!data_inside_bbox,
-                              on = c("x", "y")],
-             aes(x, y),
-             col = "grey",
-             shape = 4,
-             size = 0.2)+
-  geom_path(data = data,
-            aes(x, y),
-            col = "grey20",
-            alpha = 1)+
-  geom_hline(yintercept = c(0.5, 1),
-             col = "grey",
-             lty = 2)+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        plot.title = element_text(
-          face = "bold",
-          margin = margin(t = 30, b = -30),
[…]

### 2021-02-03 R/_03_smoothing_tracks.R

@@ -22,15 +22,16 @@ data_errors[, window_size := 0]
 ## -----------------------------------------------------------------------------
 # smooth the data over four K values
 list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
-  
   data_copy <- copy(data_errors)
-  
-  data_copy <- atl_median_smooth(data = data_copy,
-                    x = "x", 
-                    y = "y",
-                    time = "time",
-                    moving_window = K)
-  
+
+  data_copy <- atl_median_smooth(
+    data = data_copy,
+    x = "x",
+    y = "y",
+    time = "time",
+    moving_window = K
+  )
+
   data_copy[, window_size := K]
 })
 
@@ -44,7 +45,8 @@ fwrite(list_of_smooths[[3]], file = "data/data_smooth.csv")
 data_plot <- mapply(function(df, offset) {
   df[, x := x + offset]
 }, list_of_smooths, seq(0.4, 1.25, length.out = 4),
-SIMPLIFY = F)
+SIMPLIFY = F
+)
 
 data_plot <- rbindlist(data_plot)
 
@@ -53,40 +55,56 @@ data_plot <- rbindlist(data_plot)
 # prepare data to plot
 # make list of data to plot
 figure_median_smooth <-
-  ggplot()+
-  geom_point(data = data_errors,
-             aes(x, y),
-             col = pal[3],
-             size = 0.2)+
-  geom_path(data = data,
-            aes(x, y),
-            col = "grey20",
-            lwd = 0.5)+          
-  geom_path(data = data_plot,
-            aes(x, y,
-                col = window_size,
-                group = window_size),
-            show.legend = F,
-            lwd = 0.5)+
-  coord_equal(expand = F,
-              ylim = c(0.6, 0.85),
-              xlim = c(NA, 2.3),
-              ratio = 1.75)+
-  annotate(geom = "text",
-           x = c(0.75, seq(1.1, 2, length.out = 4)),
-           y = 0.82,
-           label = sprintf("(%s)", letters[seq(5)]),
-           fontface = "bold")+
-  scale_colour_distiller(palette = "Blues", direction = 1,
-                         values = c(-0.5, 1))+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())
+  ggplot() +
+  geom_point(
+    data = data_errors,
+    aes(x, y),
+    col = pal[3],
+    size = 0.2
+  ) +
+  geom_path(
+    data = data,
[…]

### 2021-02-03 R/_04_residence_patch.R

@@ -23,7 +23,8 @@ setnames(data, old = c("x", "y"), new = c("y", "x"))
 
 # do recurse
 data_recurse <- getRecursions(data[, list(x, y, time, id)],
-                              radius = 1)
+  radius = 1
+)
 
 # assign residence time
 data[, residence_time := data_recurse$residenceTime]
@@ -54,111 +55,140 @@ data[, residence_time := data_recurse$residenceTime]
 
 ## -----------------------------------------------------------------------------
 # make residence patch
-patch <- atl_res_patch(data, 
-                       buffer_radius = 0.1, 
-                       lim_spat_indep = 1, 
-                       lim_time_indep = 5)
+patch <- atl_res_patch(data,
+  buffer_radius = 0.1,
+  lim_spat_indep = 1,
+  lim_time_indep = 5
+)
 
 
 
 # get spatial representation
-patch_sf <- atl_patch_summary(patch_data = patch,
-                              which_data = "spatial", 
-                              buffer_radius = 0.15)
+patch_sf <- atl_patch_summary(
+  patch_data = patch,
+  which_data = "spatial",
+  buffer_radius = 0.15
+)
 
 # get summary data
-patch_summary <- atl_patch_summary(patch_data = patch,
-                                   which_data = "summary")
+patch_summary <- atl_patch_summary(
+  patch_data = patch,
+  which_data = "summary"
+)
 
 
 ## ----echo=FALSE---------------------------------------------------------------
 # plot_patches <-
 fig_res_patches <-
-  ggplot()+
-  geom_sf(data = patch_sf,
-          aes(fill = patch),
-          colour = "grey20",
-          alpha = 0.5,
-          lwd = 0.2,
-          show.legend = FALSE)+
-  geom_path(data = data,
-            aes(x, y),
-            lwd = 0.2,
-            col = "grey20")+
-  geom_point(data = data,
-            aes(x, y),
-            size = 0.4,
-            col = "grey20",
-            alpha = 0.2)+
-  scale_colour_manual(values = c("grey", pal[3]))+
-  
-  geom_path(data = patch_summary,
-            aes(x_median + 2, y_median),
-            col = "grey0",
-            size = 0.5)+
-  geom_point(data = patch_summary,
-             aes(x_median + 2,
-                 y_median,
[…]

### 2021-02-03 docs/ms_atlas_preproc_main_text_2021-02-03.tex
add updated tex

--- 2021-02-01 docs/ms_atlas_preproc_main_text_2021-02-01.tex
@@
     \centering
-    \includegraphics[width=0.95\textwidth]{figures/fig_02_bounds.png}
+    \includegraphics[width=0.95\textwidth]{figures/fig_02_filtering_data.png}
     \caption{\textbf{(a)} A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the \texttt{smoove} package (Gurarie et al.
@@
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.95\textwidth]{figures/fig_03_filter_speed.png}
-    \caption{Reducing large-scale outliers in a movement track.
+% \begin{figure}[h!]
+%     \centering
+%     \includegraphics[width=0.95\textwidth]{figures/fig_03_filter_speed.png}
+%     \caption{Reducing large-scale outliers in a movement track.
 The goal is to remove positions far away from the canonical track (black line), and retain those with only reasonable displacements from the canonical track (green points).
-    \textbf{(a)} location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles).
-    The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
-    Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
-    Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
-    \textbf{(b)} The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
-    While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult.
-    Users are cautioned to frequently check this and similar methods' results.}
-    \label{fig:figure_filter_bounds}
-\end{figure}
+%     \textbf{(a)} location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles).
+%     The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
+%     Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
+%     Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
+%     \textbf{(b)} The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
+%     While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult.
+%     Users are cautioned to frequently check this and similar methods' results.}
+%     \label{fig:figure_filter_bounds}
+% \end{figure}
 
@@
     \centering
-    \includegraphics[width=0.95\textwidth]{figures/fig_04_median_smooth.png}
+    \includegraphics[width=0.95\textwidth]{figures/fig_03_median_smooth.png}
     \caption{Median smoothing position coordinates reduces small-scale location error in tracking data. 
@@
     \centering
-    \includegraphics[width=0.95\textwidth]{figures/fig_05_thinning.png}
+    \includegraphics[width=0.95\textwidth]{figures/fig_04_thinning.png}
     \caption{Thinning a movement track using aggregation preserves track structure while reducing data volume, but thinning without removing gross location errors and unrealistic movement affects track metrics such as speed.
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{figures/fig_06_residence.png}
+    \includegraphics[width=0.75\textwidth]{figures/fig_05_residence.png}
     \caption{A rotation-advection movement track, similar to the movement of a soaring bird (Gurarie et al. 2017) classified into residence patches with the \texttt{atlastools} function.
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{figures/fig_07_calib_residence_patch.png}
+    \includegraphics[width=0.75\textwidth]{figures/fig_06_calib_residence_patch.png}
     \caption{Pre-processing steps for WATLAS calibration data showing filtering filtering on speed, median smoothing and thinning by aggregation, and making residence patches.
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{figures/fig_08_bats.png}
+    \includegraphics[width=0.75\textwidth]{figures/fig_07_bats.png}
     \caption{Synthesising animal tracks into residence patches can reveal how individuals in a population move in relation to landscape features, prior exploration, and other individuals.

### 2021-02-03 docs/ms_atlas_preproc_main_text_2021-02-03.tex
add caption tex update

@@
     The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (black line) as closely as possible.  
-    \textbf{(b)} Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the \texttt{atlastools} function  \texttt{atl\_filter\_bounds}.
-    Setting the function argument \texttt{remove\_inside = FALSE} retains positions within user supplied bounds (dashed grey lines; green points), and excludes those outside (grey points).}
+    Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the \texttt{atlastools} function  \texttt{atl\_filter\_bounds}.
+    Setting the function argument \texttt{remove\_inside = FALSE} retains positions within user supplied bounds (dashed grey lines; green points), and excludes those outside (grey points).
+    Reducing large-scale outliers in a movement track.
+The goal is to remove positions far away from the canonical track (black line), and retain those with only reasonable displacements from the canonical track (green points).
+    \textbf{(b)} Location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles).
+    The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
+    Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
+    Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
+    The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
+    While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult.
+    Users are cautioned to frequently check this and similar methods' results.
+    }
     \label{fig:figure_simulated_data}

### 2021-02-03 copy_tex.sh
add a tex copier

@@ -1,0 +1,3 @@
+#!/bin/bash
+# copy tex and replace old tex
+cp -p atlas-manuscript/arxiv_article.tex docs/ms_atlas_preproc_main_text_`date -I`.tex

### 2021-02-03 05_calibration_data.Rmd
rename figure 06

@@
 fig_data_raw <-
-  ggplot(data)+
+  ggplot(data) +
   geom_path(aes(x, y),
-             col = "grey", alpha = 1, size = 0.2)+
+    col = "grey", alpha = 1, size = 0.2
+  ) +
   geom_point(aes(x, y),
-             col = "grey", alpha = 0.2, size = 0.2)+
-  ggthemes::theme_few()+
-  theme(axis.title = element_blank(),
-        axis.text = element_blank())+
+    col = "grey", alpha = 0.2, size = 0.2
+  ) +
+  ggthemes::theme_few() +
+  theme(
+    axis.title = element_blank(),
+    axis.text = element_blank()
+  ) +
   coord_sf(crs = 32631)
@@
 # save figure
-ggsave(fig_data_raw, filename = "figures/fig_calibration_raw.png",
-       width = 185 / 25)
+ggsave(fig_data_raw,
+  filename = "figures/fig_calibration_raw.png",
+  width = 185 / 25
+)
 ```
@@
 # remove inside must be set to falses
-data <- atl_filter_bounds(data = data, 
+data <- atl_filter_bounds(
+  data = data,
   x = "x", y = "y",
   x_range = c(645000, max(data$x)),
-                          remove_inside = FALSE)
+  remove_inside = FALSE
+)
 ```
@@
 fig_data_bbox <-
-  ggplot()+
-  geom_path(data = data_raw,
+  ggplot() +
+  geom_path(
+    data = data_raw,
     aes(x, y),
-             col = "grey", alpha = 0.5, size = 0.3)+
-  geom_point(data = data_raw,
+    col = "grey", alpha = 0.5, size = 0.3
+  ) +
+  geom_point(
+    data = data_raw,
     aes(x, y,
-                 col = x > 645000),
+      col = x > 645000
+    ),
     alpha = ifelse(data_raw$x > 645000, 1, 1),
     size = ifelse(data_raw$x > 645000, 0.3, 2),
-             show.legend = F)+
-  scale_colour_manual(values = c("black", pal[3]))+
-  geom_vline(xintercept = 645000,
-             col = "grey", lty = 2)+
-  ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
+    show.legend = F
+  ) +
+  scale_colour_manual(values = c("black", pal[3])) +
+  geom_vline(
+    xintercept = 645000,
+    col = "grey", lty = 2
+  ) +
+  ggspatial::annotation_scale(location = "br") +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  ) +
   coord_sf(crs = 32631)
@@
 # save result
-ggsave(fig_data_bbox, filename = "figures/fig_calib_bbox.png",
-       width = 185 / 25)
+ggsave(fig_data_bbox,
+  filename = "figures/fig_calib_bbox.png",
+  width = 185 / 25
+)
 ```
@@
 # add incoming and outgoing speed
-data[, `:=` (speed_in = atl_get_speed(data, 
+data[, `:=`(
+  speed_in = atl_get_speed(data,
     x = "x",
     y = "y",
-                                      time = "time"),
-             speed_out = atl_get_speed(data, type = "out"))]
+    time = "time"
+  ),
+  speed_out = atl_get_speed(data, type = "out")
+)]
 
@@
   sapply(data[, list(speed_in, speed_out, angle)],
-       quantile, probs = 0.9, na.rm = T)
+    quantile,
+    probs = 0.9, na.rm = T
+  )
 ```
@@
 fig_speed_outliers <-
-  ggplot()+
-  geom_point(data = data,
+  ggplot() +
+  geom_point(
+    data = data,
     aes(x, y,
-                 col = speed_in < 25.6),
+      col = speed_in < 25.6
+    ),
     size = 0.3, alpha = 0.5,
-             show.legend = F)+
-  annotate(geom = "rect",
+    show.legend = F
+  ) +
+  annotate(
+    geom = "rect",
     xmin = 650785, xmax = 653250,
     ymin = 5904450, ymax = 5906133,
-           fill = NA, col = "grey20")+
-  annotate(geom = "rect",
+    fill = NA, col = "grey20"
+  ) +
+  annotate(
+    geom = "rect",
     ymin = 5901650, ymax = 5903100,
@@
     fill = NA, col = "grey20"
-           )+
-  scale_color_manual(values = c("grey", pal[3]))+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  ggspatial::annotation_scale(location = "tl")+
-  coord_sf(crs = 32631)+
+  ) +
+  scale_color_manual(values = c("grey", pal[3])) +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  ) +
+  ggspatial::annotation_scale(location = "tl") +
+  coord_sf(crs = 32631) +
   labs(colour = "speed (m/s)")
@@
 # save
-ggsave(fig_speed_outliers, filename = "figures/fig_speed_outlier.png",
-       width = 170 / 25, height = 170 / 25)
+ggsave(fig_speed_outliers,
+  filename = "figures/fig_speed_outlier.png",
+  width = 170 / 25, height = 170 / 25
+)
 ```
@@
 # remove speed outliers
-data <- atl_filter_covariates(data = data,
-            filters = c("(speed_in < 15 & speed_out < 15)"))
+data <- atl_filter_covariates(
+  data = data,
+  filters = c("(speed_in < 15 & speed_out < 15)")
+)
 
 # recalculate speed and angle
-data[, `:=` (speed_in = atl_get_speed(data, 
+data[, `:=`(
+  speed_in = atl_get_speed(data,
     x = "x",
     y = "y",
-                                      time = "time"),
-             speed_out = atl_get_speed(data, type = "out"))]
+    time = "time"
+  ),
+  speed_out = atl_get_speed(data, type = "out")
+)]
 
@@
 # now apply the smooth
-atl_median_smooth(data = data,
+atl_median_smooth(
+  data = data,
   x = "x", y = "y", time = "time",
-                  moving_window = 5)
+  moving_window = 5
+)
 ```
@@
 fig_smooth <-
-  ggplot()+
-  geom_point(data = data_raw[!data_unproc, on = c("x", "y")],
+  ggplot() +
+  geom_point(
+    data = data_raw[!data_unproc, on = c("x", "y")],
     aes(x, y),
@@
     shape = 4,
-             size = 0.4)+
-  geom_point(data = data,
+    size = 0.4
+  ) +
+  geom_point(
+    data = data,
     aes(x, y),
@@
     shape = 19,
-             alpha = 0.5) +
+    alpha = 0.5
+  ) +
   coord_cartesian(
@@
     expand = F
-  )+
-  ggthemes::theme_few()+
-  theme(axis.title = element_blank(),
-        axis.text = element_blank())+
+  ) +
+  ggthemes::theme_few() +
+  theme(
+    axis.title = element_blank(),
+    axis.text = element_blank()
+  ) +
   ggspatial::annotation_scale(location = "br")
@@
   filename = "figures/fig_calib_median_smooth.png",
-       width = 90 / 25, height = 90 / 25)
+  width = 90 / 25, height = 90 / 25
+)
 ```
@@
 # remove columns we don't need
-data <- data[, setdiff(colnames(data), 
-                       c("tID", "Timestamp", "id", "TIME", "UTCtime")), 
-             with = FALSE]
+data <- data[, setdiff(
+  colnames(data),
+  c("tID", "Timestamp", "id", "TIME", "UTCtime")
+),
+with = FALSE
+]
 
 # thin to a 30s interval
-data_thin <- atl_thin_data(data = data,
+data_thin <- atl_thin_data(
+  data = data,
   interval = 30,
   method = "aggregate",
-                           id_columns = "TAG")
+  id_columns = "TAG"
+)
 ```
@@
 fig_smooth_thin <-
-  ggplot()+
-  geom_point(data = data_raw[!data_unproc, on = c("x", "y")],
+  ggplot() +
+  geom_point(
+    data = data_raw[!data_unproc, on = c("x", "y")],
     aes(x, y),
@@
     shape = 4,
-             size = 0.4)+
-
-  geom_point(data = data_thin,
+    size = 0.4
+  ) +
+  geom_point(
+    data = data_thin,
     aes(x, y, size = count),
@@
     alpha = 0.8,
-             shape = 0, show.legend = F)+
-  geom_point(data = data_unproc,
+    shape = 0, show.legend = F
+  ) +
+  geom_point(
+    data = data_unproc,
     aes(x, y),
@@
     col = pal[3],
-             shape = 16) +
-  scale_size(range = c(2, 6))+
+    shape = 16
+  ) +
+  scale_size(range = c(2, 6)) +
   coord_cartesian(
@@
   filename = "figures/fig_calib_smooth_thin.png",
-       width = 90 / 25, height = 90 / 25, dpi = 300)
+  width = 90 / 25, height = 90 / 25, dpi = 300
+)
 ```
@@
     list(
-    fig_speed_outliers, fig_smooth_thin),
-  design = "AB")+
-  plot_annotation(tag_levels = "a",
+      fig_speed_outliers, fig_smooth_thin
+    ),
+    design = "AB"
+  ) +
+    plot_annotation(
+      tag_levels = "a",
       tag_prefix = "(",
-                  tag_suffix = ")")&
+      tag_suffix = ")"
+    ) &
     theme(plot.tag = element_text(face = "bold"))
@@
   filename = "figures/fig_walkthrough.png",
-       height = 90, width = 170, units = "mm")
+  height = 90, width = 170, units = "mm"
+)
 ```
@@
 # get centroid
-data_res_summary <- data_res[, list(x_median = median(x),
+data_res_summary <- data_res[, list(
+  x_median = median(x),
   y_median = median(y),
-                                    t_median = median(time)),
-                             by = "tID"]
+  t_median = median(time)
+),
+by = "tID"
+]
 
 # now get times 10 mins before and after
-data_res_summary[, `:=`(t_min = t_median - (10 * 60),
-                        t_max = t_median + (10 * 60))]
+data_res_summary[, `:=`(
+  t_min = t_median - (10 * 60),
+  t_max = t_median + (10 * 60)
+)]
 
@@
   return(diff(range(tmp_data$time)))
-
 }, data_res_summary$t_min, data_res_summary$t_max,
 data_res_summary$x_median, data_res_summary$y_median,
-SIMPLIFY = TRUE)
+SIMPLIFY = TRUE
+)
 ```
@@
 recurse_stats <- getRecursions(data_for_patch,
-                               radius = 50, timeunits = "mins")
+  radius = 50, timeunits = "mins"
+)
 
@@
   lim_time_indep = 5,
-                                min_fixes = 3)
+  min_fixes = 3
+)
 ```
@@
   which_data = "spatial",
-                                   buffer_radius = 20)
+  buffer_radius = 20
+)
 
@@
 patch_summary_data <- atl_patch_summary(patch_res_known,
-                                        which_data = "summary")
+  which_data = "summary"
+)
 ```
@@
 fig_basic_residence <-
-  ggplot()+
-  geom_sf(data = griend, fill = "antiquewhite")+
-  geom_point(data = data_res_summary, 
+  ggplot() +
+  geom_sf(data = griend, fill = "antiquewhite") +
+  geom_point(
+    data = data_res_summary,
     aes(x_median, y_median),
@@
     col = pal[4],
-             alpha = 1)+
-  geom_sf(data = hut,
+    alpha = 1
+  ) +
+  geom_sf(
+    data = hut,
     size = 10, shape = 7,
     stroke = 1,
-          col = pal[1])+
-  geom_path(data = data_for_patch,
+    col = pal[1]
+  ) +
+  geom_path(
+    data = data_for_patch,
     aes(x, y),
@@
     size = 0.3,
-            col = "grey")+
-  geom_point(data = data_for_patch,
+    col = "grey"
+  ) +
+  geom_point(
+    data = data_for_patch,
     aes(x, y),
     show.legend = F,
-            shape = 1, col = "grey50")+
-  geom_sf(data = patch_sf_data[patch_sf_data$duration >= 5 * 60, ],
+    shape = 1, col = "grey50"
+  ) +
+  geom_sf(
+    data = patch_sf_data[patch_sf_data$duration >= 5 * 60, ],
     fill = pal[3],
     col = pal[3],
-          alpha = 0.3)+
-  scale_colour_manual(values = c("grey50", pal[3]))+
-  scale_shape_manual(values = c(4, 19))+
-  ggspatial::annotation_scale(location = "tl")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
+    alpha = 0.3
+  ) +
+  scale_colour_manual(values = c("grey50", pal[3])) +
+  scale_shape_manual(values = c(4, 19)) +
+  ggspatial::annotation_scale(location = "tl") +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
     axis.title = element_blank(),
-        panel.background = element_rect(fill = "aliceblue"))+
-  coord_sf(crs = 32631,
+    panel.background = element_rect(fill = "aliceblue")
+  ) +
+  coord_sf(
+    crs = 32631,
     expand = F,
     ylim = c(5901650, 5903100),
-           xlim = c(650000, 650600))
+    xlim = c(650000, 650600)
+  )
 ```
@@
   filename = "figures/fig_calib_residence.png",
-       height = 170, width = 80, units = "mm")
+  height = 170, width = 80, units = "mm"
+)
 ```
@@
 # get waypoint summary
-patch_summary_real <- data_res[, list(nfixes_real = .N,
+patch_summary_real <- data_res[, list(
+  nfixes_real = .N,
   x_median = round(median(x), digits = -2),
-                                      y_median = round(median(y), digits = -2)), 
-                               by = "tID"]
+  y_median = round(median(y), digits = -2)
+),
+by = "tID"
+]
 
@@
 patch_summary_inferred <-
-  patch_summary_data[, 
-                     c("x_median", "y_median", 
-                       "nfixes", "duration", "patch")
-                     ][, `:=`(x_median = round(x_median, digits = -2),
-                              y_median = round(y_median, digits = -2))]
+  patch_summary_data[
+    ,
+    c(
+      "x_median", "y_median",
+      "nfixes", "duration", "patch"
+    )
+  ][, `:=`(
+    x_median = round(x_median, digits = -2),
+    y_median = round(y_median, digits = -2)
+  )]
 
@@
     on = c("x_median", "y_median"),
-        all.x = TRUE, all.y = TRUE)
+    all.x = TRUE, all.y = TRUE
+  )
 
@@
 model_duration <- lm(duration_real ~ duration,
-                     data = patch_summary_compare)
+  data = patch_summary_compare
+)
 
@@
 figure_lm_duration <-
-  ggplot()+
-  geom_point(data = patch_summary_compare,
-             aes(duration, 
-                 duration_real),
+  ggplot() +
+  geom_point(
+    data = patch_summary_compare,
+    aes(
+      duration,
+      duration_real
+    ),
     show.legend = F,
-            size = 2, col = "grey")+
-  geom_smooth(data = patch_summary_compare,
-             aes(duration, 
-                 duration_real),
+    size = 2, col = "grey"
+  ) +
+  geom_smooth(
+    data = patch_summary_compare,
+    aes(
+      duration,
+      duration_real
+    ),
     se = F,
@@
     method = "lm",
-             show.legend = F)+
-  annotate(geom = "text",
+    show.legend = F
+  ) +
+  annotate(
+    geom = "text",
     x = 600, y = 1020,
@@
     label = "R^2 == 0.908",
-           parse = T, family = "Arial")+
+    parse = T, family = "Arial"
+  ) +
   scale_x_continuous(
@@
     labels = seq(3, 18, 2)
-    )+
+  ) +
   scale_y_continuous(
@@
     labels = seq(5, 20, 2)
-  )+
-  coord_fixed(expand = T,
+  ) +
+  coord_fixed(
+    expand = T,
     ratio = 1.115
@@
     #             ylim = c(250, 1050)
-  )+
-  geom_abline(slope = 1, lty = 2,
-              col = "black")+
-  ggthemes::theme_few()+
-  labs(x = "inferred duration (min)",
-       y = "real duration (min)")
+  ) +
+  geom_abline(
+    slope = 1, lty = 2,
+    col = "black"
+  ) +
+  ggthemes::theme_few() +
+  labs(
+    x = "inferred duration (min)",
+    y = "real duration (min)"
+  )
 
-ggsave(figure_lm_duration, filename = "figures/fig_calib_lm_duration.png",
-       width = 80 / 25, height = 80 / 25)
+ggsave(figure_lm_duration,
+  filename = "figures/fig_calib_lm_duration.png",
+  width = 80 / 25, height = 80 / 25
+)
 ```
@@
     encoding = "UTF-8"
-  ), sep = "\n"
+  ),
+  sep = "\n"
 )
@@
 figure_res_patch <-
-  wrap_plots(list(fig_speed_outliers,
+  wrap_plots(list(
+    fig_speed_outliers,
     fig_smooth_thin,
-                  fig_basic_residence),
+    fig_basic_residence
+  ),
   design = "AC\nAC\nBC\nBC"
-             )+
-  plot_annotation(tag_levels = "a",
+  ) +
+    plot_annotation(
+      tag_levels = "a",
       tag_prefix = "(",
-                  tag_suffix = ")") &
+      tag_suffix = ")"
+    ) &
     theme(plot.tag = element_text(face = "bold"))
@@
 ggsave(figure_res_patch,
-       filename = "figures/fig_07_calib_residence_patch.png",
-       height = 170, width = 170, units = "mm")
+  filename = "figures/fig_06_calib_residence_patch.png",
+  height = 170, width = 170, units = "mm"
+)
 ```

### 2021-02-03 _01_background.Rmd

@@
   dt[, time := seq(.N)]
-}, c(7, 7, 7), sprintf('patch_%i', seq(3)))
+}, c(7, 7, 7), sprintf("patch_%i", seq(3)))
 
 # rescale x and y
-data_thermals <- Map(function(df, lim1, lim2, offset) {
+data_thermals <- Map(
+  function(df, lim1, lim2, offset) {
     df[, x := x + lim1]
@@
   }, data_thermals, c(1, 5, 9), c(0, 2.5, 5),
-   c(100, 800, 1500))
+  c(100, 800, 1500)
+)
 
@@
 transit[, d := Map(function(x1, x2, y1, y2) {
-  as.data.table(approx(x = c(x1, x2), 
+  as.data.table(approx(
+    x = c(x1, x2),
     y = c(y1, y2),
-                       n = 20))
+    n = 20
+  ))
 }, x1, x2, y1, y2)]
@@
 data_move <- merge(data_thermals, transit_data,
-                   on = "time", all = T)
+  on = "time", all = T
+)
 setorder(data_move, time)

### 2021-02-03 _02_filtering_data.Rmd

@@
 # add a prolonged spike or reflection to 300 positions
-data_copy[500:800, `:=`(x = x + 0.25,
-                        y = y + 0.25)]
+data_copy[500:800, `:=`(
+  x = x + 0.25,
+  y = y + 0.25
+)]
 
 # add normal error
-data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
-                 y = do_add_error(y, std_dev = 0.005))]
+data_copy[, `:=`(
+  x = do_add_error(x, std_dev = 0.01),
+  y = do_add_error(y, std_dev = 0.005)
+)]
 
@@
 # NB: set remove_inside to FALSE
-data_inside_bbox <- atl_filter_bounds(data = data_copy,
+data_inside_bbox <- atl_filter_bounds(
+  data = data_copy,
   y_range = c(0.5, 1),
-                                      remove_inside = FALSE)
+  remove_inside = FALSE
+)
 ```
@@
 fig_filter_bounds <-
-  ggplot()+
-  geom_point(data = data_inside_bbox,
+  ggplot() +
+  geom_point(
+    data = data_inside_bbox,
     aes(x, y),
     col = pal[3],
-             alpha = 1, size = 0.2)+
-  geom_point(data = data_copy[!data_inside_bbox,
-                              on = c("x", "y")],
+    alpha = 1, size = 0.2
+  ) +
+  geom_point(
+    data = data_copy[!data_inside_bbox,
+      on = c("x", "y")
+    ],
     aes(x, y),
@@
     shape = 4,
-             size = 0.2)+
-  geom_path(data = data,
+    size = 0.2
+  ) +
+  geom_path(
+    data = data,
     aes(x, y),
     col = "grey20",
-            alpha = 1)+
-  geom_hline(yintercept = c(0.5, 1),
+    alpha = 1
+  ) +
+  geom_hline(
+    yintercept = c(0.5, 1),
     col = "grey",
-             lty = 2)+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
+    lty = 2
+  ) +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
     axis.title = element_blank(),
@@
       hjust = 0.1
-        ))+
-  theme(plot.background = element_rect(fill = NA))+
-  coord_equal(expand = T)+
+    )
+  ) +
+  theme(plot.background = element_rect(fill = NA)) +
+  coord_equal(expand = T) +
   labs(
@@
 # get speed and turning angle
-data_copy[, `:=`(in_speed = atl_get_speed(data_copy,
-                                          type = "in"),
+data_copy[, `:=`(
+  in_speed = atl_get_speed(data_copy,
+    type = "in"
+  ),
   out_speed = atl_get_speed(data_copy,
-                                           type = "out"),
-                 angle = atl_turning_angle(data_copy))]
+    type = "out"
+  ),
+  angle = atl_turning_angle(data_copy)
+)]
 ```
@@
 data_filtered <- atl_filter_covariates(data_copy,
-    filters = c("(in_speed < 0.025 & out_speed < 0.025) | angle < 35"))
+  filters = c("(in_speed < 0.025 & out_speed < 0.025) | angle < 35")
+)
 ```
@@
 fig_outlier_remove <-
-  ggplot()+
-  geom_path(data = data_copy,
+  ggplot() +
+  geom_path(
+    data = data_copy,
     aes(x, y),
     col = "grey",
-            lwd = 0.2)+
-  geom_point(data = data_copy[500:800, ],
+    lwd = 0.2
+  ) +
+  geom_point(
+    data = data_copy[500:800, ],
     aes(x, y),
@@
     alpha = 0.6,
-             shape = 4, col = pal[4])+
-  geom_point(data = data_copy[!data_copy[500:800, ],
-                              on = c("x", "y")],
+    shape = 4, col = pal[4]
+  ) +
+  geom_point(
+    data = data_copy[!data_copy[500:800, ],
+      on = c("x", "y")
+    ],
     aes(x, y,
@@
       shape = (in_speed >= 0.03 & out_speed >= 0.03),
-                 size = (in_speed >= 0.03 & out_speed >= 0.03)),
+      size = (in_speed >= 0.03 & out_speed >= 0.03)
+    ),
     show.legend = F,
-             alpha = 1)+
-  geom_path(data = data,
+    alpha = 1
+  ) +
+  geom_path(
+    data = data,
     aes(x, y),
     col = "grey20",
-            alpha = 1)+
-  scale_color_manual(values = c(pal[3], pal[1]))+
-  scale_shape_manual(values = c(1, 4))+
+    alpha = 1
+  ) +
+  scale_color_manual(values = c(pal[3], pal[1])) +
+  scale_shape_manual(values = c(1, 4)) +
   scale_size_manual(
     values = c(0.2, 2)
-  )+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
+  ) +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
     axis.title = element_blank(),
@@
       hjust = 0.1
-        ))+
-  coord_equal()+
-  theme(plot.background = element_rect(fill = NA))+
+    )
+  ) +
+  coord_equal() +
+  theme(plot.background = element_rect(fill = NA)) +
   labs(
@@
   point_angle_cutoff = 10,
-                          reflection_speed_cutoff = 0.025)
+  reflection_speed_cutoff = 0.025
+)
 # get reflections
 reflection <- data_filtered[!data_no_reflection,
-                                         on = c("x", "y")]
+  on = c("x", "y")
+]
 reflection <- na.omit(reflection)
@@
 # wrap plot
-figure_01 =
+figure_01 <-
   wrap_plots(
@@
   filename = "figures/fig_02_filtering_data.png",
-       width = 170, height = 150, units = "mm")
+  width = 170, height = 150, units = "mm"
+)
 ```

### 2021-02-03 _03_smoothing_tracks.Rmd
minor edit

--- 2021-02-03 _03_smoothing_tracks.Rmd
@@
 list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
-
   data_copy <- copy(data_errors)
 
-  data_copy <- atl_median_smooth(data = data_copy,
+  data_copy <- atl_median_smooth(
+    data = data_copy,
     x = "x",
@@
     time = "time",
-                    moving_window = K)
+    moving_window = K
+  )
 
@@
 }, list_of_smooths, seq(0.4, 1.25, length.out = 4),
-SIMPLIFY = F)
+SIMPLIFY = F
+)
 
@@
 figure_median_smooth <-
-  ggplot()+
-  geom_point(data = data_errors,
+  ggplot() +
+  geom_point(
+    data = data_errors,
     aes(x, y),
     col = pal[3],
-             size = 0.2)+
-  geom_path(data = data,
+    size = 0.2
+  ) +
+  geom_path(
+    data = data,
     aes(x, y),
     col = "grey20",
-            lwd = 0.5)+          
-  geom_path(data = data_plot,
+    lwd = 0.5
+  ) +
+  geom_path(
+    data = data_plot,
     aes(x, y,
       col = window_size,
-                group = window_size),
+      group = window_size
+    ),
     show.legend = F,
-            lwd = 0.5)+
-  coord_equal(expand = F,
+    lwd = 0.5
+  ) +
+  coord_equal(
+    expand = F,
     ylim = c(0.6, 0.85),
     xlim = c(NA, 2.3),
-              ratio = 1.75)+
-  annotate(geom = "text",
+    ratio = 1.75
+  ) +
+  annotate(
+    geom = "text",
     x = c(0.75, seq(1.1, 2, length.out = 4)),
@@
     label = sprintf("(%s)", letters[seq(5)]),
-           fontface = "bold")+
-  scale_colour_distiller(palette = "Blues", direction = 1,
-                         values = c(-0.5, 1))+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())
+    fontface = "bold"
+  ) +
+  scale_colour_distiller(
+    palette = "Blues", direction = 1,
+    values = c(-0.5, 1)
+  ) +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  )
 
 # save figure
-ggsave(figure_median_smooth, filename = "figures/fig_04_median_smooth.png",
-       width = 170, height = 170 / 3, units = "mm")
+ggsave(figure_median_smooth,
+  filename = "figures/fig_03_median_smooth.png",
+  width = 170, height = 170 / 3, units = "mm"
+)
 ```
@@
 list_of_agg <- lapply(c(3, 10, 30, 120), function(z) {
-
-  data_return <- atl_thin_data(data = data_agg,
+  data_return <- atl_thin_data(
+    data = data_agg,
     interval = z,
-                            method = "aggregate")
+    method = "aggregate"
+  )
 
@@
     df[, speed := atl_get_speed(df)]
-    df[, list(median = median(speed, na.rm = T),
+    df[, list(
+      median = median(speed, na.rm = T),
       sd = sd(speed, na.rm = T),
-              interval = first(interval))]
+      interval = first(interval)
+    )]
   })
@@
 fig_agg_data <-
-  lapply(list_of_agg, function(df) {
-    ggplot(df)+
+  Map(
+    function(df, title) {
+      ggplot(df) +
         # geom_path(data = data,
@@
         #           size = 0.2)+
-      geom_path(data = data_agg,
+        geom_point(
+          data = data_agg,
           aes(x, y),
           size = 0.3,
-                 col = pal[2])+
-      geom_path(aes(x,y), 
-                col = pal[4])+
-      geom_point(aes(x,y,
-                     group = interval),
+          col = pal[3]
+        ) +
+        geom_path(aes(x, y),
+          col = pal[4]
+        ) +
+        geom_point(aes(x, y,
+          group = interval
+        ),
         shape = 0,
@@
         alpha = 1,
-                 show.legend = F)+
-      ggthemes::theme_few()+
-      theme(axis.text = element_blank(),
+        show.legend = F
+        ) +
+        ggthemes::theme_few() +
+        theme(
+          axis.text = element_blank(),
           axis.title = element_blank(),
-            panel.background = element_rect(fill = "white"))+
-      coord_cartesian(ylim = c(0.6, NA))
-  })
+          plot.title = element_text(
+            face = "bold",
+            margin = margin(t = 30, b = -30),
+            hjust = 0.8
+          )
+        ) +
+        coord_cartesian(ylim = c(0.6, NA)) +
+        labs(
+          title = sprintf("(%s)", title)
+        )
+    },
+    list_of_agg,
+    c(NA, NA, "a", "b")
+  )
 ```
@@
 list_of_agg <- lapply(c(3, 10, 30, 120), function(z) {
-  data_return <- atl_thin_data(data = data_errors,
+  data_return <- atl_thin_data(
+    data = data_errors,
     interval = z,
-                            method = "aggregate")
+    method = "aggregate"
+  )
   data_return[, interval := z]
@@
 fig_agg_speed <-
-  ggplot(data_agg)+
-  geom_hline(yintercept =
-               1 + quantile(data$speed, na.rm = T,
-                        probs = c(0.5, 0.95)),
-             lty = c(1, 2))+
-  geom_errorbar(data = speed_agg_smooth,
-                aes(x = factor(interval),
+  ggplot(data_agg) +
+  geom_hline(
+    yintercept =
+      1 + quantile(data$speed,
+        na.rm = T,
+        probs = c(0.5, 0.95)
+      ),
+    lty = c(1, 2)
+  ) +
+  geom_errorbar(
+    data = speed_agg_smooth,
+    aes(
+      x = factor(interval),
       ymin = 1 + median - sd,
-                    ymax = 1 + median + sd),
+      ymax = 1 + median + sd
+    ),
     width = 0.2,
-                position = position_nudge(x = 0.25)) +
-  geom_point(data = speed_agg_smooth,
-             aes(x = factor(interval),
-                 y = 1 + median),
+    position = position_nudge(x = 0.25)
+  ) +
+  geom_point(
+    data = speed_agg_smooth,
+    aes(
+      x = factor(interval),
+      y = 1 + median
+    ),
     shape = 21,
     size = 3,
-             fill = pal[2],
-             position = position_nudge(x = 0.25)) +
+    fill = pal[3],
+    position = position_nudge(x = 0.25)
+  ) +
   geom_boxplot(aes(factor(interval), 1 + speed),
-               position = position_nudge(x = -0.25,),
+    position = position_nudge(x = -0.25, ),
     fill = "grey",
@@
     width = 0.25,
-               outlier.size = 0.2)+
-  scale_y_log10(label = scales::comma,
-                limits = c(NA, 1.005))+
-  ggthemes::theme_few()+
-  theme(axis.text.y = element_blank())+
-  labs(x = "interval (s)",
-       y = "speed")
+    outlier.size = 0.2
+  ) +
+  scale_y_log10(
+    label = scales::comma,
+    limits = c(NA, 1.005)
+  ) +
+  ggthemes::theme_few() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.text = element_blank(),
+    axis.title = element_blank(),
+    plot.title = element_text(
+      face = "bold",
+      margin = margin(t = 30, b = -30),
+      hjust = 0.8
+    )
+  ) +
+  labs(
+    x = "interval (s)",
+    y = "speed",
+    title = "(c)"
+  )
 ```
@@
 # make combined figure
-# fig_aggregate <-
-  wrap_plots(append(fig_agg_data[3:4], list(fig_agg_speed)),
-             design = "AABBC")+
-  plot_annotation(tag_levels = "a",
-                  tag_prefix = "(",
-                  tag_suffix = ")") &
-  theme(plot.tag = element_text(face = "bold"))
+fig_aggregate <-
+  wrap_plots(
+    append(fig_agg_data[3:4], list(fig_agg_speed)),
+    design = "AABBC"
+  )
 
@@
 ggsave(fig_aggregate,
-       filename = "figures/fig_05_thinning.png",
-       width = 170, height = 85, units = "mm")
+  filename = "figures/fig_04_thinning.png",
+  width = 170, height = 85, units = "mm"
+)
 ```

### 2021-02-03 docs/ms_atlas_preproc_supplement_2021-02-03.tex

--- 2020-12-15 docs/supplementary_material.tex
@@
 \author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Sivan Toledo \and Ran Nathan \and Allert I. Bijleveld}
-\date{2020-12-15}
+\date{2021-02-03}
 
@@
 \CommentTok{# remove inside must be set to falses}
-\NormalTok{data <-}\StringTok{ }\KeywordTok{atl_filter_bounds}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data, }
-                          \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
-                          \DataTypeTok{x_range =} \KeywordTok{c}\NormalTok{(}\DecValTok{645000}\NormalTok{, }\KeywordTok{max}\NormalTok{(data}\OperatorTok{$}\NormalTok{x)), }
-                          \DataTypeTok{remove_inside =} \OtherTok{FALSE}\NormalTok{)}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{atl_filter_bounds}\NormalTok{(}
+  \DataTypeTok{data =}\NormalTok{ data,}
+  \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{,}
+  \DataTypeTok{x_range =} \KeywordTok{c}\NormalTok{(}\DecValTok{645000}\NormalTok{, }\KeywordTok{max}\NormalTok{(data}\OperatorTok{$}\NormalTok{x)),}
+  \DataTypeTok{remove_inside =} \OtherTok{FALSE}
+\NormalTok{)}
 \end{Highlighting}
@@
 \CommentTok{# add incoming and outgoing speed}
-\NormalTok{data[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{ (}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(data, }
-                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
-                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
-                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
-             \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
+\NormalTok{data[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}
+  \DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(data,}
+    \DataTypeTok{x =} \StringTok{"x"}\NormalTok{,}
+    \DataTypeTok{y =} \StringTok{"y"}\NormalTok{,}
+    \DataTypeTok{time =} \StringTok{"time"}
+\NormalTok{  ),}
+  \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{)}
+\NormalTok{)]}
 
@@
 \CommentTok{# use sapply}
-\NormalTok{speed_angle_thresholds <-}\StringTok{ }
-\StringTok{  }\KeywordTok{sapply}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(speed_in, speed_out, angle)], }
-\NormalTok{       quantile, }\DataTypeTok{probs =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{na.rm =}\NormalTok{ T)}
+\NormalTok{speed_angle_thresholds <-}
+\StringTok{  }\KeywordTok{sapply}\NormalTok{(data[, }\KeywordTok{list}\NormalTok{(speed_in, speed_out, angle)],}
+\NormalTok{    quantile,}
+    \DataTypeTok{probs =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{na.rm =}\NormalTok{ T}
+\NormalTok{  )}
 \end{Highlighting}
@@
 \CommentTok{# remove speed outliers}
-\NormalTok{data <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
-            \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(speed_in < 15 & speed_out < 15)"}\NormalTok{))}
+\NormalTok{data <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}
+  \DataTypeTok{data =}\NormalTok{ data,}
+  \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"(speed_in < 15 & speed_out < 15)"}\NormalTok{)}
+\NormalTok{)}
 
 \CommentTok{# recalculate speed and angle}
-\NormalTok{data[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{ (}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(data, }
-                                      \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }
-                                      \DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }
-                                      \DataTypeTok{time =} \StringTok{"time"}\NormalTok{),}
-             \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
+\NormalTok{data[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}
+  \DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(data,}
+    \DataTypeTok{x =} \StringTok{"x"}\NormalTok{,}
+    \DataTypeTok{y =} \StringTok{"y"}\NormalTok{,}
+    \DataTypeTok{time =} \StringTok{"time"}
+\NormalTok{  ),}
+  \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(data, }\DataTypeTok{type =} \StringTok{"out"}\NormalTok{)}
+\NormalTok{)]}
 
@@
 \CommentTok{# now apply the smooth}
-\KeywordTok{atl_median_smooth}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+\KeywordTok{atl_median_smooth}\NormalTok{(}
+  \DataTypeTok{data =}\NormalTok{ data,}
   \DataTypeTok{x =} \StringTok{"x"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"y"}\NormalTok{, }\DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
-                  \DataTypeTok{moving_window =} \DecValTok{5}\NormalTok{)}
+  \DataTypeTok{moving_window =} \DecValTok{5}
+\NormalTok{)}
 \end{Highlighting}
@@
 \CommentTok{# remove columns we don't need}
-\NormalTok{data <-}\StringTok{ }\NormalTok{data[, }\KeywordTok{setdiff}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(data), }
-                       \KeywordTok{c}\NormalTok{(}\StringTok{"tID"}\NormalTok{, }\StringTok{"Timestamp"}\NormalTok{, }\StringTok{"id"}\NormalTok{, }\StringTok{"TIME"}\NormalTok{, }\StringTok{"UTCtime"}\NormalTok{)), }
-\NormalTok{             with =}\StringTok{ }\OtherTok{FALSE}\NormalTok{]}
+\NormalTok{data <-}\StringTok{ }\NormalTok{data[, }\KeywordTok{setdiff}\NormalTok{(}
+  \KeywordTok{colnames}\NormalTok{(data),}
+  \KeywordTok{c}\NormalTok{(}\StringTok{"tID"}\NormalTok{, }\StringTok{"Timestamp"}\NormalTok{, }\StringTok{"id"}\NormalTok{, }\StringTok{"TIME"}\NormalTok{, }\StringTok{"UTCtime"}\NormalTok{)}
+\NormalTok{),}
+\NormalTok{with =}\StringTok{ }\OtherTok{FALSE}
+\NormalTok{]}
 
 \CommentTok{# thin to a 30s interval}
-\NormalTok{data_thin <-}\StringTok{ }\KeywordTok{atl_thin_data}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ data,}
+\NormalTok{data_thin <-}\StringTok{ }\KeywordTok{atl_thin_data}\NormalTok{(}
+  \DataTypeTok{data =}\NormalTok{ data,}
   \DataTypeTok{interval =} \DecValTok{30}\NormalTok{,}
   \DataTypeTok{method =} \StringTok{"aggregate"}\NormalTok{,}
-                           \DataTypeTok{id_columns =} \StringTok{"TAG"}\NormalTok{)}
+  \DataTypeTok{id_columns =} \StringTok{"TAG"}
+\NormalTok{)}
 \end{Highlighting}
@@
 \CommentTok{# get centroid}
-\NormalTok{data_res_summary <-}\StringTok{ }\NormalTok{data_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{x_median =} \KeywordTok{median}\NormalTok{(x),}
+\NormalTok{data_res_summary <-}\StringTok{ }\NormalTok{data_res[, }\KeywordTok{list}\NormalTok{(}
+  \DataTypeTok{x_median =} \KeywordTok{median}\NormalTok{(x),}
   \DataTypeTok{y_median =} \KeywordTok{median}\NormalTok{(y),}
-                                    \DataTypeTok{t_median =} \KeywordTok{median}\NormalTok{(time)),}
-\NormalTok{                             by =}\StringTok{ "tID"}\NormalTok{]}
+  \DataTypeTok{t_median =} \KeywordTok{median}\NormalTok{(time)}
+\NormalTok{),}
+\NormalTok{by =}\StringTok{ "tID"}
+\NormalTok{]}
 
 \CommentTok{# now get times 10 mins before and after}
-\NormalTok{data_res_summary[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{t_min =}\NormalTok{ t_median }\OperatorTok{-}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{),}
-                        \DataTypeTok{t_max =}\NormalTok{ t_median }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{))]}
+\NormalTok{data_res_summary[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}
+  \DataTypeTok{t_min =}\NormalTok{ t_median }\OperatorTok{-}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{),}
+  \DataTypeTok{t_max =}\NormalTok{ t_median }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\DecValTok{60}\NormalTok{)}
+\NormalTok{)]}
 
@@
   \KeywordTok{return}\NormalTok{(}\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(tmp_data}\OperatorTok{$}\NormalTok{time)))}
-
 \NormalTok{\}, data_res_summary}\OperatorTok{$}\NormalTok{t_min, data_res_summary}\OperatorTok{$}\NormalTok{t_max,}
-\NormalTok{   data_res_summary}\OperatorTok{$}\NormalTok{x_median, data_res_summary}\OperatorTok{$}\NormalTok{y_median,}
-\DataTypeTok{SIMPLIFY =} \OtherTok{TRUE}\NormalTok{)}
+\NormalTok{data_res_summary}\OperatorTok{$}\NormalTok{x_median, data_res_summary}\OperatorTok{$}\NormalTok{y_median,}
+\DataTypeTok{SIMPLIFY =} \OtherTok{TRUE}
+\NormalTok{)}
 \end{Highlighting}
@@
 \NormalTok{recurse_stats <-}\StringTok{ }\KeywordTok{getRecursions}\NormalTok{(data_for_patch,}
-                               \DataTypeTok{radius =} \DecValTok{50}\NormalTok{, }\DataTypeTok{timeunits =} \StringTok{"mins"}\NormalTok{)}
+  \DataTypeTok{radius =} \DecValTok{50}\NormalTok{, }\DataTypeTok{timeunits =} \StringTok{"mins"}
+\NormalTok{)}
 
@@
 \CommentTok{# on known residence points}
-\NormalTok{patch_res_known <-}\StringTok{ }\KeywordTok{atl_res_patch}\NormalTok{(data_for_patch[res_time }\OperatorTok{>=}\StringTok{ }\DecValTok{5}\NormalTok{, ], }
+\NormalTok{patch_res_known <-}\StringTok{ }\KeywordTok{atl_res_patch}\NormalTok{(data_for_patch[res_time }\OperatorTok{>=}\StringTok{ }\DecValTok{5}\NormalTok{, ],}
   \DataTypeTok{buffer_radius =} \DecValTok{5}\NormalTok{,}
@@
   \DataTypeTok{lim_time_indep =} \DecValTok{5}\NormalTok{,}
-                                \DataTypeTok{min_fixes =} \DecValTok{3}\NormalTok{)}
+  \DataTypeTok{min_fixes =} \DecValTok{3}
+\NormalTok{)}
 \end{Highlighting}
@@
 \CommentTok{# for the known and unkniwn patches}
-\NormalTok{patch_sf_data <-}\StringTok{ }\KeywordTok{atl_patch_summary}\NormalTok{(patch_res_known, }
+\NormalTok{patch_sf_data <-}\StringTok{ }\KeywordTok{atl_patch_summary}\NormalTok{(patch_res_known,}
   \DataTypeTok{which_data =} \StringTok{"spatial"}\NormalTok{,}
-                                   \DataTypeTok{buffer_radius =} \DecValTok{20}\NormalTok{)}
+  \DataTypeTok{buffer_radius =} \DecValTok{20}
+\NormalTok{)}
 
@@
 \CommentTok{# get summary data}
-\NormalTok{patch_summary_data <-}\StringTok{ }\KeywordTok{atl_patch_summary}\NormalTok{(patch_res_known, }
-                                        \DataTypeTok{which_data =} \StringTok{"summary"}\NormalTok{)}
+\NormalTok{patch_summary_data <-}\StringTok{ }\KeywordTok{atl_patch_summary}\NormalTok{(patch_res_known,}
+  \DataTypeTok{which_data =} \StringTok{"summary"}
+\NormalTok{)}
 \end{Highlighting}
@@
 \CommentTok{# get waypoint summary}
-\NormalTok{patch_summary_real <-}\StringTok{ }\NormalTok{data_res[, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{nfixes_real =}\NormalTok{ .N,}
+\NormalTok{patch_summary_real <-}\StringTok{ }\NormalTok{data_res[, }\KeywordTok{list}\NormalTok{(}
+  \DataTypeTok{nfixes_real =}\NormalTok{ .N,}
   \DataTypeTok{x_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(x), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{),}
-                                      \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{)), }
-\NormalTok{                               by =}\StringTok{ "tID"}\NormalTok{]}
+  \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(}\KeywordTok{median}\NormalTok{(y), }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{)}
+\NormalTok{),}
+\NormalTok{by =}\StringTok{ "tID"}
+\NormalTok{]}
 
@@
 \CommentTok{# round median coordinate for inferred patches}
-\NormalTok{patch_summary_inferred <-}\StringTok{ }
-\StringTok{  }\NormalTok{patch_summary_data[, }
-                     \KeywordTok{c}\NormalTok{(}\StringTok{"x_median"}\NormalTok{, }\StringTok{"y_median"}\NormalTok{, }
-                       \StringTok{"nfixes"}\NormalTok{, }\StringTok{"duration"}\NormalTok{, }\StringTok{"patch"}\NormalTok{)}
-\NormalTok{                     ][, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{x_median =} \KeywordTok{round}\NormalTok{(x_median, }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{),}
-                              \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(y_median, }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{))]}
+\NormalTok{patch_summary_inferred <-}
+\StringTok{  }\NormalTok{patch_summary_data[}
+\NormalTok{    ,}
+    \KeywordTok{c}\NormalTok{(}
+      \StringTok{"x_median"}\NormalTok{, }\StringTok{"y_median"}\NormalTok{,}
+      \StringTok{"nfixes"}\NormalTok{, }\StringTok{"duration"}\NormalTok{, }\StringTok{"patch"}
+\NormalTok{    )}
+\NormalTok{  ][, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}
+    \DataTypeTok{x_median =} \KeywordTok{round}\NormalTok{(x_median, }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{),}
+    \DataTypeTok{y_median =} \KeywordTok{round}\NormalTok{(y_median, }\DataTypeTok{digits =} \DecValTok{-2}\NormalTok{)}
+\NormalTok{  )]}
 
 \CommentTok{# join with respatch summary}
-\NormalTok{patch_summary_compare <-}\StringTok{ }
+\NormalTok{patch_summary_compare <-}
 \StringTok{  }\KeywordTok{merge}\NormalTok{(patch_summary_real,}
-\NormalTok{        patch_summary_inferred, }
+\NormalTok{    patch_summary_inferred,}
     \DataTypeTok{on =} \KeywordTok{c}\NormalTok{(}\StringTok{"x_median"}\NormalTok{, }\StringTok{"y_median"}\NormalTok{),}
-        \DataTypeTok{all.x =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{all.y =} \OtherTok{TRUE}\NormalTok{)}
+    \DataTypeTok{all.x =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{all.y =} \OtherTok{TRUE}
+\NormalTok{  )}
 
@@
 \NormalTok{model_duration <-}\StringTok{ }\KeywordTok{lm}\NormalTok{(duration_real }\OperatorTok{~}\StringTok{ }\NormalTok{duration,}
-                     \DataTypeTok{data =}\NormalTok{ patch_summary_compare)}
+  \DataTypeTok{data =}\NormalTok{ patch_summary_compare}
+\NormalTok{)}
 
@@
     \DataTypeTok{encoding =} \StringTok{"UTF-8"}
-\NormalTok{  ), }\DataTypeTok{sep =} \StringTok{"}\CharTok{\textbackslash{}n}\StringTok{"}
+\NormalTok{  ),}
+  \DataTypeTok{sep =} \StringTok{"}\CharTok{\textbackslash{}n}\StringTok{"}
 \NormalTok{)}
@@
 \CommentTok{# prepare the connection}
-\NormalTok{con <-}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}\DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(), }
-                 \DataTypeTok{dbname =} \StringTok{"data/Three_example_bats.sql"}\NormalTok{)}
+\NormalTok{con <-}\StringTok{ }\KeywordTok{dbConnect}\NormalTok{(}
+  \DataTypeTok{drv =} \KeywordTok{SQLite}\NormalTok{(),}
+  \DataTypeTok{dbname =} \StringTok{"data/Three_example_bats.sql"}
+\NormalTok{)}
 
@@
 \NormalTok{data_split <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_split, }\ControlFlowTok{function}\NormalTok{(dt) \{}
-
 \NormalTok{  dt <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}
     \DataTypeTok{data =}\NormalTok{ dt,}
-    \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"SD <= 20"}\NormalTok{,}
-                \StringTok{"NBS > 3"}\NormalTok{)}
-
+    \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}
+      \StringTok{"SD <= 20"}\NormalTok{,}
+      \StringTok{"NBS > 3"}
+\NormalTok{    )}
 \NormalTok{  )}
@@
 
-\NormalTok{    dt[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(dt, }
-                                       \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{, }
+\NormalTok{    dt[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}
+      \DataTypeTok{speed_in =} \KeywordTok{atl_get_speed}\NormalTok{(dt,}
+        \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{,}
         \DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
-                                       \DataTypeTok{type =} \StringTok{"in"}\NormalTok{),}
-              \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(dt, }
-                                       \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{, }
+        \DataTypeTok{type =} \StringTok{"in"}
+\NormalTok{      ),}
+      \DataTypeTok{speed_out =} \KeywordTok{atl_get_speed}\NormalTok{(dt,}
+        \DataTypeTok{x =} \StringTok{"X"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"Y"}\NormalTok{,}
         \DataTypeTok{time =} \StringTok{"time"}\NormalTok{,}
-                                       \DataTypeTok{type =} \StringTok{"out"}\NormalTok{))]}
+        \DataTypeTok{type =} \StringTok{"out"}
+\NormalTok{      )}
+\NormalTok{    )]}
 \NormalTok{  \})}
@@
 
-\NormalTok{  dt <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dt,}
-                              \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}\StringTok{"speed_in <= 20"}\NormalTok{,}
-                                          \StringTok{"speed_out <= 20"}\NormalTok{))}
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{atl_filter_covariates}\NormalTok{(}
+    \DataTypeTok{data =}\NormalTok{ dt,}
+    \DataTypeTok{filters =} \KeywordTok{c}\NormalTok{(}
+      \StringTok{"speed_in <= 20"}\NormalTok{,}
+      \StringTok{"speed_out <= 20"}
+\NormalTok{    )}
+\NormalTok{  )}
 \NormalTok{\})}
@@
 
-  \KeywordTok{lapply}\NormalTok{(}\DataTypeTok{X =}\NormalTok{ data_smooth, }
+  \KeywordTok{lapply}\NormalTok{(}
+    \DataTypeTok{X =}\NormalTok{ data_smooth,}
     \DataTypeTok{FUN =}\NormalTok{ atl_median_smooth,}
-         \DataTypeTok{time =} \StringTok{"time"}\NormalTok{, }\DataTypeTok{moving_window =} \DecValTok{5}\NormalTok{)}
+    \DataTypeTok{time =} \StringTok{"time"}\NormalTok{, }\DataTypeTok{moving_window =} \DecValTok{5}
+\NormalTok{  )}
 \NormalTok{)}
@@
 
-\NormalTok{  dt <-}\StringTok{ }\KeywordTok{merge}\NormalTok{(dt, }
-\NormalTok{              dt_recurse[, }\KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{, }\StringTok{"resTime"}\NormalTok{)], }
-              \DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{))}
+\NormalTok{  dt <-}\StringTok{ }\KeywordTok{merge}\NormalTok{(dt,}
+\NormalTok{    dt_recurse[, }\KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{, }\StringTok{"resTime"}\NormalTok{)],}
+    \DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"coordIdx"}\NormalTok{)}
+\NormalTok{  )}
 
@@
 \CommentTok{# add an id column}
-\NormalTok{data_residence[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}\DataTypeTok{id =}\NormalTok{ TAG,}
-                      \DataTypeTok{x =}\NormalTok{ X, }\DataTypeTok{y =}\NormalTok{ Y)]}
+\NormalTok{data_residence[, }\StringTok{`}\DataTypeTok{:=}\StringTok{`}\NormalTok{(}
+  \DataTypeTok{id =}\NormalTok{ TAG,}
+  \DataTypeTok{x =}\NormalTok{ X, }\DataTypeTok{y =}\NormalTok{ Y}
+\NormalTok{)]}
 
@@
 \NormalTok{data_patches <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(data_residence, atl_res_patch,}
-                       \DataTypeTok{buffer_radius =} \DecValTok{25}\NormalTok{)}
+  \DataTypeTok{buffer_radius =} \DecValTok{25}
+\NormalTok{)}
 \end{Highlighting}
@@
   \DataTypeTok{which_data =} \StringTok{"spatial"}\NormalTok{,}
-                        \DataTypeTok{buffer_radius =} \DecValTok{25}\NormalTok{)}
+  \DataTypeTok{buffer_radius =} \DecValTok{25}
+\NormalTok{)}
 
@@
 \KeywordTok{st_write}\NormalTok{(data_spatials,}
-         \DataTypeTok{dsn =} \StringTok{"data/data_bat_residence_patches.gpkg"}\NormalTok{)}
+  \DataTypeTok{dsn =} \StringTok{"data/data_bat_residence_patches.gpkg"}
+\NormalTok{)}
 \end{Highlighting}
@@
 \NormalTok{data_clean <-}\StringTok{ }\KeywordTok{fwrite}\NormalTok{(}\KeywordTok{rbindlist}\NormalTok{(data_smooth),}
-                     \DataTypeTok{file =} \StringTok{"data/data_bat_smooth.csv"}\NormalTok{)}
+  \DataTypeTok{file =} \StringTok{"data/data_bat_smooth.csv"}
+\NormalTok{)}
 \end{Highlighting}
@@
 \CommentTok{# write}
-\KeywordTok{fwrite}\NormalTok{(patch_summary,}
-       \StringTok{"data/data_bat_patch_summary.csv"}\NormalTok{)}
+\KeywordTok{fwrite}\NormalTok{(}
+\NormalTok{  patch_summary,}
+  \StringTok{"data/data_bat_patch_summary.csv"}
+\NormalTok{)}
 \end{Highlighting}
@@
 \centering
-\includegraphics{figures/fig_08_bats.png}
+\includegraphics{figures/fig_07_bats.png}
 \caption{A visual examination of plots of the bats' residence patches and linear approximations of paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site over the three nights \textbf{(a)}.

### 2021-02-04 _02_filtering_data.Rmd
change to seagreen

@@
 # define a four colour palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
+pal <- RColorBrewer::brewer.pal(5, "Set1")
+pal[3] = "seagreen"
 ```
@@
     aes(x, y),
-    size = 0.5,
+    # size = 0.5,
     alpha = 0.6,
-    shape = 4, col = pal[4]
+    shape = 4, col = pal[5]
   ) +

### 2021-02-04 _03_smoothing_tracks.Rmd
change green and blue

@@
 
-# prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
+# define a four colour palette
+pal <- RColorBrewer::brewer.pal(5, "Set1")
+pal[3] = "seagreen"
+pal[2] = RColorBrewer::brewer.pal(5, "Blues")[5]
 ```
@@
 
+### Plot of aggregation after smoothing
+
 ```{r echo=FALSE}
+# prepare data
+data_agg_smooth = copy(list_of_agg[[3]]) # 30s aggregate
 ### plot figures
-fig_agg_data <-
-  Map(
-    function(df, title) {
-      ggplot(df) +
-        # geom_path(data = data,
-        #           aes(x, y),
-        #           col = "grey20",
-        #           size = 0.2)+
+fig_agg_data_smooth <-
+      ggplot(data_agg_smooth) +
         geom_point(
           data = data_agg,
+          size = 0.2,
           aes(x, y),
-          size = 0.3,
           col = pal[3]
@@
         geom_path(aes(x, y),
-          col = pal[4]
+          col = pal[2],
+          lwd = 0.2
         ) +
@@
         ),
-        shape = 0,
-        col = pal[4],
+        shape = 1,
+        size = 2,
+        col = pal[2],
         alpha = 1,
@@
             margin = margin(t = 30, b = -30),
-            hjust = 0.8
+            hjust = 0.9
           )
@@
         labs(
-          title = sprintf("(%s)", title)
-        )
-    },
-    list_of_agg,
-    c(NA, NA, "a", "b")
+          title = "(a)"
         )
@@
 # aggregate before correction
-list_of_agg <- lapply(c(3, 10, 30, 120), function(z) {
+list_of_agg_errors <- lapply(c(3, 10, 30, 120), function(z) {
   data_return <- atl_thin_data(
@@
 
+### Figure aggregation before smoothing
+
+```{r echo=FALSE}
+# prepare data
+data_agg_error = copy(list_of_agg_errors[[3]]) # 30s aggregate
+### plot figures
+fig_agg_data_error <-
+      ggplot(data_agg_error) +
+        geom_point(
+          data = data_errors,
+          size = 0.1,
+          aes(x, y),
+          col = "grey"
+        ) +
+        geom_path(aes(x, y),
+          col = pal[2]
+          # lwd = 0.1
+        ) +
+        geom_point(aes(x, y,
+          group = interval
+        ),
+        shape = 1,
+        size = 2,
+        col = pal[2],
+        alpha = 1,
+        show.legend = F
+        ) +
+        ggthemes::theme_few() +
+        theme(
+          axis.text = element_blank(),
+          axis.title = element_blank(),
+          plot.title = element_text(
+            face = "bold",
+            margin = margin(t = 30, b = -30),
+            hjust = 0.9
+          )
+        ) +
+        coord_cartesian(ylim = c(0.6, NA)) +
+        labs(
+          title = "(b)"
+        )
+```
+
 ### Comparing median aggregation with and without smoothing
@@
 # now plot distribution of speed
-data_agg <- rbindlist(list_of_agg)
+data_agg <- rbindlist(list_of_agg_errors)
 ```
@@
     axis.text.y = element_blank(),
-    axis.text = element_blank(),
-    axis.title = element_blank(),
     plot.title = element_text(
@@
   wrap_plots(
-    append(fig_agg_data[3:4], list(fig_agg_speed)),
+    fig_agg_data_smooth, fig_agg_data_error, fig_agg_speed,
     design = "AABBC"

### 2021-02-04 _02_filtering_data.Rmd
return to outside panel title

@@
 # make a copy
-data_copy <- copy(data)
-
-# add a prolonged spike or reflection to 300 positions
-data_copy[500:800, `:=`(
-  x = x + 0.25,
-  y = y + 0.25
-)]
-
-# add normal error
-data_copy[, `:=`(
-  x = do_add_error(x, std_dev = 0.01),
-  y = do_add_error(y, std_dev = 0.005)
-)]
-
-# add 100 outliers
-data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
+# data_copy <- copy(data)
+# 
+# # add a prolonged spike or reflection to 300 positions
+# data_copy[500:800, `:=`(
+#   x = x + 0.25,
+#   y = y + 0.25
+# )]
+# 
+# # add normal error
+# data_copy[, `:=`(
+#   x = do_add_error(x, std_dev = 0.01),
+#   y = do_add_error(y, std_dev = 0.005)
+# )]
+# 
+# # add 100 outliers
+# data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
 ```
@@
 ```{r eval=FALSE}
-fwrite(data_copy, file = "data/data_errors.csv")
+# fwrite(data_copy, file = "data/data_errors.csv")
 ```
@@
     axis.text = element_blank(),
-    axis.title = element_blank(),
-    plot.title = element_text(
-      face = "bold",
-      margin = margin(t = 30, b = -30),
-      hjust = 0.1
-    )
+    axis.title = element_blank()
   ) +
   theme(plot.background = element_rect(fill = NA)) +
-  coord_equal(expand = T) +
-  labs(
-    title = "(a)"
-  )
+  coord_equal(expand = T)
 ```
@@
     alpha = 0.6,
-    shape = 4, col = pal[5]
+    shape = 4, col = "black"
   ) +
@@
   ) +
-  scale_color_manual(values = c(pal[3], pal[1])) +
+  scale_color_manual(values = c(pal[3], "grey")) +
   scale_shape_manual(values = c(1, 4)) +
@@
     axis.text = element_blank(),
-    axis.title = element_blank(),
-    plot.title = element_text(
-      face = "bold",
-      margin = margin(t = 30, b = -30),
-      hjust = 0.1
-    )
+    axis.title = element_blank()
   ) +
   coord_equal() +
-  theme(plot.background = element_rect(fill = NA)) +
-  labs(
-    title = "(b)"
-  )
+  theme(plot.background = element_rect(fill = NA))
 ```
@@
     fig_outlier_remove
-  )
+  )+
+    plot_annotation(
+      tag_levels = "a",
+      tag_prefix = "(",
+      tag_suffix = ")"
+    ) &
+  theme(plot.tag = element_text(face = "bold"))
 

### 2021-02-04 _03_smoothing_tracks.Rmd
outside panel titles

@@
 pal[3] = "seagreen"
-pal[2] = RColorBrewer::brewer.pal(5, "Blues")[5]
 ```
@@
   scale_colour_distiller(
-    palette = "Blues", direction = 1,
+    palette = "Purples", direction = 1,
     values = c(-0.5, 1)
@@
         geom_path(aes(x, y),
-          col = pal[2],
+          col = pal[4],
           lwd = 0.2
@@
         ),
-        shape = 1,
+        shape = 0,
         size = 2,
-        col = pal[2],
+        col = pal[4],
         alpha = 1,
@@
           axis.text = element_blank(),
-          axis.title = element_blank(),
-          plot.title = element_text(
-            face = "bold",
-            margin = margin(t = 30, b = -30),
-            hjust = 0.9
-          )
+          axis.title = element_blank()
         ) +
-        coord_cartesian(ylim = c(0.6, NA)) +
-        labs(
-          title = "(a)"
-        )
+        coord_cartesian(ylim = c(0.6, NA))
 ```
@@
         geom_path(aes(x, y),
-          col = pal[2]
+          col = pal[4]
           # lwd = 0.1
@@
           axis.text = element_blank(),
-          axis.title = element_blank(),
-          plot.title = element_text(
-            face = "bold",
-            margin = margin(t = 30, b = -30),
-            hjust = 0.9
-          )
+          axis.title = element_blank()
         ) +
-        coord_cartesian(ylim = c(0.6, NA)) +
-        labs(
-          title = "(b)"
-        )
+        coord_cartesian(ylim = c(0.6, NA))
 ```
@@
   theme(
-    axis.text.y = element_blank(),
-    plot.title = element_text(
-      face = "bold",
-      margin = margin(t = 30, b = -30),
-      hjust = 0.8
-    )
+    axis.text.y = element_blank()
   ) +
@@
     x = "interval (s)",
-    y = "speed",
-    title = "(c)"
+    y = "speed"
   )
@@
     design = "AABBC"
-  )
+  )+
+    plot_annotation(
+      tag_levels = "a",
+      tag_prefix = "(",
+      tag_suffix = ")"
+    ) &
+  theme(plot.tag = element_text(face = "bold"))
 

### 2021-02-04 05_calibration_data.Rmd
edits colours only

@@
 
-# prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
+pal <- RColorBrewer::brewer.pal(5, "Set1")
+pal[3] = "seagreen"
+pal[2] = RColorBrewer::brewer.pal(5, "Blues")[5]
 ```
@@
     alpha = ifelse(data_raw$x > 645000, 1, 1),
-    size = ifelse(data_raw$x > 645000, 0.3, 2),
+    size = ifelse(data_raw$x > 645000, 0.3, 4),
+    shape = ifelse(data_raw$x > 645000, 16, 4),
     show.legend = F
   ) +
-  scale_colour_manual(values = c("black", pal[3])) +
+  scale_colour_manual(values = c(pal[1], pal[3])) +
   geom_vline(
@@
     axis.title = element_blank()
-  ) +
+  )+
   ggspatial::annotation_scale(location = "tl") +
-  coord_sf(crs = 32631) +
-  labs(colour = "speed (m/s)")
-
+  coord_sf(crs = 32631)
 # save
@@
   ggplot() +
+  geom_path(
+    data = data_raw[!data_unproc, on = c("x", "y")],
+    aes(x, y),
+    col = "grey90",
+    size = 0.1
+  ) +
   geom_point(
@@
     col = pal[3],
-    shape = 19,
+    shape = 1,
+    stroke = 1,
     alpha = 0.5
@@
     aes(x, y),
-    alpha = 0.6,
     col = pal[3],
@@
     design = "AB"
-  ) +
-    plot_annotation(
-      tag_levels = "a",
-      tag_prefix = "(",
-      tag_suffix = ")"
-    ) &
-    theme(plot.tag = element_text(face = "bold"))
+  )
 
@@
 
+## Plot figure 7
+
 ```{r echo=FALSE}

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
add main text tex

--- 2021-02-03 docs/ms_atlas_preproc_main_text_2021-02-03.tex
@@
     \caption{A general, modular pipeline for pre-processing high-throughput tracking data from raw localisations to cleaned data, and optionally into residence patches. 
-    Users should begin by plotting their data to determine its status, for instance whether it needs to be susbet within a certain time interval, or whether there are obvious location errors.
-    Users can then pre-process their data using the appropriate module, following the pre-processing steps corresponding to each module.
+    Users should determine the appropriate pre-processing module, and work their way through the corresponding pre-processing steps, until the data are suitable for their intended purpose.
     The \texttt{atlastools} function that may be used to implement each pre-processing step is shown in the grey boxes underneath each step.
-    Users may use the pipeline in full or just the module most appropriate for their data.
-    In all cases, users are strongly encouraged to visualise their data and scan it for location errors as a first step, and to visualise it along the way.
+    Popular \texttt{R} packages in movement ecology, or statistical methods are shown underneath suggested analyses (yellow boxes).
+    Users are strongly encouraged to visualise their data and scan it for location errors as a first step, and as they work through the pipeline.
     }
@@
 
-    \item After data access, we first encourage users to visualise their data to check for evident location errors \citep{slingsby2016}.
+    \item Users should visualise their data to check for evident location errors \citep{slingsby2016}.
     \item Users can install \texttt{atlastools} using the \texttt{install\_github} function from the \texttt{R} package \texttt{remotes}.

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit numbered overview

@@
     \item Users should visualise their data to check for evident location errors \citep{slingsby2016}.
-    \item Users can install \texttt{atlastools} using the \texttt{install\_github} function from the \texttt{R} package \texttt{remotes}.
-    The package can be installed using the \texttt{R} command
+    \item Users can install \texttt{atlastools} using the \texttt{install\_github} function from the \texttt{R} package \texttt{remotes}:
 
@@
 
-    Package functions are prefixed `\texttt{atl\_}', and rely heavily on the \texttt{data.table} package (Dowle and Srinivasan 2020).
+    Package functions are prefixed `\texttt{atl\_}', and are based on the \texttt{data.table} package (Dowle and Srinivasan 2020).
     Certain functions modify the input data "in place", and users must work on a copy of their data to preserve both the original and cleaned data (details in the Supplementary Material).
-    Since ATLAS systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} assumes a coordinate system in metres, and users must convert their geographic coordinates to metres.
-    \item Users with a specific area or time of interest can use simple spatio-temporal range filters to select positions (see \textsc{Spatio-Temporal Filtering}).
-    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by an error measure, or by the plausibility of associated movement metrics (see \textsc{Filtering to Reduce Location Errors}).
+    Since high-throughput tracking systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} users must convert their geographic coordinates to metres as this is assumed by all functions.
+    \item Users can use simple spatio-temporal range filters to select positions within a certain time or area (see \textsc{Spatio-Temporal Filtering}).
+    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by a system-specific error measure, or by the plausibility of associated movement metrics (see \textsc{Filtering to Reduce Location Errors}).
     \item Users should then reduce small-scale location errors by applying a median smooth (see \textsc{Smoothing and Thinning Data}).
-    \item Users who need uniformly thinned data can then thin by either aggregating or resampling (see \textsc{Smoothing and Thinning Data}).
-    At this stage, the data are ready for a number of popular statistical treatments such as Hidden Markov Model-based classification \citep{michelot2016}.
+    \item Users who need uniformly thinned data can either aggregate or resample it (see \textsc{Smoothing and Thinning Data}).
+    At this stage, the data are ready for a number of popular statistical treatments such as Hidden Markov Model-based classification \citep{michelot2016,langrock2012}.
     \item Finally, users wishing to study prolonged animal residence in an area can classify their data into residence patches based on the movement ecology of their study system, after filtering out non-stationary positions (see \textsc{Synthesising Movement Tracks into Residence Patches}).

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit figure 2 caption

@@
     \item Finally, users wishing to study prolonged animal residence in an area can classify their data into residence patches based on the movement ecology of their study system, after filtering out non-stationary positions (see \textsc{Synthesising Movement Tracks into Residence Patches}).
-    % Residence patches are spatio-temporally independent clusters of positions that represent prolonged residence of a tracked individual in an area, and which can indicate aspects of space use such resource exploitation \citep{barraquand2008, bijleveld2016, oudman2018}.
 
@@
     \includegraphics[width=0.95\textwidth]{figures/fig_02_filtering_data.png}
-    \caption{\textbf{(a)} A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the \texttt{smoove} package (Gurarie et al.
-2017).
-Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track.
-    The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (black line) as closely as possible.  
-    Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the \texttt{atlastools} function  \texttt{atl\_filter\_bounds}.
-    Setting the function argument \texttt{remove\_inside = FALSE} retains positions within user supplied bounds (dashed grey lines; green points), and excludes those outside (grey points).
-    Reducing large-scale outliers in a movement track.
-The goal is to remove positions far away from the canonical track (black line), and retain those with only reasonable displacements from the canonical track (green points).
-    \textbf{(b)} Location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles).
-    The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
-    Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
-    Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
-    The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
-    While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult.
-    Users are cautioned to frequently check this and similar methods' results.
+    \caption{\textbf{(a)} Simulated movement data (black line), with three kinds of artificially added errors: (1) normally distributed small-scale error on each position, (2) large-scale error added to 0.5\% of positions, and (3) 300 consecutive positions displaced to simulate a gross distortion affecting a continuous subset of the track.
+    Tracks can be quickly filtered by spatial bounds (dashed grey lines) to exclude broad regions.
+    \textbf{(b)} 
+    Location error may affect single observations resulting in point outliers or `spikes' (grey crosses), or continuous subsets of a track, called a `prolonged spike' (black crosses, top right), and both represent unrealistic movement.
+    While spikes can be removed by filtering out positions with high incoming and outgoing speeds and turning angle, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
+    These and other distortions that affect continuous track segments can be removed by conceptualising algorithms that find the bounds of the distortion and remove points in between them.
+    Users should frequently check the outputs of such algorithms to avoid rejecting valid data.
     }

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit figure 4 aggregation  caption

@@
     }
-    \label{fig:figure_simulated_data}
+    \label{fig:figure_filtering_data}
 \end{figure}
@@
     \includegraphics[width=0.95\textwidth]{figures/fig_04_thinning.png}
-    \caption{Thinning a movement track using aggregation preserves track structure while reducing data volume, but thinning without removing gross location errors and unrealistic movement affects track metrics such as speed.
-    (a, b) Movement track positions with an interval of 1 second (green points) aggregated over intervals of (a) 10 and (b) 30 seconds (purple circles), after removing large location errors or filtering unrealistic movement.
-    The canonical track is shown as a black line in both panels, while the red line shows the track connecting thinned positions.
-    (c) Boxplots show the median and interquartile ranges for speed estimates of tracks aggregated over intervals of 3, 10, 30, and 120 seconds, but without the removal of location errors.
+    \caption{\textbf{(a)} Aggregating a filtered and smoothed movement track (green points) preserves track structure while reducing data volume, but \textbf{(b)} aggregating before filtering gross location errors and unrealistic movement (grey points) leads to the persistence of large-scale errors (such as prolonged spikes).
+    The thinned track is shown as purple squares and lines in both panels.
+    \textbf{(c)} Thinning before data cleaning can lead to significant misestimations of essential movement metrics such as speed at lower intervals.
+    Boxplots show the median and interquartile ranges for speed estimates of tracks aggregated over intervals of 3, 10, 30, and 120 seconds, but without the removal of location errors.
     For comparison, the median and 95\textsuperscript{th} percentile of speed of the canonical track are shown as solid and dashed horizontal lines, respectively.
-    Green circles and associated error bars show, in contrast, the median ($\pm$ standard deviation) speed for the same aggregation interval, but after removing large- and small-scale location errors (median smooth $K$ = 11).
-    Removing location errors and unrealistic movement before aggregation substantially improves speed estimates, especially at smaller aggregation intervals, as it removes the apparent spikes that contribute to over estimation of speed if not filtered.}
+    Green circles and associated error bars show, in contrast, the median ($\pm$ standard deviation) speed for the same aggregation interval, but after removing large- and small-scale location errors (median smooth $K$ = 11).}
     \label{fig:figure_thinning}
@@
 Furthermore, metrics such as the distance travelled within and between patches can help compare broad-scale individual movement strategies.
-% Here, high-throughput tracking every 1s -- 30s has its advantages: users may be permitted the assumption that an individual is truly absent from a location where it was not reported, and that at least during the tracking period, such absences are not simply artefacts of the sampling interval.
 

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit figure res patch concept caption

@@
     \includegraphics[width=0.75\textwidth]{figures/fig_05_residence.png}
-    \caption{A rotation-advection movement track, similar to the movement of a soaring bird (Gurarie et al. 2017) classified into residence patches with the \texttt{atlastools} function.
-    (a) The goal is to segment-cluster the track into areas of prolonged residence (clusters of points) while leaving out the transit between them.
-    (b) A plot of residence time against time (solid line; Bracis et al. 2018) shows how the residence patch algorithm segments and clusters positions of prolonged residence. 
+    \caption{Movement tracks can be classified into residence patches, whcih are areas of prolonged residence (clusters of points), while leaving out the transit between them.
+    \textbf{(a)} The residence patch method correctly identifies clusters of positions where the individual is relatively stationary (orange and green patches), as well as positions where it is slowly moving (blue patch).
+    This is especially useful when studying more complex behaviour such as area-restricted search, which may have a directional component.
+    The residence patch method loses the details of movement between patches, but can efficiently represent the general pattern of space-use (see coloured points representing patch centroids, and lines joining them).
+    \textbf{(b)} A plot of residence time against time (solid line; Bracis et al. 2018) shows how the residence patch algorithm segments and clusters positions of prolonged residence. 
     Regions are shaded by the temporal bounds of each residence patch.
     The arguments passed to \texttt{atl\_res\_patch} determine the clustering, and can be adjusted to get a result that fits the study system.
-    Users are cautioned that there are no `correct' arguments.
-    (c) The residence patch method correctly identifies clusters of positions where the individual is relatively stationary (orange and green patches), as well as positions where it is slowly moving (blue patch).
-    This is especially useful when studying more complex behaviour such as area-restricted search, which may have a directional component.
-    The function \texttt{atl\_patch\_summary} conveniently presents the classified data for further use, either as an \texttt{sf} object (left: coloured polygons around clustered points), or as summary data of each patch (right: coloured circles).
-    While the \texttt{sf} object retains the patch geometry as well as the patch attributes (such as duration), the summary data discards the patch geometry.
-    Both spatial and summary objects lose the details of movement between patches (compare black lines of the canonical track [left], and lines connecting patch centroids [right]), but are helpul for questions where the general structure is more relevant.}
+    Users are cautioned that there are no `correct' arguments, and the best guide is the biology of the tracked individual.}
     \label{fig:figure_residence_patch}

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit figure calibration  caption

@@
     \caption{Pre-processing steps for WATLAS calibration data showing filtering filtering on speed, median smoothing and thinning by aggregation, and making residence patches.
-    \textbf{(a)} Positions with incoming and outgoing speed $\geq$ the 15 m/s are removed (N removed = 13,491, 26.6\%; grey crosses = removed, green points = retained).
+    \textbf{(a)} Positions with incoming and outgoing speed $\geq$ 15 m/s are removed (grey crosses = removed, green points = retained).
     Rectangles show the region expanded in subfigures (b) and (c).
@@
     Square size corresponds to the number of positions used to calculate the averaged position during thinning, with no significant differences.
-    Median smoothing retains all 37,324 observations (though changing their coordinates). Thinning aggregates these positions into 1,803 aggregates (4.8\% of the smoothed data volume).
+    While median smoothing retains all observations (though changing their coordinates), thinning aggregates these positions into approx. 5\% of the smoothed data volume.
     Median smoothing recovers a good estimate of the true track from significantly over-dispersed raw data, while thinning reduces the data volume but preserves track structure.

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit figure bat caption

@@
     \includegraphics[width=0.75\textwidth]{figures/fig_07_bats.png}
-    \caption{Synthesising animal tracks into residence patches can reveal how individuals in a population move in relation to landscape features, prior exploration, and other individuals.
-    \textbf{(a)} Linear approximations of the paths (coloured lines) between residence patches (circles) show the movement of three Egyptian fruit bats (\textit{Rousettus aegyptiacus}), tracked over three nights in the Hula Valley, Israel.
-    Bat tracks are are shown as thin lines below the linear approximations.
-    Colours show bat identity. The grey hexagon represents a roost site.
+    \caption{Synthesising animal tracks into residence patches can reveal movement in relation to landscape features, prior exploration, and other individuals.
+    \textbf{(a)} Linear approximations of the paths (coloured lines) between residence patches (circles) of three Egyptian fruit bats (\textit{Rousettus aegyptiacus}), tracked over three nights in the Hula Valley, Israel.
+    Bat tracks are are shown as thin lines below the linear approximations, and colours show bat identity. The grey hexagon represents a roost site.
     Dark green points represent known fruit trees.

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit out SD calc

@@
 Tracking data covariates can be good indicators of the reliability of calculated positions (Beardsworth et al. \textit{in prep.}).
-For intance, in ATLAS systems the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
+For intance, in reverse-GPS ATLAS systems the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
+Reverse-GPS systems also calculate direct measures of location error during localisation \citep{maccurdy2009, maccurdy2019, weiser2016}.
 GPS systems' location error may be similarly indicated indirectly by the number of satellites involved in the localisation, or directly by an error measure such as the Horizontal Dilution of Precision (HDOP).
-ATLAS and other TOA systems also calculate direct measures of location error during localisation: \texttt{VARX}, \texttt{VARY}, and \texttt{COVXY}, which are the variance of the X and Y coordinates, and the covariance of the X and Y coordinates, respectively \citep{maccurdy2009, maccurdy2019, weiser2016}.
-A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
-\begin{linenomath*}
-    \begin{equation*}
-        SD = \sqrt{{VARX} + {VARY} + 2 \times {COVXY}}
-     \end{equation*}
-\end{linenomath*}
-Filtering on the position-specific standard deviation allows removing unreliable positions, and the filter can be applied using \texttt{atl\_filter\_covariates}.
+ATLAS and other TOA systems 
+% A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
+% \begin{linenomath*}
+%     \begin{equation*}
+%         SD = \sqrt{{VARX} + {VARY} + 2 \times {COVXY}}
+%      \end{equation*}
+% \end{linenomath*}
+Unreliable positions can be removed by filtering on direct or indirect measures of quality using \texttt{atl\_filter\_covariates}.
 
 \begin{lstlisting}[float, language=R, style=customR, caption = {
-    Filtering ATLAS data on position covariates. 
+    An example of filtering ATLAS data on position covariates. 
     The \texttt{filters} argument accepts a character vector with the logical statements. 
@@
 \end{lstlisting}
-
-% \begin{figure}[h!]
-%     \centering
-%     \includegraphics[width=0.95\textwidth]{figures/fig_03_filter_speed.png}
-%     \caption{Reducing large-scale outliers in a movement track.
-The goal is to remove positions far away from the canonical track (black line), and retain those with only reasonable displacements from the canonical track (green points).
-%     \textbf{(a)} location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles).
-%     The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
-%     Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
-%     Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
-%     \textbf{(b)} The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
-%     While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult.
-%     Users are cautioned to frequently check this and similar methods' results.}
-%     \label{fig:figure_filter_bounds}
-% \end{figure}
 

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit out SD in thinning

@@
 When users want to account for location error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location error when thinning is important.
-In ATLAS systems the location error (\texttt{SD}; see \textsc{Filtering on Position Covariates}) is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
-The error around each coordinate (\texttt{VARX} or \texttt{VARY} in ATLAS systems) can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
+% In reverse-GPS systems systems the location error is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
+The location error around each coordinate provided by either GPS or reverse-GPS system can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
 \begin{linenomath*}
@@
 \end{linenomath*}
-Similarly, the overall location error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance ($SD ^ 2$):
-\begin{linenomath*}
-    \begin{equation*}
-        SD_{agg} = \sqrt{ \left( \sum_{i=1}^{i=N} SD_i^2 \right) / N ^ 2  }
-    \end{equation*}    
-\end{linenomath*}
+% Similarly, the overall location error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance ($SD ^ 2$):
+% \begin{linenomath*}
+%     \begin{equation*}
+%         SD_{agg} or HDOP_{agg} = \sqrt{ \left( \sum_{i=1}^{i=N} SD_i^2 \right) / N ^ 2  }
+%     \end{equation*}    
+% \end{linenomath*}
 
 Users may question why thinning should be implemented after cleaning steps, when aggregation can obtain consensus positions over an interval, correctly propagate the location error, and also reduce data volumes.
+%%
 We caution users that thinning causes an extensive loss of small-scale detail in the data.

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
redo abstract

@@
     \item Modern, high-throughput animal tracking studies collect increasingly large volumes of data at very fine temporal scales.
-    At these scales, location error can exceed the animal's step size, confounding inferences from tracking data.
-    `Cleaning' the data to exclude positions with large location errors prior to analyses is one of the main ways movement ecologists deal with location errors.
-    Cleaning data to reduce location error before making biological inferences is widely recommended, and ecologists routinely consider cleaned data to be the ground-truth.
-    Nonetheless, uniform guidance on this crucial step is scarce.
+    At these scales, location error can exceed the animal's step size, leading to mis-estimation of movement metrics such as speed.
+    % Sophisticated methods that estimate both the error and the movement metrics are not easily scalable to massive datasets, and can be difficult to use for both novice and veteran animal ecologists.
+    `Cleaning' the data to reduce location errors prior to analyses is one of the main ways movement ecologists deal with noisy data, and has the advantage of being more scalable to massive datasets than more complex methods.
+    Though data cleaning is widely recommended, and ecologists routinely consider cleaned data to be the ground-truth, uniform guidance on this crucial step is scarce.
 
-    \item Cleaning high-throughput data must strike a balance between rejecting location errors without discarding valid animal movements.
-    Additionally, users of high-throughput systems face challenges resulting from the high volume of data itself, since processing large data volumes is computationally intensive and difficult without a common set of efficient tools.
-    Furthermore, many methods that cluster movement tracks for ecological inference are based on statistical phenomena, and may not be intuitive to understand in terms of the tracked animal's biology.
+    \item A pipeline for cleaning massive high-throughput datasets must similarly balance ease of use, computational efficiency, and the ability to prepare movement data for a range of analyses.
+    At its core, cleaning high-throughput data must strike a balance between rejecting location errors without discarding valid animal movements.
+    Additionally, a pre-processing pipeline must be capable of efficiently segmenting and clustering location data for statistical methods, while also being scalable to large datasets and robust to imperfect sampling.
+    Manual methods being prohibitively time consuming, and to boost reproducibility, a robust pre-processing pipeline must be automated.
 
     \item In this article we introduce a pipeline to pre-process high-throughput animal tracking data in order to prepare it for subsequent analysis.
+    This pipeline, consisting of removing outliers, smoothing the filtered result, and thinning it to a uniform sampling interval, is easily scaled to very large datasets.
     We demonstrate this pipeline on simulated movement data to which we have randomly added location errors.
-    We further suggest how large volumes of cleaned data may be synthesized into biologically meaningful ‘residence patches’.
-    We then use calibration data to show how the pipeline improves its quality, and to verify that the residence patch synthesis accurately captures animal space-use.
-    Finally, turning to real tracking data from Egyptian fruit bats (\textit{Rousettus aegyptiacus}), we demonstrate the pre-processing pipeline and residence patch method in a fully worked out example.
+    We further suggest how large volumes of cleaned data can be segmented-clustered into biologically meaningful ‘residence patches’.
+    We use calibration data to show how the pipeline improves its quality, and to verify that the residence patch synthesis accurately captures animal space-use.
+    Finally, turning to tracking data from Egyptian fruit bats (\textit{Rousettus aegyptiacus}), we demonstrate the pre-processing pipeline and residence patch method in a fully worked out example.
 
-    \item To help with fast implementations of our pipeline, and to help standardise methods, we developed the \texttt{R} package \texttt{atlastools}, which we introduce here.
-    Our pre-processing pipeline and \texttt{atlastools} can be used with any high-throughput animal movement data in which the high data volume combined with knowledge of the tracked individuals’ biology can be used to reduce location errors.
-    The use of common pre-processing steps that are simple yet robust promotes standardised methods in the field of movement ecology and better inferences from data.
+    \item To help with fast implementation, and to help standardise methods, we developed the \texttt{R} package \texttt{atlastools}, which we also introduce here.
+    Our pre-processing pipeline and \texttt{atlastools} can be used with any high-throughput animal movement data in which the high data volume combined with knowledge of the tracked individuals’ movement capacity can be used to reduce location errors.
+    The \texttt{atlastools} function is easy to use for beginners, while providing a template for further development.
+    The use of common pre-processing steps that are simple yet robust promotes standardised methods in the field of movement ecology and leads to better inferences from data.
 \end{enumerate}
 
-\textbf{Keywords} ATLAS, data cleaning, movement ecology, high-throughput tracking, \textit{R} package \textit{atlastools}, residence patch
+\textbf{Keywords} ATLAS, data cleaning, movement ecology, high-throughput tracking, \textit{atlastools}, residence patch
 
@@
     Since high-throughput tracking systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} users must convert their geographic coordinates to metres as this is assumed by all functions.
-    \item Users can use simple spatio-temporal range filters to select positions within a certain time or area (see \textsc{Spatio-Temporal Filtering}).
-    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by a system-specific error measure, or by the plausibility of associated movement metrics (see \textsc{Filtering to Reduce Location Errors}).
-    \item Users should then reduce small-scale location errors by applying a median smooth (see \textsc{Smoothing and Thinning Data}).
-    \item Users who need uniformly thinned data can either aggregate or resample it (see \textsc{Smoothing and Thinning Data}).
+    \item Users can use simple spatio-temporal range filters to select positions within a certain time or area.
+    %  (see \textsc{Spatio-Temporal Filtering}).
+    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by a system-specific error measure, or by the plausibility of associated movement metrics.
+    %  (see \textsc{Filtering to Reduce Location Errors}).
+    \item Users should then reduce small-scale location errors by applying a median smooth.
+    %  (see \textsc{Smoothing and Thinning Data}).
+    \item Users who need uniformly thinned data can either aggregate or resample it.
+    %  (see \textsc{Smoothing and Thinning Data}).
     At this stage, the data are ready for a number of popular statistical treatments such as Hidden Markov Model-based classification \citep{michelot2016,langrock2012}.

### 2021-02-04 05_calibration_data.Rmd
more style

@@
 pal <- RColorBrewer::brewer.pal(5, "Set1")
-pal[3] = "seagreen"
-pal[2] = RColorBrewer::brewer.pal(5, "Blues")[5]
+pal[3] <- "seagreen"
+pal[2] <- RColorBrewer::brewer.pal(5, "Blues")[5]
 ```
@@
     axis.title = element_blank()
-  )+
+  ) +
   ggspatial::annotation_scale(location = "tl") +

### 2021-02-04 R/_03_smoothing_tracks.R
add output scripts

@@ -5,8 +5,9 @@ library(atlastools)
 library(ggplot2)
 library(patchwork)
 
-# prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
+# define a four colour palette
+pal <- RColorBrewer::brewer.pal(5, "Set1")
+pal[3] <- "seagreen"
 
 
 ## ----read_sim_data_2----------------------------------------------------------
@@ -91,7 +92,7 @@ figure_median_smooth <-
     fontface = "bold"
   ) +
   scale_colour_distiller(
-    palette = "Blues", direction = 1,
+    palette = "Purples", direction = 1,
     values = c(-0.5, 1)
   ) +
   ggthemes::theme_few() +
@@ -141,50 +142,36 @@ speed_agg_smooth <- rbindlist(speed_agg_smooth)
 
 
 ## ----echo=FALSE---------------------------------------------------------------
+# prepare data
+data_agg_smooth <- copy(list_of_agg[[3]]) # 30s aggregate
 ### plot figures
-fig_agg_data <-
-  Map(
-    function(df, title) {
-      ggplot(df) +
-        # geom_path(data = data,
-        #           aes(x, y),
-        #           col = "grey20",
-        #           size = 0.2)+
-        geom_point(
-          data = data_agg,
-          aes(x, y),
-          size = 0.3,
-          col = pal[3]
-        ) +
-        geom_path(aes(x, y),
-          col = pal[4]
-        ) +
-        geom_point(aes(x, y,
-          group = interval
-        ),
-        shape = 0,
-        col = pal[4],
-        alpha = 1,
-        show.legend = F
-        ) +
-        ggthemes::theme_few() +
-        theme(
-          axis.text = element_blank(),
-          axis.title = element_blank(),
-          plot.title = element_text(
-            face = "bold",
-            margin = margin(t = 30, b = -30),
-            hjust = 0.8
-          )
-        ) +
-        coord_cartesian(ylim = c(0.6, NA)) +
-        labs(
-          title = sprintf("(%s)", title)
-        )
-    },
-    list_of_agg,
-    c(NA, NA, "a", "b")
-  )
+fig_agg_data_smooth <-
+  ggplot(data_agg_smooth) +
+  geom_point(
+    data = data_agg,
+    size = 0.2,
+    aes(x, y),
[…]

### 2021-02-04 _02_filtering_data.Rmd

@@
 pal <- RColorBrewer::brewer.pal(5, "Set1")
-pal[3] = "seagreen"
+pal[3] <- "seagreen"
 ```
@@
     fig_outlier_remove
-  )+
+  ) +
     plot_annotation(

### 2021-02-04 _03_smoothing_tracks.Rmd

@@
 # prepare data
-data_agg_smooth = copy(list_of_agg[[3]]) # 30s aggregate
+data_agg_smooth <- copy(list_of_agg[[3]]) # 30s aggregate
 ### plot figures
@@
 # prepare data
-data_agg_error = copy(list_of_agg_errors[[3]]) # 30s aggregate
+data_agg_error <- copy(list_of_agg_errors[[3]]) # 30s aggregate
 ### plot figures
@@
     design = "AABBC"
-  )+
+  ) +
     plot_annotation(

### 2021-02-04 _02_filtering_data.Rmd
change colour

@@
     alpha = 0.6,
-    shape = 4, col = "black"
+    shape = 4, col = "grey30"
   ) +

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
reduce listings

@@
     The \texttt{atl\_filter\_bounds} function removes positions outside an area defined by coordinate ranges, a polygon, or all three (\texttt{remove\_inside = FALSE}), or positions inside the area (\texttt{remove\_inside = TRUE}).
-    The arguments \texttt{x} and \texttt{y} determine which columns are considered the X and Y coordinates, the arguments \texttt{x\_range} and \texttt{y\_range} determine the acceptable range of coordinates in a reference system based in metres, while the \texttt{sf\_polygon} argument allows the data to be filtered by the user specified \texttt{sf-(MULTI)POLYGON} object.
+    The arguments \texttt{x} and \texttt{y} determine the X and Y coordinate columns, \texttt{x\_range} and \texttt{y\_range} are the filter bounds in a CRS in metres, and the data can be filtered by an \texttt{sf-(MULTI)POLYGON} can be passed using the \texttt{sf\_polygon} argument.
     \texttt{atl\_filter\_bounds} returns a filtered \texttt{data.table}, which must be saved as an object (here, \texttt{filtered\_data}).}]
@@
 A spatio-temporal filter that retains positions inside the bounds of roosts during high-tide can quickly return data that reveals individual presence at high-tide roosts, while excluding positions that erroneously place individuals on the inundated mudflats.
-% Howevever, in this example, such a filter would also exclude potentially exploratory movements over the flooded mudflats.
-% We caution users to be certain that filters do not exclude positions that reveal interesting aspects of their animals' movement.
 Users should apply filters in sequence rather than all at once, and visualise the output after each filtering step (`sanity checks').
@@
 \begin{lstlisting}[float, language=R, style=customR, caption = {
-    The \texttt{atl\_filter\_covariates} function can be used both as a simple temporal filter, and also as a combined spatio-temporal filter. 
-    Filter predicates are passed to the \texttt{filters} argument as a character vector, each of which is then evaluated as an \texttt{R} expression in the context of the data supplied. 
+    Data can be filtered by a temporal or a spatio-temporal range using {atl\_filter\_covariates}. 
+    Filter predicates are passed to the \texttt{filters} argument as a character vector. 
     Only rows in the data satisfying all the conditions passed as filters are retained. 
-    Users must make sure the filter variables exist in their dataset.
-    Here, the first example shows how nighttime data can be retained using a predicate (\texttt{inrange} from \texttt{data.table}) that determine whether the value of `hour' is between 6 and 18. 
-    The \texttt{!} sign indicates that the negative of the predicate should be returned, i.e., \texttt{TRUE} when the hour is between 6 PM and 6 AM, but \texttt{FALSE} when the hour is between 6 AM and 6 PM.
-    The second example shows the use of multiple filter statements; this data will be filtered to be within the range of times bounded by \texttt{t\_min} and \texttt{t\_max}, and with X coordinates between \texttt{x\_min} and \texttt{x\_max}.
-    The \texttt{between} function is from \texttt{data.table}.}]
+    Here, the first example shows how nighttime data can be retained using a predicate (\texttt{inrange} from \texttt{data.table}) that determine whether the value of `hour' is between 6 and 18, and also within a range of X coordinates.
+    The second example retains calculated using $>$ 3 base stations (\texttt{NBS}), with location error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.
+    }]
 night_data <- atl_filter_covariates(data = dataset,
-                filters = c("!inrange(hour, 6, 18)"))
+                filters = c("!inrange(hour, 6, 18)",
+                            "between(x, x_min, x_max)"))
 
-data_in_area <- atl_filter_covariates(data = dataset,
-                    filters = c("between(time, t_min, t_max)",
-                                "between(x, x_min, x_max)"))
+filtered_data <- atl_filter_covariates(data = data,
+                    filters = c("NBS > 3",
+                                "SD < 100",
+                                "between(day, 5, 8)"))                            
 \end{lstlisting}
@@
 Unreliable positions can be removed by filtering on direct or indirect measures of quality using \texttt{atl\_filter\_covariates}.
-
-\begin{lstlisting}[float, language=R, style=customR, caption = {
-    An example of filtering ATLAS data on position covariates. 
-    The \texttt{filters} argument accepts a character vector with the logical statements. 
-    The function only retains data for which \textit{all} the conditions are satisfied; here that is positions calculated using $>$ 3 base stations (\texttt{NBS}), with location error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.}]
-filtered_data <- atl_filter_covariates(data = data,
-                    filters = c("NBS > 3",
-                                "SD < 100",
-                                "between(day, 5, 8)"))
-\end{lstlisting}
 

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
more listing edits

@@
 
-\begin{lstlisting}[float, language=R, style=customR, caption = {
-    Removing prolonged spikes from a movement track. 
-    The important function arguments here are \texttt{point\_angle\_cutoff} ($A$), \texttt{reflection\_speed\_cutoff} ($S$), and \texttt{est\_ref\_len}, the maximum number of positions after the inner bound that are candidates for the end of the prolonged spike, i.e., the outer bound.
-    If the prolonged spike ends after less than $N$ positions, the true end point is used as the outer bound of the spike.
-    However, the algorithm behind this function fails when the prolonged spike ends after more than $N$ positions. 
-    Users are advised to use a liberally large value of N in the \texttt{est\_ref\_len} argument; 1,000 may be appropriate for 3s interval data.
-    Further, users are cautioned against relying on such algorithms for severely distorted data.}]
-filtered_data <- atl_remove_reflections(data = track_data,
-                       x = "x", y = "y", time = "time",
-                       point_angle_cutoff = A,
-                       reflection_speed_cutoff = S,
-                       est_ref_len = N)
-\end{lstlisting}
+% \begin{lstlisting}[float, language=R, style=customR, caption = {
+%     Removing unrealistic movement in the form of prolonged spikes from a movement track. 
+%     The important function arguments here are \texttt{point\_angle\_cutoff} ($A$), \texttt{reflection\_speed\_cutoff} ($S$), and \texttt{est\_ref\_len}, the maximum number of positions after the inner bound that are candidates for the end of the prolonged spike, i.e., the outer bound.
+%     If the prolonged spike ends after less than $N$ positions, the true end point is used as the outer bound of the spike.
+%     However, the algorithm behind this function fails when the prolonged spike ends after more than $N$ positions. 
+%     Users are advised to use a liberally large value of N in the \texttt{est\_ref\_len} argument; 1,000 may be appropriate for 3s interval data.
+%     Further, users are cautioned against relying on such algorithms for severely distorted data.}]
+% filtered_data <- atl_remove_reflections(data = track_data,
+%                        x = "x", y = "y", time = "time",
+%                        point_angle_cutoff = A,
+%                        reflection_speed_cutoff = S,
+%                        est_ref_len = N)
+% \end{lstlisting}
 

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
listings

@@
 Users can then retain only those positions whose incoming and outgoing speeds are both $< S$ which or satisfy the condition $\theta < A$, where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
-The removal of spikes is implemented using the \texttt{atl\_filter\_covariates} function (Listing 4).
+The removal of spikes is implemented using the \texttt{atl\_filter\_covariates} function (Listing 3).
 Many other track metrics may be used to identify implausible movement and on which data may be filtered \citep{seidel2018}.
@@
 In the case of prolonged spikes, one relatively simple solution is identifying the bounds and removing positions between them.
-We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
-Users are strongly encouraged to visualise their data before and after applying this method, and we caution against relying on this method if data are heavily distorted by errors affecting entire track-subsets.
+% We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
+Users are strongly encouraged to visualise their data before and after applying such algorithms, and we caution against relying on these method if data are heavily distorted by errors affecting entire track-subsets.
 
@@
 % In general, large $K$s lead to unrealistic tracks, both because the first and last $(K - 1) / 2$ positions are handled as special cases, and because the local median approaches the global median at larger $K$.
-Median smoothing is provided by the \texttt{atlastools} function \texttt{atl\_median\_smooth}, with the only option being the moving window size, which must be an odd integer (Listing 6).
+Median smoothing is provided by the \texttt{atlastools} function \texttt{atl\_median\_smooth}, with the only option being the moving window size, which must be an odd integer (Listing 4).
 
@@
 If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
-Grouping variables' names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 7).
+Grouping variables' names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 5).
 
@@
 Users are encouraged to base these thresholds in the movement habits of their study species.
-For example, residence patch classification of red knot movement tracks considers consecutive stationary positions independent if they are 20m apart and considers consecutive bouts independent if the distance between them $\geq$ 100m, or the time difference between them $\geq$ 30 minutes (Listing 8).
+For example, residence patch classification of red knot movement tracks considers consecutive stationary positions independent if they are 20m apart and considers consecutive bouts independent if the distance between them $\geq$ 100m, or the time difference between them $\geq$ 30 minutes (Listing 6).
 The 20m distance represents a maximum speed of 6.667 m/s between positions, above which the individual is more likely in transit.

### 2021-02-05 docs/ms_atlas_preproc_main_text.tex
edit intro and filtering

@@
 Investigating fine-scale environmental and social drivers and the consequences of movement requires position data from many individuals at high temporal and spatial resolution.
-Such high-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}, yet ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015} routinely produce high-throughput tracking data at much lower costs.
-Although high-resolution tracking inherently provides more detailed information about the true path of the tracked animal, high-throughput data presents a two-fold challenge to ecologists.
+Such high-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}, yet ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput tracking data at much lower costs.
+Although high-resolution tracking provides a massive amount of data on the path of a tracked animal, this data presents a two-fold challenge to ecologists.
 First, the location error of each position may approach or exceed the true step size of the animal compared to low-resolution tracking with the same measurement error.
@@
 other tracked individuals).
-Users have two main options to improve data quality: making inferences after modelling the system-specific location error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
+Users have two main options to improve data quality: making inferences after modelling the system-specific location error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008,aspillaga2021}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
 The first approach may be limited by the animal movement models it can fit to the data \citep{fleming2014a, noonan2019, fleming2020}, and may be beyond the computational capacity of common hardware, leading users to prefer data cleaning instead.
@@
 Ecologists recognise that tracking data are imperfect observations of the underlying process of animal movement, yet they implicitly consider cleaned data equivalent to be the ground-truth.
-This assumption is reflected by popular statistical methods in movement ecology such as Hidden Markov Models (HMMs) \citep{langrock2012}, stationary-phase identification methods \citep{patin2020a}, or step-selection functions (SSFs) \citep{barnett2008, signer2017, avgar2016}, which expect low location errors relative to real animal movement (i.e., a high signal-to-noise ratio).
+This assumption is reflected in popular statistical methods in movement ecology such as Hidden Markov Models (HMMs) \citep{langrock2012}, stationary-phase identification methods \citep{patin2020a}, or step-selection functions (SSFs) \citep{barnett2008, signer2017, avgar2016}, which expect low location errors relative to real animal movement (i.e., a high signal-to-noise ratio).
 This makes the reproducible, standardised removal of location errors crucial to any animal tracking study.
 While gross errors are often removed by positioning-system algorithms, ‘reasonable’ errors often remain to confront end users \citep{fischler1981, weiser2016, ranacher2016}.
-Location errors may be exacerbated by the conditions of deployment, with systematic differences in location error among different landscape types  \citep[see][]{lewis2007, deon2002, frair2004}. 
-% For instance, the physical features of an animal’s habitat such as hills or woodland may impede the transmissions of radio-telemetry tags (Beardsworth et al.
+% Location errors may be exacerbated by the conditions of deployment, with systematic differences in location error among different landscape types  \citep[see][]{lewis2007, deon2002, frair2004}. 
+% % For instance, the physical features of an animal’s habitat such as hills or woodland may impede the transmissions of radio-telemetry tags (Beardsworth et al.
 \textit{in prep.}), while canopy cover can lead to location errors in GPS tracking \citep{lewis2007}.
-Standardised pre-processing steps should thus be general enough to tackle animal movement data recovered from a range of environments.
+Further, as high-throughput tracking is deployed in more regions, standardised pre-processing steps should be general enough to tackle animal movement data recovered from a range of environments.
 
@@
 Furthermore, filtering out positions on their location error estimates may not be sufficient to recover a good track estimate when location error does not correspond to biological realism in movement \citep{weiser2016, ranacher2016}.
-This makes identifying and removing implausible movements from a track an important component of recovering true animal movement \citep{bjorneraas2010}.
-Even after removing large location errors and evidently unrealistic movement, a track may comprise of positions that are distributed around the true animal location.
+This makes identifying and removing implausible locations from a track an important component of recovering true animal movement \citep{bjorneraas2010}.
+Even after removing outliers, a track may comprise of positions that are distributed around the true animal location.
 Discarding these positions as `untrue' would lead to excessive data loss, but treating them as `real' locations would lead to unrealistic estimates of metrics such as speed \citep{noonan2019}.
@@
 We begin by showing how the pipeline can clean simulated tracks to which we have artificially introduced location error \citep{gurarie2017}.
-Beginning with basic spatio-temporal filtering of positions, we cover filtering movement tracks, and reducing the effect of location error with a median smooth.
+Beginning with basic spatio-temporal filtering of positions, we cover filtering outliers from movement tracks, and reducing the effect of location error with a median smooth.
 Then, we suggest one solution to issues of computational tractability, which is to synthesise residence patches \citep[\textit{sensu}][]{bijleveld2016, oudman2018, barraquand2008} from clusters of spatio-temporally proximate positions, and so reveal important aspects of animal space use.
@@
     The \texttt{atlastools} function that may be used to implement each pre-processing step is shown in the grey boxes underneath each step.
-    Popular \texttt{R} packages in movement ecology, or statistical methods are shown underneath suggested analyses (yellow boxes).
-    Users are strongly encouraged to visualise their data and scan it for location errors as a first step, and as they work through the pipeline.
+    Popular \texttt{R} packages in movement ecology, or statistical methods are shown underneath possible analyses (yellow boxes).
+    Users are strongly encouraged to visualise their data and scan it for location errors as they work through the pipeline, always asking the question, could the animal plausibly move this way?
     }
@@
     At this stage, the data are ready for a number of popular statistical treatments such as Hidden Markov Model-based classification \citep{michelot2016,langrock2012}.
-    \item Finally, users wishing to study prolonged animal residence in an area can classify their data into residence patches based on the movement ecology of their study system, after filtering out non-stationary positions (see \textsc{Synthesising Movement Tracks into Residence Patches}).
+    \item Finally, users wishing simple segmentation-clustering can classify their data into residence patches based on the movement ecology of their study system, after filtering out non-stationary positions (see \textsc{Synthesising Movement Tracks into Residence Patches}).
 
@@
 
-First, data can be filtered to exclude positions outside the spatial bounds of a study area.
-This simply involves comparing position coordinates with the range of acceptable coordinates (the bounding box), and removing those positions outside them (Fig. 2.b; Listing 1). 
-A bounding box filter retains data from within a subset of the tracking range without the need for a geospatial representation such as a shapefile.
-This can help remove unreliable data from a tracking system that is less accurate beyond a certain range (e.g. ATLAS; Beardsworth et al. \textit{in prep.}).
-In some special cases, users may wish to \textit{remove} positions inside a bounding box, either because movement behaviour within an area is not the focus of a study, or because positions recorded within an area are known to be erroneous.
+First, users should exclude positions outside the spatial bounds of a study area by comparing position coordinates with the range of acceptable coordinates (the bounding box), and removing positions outside them (Fig.
+2.b; Listing 1).
+A bounding box filter does not require a geospatial representation such as a shapefile, and can help remove unreliable data from a tracking system that is less accurate beyond a certain range (e.g.
+ATLAS; Beardsworth et al.
+\textit{in prep.}).
+In some special cases, users may wish to remove positions \textit{inside} a bounding box, either because movement behaviour within an area is not the focus of a study, or because positions recorded within an area are known to be erroneous.
 An example of the former is studies seeking to study transit behaviour between features which can be approximated by their bounding boxes. 
 Instances of the latter are likely to be system specific, but are known from ATLAS systems (Bijleveld et al. \textit{in prep.}). 
-One drawback to bounding box filters is that they are restricted to rectangular areas.
-Users seeking to filter for areas with other geometries, such as a circular or irregular study area, need a geometric intersection between their data and a spatial representation of the area of interest (e.g.
+Bounding boxes are by definition rectangular, and users seeking to filter for other geometries, such as a circular or irregularly-shaped study area, need a geometric intersection between their data and a spatial representation of the area of interest (e.g.
 shapefile, geopackage, or \texttt{sf}-object in \texttt{R}).
-The \texttt{atlastools} function \texttt{atl\_filter\_bounds} implements bounding box and explicit spatial filters and accepts X and Y coordinate ranges, an \texttt{sf}-polygon object \citep{pebesma2018}, or any combination of the three to filter the data (Listing 1).
-Multipolygon objects are supported, allowing data from different areas to be selected.
-When both coordinate ranges and a polygon are provided, the data is first filtered by the ranges and then the polygon.
-The boolean function argument \texttt{remove\_inside} determines whether positions inside the bounds are retained (\texttt{FALSE}) or removed (\texttt{TRUE}).
+The \texttt{atlastools} function \texttt{atl\_filter\_bounds} implements bounding box and explicit spatial filters and accepts X and Y coordinate ranges, an \texttt{sf}-polygon or multi-polygon object \citep{pebesma2018}, or any combination of the three to filter the data (Listing 1).
+% When both coordinate ranges and a polygon are provided, the data is first filtered by the ranges and then the polygon.
+% The boolean function argument \texttt{remove\_inside} determines whether positions inside the bounds are retained (\texttt{FALSE}) or removed (\texttt{TRUE}).
 

### 2021-02-05 docs/ms_atlas_preproc_main_text.tex
filter covariates

@@
     \textbf{(b)} 
-    Location error may affect single observations resulting in point outliers or `spikes' (grey crosses), or continuous subsets of a track, called a `prolonged spike' (black crosses, top right), and both represent unrealistic movement.
+    Location error may affect single observations resulting in point outliers or `spikes' (grey crosses), or continuous subsets of a track, called a `prolonged spike' (dark crosses, top right), and both represent unrealistic movement.
     While spikes can be removed by filtering out positions with high incoming and outgoing speeds and turning angle, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
@@
 
-Temporal filtering can exclude positions from periods during tracking when data are expected to be unreliable, either due to abnormal movement behaviour or poor tracking quality. 
-An animal’s movement may be non-representative when it has been fitted with a tracker but has not been released, leading to artificial stationary positions, or in the time shortly after release when its movement may be influenced by the stress of capture and handling (see example implementation in \texttt{amt}; \citet{signer2019}).
-In these cases, all positions from a short time after release (e.g. 24 hours) can be excluded. 
+An animal’s movement may be non-representative when it has been fitted with a tracker but has not been released, leading to artificial stationary positions, or in the time shortly after release when its movement may be influenced by the stress of capture and handling.
 Periods of poor tracking quality may result from system malfunctions, and temporal filters can be applied to exclude such chunks from the datastream.
+Temporal filtering can exclude positions from such periods when data are expected to be unreliable, either due to abnormal movement behaviour or system-specific issues.  
 Temporal filters can be combined with spatial filters to exclude time-location combinations which would be unrealistic for the tracked animal. 
@@
 
-The atlastools function \texttt{atl\_filter\_covariates} allows convenient filtering of a dataset by any number of logical statements (Listing 2).
-This function can be used to easily filter timestamps in a range, as well as combine simple spatial and temporal filters.
-It accepts a character vector of \texttt{R} expressions that each return a logical vector (i.e., \texttt{TRUE} or \texttt{FALSE}; Listing 2).
-The function returns only those data which satisfy each of the filter conditions. 
-Users must ensure that the filtering variables exist in their dataset in order to avoid errors.
+The atlastools function \texttt{atl\_filter\_covariates} allows convenient filtering of a dataset by any number of logical statements, including querying data within a spatio-temporal range (Listing 2).
+% This function can be used to easily filter timestamps in a range, as well as combine simple spatial and temporal filters.
+% It accepts a character vector of \texttt{R} expressions that each return a logical vector (i.e., \texttt{TRUE} or \texttt{FALSE}; Listing 2).
+The function returns only those data which satisfy each of the filter conditions, and users must ensure that the filtering variables exist in their dataset in order to avoid errors.
 
@@
 
-Tracking data covariates can be good indicators of the reliability of calculated positions (Beardsworth et al. \textit{in prep.}).
+Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system (Beardsworth et al. \textit{in prep.}).
 For intance, in reverse-GPS ATLAS systems the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
-Reverse-GPS systems also calculate direct measures of location error during localisation \citep{maccurdy2009, maccurdy2019, weiser2016}.
-GPS systems' location error may be similarly indicated indirectly by the number of satellites involved in the localisation, or directly by an error measure such as the Horizontal Dilution of Precision (HDOP).
-ATLAS and other TOA systems 
+GPS systems' location error may be similarly indicated indirectly by the number of satellites involved in the localisation., 
+Both GPS and reverse-GPS systems may also provide direct measures of location error during localisation \citep{maccurdy2009, maccurdy2019, weiser2016}.
 % A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables

### 2021-02-05 docs/ms_atlas_preproc_main_text.tex
some edits

@@
 
-An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 3.a).
+An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 2.a, b).
 Point outliers are characterised by artificially high speeds between the outlier and the positions before and after \citep[called incoming and outgoing speed, respectively][]{bjorneraas2010}, lending a `spiky' appearance to the track.
@@
 Users can then retain only those positions whose incoming and outgoing speeds are both $< S$ which or satisfy the condition $\theta < A$, where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
-The removal of spikes is implemented using the \texttt{atl\_filter\_covariates} function (Listing 3).
+Spike removal can be implemented using the \texttt{atl\_filter\_covariates} function (Listing 3).
 Many other track metrics may be used to identify implausible movement and on which data may be filtered \citep{seidel2018}.
@@
 Sometimes entire susbets of the track may be affected by the same large-scale location error.
-For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 3.a, b).
+For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 2.a, b).
 These cannot be corrected by targeted removal of individual positions, as in Bjorneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
@@
 % We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
-Users are strongly encouraged to visualise their data before and after applying such algorithms, and we caution against relying on these method if data are heavily distorted by errors affecting entire track-subsets.
+Users are strongly encouraged to visualise their data before and after applying such algorithms, 
+We caution that these methods are not foolproof, and data that are heavily distorted by errors affecting entire track-subsets should be used with care when making further inferences.
 
@@
 Most data at this stage is technically ‘clean’, yet its volume alone may pose challenges for lower-specification or older hardware and software if these are not optimised for efficient computation.
-Thinning data need not compromise researchers' ability to answer scientific questions with them; for instance, social interactions lasting 1 -- 2 minutes would still be detected on thinning from a sampling interval of 1 second to 1 minute.
+Thinning data need not compromise researchers' ability to answer scientific questions with them; for instance, proximity-based social interactions lasting 1 -- 2 minutes would still be detected on thinning from a sampling interval of 1 second to 1 minute.
 Indeed, temporal auto-correlation may hinder some methods such as the estimation of home-ranges or step-selection functions \citep{fleming2014a, dupke2017}.
@@
 \end{linenomath*}
-% Similarly, the overall location error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance ($SD ^ 2$):
-% \begin{linenomath*}
-%     \begin{equation*}
-%         SD_{agg} or HDOP_{agg} = \sqrt{ \left( \sum_{i=1}^{i=N} SD_i^2 \right) / N ^ 2  }
-%     \end{equation*}    
-% \end{linenomath*}
+Similarly, the overall location error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance:
+\begin{linenomath*}
+    \begin{equation*}
+        SD_{agg} \ or \ HDOP_{agg} = \sqrt{ \left( \sum_{i=1}^{i=N} SD_i^2 \ or \ HDOP_i^2 \right) / N ^ 2  }
+    \end{equation*}    
+\end{linenomath*}
+We caution users that thinning causes an extensive loss of small-scale detail in the data, and should be used carefully.
 
-Users may question why thinning should be implemented after cleaning steps, when aggregation can obtain consensus positions over an interval, correctly propagate the location error, and also reduce data volumes.
+Users may question why aggregation, which can obtain consensus positions over an interval, correctly propagate the location error, and also reduce data volumes, should not be used directly on the raw data.
 %%
-We caution users that thinning causes an extensive loss of small-scale detail in the data.
-Additionally, thinning prior to excluding unrealistic movement and smoothing can lead to estimates of essential metrics --- such as speed --- that are substantially different from the true value (Noonan et al.
+We caution that hinning prior to excluding unrealistic movement and smoothing can lead to estimates of essential metrics --- such as speed --- that are substantially different from the true value (Noonan et al.
 2019; see Fig.
@@
 However, thinning before data-cleaning may have its place as a useful step before exploratory visualisation of the movement track, since reduced data volumes are easier to handle for plotting software.
-
+%
 Thinning is implemented in \texttt{atlastools} using the \texttt{atl\_thin\_data} function, with either aggregation or resampling (specified by the \texttt{method} argument) over an interval using the \texttt{interval} argument.
-The column of timestamps must be named `time' and column classes except the identity columns (\textit{see below}) should allow averaging.
-The `resample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
-Using `resample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
-In both cases, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
-Due to its development for ATLAS systems, the `aggregate' assumes the variance around coordinates is named \texttt{VARX} and \texttt{VARY}, and standard deviation around each position is named \texttt{SD}.
-These columns must be present together for the function to correctly handle the error.
-If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
+% The column of timestamps must be named `time' and column classes except the identity columns (\textit{see below}) should allow averaging.
+% The `resample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
+% Using `resample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
+% In both aggregation and resampling, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
+% Due to its development for ATLAS systems, the `aggregate' assumes the variance around coordinates is named \texttt{VARX} and \texttt{VARY}, and standard deviation around each position is named \texttt{SD}.
+% These columns must be present together for the function to correctly handle the error.
+% If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
 Grouping variables' names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 5).

### 2021-02-08 docs/ms_atlas_preproc_main_text.tex
some edits to text and fix fig refereces

@@
 \usepackage{helvet}
+\usepackage{xfrac}
 \usepackage{amsmath}
@@
 
-    \item A pipeline for cleaning massive high-throughput datasets must similarly balance ease of use, computational efficiency, and the ability to prepare movement data for a range of analyses.
-    At its core, cleaning high-throughput data must strike a balance between rejecting location errors without discarding valid animal movements.
-    Additionally, a pre-processing pipeline must be capable of efficiently segmenting and clustering location data for statistical methods, while also being scalable to large datasets and robust to imperfect sampling.
+    \item A pipeline for cleaning massive high-throughput datasets must balance ease of use, computational efficiency, and the ability to prepare movement data for a range of analyses.
+    Data cleaning must also reject location errors without discarding valid animal movements.
+    Another useful feature of a pre-processing pipeline is efficiently segmenting and clustering location data for statistical methods, while also being scalable to large datasets and robust to imperfect sampling.
     Manual methods being prohibitively time consuming, and to boost reproducibility, a robust pre-processing pipeline must be automated.
@@
 Tracking individual animals is the methodological mainstay of movement ecology, which seeks to link animal movement with internal causes, environmental factors, and resulting consequences \citep{nathan2008a, holyoak2008}.
-Investigating fine-scale environmental and social drivers and the consequences of movement requires position data from many individuals at high temporal and spatial resolution.
-Such high-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}, yet ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput tracking data at much lower costs.
+Investigating fine-scale environmental and social drivers and the consequences of movement requires position data from many individuals at high temporal and spatial resolution relative to the phenomenon of interest (high-throughput tracking).
+High-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}, yet ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput tracking data at much lower costs.
 Although high-resolution tracking provides a massive amount of data on the path of a tracked animal, this data presents a two-fold challenge to ecologists.
@@
 other tracked individuals).
-Users have two main options to improve data quality: making inferences after modelling the system-specific location error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008,aspillaga2021}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
+Users have two main options to improve data quality: making inferences after modelling the system-specific location error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008, aspillaga2021}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
 The first approach may be limited by the animal movement models it can fit to the data \citep{fleming2014a, noonan2019, fleming2020}, and may be beyond the computational capacity of common hardware, leading users to prefer data cleaning instead.
@@
 
-Pre-processing steps must justifiably discard large location errors, also called outliers, from tracks (analogous to reducing false positives) while avoiding the overzealous rejection of valid animal movements (analogous to reducing false negatives).
+Pre-processing steps must reliably discard large location errors, also called outliers, from tracks (analogous to reducing false positives) while avoiding the overzealous rejection of valid animal movements (analogous to reducing false negatives).
 How well researchers balance these imperatives has consequences for downstream analyses \citep{stine2001}.
@@
 Despite the importance and ubiquity of reducing location errors in tracking data, movement ecologists lack formal guidance on this crucial step.
-Pre-processing protocols are often system-specific, or have limited availability, or may not be easily tractable for mainstream computing hardware and software.
+Pre-processing protocols are not often reported in the literature, or may not be easily tractable for mainstream computing hardware and software.
 Furthermore, filtering out positions on their location error estimates may not be sufficient to recover a good track estimate when location error does not correspond to biological realism in movement \citep{weiser2016, ranacher2016}.
@@
 With these in mind we have developed the \texttt{R} package \texttt{atlastools} \citep{gupte2020a, rcoreteam2020}, which implements important steps of the pre-processing pipeline.
-\texttt{R} is the computational environment of choice in movement ecology \citep{joo2020} and formalising tools as an \texttt{R} package improves portability and reproducibility \citep{marwick2018}.
+\texttt{R} is the computational environment of choice in movement ecology \citep{joo2020} and formalising tools as an \texttt{R} package improves portability and reproducibility \citep{marwick2018, }.
 \texttt{atlastools} functions are easy to use and understand, and rely on the efficiency of the \texttt{R} package \texttt{data.table} \citep{dowle2020}.
@@
     \caption{A general, modular pipeline for pre-processing high-throughput tracking data from raw localisations to cleaned data, and optionally into residence patches. 
-    Users should determine the appropriate pre-processing module, and work their way through the corresponding pre-processing steps, until the data are suitable for their intended purpose.
+    Users should apply the appropriate pre-processing modules and the steps therein until the data are suitable for their intended analysis, some of which are suggested here.
     The \texttt{atlastools} function that may be used to implement each pre-processing step is shown in the grey boxes underneath each step.
@@
 
-    Package functions are prefixed `\texttt{atl\_}', and are based on the \texttt{data.table} package (Dowle and Srinivasan 2020).
-    Certain functions modify the input data "in place", and users must work on a copy of their data to preserve both the original and cleaned data (details in the Supplementary Material).
+    % Package functions are prefixed `\texttt{atl\_}', and are based on the \texttt{data.table} package (Dowle and Srinivasan 2020).
+    % Certain functions modify the input data "in place", and users must work on a copy of their data to preserve both the original and cleaned data (details in the Supplementary Material).
     Since high-throughput tracking systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} users must convert their geographic coordinates to metres as this is assumed by all functions.
@@
     %  (see \textsc{Spatio-Temporal Filtering}).
-    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by a system-specific error measure, or by the plausibility of associated movement metrics.
+    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by a system-specific error measure, or by the plausibility of associated movement metrics, such as speed and turning-angle \citep{seidel2018, calenge2009}.
     %  (see \textsc{Filtering to Reduce Location Errors}).
@@
 2017), and connected them together with a roughly linear path (see Fig.
-6.a).
+5.a).
 RACVM models approximate the tracks of soaring birds which circle on thermals over a relatively small area, and move between thermals \citep[`thermalling'; ][]{gurarie2017, harel2016}.
@@
 First, users should exclude positions outside the spatial bounds of a study area by comparing position coordinates with the range of acceptable coordinates (the bounding box), and removing positions outside them (Fig.
-2.b; Listing 1).
+2.a; Listing 1).
 A bounding box filter does not require a geospatial representation such as a shapefile, and can help remove unreliable data from a tracking system that is less accurate beyond a certain range (e.g.
@@
 \begin{lstlisting}[float,floatplacement=h!,language=R, style=customR, caption = {
-    The \texttt{atl\_filter\_bounds} function removes positions outside an area defined by coordinate ranges, a polygon, or all three (\texttt{remove\_inside = FALSE}), or positions inside the area (\texttt{remove\_inside = TRUE}).
-    The arguments \texttt{x} and \texttt{y} determine the X and Y coordinate columns, \texttt{x\_range} and \texttt{y\_range} are the filter bounds in a CRS in metres, and the data can be filtered by an \texttt{sf-(MULTI)POLYGON} can be passed using the \texttt{sf\_polygon} argument.
-    \texttt{atl\_filter\_bounds} returns a filtered \texttt{data.table}, which must be saved as an object (here, \texttt{filtered\_data}).}]
+    The \texttt{atl\_filter\_bounds} function filters on an area defined by coordinate ranges, a polygon, or all three; it can remove positions outside (\texttt{remove\_inside = FALSE}), or within the area (\texttt{remove\_inside = TRUE}).
+    The arguments \texttt{x} and \texttt{y} determine the X and Y coordinate columns, \texttt{x\_range} and \texttt{y\_range} are the filter bounds in a coordinate reference system in metres, and the data can be filtered by an \texttt{sf-(MULTI)POLYGON} can be passed using the \texttt{sf\_polygon} argument.
+    The output is a \texttt{data.table}, which must be saved as an object (here, \texttt{filtered\_data}).}]
 filtered_data <- atl_filter_bounds(data = data,
@@
     While spikes can be removed by filtering out positions with high incoming and outgoing speeds and turning angle, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
-    These and other distortions that affect continuous track segments can be removed by conceptualising algorithms that find the bounds of the distortion and remove points in between them.
+    These and other distortions that affect continuous track segments should be removed by conceptualising algorithms that find the bounds of the distortion and remove points in between them.
     Users should frequently check the outputs of such algorithms to avoid rejecting valid data.
@@
 An animal’s movement may be non-representative when it has been fitted with a tracker but has not been released, leading to artificial stationary positions, or in the time shortly after release when its movement may be influenced by the stress of capture and handling.
-Periods of poor tracking quality may result from system malfunctions, and temporal filters can be applied to exclude such chunks from the datastream.
-Temporal filtering can exclude positions from such periods when data are expected to be unreliable, either due to abnormal movement behaviour or system-specific issues.  
+Periods of poor tracking quality may result from system malfunctions, and users may also wish to exclude these data.
+Temporal filtering can exclude positions from such periods when data are expected to be unreliable for ecological inference, either due to abnormal movement behaviour or system-specific issues.  
 Temporal filters can be combined with spatial filters to exclude time-location combinations which would be unrealistic for the tracked animal. 
@@
 \begin{lstlisting}[float, language=R, style=customR, caption = {
-    Data can be filtered by a temporal or a spatio-temporal range using {atl\_filter\_covariates}. 
-    Filter predicates are passed to the \texttt{filters} argument as a character vector. 
-    Only rows in the data satisfying all the conditions passed as filters are retained. 
-    Here, the first example shows how nighttime data can be retained using a predicate (\texttt{inrange} from \texttt{data.table}) that determine whether the value of `hour' is between 6 and 18, and also within a range of X coordinates.
-    The second example retains calculated using $>$ 3 base stations (\texttt{NBS}), with location error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.
+    Data can be filtered by a temporal or a spatio-temporal range using \texttt{atl\_filter\_covariates}. 
+    Filter conditions are passed to the \texttt{filters} argument as a character vector. 
+    Only rows in the data satisfying \textit{all} the conditions are retained. 
+    Here, the first example shows how nighttime data can be retained using a predicate that determines whether the value of `hour' is between 6 and 18, and also within a range of X coordinates.
+    The second example retains ATLAS locations calculated using $>$ 3 base stations (\texttt{NBS}), with location error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.
     }]
@@
 Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system (Beardsworth et al. \textit{in prep.}).
-For intance, in reverse-GPS ATLAS systems the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
-GPS systems' location error may be similarly indicated indirectly by the number of satellites involved in the localisation., 
+For instance, in reverse-GPS ATLAS systems the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
+GPS systems' location error may be similarly indicated indirectly by the number of satellites involved in the localisation.
 Both GPS and reverse-GPS systems may also provide direct measures of location error during localisation \citep{maccurdy2009, maccurdy2019, weiser2016}.
@@
 % \end{linenomath*}
-Unreliable positions can be removed by filtering on direct or indirect measures of quality using \texttt{atl\_filter\_covariates}.
+Unreliable positions can be removed by filtering on direct or indirect measures of quality using \texttt{atl\_filter\_covariates} (Listing 2).
 
@@
 Filtering on system-generated measures of error may not result in the removal of all erroneous positions, and data may remain which would require biologically implausible movement.
-Users are encouraged to visualise their tracks before and after filtering point locations, and especially to `join the dots' and connect consecutive positions with lines.
+Users are encouraged to visualise their tracks before and after filtering point locations, and especially to `join the dots' and connect consecutive positions with lines (Figure 2.b).
 Whether the resulting track looks realistic is ultimately a subjective human judgement, but one which implicitly integrates prior knowledge of the movement ecology of the study species to ask, `Does the animal move this way?'.
@@
 Since it is both difficult and prohibitively time consuming to exactly reproduce expert judgement when dealing with large volumes of tracking data from multiple individuals, some automation is necessary.
-Users should first manually examine a representative subset of tracks and attempt to visually identify problems --- either with individual positions, or with subsets of the track --- that persist after basic filtering.
+Users should first manually examine a representative subset of tracks and attempt to visually identify problems --- either with individual positions, or with subsets of the track --- that persist after filtering on quality covariates.
 Once such problems are identified, users can conceptualise algorithms that can be applied to their data to resolve them.
 
-An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 2.a, b).
+An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 2.b).
 Point outliers are characterised by artificially high speeds between the outlier and the positions before and after \citep[called incoming and outgoing speed, respectively][]{bjorneraas2010}, lending a `spiky' appearance to the track.
@@
 Here, it is important to remember that speed estimates are scale-dependent; high-throughput tracking typically overestimates the speed between positions where the animal is stationary or moving slowly due to small-scale location errors \citep{ranacher2016, noonan2019}.
-Even after data with large location errors have been removed by filters, it is advisable to begin with a liberal (high) speed threshold that excludes only the most unlikely of speeds.
+Even after data with large location errors have been removed, it is advisable to begin with a liberal (high) speed threshold that excludes only the most unlikely of speeds.
 Estimates of maximum speed may not always be readily obtained for all species, and an alternative is to use a data-driven threshold such as the 95\textsuperscript{th} percentile of speeds from the track.
 Once a speed threshold $S$ has been chosen, positions with incoming \textit{and} outgoing speeds $\geq S$ may be identified as spikes and removed.
-Some species can realistically achieve speeds $\geq S$ in fast transit segments when assisted by environmental factors, such as birds with tailwinds, and a simple filter on incoming and outgoing speeds would exclude this valid data.
-To avoid removing real, fast transit segments while still excluding spikes we suggest combining the speed filter with a filter on the turning angles of each position \citep[see][]{calenge2009}.
+Some species can realistically achieve speeds $\geq S$ in fast transit segments when assisted by their environment, such as birds with tailwinds, and a simple filter on incoming and outgoing speeds would exclude this valid data.
+To avoid removing valid, fast transit segments while still excluding spikes we suggest combining the speed filter with a filter on the turning angles of each position \citep[see][]{calenge2009}.
 This combined filter assumes that positions in high-throughput tracking with both high speeds and large turning angles are likely to be due to location errors, since most species are unable to turn sharply at high speed.
@@
 
-Sometimes entire susbets of the track may be affected by the same large-scale location error.
-For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 2.a, b).
+Sometimes entire subsets of the track may be affected by the same large-scale location error.
+For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 2.b).
 These cannot be corrected by targeted removal of individual positions, as in Bjorneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
@@
 \end{linenomath*}
-Users can apply a median smooth with an appropriate $K$ independently to the X and Y coordinates of a movement track to smooth it (see Fig. 4.a -- e). 
+Users can apply a median smooth with an appropriate $K$ independently to the X and Y coordinates of a movement track to smooth it (see Fig. 3.a -- e). 
 % The first and last $(K - 1)/2$ values for a moving window size of $K$ are assigned a value using Tukey's (1977) end-point smoothing rule as implemented in \texttt{R} \citep{hardle1995}.
 The median smooth is robust to even very large temporal and spatial gaps, and does not interpolate between positions when data are missing. 
-Thus it is not necessary to split the data into segments separated by periods of missing observations when applying the filter (see Fig. 4).
+Thus it is not necessary to split the data into segments separated by periods of missing observations when applying the filter (see Fig. 3).
 
@@
 Furthermore, while larger $K$ may result in smoother tracks (Fig.
-4.b -- e), one drawback of using a large $K$ is that short, quick forays away from the main track are likely to be smoothed away, leading to a loss in detail of the individual's small-scale movement.
+3.b -- e), one drawback of using a large $K$ is that short, quick forays away from the main track are likely to be smoothed away, leading to a loss in detail of the individual's small-scale movement.
 Users must themselves judge how best to trade large-scale and small-scale accuracy, and choose $K$ accordingly.
@@
     The canonical track is shown `zoomed in' for better representation, and the missing data at the top of the track (line with no points) is from where the prolonged spike was earlier removed.
-    Median smoothing the given position coordinates ((a) green points) over a moving window ($K$) of \textbf{(b)} 3, \textbf{(c)} 5, \textbf{(d)} 11, and \textbf{(e)} 21 positions respectively, yields approximations (purple lines) of the canonical track.
+    Median smoothing the given position coordinates (green points, in \textbf{(a)}) over a moving window ($K$) of \textbf{(b)} 3, \textbf{(c)} 5, \textbf{(d)} 11, and \textbf{(e)} 21 positions respectively, yields approximations (purple lines) of the canonical track.
     Users are cautioned that there is no correct $K$, and they must subjectively choose a $K$ which most usefully trades small-scale details of the track for large-scale accuracy.}
@@
 Aggregation involves computing the mean values of all covariates for positions within a time-interval group.
-Both approaches yield one position per time-interval group.
+Both approaches yield one position per time-interval group (Fig. 4.a).
 Categorical variables, such as the habitat type associated with each position can be aggregated using a suitable measure such as the mode.
@@
 %%
-We caution that hinning prior to excluding unrealistic movement and smoothing can lead to estimates of essential metrics --- such as speed --- that are substantially different from the true value (Noonan et al.
+We caution that thinning prior to excluding unrealistic movement and smoothing (Fig 4.b) can lead to preserving artefacts in the data, and estimates of essential metrics --- such as speed --- that are substantially different from the true value (Noonan et al.
 2019; see Fig.
-5.c).
+4.c).
+In our example for instance, the data with errors included would have to be thinned to \sfrac{1}{30}\textsuperscript{th} of its volume for the median speed to be correctly estimated; an undesirable step if the aim is fine-scale tracking (Fig.
+4.c).
 The mis-estimation of track metrics could have knock-on consequences for the implementation of subsequent filters based on detecting unrealistic movement.
@@
 The time interval is specified in seconds, while the \texttt{id\_columns} allows a character vector of column names to be passed to the function, with these columns used as identity variables.
-Both methods return a dataset with as many rows as there are time-intervals.
-While the resampling method retains all columns, the aggregation method drops the ATLAS specific columns specifying covariance in X and Y coordinates (COVXY), and location error (SD).}]
+Both methods return a dataset with one rows per time-interval.}]
 thinned_data <- atl_thin_data(data,
@@
     For comparison, the median and 95\textsuperscript{th} percentile of speed of the canonical track are shown as solid and dashed horizontal lines, respectively.
-    Green circles and associated error bars show, in contrast, the median ($\pm$ standard deviation) speed for the same aggregation interval, but after removing large- and small-scale location errors (median smooth $K$ = 11).}
+    Green circles and associated error bars show, in contrast, the median ($\pm$ standard deviation) speed for the same aggregation interval, but after removing large- and small-scale location errors (median smooth $K$ = 11).
+    The unfiltered data would have to be thinned to \sfrac{1}{30}\textsuperscript{th} volume to correctly estimate median speed.}
     \label{fig:figure_thinning}
@@
 
-\section{Synthesising Movement Tracks into Residence Patches}
+\section{Segmenting and Clustering Movement Tracks into Residence Patches}
 
@@
 
-Tracking data that are still too large for statistical packages, or have strong autocorrelation, can benefit from segmentation-clustering into `residence patches' \citep{bijleveld2016, oudman2018, barraquand2008}.
+An important and common analysis with animal tracking data is to link space-use with environmental covariates.
+This is difficult even with smoothed and thinned high-throughput data, as these may be too large for statistical packages, or have strong autocorrelation (Fig. 5).
+Users aiming for such analyses can benefit from segmentation-clustering the data into spatio-temporally independent `residence patches', and link the attributes of movement segments (such as duration or distance), with environmental data \citep{bijleveld2016, oudman2018, barraquand2008}.
 Making the patch the unit of observation conveniently sidesteps pseudo-replication while and reduces computational requirements.
@@
 
-A cleaned movement track can be classified into residence patches using the function \texttt{atl\_res\_patch} (see Fig. 6.c).
+A cleaned movement track can be classified into residence patches using the function \texttt{atl\_res\_patch} (see Fig. 5.b).
 \texttt{atl\_res\_patch} requires three parameters: (1) the distance threshold between positions (called \texttt{buffer\_size}), (2) the distance threshold between clusters of positions (called \texttt{lim\_spat\_indep}), and (3) the time interval between clusters (called \texttt{lim\_time\_indep}).
@@
 % The exclusion of clusters with few positions can help in removing bias due to short stops, but if such short stops are also of interest, they can be included by reducing the \texttt{min\_fixes} argument.
-Our residence patch algorithm is capable of correctly identifying clusters of related residence points from a movement track (Fig. 7.a, 7.b).
-This includes clusters where the animal is relatively stationary (orange and green patches, Fig. 7.c), as well as clusters where the animal is moving slowly (blue patch, Fig. 7.c).
+Our residence patch algorithm is capable of correctly identifying clusters of related points from a movement track (Fig. 5).
+This includes clusters where the animal is relatively stationary (orange and green patches, Fig. 5.b), as well as clusters where the animal is moving slowly (blue patch, Fig. 5.b).
 This flexibility is especially useful when studying movements that may represent two different modes of the same behaviour, for instance, area-restricted search, as well as slow, searching behaviour with a directional component.
@@
     \includegraphics[width=0.75\textwidth]{figures/fig_05_residence.png}
-    \caption{Movement tracks can be classified into residence patches, whcih are areas of prolonged residence (clusters of points), while leaving out the transit between them.
-    \textbf{(a)} The residence patch method correctly identifies clusters of positions where the individual is relatively stationary (orange and green patches), as well as positions where it is slowly moving (blue patch).
+    \caption{Movement tracks can be classified into residence patches, which are areas of prolonged residence (clusters of points), while leaving out the transit between them.
+    \textbf{(a)} The residence patch method correctly identifies clusters of positions where the individual is relatively stationary (orange and green patches), as well as positions where it is moving slowly (blue patch).
     This is especially useful when studying more complex behaviour such as area-restricted search, which may have a directional component.
@@
 We applied the pre-processing pipeline using \texttt{atlastools} functions described above to a calibration dataset to verify that the residence patch method could correctly identify known stopping points (see Fig.
-7).
+6).
 We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 (Beardsworth et al.
@@
 We clustered these positions into residence patches with a buffer radius of 5m, spatial independence limit of 50m, temporal independence limit of 5 minutes, and a minimum of 3 positions per patch.
-Inferred residence patches corresponded well to the locations of stops (see Fig. 7.c).
+Inferred residence patches corresponded well to the locations of stops (see Fig. 6.c).
 However, the residence patch algorithm detected more stops than were logged as waypoints (n = 28, n waypoints = 21).
-One of these was the field station on Griend where the tag was stored between trips (red crossed-square, Fig. 7.c).
+One of these was the field station on Griend where the tag was stored between trips (red crossed-square, Fig. 6.c).
 The method also did not detect two stops of 105 and 563 seconds (1.75 and 9.4 minutes) since they were data poor and aggregated away in the thinning step (n positions = 6, 15).
@@
 We extracted summary data and spatial polygons from the constructed residence patches.
-Plotting the bats' residence patches and the linear paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site (Fig. 9.a).
-Bats tended to show prolonged residence near known food sources (fruit trees), travelling repeatedly between previously visited areas (Fig. 9.b).
+Plotting the bats' residence patches and the linear paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site (Fig. 7.a).
+Bats tended to show prolonged residence near known food sources (fruit trees), travelling repeatedly between previously visited areas (Fig. 7.b).
 However, bats also appeared to spend some time at locations where no fruit trees were recorded, prompting questions about their use of other food sources, or another behaviour entirely (Fig.
-9.b, 9.c).
-Bats occurring close together did not have strongly overlapping residence patches, and their paths to and from area of co-occurrence were different (Fig. 9.a, 9.c).
+7.b, 7.c).
+Bats occurring close together did not have strongly overlapping residence patches, and their paths to and from area of co-occurrence were different (Fig. 7.a, 7.c).
 Constructing residence patches for multiple individuals over multiple nights suggests interesting dynamics of within- and between-individual overlap.
@@
 The open-source \texttt{R} package \texttt{atlastools} serves as a starting point for methodological collaboration among movement ecologists.
-Efficient location error modelling approaches \citep{fleming2020} may eventually make data-cleaning optional.
+Efficient location error modelling approaches \citep{fleming2020,aspillaga2021} may eventually make data-cleaning optional.
 Yet cleaning tracking data even partially before modelling location error will be faster than error-modelling on the full data, and the removal of large location errors may improve model fits.

### 2021-02-08 _02_filtering_data.Rmd
remove extra figure
remove figure cap

@@
 ```
-
-![](figures/fig_raw_bounds.png)
-
-**(a)** A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the `smoove` package (Gurarie et al.
-2017).
-Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track.
-The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (solid black line) as closely as possible.
-**(b)** Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the `atlastools` function  `atl_filter_bounds`.
-Setting the function argument `remove_inside = FALSE` retains positions within user supplied bounds (blue points), and excludes those outside (red points).
 

### 2021-02-08 05_calibration_data.Rmd
spelling

@@
 We began by visualising the data to check for location errors, and found a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2; main text Fig. 7.a).
-This outlier was removed by filtering data by the X coordinate bounds using the function `atl_filter\_bounds`; X coordinate bounds $\leq$ 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
+This outlier was removed by filtering data by the X coordinate bounds using the function `atl_filter_bounds`; X coordinate bounds $\leq$ 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
 1.2).
@@
 
+Plotting code is not shown in PDF and HTML form, see the `.Rmd` file.
+
 ```{r echo=FALSE}

### 2021-02-08 docs/ms_atlas_preproc_supplement.tex
new pdf and tex
add new tex and pdf

--- 2021-02-03 docs/ms_atlas_preproc_supplement_2021-02-03.tex
@@
 
-\usepackage{fontspec}
+% \usepackage{fontspec}
 % use nice fonts if available else use boring defaults
@@
 \usepackage{lineno}
-\KOMAoption{fontsize}{10pt}
-
-\IfFontExistsTF{Times New Roman}{\setmainfont[]{Times New Roman}}{} 
-\IfFontExistsTF{Arial}{\setsansfont[]{Arial}}{}
-\IfFontExistsTF{Fira Code}{\setmonofont{Fira Code}}
+% \KOMAoption{fontsize}{10pt}
+
+% \IfFontExistsTF{Palatino}{\setmainfont[]{Palatino}}{} 
+\usepackage{mathpazo}
+\usepackage{helvet}
+% \IfFontExistsTF{Arial}{\setsansfont[]{Arial}}{}
+% \IfFontExistsTF{Fira Code}{\setmonofont{Fira Code}}
 
@@
 \author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Sivan Toledo \and Ran Nathan \and Allert I. Bijleveld}
-\date{2021-02-03}
+\date{2021-02-08}
 
@@
 We began by visualising the data to check for location errors, and found a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2; main text Fig. 7.a).
-This outlier was removed by filtering data by the X coordinate bounds using the function \texttt{atl\_filter\textbackslash{}\_bounds}; X coordinate bounds \(\leq\) 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
+This outlier was removed by filtering data by the X coordinate bounds using the function \texttt{atl\_filter\_bounds}; X coordinate bounds \(\leq\) 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
 1.2).
@@
 
-\CommentTok{# prepare a palette}
-\NormalTok{pal <-}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
+\NormalTok{pal <-}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{5}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
+\NormalTok{pal[}\DecValTok{3}\NormalTok{] <-}\StringTok{ "seagreen"}
+\NormalTok{pal[}\DecValTok{2}\NormalTok{] <-}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{5}\NormalTok{, }\StringTok{"Blues"}\NormalTok{)[}\DecValTok{5}\NormalTok{]}
 \end{Highlighting}
@@
 
+\hypertarget{plot-figure-7}{%
+\section{Plot figure 7}\label{plot-figure-7}}
+
+Plotting code is not shown in PDF and HTML form, see the \texttt{.Rmd} file.
+
 \hypertarget{processing-egyptian-fruit-bat-tracks}{%

### 2021-02-08 docs/ms_atlas_preproc_main_text.tex
minor edits

@@
 With these in mind we have developed the \texttt{R} package \texttt{atlastools} \citep{gupte2020a, rcoreteam2020}, which implements important steps of the pre-processing pipeline.
-\texttt{R} is the computational environment of choice in movement ecology \citep{joo2020} and formalising tools as an \texttt{R} package improves portability and reproducibility \citep{marwick2018, }.
+\texttt{R} is the computational environment of choice in movement ecology \citep{joo2020} and formalising tools as an \texttt{R} package improves portability and reproducibility \citep{marwick2018}.
 \texttt{atlastools} functions are easy to use and understand, and rely on the efficiency of the \texttt{R} package \texttt{data.table} \citep{dowle2020}.
@@
 GPS systems' location error may be similarly indicated indirectly by the number of satellites involved in the localisation.
-Both GPS and reverse-GPS systems may also provide direct measures of location error during localisation \citep{maccurdy2009, maccurdy2019, weiser2016}.
+Both GPS and reverse-GPS systems may also provide direct measures of location error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS, and standard deviation SD in reverse-GPS]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
 % A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
@@
 This is difficult even with smoothed and thinned high-throughput data, as these may be too large for statistical packages, or have strong autocorrelation (Fig. 5).
-Users aiming for such analyses can benefit from segmentation-clustering the data into spatio-temporally independent `residence patches', and link the attributes of movement segments (such as duration or distance), with environmental data \citep{bijleveld2016, oudman2018, barraquand2008}.
-Making the patch the unit of observation conveniently sidesteps pseudo-replication while and reduces computational requirements.
-Furthermore, metrics such as the distance travelled within and between patches can help compare broad-scale individual movement strategies.
+Users aiming for such analyses can benefit from segmenting and clustering the data into spatio-temporally independent `residence patches', and metrics such as the distance travelled within and between patches can help compare broad-scale individual movement strategies, or be linked with environmental data \citep{bijleveld2016, oudman2018, barraquand2008}.
+Common methods of segmentation-clustering are either not based in the biology of the tracked individual, or are not scalable to very large yet often gappy datasets \citep{patin2020a}, making a first-principles approach an essential part of a pre-processing pipeline.
+Furthermore, making the patch the unit of observation conveniently sidesteps pseudo-replication while and reduces computational requirements.
 

### 2021-02-09 docs/ms_atlas_preproc_main_text.tex
add tex and pdf
edit intro and abstract

@@
 \\
-\textbf{2} Deptartment of Coastal Systems, NIOZ Royal Netherlands Institute for Sea Research, The Netherlands.
+\textbf{2} Department of Coastal Systems, NIOZ Royal Netherlands Institute for Sea Research, The Netherlands.
 \\
@@
 
-    \item A pipeline for cleaning massive high-throughput datasets must balance ease of use, computational efficiency, and the ability to prepare movement data for a range of analyses.
-    Data cleaning must also reject location errors without discarding valid animal movements.
+    \item A pipeline for cleaning massive high-throughput datasets must balance ease of use, computational efficiency.
+    Data cleaning must also reject location errors discarding valid animal movements.
     Another useful feature of a pre-processing pipeline is efficiently segmenting and clustering location data for statistical methods, while also being scalable to large datasets and robust to imperfect sampling.
@@
     This pipeline, consisting of removing outliers, smoothing the filtered result, and thinning it to a uniform sampling interval, is easily scaled to very large datasets.
-    We demonstrate this pipeline on simulated movement data to which we have randomly added location errors.
+    We demonstrate this pipeline on simulated movement data with location errors.
     We further suggest how large volumes of cleaned data can be segmented-clustered into biologically meaningful ‘residence patches’.
@@
 
-\textbf{Keywords} ATLAS, data cleaning, movement ecology, high-throughput tracking, \textit{atlastools}, residence patch
+\textbf{Keywords} high-throughput movement ecology (HTME), data cleaning, movement ecology, high-throughput tracking, \textit{atlastools}, residence patch
 
@@
 Investigating fine-scale environmental and social drivers and the consequences of movement requires position data from many individuals at high temporal and spatial resolution relative to the phenomenon of interest (high-throughput tracking).
-High-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}, yet ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput tracking data at much lower costs.
+High-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}; additionally, ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput tracking data at much lower costs.
+
 Although high-resolution tracking provides a massive amount of data on the path of a tracked animal, this data presents a two-fold challenge to ecologists.
@@
 Pre-processing protocols are not often reported in the literature, or may not be easily tractable for mainstream computing hardware and software.
-Furthermore, filtering out positions on their location error estimates may not be sufficient to recover a good track estimate when location error does not correspond to biological realism in movement \citep{weiser2016, ranacher2016}.
+Furthermore, filtering out positions on their location error estimates may not be be appropriate for outliers which represent unrealistic movement but have low error measures \citep{weiser2016, ranacher2016}.
 This makes identifying and removing implausible locations from a track an important component of recovering true animal movement \citep{bjorneraas2010}.
-Even after removing outliers, a track may comprise of positions that are distributed around the true animal location.
-Discarding these positions as `untrue' would lead to excessive data loss, but treating them as `real' locations would lead to unrealistic estimates of metrics such as speed \citep{noonan2019}.
+Even after removing outliers, a track may comprise of positions that are distributed around the true animal location \citep{noonan2019}.
 The large data-volumes of high-throughput tracking allow for a neat solution: tracks can be `smoothed' to reduce small location errors that have remained undetected.
-This large data volume becomes a challenge when users seek to examine animal space-use in relation to relevant environmental covariates, such as the predictors of prolonged residence in an area \citep[see][]{bracis2018}.
-First, statistical modelling using high-throughput data may suffer from pseudo-replication since both the response (e.g.
+This large data volume becomes a challenge when users seek to examine animal space-use in relation to relevant environmental covariates, such as the predictors of prolonged residence in an area \citep[see][]{bracis2018, aarts2008, fleming2014a,bijleveld2016, oudman2018, harel2016}.
+% First, statistical modelling using high-throughput data may suffer from pseudo-replication since both the response (e.g.
 step-length) and the predictors (e.g.
-resource landscape values) are spatio-temporally auto-correlated and hence non-independent \citep{aarts2008, fleming2014a,bijleveld2016, oudman2018, harel2016}.
-Second, fitting statistical models to datasets with many millions of observations may require significant innovation in mainstream statistical software \citep[e.g.
+resource landscape values) are spatio-temporally auto-correlated and hence non-independent \citep{}.
+% Second, fitting statistical models to datasets with many millions of observations may require significant innovation in mainstream statistical software \citep[e.g.
 ][]{wood2015}, and users may instead want to reduce data volumes by thinning or clustering.
@@
 We take two important considerations into account, (1) that methods must be computationally efficient, and (2) that the pre-processing steps should be easily understood and reproduced.
-With these in mind we have developed the \texttt{R} package \texttt{atlastools} \citep{gupte2020a, rcoreteam2020}, which implements important steps of the pre-processing pipeline.
 \texttt{R} is the computational environment of choice in movement ecology \citep{joo2020} and formalising tools as an \texttt{R} package improves portability and reproducibility \citep{marwick2018}.
+With these considerations in mind we have developed the \texttt{R} package \texttt{atlastools} \citep{gupte2020a, rcoreteam2020}, which implements important steps of the pre-processing pipeline.
 \texttt{atlastools} functions are easy to use and understand, and rely on the efficiency of the \texttt{R} package \texttt{data.table} \citep{dowle2020}.
-We begin by showing how the pipeline can clean simulated tracks to which we have artificially introduced location error \citep{gurarie2017}.
+% We begin by showing how the pipeline can clean simulated tracks to which we have artificially introduced location error \citep{gurarie2017}.
 Beginning with basic spatio-temporal filtering of positions, we cover filtering outliers from movement tracks, and reducing the effect of location error with a median smooth.

### 2021-02-09 docs/ms_atlas_preproc_main_text.tex
edit pipeline overview

@@
 Users may follow the pipeline in full, or implement the module most suitable for their data.
+Users are advised to visualise their data throughout their workflow to check for evident location errors \citep{slingsby2016}.
 
@@
 
-    \item Users should visualise their data to check for evident location errors \citep{slingsby2016}.
-    \item Users can install \texttt{atlastools} using the \texttt{install\_github} function from the \texttt{R} package \texttt{remotes}:
+    % \item Users can install \texttt{atlastools} using the \texttt{install\_github} function from the \texttt{R} package \texttt{remotes}:
 
-    \texttt{remotes::install\_github("pratikunterwegs/atlastools")}.
+    % \texttt{remotes::install\_github("pratikunterwegs/atlastools")}.
 
@@
     % Certain functions modify the input data "in place", and users must work on a copy of their data to preserve both the original and cleaned data (details in the Supplementary Material).
-    Since high-throughput tracking systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} users must convert their geographic coordinates to metres as this is assumed by all functions.
-    \item Users can use simple spatio-temporal range filters to select positions within a certain time or area.
+    % Since high-throughput tracking systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} users must convert their geographic coordinates to metres as this is assumed by all functions.
+    \item Users can use simple spatio-temporal filters to select positions within a certain time or area.
     %  (see \textsc{Spatio-Temporal Filtering}).
@@
 
-\subsection{Simulating Movement Tracks}
+\subsection{Demonstrating the Pipeline with Simulated Data}
 

### 2021-02-10 _02_filtering_data.Rmd
update code and fig name in rmd
edit figure 2

@@
 library(patchwork)
+library(ggthemes)
 ```
@@
   ggplot() +
+  geom_path(
+    data = data_copy,
+    aes(x, y),
+    col = "grey",
+    lwd = 0.2
+  ) +
   geom_point(
@@
 
+Histogram of speed
+
+```{r}
+fig_hist_speed =
+  ggplot()+
+  geom_histogram(
+    data = data_copy,
+    aes(x = in_speed,
+        fill = ifelse(in_speed > 0.05, "outlier", "good")),
+    position = position_identity(),
+    bins = 60
+  )+
+  geom_histogram(
+    data = data_copy[500:800, ],
+    aes(x = in_speed),
+    fill = pal[4],
+    bins = 60
+  )+
+  geom_vline(
+    xintercept = 0.05,
+    # size = 0.2,
+    lty = 2
+  )+
+  annotate(
+    geom = "text",
+    x = 0.11,
+    y = 150,
+    label = "Threshold",
+    size = 4,
+    angle = 90
+  )+
+  scale_fill_manual(
+    values = c(
+      "outlier" = pal[1],
+      "good" = "grey"
+    )
+  )+
+  scale_y_sqrt()+
+  scale_x_log10(
+  )+
+  theme_test(base_family = "sans")+
+  theme(
+    legend.position = "none",
+    axis.text = element_blank(),
+    axis.ticks = element_blank(),
+    axis.title.y = element_blank(),
+    plot.title = element_text(
+      face = "bold"
+    ),
+    plot.background = element_blank()
+  )+
+  labs(
+    title = "(c)",
+    x = "Speed"
+  )
+```
+
 ```{r echo=FALSE}
@@
     data = data_copy,
-    aes(x, y),
-    col = "grey",
+    aes(x, y,
+        col = ifelse(in_speed >= 0.05 | out_speed >= 0.05, 
+                     "outlier_path", "good_path"),
+        group = NA),
     lwd = 0.2
@@
     alpha = 0.6,
-    shape = 4, col = "grey30"
+    # stroke = 1,
+    shape = 4, col = pal[4]
   ) +
@@
     aes(x, y,
-      col = (in_speed >= 0.03 & out_speed >= 0.03),
+      col = ifelse(in_speed >= 0.03 & out_speed >= 0.03, "outlier", "good"),
       shape = (in_speed >= 0.03 & out_speed >= 0.03),
@@
   ) +
-  scale_color_manual(values = c(pal[3], "grey")) +
+  annotation_custom(
+    grob = ggplotGrob(
+      fig_hist_speed
+    ),
+    xmin = 0.80,
+    xmax = 1.08,
+    ymin = 0.25,
+    ymax = 0.6
+  )+
+  scale_color_manual(values = c(
+    "outlier_path" = "red",
+    "good_path" = "grey",
+    "good" = pal[3], 
+    "outlier" = "red")) +
   scale_shape_manual(values = c(1, 4)) +
@@
   ) +
-  ggthemes::theme_few() +
+  ggthemes::theme_few(
+    base_family = "Arial"
+  ) +
   theme(
     axis.text = element_blank(),
-    axis.title = element_blank()
+    axis.title = element_blank(),
+    legend.position = "none"
   ) +
@@
 
-![](figures/fig_03_filter_speed.png)
+![](figures/fig_02_filtering_data.png)
 

### 2021-02-10 _03_smoothing_tracks.Rmd
add speed histogram code

@@
 data[, window_size := NA]
+data$speed_in = atl_get_speed(data)
 
@@
 data_plot <- mapply(function(df, offset) {
+  df = copy(df)
   df[, x := x + offset]
+  df$speed_in = atl_get_speed(df)
+  return(df)
 }, list_of_smooths, seq(0.4, 1.25, length.out = 4),
@@
 data_plot <- rbindlist(data_plot)
+
+# add offset to data errors
+data_errors = lapply(c(0, seq(0.4, 1.25, length.out = 4)),
+                     function(offset) {
+  df = copy(data_errors)
+  df[,x := x + offset]
+  df[, offset := offset]
+})
+
+# bind list
+data_errors = rbindlist(data_errors)
+```
+
+```{r}
+pal2 = RColorBrewer::brewer.pal(4, "RdPu")[c(2, 4)]
+
+fig_hist_smooth =
+  ggplot()+
+  geom_histogram(
+    data = data_errors[offset == 0,],
+    aes(x = in_speed,
+        fill = "errors"),
+    bins = 80,
+    alpha = 1
+  )+
+  geom_histogram(
+    data = data_plot[window_size %in% c(5, 21)],
+    aes(x = speed_in,
+        fill = as.factor(window_size),
+        group = window_size),
+    bins= 80
+  )+
+  geom_histogram(
+    data = data,
+    aes(x = speed_in,
+        fill = "real"),
+    # col = pal[1],
+    alpha = 0.8,
+    bins = 80
+  )+
+  scale_x_log10()+
+  scale_fill_manual(
+    values = c(
+      "real" = alpha("grey20", .8),
+      "errors" = "grey",
+      "5" = pal2[1],
+      "21" = pal2[2]
+    ),
+    labels = c(
+      "real" = "real track",
+      "errors" = "track w/\nerrors",
+      "5" = "K = 5",
+      "21" = "K = 21"
+    )
+  )+
+  coord_cartesian(xlim = c(1e-5, .05))+
+  theme_test(base_family = "sans")+
+  theme(
+    legend.position = c(0.2, 0.65),
+    axis.text = element_blank(),
+    # axis.ticks.y = element_blank(),
+    axis.title.y = element_blank(),
+    plot.title = element_text(
+      face = "bold"
+    ),
+    plot.background = element_blank()
+  )+
+  labs(x= "Speed [log10 scale]",
+       fill = "Smoothing",
+       title = "(f)")
 ```
@@
   ggplot() +
+  geom_path(
+    data = data_errors,
+    aes(x, y,
+        group = offset),
+    col = "grey60",
+    size = 0.1
+  ) +
   geom_point(
-    data = data_errors,
+    data = data_errors[offset == 0,],
     aes(x, y),
@@
     aes(x, y,
-      col = window_size,
-      group = window_size
+      group = window_size,
+      col = window_size
     ),
     show.legend = F,
-    lwd = 0.5
+    lwd = 0.3
   ) +
@@
   scale_colour_distiller(
-    palette = "Purples", direction = 1,
-    values = c(-0.5, 1)
+    palette = "RdPu", direction = 1,
+    values = c(-1, 1)
   ) +
@@
   )
+
+# # wrap figures
+# figure_median_smooth =
+#   wrap_plots(figure_median_smooth,
+#              fig_hist_smooth,
+#              design = "AAAA\nBB##")
 

### 2021-02-10 docs/ms_atlas_preproc_main_text.tex
edits post orr comments

@@
     \includegraphics[width=0.95\textwidth]{figures/fig_02_filtering_data.png}
-    \caption{\textbf{(a)} Simulated movement data (black line), with three kinds of artificially added errors: (1) normally distributed small-scale error on each position, (2) large-scale error added to 0.5\% of positions, and (3) 300 consecutive positions displaced to simulate a gross distortion affecting a continuous subset of the track.
-    Tracks can be quickly filtered by spatial bounds (dashed grey lines) to exclude broad regions.
+    \caption{Simulated movement data (black line), with three kinds of artificially added errors: (1) normally distributed small-scale error on each position, (2) large-scale error added to 0.5\% of positions, and (3) 300 consecutive positions displaced to simulate a gross distortion affecting a continuous subset of the track.
+    \textbf{(a)} Tracks can be quickly filtered by spatial bounds (dashed grey lines) to exclude broad regions.
     \textbf{(b)} 
@@
 Temporal filtering can exclude positions from such periods when data are expected to be unreliable for ecological inference, either due to abnormal movement behaviour or system-specific issues.  
-Temporal filters can be combined with spatial filters to exclude time-location combinations which would be unrealistic for the tracked animal. 
-For example, red knots in the Dutch Wadden Sea congregate at communal roosts during high-tide, when their foraging grounds on the inter-tidal mudflats are inaccessible \citep{vangils2006}.
-A spatio-temporal filter that retains positions inside the bounds of roosts during high-tide can quickly return data that reveals individual presence at high-tide roosts, while excluding positions that erroneously place individuals on the inundated mudflats.
+Temporal filters can be combined with spatial filters to find specific time-location combinations. 
+For example, users who want to study the response of individuals to the stress of capture can select for positions shortly after release and within a certain distance of the release location.
 Users should apply filters in sequence rather than all at once, and visualise the output after each filtering step (`sanity checks').
@@
 % It accepts a character vector of \texttt{R} expressions that each return a logical vector (i.e., \texttt{TRUE} or \texttt{FALSE}; Listing 2).
-The function returns only those data which satisfy each of the filter conditions, and users must ensure that the filtering variables exist in their dataset in order to avoid errors.
+The function keeps only those data which satisfy each of the filter conditions, and users must ensure that the filtering variables exist in their dataset in order to avoid errors.
 
@@
 Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system (Beardsworth et al. \textit{in prep.}).
-For instance, in reverse-GPS ATLAS systems the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
-GPS systems' location error may be similarly indicated indirectly by the number of satellites involved in the localisation.
-Both GPS and reverse-GPS systems may also provide direct measures of location error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS, and standard deviation SD in reverse-GPS]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
+GPS systems provide direct measures of location error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS]{ranacher2016}, while  reverse-GPS systems provide a similar estimate \citep[called Standard Deviation, SD][]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
+Tracking data can also include indirect indicators of data quality.
+For instance, GPS systems' location error may be indicated indirectly by the number of satellites involved in the localisation.
+In reverse-GPS systems too the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
 % A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
@@
 Once a speed threshold $S$ has been chosen, positions with incoming \textit{and} outgoing speeds $\geq S$ may be identified as spikes and removed.
+
 Some species can realistically achieve speeds $\geq S$ in fast transit segments when assisted by their environment, such as birds with tailwinds, and a simple filter on incoming and outgoing speeds would exclude this valid data.
@@
 This combined filter assumes that positions in high-throughput tracking with both high speeds and large turning angles are likely to be due to location errors, since most species are unable to turn sharply at high speed.
-Users can then retain only those positions whose incoming and outgoing speeds are both $< S$ which or satisfy the condition $\theta < A$, where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
+Users can then remove those positions whose incoming and outgoing speeds are both $> S$ and $\theta > A$ (sharp, high-speed turns), where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
 Spike removal can be implemented using the \texttt{atl\_filter\_covariates} function (Listing 3).
@@
     \textbf{(a)} The goal of this step is to approximate the simulated canonical track (black line), given positions with small-scale error remaining from previous steps (green points). 
-    The canonical track is shown `zoomed in' for better representation, and the missing data at the top of the track (line with no points) is from where the prolonged spike was earlier removed.
+    % The canonical track is shown `zoomed in' for better representation, and the missing data at the top of the track (line with no points) is from where the prolonged spike was earlier removed.
     Median smoothing the given position coordinates (green points, in \textbf{(a)}) over a moving window ($K$) of \textbf{(b)} 3, \textbf{(c)} 5, \textbf{(d)} 11, and \textbf{(e)} 21 positions respectively, yields approximations (purple lines) of the canonical track.
@@
 Thinning data need not compromise researchers' ability to answer scientific questions with them; for instance, proximity-based social interactions lasting 1 -- 2 minutes would still be detected on thinning from a sampling interval of 1 second to 1 minute.
-Indeed, temporal auto-correlation may hinder some methods such as the estimation of home-ranges or step-selection functions \citep{fleming2014a, dupke2017}.
+Indeed, some analyses require that temporal auto-correlation in the data be broken by subsampling the data to a lower resolution \citep[such as the estimation of home-ranges or step-selection functions][]{fleming2014a, dupke2017}.
 Added to the requirement of uniform sampling intervals by many methods in the field, evenly reducing data volumes is a worthwhile endeavour \citep[e.g.][]{fleming2014a, michelot2016, avgar2016}.
@@
 Categorical variables, such as the habitat type associated with each position can be aggregated using a suitable measure such as the mode.
+We caution users that thinning causes an extensive loss of small-scale detail in the data, and should be used carefully.
 
 The aggregation method is less sensitive to selecting point outliers by chance than resampling.
-When users want to account for location error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location error when thinning is important.
+When users want to account for location error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location error is important.
+Resampling directly propagates these errors, while durin aggregation,
 % In reverse-GPS systems systems the location error is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
-The location error around each coordinate provided by either GPS or reverse-GPS system can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
+the location error around each coordinate provided by either GPS or reverse-GPS system can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
 \begin{linenomath*}
@@
 \end{linenomath*}
-We caution users that thinning causes an extensive loss of small-scale detail in the data, and should be used carefully.
 
@@
     \includegraphics[width=0.75\textwidth]{figures/fig_06_calib_residence_patch.png}
-    \caption{Pre-processing steps for WATLAS calibration data showing filtering filtering on speed, median smoothing and thinning by aggregation, and making residence patches.
+    \caption{Pre-processing steps for WATLAS calibration data showing filtering on speed, median smoothing and thinning by aggregation, and making residence patches.
     \textbf{(a)} Positions with incoming and outgoing speed $\geq$ 15 m/s are removed (grey crosses = removed, green points = retained).
-    Rectangles show the region expanded in subfigures (b) and (c).
+    Rectangles show the region expanded in panels (b) and (c).
     \textbf{(b)} Expanded view of upper rectangle in (a) showing the raw data (grey crosses), the median smoothed positions (green circles; moving window $K$ = 5), and the smoothed track thinned by aggregation to a 30 second interval (purple squares).
-    Square size corresponds to the number of positions used to calculate the averaged position during thinning, with no significant differences.
-    While median smoothing retains all observations (though changing their coordinates), thinning aggregates these positions into approx. 5\% of the smoothed data volume.
-    Median smoothing recovers a good estimate of the true track from significantly over-dispersed raw data, while thinning reduces the data volume but preserves track structure.
+    Square size corresponds to the number of positions used to calculate the averaged position during thinning.
+    % While median smoothing retains all observations (though changing their coordinates), thinning aggregates these positions into approx. 5\% of the smoothed data volume.
+    % Median smoothing recovers a good estimate of the true track from significantly over-dispersed raw data, while thinning reduces the data volume but preserves track structure.
     \textbf{(c)} Classifying thinned data into residence patches yields robust estimates of the duration of known stops. The island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) is shown in beige.
     Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
-    However, the algorithm identified all areas with prolonged residence, including those which were not intended stops, such as stops at the field station (n = 12; green polygon over red crossed-squares).
+    The algorithm identified all areas with prolonged residence, including those which we had not intended to be recorded, such as stops at the field station (n = 12; green polygon over red crossed-squares).
     The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).}
@@
     In addition, the linear approximation of movement between patches (blue paths in panel (a)) can be contrasted with the estimated real path between patches (green lines).
-    \textbf{(c)} Residence patch polygons can also show how individuals partition space and resources (black circles are fruit trees), and the combined influence of resource and potential social cues on movement between patches (coloured paths).
-    Patches and paths are coloured by bat identity.}
+    \textbf{(c)} Fine scale tracks (thin coloured lines), large scale movement (thick lines), residence patch polygons, and fruit tree locations show how high-throughput data can be used to study movement across scales.
+    Patches and lines are coloured by bat identity.}
     \label{fig:figure_bats}

### 2021-02-10 _03_smoothing_tracks.Rmd
boxplots instead of error bars fig 4

@@
 ```
+
+## Histogram of speeds after smoothing
 
@@
 
-# # wrap figures
-# figure_median_smooth =
-#   wrap_plots(figure_median_smooth,
-#              fig_hist_smooth,
-#              design = "AAAA\nBB##")
-
 # save figure
@@
     df[, speed := atl_get_speed(df)]
-    df[, list(
-      median = median(speed, na.rm = T),
-      sd = sd(speed, na.rm = T),
-      interval = first(interval)
-    )]
+    # df[, list(
+    #   median = median(speed, na.rm = T),
+    #   sd = sd(speed, na.rm = T),
+    #   interval = first(interval)
+    # )]
   })
@@
   ) +
-  geom_errorbar(
+  geom_boxplot(
     data = speed_agg_smooth,
     aes(
-      x = factor(interval),
-      ymin = 1 + median - sd,
-      ymax = 1 + median + sd
+      x = as.factor(interval),
+      y = 1 + speed
     ),
-    width = 0.2,
-    position = position_nudge(x = 0.25)
-  ) +
-  geom_point(
-    data = speed_agg_smooth,
-    aes(
-      x = factor(interval),
-      y = 1 + median
-    ),
-    shape = 21,
-    size = 3,
     fill = pal[3],
+    show.legend = F,
+    width = 0.25,
+    outlier.size = 0.2,
     position = position_nudge(x = 0.25)
@@
   labs(
-    x = "interval (s)",
-    y = "speed"
+    x = "Interval (s)",
+    y = "Speed"
   )
@@
 
-![](figures/fig_05_thinning.png)
+![](figures/fig_04_thinning.png)
 

### 2021-02-10 05_calibration_data.Rmd
change figure symbols

@@
     aes(x_median, y_median),
-    size = 10, shape = 7,
+    size = 10, 
+    shape = 2,
     stroke = 1,
@@
     data = hut,
-    size = 10, shape = 7,
+    size = 10, shape = 2,
     stroke = 1,

### 2021-02-10 docs/ms_atlas_preproc_main_text.tex
add updated tex

@@
     Square size corresponds to the number of positions used to calculate the averaged position during thinning.
-    % While median smoothing retains all observations (though changing their coordinates), thinning aggregates these positions into approx. 5\% of the smoothed data volume.
-    % Median smoothing recovers a good estimate of the true track from significantly over-dispersed raw data, while thinning reduces the data volume but preserves track structure.
     \textbf{(c)} Classifying thinned data into residence patches yields robust estimates of the duration of known stops. The island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) is shown in beige.
-    Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
-    The algorithm identified all areas with prolonged residence, including those which we had not intended to be recorded, such as stops at the field station (n = 12; green polygon over red crossed-squares).
+    Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple triangles).
+    The algorithm identified all areas with prolonged residence, including those which we had not intended to be recorded, such as stops at the field station (n = 12; green polygon over red triangles).
     The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).}
@@
     Bat tracks are are shown as thin lines below the linear approximations, and colours show bat identity. The grey hexagon represents a roost site.
-    Dark green points represent known fruit trees.
-    Background is shaded by elevation at 30 metre resolution (SRTM).    
+    Black points represent known fruit trees.
+    Background is shaded by elevation at 30 metre resolution.
     \textbf{(b)} Spatial representations of an individual bat's residence patches (green polygons) can be used to study site-fidelity by examining overlaps between patches, or to study resource selection by inspecting overlaps with known resources (here, black circles show the location of clusters of fruit trees).

### 2021-02-11 05_calibration_data.Rmd
tense change

@@
 
-We began by visualising the data to check for location errors, and found a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2; main text Fig. 7.a).
-This outlier was removed by filtering data by the X coordinate bounds using the function `atl_filter_bounds`; X coordinate bounds $\leq$ 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
+We begin by visualising the data to check for location errors, and find a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2).
+This outlier is removed by filtering data by the X coordinate bounds using the function `atl_filter_bounds`; X coordinate bounds $\leq$ 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
 1.2).
-We then calculated the incoming and outgoing speed, as well as the turning angle at each position using the functions `atl_get_speed` and `atl_turning_angle` respectively, as a precursor to targeting large-scale location errors in the form of point outliers.
-We used the function `atl_filter_covariates` to remove positions with incoming and outgoing speeds $\geq$ the speed threshold of 15 m/s (n = 13,491, 26.5%; remaining positions = 37,324, 73.5%; Fig.
+We then calculate the incoming and outgoing speed, as well as the turning angle at each position using the functions `atl_get_speed` and `atl_turning_angle` respectively, as a precursor to targeting large-scale location errors in the form of point outliers.
+We use the function `atl_filter_covariates` to remove positions with incoming and outgoing speeds $\geq$ the speed threshold of 15 m/s (n = 13,491, 26.5%; remaining positions = 37,324, 73.5%; Fig.
 1.3; main text Fig.
 7.b).
-This speed threshold was chosen as 5 m/s faster than the known boat speed, 10 m/s.
-Finally, we targeted small-scale location errors by applying a median smooth with a moving window size $K$ = 5 using the function `atl_median_smooth` (Fig. 1.4; main text Fig. 7.c).
-Smoothing did not reduce the number of positions.
-We thinned the data to a 30 second interval leaving 1,803 positions (4.8\% positions of the smoothed track)
+This speed threshold is chosen as 5 m/s faster than the known boat speed, 10 m/s.
+Finally, we target small-scale location errors by applying a median smooth with a moving window size $K$ = 5 using the function `atl_median_smooth` (Fig. 1.4; main text Fig. 7.c).
+Smoothing does not reduce the number of positions.
+We thin the data to a 30 second interval leaving 1,803 positions (4.8\% positions of the smoothed track)
 

### 2021-02-11 05_calibration_data.Rmd
median smoother

@@
 This speed threshold is chosen as 5 m/s faster than the known boat speed, 10 m/s.
-Finally, we target small-scale location errors by applying a median smooth with a moving window size $K$ = 5 using the function `atl_median_smooth` (Fig. 1.4; main text Fig. 7.c).
+Finally, we target small-scale location errors by applying a median smoother with a moving window size $K$ = 5 using the function `atl_median_smooth` (Fig. 1.4; main text Fig. 7.c).
 Smoothing does not reduce the number of positions.

### 2021-02-11 05_calibration_data.Rmd
add install step here

@@
 
-We begin by visualising the data to check for location errors, and find a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2).
+We begin by preparing the libraries we need, and installing `atlastools` from Github.
+After installing `atlastools`, we visualise the data to check for location errors, and find a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2).
 This outlier is removed by filtering data by the X coordinate bounds using the function `atl_filter_bounds`; X coordinate bounds $\leq$ 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
@@
 pal[2] <- RColorBrewer::brewer.pal(5, "Blues")[5]
+```
+
+## Install `atlastools` from Github.
+
+`atlastools` is available from Github and is archived on Zenodo [@gupte2020a].
+It can be installed using `remotes` or `devtools`. Here we use the `remotes` function `install_github`.
+
+```{r install_atlastools}
+install.packages("remotes")
+
+# installation using remotes
+remotes::install_github("pratikunterwegs/atlastools")
 ```

### 2021-02-11 05_calibration_data.Rmd
show one plot code chunk

@@
 
-```{r echo=FALSE}
+Here we show how data can be easily visualised using the popular plotting package `ggplot2`.
+Note that we plot both the points (`geom_point`) and the inferred path between them (`geom_path`), and specify a geospatial coordinate system in metres, suitable for the Dutch Wadden Sea (UTM 31N; ESPG code:32631; `coord_sf`).
+We save the output to file for future reference.
+
+Since plot code can become very lengthy and complicated, we omit showing further plot code in versions of this document rendered as PDF or HTML; it can however be seen in the online `.Rmd` version.
+
+```{r echo=TRUE}
 # plot data

### 2021-02-11 05_calibration_data.Rmd
add double conversion

@@
 ![Removal of a point outlier using the function `atl_filter_bounds`.
-The point outlier (black point) is removed based on its X coordinate value, with the data filtered to exclude positions with an X coordinate < 645,000 in the UTM 31N system.
+The point outlier (black point) is removed based on its X coordinate value, with the data filtered to exclude positions with an X coordinate < 645,000 in the UTM 31N coordinate system.
 Positions that are retained are shown in green.](figures/fig_calib_bbox.png)
@@
 
-Time in ATLAS tracking is counted in milliseconds and is represented by a 64-bit integer (type `long`), which is not natively supported in R; it will instead be converted to a `numeric`, or `double`.
+Time in ATLAS tracks is represented by 64-bit integers (type `long`) that specify time in milliseconds, starting from the beginning of 1970 (the UNIX epoch).
+This representation of time is called `POSIX` time and is usually specified in seconds, not milliseconds.
 
-This is not what is intended, but it works. The `bit64` package can help handle 64-bit integers if you want to keep to intended type.
+Since about 1.6 billion seconds have passed since the beginning of 1970, current `POSIX` times in milliseconds cannot be represented by R's built-in 32-bit integers.
+A naive conversion results in truncation of out-of-range numbers leading to huge errors (dates many thousands of years in the future).
 
-A further issue is that 64-bit integers (whether represented as `bit64` or `double`) do not yield meaninful results when you try to convert them to a date-time object, such as of the class `POSIXct`.
+R does not natively support 64-bit integers. One option is to use the bit64 package, which adds 64-bit integer support to R.
 
-This is because `as.POSIXct` fails when trying to work with 64-bit integers (it cannot interpret this type), and returns a date many thousands of years in the future (approx.
-52,000 CE) if the time column is converted to `numeric`.
+A simpler solution is to convert the times to R's built in `double` data type (also called `numeric`), which uses a 64-bit floating point representation.
+This representation can represent integers with up to 16 digits without error; we only need 13 digits to represent the number of milliseconds since 1970, so the conversion is error free.
+We can also perform the conversion and then divide by 1000 so that times are represented in seconds, not milliseconds; this simplifies speed estimation.
 
-There are two possible solutions. The parsimonious one is to convert the 64-bit number to a 32-bit short integer (dividing by 1000), or to use the `nanotime` package.
-
-The conversion method loses an imperceptible amount of precision. The `nanotime` requires installing another package. The first method is shown here.
-
-In the spirit of not destroying data, we create a second lower-case column called `time`.
+If second-resolution is accurate enough (it is for our purposes), the solution that we use is to divide times by 1000 to reduce the resolution from milliseconds to seconds and then to convert the time stamps to R integers.
+In the spirit of not destroying data, we create a second lower-case column called `time` to store this
 
@@
 # divide by 1000, convert to integer, then convert to POSIXct
-data[, time := as.integer(TIME / 1000)]
+data[, time := as.integer(
+  as.numeric(TIME) / 1000
+)]
 ```

### 2021-02-11 05_calibration_data.Rmd
rationalise lib calls

@@
 7.b).
-This speed threshold is chosen as 5 m/s faster than the known boat speed, 10 m/s.
+This speed threshold is chosen as the fastest boat speed during the experiment, 15 m/s.
 Finally, we target small-scale location errors by applying a median smoother with a moving window size $K$ = 5 using the function `atl_median_smooth` (Fig. 1.4; main text Fig. 7.c).
@@
 ```{r}
-# load libs
+# for data handling
 library(data.table)
 library(atlastools)
+
+# for recursion analysis
+library(recurse)
+
+# for plotting
 library(ggplot2)
@@
 
+# making a colour palette
 pal <- RColorBrewer::brewer.pal(5, "Set1")
 pal[3] <- "seagreen"
-pal[2] <- RColorBrewer::brewer.pal(5, "Blues")[5]
 ```
@@
 ```{r}
-# load recurse
-library(recurse)
-```
-
-```{r}
 # get 4 column data

### 2021-02-11 06_bat_data.Rmd
atlastools call
remove install code

@@
 
+# atlastools
+library(atlastools)
+
 # prepare a palette
 pal <- RColorBrewer::brewer.pal(4, "Set1")
-```
-
-## Install `atlastools` from Github.
-
-`atlastools` is available from Github and is archived on Zenodo [@gupte2020a].
-It can be installed using `remotes` or `devtools`. Here we use the `remotes` function `install_github`.
-
-```{r install_atlastools}
-install.packages("remotes")
-
-# installation using remotes
-remotes::install_github("pratikunterwegs/atlastools")
 ```
@@
 ```{r}
-data_clean <- fwrite(rbindlist(data_smooth),
+fwrite(rbindlist(data_smooth),
   file = "data/data_bat_smooth.csv"

### 2021-02-11 06_bat_data.Rmd
package lib arrange

@@
 ```{r}
-# load libs
+# libs for data
 library(data.table)
 library(RSQLite)
+library(atlastools)
+
+# libs for plotting
 library(ggplot2)
@@
 
-# atlastools
-library(atlastools)
+# recursion analysis
+library(recurse)
 
@@
 ```{r}
-# load recurse
-library(recurse)
-
 # split the data

### 2021-02-11 docs/ms_atlas_preproc_main_text.tex
update tex

@@
   breaklines=true,
-  numbers=left,
+%   numbers=left,
   frame=L,
@@
 Efficient location error modelling approaches \citep{fleming2020,aspillaga2021} may eventually make data-cleaning optional.
-Yet cleaning tracking data even partially before modelling location error will be faster than error-modelling on the full data, and the removal of large location errors may improve model fits.
+Yet cleaning tracking data even partially before modelling location error is faster than error-modelling on the full data, and the removal of large location errors may improve model fits.
 Thus we see our pipeline as complementary to these approaches \citep{fleming2014a,fleming2020}.

### 2021-02-11 _02_filtering_data.Rmd
new section
style code

@@
 
-We prepare the data for plotting.
-
-Histogram of speed
+### Histogram of speed
 
 ```{r}
-fig_hist_speed =
-  ggplot()+
+fig_hist_speed <-
+  ggplot() +
   geom_histogram(
     data = data_copy,
-    aes(x = in_speed,
-        fill = ifelse(in_speed > 0.05, "outlier", "good")),
+    aes(
+      x = in_speed,
+      fill = ifelse(in_speed > 0.05, "outlier", "good")
+    ),
     position = position_identity(),
     bins = 60
-  )+
+  ) +
   geom_histogram(
@@
     bins = 60
-  )+
+  ) +
   geom_vline(
@@
     lty = 2
-  )+
+  ) +
   annotate(
@@
     angle = 90
-  )+
+  ) +
   scale_fill_manual(
@@
     )
-  )+
-  scale_y_sqrt()+
-  scale_x_log10(
-  )+
-  theme_test(base_family = "sans")+
+  ) +
+  scale_y_sqrt() +
+  scale_x_log10() +
+  theme_test(base_family = "sans") +
   theme(
@@
     plot.background = element_blank()
-  )+
+  ) +
   labs(
@@
 ```
+
+## Figure outlier removal
 
@@
       col = ifelse(in_speed >= 0.05 | out_speed >= 0.05,
-                     "outlier_path", "good_path"),
-        group = NA),
+        "outlier_path", "good_path"
+      ),
+      group = NA
+    ),
     lwd = 0.2
@@
     ymax = 0.6
-  )+
+  ) +
   scale_color_manual(values = c(
@@
     "good" = pal[3],
-    "outlier" = "red")) +
+    "outlier" = "red"
+  )) +
   scale_shape_manual(values = c(1, 4)) +

### 2021-02-11 _03_smoothing_tracks.Rmd
style code

@@
 data[, window_size := NA]
-data$speed_in = atl_get_speed(data)
+data$speed_in <- atl_get_speed(data)
 
@@
 data_plot <- mapply(function(df, offset) {
-  df = copy(df)
+  df <- copy(df)
   df[, x := x + offset]
-  df$speed_in = atl_get_speed(df)
+  df$speed_in <- atl_get_speed(df)
   return(df)
@@
 # add offset to data errors
-data_errors = lapply(c(0, seq(0.4, 1.25, length.out = 4)),
+data_errors <- lapply(
+  c(0, seq(0.4, 1.25, length.out = 4)),
   function(offset) {
-  df = copy(data_errors)
-  df[,x := x + offset]
+    df <- copy(data_errors)
+    df[, x := x + offset]
     df[, offset := offset]
-})
+  }
+)
 
 # bind list
-data_errors = rbindlist(data_errors)
+data_errors <- rbindlist(data_errors)
 ```
@@
 ```{r}
-pal2 = RColorBrewer::brewer.pal(4, "RdPu")[c(2, 4)]
+pal2 <- RColorBrewer::brewer.pal(4, "RdPu")[c(2, 4)]
 
-fig_hist_smooth =
-  ggplot()+
+fig_hist_smooth <-
+  ggplot() +
   geom_histogram(
-    data = data_errors[offset == 0,],
-    aes(x = in_speed,
-        fill = "errors"),
+    data = data_errors[offset == 0, ],
+    aes(
+      x = in_speed,
+      fill = "errors"
+    ),
     bins = 80,
     alpha = 1
-  )+
+  ) +
   geom_histogram(
     data = data_plot[window_size %in% c(5, 21)],
-    aes(x = speed_in,
+    aes(
+      x = speed_in,
       fill = as.factor(window_size),
-        group = window_size),
-    bins= 80
-  )+
+      group = window_size
+    ),
+    bins = 80
+  ) +
   geom_histogram(
     data = data,
-    aes(x = speed_in,
-        fill = "real"),
+    aes(
+      x = speed_in,
+      fill = "real"
+    ),
     # col = pal[1],
@@
     bins = 80
-  )+
-  scale_x_log10()+
+  ) +
+  scale_x_log10() +
   scale_fill_manual(
@@
     )
-  )+
-  coord_cartesian(xlim = c(1e-5, .05))+
-  theme_test(base_family = "sans")+
+  ) +
+  coord_cartesian(xlim = c(1e-5, .05)) +
+  theme_test(base_family = "sans") +
   theme(
@@
     plot.background = element_blank()
-  )+
-  labs(x= "Speed [log10 scale]",
+  ) +
+  labs(
+    x = "Speed [log10 scale]",
     fill = "Smoothing",
-       title = "(f)")
+    title = "(f)"
+  )
 ```
@@
     aes(x, y,
-        group = offset),
+      group = offset
+    ),
     col = "grey60",
@@
   geom_point(
-    data = data_errors[offset == 0,],
+    data = data_errors[offset == 0, ],
     aes(x, y),

### 2021-02-11 R/06_bat_data.R
new scripts

@@ -1,19 +1,18 @@
 ## -----------------------------------------------------------------------------
-# load libs
+# libs for data
 library(data.table)
 library(RSQLite)
+library(atlastools)
+
+# libs for plotting
 library(ggplot2)
 library(patchwork)
 
+# recursion analysis
+library(recurse)
+
 # prepare a palette
 pal <- RColorBrewer::brewer.pal(4, "Set1")
-
-
-## ----install_atlastools-------------------------------------------------------
-install.packages("remotes")
-
-# installation using remotes
-remotes::install_github("pratikunterwegs/atlastools")
 
 
 ## ----read_bat_data------------------------------------------------------------
@@ -323,9 +322,6 @@ ggsave(fig_bat_smooth,
 
 
 ## -----------------------------------------------------------------------------
-# load recurse
-library(recurse)
-
 # split the data
 data_smooth <- split(data_smooth, data_smooth$TAG)
 
@@ -459,7 +455,7 @@ st_write(data_spatials,
 
 
 ## -----------------------------------------------------------------------------
-data_clean <- fwrite(rbindlist(data_smooth),
+fwrite(rbindlist(data_smooth),
   file = "data/data_bat_smooth.csv"
 )
 

### 2021-02-11 R/_03_smoothing_tracks.R

@@ -14,6 +14,7 @@ pal[3] <- "seagreen"
 # read in the data and set the window size variable
 data <- fread("data/data_sim.csv")[5000:10000, ]
 data[, window_size := NA]
+data$speed_in <- atl_get_speed(data)
 
 # data with small scale errors but no reflections or outliers
 data_errors <- fread("data/data_no_reflection.csv")
@@ -44,12 +45,95 @@ fwrite(list_of_smooths[[3]], file = "data/data_smooth.csv")
 ## -----------------------------------------------------------------------------
 # bind list after offset
 data_plot <- mapply(function(df, offset) {
+  df <- copy(df)
   df[, x := x + offset]
+  df$speed_in <- atl_get_speed(df)
+  return(df)
 }, list_of_smooths, seq(0.4, 1.25, length.out = 4),
 SIMPLIFY = F
 )
 
 data_plot <- rbindlist(data_plot)
+
+# add offset to data errors
+data_errors <- lapply(
+  c(0, seq(0.4, 1.25, length.out = 4)),
+  function(offset) {
+    df <- copy(data_errors)
+    df[, x := x + offset]
+    df[, offset := offset]
+  }
+)
+
+# bind list
+data_errors <- rbindlist(data_errors)
+
+
+## -----------------------------------------------------------------------------
+pal2 <- RColorBrewer::brewer.pal(4, "RdPu")[c(2, 4)]
+
+fig_hist_smooth <-
+  ggplot() +
+  geom_histogram(
+    data = data_errors[offset == 0, ],
+    aes(
+      x = in_speed,
+      fill = "errors"
+    ),
+    bins = 80,
+    alpha = 1
+  ) +
+  geom_histogram(
+    data = data_plot[window_size %in% c(5, 21)],
+    aes(
+      x = speed_in,
+      fill = as.factor(window_size),
+      group = window_size
+    ),
+    bins = 80
+  ) +
+  geom_histogram(
+    data = data,
+    aes(
+      x = speed_in,
+      fill = "real"
+    ),
+    # col = pal[1],
+    alpha = 0.8,
+    bins = 80
+  ) +
+  scale_x_log10() +
+  scale_fill_manual(
+    values = c(
+      "real" = alpha("grey20", .8),
+      "errors" = "grey",
+      "5" = pal2[1],
+      "21" = pal2[2]
+    ),
+    labels = c(
+      "real" = "real track",
[…]

### 2021-02-11 docs/ms_atlas_preproc_supplement.tex

@@
 \author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Sivan Toledo \and Ran Nathan \and Allert I. Bijleveld}
-\date{2021-02-08}
+\date{2021-02-11}
 
@@
 
-We began by visualising the data to check for location errors, and found a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2; main text Fig. 7.a).
-This outlier was removed by filtering data by the X coordinate bounds using the function \texttt{atl\_filter\_bounds}; X coordinate bounds \(\leq\) 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
+We begin by preparing the libraries we need, and installing \texttt{atlastools} from Github.
+After installing \texttt{atlastools}, we visualise the data to check for location errors, and find a single outlier position approx. 15km away from the study area (Fig. 1.1, 1.2).
+This outlier is removed by filtering data by the X coordinate bounds using the function \texttt{atl\_filter\_bounds}; X coordinate bounds \(\leq\) 645,000 in the UTM 31N coordinate reference system were removed (n = 1; remaining positions = 50,815; Fig.
 1.2).
-We then calculated the incoming and outgoing speed, as well as the turning angle at each position using the functions \texttt{atl\_get\_speed} and \texttt{atl\_turning\_angle} respectively, as a precursor to targeting large-scale location errors in the form of point outliers.
-We used the function \texttt{atl\_filter\_covariates} to remove positions with incoming and outgoing speeds \(\geq\) the speed threshold of 15 m/s (n = 13,491, 26.5\%; remaining positions = 37,324, 73.5\%; Fig.
+We then calculate the incoming and outgoing speed, as well as the turning angle at each position using the functions \texttt{atl\_get\_speed} and \texttt{atl\_turning\_angle} respectively, as a precursor to targeting large-scale location errors in the form of point outliers.
+We use the function \texttt{atl\_filter\_covariates} to remove positions with incoming and outgoing speeds \(\geq\) the speed threshold of 15 m/s (n = 13,491, 26.5\%; remaining positions = 37,324, 73.5\%; Fig.
 1.3; main text Fig.
 7.b).
-This speed threshold was chosen as 5 m/s faster than the known boat speed, 10 m/s.
-Finally, we targeted small-scale location errors by applying a median smooth with a moving window size \(K\) = 5 using the function \texttt{atl\_median\_smooth} (Fig. 1.4; main text Fig. 7.c).
-Smoothing did not reduce the number of positions.
-We thinned the data to a 30 second interval leaving 1,803 positions (4.8\% positions of the smoothed track)
+This speed threshold is chosen as the fastest boat speed during the experiment, 15 m/s.
+Finally, we target small-scale location errors by applying a median smoother with a moving window size \(K\) = 5 using the function \texttt{atl\_median\_smooth} (Fig. 1.4; main text Fig. 7.c).
+Smoothing does not reduce the number of positions.
+We thin the data to a 30 second interval leaving 1,803 positions (4.8\% positions of the smoothed track)
 
@@
 \begin{Highlighting}[]
-\CommentTok{# load libs}
+\CommentTok{# for data handling}
 \KeywordTok{library}\NormalTok{(data.table)}
 \KeywordTok{library}\NormalTok{(atlastools)}
+
+\CommentTok{# for recursion analysis}
+\KeywordTok{library}\NormalTok{(recurse)}
+
+\CommentTok{# for plotting}
 \KeywordTok{library}\NormalTok{(ggplot2)}
@@
 
+\CommentTok{# making a colour palette}
 \NormalTok{pal <-}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{5}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
 \NormalTok{pal[}\DecValTok{3}\NormalTok{] <-}\StringTok{ "seagreen"}
-\NormalTok{pal[}\DecValTok{2}\NormalTok{] <-}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{5}\NormalTok{, }\StringTok{"Blues"}\NormalTok{)[}\DecValTok{5}\NormalTok{]}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{install-atlastools-from-github}{%
+\section{\texorpdfstring{Install \texttt{atlastools} from Github}{Install atlastools from Github}}\label{install-atlastools-from-github}}
+
+\texttt{atlastools} is available from Github and is archived on Zenodo (Gupte \protect\hyperlink{ref-gupte2020a}{2020}).
+It can be installed using \texttt{remotes} or \texttt{devtools}. Here we use the \texttt{remotes} function \texttt{install\_github}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{install.packages}\NormalTok{(}\StringTok{"remotes"}\NormalTok{)}
+
+\CommentTok{# installation using remotes}
+\NormalTok{remotes}\OperatorTok{::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
 \end{Highlighting}
@@
 \NormalTok{data_raw <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+\end{Highlighting}
+\end{Shaded}
+
+Here we show how data can be easily visualised using the popular plotting package \texttt{ggplot2}.
+Note that we plot both the points (\texttt{geom\_point}) and the inferred path between them (\texttt{geom\_path}), and specify a geospatial coordinate system in metres, suitable for the Dutch Wadden Sea (UTM 31N; ESPG code:32631; \texttt{coord\_sf}).
+We save the output to file for future reference.
+
+Since plot code can become very lengthy and complicated, we omit showing further plot code in versions of this document rendered as PDF or HTML; it can however be seen in the online \texttt{.Rmd} version.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# plot data}
+\NormalTok{fig_data_raw <-}
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(data) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_path}\NormalTok{(}\KeywordTok{aes}\NormalTok{(x, y),}
+    \DataTypeTok{col =} \StringTok{"grey"}\NormalTok{, }\DataTypeTok{alpha =} \DecValTok{1}\NormalTok{, }\DataTypeTok{size =} \FloatTok{0.2}
+\NormalTok{  ) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(x, y),}
+    \DataTypeTok{col =} \StringTok{"grey"}\NormalTok{, }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{, }\DataTypeTok{size =} \FloatTok{0.2}
+\NormalTok{  ) }\OperatorTok{+}
+\StringTok{  }\NormalTok{ggthemes}\OperatorTok{::}\KeywordTok{theme_few}\NormalTok{() }\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}
+    \DataTypeTok{axis.title =} \KeywordTok{element_blank}\NormalTok{(),}
+    \DataTypeTok{axis.text =} \KeywordTok{element_blank}\NormalTok{()}
+\NormalTok{  ) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{coord_sf}\NormalTok{(}\DataTypeTok{crs =} \DecValTok{32631}\NormalTok{)}
+
+\CommentTok{# save figure}
+\KeywordTok{ggsave}\NormalTok{(fig_data_raw,}
+  \DataTypeTok{filename =} \StringTok{"figures/fig_calibration_raw.png"}\NormalTok{,}
+  \DataTypeTok{width =} \DecValTok{185} \OperatorTok{/}\StringTok{ }\DecValTok{25}
+\NormalTok{)}
 \end{Highlighting}
@@
 \caption{Removal of a point outlier using the function \texttt{atl\_filter\_bounds}.
-The point outlier (black point) is removed based on its X coordinate value, with the data filtered to exclude positions with an X coordinate \textless{} 645,000 in the UTM 31N system.
+The point outlier (black point) is removed based on its X coordinate value, with the data filtered to exclude positions with an X coordinate \textless{} 645,000 in the UTM 31N coordinate system.
 Positions that are retained are shown in green.}
@@
 
-Time in ATLAS tracking is counted in milliseconds and is represented by a 64-bit integer (type \texttt{long}), which is not natively supported in R; it will instead be converted to a \texttt{numeric}, or \texttt{double}.
-
-This is not what is intended, but it works. The \texttt{bit64} package can help handle 64-bit integers if you want to keep to intended type.
-
-A further issue is that 64-bit integers (whether represented as \texttt{bit64} or \texttt{double}) do not yield meaninful results when you try to convert them to a date-time object, such as of the class \texttt{POSIXct}.
-
-This is because \texttt{as.POSIXct} fails when trying to work with 64-bit integers (it cannot interpret this type), and returns a date many thousands of years in the future (approx.
-52,000 CE) if the time column is converted to \texttt{numeric}.
-
-There are two possible solutions. The parsimonious one is to convert the 64-bit number to a 32-bit short integer (dividing by 1000), or to use the \texttt{nanotime} package.
-
-The conversion method loses an imperceptible amount of precision. The \texttt{nanotime} requires installing another package. The first method is shown here.
-
-In the spirit of not destroying data, we create a second lower-case column called \texttt{time}.
+Time in ATLAS tracks is represented by 64-bit integers (type \texttt{long}) that specify time in milliseconds, starting from the beginning of 1970 (the UNIX epoch).
+This representation of time is called \texttt{POSIX} time and is usually specified in seconds, not milliseconds.
+
+Since about 1.6 billion seconds have passed since the beginning of 1970, current \texttt{POSIX} times in milliseconds cannot be represented by R's built-in 32-bit integers.
+A naive conversion results in truncation of out-of-range numbers leading to huge errors (dates many thousands of years in the future).
+
+R does not natively support 64-bit integers. One option is to use the bit64 package, which adds 64-bit integer support to R.
+
+A simpler solution is to convert the times to R's built in \texttt{double} data type (also called \texttt{numeric}), which uses a 64-bit floating point representation.
+This representation can represent integers with up to 16 digits without error; we only need 13 digits to represent the number of milliseconds since 1970, so the conversion is error free.
+We can also perform the conversion and then divide by 1000 so that times are represented in seconds, not milliseconds; this simplifies speed estimation.
+
+If second-resolution is accurate enough (it is for our purposes), the solution that we use is to divide times by 1000 to reduce the resolution from milliseconds to seconds and then to convert the time stamps to R integers.
+In the spirit of not destroying data, we create a second lower-case column called \texttt{time} to store this
 
@@
 \CommentTok{# divide by 1000, convert to integer, then convert to POSIXct}
-\NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(TIME }\OperatorTok{/}\StringTok{ }\DecValTok{1000}\NormalTok{)]}
+\NormalTok{data[, time }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(}
+  \KeywordTok{as.numeric}\NormalTok{(TIME) }\OperatorTok{/}\StringTok{ }\DecValTok{1000}
+\NormalTok{)]}
 \end{Highlighting}
@@
 We save this data to file for later use.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# load recurse}
-\KeywordTok{library}\NormalTok{(recurse)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 \begin{Highlighting}[]
-\CommentTok{# load libs}
+\CommentTok{# libs for data}
 \KeywordTok{library}\NormalTok{(data.table)}
 \KeywordTok{library}\NormalTok{(RSQLite)}
+\KeywordTok{library}\NormalTok{(atlastools)}
+
+\CommentTok{# libs for plotting}
 \KeywordTok{library}\NormalTok{(ggplot2)}
@@
 
+\CommentTok{# recursion analysis}
+\KeywordTok{library}\NormalTok{(recurse)}
+
 \CommentTok{# prepare a palette}
 \NormalTok{pal <-}\StringTok{ }\NormalTok{RColorBrewer}\OperatorTok{::}\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{4}\NormalTok{, }\StringTok{"Set1"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{install-atlastools-from-github.}{%
-\section{\texorpdfstring{Install \texttt{atlastools} from Github.}{Install atlastools from Github.}}\label{install-atlastools-from-github.}}
-
-\texttt{atlastools} is available from Github and is archived on Zenodo (Gupte \protect\hyperlink{ref-gupte2020a}{2020}).
-It can be installed using \texttt{remotes} or \texttt{devtools}. Here we use the \texttt{remotes} function \texttt{install\_github}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{install.packages}\NormalTok{(}\StringTok{"remotes"}\NormalTok{)}
-
-\CommentTok{# installation using remotes}
-\NormalTok{remotes}\OperatorTok{::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{# load recurse}
-\KeywordTok{library}\NormalTok{(recurse)}
-
 \CommentTok{# split the data}
@@
 \begin{Highlighting}[]
-\NormalTok{data_clean <-}\StringTok{ }\KeywordTok{fwrite}\NormalTok{(}\KeywordTok{rbindlist}\NormalTok{(data_smooth),}
+\KeywordTok{fwrite}\NormalTok{(}\KeywordTok{rbindlist}\NormalTok{(data_smooth),}
   \DataTypeTok{file =} \StringTok{"data/data_bat_smooth.csv"}

### 2021-02-11 docs/ms_atlas_preproc_main_text.tex
add validation ref

@@
 2.a; Listing 1).
-A bounding box filter does not require a geospatial representation such as a shapefile, and can help remove unreliable data from a tracking system that is less accurate beyond a certain range (e.g.
-ATLAS; Beardsworth et al.
-\textit{in prep.}).
+A bounding box filter does not require a geospatial representation such as a shapefile, and can help remove unreliable data from a tracking system that is less accurate beyond a certain range \citep[e.g.
+in ATLAS systems][]{beardsworth2021}.
 In some special cases, users may wish to remove positions \textit{inside} a bounding box, either because movement behaviour within an area is not the focus of a study, or because positions recorded within an area are known to be erroneous.
@@
 
-Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system (Beardsworth et al. \textit{in prep.}).
+Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system \citep{beardsworth2021}.
 GPS systems provide direct measures of location error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS]{ranacher2016}, while  reverse-GPS systems provide a similar estimate \citep[called Standard Deviation, SD][]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
@@
 6).
-We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 (Beardsworth et al.
-\textit{in prep.}; Bijleveld et al.
-\textit{in prep.}).
+We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 \citep[][; Bijleveld et al.
+\textit{in prep.}]{beardsworth2021}.
 Stops in the calibration track were recorded as waypoints using a handheld GPS device (Garmin Dakota 10) at each stop.

### 2021-02-11 docs/ms_atlas_preproc_main_text.tex
format listings

@@
 \usepackage{helvet}
+\usepackage[T1]{fontenc}
+\usepackage{inconsolata}%
 \usepackage{xfrac}
@@
   basicstyle=\small\ttfamily\color{black},
-%   keywordstyle=\bfseries\color{blue!40!black},
-%   commentstyle=\itshape\color{purple!40!black},
-%   identifierstyle=\color{green!40!black},
-%   stringstyle=\color{red}
-%   keywordstyle=\color{keywordstyleCol},      % keyword style
+  keywordstyle=\ttfamily,
   commentstyle=\color{commentstyleCol},   % comment style
-  stringstyle=\color{stringstyleCol},      % string literal style
-  literate=%
-   *{0}{{{\color{NumCol}0}}}1
-    {1}{{{\color{NumCol}1}}}1
-    {2}{{{\color{NumCol}2}}}1
-    {3}{{{\color{NumCol}3}}}1
-    {4}{{{\color{NumCol}4}}}1
-    {5}{{{\color{NumCol}5}}}1
-    {6}{{{\color{NumCol}6}}}1
-    {7}{{{\color{NumCol}7}}}1
-    {8}{{{\color{NumCol}8}}}1
-    {9}{{{\color{NumCol}9}}}1
+  stringstyle=\color{stringstyleCol}      % string literal style
 }
@@
 \\
-\textbf{2} Department of Coastal Systems, NIOZ Royal Netherlands Institute for Sea Research, The Netherlands.
+\textbf{2} Department of Coastal Systems, NIOZ Royal Netherlands Institute for Sea Research, 1790 AB, Den Burg, The Netherlands.
 \\
@@
     At these scales, location error can exceed the animal's step size, leading to mis-estimation of movement metrics such as speed.
-    % Sophisticated methods that estimate both the error and the movement metrics are not easily scalable to massive datasets, and can be difficult to use for both novice and veteran animal ecologists.
     `Cleaning' the data to reduce location errors prior to analyses is one of the main ways movement ecologists deal with noisy data, and has the advantage of being more scalable to massive datasets than more complex methods.
@@
     We demonstrate this pipeline on simulated movement data with location errors.
-    We further suggest how large volumes of cleaned data can be segmented-clustered into biologically meaningful ‘residence patches’.
-    We use calibration data to show how the pipeline improves its quality, and to verify that the residence patch synthesis accurately captures animal space-use.
+    We also show a case study of how large volumes of cleaned data can be quickly segmented-clustered into biologically meaningful ‘residence patches’.
+    We use calibration data to show how pre-processing improves its quality, and to verify that the residence patch synthesis accurately captures animal space-use.
     Finally, turning to tracking data from Egyptian fruit bats (\textit{Rousettus aegyptiacus}), we demonstrate the pre-processing pipeline and residence patch method in a fully worked out example.

### 2021-02-11 docs/ms_atlas_preproc_main_text.tex
add allert edits

@@
 We opted for a first-principles approach and selected as residence positions any locations with a residence time $>$ 5 minutes, reasoning that a flying animal stopping for $>$ 5 minutes at a location should plausibly indicate resource use or another interesting behaviour.
-This step retained 7,819 positions per bat on average (75.6\%) of the smoothed data; suggesting the extent to which roosting biases the data.
+This step retained 7,819 positions per bat on average (75.6\%) of the smoothed data, showing that bats are actually relatively stationary at the daily scale, and move only mostly between their roost and foraging sites (see Fig.
+7).
 

### 2021-02-11 R/05_calibration_data.R
add new script

@@ -1,3 +1,10 @@
+## ----install_atlastools-------------------------------------------------------
+install.packages("remotes")
+
+# installation using remotes
+remotes::install_github("pratikunterwegs/atlastools")
+
+
 ## -----------------------------------------------------------------------------
 # for data handling
 library(data.table)
@@ -13,13 +20,6 @@ library(patchwork)
 # making a colour palette
 pal <- RColorBrewer::brewer.pal(5, "Set1")
 pal[3] <- "seagreen"
-
-
-## ----install_atlastools-------------------------------------------------------
-install.packages("remotes")
-
-# installation using remotes
-remotes::install_github("pratikunterwegs/atlastools")
 
 
 ## -----------------------------------------------------------------------------

### 2021-02-11 05_calibration_data.Rmd
title head to full stop
style code
reorder section
add inset labels

@@
 
+## Install `atlastools` from Github
+
+`atlastools` is available from Github and is archived on Zenodo [@gupte2020a].
+It can be installed using `remotes` or `devtools`. Here we use the `remotes` function `install_github`.
+
+```{r install_atlastools, eval=FALSE}
+install.packages("remotes")
+
+# installation using remotes
+remotes::install_github("pratikunterwegs/atlastools")
+```
+
 ## Prepare libraries
@@
 pal[3] <- "seagreen"
-```
-
-## Install `atlastools` from Github.
-
-`atlastools` is available from Github and is archived on Zenodo [@gupte2020a].
-It can be installed using `remotes` or `devtools`. Here we use the `remotes` function `install_github`.
-
-```{r install_atlastools}
-install.packages("remotes")
-
-# installation using remotes
-remotes::install_github("pratikunterwegs/atlastools")
 ```
@@
   annotate(
+    geom = "text",
+    x = 653250,
+    y = 5906133 + 500,
+    label = "(b)",
+    # size = 10,
+    fontface = "bold"
+  ) +
+  annotate(
     geom = "rect",
@@
     fill = NA, col = "grey20"
+  ) +
+  annotate(
+    geom = "text",
+    y = 5901650 - 300,
+    x = 650600,
+    label = "(c)",
+    # size = 10,
+    fontface = "bold"
   ) +

### 2021-02-11 docs/ms_atlas_preproc_main_text.tex
add new tex and pdf
spelling uniformity

@@
     \item Modern, high-throughput animal tracking studies collect increasingly large volumes of data at very fine temporal scales.
-    At these scales, location error can exceed the animal's step size, leading to mis-estimation of movement metrics such as speed.
+    At these scales, location-error can exceed the animal's step size, leading to mis-estimation of movement metrics such as speed.
     `Cleaning' the data to reduce location errors prior to analyses is one of the main ways movement ecologists deal with noisy data, and has the advantage of being more scalable to massive datasets than more complex methods.
@@
     \item A pipeline for cleaning massive high-throughput datasets must balance ease of use, computational efficiency.
-    Data cleaning must also reject location errors discarding valid animal movements.
+    Data cleaning must also reject location errors without discarding valid animal movements.
     Another useful feature of a pre-processing pipeline is efficiently segmenting and clustering location data for statistical methods, while also being scalable to large datasets and robust to imperfect sampling.
@@
     \item To help with fast implementation, and to help standardise methods, we developed the \texttt{R} package \texttt{atlastools}, which we also introduce here.
-    Our pre-processing pipeline and \texttt{atlastools} can be used with any high-throughput animal movement data in which the high data volume combined with knowledge of the tracked individuals’ movement capacity can be used to reduce location errors.
+    Our pre-processing pipeline and \texttt{atlastools} can be used with any high-throughput animal movement data in which the high data-volume combined with knowledge of the tracked individuals’ movement capacity can be used to reduce location errors.
     The \texttt{atlastools} function is easy to use for beginners, while providing a template for further development.
@@
 Investigating fine-scale environmental and social drivers and the consequences of movement requires position data from many individuals at high temporal and spatial resolution relative to the phenomenon of interest (high-throughput tracking).
-High-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}; additionally, ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput tracking data at much lower costs.
+High-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}; additionally, ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput location data at much lower costs \citep{wikelski2007}.
 
 Although high-resolution tracking provides a massive amount of data on the path of a tracked animal, this data presents a two-fold challenge to ecologists.
-First, the location error of each position may approach or exceed the true step size of the animal compared to low-resolution tracking with the same measurement error.
+First, the location-error of each position may approach or exceed the true step size of the animal compared to low-resolution tracking with the same measurement error.
 This biases derived metrics such as speed and tortuosity \citep[see][]{ranacher2016, noonan2019, hurford2009, calenge2009}.
-Additionally, the absolute location errors around a position introduces uncertainty when studying the location of an animal relative to fixed landscape features (e.g.
+Additionally, the absolute location-error around a position introduces uncertainty when studying the location of an animal relative to fixed landscape features (e.g.
 roads) or mobile elements (e.g.
 other tracked individuals).
-Users have two main options to improve data quality: making inferences after modelling the system-specific location error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008, aspillaga2021}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
-The first approach may be limited by the animal movement models it can fit to the data \citep{fleming2014a, noonan2019, fleming2020}, and may be beyond the computational capacity of common hardware, leading users to prefer data cleaning instead.
+Users have two main options to improve data quality: making inferences after modelling the system-specific location-error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008, aspillaga2021}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
+The first approach may be limited by the animal movement models that can be fit to the data \citep{fleming2014a, noonan2019, fleming2020}, and may be beyond the computational capacity of common hardware, leading users to prefer data cleaning instead.
 When attempting data cleaning, the second challenge of high-throughput tracking reveals itself: the large number of observations themselves \citep{weiser2016, toledo2020}.
@@
 For instance, small-scale resource selection functions can easily infer spurious preference and avoidance effects when there is uncertainty about an animal's true position and movement \citep{visscher2006}.
-Ecologists recognise that tracking data are imperfect observations of the underlying process of animal movement, yet they implicitly consider cleaned data equivalent to be the ground-truth.
+Ecologists recognise that tracking data are imperfect observations of the underlying process of animal movement, yet they implicitly consider cleaned data equivalent to the ground-truth.
 This assumption is reflected in popular statistical methods in movement ecology such as Hidden Markov Models (HMMs) \citep{langrock2012}, stationary-phase identification methods \citep{patin2020a}, or step-selection functions (SSFs) \citep{barnett2008, signer2017, avgar2016}, which expect low location errors relative to real animal movement (i.e., a high signal-to-noise ratio).
@@
 While gross errors are often removed by positioning-system algorithms, ‘reasonable’ errors often remain to confront end users \citep{fischler1981, weiser2016, ranacher2016}.
-% Location errors may be exacerbated by the conditions of deployment, with systematic differences in location error among different landscape types  \citep[see][]{lewis2007, deon2002, frair2004}. 
+% Location errors may be exacerbated by the conditions of deployment, with systematic differences in location-error among different landscape types  \citep[see][]{lewis2007, deon2002, frair2004}. 
 % % For instance, the physical features of an animal’s habitat such as hills or woodland may impede the transmissions of radio-telemetry tags (Beardsworth et al.
@@
 Pre-processing protocols are not often reported in the literature, or may not be easily tractable for mainstream computing hardware and software.
-Furthermore, filtering out positions on their location error estimates may not be be appropriate for outliers which represent unrealistic movement but have low error measures \citep{weiser2016, ranacher2016}.
-This makes identifying and removing implausible locations from a track an important component of recovering true animal movement \citep{bjorneraas2010}.
-Even after removing outliers, a track may comprise of positions that are distributed around the true animal location \citep{noonan2019}.
+Furthermore, filtering out positions on their location-error estimates may not be appropriate for outliers which represent unrealistic movement but have low error measures \citep{weiser2016, ranacher2016}.
+This makes identifying and removing biologically implausible locations from a track an important component of recovering true animal movement \citep{bjorneraas2010}.
+Even after removing unrealistic movement, a track may comprise of positions that are distributed around the true animal location \citep{noonan2019}.
 The large data-volumes of high-throughput tracking allow for a neat solution: tracks can be `smoothed' to reduce small location errors that have remained undetected.
-This large data volume becomes a challenge when users seek to examine animal space-use in relation to relevant environmental covariates, such as the predictors of prolonged residence in an area \citep[see][]{bracis2018, aarts2008, fleming2014a,bijleveld2016, oudman2018, harel2016}.
-% First, statistical modelling using high-throughput data may suffer from pseudo-replication since both the response (e.g.
-step-length) and the predictors (e.g.
-resource landscape values) are spatio-temporally auto-correlated and hence non-independent \citep{}.
-% Second, fitting statistical models to datasets with many millions of observations may require significant innovation in mainstream statistical software \citep[e.g.
-][]{wood2015}, and users may instead want to reduce data volumes by thinning or clustering.
+This large data-volume becomes a challenge when users seek to examine animal space-use in relation to relevant environmental covariates, such as the predictors of prolonged residence in an area \citep[see][]{bracis2018, aarts2008, fleming2014a,bijleveld2016, oudman2018, harel2016}.
+To avoid issues of non-independent observations due to strong spatio-temporal autocorrelation, and to make datasets computationally tractable for mainstream statistical software, users may want to reduce data-volumes by thinning or clustering.
 
-Here, we present a hands-on guide to pre-processing high-throughput tracking data, and demonstrate a relatively simple data cleaning pipeline to prepare these data for subsequent analyses (see Fig.
-1).
+Here, we present a hands-on guide to pre-processing high-throughput tracking data, and demonstrate a relatively simple pipeline to prepare these data for subsequent analyses (see Fig. 1). 
 We take two important considerations into account, (1) that methods must be computationally efficient, and (2) that the pre-processing steps should be easily understood and reproduced.
@@
 \texttt{atlastools} functions are easy to use and understand, and rely on the efficiency of the \texttt{R} package \texttt{data.table} \citep{dowle2020}.
-% We begin by showing how the pipeline can clean simulated tracks to which we have artificially introduced location error \citep{gurarie2017}.
-Beginning with basic spatio-temporal filtering of positions, we cover filtering outliers from movement tracks, and reducing the effect of location error with a median smooth.
-Then, we suggest one solution to issues of computational tractability, which is to synthesise residence patches \citep[\textit{sensu}][]{bijleveld2016, oudman2018, barraquand2008} from clusters of spatio-temporally proximate positions, and so reveal important aspects of animal space use.
-Using calibration data from a manually transported ATLAS tag, we demonstrate how the residence patch method in \texttt{atlastools} accurately identifies areas of prolonged residence under real field conditions.
+% We begin by showing how the pipeline can clean simulated tracks to which we have artificially introduced location-error \citep{gurarie2017}.
+Beginning with basic spatio-temporal filtering of positions, we cover filtering outliers from movement tracks, and reducing the effect of location-error with a median smooth.
+Then, we suggest one solution to issues of computational tractability, which is to synthesise residence patches from clusters of spatio-temporally proximate positions, and so reveal important aspects of animal space use \citep[\textit{sensu}][]{bijleveld2016, oudman2018, barraquand2008}.
+Using calibration data from a manually transported ATLAS tag, we demonstrate how the residence patch segmentation-clustering method can be used to accurately identify areas of prolonged residence under real field conditions.
 Finally, we turn to data from Egyptian fruit bats (\textit{Rousettus aegyptiacus}) tracked in the Hula Valley, Israel, to show a fully worked out example of the pre-processing pipeline and the residence patch method.
@@
     The \texttt{atlastools} function that may be used to implement each pre-processing step is shown in the grey boxes underneath each step.
-    Popular \texttt{R} packages in movement ecology, or statistical methods are shown underneath possible analyses (yellow boxes).
+    Popular statistical methods are shown underneath possible analyses (yellow boxes).
     Users are strongly encouraged to visualise their data and scan it for location errors as they work through the pipeline, always asking the question, could the animal plausibly move this way?
@@
     %  (see \textsc{Spatio-Temporal Filtering}).
-    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by a system-specific error measure, or by the plausibility of associated movement metrics, such as speed and turning-angle \citep{seidel2018, calenge2009}.
+    \item Next, users should reduce gross location errors by removing unreliable positions based on a system-specific error measure, or by the plausibility of associated movement metrics, such as speed and turning-angle \citep{seidel2018, calenge2009}.
     %  (see \textsc{Filtering to Reduce Location Errors}).
@@
     At this stage, the data are ready for a number of popular statistical treatments such as Hidden Markov Model-based classification \citep{michelot2016,langrock2012}.
-    \item Finally, users wishing simple segmentation-clustering can classify their data into residence patches based on the movement ecology of their study system, after filtering out non-stationary positions (see \textsc{Synthesising Movement Tracks into Residence Patches}).
+    \item Finally, users wishing simple, efficient segmentation-clustering to identify areas of prolonged residence can classify their data into residence patches based on the movement ecology of their study species, after filtering out travelling segments (see \textsc{Segmenting and Clustering Movement Tracks into Residence Patches}).
 
@@
 In some special cases, users may wish to remove positions \textit{inside} a bounding box, either because movement behaviour within an area is not the focus of a study, or because positions recorded within an area are known to be erroneous.
-An example of the former is studies seeking to study transit behaviour between features which can be approximated by their bounding boxes. 
+An example of the former is studies of transit behaviour between features which can be approximated by their bounding boxes. 
 Instances of the latter are likely to be system specific, but are known from ATLAS systems (Bijleveld et al. \textit{in prep.}). 
@@
 shapefile, geopackage, or \texttt{sf}-object in \texttt{R}).
-The \texttt{atlastools} function \texttt{atl\_filter\_bounds} implements bounding box and explicit spatial filters and accepts X and Y coordinate ranges, an \texttt{sf}-polygon or multi-polygon object \citep{pebesma2018}, or any combination of the three to filter the data (Listing 1).
+The \texttt{atlastools} function \texttt{atl\_filter\_bounds} implements both bounding box and explicit spatial filters, and accepts X and Y coordinate ranges, an \texttt{sf}-polygon or multi-polygon object \citep{pebesma2018}, or any combination of the three to filter the data (Listing 1).
 % When both coordinate ranges and a polygon are provided, the data is first filtered by the ranges and then the polygon.
@@
     \textbf{(b)} 
-    Location error may affect single observations resulting in point outliers or `spikes' (grey crosses), or continuous subsets of a track, called a `prolonged spike' (dark crosses, top right), and both represent unrealistic movement.
+    Location-error may affect single observations resulting in point outliers or `spikes' (grey crosses), or continuous subsets of a track, called a `prolonged spike' (dark crosses, top right), and both represent unrealistic movement.
     While spikes can be removed by filtering out positions with high incoming and outgoing speeds and turning angle, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
@@
 Periods of poor tracking quality may result from system malfunctions, and users may also wish to exclude these data.
-Temporal filtering can exclude positions from such periods when data are expected to be unreliable for ecological inference, either due to abnormal movement behaviour or system-specific issues.  
-Temporal filters can be combined with spatial filters to find specific time-location combinations. 
-For example, users who want to study the response of individuals to the stress of capture can select for positions shortly after release and within a certain distance of the release location.
+Temporal filtering can exclude positions from intervals when data are expected to be unreliable for ecological inference, either due to abnormal movement behaviour or system-specific issues.  
+Temporal filters can be combined with spatial filters to select specific time-location combinations. 
+For example, users who want to study the response of individuals to the stress of capture can select positions shortly after release and within a certain distance of the release location.
 Users should apply filters in sequence rather than all at once, and visualise the output after each filtering step (`sanity checks').
-
 The atlastools function \texttt{atl\_filter\_covariates} allows convenient filtering of a dataset by any number of logical statements, including querying data within a spatio-temporal range (Listing 2).
@@
     Here, the first example shows how nighttime data can be retained using a predicate that determines whether the value of `hour' is between 6 and 18, and also within a range of X coordinates.
-    The second example retains ATLAS locations calculated using $>$ 3 base stations (\texttt{NBS}), with location error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.
+    The second example retains ATLAS locations calculated using $>$ 3 base stations (\texttt{NBS}), with location-error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.
     }]
@@
 Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system \citep{beardsworth2021}.
-GPS systems provide direct measures of location error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS]{ranacher2016}, while  reverse-GPS systems provide a similar estimate \citep[called Standard Deviation, SD][]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
+GPS systems provide direct measures of location-error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS]{ranacher2016}, while  reverse-GPS systems provide a similar estimate \citep[called Standard Deviation, SD;][]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
 Tracking data can also include indirect indicators of data quality.
-For instance, GPS systems' location error may be indicated indirectly by the number of satellites involved in the localisation.
-In reverse-GPS systems too the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
-% A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
+For instance, GPS systems' location-error may be indicated indirectly by the number of satellites involved in the localisation.
+In reverse-GPS systems too, the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
+% A location-error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
 % \begin{linenomath*}
@@
 An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 2.b).
-Point outliers are characterised by artificially high speeds between the outlier and the positions before and after \citep[called incoming and outgoing speed, respectively][]{bjorneraas2010}, lending a `spiky' appearance to the track.
+Point outliers are characterised by artificially high speeds between the outlier and the positions before and after \citep[called incoming and outgoing speed, respectively;][]{bjorneraas2010}, lending a `spiky' appearance to the track.
 Removing spikes is simple: remove positions with extreme incoming and outgoing speeds.
-Users must first define plausible upper limits for speed and turning angle for the study species \citep{calenge2009, seidel2018}.
-Here, it is important to remember that speed estimates are scale-dependent; high-throughput tracking typically overestimates the speed between positions where the animal is stationary or moving slowly due to small-scale location errors \citep{ranacher2016, noonan2019}.
+Users must first define plausible upper limits of the study species' speed \citep{calenge2009, seidel2018}.
+Here, it is important to remember that speed estimates are scale-dependent; high-throughput tracking typically overestimates the speed between positions where the animal is stationary or moving slowly, due to small-scale location errors \citep{ranacher2016, noonan2019}.
 Even after data with large location errors have been removed, it is advisable to begin with a liberal (high) speed threshold that excludes only the most unlikely of speeds.
 Estimates of maximum speed may not always be readily obtained for all species, and an alternative is to use a data-driven threshold such as the 95\textsuperscript{th} percentile of speeds from the track.
-Once a speed threshold $S$ has been chosen, positions with incoming \textit{and} outgoing speeds $\geq S$ may be identified as spikes and removed.
+Once a speed threshold $S$ has been chosen, positions with incoming \textit{and} outgoing speeds $> S$ may be identified as spikes and removed.
 
-Some species can realistically achieve speeds $\geq S$ in fast transit segments when assisted by their environment, such as birds with tailwinds, and a simple filter on incoming and outgoing speeds would exclude this valid data.
+Some species can realistically achieve speeds $> S$ in fast transit segments when assisted by their environment, such as birds with tailwinds, and a simple filter on incoming and outgoing speeds would exclude this valid data.
 To avoid removing valid, fast transit segments while still excluding spikes we suggest combining the speed filter with a filter on the turning angles of each position \citep[see][]{calenge2009}.
 This combined filter assumes that positions in high-throughput tracking with both high speeds and large turning angles are likely to be due to location errors, since most species are unable to turn sharply at high speed.
-Users can then remove those positions whose incoming and outgoing speeds are both $> S$ and $\theta > A$ (sharp, high-speed turns), where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
+Users can then remove those positions whose incoming and outgoing speeds are both $> S$, and where $\theta > A$ (sharp, high-speed turns), where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
+Many other track metrics may be used to identify implausible movement, and used to filter data \citep{seidel2018}.
 Spike removal can be implemented using the \texttt{atl\_filter\_covariates} function (Listing 3).
-Many other track metrics may be used to identify implausible movement and on which data may be filtered \citep{seidel2018}.
 
@@
 
-Sometimes entire subsets of the track may be affected by the same large-scale location error.
+Sometimes, entire subsets of the track may be affected by the same large-scale location-error.
 For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 2.b).
@@
 Track-subset algorithms are likely to be system-specific, and may be challenging to conceptualise or implement.
-In the case of prolonged spikes, one relatively simple solution is identifying the bounds and removing positions between them.
+In the case of prolonged spikes, one relatively simple solution is identifying the bounds of displaced segments, and removing positions between them.
 % We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
@@
 
-After filtering out large location errors may, the track may still look ‘spiky’ at small scales, and this is due to smaller location errors.
+After filtering out large location errors, the track may still look ‘spiky’ at small scales, and this is due to smaller location errors.
+These are caused by errors in sampling the true animal position, and are especially noticeable when the individual is stationary or moving slowly \citep{noonan2019}.
 These smaller errors are challenging to remove since their covariates (such as speed and turning angles) are within the expected range of movement behaviour for the study species. 
-The large data volumes of high-throughput tracking allow users to resolve this problem by smoothing the positions. 
+The large data-volumes of high-throughput tracking allow users to resolve this problem by smoothing the positions. 
 A ‘smooth’ works by approximating the value of an observation based on neighbouring values.
@@
 Smoothing does not change the number of observations, but does decouple the coordinates from some of their covariates.
-For instance, smoothing breaks the relationship between a coordinate and the location error estimate around it (\texttt{VARX}, \texttt{VARY}, and \texttt{SD} in ATLAS systems, or HDOP in GPS tracking).
+For instance, smoothing breaks the relationship between a coordinate and the location-error estimate around it (VARX, VARY, and SD in ATLAS systems, or HDOP in GPS tracking).
 This makes subsequent filtering on covariates of data quality unreliable, and smoothed data are unsuitable for use with methods that model location uncertainty \citep{noonan2019, fleming2014a, fleming2020, calabrese2016}.
@@
 3.b -- e), one drawback of using a large $K$ is that short, quick forays away from the main track are likely to be smoothed away, leading to a loss in detail of the individual's small-scale movement.
-Users must themselves judge how best to trade large-scale and small-scale accuracy, and choose $K$ accordingly.
+Users must themselves judge how best to balance large-scale and small-scale accuracy, and choose $K$ accordingly.
 One empirical way to compare $K$ is by calculating the root mean squared error (RMSE) for different $K$ on the same data.
@@
     \includegraphics[width=0.95\textwidth]{figures/fig_03_median_smooth.png}
-    \caption{Median smoothing position coordinates reduces small-scale location error in tracking data. 
-    \textbf{(a)} The goal of this step is to approximate the simulated canonical track (black line), given positions with small-scale error remaining from previous steps (green points). 
+    \caption{Median smoothing position coordinates reduces small-scale location-error in tracking data. 
+    \textbf{(a)} The goal of this step is to approximate the simulated canonical track (black line), given positions with small-scale error remaining from previous steps (green points; grey lines). 
     % The canonical track is shown `zoomed in' for better representation, and the missing data at the top of the track (line with no points) is from where the prolonged spike was earlier removed.
     Median smoothing the given position coordinates (green points, in \textbf{(a)}) over a moving window ($K$) of \textbf{(b)} 3, \textbf{(c)} 5, \textbf{(d)} 11, and \textbf{(e)} 21 positions respectively, yields approximations (purple lines) of the canonical track.
+    Grey lines show the track without smoothing.
     Users are cautioned that there is no correct $K$, and they must subjectively choose a $K$ which most usefully trades small-scale details of the track for large-scale accuracy.}
@@
 Thinning data need not compromise researchers' ability to answer scientific questions with them; for instance, proximity-based social interactions lasting 1 -- 2 minutes would still be detected on thinning from a sampling interval of 1 second to 1 minute.
-Indeed, some analyses require that temporal auto-correlation in the data be broken by subsampling the data to a lower resolution \citep[such as the estimation of home-ranges or step-selection functions][]{fleming2014a, dupke2017}.
-Added to the requirement of uniform sampling intervals by many methods in the field, evenly reducing data volumes is a worthwhile endeavour \citep[e.g.][]{fleming2014a, michelot2016, avgar2016}.
+Indeed, some analyses require that temporal auto-correlation in the data be broken by subsampling the data to a lower resolution \citep[such as the estimation of home-ranges or step-selection functions;][]{fleming2014a, dupke2017}.
+Added to the requirement of uniform sampling intervals by many methods in the field, evenly reducing data-volumes is a worthwhile endeavour \citep[e.g.][]{fleming2014a, michelot2016, avgar2016}.
 Two plausible approaches here are resampling and aggregation, and both approaches begin with identifying time-interval groups (e.g. of 1 minute).
-Resampling picks one position from each time-interval group.
-Aggregation involves computing the mean values of all covariates for positions within a time-interval group.
+Resampling picks one position from each time-interval group while aggregation involves computing the mean values of all covariates for positions within a time-interval group.
 Both approaches yield one position per time-interval group (Fig. 4.a).
-Categorical variables, such as the habitat type associated with each position can be aggregated using a suitable measure such as the mode.
+Categorical variables, such as the habitat type associated with each position, can be aggregated using a suitable measure such as the mode.
 We caution users that thinning causes an extensive loss of small-scale detail in the data, and should be used carefully.
 
+Both aggregation and resampling have their relative advantages. 
 The aggregation method is less sensitive to selecting point outliers by chance than resampling.
-When users want to account for location error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location error is important.
-Resampling directly propagates these errors, while durin aggregation,
-% In reverse-GPS systems systems the location error is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
-the location error around each coordinate provided by either GPS or reverse-GPS system can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
+However, when users want to account for location-error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location-error is important, and resampling
+directly propagates these errors without further processing.
+% In reverse-GPS systems systems the location-error is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
+In the aggregation method, the location-error around each coordinate provided by either GPS or reverse-GPS system can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
 \begin{linenomath*}
@@
 \end{linenomath*}
-Similarly, the overall location error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance:
+Similarly, the overall location-error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance:
 \begin{linenomath*}
@@
 
-Users may question why aggregation, which can obtain consensus positions over an interval, correctly propagate the location error, and also reduce data volumes, should not be used directly on the raw data.
+Users may question why thinning, which can obtain consensus positions over an interval and also reduce data-volumes should not be used directly on the raw data.
 %%
@@
 The mis-estimation of track metrics could have knock-on consequences for the implementation of subsequent filters based on detecting unrealistic movement.
-However, thinning before data-cleaning may have its place as a useful step before exploratory visualisation of the movement track, since reduced data volumes are easier to handle for plotting software.
+However, thinning before data-cleaning may have its place as a useful step before exploratory visualisation of the movement track, since reduced data-volumes are easier to handle for plotting software.
 %
@@
     \includegraphics[width=0.95\textwidth]{figures/fig_04_thinning.png}
-    \caption{\textbf{(a)} Aggregating a filtered and smoothed movement track (green points) preserves track structure while reducing data volume, but \textbf{(b)} aggregating before filtering gross location errors and unrealistic movement (grey points) leads to the persistence of large-scale errors (such as prolonged spikes).
+    \caption{\textbf{(a)} Aggregating a filtered and smoothed movement track (green points) preserves track structure while reducing data-volume, but \textbf{(b)} aggregating before filtering gross location errors and unrealistic movement (grey points) leads to the persistence of large-scale errors (such as prolonged spikes).
     The thinned track is shown as purple squares and lines in both panels.
@@
 An important and common analysis with animal tracking data is to link space-use with environmental covariates.
-This is difficult even with smoothed and thinned high-throughput data, as these may be too large for statistical packages, or have strong autocorrelation (Fig. 5).
+This is difficult even with smoothed and thinned high-throughput data, as these may be too large for statistical packages, or have strong autocorrelation.
 Users aiming for such analyses can benefit from segmenting and clustering the data into spatio-temporally independent `residence patches', and metrics such as the distance travelled within and between patches can help compare broad-scale individual movement strategies, or be linked with environmental data \citep{bijleveld2016, oudman2018, barraquand2008}.
-Common methods of segmentation-clustering are either not based in the biology of the tracked individual, or are not scalable to very large yet often gappy datasets \citep{patin2020a}, making a first-principles approach an essential part of a pre-processing pipeline.
-Furthermore, making the patch the unit of observation conveniently sidesteps pseudo-replication while and reduces computational requirements.
+Common methods of segmentation-clustering are either not based in the biology of the tracked individual, or are not scalable to very large, yet often gappy datasets \citep{patin2020a}, making a first-principles approach an essential part of a pre-processing pipeline.
+Furthermore, making the patch the unit of observation conveniently sidesteps pseudo-replication and reduces computational requirements.
 
 First, users should identify positions representing bouts of stationary behaviour, for instance on their speed or residence time \citep{bracis2018}.
-Patch identification assesses whether consecutive positions and the bouts they comprise are spatio-temporally independent, and clusters them together if they are not.
+Patch identification assesses whether consecutive stationary positions and the bouts they comprise are spatio-temporally independent, and clusters them together if they are not.
 The splitting and lumping of positions into bouts, and bouts into patches is based on simple user specified thresholds --- the distance and the time interval between positions (and bouts) beyond which they should be considered independent.
@@
 
-A cleaned movement track can be classified into residence patches using the function \texttt{atl\_res\_patch} (see Fig. 5.b).
-\texttt{atl\_res\_patch} requires three parameters: (1) the distance threshold between positions (called \texttt{buffer\_size}), (2) the distance threshold between clusters of positions (called \texttt{lim\_spat\_indep}), and (3) the time interval between clusters (called \texttt{lim\_time\_indep}).
+A cleaned movement track can be classified into residence patches using the function \texttt{atl\_res\_patch} (see Fig.
+5.b), which requires three parameters: (1) the distance threshold between positions (called \texttt{buffer\_size}), (2) the distance threshold between clusters of positions (called \texttt{lim\_spat\_indep}), and (3) the time interval between clusters (called \texttt{lim\_time\_indep}).
 Clusters formed of fewer than a minimum number of positions can be excluded.
 % The exclusion of clusters with few positions can help in removing bias due to short stops, but if such short stops are also of interest, they can be included by reducing the \texttt{min\_fixes} argument.
-Our residence patch algorithm is capable of correctly identifying clusters of related points from a movement track (Fig. 5).
+This residence patch algorithm is capable of correctly identifying clusters of related points from a movement track (Fig. 5).
 This includes clusters where the animal is relatively stationary (orange and green patches, Fig. 5.b), as well as clusters where the animal is moving slowly (blue patch, Fig. 5.b).
@@
 6).
-We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 \citep[][; Bijleveld et al.
+We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 \citep[][Bijleveld et al.
 \textit{in prep.}]{beardsworth2021}.
@@
     \caption{Pre-processing steps for WATLAS calibration data showing filtering on speed, median smoothing and thinning by aggregation, and making residence patches.
-    \textbf{(a)} Positions with incoming and outgoing speed $\geq$ 15 m/s are removed (grey crosses = removed, green points = retained).
-    Rectangles show the region expanded in panels (b) and (c).
+    \textbf{(a)} Positions with incoming and outgoing speed $>$ 15 m/s are removed (grey crosses = removed, green points = retained).
     \textbf{(b)} Expanded view of upper rectangle in (a) showing the raw data (grey crosses), the median smoothed positions (green circles; moving window $K$ = 5), and the smoothed track thinned by aggregation to a 30 second interval (purple squares).
@@
     \caption{Synthesising animal tracks into residence patches can reveal movement in relation to landscape features, prior exploration, and other individuals.
-    \textbf{(a)} Linear approximations of the paths (coloured lines) between residence patches (circles) of three Egyptian fruit bats (\textit{Rousettus aegyptiacus}), tracked over three nights in the Hula Valley, Israel.
-    Bat tracks are are shown as thin lines below the linear approximations, and colours show bat identity. The grey hexagon represents a roost site.
+    \textbf{(a)} Linear approximations of the paths (coloured straight lines) between residence patches (circles) of three Egyptian fruit bats (\textit{Rousettus aegyptiacus}), tracked over three nights in the Hula Valley, Israel.
+    Real bat tracks are are shown as thin lines below the linear approximations, and colours show bat identity. The grey hexagon represents a roost site.
     Black points represent known fruit trees.
     Background is shaded by elevation at 30 metre resolution.
-    \textbf{(b)} Spatial representations of an individual bat's residence patches (green polygons) can be used to study site-fidelity by examining overlaps between patches, or to study resource selection by inspecting overlaps with known resources (here, black circles show the location of clusters of fruit trees).
-    In addition, the linear approximation of movement between patches (blue paths in panel (a)) can be contrasted with the estimated real path between patches (green lines).
-    \textbf{(c)} Fine scale tracks (thin coloured lines), large scale movement (thick lines), residence patch polygons, and fruit tree locations show how high-throughput data can be used to study movement across scales.
+    \textbf{(b)} Spatial representations of an individual bat's residence patches (green polygons) can be used to study site-fidelity by examining overlaps between patches, or to study resource selection by inspecting overlaps with known resources such as fruit trees (black circles).
+    In addition, the linear approximation of movement between patches (straight green lines) can be contrasted with the estimated real path between patches (irregular green lines).
+    \textbf{(c)} Fine scale tracks (thin coloured lines), large-scale movement (thick lines), residence patch polygons, and fruit tree locations show how high-throughput data can be used to study movement across scales.
     Patches and lines are coloured by bat identity.}
@@
 
-Our guide anticipates high-throughput animal tracking data becoming increasingly common.
+Our guide anticipates that high-throughput animal tracking data will become increasingly common.
 A uniform pipeline and toolset for data cleaning promotes reproducibility and standardisation across studies, making comparative inferences more robust.
 The open-source \texttt{R} package \texttt{atlastools} serves as a starting point for methodological collaboration among movement ecologists.
-Efficient location error modelling approaches \citep{fleming2020,aspillaga2021} may eventually make data-cleaning optional.
-Yet cleaning tracking data even partially before modelling location error is faster than error-modelling on the full data, and the removal of large location errors may improve model fits.
-Thus we see our pipeline as complementary to these approaches \citep{fleming2014a,fleming2020}.
-Finally, we recognise that the diversity and complexity of animal movement and data collection techniques often requires bespoke pre-processing solutions.
+Efficient location-error modelling approaches \citep{fleming2020, aspillaga2021} may eventually make data-cleaning optional.
+Yet cleaning tracking data even partially before modelling location-error is faster than error-modelling on the full data, and the removal of large location errors may improve model fits.
+Thus we see our pipeline as complementary to these approaches \citep{fleming2014a, fleming2020}.
+Finally, we recognise that the diversity and complexity of animal movement and data collection techniques often requires system-specific, even bespoke, pre-processing solutions.
 Though the principles outlined here are readily generalised, users' requirements will eventually exceed the particular tools we provide.
 We see this as an incentive for more users to be involved in developing methods for their systems.
-We offer our pipeline and package as a foundation for system specific tools in the belief that simple, robust concepts are key to methods development that balances system-specificity and broad applicability.
+We offer our pipeline and package as a foundation for system-specific tools in the belief that simple, robust concepts are key to methods development that balances system-specificity and broad applicability.
 

### 2021-02-11 docs/ms_atlas_preproc_main_text.tex
new figure captions

@@
     \caption{Simulated movement data (black line), with three kinds of artificially added errors: (1) normally distributed small-scale error on each position, (2) large-scale error added to 0.5\% of positions, and (3) 300 consecutive positions displaced to simulate a gross distortion affecting a continuous subset of the track.
-    \textbf{(a)} Tracks can be quickly filtered by spatial bounds (dashed grey lines) to exclude broad regions.
-    \textbf{(b)} 
-    Location-error may affect single observations resulting in point outliers or `spikes' (grey crosses), or continuous subsets of a track, called a `prolonged spike' (dark crosses, top right), and both represent unrealistic movement.
-    While spikes can be removed by filtering out positions with high incoming and outgoing speeds and turning angle, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
-    These and other distortions that affect continuous track segments should be removed by conceptualising algorithms that find the bounds of the distortion and remove points in between them.
+    \textbf{(a)} Tracks can be quickly filtered by spatial bounds (dashed grey lines) to exclude broad regions (green = retained; grey = removed).
+    \textbf{(b)} Location-error may affect single observations resulting in point outliers or `spikes' (red crosses and track segments), or continuous subsets of a track, called a `prolonged spike' (purple crosses, top right), and both represent unrealistic movement.
+    \textbf{(c)} Histograms of speed for the track (grey = small-scale errors, red = spikes), and the prolonged spike (purple).
+    While spikes can be removed by filtering out positions with high incoming and outgoing speeds and turning angle, prolonged spikes cannot be removed in this way, and should be resolved by conceptualising algorithms that find the bounds of the distortion instead.
     Users should frequently check the outputs of such algorithms to avoid rejecting valid data.
@@
 Users must themselves judge how best to balance large-scale and small-scale accuracy, and choose $K$ accordingly.
-One empirical way to compare $K$ is by calculating the root mean squared error (RMSE) for different $K$ on the same data.
+% One empirical way to compare $K$ is by calculating the root mean squared error (RMSE) for different $K$ on the same data.
 % In general, large $K$s lead to unrealistic tracks, both because the first and last $(K - 1) / 2$ positions are handled as special cases, and because the local median approaches the global median at larger $K$.
@@
     \textbf{(c)} Thinning before data cleaning can lead to significant misestimations of essential movement metrics such as speed at lower intervals.
-    Boxplots show the median and interquartile ranges for speed estimates of tracks aggregated over intervals of 3, 10, 30, and 120 seconds, but without the removal of location errors.
+    Boxplots show the median and interquartile ranges for speed estimates of tracks aggregated over intervals of 3, 10, 30, and 120 seconds (grey = thinning without removing location errors; green = thinning after filtering and smoothing).
     For comparison, the median and 95\textsuperscript{th} percentile of speed of the canonical track are shown as solid and dashed horizontal lines, respectively.
-    Green circles and associated error bars show, in contrast, the median ($\pm$ standard deviation) speed for the same aggregation interval, but after removing large- and small-scale location errors (median smooth $K$ = 11).
     The unfiltered data would have to be thinned to \sfrac{1}{30}\textsuperscript{th} volume to correctly estimate median speed.}
@@
     Regions are shaded by the temporal bounds of each residence patch.
-    The arguments passed to \texttt{atl\_res\_patch} determine the clustering, and can be adjusted to get a result that fits the study system.
+    The arguments passed to \texttt{atl\_res\_patch} determine the clustering, and can be adjusted to fit the study system.
     Users are cautioned that there are no `correct' arguments, and the best guide is the biology of the tracked individual.}
@@
     \textbf{(a)} Positions with incoming and outgoing speed $>$ 15 m/s are removed (grey crosses = removed, green points = retained).
-    \textbf{(b)} Expanded view of upper rectangle in (a) showing the raw data (grey crosses), the median smoothed positions (green circles; moving window $K$ = 5), and the smoothed track thinned by aggregation to a 30 second interval (purple squares).
+    \textbf{(b)} Raw data (grey crosses), median smoothed positions (green circles; moving window $K$ = 5), and the smoothed track thinned by aggregation to a 30 second interval (purple squares).
     Square size corresponds to the number of positions used to calculate the averaged position during thinning.
-    \textbf{(c)} Classifying thinned data into residence patches yields robust estimates of the duration of known stops. The island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) is shown in beige.
-    Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple triangles).
-    The algorithm identified all areas with prolonged residence, including those which we had not intended to be recorded, such as stops at the field station (n = 12; green polygon over red triangles).
-    The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).}
+    \textbf{(c)} Clustering thinned data into residence patches (green polygons) yields robust estimates of the location of known stops (purple triangles).
+    The algorithm identified all areas with prolonged residence, including those which we had not intended to be recorded, such as stops at the field station (n = 12; red triangle).
+    The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step; one of these is shown here (purple triangle without green polygon).}
     \label{fig:figure_calibration}

### 2021-02-11 R/05_calibration_data.R
edits to script

@@ -1,8 +1,8 @@
-## ----install_atlastools-------------------------------------------------------
-install.packages("remotes")
-
-# installation using remotes
-remotes::install_github("pratikunterwegs/atlastools")
+## ----install_atlastools, eval=FALSE-------------------------------------------
+## install.packages("remotes")
+## 
+## # installation using remotes
+## remotes::install_github("pratikunterwegs/atlastools")
 
 
 ## -----------------------------------------------------------------------------
@@ -156,10 +156,26 @@ fig_speed_outliers <-
     fill = NA, col = "grey20"
   ) +
   annotate(
+    geom = "text",
+    x = 653250,
+    y = 5906133 + 500,
+    label = "(b)",
+    # size = 10,
+    fontface = "bold"
+  ) +
+  annotate(
     geom = "rect",
     ymin = 5901650, ymax = 5903100,
     xmin = 650000, xmax = 650600,
     fill = NA, col = "grey20"
+  ) +
+  annotate(
+    geom = "text",
+    y = 5901650 - 300,
+    x = 650600,
+    label = "(c)",
+    # size = 10,
+    fontface = "bold"
   ) +
   scale_color_manual(values = c("grey", pal[3])) +
   ggthemes::theme_few() +

### 2021-02-11 R/_03_smoothing_tracks.R

@@ -184,8 +184,8 @@ figure_median_smooth <-
     fontface = "bold"
   ) +
   scale_colour_distiller(
-    palette = "RdPu", direction = 1,
-    values = c(-1, 1)
+    palette = "BuPu", direction = 1,
+    values = c(-2, 1)
   ) +
   ggthemes::theme_few() +
   theme(

### 2021-02-11 docs/ms_atlas_preproc_supplement.tex
add new tex and pdf
add pdfs and tex

@@
 \usepackage{helvet}
+\usepackage{inconsolata}
 % \IfFontExistsTF{Arial}{\setsansfont[]{Arial}}{}
@@
 
+\hypertarget{install-atlastools-from-github}{%
+\section{\texorpdfstring{Install \texttt{atlastools} from Github}{Install atlastools from Github}}\label{install-atlastools-from-github}}
+
+\texttt{atlastools} is available from Github and is archived on Zenodo (Gupte \protect\hyperlink{ref-gupte2020a}{2020}).
+It can be installed using \texttt{remotes} or \texttt{devtools}. Here we use the \texttt{remotes} function \texttt{install\_github}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{install.packages}\NormalTok{(}\StringTok{"remotes"}\NormalTok{)}
+
+\CommentTok{# installation using remotes}
+\NormalTok{remotes}\OperatorTok{::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
 \hypertarget{prepare-libraries}{%
@@
 \NormalTok{pal[}\DecValTok{3}\NormalTok{] <-}\StringTok{ "seagreen"}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{install-atlastools-from-github}{%
-\section{\texorpdfstring{Install \texttt{atlastools} from Github}{Install atlastools from Github}}\label{install-atlastools-from-github}}
-
-\texttt{atlastools} is available from Github and is archived on Zenodo (Gupte \protect\hyperlink{ref-gupte2020a}{2020}).
-It can be installed using \texttt{remotes} or \texttt{devtools}. Here we use the \texttt{remotes} function \texttt{install\_github}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{install.packages}\NormalTok{(}\StringTok{"remotes"}\NormalTok{)}
-
-\CommentTok{# installation using remotes}
-\NormalTok{remotes}\OperatorTok{::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"pratikunterwegs/atlastools"}\NormalTok{)}
 \end{Highlighting}
@@
 The island of Griend (53.25\(^{\circ}\)N, 5.25\(^{\circ}\)E) is shown in beige.
-Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
-However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without crossed-squares).
-The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).
+Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple triangles).
+However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without triangles).
+The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (triangle without green polygon shows one of these).
 The area shown is the lower rectangle in Fig.

### 2021-02-11 05_calibration_data.Rmd
minor edit

@@
 The island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) is shown in beige.
-Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
-However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without crossed-squares).
-The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).
+Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple triangles).
+However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without triangles).
+The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (triangle without green polygon shows one of these).
 The area shown is the lower rectangle in Fig.

### 2021-02-11 _03_smoothing_tracks.Rmd

@@
   scale_colour_distiller(
-    palette = "RdPu", direction = 1,
-    values = c(-1, 1)
+    palette = "BuPu", direction = 1,
+    values = c(-2, 1)
   ) +

### 2021-02-11 docs/ms_atlas_preproc_main_text.tex

@@
     The output is a \texttt{data.table}, which must be saved as an object (here, \texttt{filtered\_data}).}]
-filtered_data <- atl_filter_bounds(data = data,
+filtered_data <- atl_filter_bounds(
+                        data = data,
                         x = "X", y = "Y",
@@
                         sf_polygon = your_polygon,
-                    remove_inside = FALSE)
+                        remove_inside = FALSE
+                    )
 \end{lstlisting}
@@
     }]
-night_data <- atl_filter_covariates(data = dataset,
-                filters = c("!inrange(hour, 6, 18)",
-                            "between(x, x_min, x_max)"))
+night_data <- atl_filter_covariates(
+                    data = dataset,
+                    filters = c(
+                        "!inrange(hour, 6, 18)",
+                        "between(x, x_min, x_max)"
+                    )
+                )
 
-filtered_data <- atl_filter_covariates(data = data,
-                    filters = c("NBS > 3",
+filtered_data <- atl_filter_covariates(
+                        data = data,
+                        filters = c(
+                            "NBS > 3",
                             "SD < 100",
-                                "between(day, 5, 8)"))                            
+                            "between(day, 5, 8)"
+                        )
+                    )                            
 \end{lstlisting}
@@
     The filter step only retains positions with speeds below the speed threshold $S$ \textit{or} angles above the turning angle threshold $\theta$, i.e., positions where the animal is slow but makes sharp turns, and data where the animal moves quickly in a relatively straight line.}]
-data$speed_in <- atl_get_speed(data,
+data$speed_in <- atl_get_speed(
+                    data,
                     x = "x", y = "y",
-                time = "time", type = c("in"))
+                    time = "time", 
+                    type = c("in")
+                )
 
-data$angle <- atl_turning_angle(data,
-                  x = "x", y = "y", time = "time")
+data$angle <- atl_turning_angle(
+                    data,
+                    x = "x", y = "y", 
+                    time = "time"
+                )
 
-filtered_data <- atl_filter_covariates(data = data,
-        filters = c("(speed_in < S & speed_out < S) | angle < A"))
+filtered_data <- atl_filter_covariates(
+                    data = data,
+                    filters = c(
+                        "(speed_in < S & speed_out < S) | angle < A"
+                    )
+                )
 \end{lstlisting}
@@
 The median smoothed value of the X coordinate, for instance, is
+%
 \begin{linenomath*}
@@
 \end{linenomath*}
+%
 Users can apply a median smooth with an appropriate $K$ independently to the X and Y coordinates of a movement track to smooth it (see Fig. 3.a -- e).
-% The first and last $(K - 1)/2$ values for a moving window size of $K$ are assigned a value using Tukey's (1977) end-point smoothing rule as implemented in \texttt{R} \citep{hardle1995}.
 The median smooth is robust to even very large temporal and spatial gaps, and does not interpolate between positions when data are missing. 
@@
 % One empirical way to compare $K$ is by calculating the root mean squared error (RMSE) for different $K$ on the same data.
-% In general, large $K$s lead to unrealistic tracks, both because the first and last $(K - 1) / 2$ positions are handled as special cases, and because the local median approaches the global median at larger $K$.
 Median smoothing is provided by the \texttt{atlastools} function \texttt{atl\_median\_smooth}, with the only option being the moving window size, which must be an odd integer (Listing 4).
@@
     Larger values of $K$ yield smoother tracks, but $K$ should always be some orders of magnitude lower than the number of observations.}]
-atl_median_smooth(data = track_data,
+atl_median_smooth(
+    data = track_data,
     x = "x", y = "y",
     time = "time",
-                  moving_window = 5)
+    moving_window = 5
+)
 \end{lstlisting}
@@
 Both methods return a dataset with one rows per time-interval.}]
-thinned_data <- atl_thin_data(data,
+thinned_data <- atl_thin_data(
+                    data,
                     interval = 60,
                     id_columns = c("animal_id"),
-                    method = "aggregate")
+                    method = "aggregate"
+                )
 \end{lstlisting}
@@
 The \texttt{atl\_patch\_summary} function is used to access the classified patch in one of three ways, here using the \texttt{summary} option which returns a table of patch-wise summary statistics.}]
-patches <- atl_res_patch(data = track_data,
+patches <- atl_res_patch(
+                data = track_data,
                 buffer_radius = 10,
@@
                 summary_variables = c("speed"),
-                summary_functions = c("mean", "sd"))
+                summary_functions = c("mean", "sd")
+            )
 
-patch_summary <- atl_patch_summary(patch_data = patches,
+patch_summary <- atl_patch_summary(
+                    patch_data = patches,
                     which_data = "summary",
-                    buffer_radius = 10)
+                    buffer_radius = 10
+                )
 \end{lstlisting}

### 2021-02-15 05_calibration_data.Rmd
add ref to christine and link to data
add note on dt syntx

@@
 Here we show how the residence patch method [@barraquand2008; @bijleveld2016; @oudman2018] accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
+These data can be accessed from the `data` folder at this link: https://doi.org/10.5281/zenodo.4287462.
+These data are more fully reported in [@beardsworth2021].
 
@@
 
+### A Note on `:=` {-}
+
+The `atlastools` package is based on `data.table`, to be fast and efficient [@dowle2020].
+A key feature is modification in place, where data is changed without making a copy.
+This is already implemented in `R` and will be familiar to many users as `data_frame$column_name <- values`.
+
+The `data.table` way of writing this assignment would be `data_frame[, column_name := values]`.
+We use this syntax throughout, as it provides many useful shortcuts, such as multiple assignment:
+
+`data_frame[, c("col_a", "col_b) := list(values_a, values_b)]`
+
+Users can use this special syntax, and will find it convenient with practice, but there are _no_ cases where users _must_ use the `data.table` syntax, and can simply treat the data as a regular `data.frame`.
+However, users are advised to convert their `data.frame` to a `data.table` using the function `data.table::setDT()`.
+
 ## Prepare libraries
@@
 First we access the data from a local file using the `data.table` package [@dowle2020].
+We look at the first few rows, using `head()`.
 We then visualise the raw data.
@@
 data_raw <- copy(data)
+
+# see raw data
+head(data_raw)
 ```

### 2021-02-15 05_calibration_data.Rmd
add some explanation for summary vars

@@
 
-An indicator of individual residence at or near a position can be useful when attempting to identify residence patches. Positions can be filtered on a metric such as residence time [@bracis2018].
+First, we filter data where we know the animal spent some time at or near a position, as this is the first step to identify residence patches. 
+One way of doing this is to filter out positions with speeds above which the animal is likely to be in transit, but a better way of doing this by filtering on a metric such as residence time [@bracis2018].
 
@@
 ```
+
+### A note on summary statistics {-}
+
+Users specifying a `summary_variable` should make sure that the variable for which they want a summary statistic is present in the data.
+For instance, requesting mean speed by passing `summary_variable = "speed"` and `summary_function = "mean"` to `atl_res_patch`, should make sure that their data includes a column called `speed`.
 

### 2021-02-15 docs/ms_atlas_preproc_main_text.tex
update tex

@@
 Aparajitha Ramesh and Franjo Weissing for discussions on error propagation;
-Geert Aarts, Jacob RL Gismann, Evy Gobbens for feedback that improved the manuscript; 
+Geert Aarts, Jacob RL Gismann, Evy Gobbens, and Roos Kentie for feedback that improved the manuscript; 
 members of the Modelling Adaptive Response Mechanisms Group (Weissing Lab), and the Theoretical Biology department at the University of Groningen for helpful discussions on \texttt{atlastools} and the manuscript.
-All authors thank the attendees of ATLAS workshops held in May and June 2020 at the Hebrew University of Jerusalem for helpful comments on the pipeline and \texttt{atlastools}.
+We thank the many volunteers, students, and NIOZ staff involved in operating the WATLAS tracking system, and most importantly Frank van Maarseveen, Bas Denissen and Anne Dekinga.
+We also thank the attendees of ATLAS workshops held in May and June 2020 at the Hebrew University of Jerusalem for helpful comments on the pipeline and \texttt{atlastools}.
+This work was partly funded by the Dutch Research Council grant VI.Veni.192.051 awarded to AIB.
 

### 2021-02-15 docs/ms_atlas_preproc_supplement.tex
update tex

@@
 \author{Pratik R. Gupte \and Christine E. Beardsworth \and Orr Spiegel \and Emmanuel Lourie \and Sivan Toledo \and Ran Nathan \and Allert I. Bijleveld}
-\date{2021-02-11}
+\date{2021-02-15}
 
@@
 \protect\hyperlink{ref-oudman2018}{2018}) accurately estimates the duration of known stops in a track collected as part of a calibration exercise in the Wadden Sea.
+These data can be accessed from the \texttt{data} folder at this link: \url{https://doi.org/10.5281/zenodo.4287462}.
+These data are more fully reported in (Beardsworth et al.
+\protect\hyperlink{ref-beardsworth2021}{2021}).
 
@@
 
+\hypertarget{a-note-on}{%
+\subsection*{\texorpdfstring{A Note on \texttt{:=}}{A Note on :=}}\label{a-note-on}}
+\addcontentsline{toc}{subsection}{A Note on \texttt{:=}}
+
+The \texttt{atlastools} package is based on \texttt{data.table}, to be fast and efficient (Dowle and Srinivasan \protect\hyperlink{ref-dowle2020}{2020}).
+A key feature is modification in place, where data is changed without making a copy.
+This is already implemented in \texttt{R} and will be familiar to many users as \texttt{data\_frame\$column\_name\ \textless{}-\ values}.
+
+The \texttt{data.table} way of writing this assignment would be \texttt{data\_frame{[},\ column\_name\ :=\ values{]}}.
+We use this syntax throughout, as it provides many useful shortcuts, such as multiple assignment:
+
+\texttt{data\_frame{[},\ c("col\_a",\ "col\_b)\ :=\ list(values\_a,\ values\_b){]}}
+
+Users can use this special syntax, and will find it convenient with practice, but there are \emph{no} cases where users \emph{must} use the \texttt{data.table} syntax, and can simply treat the data as a regular \texttt{data.frame}.
+However, users are advised to convert their \texttt{data.frame} to a \texttt{data.table} using the function \texttt{data.table::setDT()}.
+
 \hypertarget{prepare-libraries}{%
@@
 First we access the data from a local file using the \texttt{data.table} package (Dowle and Srinivasan \protect\hyperlink{ref-dowle2020}{2020}).
+We look at the first few rows, using \texttt{head()}.
 We then visualise the raw data.
@@
 \NormalTok{data_raw <-}\StringTok{ }\KeywordTok{copy}\NormalTok{(data)}
+
+\CommentTok{# see raw data}
+\KeywordTok{head}\NormalTok{(data_raw)}
 \end{Highlighting}
@@
 
-An indicator of individual residence at or near a position can be useful when attempting to identify residence patches.
-Positions can be filtered on a metric such as residence time (Bracis, Bildstein, and Mueller \protect\hyperlink{ref-bracis2018}{2018}).
+First, we filter data where we know the animal spent some time at or near a position, as this is the first step to identify residence patches.
+One way of doing this is to filter out positions with speeds above which the animal is likely to be in transit, but a better way of doing this by filtering on a metric such as residence time (Bracis, Bildstein, and Mueller \protect\hyperlink{ref-bracis2018}{2018}).
 
@@
 
+\hypertarget{a-note-on-summary-statistics}{%
+\subsection*{A note on summary statistics}\label{a-note-on-summary-statistics}}
+\addcontentsline{toc}{subsection}{A note on summary statistics}
+
+Users specifying a \texttt{summary\_variable} should make sure that the variable for which they want a summary statistic is present in the data.
+For instance, requesting mean speed by passing \texttt{summary\_variable\ =\ "speed"} and \texttt{summary\_function\ =\ "mean"} to \texttt{atl\_res\_patch}, should make sure that their data includes a column called \texttt{speed}.
+
 \hypertarget{get-spatial-and-summary-objects}{%
@@
 \end{Shaded}
+
+At this stage, users have successfully pre-processed their data from raw positions to residence patches.
+Residence patches are essentially \texttt{sf} objects and can be visualised using the \texttt{sf} method for \texttt{plot}; for instance \texttt{plot(patch\_sf\_data)}.
+Further sections reproduce the analyses in the main manuscript.
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
 
@@
 
-\hypertarget{plot-figure-7}{%
-\section{Plot figure 7}\label{plot-figure-7}}
+\hypertarget{main-text-figure-6}{%
+\section{Main text Figure 6}\label{main-text-figure-6}}
 
@@
 
+\leavevmode\hypertarget{ref-beardsworth2021}{}%
+Beardsworth, Christine E., Evy Gobbens, Frank van Maarseveen, Bas Denissen, Anne Dekinga, Ran Nathan, Sivan Toledo, and Allert I.
+Bijleveld.
+2021.
+``Validating a High-Throughput Tracking System: ATLAS as a Regional-Scale Alternative to GPS.'' \emph{bioRxiv}, February, 2021.02.09.430514.
+\url{https://doi.org/10.1101/2021.02.09.430514}.
+
 \leavevmode\hypertarget{ref-bijleveld2016}{}%

### 2021-02-16 05_calibration_data.Rmd
add breaker for patch emthod

@@
 
+At this stage, users have successfully pre-processed their data from raw positions to residence patches.
+Residence patches are essentially `sf` objects and can be visualised using the `sf` method for `plot`; for instance `plot(patch_sf_data)`.
+Further sections reproduce the analyses in the main manuscript.
+
+---
+
 ### Prepare to plot data
@@
 
-## Plot figure 7
+## Main text Figure 6
 

### 2021-02-16 docs/ms_atlas_preproc_main_text.tex

@@
 2.a).
-To demonstrate the residence patch method, we chose to simulate three independent rotational/advective correlated velocity movement (RACVM) tracks of 500 positions each ($\omega$ = 7, initial velocity = 0.1, $\mu$ = 0; see Gurarie et al.
+To demonstrate the residence patch method, we chose to simulate three independent rotational-advective correlated velocity movement (RACVM) tracks of 500 positions each ($\omega$ = 7, initial velocity = 0.1, $\mu$ = 0; see Gurarie et al.
 2017), and connected them together with a roughly linear path (see Fig.
@@
 For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 2.b).
-These cannot be corrected by targeted removal of individual positions, as in Bjorneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
+These cannot be corrected by targeted removal of individual positions, as in Bjørneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
 Since filtering individual positions will not suffice, algorithms to correct such errors must take a track-level view, and target the displaced sequence overall.
@@
 The aggregation method is less sensitive to selecting point outliers by chance than resampling.
-However, when users want to account for location-error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location-error is important, and resampling
-directly propagates these errors without further processing.
+However, when users want to account for location-error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location-error is important, and resampling directly propagates these errors without further processing.
 % In reverse-GPS systems systems the location-error is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
@@
 %%
-We caution that thinning prior to excluding unrealistic movement and smoothing (Fig 4.b) can lead to preserving artefacts in the data, and estimates of essential metrics --- such as speed --- that are substantially different from the true value (Noonan et al.
-2019; see Fig.
-4.c).
+We caution that thinning prior to excluding unrealistic movement and smoothing (Fig 4.b) can lead to preserving artefacts in the data, and estimates of essential metrics --- such as speed --- that are substantially different from the true value \citep[see Fig.
+4.c;][]{noonan2019}.
 In our example for instance, the data with errors included would have to be thinned to \sfrac{1}{30}\textsuperscript{th} of its volume for the median speed to be correctly estimated; an undesirable step if the aim is fine-scale tracking (Fig.
@@
 % If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
-Grouping variables' names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 5).
+Grouping variable names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 5).
 

### 2021-02-16 05_calibration_data.Rmd
show data head

@@
 
-```{r}
+```{r eval=TRUE}
 # for data handling
@@
 
-```{r}
+```{r, eval=TRUE}
 # read and plot example data

### 2021-02-16 R/05_calibration_data.R
script data head

@@ -5,7 +5,7 @@
 ## remotes::install_github("pratikunterwegs/atlastools")
 
 
-## -----------------------------------------------------------------------------
+## ----eval=TRUE----------------------------------------------------------------
 # for data handling
 library(data.table)
 library(atlastools)
@@ -22,7 +22,7 @@ pal <- RColorBrewer::brewer.pal(5, "Set1")
 pal[3] <- "seagreen"
 
 
-## -----------------------------------------------------------------------------
+## ---- eval=TRUE---------------------------------------------------------------
 # read and plot example data
 data <- fread("data/atlas1060_allTrials_annotated.csv")
 data_raw <- copy(data)

### 2021-02-16 render_books.sh
vanilla render

@@ -11,7 +11,7 @@ cp figures/fig_0*.png atlas-manuscript/figures
 # style files
 Rscript --vanilla --slave -e 'styler::style_dir(".", filetype = "Rmd", recursive = FALSE)'
 
-Rscript --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2")'
+Rscript --vanilla --slave -e 'bookdown::render_book("index.rmd", "bookdown::pdf_document2")'
 
 # to render source files
 Rscript --slave -e 'lapply(list.files(pattern = "\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'

### 2021-02-17 _03_smoothing_tracks.Rmd
edits to figure 3

@@
 # smooth the data over four K values
-list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
+list_of_smooths <- lapply(c(5, 11, 21, 101), function(K) {
   data_copy <- copy(data_errors)
@@
 ```{r}
-fwrite(list_of_smooths[[3]], file = "data/data_smooth.csv")
+fwrite(list_of_smooths[[2]], file = "data/data_smooth.csv")
 ```
@@
       group = window_size,
-      col = window_size
+      col = factor(window_size)
     ),
     show.legend = F,
-    lwd = 0.3
+    lwd = 0.35
   ) +
@@
     expand = F,
-    ylim = c(0.6, 0.85),
+    ylim = c(0.6, 0.83),
     xlim = c(NA, 2.3),
@@
     x = c(0.75, seq(1.1, 2, length.out = 4)),
-    y = 0.82,
+    y = 0.81,
     label = sprintf("(%s)", letters[seq(5)]),
@@
   ) +
-  scale_colour_distiller(
-    palette = "BuPu", direction = 1,
-    values = c(-2, 1)
+  scale_colour_manual(
+    values = c(colorspace::sequential_hcl(3, l = 40, palette = "PuBu",
+                                          rev = T), "sienna")
   ) +

### 2021-02-17 _03_smoothing_tracks.Rmd
edits to smoothing values
edit thinning figure

@@
 
-We apply a median smooth with four different values of $K$, the moving window size --- 3, 5, 11, and 21 positions.
+We apply a median smooth with four different values of $K$, the moving window size --- 5, 11, 21 and 101 positions.
 $K$ must be an odd number, but apart from that there is no correct magnitude of $K$.
@@
 We save the 11 point smoothed data.
-
-![](figures/fig_04_median_smooth.png)
 
@@
     fill = pal[3],
+    size = 0.3,
     show.legend = F,
@@
     outlier.size = 0.2,
-    position = position_nudge(x = 0.25)
+    position = position_nudge(x = 0.15)
   ) +
   geom_boxplot(aes(factor(interval), 1 + speed),
-    position = position_nudge(x = -0.25, ),
+    position = position_nudge(x = -0.15, ),
     fill = "grey",
+    size = 0.3,
     alpha = 0.5,
@@
     fig_agg_data_smooth, fig_agg_data_error, fig_agg_speed,
-    design = "AABBC"
+    design = "AABBCC"
   ) +

### 2021-02-17 _03_smoothing_tracks.Rmd
edit col fig 3

@@
   scale_colour_manual(
-    values = c(colorspace::sequential_hcl(3, l = 40, palette = "PuBu",
-                                          rev = T), "sienna")
+    values = c(colorspace::sequential_hcl(3,
+      l = 40, palette = "PuBu",
+      rev = T
+    ), "darkred")
   ) +

### 2021-02-17 R/_03_smoothing_tracks.R
update fig 3 script

@@ -23,7 +23,7 @@ data_errors[, window_size := 0]
 
 ## -----------------------------------------------------------------------------
 # smooth the data over four K values
-list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
+list_of_smooths <- lapply(c(5, 11, 21, 101), function(K) {
   data_copy <- copy(data_errors)
 
   data_copy <- atl_median_smooth(
@@ -39,7 +39,7 @@ list_of_smooths <- lapply(c(3, 5, 11, 21), function(K) {
 
 
 ## -----------------------------------------------------------------------------
-fwrite(list_of_smooths[[3]], file = "data/data_smooth.csv")
+fwrite(list_of_smooths[[2]], file = "data/data_smooth.csv")
 
 
 ## -----------------------------------------------------------------------------
@@ -165,27 +165,29 @@ figure_median_smooth <-
     data = data_plot,
     aes(x, y,
       group = window_size,
-      col = window_size
+      col = factor(window_size)
     ),
     show.legend = F,
-    lwd = 0.3
+    lwd = 0.35
   ) +
   coord_equal(
     expand = F,
-    ylim = c(0.6, 0.85),
+    ylim = c(0.6, 0.83),
     xlim = c(NA, 2.3),
     ratio = 1.75
   ) +
   annotate(
     geom = "text",
     x = c(0.75, seq(1.1, 2, length.out = 4)),
-    y = 0.82,
+    y = 0.81,
     label = sprintf("(%s)", letters[seq(5)]),
     fontface = "bold"
   ) +
-  scale_colour_distiller(
-    palette = "BuPu", direction = 1,
-    values = c(-2, 1)
+  scale_colour_manual(
+    values = c(colorspace::sequential_hcl(3,
+      l = 40, palette = "PuBu",
+      rev = T
+    ), "sienna")
   ) +
   ggthemes::theme_few() +
   theme(
@@ -343,14 +345,16 @@ fig_agg_speed <-
       y = 1 + speed
     ),
     fill = pal[3],
+    size = 0.3,
     show.legend = F,
     width = 0.25,
     outlier.size = 0.2,
-    position = position_nudge(x = 0.25)
+    position = position_nudge(x = 0.15)
   ) +
   geom_boxplot(aes(factor(interval), 1 + speed),
-    position = position_nudge(x = -0.25, ),
+    position = position_nudge(x = -0.15, ),
[…]

### 2021-02-17 docs/ms_atlas_preproc_main_text.tex
edit keywords
main text after ran comments

@@
     \item Modern, high-throughput animal tracking studies collect increasingly large volumes of data at very fine temporal scales.
-    At these scales, location-error can exceed the animal's step size, leading to mis-estimation of movement metrics such as speed.
+    At these scales, location error can exceed the animal's step size, leading to mis-estimation of key movement metrics such as speed.
     `Cleaning' the data to reduce location errors prior to analyses is one of the main ways movement ecologists deal with noisy data, and has the advantage of being more scalable to massive datasets than more complex methods.
-    Though data cleaning is widely recommended, and ecologists routinely consider cleaned data to be the ground-truth, uniform guidance on this crucial step is scarce.
+    Though data cleaning is widely recommended, and ecologists routinely consider cleaned data to be the ground-truth, inclusive uniform guidance on this crucial step is still rather scarce.
 
-    \item A pipeline for cleaning massive high-throughput datasets must balance ease of use, computational efficiency.
-    Data cleaning must also reject location errors without discarding valid animal movements.
+    \item A pipeline for cleaning massive high-throughput datasets must balance ease of use and computationally efficient signal vs.
+noise screening, in which location errors are rejected without discarding valid animal movements.
     Another useful feature of a pre-processing pipeline is efficiently segmenting and clustering location data for statistical methods, while also being scalable to large datasets and robust to imperfect sampling.
@@
     This pipeline, consisting of removing outliers, smoothing the filtered result, and thinning it to a uniform sampling interval, is easily scaled to very large datasets.
-    We demonstrate this pipeline on simulated movement data with location errors.
-    We also show a case study of how large volumes of cleaned data can be quickly segmented-clustered into biologically meaningful ‘residence patches’.
-    We use calibration data to show how pre-processing improves its quality, and to verify that the residence patch synthesis accurately captures animal space-use.
+    We apply this pipeline on simulated movement data with location errors, and also show a case study of how large volumes of cleaned data can be quickly segmented-clustered into biologically meaningful ‘residence patches’.
+    We use calibration data to illustrate how pre-processing improves its quality, and to verify that the residence patch synthesis accurately captures animal space use.
     Finally, turning to tracking data from Egyptian fruit bats (\textit{Rousettus aegyptiacus}), we demonstrate the pre-processing pipeline and residence patch method in a fully worked out example.
 
-    \item To help with fast implementation, and to help standardise methods, we developed the \texttt{R} package \texttt{atlastools}, which we also introduce here.
+    \item To help with fast implementation of standardise methods, we developed the \texttt{R} package \texttt{atlastools}, which we also introduce here.
     Our pre-processing pipeline and \texttt{atlastools} can be used with any high-throughput animal movement data in which the high data-volume combined with knowledge of the tracked individuals’ movement capacity can be used to reduce location errors.
@@
 
-\textbf{Keywords} high-throughput movement ecology (HTME), data cleaning, movement ecology, high-throughput tracking, \textit{atlastools}, residence patch
+\textbf{Keywords} high-throughput movement ecology (HTME), biotelemetry, data cleaning, ATLAS tracking, \textit{atlastools}, residence patch
 
@@
 
-Tracking individual animals is the methodological mainstay of movement ecology, which seeks to link animal movement with internal causes, environmental factors, and resulting consequences \citep{nathan2008a, holyoak2008}.
+Tracking individual animals is the methodological mainstay of movement ecology, which seeks to link animal movement with internal causes, life-history capacities and constraints, biotic interactions and other environmental factors, and resulting consequences \citep{nathan2008a, holyoak2008}.
 Investigating fine-scale environmental and social drivers and the consequences of movement requires position data from many individuals at high temporal and spatial resolution relative to the phenomenon of interest (high-throughput tracking).
-High-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}; additionally, ‘reverse-GPS’ systems developed to track animals over land \citep{toledo2014, weiser2016, toledo2016,toledo2020, maccurdy2009, maccurdy2019} and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput location data at much lower costs \citep{wikelski2007}.
+High-throughput tracking is possible using GPS tags \citep[see recent examples in][]{strandburg-peshkin2015, papageorgiou2019, harel2016}; additionally, ‘reverse-GPS’ systems such as ATLAS (Advanced Tracking and Localization of Animals in real-life Systems) developed to track animals over land  \citep[][see also \citealt{maccurdy2009, maccurdy2019}]{toledo2014, weiser2016, toledo2016,toledo2020}, and in aquatic systems \citep{hussey2015, baktoft2019, baktoft2017,jung2015, aspillaga2021} routinely produce high-throughput location data at much lower costs \citep{wikelski2007}.
 
 Although high-resolution tracking provides a massive amount of data on the path of a tracked animal, this data presents a two-fold challenge to ecologists.
-First, the location-error of each position may approach or exceed the true step size of the animal compared to low-resolution tracking with the same measurement error.
+First, the location error of each position may approach or exceed the true step size of the animal compared to low-resolution tracking with the same measurement error.
 This biases derived metrics such as speed and tortuosity \citep[see][]{ranacher2016, noonan2019, hurford2009, calenge2009}.
-Additionally, the absolute location-error around a position introduces uncertainty when studying the location of an animal relative to fixed landscape features (e.g.
+Additionally, the absolute location error around a position introduces uncertainty when studying the location of an animal relative to fixed landscape features (e.g.
 roads) or mobile elements (e.g.
 other tracked individuals).
-Users have two main options to improve data quality: making inferences after modelling the system-specific location-error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008, aspillaga2021}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
-The first approach may be limited by the animal movement models that can be fit to the data \citep{fleming2014a, noonan2019, fleming2020}, and may be beyond the computational capacity of common hardware, leading users to prefer data cleaning instead.
+Users have two main options to improve data quality: making inferences after modelling the system-specific location error \citep{fleming2014a, fleming2020, jonsen2003, jonsen2005, johnson2008, patterson2008, aspillaga2021}, or pre-processing data to clean it of positions with large location errors \citep{bjorneraas2010}.
+The first approach may be limited by the animal movement models that can be fitted to the data \citep{fleming2014a, noonan2019, fleming2020}, and may be beyond the computational capacity of common hardware, leading users to prefer data cleaning instead.
 When attempting data cleaning, the second challenge of high-throughput tracking reveals itself: the large number of observations themselves \citep{weiser2016, toledo2020}.
@@
 While gross errors are often removed by positioning-system algorithms, ‘reasonable’ errors often remain to confront end users \citep{fischler1981, weiser2016, ranacher2016}.
-% Location errors may be exacerbated by the conditions of deployment, with systematic differences in location-error among different landscape types  \citep[see][]{lewis2007, deon2002, frair2004}. 
+% Location errors may be exacerbated by the conditions of deployment, with systematic differences in location error among different landscape types  \citep[see][]{lewis2007, deon2002, frair2004}. 
 % % For instance, the physical features of an animal’s habitat such as hills or woodland may impede the transmissions of radio-telemetry tags (Beardsworth et al.
 \textit{in prep.}), while canopy cover can lead to location errors in GPS tracking \citep{lewis2007}.
-Further, as high-throughput tracking is deployed in more regions, standardised pre-processing steps should be general enough to tackle animal movement data recovered from a range of environments.
+Further, as high-throughput tracking is deployed in more regions and for more species, standardised pre-processing steps should be general enough to tackle animal movement data recovered from a range of environments, so as to enable sound comparisons across species and ecosystems.
 
@@
 Pre-processing protocols are not often reported in the literature, or may not be easily tractable for mainstream computing hardware and software.
-Furthermore, filtering out positions on their location-error estimates may not be appropriate for outliers which represent unrealistic movement but have low error measures \citep{weiser2016, ranacher2016}.
+Although GPS tracking data are autonomously pre-processed without allowing access to the raw data (REF), substantial location errors could still exist.
+For other tracking technologies, the availability of raw data enables the researcher to better control the data processing stage, and to substantially improve data quality without extrapolating beyond the relevant range.
+Furthermore, filtering out positions on their location error estimates may not be appropriate for outliers which represent unrealistic movement but have low error measures \citep{weiser2016, ranacher2016}.
 This makes identifying and removing biologically implausible locations from a track an important component of recovering true animal movement \citep{bjorneraas2010}.
@@
 The large data-volumes of high-throughput tracking allow for a neat solution: tracks can be `smoothed' to reduce small location errors that have remained undetected.
-This large data-volume becomes a challenge when users seek to examine animal space-use in relation to relevant environmental covariates, such as the predictors of prolonged residence in an area \citep[see][]{bracis2018, aarts2008, fleming2014a,bijleveld2016, oudman2018, harel2016}.
+This large data-volume becomes a challenge when users seek to examine animal space use in relation to relevant environmental covariates, such as the predictors of prolonged residence in an area \citep[see][]{bracis2018, aarts2008, fleming2014a,bijleveld2016, oudman2018, harel2016}.
 To avoid issues of non-independent observations due to strong spatio-temporal autocorrelation, and to make datasets computationally tractable for mainstream statistical software, users may want to reduce data-volumes by thinning or clustering.
@@
 \texttt{atlastools} functions are easy to use and understand, and rely on the efficiency of the \texttt{R} package \texttt{data.table} \citep{dowle2020}.
-% We begin by showing how the pipeline can clean simulated tracks to which we have artificially introduced location-error \citep{gurarie2017}.
-Beginning with basic spatio-temporal filtering of positions, we cover filtering outliers from movement tracks, and reducing the effect of location-error with a median smooth.
+% We begin by showing how the pipeline can clean simulated tracks to which we have artificially introduced location error \citep{gurarie2017}.
+Beginning with basic spatio-temporal filtering of positions, we cover filtering outliers from movement tracks, and reduce the effect of location error with a median smooth.
 Then, we suggest one solution to issues of computational tractability, which is to synthesise residence patches from clusters of spatio-temporally proximate positions, and so reveal important aspects of animal space use \citep[\textit{sensu}][]{bijleveld2016, oudman2018, barraquand2008}.
@@
     %  (see \textsc{Smoothing and Thinning Data}).
-    \item Users who need uniformly thinned data can either aggregate or resample it.
+    \item Users who need uniformly thinned data can either aggregate or subsample it.
     %  (see \textsc{Smoothing and Thinning Data}).
@@
 Instances of the latter are likely to be system specific, but are known from ATLAS systems (Bijleveld et al. \textit{in prep.}). 
-Bounding boxes are by definition rectangular, and users seeking to filter for other geometries, such as a circular or irregularly-shaped study area, need a geometric intersection between their data and a spatial representation of the area of interest (e.g.
+Bounding boxes are typically rectangular, and users seeking to filter for other geometries, such as a circular or irregularly-shaped study area, need a geometric intersection between their data and a spatial representation of the area of interest (e.g.
 shapefile, geopackage, or \texttt{sf}-object in \texttt{R}).
@@
     \textbf{(a)} Tracks can be quickly filtered by spatial bounds (dashed grey lines) to exclude broad regions (green = retained; grey = removed).
-    \textbf{(b)} Location-error may affect single observations resulting in point outliers or `spikes' (red crosses and track segments), or continuous subsets of a track, called a `prolonged spike' (purple crosses, top right), and both represent unrealistic movement.
+    \textbf{(b)} location error may affect single observations resulting in point outliers or `spikes' (red crosses and track segments), or continuous subsets of a track, called a `prolonged spike' (purple crosses, top right), and both represent unrealistic movement.
     \textbf{(c)} Histograms of speed for the track (grey = small-scale errors, red = spikes), and the prolonged spike (purple).
@@
 
-An animal’s movement may be non-representative when it has been fitted with a tracker but has not been released, leading to artificial stationary positions, or in the time shortly after release when its movement may be influenced by the stress of capture and handling.
-Periods of poor tracking quality may result from system malfunctions, and users may also wish to exclude these data.
+Tag-recorded data might fail to properly represent an animal’s movement for various technical system-level reasons.
+Data recorded before release, for example, add irrelevant stationary positions, whereas data recorded shortly after release might be influenced by the stress of capture and handling.
+Periods of poor tracking quality may result from system malfunctions and unusual disturbances, and users may also wish to exclude these data.
 Temporal filtering can exclude positions from intervals when data are expected to be unreliable for ecological inference, either due to abnormal movement behaviour or system-specific issues.  
@@
     Here, the first example shows how nighttime data can be retained using a predicate that determines whether the value of `hour' is between 6 and 18, and also within a range of X coordinates.
-    The second example retains ATLAS locations calculated using $>$ 3 base stations (\texttt{NBS}), with location-error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.
+    The second example retains ATLAS locations calculated using $>$ 3 base stations (\texttt{NBS}), with location error (\texttt{SD}) $<$ 100, and data between an arbitrary day 5 and day 8.
     }]
@@
 Tracking data covariates can be good indicators of the reliability of positions calculated by a tracking system \citep{beardsworth2021}.
-GPS systems provide direct measures of location-error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS]{ranacher2016}, while  reverse-GPS systems provide a similar estimate \citep[called Standard Deviation, SD;][]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
+GPS systems provide direct measures of location error during localisation \citep[Horizontal Dilution of Precision, HDOP in GPS]{ranacher2016}, while  reverse-GPS systems provide a similar estimate \citep[called Standard Deviation, SD;][]{maccurdy2009, maccurdy2019, weiser2016, ranacher2016}.
 Tracking data can also include indirect indicators of data quality.
-For instance, GPS systems' location-error may be indicated indirectly by the number of satellites involved in the localisation.
+For instance, GPS systems' location error may be indicated indirectly by the number of satellites involved in the localisation.
 In reverse-GPS systems too, the number of base stations involved in each localisation is an indirect indicator of data quality, and positions localised using more receivers are usually more reliable \citep[the minimum required for an ATLAS localisation is 3; see][]{weiser2016}.
-% A location-error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
+% A location error measure associated with each coordinate pair (similar to GPS HDOP) can be calculated and assigned to a new column \texttt{SD} using the formula for the sum of correlated random variables
 % \begin{linenomath*}
@@
 
-An example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 2.b).
+A common example of a problem with individual positions is that of point outliers or `spikes' \citep{bjorneraas2010}, where a single position is displaced far from the track (see Fig. 2.b).
 Point outliers are characterised by artificially high speeds between the outlier and the positions before and after \citep[called incoming and outgoing speed, respectively;][]{bjorneraas2010}, lending a `spiky' appearance to the track.
@@
 To avoid removing valid, fast transit segments while still excluding spikes we suggest combining the speed filter with a filter on the turning angles of each position \citep[see][]{calenge2009}.
-This combined filter assumes that positions in high-throughput tracking with both high speeds and large turning angles are likely to be due to location errors, since most species are unable to turn sharply at high speed.
+This combined filter assumes that positions in high-throughput tracking with both high speeds and large turning angles are likely to be due to location errors, since most species are unable to turn sharply at very high speed.
 Users can then remove those positions whose incoming and outgoing speeds are both $> S$, and where $\theta > A$ (sharp, high-speed turns), where $\theta$ is the turning angle, and $A$ is the turning angle threshold.
@@
 
-Sometimes, entire subsets of the track may be affected by the same large-scale location-error.
+Sometimes, entire subsets of the track may be affected by the same large-scale location error.
 For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 2.b).
@@
 % We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
-Users are strongly encouraged to visualise their data before and after applying such algorithms, 
+Users are strongly encouraged to visualise their data before and after applying such algorithms. 
 We caution that these methods are not foolproof, and data that are heavily distorted by errors affecting entire track-subsets should be used with care when making further inferences.
@@
 These smaller errors are challenging to remove since their covariates (such as speed and turning angles) are within the expected range of movement behaviour for the study species. 
-The large data-volumes of high-throughput tracking allow users to resolve this problem by smoothing the positions. 
-A ‘smooth’ works by approximating the value of an observation based on neighbouring values.
+The large data volumes of high-throughput tracking allow users to resolve this problem by smoothing the positions. 
+The most basic ‘smooths’ work by approximating the value of an observation based on neighbouring values.
 For a one-dimensional series of observations, the neighbouring values are the $K$ observations centred on each index value $i$.
@@
 A common smooth is nearest neighbour averaging, in which the value of an observation $x_i$ is the average of the moving window $K$.
-The median smooth is a variant of nearest neighbour averaging which uses the median rather than the mean, and it is robust to outliers (\citeauthor{tukey1977} 1977).
+The median smooth is a variant of nearest neighbour averaging which uses the median rather than the mean, and is more robust to outliers (\citeauthor{tukey1977} 1977).
 The median smoothed value of the X coordinate, for instance, is
@@
     \begin{equation*}
-        X_i = Median(X_{i - (K-1)/2} \ldots X_{i + (K-1)/2})
+        X_i = Median(X_{i - (K-1)/2} \ldots X_{i + (K-1)/2}).
      \end{equation*}
@@
 
+Some data sources, such as GPS, provide tracks that have already been smoothed in quite sophisticated ways, such as with a Kalman filter \citep{kaplan2005}.
+Thus a median smooth is unnecessary on GPS tracks.
+Furthermore, smoothing is not a panacea for data quality issues, and has its drawbacks.
 Smoothing does not change the number of observations, but does decouple the coordinates from some of their covariates.
-For instance, smoothing breaks the relationship between a coordinate and the location-error estimate around it (VARX, VARY, and SD in ATLAS systems, or HDOP in GPS tracking).
+For instance, smoothing breaks the relationship between a coordinate and the location error estimate around it (VARX, VARY, and SD in ATLAS systems).
 This makes subsequent filtering on covariates of data quality unreliable, and smoothed data are unsuitable for use with methods that model location uncertainty \citep{noonan2019, fleming2014a, fleming2020, calabrese2016}.
-Furthermore, while larger $K$ may result in smoother tracks (Fig.
-3.b -- e), one drawback of using a large $K$ is that short, quick forays away from the main track are likely to be smoothed away, leading to a loss in detail of the individual's small-scale movement.
+Thus, when applying location error modelling methods, users should be sure that the error measure bears a mechanistic relationship with the location estimate \citep[see][ for more details]{fleming2020, noonan2019}.
+Additionally, excessively large $K$ may result in a loss in detail of the individual's small-scale movement (compare Fig. 3.e with 3.a).
 Users must themselves judge how best to balance large-scale and small-scale accuracy, and choose $K$ accordingly.
@@
     \textbf{(a)} The goal of this step is to approximate the simulated canonical track (black line), given positions with small-scale error remaining from previous steps (green points; grey lines).
-    % The canonical track is shown `zoomed in' for better representation, and the missing data at the top of the track (line with no points) is from where the prolonged spike was earlier removed.
-    Median smoothing the given position coordinates (green points, in \textbf{(a)}) over a moving window ($K$) of \textbf{(b)} 3, \textbf{(c)} 5, \textbf{(d)} 11, and \textbf{(e)} 21 positions respectively, yields approximations (purple lines) of the canonical track.
+    Median smoothing the given position coordinates (green points, in \textbf{(a)}) over a moving window ($K$) of \textbf{(b)} 5, \textbf{(c)} 11, and \textbf{(d)} 21 positions respectively, yields approximations (coloured lines) of the canonical track.
+    \textbf{(e)} However, extremely large $K$ (101) may lead to a loss of both large and small scale detail.
     Grey lines show the track without smoothing.
-    Users are cautioned that there is no correct $K$, and they must subjectively choose a $K$ which most usefully trades small-scale details of the track for large-scale accuracy.}
+    Users are cautioned that there is no correct $K$, and they must subjectively choose a $K$ which most usefully trades small-scale details of the track for large-scale accuracy.
+    }
     \label{fig:figure_median_smooth}
@@
 Thinning data need not compromise researchers' ability to answer scientific questions with them; for instance, proximity-based social interactions lasting 1 -- 2 minutes would still be detected on thinning from a sampling interval of 1 second to 1 minute.
+Thinning data also does not imply that efforts to collect high-throughput movement data are ‘wasted’, as rich movement data-sets enable more detailed and more accurate representation of the true track, as elaborated above.
 Indeed, some analyses require that temporal auto-correlation in the data be broken by subsampling the data to a lower resolution \citep[such as the estimation of home-ranges or step-selection functions;][]{fleming2014a, dupke2017}.
 Added to the requirement of uniform sampling intervals by many methods in the field, evenly reducing data-volumes is a worthwhile endeavour \citep[e.g.][]{fleming2014a, michelot2016, avgar2016}.
-Two plausible approaches here are resampling and aggregation, and both approaches begin with identifying time-interval groups (e.g. of 1 minute).
-Resampling picks one position from each time-interval group while aggregation involves computing the mean values of all covariates for positions within a time-interval group.
+Two plausible approaches here are subsampling and aggregation, and both approaches begin with identifying time-interval groups (e.g. of 1 minute).
+subsampling picks one position from each time-interval group while aggregation involves computing the mean values of all covariates for positions within a time-interval group.
 Both approaches yield one position per time-interval group (Fig. 4.a).
@@
 
-Both aggregation and resampling have their relative advantages. 
-The aggregation method is less sensitive to selecting point outliers by chance than resampling.
-However, when users want to account for location-error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location-error is important, and resampling directly propagates these errors without further processing.
-% In reverse-GPS systems systems the location-error is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
-In the aggregation method, the location-error around each coordinate provided by either GPS or reverse-GPS system can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
+Both aggregation and subsampling have their relative advantages. 
+The aggregation method is less sensitive to selecting point outliers by chance than subsampling.
+However, when users want to account for location error with methods such as state-space models \citep{jonsen2003, jonsen2005, johnson2008}, or continuous time movement models \citep{fleming2014a, noonan2019, gurarie2017, calabrese2016, fleming2020}, correctly propagating the location error is important, and subsampling directly propagates these errors without further processing.
+% In reverse-GPS systems systems the location error is calculated from the variance-covariance matrix of the coordinates of candidate positions considered by the location solver \citep{weiser2016}; this is equivalent to GPS systems' HDOP \citep{ranacher2016}.
+In the aggregation method, the location error around each coordinate provided by either GPS or reverse-GPS systems can be propagated to the averaged position as the sum of errors divided by the square of the number of observations contributing to each average ($N$):
 \begin{linenomath*}
@@
 \end{linenomath*}
-Similarly, the overall location-error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance:
+Similarly, the overall location error estimate for the average of $N$ positions in a time-interval can be calculated by treating it as a variance.
+For instance, the ATLAS error and GPS error measures (SD and HDOP, respectively) can be aggregated as:
 \begin{linenomath*}
@@
 %
-Thinning is implemented in \texttt{atlastools} using the \texttt{atl\_thin\_data} function, with either aggregation or resampling (specified by the \texttt{method} argument) over an interval using the \texttt{interval} argument.
+Thinning is implemented in \texttt{atlastools} using the \texttt{atl\_thin\_data} function, with either aggregation or subsampling (specified by the \texttt{method} argument) over an interval using the \texttt{interval} argument.
 % The column of timestamps must be named `time' and column classes except the identity columns (\textit{see below}) should allow averaging.
-% The `resample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
-% Using `resample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
-% In both aggregation and resampling, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
+% The `subsample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
+% Using `subsample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
+% In both aggregation and subsampling, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
 % Due to its development for ATLAS systems, the `aggregate' assumes the variance around coordinates is named \texttt{VARX} and \texttt{VARY}, and standard deviation around each position is named \texttt{SD}.
@@
 
-\begin{lstlisting}[float, language=R, style=customR, caption = {Code to thin data by aggregation in \texttt{atlastools}. The method can be either "aggregate" or "resample". 
+\begin{lstlisting}[float, language=R, style=customR, caption = {Code to thin data by aggregation in \texttt{atlastools}. The method can be either "aggregate" or "subsample". 
 The time interval is specified in seconds, while the \texttt{id\_columns} allows a character vector of column names to be passed to the function, with these columns used as identity variables.
@@
 
-An important and common analysis with animal tracking data is to link space-use with environmental covariates.
+An important and common analysis with animal tracking data is to link space use with environmental covariates.
 This is difficult even with smoothed and thinned high-throughput data, as these may be too large for statistical packages, or have strong autocorrelation.
@@
 Patch identification assesses whether consecutive stationary positions and the bouts they comprise are spatio-temporally independent, and clusters them together if they are not.
-The splitting and lumping of positions into bouts, and bouts into patches is based on simple user specified thresholds --- the distance and the time interval between positions (and bouts) beyond which they should be considered independent.
+The splitting and lumping of positions into bouts, and bouts into patches is based on simple user-specified thresholds --- the distance and the time interval between positions (and bouts) beyond which they should be considered independent.
 % Patches are considered independent if either the distance or time between them exceeds their respective thresholds.
-Users are encouraged to base these thresholds in the movement habits of their study species.
+Users are encouraged to base these thresholds on the movement habits of their study species.
 For example, residence patch classification of red knot movement tracks considers consecutive stationary positions independent if they are 20m apart and considers consecutive bouts independent if the distance between them $\geq$ 100m, or the time difference between them $\geq$ 30 minutes (Listing 6).
@@
     This is especially useful when studying more complex behaviour such as area-restricted search, which may have a directional component.
-    The residence patch method loses the details of movement between patches, but can efficiently represent the general pattern of space-use (see coloured points representing patch centroids, and lines joining them).
+    The residence patch method loses the details of movement between patches, but can efficiently represent the general pattern of space use (see coloured points representing patch centroids, and lines joining them).
     \textbf{(b)} A plot of residence time against time (solid line; Bracis et al. 2018) shows how the residence patch algorithm segments and clusters positions of prolonged residence. 
@@
 6).
-We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 \citep[][Bijleveld et al.
+We collected the calibration data (n = 50,816) by walking and boating with a hand-held WATLAS tag (sampling frequency = 1 Hz) around the island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) in August 2020 \citep[WATLAS: Wadden Sea ATLAS system][Bijleveld et al.
 \textit{in prep.}]{beardsworth2021}.
@@
 Stops had a median duration of 10.28 minutes (range: 1.75 minutes -- 20 minutes; see Supplementary Material).
-We cleaned the data before constructing residence patches by (1) removing a single outlier ($>$ 15 km away), removing unrealistic movement ($\geq$ 15 m/s), smoothing the data ($K$ = 5), and (4) thinning the data by resampling over a 30 second interval.
+We cleaned the data before constructing residence patches by (1) removing a single outlier ($>$ 15 km away), removing unrealistic movement ($\geq$ 15 m/s), smoothing the data ($K$ = 5), and (4) thinning the data by subsampling over a 30 second interval.
 The cleaning steps retained 37,324 positions (74.45\%), while thinning reduced these to 1,803 positions (4.8\% positions of the smoothed track).
@@
 Bats were tracked over three nights (5\textsuperscript{th}, 6\textsuperscript{th}, and 7\textsuperscript{th} May, 2018) in the Hula Valley, Israel (33.1$^{\circ}$N, 35.6$^{\circ}$E), with an average of 13,370 positions (SD = 2,173; range = 11,195 -- 15,542; interval = 8 seconds) per individual.
-Plotting the tracks showed severe distortions (see Supplementary Material Fig. 2.1).
+Plotting the tracks showed some severe distortions (see Supplementary Material Fig. 2.1).
 We first reduced location errors by removing observations with ATLAS SD $>$ 20, and observations calculated using fewer than four base stations (mean positions remaining = 10,447 / individual; 78\% of the raw data on average).
@@
 Bats had a mean residence time at locations of 100.54 minutes (SD = 114.7); this measure was strongly biased by time spent at the roost.
-We opted for a first-principles approach and selected as residence positions any locations with a residence time $>$ 5 minutes, reasoning that a flying animal stopping for $>$ 5 minutes at a location should plausibly indicate resource use or another interesting behaviour.
-This step retained 7,819 positions per bat on average (75.6\%) of the smoothed data, showing that bats are actually relatively stationary at the daily scale, and move only mostly between their roost and foraging sites (see Fig.
+We opted for a first-principles approach and selected as residence positions any locations with a residence time $>$ 5 minutes, reasoning that a flying animal stopping for $>$ 5 minutes at a location should plausibly indicate resource use or another interesting localized behaviour.
+This step retained 7,819 positions per bat on average (75.6\%) of the smoothed data, showing that bats are actually relatively stationary at the nightly scale, and move only mostly between their roost and foraging sites (see Fig.
 7).
@@
 Bats tended to show prolonged residence near known food sources (fruit trees), travelling repeatedly between previously visited areas (Fig. 7.b).
-However, bats also appeared to spend some time at locations where no fruit trees were recorded, prompting questions about their use of other food sources, or another behaviour entirely (Fig.
-7.b, 7.c).
+However, bats also appeared to spend some time at locations where no fruit trees were recorded (Fig.
+7.b, 7.c), in line with previous evidence for their use of trees to rest, to handle and digest food, and presumably for sexual or social reasons (REF).
 Bats occurring close together did not have strongly overlapping residence patches, and their paths to and from area of co-occurrence were different (Fig. 7.a, 7.c).
@@
 The open-source \texttt{R} package \texttt{atlastools} serves as a starting point for methodological collaboration among movement ecologists.
-Efficient location-error modelling approaches \citep{fleming2020, aspillaga2021} may eventually make data-cleaning optional.
-Yet cleaning tracking data even partially before modelling location-error is faster than error-modelling on the full data, and the removal of large location errors may improve model fits.
+Efficient location error modelling approaches \citep{fleming2020, aspillaga2021} may eventually make data-cleaning optional.
+Yet cleaning tracking data even partially before modelling location error is faster than error-modelling on the full data, and the removal of large location errors may improve model fits.
 Thus we see our pipeline as complementary to these approaches \citep{fleming2014a, fleming2020}.
@@
 We thank the many volunteers, students, and NIOZ staff involved in operating the WATLAS tracking system, and most importantly Frank van Maarseveen, Bas Denissen and Anne Dekinga.
-We also thank the attendees of ATLAS workshops held in May and June 2020 at the Hebrew University of Jerusalem for helpful comments on the pipeline and \texttt{atlastools}.
+We also thank the members of the Minerva Center for Movement Ecology for their valuable support and especially the attendees of ATLAS workshops held in May and June 2020 at the Hebrew University of Jerusalem for helpful comments on the pipeline and \texttt{atlastools}.
 This work was partly funded by the Dutch Research Council grant VI.Veni.192.051 awarded to AIB.
+ATLAS development was funded by the Minerva Foundation grant to RN, and by the Israel Science Foundation grant (ISF ISF-965/15) to RN and ST.
 

### 2021-02-17 R/_03_smoothing_tracks.R
update fig 3 script

@@ -187,7 +187,7 @@ figure_median_smooth <-
     values = c(colorspace::sequential_hcl(3,
       l = 40, palette = "PuBu",
       rev = T
-    ), "sienna")
+    ), "darkred")
   ) +
   ggthemes::theme_few() +
   theme(

### 2021-02-17 docs/ms_atlas_preproc_main_text.tex
a minor edit again

@@
 
-We present a fully worked-out example of our pre-processing pipeline and residence patch method using movement data from three Egyptian fruit bats tracked using the ATLAS system (\textit{Rousettus aegyptiacus}; \citet{toledo2020}).
+We present a fully worked-out example of our pre-processing pipeline and residence patch method using movement data from three Egyptian fruit bats tracked using the ATLAS system (\textit{Rousettus aegyptiacus}; \citealt{toledo2020}).
 Code can be found in the Supplementary Material (see \textsc{Processing Egyptian Fruit Bat Tracks}).

### 2021-02-17 docs/ms_atlas_preproc_main_text.tex
add new refs

@@
 Pre-processing protocols are not often reported in the literature, or may not be easily tractable for mainstream computing hardware and software.
-Although GPS tracking data are autonomously pre-processed without allowing access to the raw data (REF), substantial location errors could still exist.
+Although GPS tracking data are autonomously pre-processed without allowing access to the raw data \citep{kaplan2005}, substantial location errors could still exist.
 For other tracking technologies, the availability of raw data enables the researcher to better control the data processing stage, and to substantially improve data quality without extrapolating beyond the relevant range.
@@
 5.a).
-RACVM models approximate the tracks of soaring birds which circle on thermals over a relatively small area, and move between thermals \citep[`thermalling'; ][]{gurarie2017, harel2016}.
+RACVM models approximate the tracks of soaring birds which circle on thermals over a relatively small area, and move between thermals \citep[`thermalling'; ][]{gurarie2017, harel2016,harel2018}.
 This complex track structure provides a suitable challenge for the residence patch method and helps to demonstrate its generality.
@@
 However, bats also appeared to spend some time at locations where no fruit trees were recorded (Fig.
-7.b, 7.c), in line with previous evidence for their use of trees to rest, to handle and digest food, and presumably for sexual or social reasons (REF).
+7.b, 7.c), in line with previous evidence for their use of trees to rest, to handle and digest food, and presumably for sexual or social reasons \citep{tsoar2010}.
 Bats occurring close together did not have strongly overlapping residence patches, and their paths to and from area of co-occurrence were different (Fig. 7.a, 7.c).

### 2021-04-16 06_bat_data.Rmd
install atlastools
eval chunk

@@
 pal <- RColorBrewer::brewer.pal(4, "Set1")
+```
+
+```{r install_atlastools_2, eval=TRUE}
+install.packages("remotes")
+
+# installation using remotes
+remotes::install_github("pratikunterwegs/atlastools")
 ```
@@
 
-```{r echo=FALSE, eval=FALSE}
+```{r echo=FALSE, eval=TRUE}
 # not shown or mentioned

### 2021-05-04 scripts/05_calibration_data.R
move code to scripts

@@ -1,0 +1,670 @@
+## ----install_atlastools, eval=FALSE-------------------------------------------
+## install.packages("remotes")
+##
+## # installation using remotes
+## remotes::install_github("pratikunterwegs/atlastools")
+
+
+## ----eval=TRUE----------------------------------------------------------------
+# for data handling
+library(data.table)
+library(atlastools)
+
+# for recursion analysis
+library(recurse)
+
+# for plotting
+library(ggplot2)
+library(patchwork)
+
+# making a colour palette
+pal <- RColorBrewer::brewer.pal(5, "Set1")
+pal[3] <- "seagreen"
+
+
+## ---- eval=TRUE---------------------------------------------------------------
+# read and plot example data
+data <- fread("data/atlas1060_allTrials_annotated.csv")
+data_raw <- copy(data)
+
+# see raw data
+head(data_raw)
+
+
+## ----echo=TRUE----------------------------------------------------------------
+# plot data
+fig_data_raw <-
+  ggplot(data) +
+  geom_path(aes(x, y),
+    col = "grey", alpha = 1, size = 0.2
+  ) +
+  geom_point(aes(x, y),
+    col = "grey", alpha = 0.2, size = 0.2
+  ) +
+  ggthemes::theme_few() +
+  theme(
+    axis.title = element_blank(),
+    axis.text = element_blank()
+  ) +
+  coord_sf(crs = 32631)
+
+# save figure
+ggsave(fig_data_raw,
+  filename = "figures/fig_calibration_raw.png",
+  width = 185 / 25
+)
+
+
+## -----------------------------------------------------------------------------
+# make a copy using the data.table copy function
+data_unproc <- copy(data)
+
+
+## -----------------------------------------------------------------------------
+# remove inside must be set to falses
+data <- atl_filter_bounds(
+  data = data,
+  x = "x", y = "y",
+  x_range = c(645000, max(data$x)),
+  remove_inside = FALSE
+)
+
+
+## ----echo=FALSE---------------------------------------------------------------
+# plot data
+fig_data_bbox <-
+  ggplot() +
+  geom_path(
+    data = data_raw,
+    aes(x, y),
+    col = "grey", alpha = 0.5, size = 0.3
+  ) +
+  geom_point(
+    data = data_raw,
+    aes(x, y,
[…]

### 2021-05-04 scripts/06_bat_data.R

@@ -1,0 +1,474 @@
+## -----------------------------------------------------------------------------
+# libs for data
+library(data.table)
+library(RSQLite)
+library(atlastools)
+
+# libs for plotting
+library(ggplot2)
+library(patchwork)
+
+# recursion analysis
+library(recurse)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+
+
+## ----read_bat_data------------------------------------------------------------
+# prepare the connection
+con <- dbConnect(
+  drv = SQLite(),
+  dbname = "data/Three_example_bats.sql"
+)
+
+# list the tables
+table_name <- dbListTables(con)
+
+# prepare to query all tables
+query <- sprintf('select * from \"%s\"', table_name)
+
+# query the database
+data <- dbGetQuery(conn = con, statement = query)
+
+# disconnect from database
+dbDisconnect(con)
+
+
+## -----------------------------------------------------------------------------
+# convert data to datatable
+setDT(data)
+
+# write data for QGIS
+fwrite(data, file = "data/bat_data.csv")
+
+
+## ----echo=FALSE---------------------------------------------------------------
+# plot bat data to examine for errors
+fig_bat_raw <-
+  ggplot(data) +
+  geom_path(aes(X, Y,
+    group = TAG,
+    colour = TAG
+  ),
+  show.legend = FALSE
+  ) +
+  scale_colour_brewer(palette = "Dark2") +
+  ggspatial::annotation_scale(location = "br") +
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  ) +
+  coord_sf(crs = 2039)
+
+
+## ----echo=FALSE---------------------------------------------------------------
+# save to figures
+ggsave(fig_bat_raw,
+  filename = "figures/fig_bat_raw.png",
+  dpi = 300,
+  width = 6, height = 4
+)
+
+
+## -----------------------------------------------------------------------------
+# split bat data by tag
+# first make a copy using the data.table function copy
+# this prevents the orignal data from being modified by atlastools
+# functions which DO MODIFY BY REFERENCE!
+data_split <- copy(data)
+
+# now split
+data_split <- split(data_split, by = "TAG")
+
+
+## -----------------------------------------------------------------------------
+# get SD.
[…]

### 2021-05-04 scripts/_01_background.R

@@ -1,0 +1,112 @@
+## ----install_package----------------------------------------------------------
+# use either devtools or remotes to install
+install.packages("devtools")
+
+# installation using devtools
+devtools::install_github("pratikunterwegs/atlastools")
+
+
+## ----prep_libs_01_02----------------------------------------------------------
+# load smoove and datatable
+library(smoove)
+library(data.table)
+library(scales)
+
+# source helper functions
+source("R/helper_functions.R")
+
+
+## ----save_data----------------------------------------------------------------
+data <- do_smoove_data()
+# save simulated data
+fwrite(data, "data/data_sim.csv")
+
+
+## -----------------------------------------------------------------------------
+# set a seed
+set.seed(0)
+
+# do smoove data using a RACVM for three different rotational speed
+data_thermals <- Map(function(rot_speed, patch) {
+  dt <- smoove::simulateRACVM(dt = 0.1, Tmax = 50, omega = rot_speed, v0 = 0.1)
+  dt <- as.data.table(dt$XY)
+  dt[, patch := patch]
+  dt[, time := seq(.N)]
+}, c(7, 7, 7), sprintf("patch_%i", seq(3)))
+
+# rescale x and y
+data_thermals <- Map(
+  function(df, lim1, lim2, offset) {
+    df[, x := x + lim1]
+    df[, y := y + lim2]
+    df[, time := time + offset]
+  }, data_thermals, c(1, 5, 9), c(0, 2.5, 5),
+  c(100, 800, 1500)
+)
+
+# get thermals data
+data_thermals <- rbindlist(data_thermals)
+
+
+## -----------------------------------------------------------------------------
+# get limits
+starts <- data_thermals[, lapply(.SD, first), by = "patch"]
+ends <- data_thermals[, lapply(.SD, last), by = "patch"]
+
+# bind
+gaps <- rbind(starts, ends)
+setorder(gaps, "time")
+gaps[, event := rep(c("start", "end"), 3)]
+
+# cast
+gaps <- dcast(gaps, patch ~ event, value.var = c("x", "y", "time"))
+
+# make a trajectory section
+transit <- data.table(
+  x1 = shift(gaps$x_end),
+  x2 = gaps$x_start,
+  y1 = shift(gaps$y_end),
+  y2 = gaps$y_start,
+  t1 = shift(gaps$time_end),
[…]

### 2021-05-04 scripts/_02_filtering_data.R

@@ -1,0 +1,329 @@
+## -----------------------------------------------------------------------------
+# to handle movement data
+library(data.table)
+library(atlastools)
+
+# to plot
+library(ggplot2)
+library(patchwork)
+library(ggthemes)
+
+
+## -----------------------------------------------------------------------------
+# source helper functions
+source("R/helper_functions.R")
+
+
+## ----read_sim_data------------------------------------------------------------
+# read in the data
+data <- fread("data/data_sim.csv")[5000:10000, ]
+
+
+## ----add_outlier, eval=FALSE--------------------------------------------------
+## # make a copy
+## # data_copy <- copy(data)
+## #
+## # # add a prolonged spike or reflection to 300 positions
+## # data_copy[500:800, `:=`(
+## #   x = x + 0.25,
+## #   y = y + 0.25
+## # )]
+## #
+## # # add normal error
+## # data_copy[, `:=`(
+## #   x = do_add_error(x, std_dev = 0.01),
+## #   y = do_add_error(y, std_dev = 0.005)
+## # )]
+## #
+## # # add 100 outliers
+## # data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
+
+
+## ----eval=FALSE---------------------------------------------------------------
+## # fwrite(data_copy, file = "data/data_errors.csv")
+
+
+## -----------------------------------------------------------------------------
+data_copy <- fread("data/data_errors.csv")
+
+
+## -----------------------------------------------------------------------------
+# define a four colour palette
+pal <- RColorBrewer::brewer.pal(5, "Set1")
+pal[3] <- "seagreen"
+
+
+## ----echo=FALSE---------------------------------------------------------------
+# make figure of canonical data with added errors
+# figure_raw <-
+#   ggplot()+
+#   geom_point(data = data_copy,
+#             aes(x, y),
+#             col = "grey",
+#             alpha = 1,
+#             shape = 4,
+#             size = 0.2)+
+#   geom_path(data = data,
+#             aes(x, y),
+#             col = "grey20",
+#             alpha = 1)+
+#   ggthemes::theme_few()+
+#   theme(axis.text = element_blank(),
[…]

### 2021-05-04 scripts/_03_smoothing_tracks.R

@@ -1,0 +1,395 @@
+## -----------------------------------------------------------------------------
+# prep libs
+library(data.table)
+library(atlastools)
+library(ggplot2)
+library(patchwork)
+
+# define a four colour palette
+pal <- RColorBrewer::brewer.pal(5, "Set1")
+pal[3] <- "seagreen"
+
+
+## ----read_sim_data_2----------------------------------------------------------
+# read in the data and set the window size variable
+data <- fread("data/data_sim.csv")[5000:10000, ]
+data[, window_size := NA]
+data$speed_in <- atl_get_speed(data)
+
+# data with small scale errors but no reflections or outliers
+data_errors <- fread("data/data_no_reflection.csv")
+data_errors[, window_size := 0]
+
+
+## -----------------------------------------------------------------------------
+# smooth the data over four K values
+list_of_smooths <- lapply(c(5, 11, 21, 101), function(K) {
+  data_copy <- copy(data_errors)
+
+  data_copy <- atl_median_smooth(
+    data = data_copy,
+    x = "x",
+    y = "y",
+    time = "time",
+    moving_window = K
+  )
+
+  data_copy[, window_size := K]
+})
+
+
+## -----------------------------------------------------------------------------
+fwrite(list_of_smooths[[2]], file = "data/data_smooth.csv")
+
+
+## -----------------------------------------------------------------------------
+# bind list after offset
+data_plot <- mapply(function(df, offset) {
+  df <- copy(df)
+  df[, x := x + offset]
+  df$speed_in <- atl_get_speed(df)
+  return(df)
+}, list_of_smooths, seq(0.4, 1.25, length.out = 4),
+SIMPLIFY = F
+)
+
+data_plot <- rbindlist(data_plot)
+
+# add offset to data errors
+data_errors <- lapply(
+  c(0, seq(0.4, 1.25, length.out = 4)),
+  function(offset) {
+    df <- copy(data_errors)
+    df[, x := x + offset]
+    df[, offset := offset]
+  }
+)
+
+# bind list
+data_errors <- rbindlist(data_errors)
+
+
+## -----------------------------------------------------------------------------
+pal2 <- RColorBrewer::brewer.pal(4, "RdPu")[c(2, 4)]
+
[…]

### 2021-05-04 scripts/_04_residence_patch.R

@@ -1,0 +1,207 @@
+## ----prep_libs_02_01----------------------------------------------------------
+library(data.table)
+library(atlastools)
+library(ggplot2)
+library(patchwork)
+
+# for residence time
+library(recurse)
+
+# prepare a palette
+pal <- RColorBrewer::brewer.pal(4, "Set1")
+
+
+## -----------------------------------------------------------------------------
+# read patch data
+data <- fread("data/data_for_res_patch.csv")
+
+# an id is required
+data[, id := "test"]
+
+# flip lengthwise
+setnames(data, old = c("x", "y"), new = c("y", "x"))
+
+# do recurse
+data_recurse <- getRecursions(data[, list(x, y, time, id)],
+  radius = 1
+)
+
+# assign residence time
+data[, residence_time := data_recurse$residenceTime]
+
+
+## ----echo=FALSE---------------------------------------------------------------
+# restime by position
+# fig_res_a <-
+#   ggplot()+
+#   geom_path(data = data,
+#             aes(x, y),
+#             lwd = 0.2,
+#             col = "grey")+
+#   geom_point(data = data,
+#              aes(x, y,
+#                  group = NA
+#                  ),
+#              size = 0.3,
+#              show.legend = F,
+#              alpha = 0.2,
+#              col = "grey20")+
+#   coord_equal()+
+#   scale_colour_manual(values = c("grey", pal[3]))+
+#   ggthemes::theme_few()+
+#   theme(axis.text = element_blank(),
+#         axis.title = element_blank())
+
+
+## -----------------------------------------------------------------------------
+# make residence patch
+patch <- atl_res_patch(data,
+  buffer_radius = 0.1,
+  lim_spat_indep = 1,
+  lim_time_indep = 5
+)
+
+
+
+# get spatial representation
+patch_sf <- atl_patch_summary(
+  patch_data = patch,
+  which_data = "spatial",
+  buffer_radius = 0.15
+)
+
+# get summary data
+patch_summary <- atl_patch_summary(
+  patch_data = patch,
+  which_data = "summary"
+)
+
+
+## ----echo=FALSE---------------------------------------------------------------
+# plot_patches <-
+fig_res_patches <-
+  ggplot() +
+  geom_sf(
[…]

### 2021-05-04 scripts/helper_functions.R

@@ -1,0 +1,50 @@
+#### Functions to simulate and add errors to movement data ####
+
+# functions are not documented and are not intended for general use
+
+# a function to use smoove to simulate movement data
+do_smoove_data <- function(t_max = 100, dt = dt) {
+  # define parameters
+  nu <- 2
+  tau <- 1
+  dt <- .01
+  ucvm1 <- smoove::simulateUCVM(nu = nu, tau = tau, T.max = 100, dt = dt)
+
+  # get the data in data.table format
+  tmp_data <- data.table::as.data.table(ucvm1$XY)
+  # rescale coordinates
+  tmp_data[, c("x", "y") := (lapply(
+    .SD,
+    scales::rescale
+  )),
+  .SDcols = c("x", "y")
+  ]
+
+  # add a time column
+  tmp_data[, time := seq_len(nrow(tmp_data))]
+
+  return(tmp_data)
+}
+
+# simple function to introduce normally distributed error
+do_add_error <- function(v, std_dev = 0.5) {
+  pos_error <- stats::rnorm(length(v), sd = std_dev)
+  return(v + pos_error)
+}
+
+# function for point outliers
+do_add_outliers <- function(data, p_data = 0.1, std_dev = 0.05) {
+  n_rows <- floor(p_data * nrow(data))
+  tmp_data <- data.table::copy(data)
+
+  tmp_data[sample(nrow(data), n_rows),
+    c("x", "y") := lapply(
+      .SD,
+      function(z) {
+        z + stats::rnorm(n_rows, sd = std_dev)
+      }
+    ),
+    .SDcols = c("x", "y")
+  ]
+  return(tmp_data)
+}

### 2021-05-04 supplement/common.r
move code to supplement

@@ -1,0 +1,20 @@
+knitr::opts_knit$set(root.dir = here::here())
+set.seed(1)
+options(digits = 3)
+
+knitr::opts_chunk$set(
+  comment = "#>",
+  collapse = TRUE,
+  cache = FALSE,
+  eval = TRUE,
+  warning = FALSE,
+  error = FALSE,
+  message = FALSE,
+  out.width = "\\textwidth", 
+  fig.align = "center",
+  fig.width = 7,
+  fig.asp = 0.618,  # 1 / phi
+  fig.show = "hold"
+)
+options(knitr.kable.NA = "")
+options(dplyr.print_min = 6, dplyr.print_max = 6)

### 2021-05-04 supplement/render_books.sh

@@ -1,0 +1,26 @@
+#!/bin/bash
+
+# remove old tex files and pdfs
+cd docs
+rm *.pdf
+
+cd ..
+
+cp figures/fig_0*.png atlas-manuscript/figures
+
+# style files
+Rscript --vanilla --slave -e 'styler::style_dir("scripts", filetype = "R", recursive = FALSE)'
+Rscript --vanilla --slave -e 'styler::style_dir("supplement", filetype = "Rmd", recursive = FALSE)'
+
+# render supplement
+Rscript --vanilla --slave -e 'bookdown::render_book("supplement/index.rmd", "bookdown::pdf_document2", output_dir = "docs", config_file = "supplement/_bookdown.yml")'
+
+# to render source files
+Rscript --slave -e 'lapply(list.files(pattern = "supplement\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
+
+# copy pdfs and name correctly
+cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text_`date -I`.pdf
+mv docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_`date -I`.pdf
+
+# build atlastools manual
+Rscript --slave -e 'devtools::build_manual(pkg = "../atlastools", path = "docs/")'

### 2021-05-04 R/05_calibration_data.R
remove old files

@@ -1,671 +1,0 @@
-## ----install_atlastools, eval=FALSE-------------------------------------------
-## install.packages("remotes")
-## 
-## # installation using remotes
-## remotes::install_github("pratikunterwegs/atlastools")
-
-
-## ----eval=TRUE----------------------------------------------------------------
-# for data handling
-library(data.table)
-library(atlastools)
-
-# for recursion analysis
-library(recurse)
-
-# for plotting
-library(ggplot2)
-library(patchwork)
-
-# making a colour palette
-pal <- RColorBrewer::brewer.pal(5, "Set1")
-pal[3] <- "seagreen"
-
-
-## ---- eval=TRUE---------------------------------------------------------------
-# read and plot example data
-data <- fread("data/atlas1060_allTrials_annotated.csv")
-data_raw <- copy(data)
-
-# see raw data
-head(data_raw)
-
-
-## ----echo=TRUE----------------------------------------------------------------
-# plot data
-fig_data_raw <-
-  ggplot(data) +
-  geom_path(aes(x, y),
-    col = "grey", alpha = 1, size = 0.2
-  ) +
-  geom_point(aes(x, y),
-    col = "grey", alpha = 0.2, size = 0.2
-  ) +
-  ggthemes::theme_few() +
-  theme(
-    axis.title = element_blank(),
-    axis.text = element_blank()
-  ) +
-  coord_sf(crs = 32631)
-
-# save figure
-ggsave(fig_data_raw,
-  filename = "figures/fig_calibration_raw.png",
-  width = 185 / 25
-)
-
-
-## -----------------------------------------------------------------------------
-# make a copy using the data.table copy function
-data_unproc <- copy(data)
-
-
-## -----------------------------------------------------------------------------
-# remove inside must be set to falses
-data <- atl_filter_bounds(
-  data = data,
-  x = "x", y = "y",
-  x_range = c(645000, max(data$x)),
-  remove_inside = FALSE
-)
-
-
-## ----echo=FALSE---------------------------------------------------------------
-# plot data
-fig_data_bbox <-
-  ggplot() +
-  geom_path(
-    data = data_raw,
-    aes(x, y),
-    col = "grey", alpha = 0.5, size = 0.3
-  ) +
-  geom_point(
-    data = data_raw,
-    aes(x, y,
[…]

### 2021-05-04 R/06_bat_data.R

@@ -1,475 +1,0 @@
-## -----------------------------------------------------------------------------
-# libs for data
-library(data.table)
-library(RSQLite)
-library(atlastools)
-
-# libs for plotting
-library(ggplot2)
-library(patchwork)
-
-# recursion analysis
-library(recurse)
-
-# prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
-
-
-## ----read_bat_data------------------------------------------------------------
-# prepare the connection
-con <- dbConnect(
-  drv = SQLite(),
-  dbname = "data/Three_example_bats.sql"
-)
-
-# list the tables
-table_name <- dbListTables(con)
-
-# prepare to query all tables
-query <- sprintf('select * from \"%s\"', table_name)
-
-# query the database
-data <- dbGetQuery(conn = con, statement = query)
-
-# disconnect from database
-dbDisconnect(con)
-
-
-## -----------------------------------------------------------------------------
-# convert data to datatable
-setDT(data)
-
-# write data for QGIS
-fwrite(data, file = "data/bat_data.csv")
-
-
-## ----echo=FALSE---------------------------------------------------------------
-# plot bat data to examine for errors
-fig_bat_raw <-
-  ggplot(data) +
-  geom_path(aes(X, Y,
-    group = TAG,
-    colour = TAG
-  ),
-  show.legend = FALSE
-  ) +
-  scale_colour_brewer(palette = "Dark2") +
-  ggspatial::annotation_scale(location = "br") +
-  ggthemes::theme_few() +
-  theme(
-    axis.text = element_blank(),
-    axis.title = element_blank()
-  ) +
-  coord_sf(crs = 2039)
-
-
-## ----echo=FALSE---------------------------------------------------------------
-# save to figures
-ggsave(fig_bat_raw,
-  filename = "figures/fig_bat_raw.png",
-  dpi = 300,
-  width = 6, height = 4
-)
-
-
-## -----------------------------------------------------------------------------
-# split bat data by tag
-# first make a copy using the data.table function copy
-# this prevents the orignal data from being modified by atlastools
-# functions which DO MODIFY BY REFERENCE!
-data_split <- copy(data)
-
-# now split
-data_split <- split(data_split, by = "TAG")
-
-
-## -----------------------------------------------------------------------------
-# get SD.
[…]

### 2021-05-04 R/_01_background.R

@@ -1,113 +1,0 @@
-## ----install_package----------------------------------------------------------
-# use either devtools or remotes to install
-install.packages("devtools")
-
-# installation using devtools
-devtools::install_github("pratikunterwegs/atlastools")
-
-
-## ----prep_libs_01_02----------------------------------------------------------
-# load smoove and datatable
-library(smoove)
-library(data.table)
-library(scales)
-
-# source helper functions
-source("R/helper_functions.R")
-
-
-## ----save_data----------------------------------------------------------------
-data <- do_smoove_data()
-# save simulated data
-fwrite(data, "data/data_sim.csv")
-
-
-## -----------------------------------------------------------------------------
-# set a seed
-set.seed(0)
-
-# do smoove data using a RACVM for three different rotational speed
-data_thermals <- Map(function(rot_speed, patch) {
-  dt <- smoove::simulateRACVM(dt = 0.1, Tmax = 50, omega = rot_speed, v0 = 0.1)
-  dt <- as.data.table(dt$XY)
-  dt[, patch := patch]
-  dt[, time := seq(.N)]
-}, c(7, 7, 7), sprintf("patch_%i", seq(3)))
-
-# rescale x and y
-data_thermals <- Map(
-  function(df, lim1, lim2, offset) {
-    df[, x := x + lim1]
-    df[, y := y + lim2]
-    df[, time := time + offset]
-  }, data_thermals, c(1, 5, 9), c(0, 2.5, 5),
-  c(100, 800, 1500)
-)
-
-# get thermals data
-data_thermals <- rbindlist(data_thermals)
-
-
-## -----------------------------------------------------------------------------
-# get limits
-starts <- data_thermals[, lapply(.SD, first), by = "patch"]
-ends <- data_thermals[, lapply(.SD, last), by = "patch"]
-
-# bind
-gaps <- rbind(starts, ends)
-setorder(gaps, "time")
-gaps[, event := rep(c("start", "end"), 3)]
-
-# cast
-gaps <- dcast(gaps, patch ~ event, value.var = c("x", "y", "time"))
-
-# make a trajectory section
-transit <- data.table(
-  x1 = shift(gaps$x_end),
-  x2 = gaps$x_start,
-  y1 = shift(gaps$y_end),
-  y2 = gaps$y_start,
-  t1 = shift(gaps$time_end),
[…]

### 2021-05-04 R/_02_filtering_data.R

@@ -1,330 +1,0 @@
-## -----------------------------------------------------------------------------
-# to handle movement data
-library(data.table)
-library(atlastools)
-
-# to plot
-library(ggplot2)
-library(patchwork)
-library(ggthemes)
-
-
-## -----------------------------------------------------------------------------
-# source helper functions
-source("R/helper_functions.R")
-
-
-## ----read_sim_data------------------------------------------------------------
-# read in the data
-data <- fread("data/data_sim.csv")[5000:10000, ]
-
-
-## ----add_outlier, eval=FALSE--------------------------------------------------
-## # make a copy
-## # data_copy <- copy(data)
-## #
-## # # add a prolonged spike or reflection to 300 positions
-## # data_copy[500:800, `:=`(
-## #   x = x + 0.25,
-## #   y = y + 0.25
-## # )]
-## #
-## # # add normal error
-## # data_copy[, `:=`(
-## #   x = do_add_error(x, std_dev = 0.01),
-## #   y = do_add_error(y, std_dev = 0.005)
-## # )]
-## #
-## # # add 100 outliers
-## # data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
-
-
-## ----eval=FALSE---------------------------------------------------------------
-## # fwrite(data_copy, file = "data/data_errors.csv")
-
-
-## -----------------------------------------------------------------------------
-data_copy <- fread("data/data_errors.csv")
-
-
-## -----------------------------------------------------------------------------
-# define a four colour palette
-pal <- RColorBrewer::brewer.pal(5, "Set1")
-pal[3] <- "seagreen"
-
-
-## ----echo=FALSE---------------------------------------------------------------
-# make figure of canonical data with added errors
-# figure_raw <-
-#   ggplot()+
-#   geom_point(data = data_copy,
-#             aes(x, y),
-#             col = "grey",
-#             alpha = 1,
-#             shape = 4,
-#             size = 0.2)+
-#   geom_path(data = data,
-#             aes(x, y),
-#             col = "grey20",
-#             alpha = 1)+
-#   ggthemes::theme_few()+
-#   theme(axis.text = element_blank(),
[…]

### 2021-05-04 R/_04_residence_patch.R

@@ -1,208 +1,0 @@
-## ----prep_libs_02_01----------------------------------------------------------
-library(data.table)
-library(atlastools)
-library(ggplot2)
-library(patchwork)
-
-# for residence time
-library(recurse)
-
-# prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
-
-
-## -----------------------------------------------------------------------------
-# read patch data
-data <- fread("data/data_for_res_patch.csv")
-
-# an id is required
-data[, id := "test"]
-
-# flip lengthwise
-setnames(data, old = c("x", "y"), new = c("y", "x"))
-
-# do recurse
-data_recurse <- getRecursions(data[, list(x, y, time, id)],
-  radius = 1
-)
-
-# assign residence time
-data[, residence_time := data_recurse$residenceTime]
-
-
-## ----echo=FALSE---------------------------------------------------------------
-# restime by position
-# fig_res_a <-
-#   ggplot()+
-#   geom_path(data = data,
-#             aes(x, y),
-#             lwd = 0.2,
-#             col = "grey")+
-#   geom_point(data = data,
-#              aes(x, y,
-#                  group = NA
-#                  ),
-#              size = 0.3,
-#              show.legend = F,
-#              alpha = 0.2,
-#              col = "grey20")+
-#   coord_equal()+
-#   scale_colour_manual(values = c("grey", pal[3]))+
-#   ggthemes::theme_few()+
-#   theme(axis.text = element_blank(),
-#         axis.title = element_blank())
-
-
-## -----------------------------------------------------------------------------
-# make residence patch
-patch <- atl_res_patch(data,
-  buffer_radius = 0.1,
-  lim_spat_indep = 1,
-  lim_time_indep = 5
-)
-
-
-
-# get spatial representation
-patch_sf <- atl_patch_summary(
-  patch_data = patch,
-  which_data = "spatial",
-  buffer_radius = 0.15
-)
-
-# get summary data
-patch_summary <- atl_patch_summary(
-  patch_data = patch,
-  which_data = "summary"
-)
-
-
-## ----echo=FALSE---------------------------------------------------------------
-# plot_patches <-
-fig_res_patches <-
-  ggplot() +
-  geom_sf(
[…]

### 2021-05-04 R/_05_calibration_data.R

@@ -1,488 +1,0 @@
-## ---------------------------------------------------------------------
-# load libs
-library(data.table)
-library(atlastools)
-library(ggplot2)
-library(patchwork)
-
-# prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
-
-
-## ---------------------------------------------------------------------
-# read and plot example data
-data <- fread("data/atlas1060_allTrials_annotated.csv")
-data_raw <- copy(data)
-
-
-## ----echo=FALSE-------------------------------------------------------
-# plot data
-fig_data_raw <-
-  ggplot(data)+
-  geom_path(aes(x, y),
-             col = "grey", alpha = 0.5, size = 0.3)+
-  geom_point(aes(x, y),
-             col = pal[3], alpha = 0.2, size = 2)+
-  ggthemes::theme_map()+
-  coord_sf(crs = 32631)
-
-# save figure
-ggsave(fig_data_raw, filename = "figures/fig_calibration_raw.png",
-       width = 185 / 25)
-
-
-## ---------------------------------------------------------------------
-data_unproc <- copy(data)
-
-
-## ---------------------------------------------------------------------
-# remove inside must be set to falses
-data <- atl_filter_bounds(data = data, 
-                          x = "x", y = "y", 
-                          x_range = c(645000, max(data$x)), 
-                          remove_inside = FALSE)
-
-
-## ----echo=FALSE-------------------------------------------------------
-# plot data
-fig_data_bbox <-
-  ggplot()+
-  geom_path(data = data_raw,
-            aes(x, y),
-             col = "grey", alpha = 0.5, size = 0.3)+
-  geom_point(data = data_raw,
-             aes(x, y,
-                 col = x > 645000),
-             alpha = ifelse(data_raw$x > 645000, 0.5, 1),
-             size = ifelse(data_raw$x > 645000, 0.3, 2),
-             show.legend = F)+
-  scale_colour_manual(values = c("grey", pal[3]))+
-  geom_vline(xintercept = 645000,
-             col = "grey", lty = 2)+
-  ggspatial::annotation_scale(location = "br")+
-  ggthemes::theme_few()+
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())+
-  coord_sf(crs = 32631)
-
-# save result
-ggsave(fig_data_bbox, filename = "figures/fig_calib_bbox.png",
-       width = 185 / 25)
-
-
-## ---------------------------------------------------------------------
-# divide by 1000, convert to integer, then convert to POSIXct
[…]

### 2021-05-04 R/helper_functions.R

@@ -1,44 +1,0 @@
-#### Functions to simulate and add errors to movement data ####
-
-# functions are not documented and are not intended for general use
-
-# a function to use smoove to simulate movement data
-do_smoove_data <- function(t_max = 100, dt = dt) {
-  # define parameters
-  nu <- 2
-  tau <- 1
-  dt <- .01 
-  ucvm1 <- smoove::simulateUCVM(nu = nu, tau = tau, T.max = 100, dt = dt)
-  
-  # get the data in data.table format
-  tmp_data <- data.table::as.data.table(ucvm1$XY)
-  # rescale coordinates
-  tmp_data[, c("x", "y") := (lapply(.SD, 
-                                scales::rescale)),
-       .SDcols = c("x", "y")]
-  
-  # add a time column
-  tmp_data[, time := seq_len(nrow(tmp_data))]
-
-  return(tmp_data)
-}
-
-# simple function to introduce normally distributed error
-do_add_error <- function(v, std_dev = 0.5) {
-  pos_error <- stats::rnorm(length(v), sd = std_dev)
-  return(v + pos_error)
-}
-
-# function for point outliers
-do_add_outliers <- function(data, p_data = 0.1, std_dev = 0.05) {
-  n_rows <- floor(p_data * nrow(data))
-  tmp_data <- data.table::copy(data)
-  
-  tmp_data[sample(nrow(data), n_rows),
-            c("x", "y") := lapply(.SD,
-                                  function(z) {
-                                    z + stats::rnorm(n_rows, sd = std_dev)
-                                  }),
-            .SDcols = c("x", "y")]
-  return(tmp_data)
-}

### 2021-05-04 supplement/06_bat_data.Rmd
move code to supplement
check for remotes

--- 2021-04-16 06_bat_data.Rmd
@@
 ```{r install_atlastools_2, eval=TRUE}
-install.packages("remotes")
+if (!require(remotes)) {
+  install.packages("remotes", repos = "http://cran.us.r-project.org")
+}
 
@@
 st_write(data_spatials,
-  dsn = "data/data_bat_residence_patches.gpkg"
+  dsn = "data/data_bat_residence_patches.gpkg",
+  append = FALSE
 )

### 2021-05-05 supplement/index.rmd

--- 
knit: "bookdown::render_book"
title: "Supplementary Material for _A Guide to Pre-processing High-Frequency Animal Tracking Data_"
author:
    - Pratik R. Gupte
    - Christine E. Beardsworth
    - Orr Spiegel
    - Emmanuel Lourie
    - Sivan Toledo
    - Ran Nathan
    - Allert I. Bijleveld
date: "`r Sys.Date()`"
output: bookdown::pdf_document2
documentclass: scrreprt
toc-depth: 1
secnumdepth: 2
geometry: "left=4cm, right=3cm, top=2.5cm, bottom=2.5cm"
bibliography: [references.bib]
link-citations: yes
github-repo: pratikunterwegs/atlas-best-practices
description: "Supplementary Material for _A Guide to Pre-processing High-Frequency Animal Tracking Data_"
---

### 2021-05-06 scripts/_05_calibration_data.R
rename

@@ -1,609 +1,0 @@
-## ---------------------------------------------------------------------
-# load libs
-library(data.table)
-library(atlastools)
-library(ggplot2)
-library(patchwork)
-
-# prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
-
-
-## ---------------------------------------------------------------------
-# read and plot example data
-data <- fread("data/atlas1060_allTrials_annotated.csv")
-data_raw <- copy(data)
-
-
-## ----echo=FALSE-------------------------------------------------------
-# plot data
-fig_data_raw <-
-  ggplot(data) +
-  geom_path(aes(x, y),
-    col = "grey", alpha = 0.5, size = 0.3
-  ) +
-  geom_point(aes(x, y),
-    col = pal[3], alpha = 0.2, size = 2
-  ) +
-  ggthemes::theme_map() +
-  coord_sf(crs = 32631)
-
-# save figure
-ggsave(fig_data_raw,
-  filename = "figures/fig_calibration_raw.png",
-  width = 185 / 25
-)
-
-
-## ---------------------------------------------------------------------
-data_unproc <- copy(data)
-
-
-## ---------------------------------------------------------------------
-# remove inside must be set to falses
-data <- atl_filter_bounds(
-  data = data,
-  x = "x", y = "y",
-  x_range = c(645000, max(data$x)),
-  remove_inside = FALSE
-)
-
-
-## ----echo=FALSE-------------------------------------------------------
-# plot data
-fig_data_bbox <-
-  ggplot() +
-  geom_path(
-    data = data_raw,
-    aes(x, y),
-    col = "grey", alpha = 0.5, size = 0.3
-  ) +
-  geom_point(
-    data = data_raw,
-    aes(x, y,
-      col = x > 645000
-    ),
-    alpha = ifelse(data_raw$x > 645000, 0.5, 1),
-    size = ifelse(data_raw$x > 645000, 0.3, 2),
-    show.legend = F
-  ) +
-  scale_colour_manual(values = c("grey", pal[3])) +
-  geom_vline(
-    xintercept = 645000,
-    col = "grey", lty = 2
-  ) +
-  ggspatial::annotation_scale(location = "br") +
-  ggthemes::theme_few() +
-  theme(
-    axis.text = element_blank(),
-    axis.title = element_blank()
-  ) +
-  coord_sf(crs = 32631)
-
-# save result
-ggsave(fig_data_bbox,
[…]

### 2021-05-06 copy_tex.sh
no copy tex

@@ -1,5 +1,3 @@
 #!/bin/bash
-# copy tex and replace old tex
-cp -p atlas-manuscript/arxiv_article.tex docs/ms_atlas_preproc_main_text.tex
-
+# transfer figures
 cp figures/fig_0*.png atlas-manuscript/figures

### 2021-05-06 render_books.sh
move to top level

@@ -1,5 +1,4 @@
 #!/bin/bash
-
 # remove old tex files and pdfs
 cd docs
 rm *.pdf
@@ -16,7 +15,7 @@ Rscript --vanilla --slave -e 'styler::style_dir("supplement", filetype = "Rmd", 
 Rscript --vanilla --slave -e 'bookdown::render_book("supplement/index.rmd", "bookdown::pdf_document2", output_dir = "docs", config_file = "supplement/_bookdown.yml")'
 
 # to render source files
-Rscript --slave -e 'lapply(list.files(pattern = "supplement\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
+Rscript --slave -e 'lapply(list.files(pattern = "supplement\\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
 
 # copy pdfs and name correctly
 cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text_`date -I`.pdf

### 2021-05-16 supplement/06_bat_data.Rmd
add figures for concept figure

@@
 
-## A First Visual Inspection
+## Exploratory Data Analysis
 
-Plot the bat data as a sanity check, and inspect it visually for errors (Fig. 2.1).
+Plot the bat data as a sanity check, and inspect it visually for errors.
 The plot code is hidden in the rendered copy (PDF) of this supplementary material, but is available in the `Rmarkdown` file "06_bat_data.Rmd".
-The saved plot is shown below as Fig. 2.1.
+
+### Heatmap of Locations
+
+```{r summarise_counts}
+data_heatmap = copy(data)
+data_heatmap[, c("xround", "yround") := list(
+  plyr::round_any(X, 200),
+  plyr::round_any(Y, 200)
+)]
+data_heatmap = data_heatmap[, .N, by = c("xround", "yround")]
+```
+
+```{r plot_heatmap}
+fig_heatmap = 
+ggplot(data_heatmap)+
+  geom_tile(
+    aes(
+      xround, yround, fill = N
+    ),
+    colour = "grey",
+    size = 0.1,
+    show.legend = F
+  )+
+  scale_fill_viridis_c(
+    option = "C",
+    direction = -1,
+    trans = "log10"
+  )+
+  ggthemes::theme_few() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank()
+  ) +
+  coord_sf(crs = 2039)
+
+ggsave(fig_heatmap,
+  filename = "supplement/figures/fig_bat_heatmap_raw.png",
+  dpi = 300,
+  width = 6, height = 4
+)
+```
+
+### Sampling Intervals
+
+```{r sampling_interval_check}
+# get intervals per individual
+data_for_intervals = data[,list(lags = list(diff(TIME/1000))), by = "TAG"]
+data_for_intervals = data_for_intervals[,list(lag = unlist(lags, recursive = F)), by = "TAG"]
+
+# make figure
+fig_lags = 
+ggplot(data_for_intervals)+
+  geom_histogram(
+    aes(
+      x = lag / 8
+    ),
+    bins = 15,
+    fill = "grey",
+    col = "black"
+  )+
+  geom_vline(
+    xintercept = 1,
+    col = "red"
+  )+
+  scale_x_log10(
+    labels = function(x) {sprintf("%ix", x)}
+  )+
+  scale_y_sqrt()+
+  ggthemes::theme_few() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.text.x = element_text(
+      face = "bold",
+      size = 16
+    ),
+    axis.title.y = element_blank(),
+    axis.title.x = element_text(
+      face = "bold",
+      size = 16
+    )
+  )+
+  labs(x = "Sampling interval")
+
+# save to figures
+ggsave(fig_lags,
+  filename = "supplement/figures/fig_lag_histogram.png",
+  dpi = 300,
+  width = 3, height = 4
+)
+```
+
+### Localisation Error Measured by Systems
+
+```{r localisation_error}
+fig_error_histogram = 
+ggplot(data)+
+  geom_histogram(
+    aes(
+      x = sqrt(VARX)
+    ),
+    bins = 15,
+    fill = "grey",
+    col = "black"
+  )+
+  scale_x_log10(
+    labels = function(x) {sprintf("%2.fm", x)}
+  )+
+  scale_y_sqrt()+
+  coord_cartesian(
+    xlim = c(0.5, 5e3)
+  )+
+  ggthemes::theme_few() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.text.x = element_text(
+      face = "bold",
+      size = 16
+    ),
+    axis.title.y = element_blank(),
+    axis.title.x = element_text(
+      face = "bold",
+      size = 16
+    )
+  )+
+  labs(x = "Variance X")
+
+# save to figures
+ggsave(fig_error_histogram,
+  filename = "supplement/figures/fig_error_histogram.png",
+  dpi = 300,
+  width = 3, height = 4
+)
+```
 
@@
 ggsave(fig_bat_raw,
-  filename = "figures/fig_bat_raw.png",
+  filename = "supplement/figures/fig_bat_raw.png",
   dpi = 300,
@@
 The bats were tracked in the Hula Valley, Israel (33.1$^{\circ}$N, 35.6$^{\circ}$E), and we use three nights of tracking (5\textsuperscript{th}, 6\textsuperscript{th}, and 7\textsuperscript{th} May, 2018), for our demonstration, with an average of 13,370 positions (SD = 2,173; range = 11,195 -- 15,542; interval = 8 seconds) per individual.
-After first plotting the individual tracks, we notice severe distortions, making pre-processing necesary](figures/fig_bat_raw.png)
+After first plotting the individual tracks, we notice severe distortions, making pre-processing necesary](supplement/figures/fig_bat_raw.png)
 
@@
     ),
-    size = 0.3,
     show.legend = FALSE
@@
 ggsave(fig_bat_filter_cov,
-  filename = "figures/fig_bat_filter_cov.png",
+  filename = "supplement/figures/fig_bat_filter_cov.png",
   dpi = 300,
@@
 Both steps used the function \texttt{atl\_filter\_covariates}.
-This filtering reduced the data to an average of 10,447 positions per individual (78% of the raw data on average). However, some point outliers remain.](figures/fig_bat_filter_cov.png)
+This filtering reduced the data to an average of 10,447 positions per individual (78% of the raw data on average). However, some point outliers remain.](supplement/figures/fig_bat_filter_cov.png)
 
@@
 ggsave(fig_bat_filter_speed,
-  filename = "figures/fig_bat_filter_speed.png",
+  filename = "supplement/figures/fig_bat_filter_speed.png",
   dpi = 300,
@@
 Grey crosses show data that were removed.
-We calculated the incoming and outgoing speed of each position using `atl_get_speed`, and filtered out positions with speeds > 20 m/s using `atl_filter_covariates`, leaving 10,337 positions per individual on average (98% from the previous step).](figures/fig_bat_filter_speed.png)
+We calculated the incoming and outgoing speed of each position using `atl_get_speed`, and filtered out positions with speeds > 20 m/s using `atl_filter_covariates`, leaving 10,337 positions per individual on average (98% from the previous step).](supplement/figures/fig_bat_filter_speed.png)
 
@@
   geom_point(
-    data = data_split,
+    data = data,
     aes(X, Y),
     col = "grey",
-    shape = 4,
+    shape = 1,
+    stroke = 0.3,
     show.legend = FALSE
@@
     ),
-    lwd = 0.3,
+    lwd = 0.6,
     show.legend = FALSE
@@
 ggsave(fig_bat_smooth,
-  filename = "figures/fig_bat_smooth.png",
+  filename = "supplement/figures/fig_bat_smooth.png",
   dpi = 300,
@@
 
-![Bat data after applying a median smooth with a moving window $K$ = 5. Grey crosses show data prior to smoothing. The smoothing step did not discard any data.](figures/fig_bat_smooth.png)
+![Bat data after applying a median smooth with a moving window $K$ = 5. Grey crosses show data prior to smoothing. The smoothing step did not discard any data.](supplement/figures/fig_bat_smooth.png)
 
@@
 ggsave(fig_bat_res_timeseries,
-  filename = "figures/fig_bat_res_timeseries.png",
+  filename = "supplement/figures/fig_bat_res_timeseries.png",
   dpi = 300,
@@
 When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar **(c)**.
-Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_07_bats.png)
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](supplement/figures/fig_07_bats.png)

### 2021-05-18 supplement/06_bat_data.Rmd
some edits to bats

@@
 data_heatmap[, c("xround", "yround") := list(
-  plyr::round_any(X, 200),
-  plyr::round_any(Y, 200)
+  plyr::round_any(X, 250),
+  plyr::round_any(Y, 250)
 )]
@@
     option = "C",
-    direction = -1,
+    direction = 1,
     trans = "log10"
@@
     aes(
-      x = lag / 8
+      x = lag / 8,
+      fill = (lag / 8 > 3)
     ),
+    position = position_identity(),
     bins = 15,
-    fill = "grey",
-    col = "black"
+    # fill = "grey",
+    col = "black",
+    show.legend = F
   )+
@@
   scale_x_log10(
-    labels = function(x) {sprintf("%ix", x)}
+    labels = function(x) {sprintf("%ix", x)},
+    breaks = c(1, 3, 10, 100, 1000)
   )+
   scale_y_sqrt()+
+  scale_fill_viridis_d(
+    option = "H",
+    begin = 0.1,
+    end = 0.8
+  )+
   ggthemes::theme_few() +
@@
     aes(
-      x = sqrt(VARX)
+      x = sqrt(VARX),
+      fill = (sqrt(VARX) > 100)
     ),
     bins = 15,
-    fill = "grey",
-    col = "black"
+    # fill = "grey",
+    col = "black",
+    show.legend = F
+  )+
+  scale_fill_viridis_d(
+    option = "H",
+    begin = 0.1,
+    end = 0.8
   )+
@@
   ggplot() +
-  geom_point(
-    data = data[!data_split,
-      on = c("TAG", "X", "Y", "TIME")
-    ],
-    aes(X, Y),
-    col = "grey",
-    shape = 4
+  geom_path(
+    data = data,
+    aes(X, Y, group = TAG),
+    col = "black",
+    size = 0.3
   ) +
@@
     ),
-    size = 0.3,
+    # col = "grey20",
+    # size = 0.3,
     show.legend = FALSE

### 2021-05-19 supplement/06_bat_data.Rmd
inset fig for concept fig

@@
 # now plot over raw data
-fig_bat_filter_cov <-
+# fig_bat_filter_cov <-
   ggplot() +
@@
     data = data,
-    aes(X, Y, group = TAG),
-    col = "black",
-    size = 0.3
-  ) +
-  geom_path(
-    data = data_split,
-    aes(X, Y,
-      col = TAG
-    ),
-    # col = "grey20",
-    # size = 0.3,
-    show.legend = FALSE
+    aes(X, Y, group = TAG, col = TAG),
+    show.legend = F
   ) +

### 2021-05-20 supplement/05_calibration_data.Rmd
check for remotes
new figure output paths

--- 2021-02-16 05_calibration_data.Rmd
@@
 
-```{r install_atlastools, eval=FALSE}
-install.packages("remotes")
+```{r install_atlastools, eval=TRUE}
+if (!require(remotes)) {
+  install.packages("remotes", repos = "http://cran.us.r-project.org")
+}
 
@@
 ggsave(fig_data_raw,
-  filename = "figures/fig_calibration_raw.png",
+  filename = "supplement/figures/fig_calibration_raw.png",
   width = 185 / 25
@@
 A handheld WATLAS tag was used to examine how ATLAS data compared to GPS tracks, and we use the WATLAS data here to demonstrate the basics of the pre-processing pipeline, as well as validate the residence patch method.
-It is immediately clear from the figure that the track shows location errors, both in the form of point outliers as well as small-scale errors around the true location.](figures/fig_calibration_raw.png)
+It is immediately clear from the figure that the track shows location errors, both in the form of point outliers as well as small-scale errors around the true location.](supplement/figures/fig_calibration_raw.png)
 
@@
 ggsave(fig_data_bbox,
-  filename = "figures/fig_calib_bbox.png",
+  filename = "supplement/figures/fig_calib_bbox.png",
   width = 185 / 25
@@
 The point outlier (black point) is removed based on its X coordinate value, with the data filtered to exclude positions with an X coordinate < 645,000 in the UTM 31N coordinate system.
-Positions that are retained are shown in green.](figures/fig_calib_bbox.png)
+Positions that are retained are shown in green.](supplement/figures/fig_calib_bbox.png)
 
@@
 
-### Get 95th percentile of speed and angle
+### Get 90th percentile of speed and angle
 
@@
   coord_sf(crs = 32631)
+
 # save
 ggsave(fig_speed_outliers,
-  filename = "figures/fig_speed_outlier.png",
+  filename = "supplement/figures/fig_speed_outlier.png",
   width = 170 / 25, height = 170 / 25
@@
 Grey positions are removed, while green positions are retained.
-Rectangles indicate areas expanded for visualisation in following figures.](figures/fig_speed_outlier.png)
+Rectangles indicate areas expanded for visualisation in following figures.](supplement/figures/fig_speed_outlier.png)
 
@@
 ggsave(fig_smooth,
-  filename = "figures/fig_calib_median_smooth.png",
+  filename = "supplement/figures/fig_calib_median_smooth.png",
   width = 90 / 25, height = 90 / 25
@@
 The area shown is the upper rectangle from Fig.
-1.3.](figures/fig_calib_median_smooth.png)
+1.3.](supplement/figures/fig_calib_median_smooth.png)
 
@@
 ggsave(fig_smooth_thin,
-  filename = "figures/fig_calib_smooth_thin.png",
+  filename = "supplement/figures/fig_calib_smooth_thin.png",
   width = 90 / 25, height = 90 / 25, dpi = 300
@@
 ggsave(figure_walkthrough,
-  filename = "figures/fig_walkthrough.png",
+  filename = "supplement/figures/fig_walkthrough.png",
   height = 90, width = 170, units = "mm"
@@
 The area shown is the upper rectangle in Fig.
-1.3.](figures/fig_calib_smooth_thin.png)
+1.3.](supplement/figures/fig_calib_smooth_thin.png)
 

### 2021-05-20 supplement/05_calibration_data.Rmd
res patch using smoothed speed

@@
 
-First, we filter data where we know the animal spent some time at or near a position, as this is the first step to identify residence patches. 
-One way of doing this is to filter out positions with speeds above which the animal is likely to be in transit, but a better way of doing this by filtering on a metric such as residence time [@bracis2018].
+First, we filter data where we know the animal (or in this case, the human-carried tag) spent some time at or near a position, as this is the first step to identify residence patches. 
+One way of doing this is to filter out positions with speeds above which the tag (ideally on an animal) is likely to be in transit.
+Rather than filtering on instantaneous speed estimates, filtering on a median smoothed speed estimate is more reliable.
 
-### Calculate residence time
+### Exclude transit points
 
-First we calculate the residence time with a radius of 50 metres.
-For this, we need a dataframe with coordinates, the timestamp, and the animal id.
-We save this data to file for later use.
+Here, we aim to remove locations where the tag is clearly moving, by filtering on smoothed speed, using a one-way median smooth with $K$ = 5.
+The speeds between points must be recalculated here because the speed metrics now associated with the data refer to the raw data before median smoothing.
 
@@
 # get 4 column data
-data_for_patch <- data_thin[, list(x, y, time, TAG)]
+data_for_patch <- copy(data_thin)
 
-# get recurse data for a 10m radius
-recurse_stats <- getRecursions(data_for_patch,
-  radius = 50, timeunits = "mins"
-)
+# recalculate speeds, removing speed out
+data_for_patch[, c("speed_in", "speed_out") := list(
+  atl_get_speed(data_for_patch),
+  NULL
+)]
 
-# assign to recurse data
-data_for_patch[, res_time := recurse_stats$residenceTime]
+# get smoothed speed
+data_for_patch[, speed_smooth := runmed(speed_in, k = 5)]
 
@@
 
-We subset data with a residence time > 5 minutes in order to construct residence patches.
+We subset data with a smoothed speed < 2 m/s in order to construct residence patches.
 From this subset, we construct residence patches using the parameters: `buffer_radius` = 5 metres, `lim_spat_indep` = 50 metres, `lim_time_indep` = 5 minutes, and `min_fixes` = 3.
@@
 # on known residence points
-patch_res_known <- atl_res_patch(data_for_patch[res_time >= 5, ],
+patch_res_known <- atl_res_patch(
+  data = data_for_patch[speed_smooth < 2, ],
   buffer_radius = 5,

### 2021-05-20 supplement/05_calibration_data.Rmd
minor updates to caption and figures

@@
 ggsave(fig_basic_residence,
-  filename = "figures/fig_calib_residence.png",
+  filename = "supplement/figures/fig_calib_residence.png",
   height = 170, width = 80, units = "mm"
@@
 However, the algorithm identified all areas with prolonged residence, including those which were not intended stops (n = 12; green polygons without triangles).
-The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (triangle without green polygon shows one of these).
+The field station on Griend (red triangle) was not intended to be a stop, but the tags were stored here before the trial, and the method correctly picked up this prolonged stationary data as a residence patch.
+The algorithm failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (purple triangle without green polygon shows one of these).
 The area shown is the lower rectangle in Fig.
-1.3.](figures/fig_calib_residence.png)
+1.3.](supplement/figures/fig_calib_residence.png)
 
@@
     col = pal[1],
-    lwd = 0.5,
     method = "lm",
@@
   scale_x_continuous(
-    breaks = 60 * seq(3, 18, 2),
-    labels = seq(3, 18, 2)
+    breaks = 60 * seq(5, 20, 5),
+    labels = seq(5, 20, 5)
   ) +
   scale_y_continuous(
-    breaks = 60 * seq(5, 20, 2),
-    labels = seq(5, 20, 2)
+    breaks = 60 * seq(5, 20, 5),
+    labels = seq(5, 20, 5)
   ) +
   coord_fixed(
-    expand = T,
-    ratio = 1.115
-    #             # xlim = c(400, 1050),
-    #             ylim = c(250, 1050)
+    expand = F,
+    xlim = c(5, 20) * 60,
+    ylim = c(5, 20) * 60,
+    ratio = 1.0
   ) +
@@
   ) +
-  ggthemes::theme_few() +
+  theme_classic() +
   labs(
-    x = "inferred duration (min)",
-    y = "real duration (min)"
+    x = "Inferred duration (min)",
+    y = "Real duration (min)"
   )
@@
 ggsave(figure_lm_duration,
-  filename = "figures/fig_calib_lm_duration.png",
+  filename = "supplement/figures/fig_calib_lm_duration.png",
   width = 80 / 25, height = 80 / 25
@@
 ![The inferred duration of residence patches corresponds very closely to the real duration (grey circles, red line shows linear model fit), with an underestimation of the true duration of around 2%.
-The dashed black line represents $y = x$ for reference.](figures/fig_calib_lm_duration.png)
+The dashed black line represents $y = x$ for reference.](supplement/figures/fig_calib_lm_duration.png)
 
@@
 ggsave(figure_res_patch,
-  filename = "figures/fig_06_calib_residence_patch.png",
+  filename = "figures/fig_06.png",
   height = 170, width = 170, units = "mm"

### 2021-05-20 supplement/06_bat_data.Rmd
minor mods to bats code

@@
 
-## Exploratory Data Analysis
+## Exploratory Data Analysis Panels for Main Text Figure 1
+
+Here, we make some basic figures for exploratory data analysis shown in Figure 1 of the main text.
 
 Plot the bat data as a sanity check, and inspect it visually for errors.
-The plot code is hidden in the rendered copy (PDF) of this supplementary material, but is available in the `Rmarkdown` file "06_bat_data.Rmd".
+The plot code is hidden in the rendered copy (PDF) of this supplementary material, but is available in the `Rmarkdown` file "supplement/06_bat_data.Rmd".
 
 ### Heatmap of Locations
+
+Here we demonstrate a basic heatmap of locations, aggregating over all individuals.
+In this one instance, the plotting code is also shown as a guide for readers, but in general, plotting code is hidden throughout this document.
 
@@
 
-```{r sampling_interval_check}
+Here, we create the histogram of sampling intervals shown in Figure 1 of the main text.
+The plotting code is hidden in the PDF version, but available in the source code.
+
+```{r sampling_interval_check, echo=FALSE}
 # get intervals per individual
@@
 
-```{r localisation_error}
+Here, we create the histogram of location error (variance in X) [@weiser2016] shown in Figure 1 of the main text.
+The plotting code is hidden in the PDF version, but available in the source code.
+
+```{r localisation_error, echo=FALSE}
 fig_error_histogram = 
@@
 ```{r echo=FALSE}
+
+```
+
+### Plot paths from raw tracking data
+
+Here, we plot the paths of individual bats from the raw tracking data to visually inspect them for errors.
+
+```{r echo=FALSE}
 # plot bat data to examine for errors
@@
   coord_sf(crs = 2039)
+```
+
+```{r plot_focus_bad_speed}
+# this figure for the panel in main text figure 1
+fig_bat_focus_bad_speed = 
+  fig_bat_raw +
+  coord_sf(
+    crs = 2039,
+    xlim = c(253000, NA),
+    ylim = c(772000, NA)
+  )
+
+# save to supplement figures
+ggsave(fig_bat_focus_bad_speed,
+  filename = "supplement/figures/fig_bat_focus_bad_speed.png",
+  dpi = 300,
+  width = 4, height = 6
+)
 ```
@@
 into a list of data.tables to avoid mixups among individuals.
+
+This is a very rudimentary demonstration of the principle behind **batch processing** --- splitting data into smaller, independent subsets, and applying the same steps to each subset.
 
@@
 We apply the filter to each individual's data using an `lapply`.
+
+This is another basic example of the principles behind batch-processing, and could be parallelised using the `R` package `furrr` (see https://CRAN.R-project.org/package=furrr).
 
@@
   ggplot() +
+  geom_point(
+    data = data[!data_split,
+      on = c("TAG", "X", "Y", "TIME")
+    ],
+    aes(X, Y),
+    col = "grey",
+    shape = 4
+  ) +
   geom_path(
-    data = data,
+    data = data_split,
     aes(X, Y, group = TAG, col = TAG),
@@
 
-![Bat data after applying a median smooth with a moving window $K$ = 5. Grey crosses show data prior to smoothing. The smoothing step did not discard any data.](supplement/figures/fig_bat_smooth.png)
+![Bat data after applying a median smooth with a moving window $K$ = 5. Grey circles show data prior to smoothing. The smoothing step did not discard any data.](supplement/figures/fig_bat_smooth.png)
 

### 2021-05-20 supplement/06_bat_data.Rmd
use smoothed speed not recurse

@@
 
-### Calculating residence time
+### Excluding transit points
 
-First, the data is put through the `recurse` package to get residence time [@bracis2018].
+First, we calculate the instantaneous speeds, and then apply a one-way median smooth to the speeds with $K$ = 5. This procedure is also applied to each individual *separately*.
 
@@
 # split the data
-data_smooth <- split(data_smooth, data_smooth$TAG)
+data_residence <- split(data_smooth, data_smooth$TAG)
 ```
 
-We calculated residence time, but since bats may revisit the same features, we want to prevent confusion between frequent revisits and prolonged residence.
-
-For this, we stop summing residence times within $Z$ metres of a location if the animal exited the area for one hour or more. The value of $Z$ (radius, in `recurse` parameter terms) was chosen as 50m.
-
-This step is relatively complicated and is only required for individuals which frequently return to the same location, or pass over the same areas repeatedly, and for which revisits (cumulative time spent) may be confused for residence time in a single visit.
-
-While a simpler implementation using total residence time divided by the number of revisits is also possible, this does assume that each revisit had the same residence time.
+We apply this step as with the calibration data in the previous example.
 
 ```{r}
-# get residence times
+# get smoothed speeds per individual
+data_residence <- lapply(data_residence, function(dt) {
 
-data_residence <- lapply(data_smooth, function(dt) {
-  # do basic recurse
-  dt_recurse <- getRecursions(
-    x = dt[, c("X", "Y", "time", "TAG")],
-    radius = 50,
-    timeunits = "mins"
-  )
+  # first get the speed (in) between smoothed points
+  dt[, c("speed_in", "speed_out") := list(
+    atl_get_speed(dt, x = "X", y = "Y", time = "time"),
+    NULL
+  )]
 
-  # get revisit stats
-  dt_recurse <- setDT(
-    dt_recurse[["revisitStats"]]
-  )
+  # then smooth the speeds in a new column called speed_smoothed
+  dt[, speed_smoothed := runmed(speed_in, k = 5)]
 
-  # count long absences from the area
-  dt_recurse[, timeSinceLastVisit :=
-    ifelse(is.na(timeSinceLastVisit), -Inf, timeSinceLastVisit)]
-  dt_recurse[, longAbsenceCounter := cumsum(timeSinceLastVisit > 60),
-    by = .(coordIdx)
-  ]
-  # get data before the first long absence of 60 mins
-  dt_recurse <- dt_recurse[longAbsenceCounter < 1, ]
+  # filter for speeds less than 2 m/s
+  dt = dt[speed_smoothed < 2, ]
 
-  dt_recurse <- dt_recurse[, list(
-    resTime = sum(timeInside),
-    fpt = first(timeInside),
-    revisits = max(visitIdx)
-  ),
-  by = .(coordIdx, x, y)
-  ]
-
-  # prepare and merge existing data with recursion data
-  dt[, coordIdx := seq(nrow(dt))]
-
-  dt <- merge(dt,
-    dt_recurse[, c("coordIdx", "resTime")],
-    by = c("coordIdx")
-  )
-
-  setorderv(dt, "time")
+  dt
 })
@@
     data = data_residence,
-    aes(ts, resTime,
+    aes(ts, speed_smoothed,
       colour = TAG
@@
     x = NULL,
-    y = "residence time (mins)"
+    y = "Speed (m/s)"
   )
@@
 Some preparation is required. First, the function requires columns `x`, `y`,
-`time`, and `id`, which we assign using the `data.table` syntax.
+`time`, and `id`, which we assign using the `data.table` syntax. The `time` column is already present, but the other columns need to be renamed to lower case.
 Then we subset the data to only work with positions where the individual had a residence time of more than 5 minutes.
@@
 
-# filter for residence time > 5 minutes
-data_residence <- data_residence[resTime > 5, ]
-
 # split the data
@@
 We apply the residence patch method, using the default argument values (`lim_spat_indep = 100` (metres), `lim_time_indep = 30` (minutes), and `min_fixes = 3`).
-We change the `buffer_radius` to 25 metres (twice the buffer radius is used, so points must be separated by 50m to be independent bouts).
+We change the `buffer_radius` to 30 metres (twice the buffer radius is used, so points must be separated by 50m to be independent bouts).
 
@@
 data_patches <- lapply(data_residence, atl_res_patch,
-  buffer_radius = 25
+  buffer_radius = 30
 )
@@
   which_data = "spatial",
-  buffer_radius = 25
+  buffer_radius = 30
 )
@@
 When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar **(c)**.
-Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](supplement/figures/fig_07_bats.png)
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_07.png)

### 2021-05-20 scripts/02_filtering_data.R
relabel export figures

@@ -320,7 +320,7 @@ figure_01 <-
 
 # save figure
 ggsave(figure_01,
-  filename = "figures/fig_02_filtering_data.png",
+  filename = "figures/fig_03.png",
   width = 170, height = 150, units = "mm"
 )
 

### 2021-05-20 scripts/04_residence_patch.R

@@ -202,6 +202,6 @@ fig_residence_patch <-
 
 # save the figure
 ggsave(fig_residence_patch,
-  filename = "figures/fig_05_residence.png",
+  filename = "figures/fig_06.png",
   height = 170, width = 150, units = "mm"
 )

### 2021-05-20 supplement/06_bat_data.Rmd
return to using recurse

--- 2021-05-20 supplement/06_bat_data.Rmd
@@
   setorderv(dt, "time")
+
+  # print message when done
+  message(sprintf("TAG %i residence times done", unique(dt$TAG)))
 })
@@
 
-```{r echo=FALSE, eval=TRUE}
-# not shown or mentioned
-fig_bat_res_timeseries <-
-  ggplot() +
-  geom_path(
-    data = data_residence,
-    aes(ts, resTime,
-      colour = TAG
-    ),
-    show.legend = F
-  ) +
-  scale_color_brewer(palette = "Dark2") +
-  scale_x_datetime() +
-  ggthemes::theme_few() +
-  labs(
-    x = NULL,
-    y = "residence time (mins)"
-  )
-```
-
-```{r echo=FALSE}
-# not shown or mentioned
-# save to figures
-ggsave(fig_bat_res_timeseries,
-  filename = "supplement/figures/fig_bat_res_timeseries.png",
-  dpi = 300,
-  width = 6, height = 3
-)
-```
-
 ### Constructing residence patches
@@
 
-We apply the residence patch method, using the default argument values (`lim_spat_indep = 100` (metres), `lim_time_indep = 30` (minutes), and `min_fixes = 3`).
-We change the `buffer_radius` to 25 metres (twice the buffer radius is used, so points must be separated by 50m to be independent bouts).
+We apply the residence patch method, using the default argument values (`lim_spat_indep = 100` (metres), `lim_time_indep = 30` (minutes)).
+We change the `buffer_radius` to 25 metres (twice the buffer radius is used, so points must be separated by 50m to be independent bouts), and `min_fixes = 3`.
 
@@
 data_patches <- lapply(data_residence, atl_res_patch,
-  buffer_radius = 25
+  buffer_radius = 25,
+  min_fixes = 3
 )

### 2021-05-21 render_books.sh
edits

@@ -5,6 +5,9 @@ rm *.pdf
 
 cd ..
 
+# copy pdfs and name correctly
+cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text_`date -I`.pdf
+
 cp figures/fig_0*.png atlas-manuscript/figures
 
 # style files
@@ -17,8 +20,6 @@ Rscript --vanilla --slave -e 'bookdown::render_book("supplement/index.rmd", "boo
 # to render source files
 Rscript --slave -e 'lapply(list.files(pattern = "supplement\\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
 
-# copy pdfs and name correctly
-cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text_`date -I`.pdf
 mv docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_`date -I`.pdf
 
 # build atlastools manual

### 2021-05-21 copy_pdfs.sh
makes docs

@@ -1,0 +1,12 @@
+#!/bin/bash
+# remove old tex files and pdfs
+
+cd docs
+rm *.pdf
+
+cd ..
+
+# copy pdfs and name correctly
+cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text_`date -I`.pdf
+
+pandoc atlas-manuscript/arxiv_article.tex --bibliography=atlas-manuscript/references.bib -o docs/ms_preprocessing_`date -I`.docx

### 2021-05-21 render_books.sh
remove copy funs

@@ -1,14 +1,5 @@
 #!/bin/bash
 # remove old tex files and pdfs
-cd docs
-rm *.pdf
-
-cd ..
-
-# copy pdfs and name correctly
-cp -p atlas-manuscript/arxiv_article.pdf docs/ms_atlas_preproc_main_text_`date -I`.pdf
-
-cp figures/fig_0*.png atlas-manuscript/figures
 
 # style files
 Rscript --vanilla --slave -e 'styler::style_dir("scripts", filetype = "R", recursive = FALSE)'
@@ -24,3 +15,5 @@ mv docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_
 
 # build atlastools manual
 Rscript --slave -e 'devtools::build_manual(pkg = "../atlastools", path = "docs/")'
+
+pandoc atlas-manuscript/arxiv_article.tex --bibliography=atlas-manuscript/references.bib -o docs/ms_preprocessing_`date -I`.docx

### 2021-06-06 supplement/06_bat_data.Rmd
correct recurse code inline function

@@
 ```{r summarise_counts}
-data_heatmap = copy(data)
+data_heatmap <- copy(data)
 data_heatmap[, c("xround", "yround") := list(
@@
 )]
-data_heatmap = data_heatmap[, .N, by = c("xround", "yround")]
+data_heatmap <- data_heatmap[, .N, by = c("xround", "yround")]
 ```
@@
 ```{r plot_heatmap}
-fig_heatmap = 
-ggplot(data_heatmap)+
+fig_heatmap <-
+  ggplot(data_heatmap) +
   geom_tile(
     aes(
-      xround, yround, fill = N
+      xround, yround,
+      fill = N
     ),
@@
     show.legend = F
-  )+
+  ) +
   scale_fill_viridis_c(
@@
     trans = "log10"
-  )+
+  ) +
   ggthemes::theme_few() +
@@
 # get intervals per individual
-data_for_intervals = data[,list(lags = list(diff(TIME/1000))), by = "TAG"]
-data_for_intervals = data_for_intervals[,list(lag = unlist(lags, recursive = F)), by = "TAG"]
+data_for_intervals <- data[, list(lags = list(diff(TIME / 1000))), by = "TAG"]
+data_for_intervals <- data_for_intervals[, list(lag = unlist(lags, recursive = F)), by = "TAG"]
 
 # make figure
-fig_lags = 
-ggplot(data_for_intervals)+
+fig_lags <-
+  ggplot(data_for_intervals) +
   geom_histogram(
@@
     show.legend = F
-  )+
+  ) +
   geom_vline(
@@
     col = "red"
-  )+
+  ) +
   scale_x_log10(
-    labels = function(x) {sprintf("%ix", x)},
+    labels = function(x) {
+      sprintf("%ix", x)
+    },
     breaks = c(1, 3, 10, 100, 1000)
-  )+
-  scale_y_sqrt()+
+  ) +
+  scale_y_sqrt() +
   scale_fill_viridis_d(
@@
     end = 0.8
-  )+
+  ) +
   ggthemes::theme_few() +
@@
     )
-  )+
+  ) +
   labs(x = "Sampling interval")
@@
 ```{r localisation_error, echo=FALSE}
-fig_error_histogram = 
-ggplot(data)+
+fig_error_histogram <-
+  ggplot(data) +
   geom_histogram(
@@
     show.legend = F
-  )+
+  ) +
   scale_fill_viridis_d(
@@
     end = 0.8
-  )+
+  ) +
   scale_x_log10(
-    labels = function(x) {sprintf("%2.fm", x)}
-  )+
-  scale_y_sqrt()+
+    labels = function(x) {
+      sprintf("%2.fm", x)
+    }
+  ) +
+  scale_y_sqrt() +
   coord_cartesian(
     xlim = c(0.5, 5e3)
-  )+
+  ) +
   ggthemes::theme_few() +
@@
     )
-  )+
+  ) +
   labs(x = "Variance X")
@@
 )
-```
-
-```{r echo=FALSE}
-
 ```
@@
 # this figure for the panel in main text figure 1
-fig_bat_focus_bad_speed = 
+fig_bat_focus_bad_speed <-
   fig_bat_raw +
@@
 # get residence times
-
 data_residence <- lapply(data_smooth, function(dt) {
@@
   # print message when done
-  message(sprintf("TAG %i residence times done", unique(dt$TAG)))
+  message(sprintf("TAG %s residence times done", unique(dt$TAG)))
+
+  dt
 })

### 2021-06-06 supplement/05_calibration_data.Rmd
relabel export figures
stringi to top

@@
 library(atlastools)
+library(stringi)
 
@@
 ```{r}
-library(stringi)
 data_res <- data_unproc[stri_detect(tID, regex = "(WP)")]
@@
 ggsave(figure_res_patch,
-  filename = "figures/fig_06.png",
+  filename = "figures/fig_07.png",
   height = 170, width = 170, units = "mm"

### 2021-06-06 render_books.sh
correct render script

@@ -6,7 +6,7 @@ Rscript --vanilla --slave -e 'styler::style_dir("scripts", filetype = "R", recur
 Rscript --vanilla --slave -e 'styler::style_dir("supplement", filetype = "Rmd", recursive = FALSE)'
 
 # render supplement
-Rscript --vanilla --slave -e 'bookdown::render_book("supplement/index.rmd", "bookdown::pdf_document2", output_dir = "docs", config_file = "supplement/_bookdown.yml")'
+Rscript --vanilla --slave -e 'bookdown::render_book("supplement/")'
 
 # to render source files
 Rscript --slave -e 'lapply(list.files(pattern = "supplement\\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'

### 2021-06-06 supplement/05_calibration_data.Rmd
correct figure locations

@@
 A handheld WATLAS tag was used to examine how ATLAS data compared to GPS tracks, and we use the WATLAS data here to demonstrate the basics of the pre-processing pipeline, as well as validate the residence patch method.
-It is immediately clear from the figure that the track shows location errors, both in the form of point outliers as well as small-scale errors around the true location.](supplement/figures/fig_calibration_raw.png)
+It is immediately clear from the figure that the track shows location errors, both in the form of point outliers as well as small-scale errors around the true location.](figures/fig_calibration_raw.png)
 
@@
 The point outlier (black point) is removed based on its X coordinate value, with the data filtered to exclude positions with an X coordinate < 645,000 in the UTM 31N coordinate system.
-Positions that are retained are shown in green.](supplement/figures/fig_calib_bbox.png)
+Positions that are retained are shown in green.](figures/fig_calib_bbox.png)
 
@@
 Grey positions are removed, while green positions are retained.
-Rectangles indicate areas expanded for visualisation in following figures.](supplement/figures/fig_speed_outlier.png)
+Rectangles indicate areas expanded for visualisation in following figures.](figures/fig_speed_outlier.png)
 
@@
 The area shown is the upper rectangle from Fig.
-1.3.](supplement/figures/fig_calib_median_smooth.png)
+1.3.](figures/fig_calib_median_smooth.png)
 
@@
 The area shown is the upper rectangle in Fig.
-1.3.](supplement/figures/fig_calib_smooth_thin.png)
+1.3.](figures/fig_calib_smooth_thin.png)
 
@@
 The area shown is the lower rectangle in Fig.
-1.3.](supplement/figures/fig_calib_residence.png)
+1.3.](figures/fig_calib_residence.png)
 
@@
 ![The inferred duration of residence patches corresponds very closely to the real duration (grey circles, red line shows linear model fit), with an underestimation of the true duration of around 2%.
-The dashed black line represents $y = x$ for reference.](supplement/figures/fig_calib_lm_duration.png)
+The dashed black line represents $y = x$ for reference.](figures/fig_calib_lm_duration.png)
 

### 2021-06-06 supplement/06_bat_data.Rmd
correct figure locations

@@
 The bats were tracked in the Hula Valley, Israel (33.1$^{\circ}$N, 35.6$^{\circ}$E), and we use three nights of tracking (5\textsuperscript{th}, 6\textsuperscript{th}, and 7\textsuperscript{th} May, 2018), for our demonstration, with an average of 13,370 positions (SD = 2,173; range = 11,195 -- 15,542; interval = 8 seconds) per individual.
-After first plotting the individual tracks, we notice severe distortions, making pre-processing necesary](supplement/figures/fig_bat_raw.png)
+After first plotting the individual tracks, we notice severe distortions, making pre-processing necesary](figures/fig_bat_raw.png)
 
@@
 Both steps used the function \texttt{atl\_filter\_covariates}.
-This filtering reduced the data to an average of 10,447 positions per individual (78% of the raw data on average). However, some point outliers remain.](supplement/figures/fig_bat_filter_cov.png)
+This filtering reduced the data to an average of 10,447 positions per individual (78% of the raw data on average). However, some point outliers remain.](figures/fig_bat_filter_cov.png)
 
@@
 Grey crosses show data that were removed.
-We calculated the incoming and outgoing speed of each position using `atl_get_speed`, and filtered out positions with speeds > 20 m/s using `atl_filter_covariates`, leaving 10,337 positions per individual on average (98% from the previous step).](supplement/figures/fig_bat_filter_speed.png)
+We calculated the incoming and outgoing speed of each position using `atl_get_speed`, and filtered out positions with speeds > 20 m/s using `atl_filter_covariates`, leaving 10,337 positions per individual on average (98% from the previous step).](figures/fig_bat_filter_speed.png)
 
@@
 
-![Bat data after applying a median smooth with a moving window $K$ = 5. Grey circles show data prior to smoothing. The smoothing step did not discard any data.](supplement/figures/fig_bat_smooth.png)
+![Bat data after applying a median smooth with a moving window $K$ = 5. Grey circles show data prior to smoothing. The smoothing step did not discard any data.](figures/fig_bat_smooth.png)
 
@@
 When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar **(c)**.
-Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_07.png)
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](../figures/fig_07.png)

### 2021-06-11 supplement/05_calibration_data.Rmd
edits per reviewer comments

@@
 
-`data_frame[, c("col_a", "col_b) := list(values_a, values_b)]`
+`data_frame[, c("col_a", "col_b") := list(values_a, values_b)]`
 
@@
 First we access the data from a local file using the `data.table` package [@dowle2020].
-We look at the first few rows, using `head()`.
-We then visualise the raw data.
+
+In all, we aim to keep three versions of the data: (1) `data_raw`, the entirely unprocessed data, (2) `data`, the working version, and (3) `data_unproc`, data that has been partially processed, but which is one step behind `data`.
+This allows us to better illustrate the pre-processing steps, and prevents us from irreversibly modifying our data --- at best, we would have to re-run many pre-processing steps, and at worse, we might overwrite the data on disk.
+
+We look at the first few rows, using `head()`. We then visualise the raw data.
 
@@
 
-We first save a copy of the data, so that we can plot the raw data with the cleaned data plotted over it for comparison.
+We first save a copy of the data, so that we can plot the unprocessed data with the cleaned data plotted over it for comparison.
+Here, `data_unproc`, `data`, and `data_raw` are still the same, since no pre-processing steps have been applied yet.
 
@@
 We then plot the result of filtering, with the excluded point in black, and the points that are retained in green.
+After this stage, `data` is filtered and 'ahead' of `data_raw` and `data_unproc`, which are still the same.
+This pattern will repeat throughout this material.
 
@@
 
+Here, `data_unproc` moves 'ahead' of `data_raw`, and holds the data filtered by a bounding box --- `data` is also moving ahead, and will be filtered on speed.
+
 ```{r}
@@
 
-Next we thin the data to demonstrate thinning by median smoothing.
+Next we thin the data by aggregation to demonstrate thinning after median smoothing.
 Following this, we plot the median smooth and thinning by aggregation.
@@
 # remove columns we don't need
-data <- data[, setdiff(
-  colnames(data),
-  c("tID", "Timestamp", "id", "TIME", "UTCtime")
-),
-with = FALSE
-]
+data <- data[, !c("tID", "Timestamp", "id", "TIME", "UTCtime")]
 
@@
 # now get times 10 mins before and after
-data_res_summary[, `:=`(
-  t_min = t_median - (10 * 60),
-  t_max = t_median + (10 * 60)
+data_res_summary[, c("t_min", "t_max") := list(
+  t_median - (10 * 60),
+  t_median + (10 * 60)
 )]
 
-# make a list of positions 10min before and after
+# manually get the duration of the stops
 wp_data <- mapply(function(l, u, mx, my) {
-  tmp_data <- data_unproc[inrange(time, l, u)]
+
+  # first select all data whose timestamp is between
+  # the upper and lower bounds of the stop (l = lower, u = upper)
+  tmp_data <- data_unproc[inrange(time, l, u), ]
+
+  # calculate the distance between the positions selected above
+  # and the median X and Y coordinates of the stop (centroid)
   tmp_data[, distance := sqrt((mx - x)^2 + (my - y)^2)]
 
-  # keep within 50
+  # keep positions that are within 50m of the centroid
   tmp_data <- tmp_data[distance <= 50, ]
 
-  # get duration
+  # get the duration of the stop as the difference between
+  # the minimum and maximum times of the positions retained above
   return(diff(range(tmp_data$time)))
@@
    data_res_summary$x_median, data_res_summary$y_median,
+
+   # this specifies that a vector, rather than a list, is returned
    SIMPLIFY = TRUE
@@
 First, we filter data where we know the animal (or in this case, the human-carried tag) spent some time at or near a position, as this is the first step to identify residence patches. 
-One way of doing this is to filter out positions with speeds above which the tag (ideally on an animal) is likely to be in transit.
+One way of doing this is by filtering out positions with speeds above which the tag (ideally on an animal) is likely to be in transit.
 Rather than filtering on instantaneous speed estimates, filtering on a median smoothed speed estimate is more reliable.

### 2021-06-11 supplement/05_calibration_data.Rmd
add text explaining get data opts

@@
 
-We get spatial and summary ouput of the residence patch method using the `atl_patch_summary` function using the options `which_data` = "spatial" and `which_data` = "summary.
-We use a buffer radius here of 20 metres for the spatial buffer, despite using a buffer radius of 5 metres earlier, simply because it is easier to visualise in the output figure.
+Having classified slow-moving or stationary behavioural bouts into residence patches, many animal ecologists would most probably wish to know something about the environment at or around these patches --- more accurately, around the point locations classified into patches.
+
+How exactly this is done depends on the relative spatial scales of the residence patches and the resolution of the environmental data layer.
+For instance, a residence patch some 40m -- 50m wide or long may be overlaid on an environmental raster layer with a resolution of 250m.
+In this case, sampling the layer at the centroid of the patch is as good as sampling at all the patch's points -- the mean is unlikely to differ (except at raster pixel boundaries).
+
+On the other hand, a raster with a 10m resolution (e.g.
+Sentinel 1 and 2 data) may be worthwhile to sample at all the locations comprising a residence patch, so as to calculate the mean and variance of environmental conditions.
+
+Furthermore, many (if not all) animals integrate cues from quite a distance (10m -- 100m) when making decisions on when to settle in an area, and when to leave.
+Thus it can also be useful to sample environmental layers not at point locations, but to extract the mean and variance from an _area_, or a buffer, around the animal's point locations.
+
+We have provided a convenient function to get either _(1)_ the points (classified into patches), or _(2)_ a summary ouput of the residence patches (i.e., the median coordinates and their attributes), or finally _(3)_ a spatial buffer around the points from _(1)_.
+This function, `atl_patch_summary` implements these options using the `which_data` argument, where the options are _(1)_ "points", _(2)_ "summary", or _(3)_ "spatial".
+
+Here, we choose option (3), using a spatial buffer of 20m. The distance of the bufffer is passed to the argument `buffer_radius`.
 

### 2021-06-11 supplement/05_calibration_data.Rmd

@@
 data_res_summary <- data_res[, list(
+    nfixes_real = .N,
     x_median = median(x),
@@
 )
+
+# get waypoint summary --- rounding median coordinates to the nearest 100m
+patch_summary_real <- data_res_summary[, list(
+  nfixes_real = nfixes_real,
+  x_median = round(median(x_median), digits = -2),
+  y_median = round(median(y_median), digits = -2)
+),
+by = "tID"
+]
+
+# add real duration
+patch_summary_real[, duration_real := wp_data]
 ```
@@
 
-```{r}
+```{r message=FALSE, warning=FALSE}
 # read griend and hut
-griend <- sf::st_read("data/griend_polygon/griend_polygon.shp")
-hut <- sf::st_read("data/griend_hut.gpkg")
+griend <- sf::st_read("data/griend_polygon/griend_polygon.shp", quiet = TRUE)
+hut <- sf::st_read("data/griend_hut.gpkg", quiet = TRUE)
 ```
@@
 
-We then merge the annoated, known stop data with the calculated patch duration.
-We filter this data to exclude one exceedingly long outlier of about an hour (WP080), which how
+We filter these data to exclude one exceedingly long outlier of about an hour (WP080).
 
 ```{r}
-# get known patch summary
-data_res <- data_unproc[stringi::stri_detect(tID, regex = "(WP)"), ]
-
-# get waypoint summary
-patch_summary_real <- data_res[, list(
-  nfixes_real = .N,
-  x_median = round(median(x), digits = -2),
-  y_median = round(median(y), digits = -2)
-),
-by = "tID"
-]
-
-# add real duration
-patch_summary_real[, duration_real := wp_data]
-
 # round median coordinate for inferred patches
@@
   )]
+```
 
+We add data from the known patches, matching by X and Y median.
+
+```{r }
 # join with respatch summary

### 2021-06-11 supplement/06_bat_data.Rmd

@@
 
-## Exploratory Data Analysis Panels for Main Text Figure 1
+## Exploratory Data Analysis Panels: Main Text Figure 1
 
@@
 Then we pass the filters to `atl_filter_covariates`.
-We apply the filter to each individual's data using an `lapply`.
+We apply the filter to each individual's data using an `lapply` --  this separates the data from each individual into a separate data frame, lessening the chances of inter-individual mix-ups.
 
@@
 # here, reassignment is necessary as rows are being removed
-# the atl_filter_covariates function could have been used here
 data_split <- lapply(data_split, function(dt) {
@@
 ![Bat data with unrealistic speeds removed.
+Notice, compared with the previous figure, that spikes of unrealistic movement in all three tracks have been removed.
 Grey crosses show data that were removed.
@@
 data_residence <- lapply(data_smooth, function(dt) {
-  # do basic recurse
+
+  # do basic recurse -- refer to Bracis et al. (2018) Ecography
   dt_recurse <- getRecursions(
@@
 
-  # get revisit stats
+  # get revisit stats column provided as recurse output
   dt_recurse <- setDT(
@@
 
-  # count long absences from the area
+  # count long absences from the each position
   dt_recurse[, timeSinceLastVisit :=
@@
   ]
-  # get data before the first long absence of 60 mins
+
+  # filter data to exclude revisits after the first long absence of 60 mins
   dt_recurse <- dt_recurse[longAbsenceCounter < 1, ]
 
+  # calculate the residence time as the sum of times inside
+  # before the first 'long absence'
+  # also calculate the First Passage Time and the number of revisits
   dt_recurse <- dt_recurse[, list(
@@
 
-  # prepare and merge existing data with recursion data
+  # prepare to merge existing data with recursion data
   dt[, coordIdx := seq(nrow(dt))]
 
+  # merge the revised recursion analysis data with the tracking data
   dt <- merge(dt,
@@
 
+  # ensure the data are ordered in ascending order of time
   setorderv(dt, "time")
@@
 
+  # return the dataframe
   dt

### 2021-06-16 scripts/01_background.R
styler run

@@ -5,9 +5,9 @@
 #' ---
 #'
 #' # Code to Prepare Example Data
-#' 
-#' 
-#' 
+#'
+#'
+#'
 ## ----install_package----------------------------------------------------------
 # use either devtools or remotes to install
 install.packages("devtools")

### 2021-06-28 supplement/06_bat_data.Rmd
styler run
add sd and add some analyses

@@
 
-## Filter by covariates
+## Filter by system-generated error attributes
 
-No natural bounds suggest themselves, so instead we proceed to filter by covariates, since point outliers are obviously visible.
+No natural bounds suggest themselves, so instead we proceed to filter by system-generated attributes of error, since point outliers are obviously visible.
 
@@
 
-First we calculate the variable `SD`.
+First we calculate the variable `SD`, which for ATLAS systems is calculated as:
+
+$$SD = \sqrt{{VARX} + {VARY} + 2 \times {COVXY}}$$
 
@@
     ifelse(is.na(timeSinceLastVisit), -Inf, timeSinceLastVisit)]
-  dt_recurse[, longAbsenceCounter := cumsum(timeSinceLastVisit > 60),
+  dt_recurse[, longAbsenceCounter := cumsum(timeSinceLastVisit > 5),
     by = .(coordIdx)
@@
 data_residence[, ts := as.POSIXct(time, origin = "1970-01-01")]
+
+# get hour of day to filter for nighttime
+data_residence[, hour := data.table::hour(ts)]
+
+# filter for hour between 8pm and 5am
+data_residence = atl_filter_covariates(
+  data = data_residence,
+  filters = "hour > 20 | hour < 5"
+)
 ```
@@
 
+# get mean residence time per id
+data_residence[, list(mean_residence = mean(resTime),
+                      sd_residence = sd(resTime)), by = "TAG"]
+
+# get mean residence time for all bats pooled
+data_residence[, list(mean_residence = mean(resTime),
+                      sd_residence = sd(resTime))]
+
 # filter for residence time > 5 minutes
 data_residence <- data_residence[resTime > 5, ]
+
+# average positions per bat after removing transit points
+data_residence[, list(.N), by = "TAG"][, list(mean_positions_ = mean(N))]
 
@@
 
+### Duration at foraging sites
+
+We exclude the first and last patch of each day as being roosting related, and examine how much of the total foraging time (time between the remaining first and last patch) was spent at foraging sites.
+It follows that the remainder of the time must have been spent in transit, or otherwise not foraging.
+
+```{r}
+# make patch summary a datatable
+setDT(patch_summary)
+
+# exclude roost
+patch_summary = patch_summary[x_median > 254000, ]
+
+# get mean and sd of duration in patches
+patch_summary[, list(mean_duration = mean(duration / 60), 
+                     sd_duration = sd(duration / 60))]
+
+# assign night id
+patch_summary[, c("hour", "day") := list(
+  data.table::hour(as.POSIXct(time_start, origin = "1970-01-01")),
+  data.table::mday(as.POSIXct(time_start, origin = "1970-01-01")))]
+
+patch_summary[, night := 1 + c(0, cumsum(diff(hour) > 12)), by = "id"]
+
+# get total foraging time
+foraging_proportion = patch_summary[, list(
+  time_total_forage = (max(time_end) - min(time_start)) / 60, 
+  time_forage_site = sum(duration / 60)),
+  by = c("id", "night")
+  ]
+
+# get proporion of foraging that is at a foraging site
+foraging_proportion[, list(
+  mean_foraging_prop = mean(time_forage_site / time_total_forage),
+  sd_foraging_prop = sd(time_forage_site / time_total_forage)
+)]
+
+```
+
 ![A visual examination of plots of the bats' residence patches and linear approximations of paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site over the three nights **(a)**.

### 2021-06-29 scripts/03_smoothing_tracks.R
add legend figure 5

@@ -204,12 +204,12 @@ ggsave(figure_median_smooth,
 
 ## -----------------------------------------------------------------------------
 # choose the 11 point median smooth data
-data_agg <- fread("data/data_smooth.csv")
+data_smooth <- fread("data/data_smooth.csv")
 
 # get list of aggregated data
 list_of_agg <- lapply(c(3, 10, 30, 120), function(z) {
   data_return <- atl_thin_data(
-    data = data_agg,
+    data = data_smooth,
     interval = z,
     method = "aggregate"
   )
@@ -240,30 +240,75 @@ speed_agg_smooth <- rbindlist(speed_agg_smooth)
 data_agg_smooth <- copy(list_of_agg[[3]]) # 30s aggregate
 ### plot figures
 fig_agg_data_smooth <-
-  ggplot(data_agg_smooth) +
+  ggplot() +
   geom_point(
-    data = data_agg,
+    data = data_smooth,
     size = 0.2,
-    aes(x, y),
-    col = pal[3]
+    aes(
+      x, y,
+      colour = "smooth",
+      shape = "smooth"
+    )
   ) +
-  geom_path(aes(x, y),
-    col = pal[4],
+  geom_path(
+    data = data_agg_smooth,
+    aes(
+      x, y,
+      colour = "thin",
+      shape = "thin"
+    ),
     lwd = 0.2
   ) +
-  geom_point(aes(x, y,
-    group = interval
-  ),
-  shape = 0,
-  size = 2,
-  col = pal[4],
-  alpha = 1,
-  show.legend = F
+  geom_point(
+    data = data_agg_smooth,
+    aes(
+      x, y,
+      group = interval,
+      shape = "thin",
+      col = "thin"
+    ),
+    # shape = 0,
+    size = 2,
+    # col = pal[4],
+    alpha = 1
+    # show.legend = F
+  ) +
+  scale_colour_manual(
+    values = c(
+      thin = pal[4],
+      smooth = pal[3]
+    ),
+    labels = c(
+      thin = "Thinned data",
+      smooth = "Smoothed,\nfiltered data"
+    ),
+    name = NULL
+  ) +
+  scale_shape_manual(
+    values = c(
+      thin = 0,
+      smooth = 16
+    ),
+    labels = c(
+      thin = "Thinned data",
+      smooth = "Smoothed,\nfiltered data"
+    ),
+    name = NULL
+  ) +
+  guides(
+    color = guide_legend(
[…]

### 2021-06-29 scripts/03_smoothing_tracks.R
minor edit

@@ -467,7 +467,7 @@ fig_agg_speed <-
     ),
     breaks = c("aggunfil", "aggsmooth"),
     name = NULL
-  )+
+  ) +
   ggthemes::theme_few() +
   theme(
     axis.text.y = element_blank(),

### 2021-07-01 supplement/06_bat_data.Rmd

@@
 # filter for hour between 8pm and 5am
-data_residence = atl_filter_covariates(
+data_residence <- atl_filter_covariates(
   data = data_residence,
@@
 # get mean residence time per id
-data_residence[, list(mean_residence = mean(resTime),
-                      sd_residence = sd(resTime)), by = "TAG"]
+data_residence[, list(
+  mean_residence = mean(resTime),
+  sd_residence = sd(resTime)
+), by = "TAG"]
 
 # get mean residence time for all bats pooled
-data_residence[, list(mean_residence = mean(resTime),
-                      sd_residence = sd(resTime))]
+data_residence[, list(
+  mean_residence = mean(resTime),
+  sd_residence = sd(resTime)
+)]
 
@@
 # exclude roost
-patch_summary = patch_summary[x_median > 254000, ]
+patch_summary <- patch_summary[x_median > 254000, ]
 
 # get mean and sd of duration in patches
-patch_summary[, list(mean_duration = mean(duration / 60), 
-                     sd_duration = sd(duration / 60))]
+patch_summary[, list(
+  mean_duration = mean(duration / 60),
+  sd_duration = sd(duration / 60)
+)]
 
@@
   data.table::hour(as.POSIXct(time_start, origin = "1970-01-01")),
-  data.table::mday(as.POSIXct(time_start, origin = "1970-01-01")))]
+  data.table::mday(as.POSIXct(time_start, origin = "1970-01-01"))
+)]
 
@@
 # get total foraging time
-foraging_proportion = patch_summary[, list(
+foraging_proportion <- patch_summary[, list(
   time_total_forage = (max(time_end) - min(time_start)) / 60,
-  time_forage_site = sum(duration / 60)),
+  time_forage_site = sum(duration / 60)
+),
 by = c("id", "night")
@@
 )]
-
 ```

### 2021-07-01 render_books.sh
bash rm code for supplement

@@ -1,6 +1,9 @@
 #!/bin/bash
 # remove old tex files and pdfs
 
+rm docs/*supplement*.pdf
+rm docs/atlastools*.pdf
+
 # style files
 Rscript --vanilla --slave -e 'styler::style_dir("scripts", filetype = "R", recursive = FALSE)'
 Rscript --vanilla --slave -e 'styler::style_dir("supplement", filetype = "Rmd", recursive = FALSE)'

### 2021-07-01 supplement/05_calibration_data.Rmd

@@
 In all, we aim to keep three versions of the data: (1) `data_raw`, the entirely unprocessed data, (2) `data`, the working version, and (3) `data_unproc`, data that has been partially processed, but which is one step behind `data`.
-This allows us to better illustrate the pre-processing steps, and prevents us from irreversibly modifying our data --- at best, we would have to re-run many pre-processing steps, and at worse, we might overwrite the data on disk.
+This allows us to better illustrate the pre-processing steps, and prevents us from irreversibly modifying our data --- at best, we would have to re-run many pre-processing steps, and at worst, we might overwrite the original data on disk.
 

### 2021-09-09 bash/copy_tex.sh
update script wd

@@ -1,3 +1,3 @@
 #!/bin/bash
 # transfer figures
-cp figures/fig_0*.png atlas-manuscript/figures
+cp ../figures/fig_0*.png ../atlas-manuscript/figures

### 2021-09-09 bash/render_books.sh

@@ -1,6 +1,8 @@
 #!/bin/bash
 # remove old tex files and pdfs
 
+cd ..
+
 rm docs/*supplement*.pdf
 rm docs/atlastools*.pdf
 

### 2021-09-09 scripts/04_residence_patch.R
update figure for res patch

@@ -13,21 +13,91 @@ pal <- RColorBrewer::brewer.pal(4, "Set1")
 
 ## -----------------------------------------------------------------------------
 # read patch data
-data <- fread("data/data_for_res_patch.csv")
+data <- fread("data/data_lt_plot.csv")
+# filter for illustrative plots
+data = data[id == 3 & growth == 0.01 & type == "exploit", ]
 
-# an id is required
-data[, id := "test"]
+data[, stationary := (
+  c(0, diff(x)) == 0 & c(0, diff(y)) == 0
+), by = "interval"]
+
+# get cell r from agent
+data_r = copy(data[interval == 1,])
+data_r = data_r[, c("x_wrap", "y_wrap") := list(
+  round(x_wrap),
+  round(y_wrap)
+)]
+data_r = data_r[, list(
+  cell_r = mean(cell_r)
+), by = c("x_wrap", "y_wrap")]
+
+
+## plot initial figure
+fig_a =
+ggplot(
+  data[interval == 1,], 
+  aes(
+    x_wrap, y_wrap
+  )
+)+
+  geom_tile(
+    data = data_r,
+    aes(
+      x_wrap, y_wrap,
+      fill = cell_r
+    ),
+    alpha = 0.5
+  )+
+  geom_path(
+    aes(
+      group = interval
+    ),
+    col = "grey",
+    size = 0.3
+  )+
+  geom_point(
+    aes(
+      shape = stationary
+    ),
+    alpha = 0.8
+  )+
+  scale_fill_distiller(
+    palette = "YlGn", 
+    direction = 1,
+    breaks = c(0.1, 0.5, 0.99),
+    labels = c("Low", "Med.", "High"),
+    name = "Productivity"
+  )+
+  scale_shape_manual(
+    values = c(3, 16),
+    guide = "none"
+  )+
+  # facet_wrap(~interval)+
+  ggthemes::theme_few() +
+  theme(
+    legend.position = c(0.2, 0.8),
+    legend.key.width = unit(1, "mm"),
+    legend.background = element_blank(),
+    axis.title = element_blank(),
+    axis.text = element_blank(),
+    strip.text = element_blank(),
+    strip.background = element_blank()
+  )+
+  coord_equal()+
+  labs(
+    colour = "Productivity"
+  )
 
 # flip lengthwise
-setnames(data, old = c("x", "y"), new = c("y", "x"))
+# setnames(data, old = c("x", "y"), new = c("y", "x"))
 
-# do recurse
[…]

### 2021-09-09 bash/render_books.sh
do not render supplement as scripts

@@ -14,7 +14,7 @@ Rscript --vanilla --slave -e 'styler::style_dir("supplement", filetype = "Rmd", 
 Rscript --vanilla --slave -e 'bookdown::render_book("supplement/")'
 
 # to render source files
-Rscript --slave -e 'lapply(list.files(pattern = "supplement\\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
+# Rscript --slave -e 'lapply(list.files(pattern = "supplement\\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
 
 mv docs/supplementary_material.pdf docs/ms_atlas_preproc_supplementary_material_`date -I`.pdf
 

### 2021-09-09 supplement/05_calibration_data.Rmd
minor corrections per R1

@@
 ```{r}
-# divide by 1000, convert to integer, then convert to POSIXct
+# divide by 1000, convert to integer
 data[, time := as.integer(
@@
 patch_summary_real[, duration_real := wp_data]
+
+# write to file
+fwrite(patch_summary_real, "data/data_real_watlas_stops.csv")
 ```

### 2021-09-09 supplement/06_bat_data.Rmd
minor corrections per R1

@@
 When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar **(c)**.
-Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](../figures/fig_07.png)
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](../figures/fig_08.png)

### 2021-09-10 scripts/02_filtering_data.R
update figure plot

@@ -46,6 +46,14 @@ data <- fread("data/data_sim.csv")[5000:10000, ]
 ## -----------------------------------------------------------------------------
 data_copy <- fread("data/data_errors.csv")
 
+## remove 10% positions
+data_copy = data_copy[!sample(
+  x = seq(nrow(data_copy)),
+  size = as.integer(0.1 * nrow(data_copy)),
+  ), ]
+
+# save as dataset
+fwrite(data_copy, "data/data_errors_missing.csv")
 
 ## -----------------------------------------------------------------------------
 # define a four colour palette
@@ -96,6 +104,14 @@ data_inside_bbox <- atl_filter_bounds(
 # plot data inside and outside bbox
 fig_filter_bounds <-
   ggplot() +
+  geom_point(
+    data = data_copy[!data_inside_bbox,
+                     on = c("x", "y")
+    ],
+    aes(x, y),
+    col = "grey",
+    shape = 1
+  ) +
   geom_path(
     data = data_copy,
     aes(x, y),
@@ -108,15 +124,6 @@ fig_filter_bounds <-
     col = pal[3],
     alpha = 1, size = 0.2
   ) +
-  geom_point(
-    data = data_copy[!data_inside_bbox,
-      on = c("x", "y")
-    ],
-    aes(x, y),
-    col = "grey",
-    shape = 4,
-    size = 0.2
-  ) +
   geom_path(
     data = data,
     aes(x, y),
@@ -128,7 +135,7 @@ fig_filter_bounds <-
     col = "grey",
     lty = 2
   ) +
-  ggthemes::theme_few() +
+  theme_void() +
   theme(
     axis.text = element_blank(),
     axis.title = element_blank()
@@ -158,9 +165,9 @@ sapply(data_copy[, c("in_speed", "angle")], function(z) {
 
 
 ## -----------------------------------------------------------------------------
-# filter the copy by the 95th percentile
+# filter the copy by the 90th percentile
 data_filtered <- atl_filter_covariates(data_copy,
-  filters = c("(in_speed < 0.025 & out_speed < 0.025) | angle < 35")
+  filters = c("(in_speed < 0.05 & out_speed < 0.05) | angle < 35")
 )
 
 
@@ -174,13 +181,13 @@ fig_hist_speed <-
       fill = ifelse(in_speed > 0.05, "outlier", "good")
     ),
     position = position_identity(),
-    bins = 60
[…]

### 2021-09-10 supplement/05_calibration_data.Rmd
add fig sd nbs

@@
 data[, angle := atl_turning_angle(data = data)]
+```
+
+Compare number of receivers and SD and speed.
+
+```{r}
+data_nbs = data[, c("NBS", "SD", "speed_in", "VARX", "VARY")]
+data_nbs = melt(
+  data_nbs,
+  id.vars = c("NBS")
+)
+```
+
+```{r}
+# plot heatmaps
+fig_sd_nbs = 
+  ggplot(data_nbs)+
+  geom_jitter(
+    aes(factor(NBS), SD, col = speed_in),
+    size = 0.2,
+    alpha = 0.4
+  )+
+  geom_hline(
+    yintercept = 100,
+    lty = 2, col = "grey"
+  )+
+  geom_vline(
+    xintercept = 1.5,
+    lty = 2, col = "grey"
+  )+
+  scale_colour_viridis_c(
+    option = "G", direction = -1,
+    limits = c(0, 20),
+    na.value = "red"
+  )+
+  coord_cartesian(
+    expand = F,
+    xlim = c(0.5, 6),
+    ylim = c(1, 1e3)
+  )+
+  scale_y_log10()+
+  theme_classic()+
+  theme(
+    # panel.border = element_rect(
+    #   colour = "black",
+    #   fill = "NA"
+    # ),
+    legend.key.width = unit(2, "mm"),
+    legend.key.height = unit(5, "mm"),
+    legend.position = c(0.8, 0.85),
+    legend.background = element_blank()
+  )+
+  # facet_wrap(~variable)+
+  labs(
+    colour = "Speed (m/s)",
+    x = "# WATLAS receivers",
+    y = "Positioning error (SD)"
+  )
+
+ggsave(
+  fig_sd_nbs,
+  filename = "supplement/figures/fig_sd_nbs.png",
+  height = 5,
+  width = 3
+)
 ```

### 2021-09-13 supplement/06_bat_data.Rmd
edits for fig 1

@@
     ),
-    colour = "grey",
     size = 0.1,
@@
   scale_fill_viridis_c(
-    option = "C",
-    direction = 1,
+    option = "A",
+    direction = -1,
     trans = "log10"
   ) +
-  ggthemes::theme_few() +
-  theme(
-    axis.text = element_blank(),
-    axis.title = element_blank()
-  ) +
+  theme_void() +
   coord_sf(crs = 2039)
@@
     # fill = "grey",
-    col = "black",
+    # col = "black",
     show.legend = F
   ) +
-  geom_vline(
-    xintercept = 1,
-    col = "red"
-  ) +
   scale_x_log10(
-    labels = function(x) {
-      sprintf("%ix", x)
-    },
-    breaks = c(1, 3, 10, 100, 1000)
+    breaks = c(1, 1000),
+    labels = c("low", "high")
   ) +
@@
   scale_fill_viridis_d(
-    option = "H",
-    begin = 0.1,
+    option = "B",
+    begin = 0.4,
     end = 0.8
   ) +
-  ggthemes::theme_few() +
+  theme_void() +
   theme(
@@
     # fill = "grey",
-    col = "black",
+    # col = "black",
     show.legend = F
@@
   scale_fill_viridis_d(
-    option = "H",
-    begin = 0.1,
+    option = "B",
+    begin = 0.4,
     end = 0.8
@@
   scale_x_log10(
-    labels = function(x) {
-      sprintf("%2.fm", x)
-    }
+    breaks = c(1, 1000),
+    labels = c("low", "high")
   ) +
@@
   ) +
-  ggthemes::theme_few() +
+  theme_void()+
   theme(
@@
   ) +
-  labs(x = "Variance X")
+  labs(x = "Positioning error")
 
@@
   ggspatial::annotation_scale(location = "br") +
-  ggthemes::theme_few() +
+  theme_void()+
   theme(
@@
 )
+
+# for fig 1
+ggsave(
+  fig_bat_smooth + theme_void(),
+  filename = "supplement/figures/fig_bat_preproc_f1.png",
+  dpi = 300,
+  width = 4, height = 6
+)
 ```

### 2021-09-14 scripts/02_filtering_data.R
remove ggthemes

@@ -6,8 +6,6 @@ library(atlastools)
 # to plot
 library(ggplot2)
 library(patchwork)
-library(ggthemes)
-
 
 ## -----------------------------------------------------------------------------
 # source helper functions

### 2021-09-14 scripts/03_smoothing_tracks.R
update figure 4

@@ -53,20 +53,124 @@ data_plot <- mapply(function(df, offset) {
 SIMPLIFY = F
 )
 
-data_plot <- rbindlist(data_plot)
+pal_smooth = c(colorspace::sequential_hcl(
+  3,
+  l = 40, palette = "PuBu",
+  rev = T
+), "darkred")
 
-# add offset to data errors
-data_errors <- lapply(
-  c(0, seq(0.4, 1.25, length.out = 4)),
-  function(offset) {
-    df <- copy(data_errors)
-    df[, x := x + offset]
-    df[, offset := offset]
-  }
+plot_smooth = Map(list_of_smooths, 
+    pal_smooth,
+    f = function(df, col) {
+  ggplot(df) +
+    geom_path(
+      data = data_errors,
+      aes(x, y),
+      col = "grey70",
+      size = 0.2
+    )+
+    geom_point(
+      data = data_errors,
+      aes(x, y),
+      col = "grey60",
+      size = 0.2,
+      shape = 16
+    )+
+    geom_path(
+      data = df,
+      aes(x, y),
+      show.legend = F,
+      # lwd = 0.35,
+      col = col
+    )+
+    coord_equal(
+      expand = T,
+      ylim = c(0.6, 0.78),
+      # xlim = c(NA, 2.3),
+      ratio = 2
+    ) +
+    theme_void(base_size = 8)+
+    theme(
+      plot.background = element_rect(
+        fill = "white", colour = NA
+      ),
+      plot.title = ggtext::element_markdown()
+    )+
+    labs(
+      title = sprintf("Median smooth; *K* = %i", unique(df$window_size))
+    )
+})
+
+# plot the filtered data to show the errors
+plot_errors =
+  ggplot()+
+  geom_path(
+    data = data_errors,
+    aes(x, y),
+    col = "grey90",
+    size = 0.1
+  )+
+  geom_point(
+    data = data_errors,
+    aes(x, y),
+    col = pal[3],
+    alpha = 1,
+    size = 0.2
+  ) +
+  geom_path(
+    data = data,
+    aes(x, y),
+    col = "grey20"
+  )+
+  coord_equal(
+    expand = T,
+    ylim = c(0.6, 0.78),
+    # xlim = c(NA, 2.3),
+    ratio = 2
+  ) +
+    theme_void(base_size = 8)+
+    theme(
+      plot.background = element_rect(
+        fill = "white", colour = NA
+      ),
+      plot.title = ggtext::element_markdown()
+    )+
[…]

### 2021-09-14 scripts/02_filtering_data.R
update fig 3

@@ -138,7 +138,7 @@ fig_filter_bounds <-
     axis.text = element_blank(),
     axis.title = element_blank()
   ) +
-  theme(plot.background = element_rect(fill = NA)) +
+  theme(plot.background = element_rect(fill = "white", colour = NA))+
   coord_equal(expand = T)
 
 
@@ -313,7 +313,7 @@ ggplot() +
     legend.position = "none"
   ) +
   coord_equal() +
-  theme(plot.background = element_rect(fill = NA))
+  theme(plot.background = element_rect(fill = "white", colour = NA))
 
 ## ----echo=FALSE---------------------------------------------------------------
 # wrap plot

### 2021-09-14 scripts/03_smoothing_tracks.R
update fig 5

@@ -83,7 +83,7 @@ plot_smooth = Map(list_of_smooths,
       # lwd = 0.35,
       col = col
     )+
-    coord_equal(
+    coord_cartesian(
       expand = T,
       ylim = c(0.6, 0.78),
       # xlim = c(NA, 2.3),
@@ -171,74 +171,8 @@ ggsave(
   units = "mm"
 )
 
-
-
 ## -----------------------------------------------------------------------------
 pal2 <- RColorBrewer::brewer.pal(4, "RdPu")[c(2, 4)]
-
-fig_hist_smooth <-
-  ggplot() +
-  geom_histogram(
-    data = data_errors[offset == 0, ],
-    aes(
-      x = in_speed,
-      fill = "errors"
-    ),
-    bins = 80,
-    alpha = 1
-  ) +
-  geom_histogram(
-    data = data_plot[window_size %in% c(5, 21)],
-    aes(
-      x = speed_in,
-      fill = as.factor(window_size),
-      group = window_size
-    ),
-    bins = 80
-  ) +
-  geom_histogram(
-    data = data,
-    aes(
-      x = speed_in,
-      fill = "real"
-    ),
-    # col = pal[1],
-    alpha = 0.8,
-    bins = 80
-  ) +
-  scale_x_log10() +
-  scale_fill_manual(
-    values = c(
-      "real" = alpha("grey20", .8),
-      "errors" = "grey",
-      "5" = pal2[1],
-      "21" = pal2[2]
-    ),
-    labels = c(
-      "real" = "real track",
-      "errors" = "track w/\nerrors",
-      "5" = "K = 5",
-      "21" = "K = 21"
-    )
-  ) +
-  coord_cartesian(xlim = c(1e-5, .05)) +
-  theme_test(base_family = "sans") +
-  theme(
-    legend.position = c(0.2, 0.65),
-    axis.text = element_blank(),
-    # axis.ticks.y = element_blank(),
-    axis.title.y = element_blank(),
-    plot.title = element_text(
-      face = "bold"
-    ),
-    plot.background = element_blank()
-  ) +
-  labs(
-    x = "Speed [log10 scale]",
-    fill = "Smoothing",
-    title = "(f)"
-  )
-
 
 ## -----------------------------------------------------------------------------
 # choose the 11 point median smooth data
@@ -340,13 +274,15 @@ fig_agg_data_smooth <-
       )
     )
   ) +
-  ggthemes::theme_few() +
+  theme_void()+
   theme(
[…]

### 2021-09-14 scripts/04_residence_patch.R
update fig 6

@@ -73,8 +73,11 @@ ggplot(
     guide = "none"
   )+
   # facet_wrap(~interval)+
-  ggthemes::theme_few() +
+  theme_void()+
   theme(
+    plot.background = element_rect(
+      fill = "white", colour = NA
+    ),
     legend.position = c(0.2, 0.8),
     legend.key.width = unit(1, "mm"),
     legend.background = element_blank(),
@@ -229,8 +232,11 @@ fig_res_patches <-
     # xlim = c(-3, NA)
   ) +
   facet_wrap(~interval)+
-  ggthemes::theme_few() +
+  theme_void()+
   theme(
+    plot.background = element_rect(
+      fill = "white", colour = NA
+    ),
     axis.title = element_blank(),
     axis.text = element_blank(),
     strip.text = element_blank(),

### 2021-09-29 supplement/05_calibration_data.Rmd
style code

--- 2021-09-09 supplement/05_calibration_data.Rmd
@@
 data[, angle := atl_turning_angle(data = data)]
+```
+
+Compare number of receivers and SD and speed.
+
+```{r, eval=FALSE, echo=FALSE}
+data_nbs <- data[, c("NBS", "SD", "speed_in", "VARX", "VARY")]
+data_nbs <- melt(
+  data_nbs,
+  id.vars = c("NBS")
+)
+```
+
+```{r, eval=FALSE, echo=FALSE}
+# plot heatmaps
+fig_sd_nbs <-
+  ggplot(data_nbs) +
+  geom_jitter(
+    aes(factor(NBS), SD, col = speed_in),
+    size = 0.2,
+    alpha = 0.4
+  ) +
+  geom_hline(
+    yintercept = 100,
+    lty = 2, col = "grey"
+  ) +
+  geom_vline(
+    xintercept = 1.5,
+    lty = 2, col = "grey"
+  ) +
+  scale_colour_viridis_c(
+    option = "G", direction = -1,
+    limits = c(0, 20),
+    na.value = "red"
+  ) +
+  coord_cartesian(
+    expand = F,
+    xlim = c(0.5, 6),
+    ylim = c(1, 1e3)
+  ) +
+  scale_y_log10() +
+  theme_classic() +
+  theme(
+    # panel.border = element_rect(
+    #   colour = "black",
+    #   fill = "NA"
+    # ),
+    legend.key.width = unit(2, "mm"),
+    legend.key.height = unit(5, "mm"),
+    legend.position = c(0.8, 0.85),
+    legend.background = element_blank()
+  ) +
+  # facet_wrap(~variable)+
+  labs(
+    colour = "Speed (m/s)",
+    x = "# WATLAS receivers",
+    y = "Positioning error (SD)"
+  )
+
+ggsave(
+  fig_sd_nbs,
+  filename = "supplement/figures/fig_sd_nbs.png",
+  height = 5,
+  width = 3
+)
 ```

### 2021-09-29 supplement/06_bat_data.Rmd
style code

@@
   ) +
-  theme_void()+
+  theme_void() +
   theme(
@@
   ggspatial::annotation_scale(location = "br") +
-  theme_void()+
+  theme_void() +
   theme(

### 2021-09-29 scripts/02_filtering_data.R

@@ -45,10 +45,10 @@ data <- fread("data/data_sim.csv")[5000:10000, ]
 data_copy <- fread("data/data_errors.csv")
 
 ## remove 10% positions
-data_copy = data_copy[!sample(
+data_copy <- data_copy[!sample(
   x = seq(nrow(data_copy)),
   size = as.integer(0.1 * nrow(data_copy)),
-  ), ]
+), ]
 
 # save as dataset
 fwrite(data_copy, "data/data_errors_missing.csv")
@@ -104,7 +104,7 @@ fig_filter_bounds <-
   ggplot() +
   geom_point(
     data = data_copy[!data_inside_bbox,
-                     on = c("x", "y")
+      on = c("x", "y")
     ],
     aes(x, y),
     col = "grey",
@@ -138,7 +138,7 @@ fig_filter_bounds <-
     axis.text = element_blank(),
     axis.title = element_blank()
   ) +
-  theme(plot.background = element_rect(fill = "white", colour = NA))+
+  theme(plot.background = element_rect(fill = "white", colour = NA)) +
   coord_equal(expand = T)
 
 
@@ -244,46 +244,48 @@ setorder(data_filtered, time)
 ## ----echo=FALSE---------------------------------------------------------------
 # data plot
 fig_outlier_remove <-
-ggplot() +
-  
+  ggplot() +
   geom_path(
     data = data_copy[!reflection, on = c("x", "y")],
     aes(x, y, col = "outlier_path"),
     lwd = 0.2
-  )+
+  ) +
   geom_path(
     data = data_filtered[!reflection, on = c("x", "y")],
-    aes(x, y,col = "good_path"),
+    aes(x, y, col = "good_path"),
     lwd = 0.2
-  )+
+  ) +
   geom_point(
-    data = data_copy[!reflection, 
-                     on = c("x", "y")][
-                       !data_filtered, on = c("x", "y")],
+    data = data_copy[!reflection,
+      on = c("x", "y")
+    ][
+      !data_filtered,
+      on = c("x", "y")
+    ],
     aes(x, y, col = "outlier"),
     shape = 4, stroke = 1
-  )+
+  ) +
   geom_point(
     data = reflection,
-    aes(x, y,col = "reflection"),
+    aes(x, y, col = "reflection"),
     # size = 0.5,
     alpha = 0.6,
     # stroke = 1,
     shape = 1
-  )+
+  ) +
   geom_point(
[…]

### 2021-09-29 scripts/03_smoothing_tracks.R

@@ -53,63 +53,64 @@ data_plot <- mapply(function(df, offset) {
 SIMPLIFY = F
 )
 
-pal_smooth = c(colorspace::sequential_hcl(
+pal_smooth <- c(colorspace::sequential_hcl(
   3,
   l = 40, palette = "PuBu",
   rev = T
 ), "darkred")
 
-plot_smooth = Map(list_of_smooths, 
-    pal_smooth,
-    f = function(df, col) {
-  ggplot(df) +
-    geom_path(
-      data = data_errors,
-      aes(x, y),
-      col = "grey70",
-      size = 0.2
-    )+
-    geom_point(
-      data = data_errors,
-      aes(x, y),
-      col = "grey60",
-      size = 0.2,
-      shape = 16
-    )+
-    geom_path(
-      data = df,
-      aes(x, y),
-      show.legend = F,
-      # lwd = 0.35,
-      col = col
-    )+
-    coord_cartesian(
-      expand = T,
-      ylim = c(0.6, 0.78),
-      # xlim = c(NA, 2.3),
-      ratio = 2
-    ) +
-    theme_void(base_size = 8)+
-    theme(
-      plot.background = element_rect(
-        fill = "white", colour = NA
-      ),
-      plot.title = ggtext::element_markdown()
-    )+
-    labs(
-      title = sprintf("Median smooth; *K* = %i", unique(df$window_size))
-    )
-})
+plot_smooth <- Map(list_of_smooths,
+  pal_smooth,
+  f = function(df, col) {
+    ggplot(df) +
+      geom_path(
+        data = data_errors,
+        aes(x, y),
+        col = "grey70",
+        size = 0.2
+      ) +
+      geom_point(
+        data = data_errors,
+        aes(x, y),
+        col = "grey60",
+        size = 0.2,
+        shape = 16
+      ) +
+      geom_path(
+        data = df,
+        aes(x, y),
+        show.legend = F,
+        # lwd = 0.35,
+        col = col
+      ) +
+      coord_cartesian(
+        expand = T,
+        ylim = c(0.6, 0.78),
+        # xlim = c(NA, 2.3),
+        ratio = 2
+      ) +
+      theme_void(base_size = 8) +
+      theme(
+        plot.background = element_rect(
+          fill = "white", colour = NA
+        ),
+        plot.title = ggtext::element_markdown()
+      ) +
+      labs(
[…]

### 2021-09-29 scripts/04_residence_patch.R

@@ -15,31 +15,31 @@ pal <- RColorBrewer::brewer.pal(4, "Set1")
 # read patch data
 data <- fread("data/data_lt_plot.csv")
 # filter for illustrative plots
-data = data[id == 3 & growth == 0.01 & type == "exploit", ]
+data <- data[id == 3 & growth == 0.01 & type == "exploit", ]
 
 data[, stationary := (
   c(0, diff(x)) == 0 & c(0, diff(y)) == 0
 ), by = "interval"]
 
 # get cell r from agent
-data_r = copy(data[interval == 1,])
-data_r = data_r[, c("x_wrap", "y_wrap") := list(
+data_r <- copy(data[interval == 1, ])
+data_r <- data_r[, c("x_wrap", "y_wrap") := list(
   round(x_wrap),
   round(y_wrap)
 )]
-data_r = data_r[, list(
+data_r <- data_r[, list(
   cell_r = mean(cell_r)
 ), by = c("x_wrap", "y_wrap")]
 
 
 ## plot initial figure
-fig_a =
-ggplot(
-  data[interval == 1,], 
-  aes(
-    x_wrap, y_wrap
-  )
-)+
+fig_a <-
+  ggplot(
+    data[interval == 1, ],
+    aes(
+      x_wrap, y_wrap
+    )
+  ) +
   geom_tile(
     data = data_r,
     aes(
@@ -47,33 +47,33 @@ ggplot(
       fill = cell_r
     ),
     alpha = 0.5
-  )+
+  ) +
   geom_path(
     aes(
       group = interval
     ),
     col = "grey",
     size = 0.3
-  )+
+  ) +
   geom_point(
     aes(
       shape = stationary
     ),
     alpha = 0.8
-  )+
+  ) +
   scale_fill_distiller(
-    palette = "YlGn", 
+    palette = "YlGn",
     direction = 1,
     breaks = c(0.1, 0.5, 0.99),
     labels = c("Low", "Med.", "High"),
     name = "Productivity"
-  )+
+  ) +
   scale_shape_manual(
     values = c(3, 16),
     guide = "none"
-  )+
+  ) +
   # facet_wrap(~interval)+
-  theme_void()+
+  theme_void() +
   theme(
     plot.background = element_rect(
       fill = "white", colour = NA
@@ -85,8 +85,8 @@ ggplot(
     axis.text = element_blank(),
     strip.text = element_blank(),
     strip.background = element_blank()
-  )+
-  coord_equal()+
+  ) +
+  coord_equal() +
   labs(
[…]

### 2021-10-03 bash/render_books.sh
prep for rendering supplement as bookdown

@@ -13,6 +13,9 @@ Rscript --vanilla --slave -e 'styler::style_dir("supplement", filetype = "Rmd", 
 # render supplement
 Rscript --vanilla --slave -e 'bookdown::render_book("supplement/")'
 
+# render worked out examples for Github Pages into docs
+Rscript --vanilla --slave -e 'bookdown::render_book("supplement/")'
+
 # to render source files
 # Rscript --slave -e 'lapply(list.files(pattern = "supplement\\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'
 

### 2021-10-04 scripts/04_residence_patch.R
add new fig 6

@@ -151,8 +151,8 @@ patches <- Map(
   lim_spat_indep = 5,
   lim_time_indep = 10,
   min_fixes = 5,
-  summary_variables = "cell_r",
-  summary_functions = c("mean")
+  summary_variables = c("cell_r"),
+  summary_functions = c("mean", "sd")
 )
 
 # get spatial representation
@@ -216,14 +216,14 @@ fig_res_patches <-
     aes(fill = factor(patch)),
     alpha = 0.4,
     lwd = 0.2,
-    colour = "transparent",
+    colour = "grey",
     show.legend = FALSE
   ) +
   scale_colour_manual(
     values = c("black", "black")
   ) +
-  scale_fill_brewer(
-    palette = "Set2"
+  scale_fill_viridis_d(
+    option = "D"
   ) +
   scale_shape_manual(
     values = c(3, 16)

### 2021-10-04 bash/render_books.sh
add gitbook option

@@ -14,7 +14,7 @@ Rscript --vanilla --slave -e 'styler::style_dir("supplement", filetype = "Rmd", 
 Rscript --vanilla --slave -e 'bookdown::render_book("supplement/")'
 
 # render worked out examples for Github Pages into docs
-Rscript --vanilla --slave -e 'bookdown::render_book("supplement/")'
+Rscript --vanilla --slave -e 'bookdown::render_book("supplement/", output_format = bookdown::gitbook())'
 
 # to render source files
 # Rscript --slave -e 'lapply(list.files(pattern = "supplement\\/\\d{2}\\w+.Rmd"), function(x) knitr::purl(x, output = sprintf("R/%s", gsub(".{4}$", ".R", x))))'

### 2021-10-04 supplement/06_bat_data.Rmd
use new pal

@@
 # prepare a palette
-pal <- RColorBrewer::brewer.pal(4, "Set1")
+pal <- pals::kovesi.rainbow_bgyr_35_85_c73(3)
 ```
@@
 # installation using remotes
-remotes::install_github("pratikunterwegs/atlastools")
+if (!require(remotes)) {
+  remotes::install_github("pratikunterwegs/atlastools", upgrade = FALSE)
+}
 ```

### 2021-10-04 supplement/06_bat_data.Rmd
new pal used

@@
   ) +
-  scale_colour_brewer(palette = "Dark2") +
+  scale_colour_manual(
+    values = pal
+  ) +
   ggspatial::annotation_scale(location = "br") +
@@
   ) +
-  scale_colour_brewer(palette = "Dark2") +
+  scale_colour_manual(
+    values = pal
+  ) +
   ggspatial::annotation_scale(location = "br") +
-  ggthemes::theme_few() +
+  theme_test() +
   theme(
@@
   ) +
-  scale_colour_brewer(palette = "Dark2") +
+  scale_colour_manual(values = pal) +
   ggspatial::annotation_scale(location = "br") +
-  ggthemes::theme_few() +
+  theme_test() +
   theme(

### 2021-10-04 supplement/06_bat_data.Rmd
select movements away from roost

@@
 data_residence[, hour := data.table::hour(ts)]
+```
+
+### Movements away from the roost
+
+To focus on night-time bat foraging around fruit trees, we shall filter data _both_ on the timestamps, to select night-time positions, and on the locations, to select positions > 1 km away from the roost-cave at Har Gershom (see main text Fig.
+8).
+
+Combining these two filters allows us to exclude bat positions at the roost-cave that may be due to individual-differences in bats' departure or return times to and from their foraging areas.
+
+```{r}
+# read in roosts and select the Har Gershom roost
+roosts = fread("data/Roosts.csv")
+setnames(roosts, "Species", "roost_name")
+roosts = roosts[roost_name == "Har Gershom"]
+
+# define a simple distance function that is vectorised
+get_distance_adhoc = function(x1, y1, x2, y2) {
+  sqrt(((x2 - x1) ^2) + ((y2 - y1) ^2) )
+}
+
+# calculate distance to roost cave at Har Gershom
+data_residence[, distance_roost := get_distance_adhoc(
+  x1 = roosts$X, 
+  x2 = data_residence$X, 
+  y1 = roosts$Y, 
+  y2 = data_residence$Y
+)
+]
+
+# plot for sanity check --- this plot is not shown
+fig_roost_exclude =
+  ggplot(data_residence)+
+  geom_point(
+    aes(X, Y, col = distance_roost)
+  )+
+  geom_point(
+    data = roosts,
+    aes(X, Y),
+    shape = 21, size = 4,
+    fill = "blue",
+    alpha = 0.5
+  )+
+  scale_colour_viridis_c(
+    option = "B", direction = -1,
+    trans = "log10",
+    breaks = c(1000, 2500, 5000),
+    labels = function(x) {
+      scales::comma(
+        x,
+        scale = 0.001, suffix = "km"
+      )
+    },
+    limits = c(1000, NA),
+    na.value = "lightblue"
+  )+
+  ggspatial::annotation_scale(location = "br") +
+  theme_test() +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank(),
+    legend.position = "top",
+    legend.key.height = unit(2, "mm"),
+    legend.title = element_text(vjust = 1)
+  ) +
+  coord_sf(
+    crs = 2039
+  )+
+  labs(
+    colour = "distance (km)"
+  )
+```
 

### 2021-10-04 supplement/06_bat_data.Rmd
split by night before res patch

@@
 
+```{r}
 # filter for hour between 8pm and 5am
@@
   data = data_residence,
-  filters = "hour > 20 | hour < 5"
+  filters = c(
+    "hour > 20 | hour < 5",
+    "distance_roost > 1000")
 )
+```
+
+### Split data by night-id
+
+We assign a night-id to each position, i.e., the night-time spanning two calendar days.
+We then filter for data with a residence time > 5 minutes, as we expect that a bat stopped at a location for more than 5 minutes is likely to be foraging.
+
+```{r}
+# split data into separate nights
+data_residence[, night := 1 + c(0, cumsum(diff(hour) > 12)), by = "TAG"]
+
+# filter for residence time > 5 minutes
+data_residence <- data_residence[resTime > 5, ]
 ```

### 2021-10-04 supplement/06_bat_data.Rmd
get mean and sd night and distance roost

@@
 # split the data
-data_residence <- split(data_residence, data_residence$TAG)
+data_residence <- split(data_residence, by = c("TAG", "night"))
 ```
@@
   buffer_radius = 25,
-  min_fixes = 3
+  min_fixes = 3,
+  summary_variable = c("night", "distance_roost"),
+  summary_functions = c("mean", "sd")
 )

### 2021-10-04 supplement/06_bat_data.Rmd
minor edits

@@
 `time`, and `id`, which we assign using the `data.table` syntax. The `time` column is already present, but the other columns need to be renamed to lower case.
-Then we subset the data to only work with positions where the individual had a residence time of more than 5 minutes.
 
@@
 )]
-
-# filter for residence time > 5 minutes
-data_residence <- data_residence[resTime > 5, ]
 
@@
 
-# exclude roost
-patch_summary <- patch_summary[x_median > 254000, ]
-
 # get mean and sd of duration in patches
@@
 
-patch_summary[, night := 1 + c(0, cumsum(diff(hour) > 12)), by = "id"]
-
 # get total foraging time
@@
 ),
-by = c("id", "night")
+by = c("id", "night_mean")
 ]

### 2021-10-04 supplement/06_bat_data.Rmd
update fig code

@@
   ggspatial::annotation_scale(location = "br") +
-  theme_void() +
+  theme_test() +
   theme(

### 2021-10-05 supplement/06_bat_data.Rmd
check for atlastools
update supplement

@@
 # installation using remotes
-if (!require(remotes)) {
+if (!require(atlastools)) {
   remotes::install_github("pratikunterwegs/atlastools", upgrade = FALSE)
@@
 ]
+```
 
+Users should plot the data to examine the effect of applying filters --- this code is shown, but the figure is hidden for brevity.
+
+```{r}
 # plot for sanity check --- this plot is not shown
@@
 
+We now filter the data to exclude both day-time data, as well as data that is < 1 km from the roost.
+
 ```{r}
-# filter for hour between 8pm and 5am
+# filter for hour between 8pm and 5am and distance > 1km
 data_residence <- atl_filter_covariates(

### 2021-10-05 supplement/index.Rmd

--- 
knit: "bookdown::render_book"
title: "Supplementary Material for _A Guide to Pre-processing High-Frequency Animal Tracking Data_"
author:
    - Pratik R. Gupte
    - Christine E. Beardsworth
    - Orr Spiegel
    - Emmanuel Lourie
    - Sivan Toledo
    - Ran Nathan
    - Allert I. Bijleveld
date: "`r Sys.Date()`"
output: 
    bookdown::pdf_document2
    bookdown::gitbook
documentclass: scrreprt
toc-depth: 1
secnumdepth: 2
geometry: "left=4cm, right=3cm, top=2.5cm, bottom=2.5cm"
bibliography: [references.bib]
link-citations: yes
github-repo: pratikunterwegs/atlas-best-practices
description: "Supplementary Material for _A Guide to Pre-processing High-Frequency Animal Tracking Data_"
---

### 2021-10-05 supplement/06_bat_data.Rmd
minor updates to text

@@
 
+## Main text Figure 8
+
+See Fig. 8 in the main text, made with QGIS.
+
 ![A visual examination of plots of the bats' residence patches and linear approximations of paths between them showed that though all three bats roosted at the same site, they used distinct areas of the study site over the three nights **(a)**.
@@
 When bats did occur close together, their residence patches barely overlapped, and their paths to and from the broad area of co-occurrence were not similar **(c)**.
-Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](../figures/fig_08.png)
+Constructing residence patches for multiple individuals over multiple activity periods suggests interesting dynamics of within- and between-individual overlap **(b, c)**.](figures/fig_08.png)

### 2021-10-06 supplement/06_bat_data.Rmd
update fig bad speed

--- 2021-10-05 supplement/06_bat_data.Rmd
@@
   ggspatial::annotation_scale(location = "br") +
-  theme_test() +
+  theme_void() +
   theme(

### 2021-10-07 supplement/index.Rmd
add index

@@
 ---
+
+# Introduction {-}
+
+This is a rendering of the Supplementary Material for _A Guide to Pre-processing High-Frequency Animal Tracking Data_, and contains two fully worked out examples that could provide a useful template for structuring pre-processing pipelines for your own high-throughput tracking data.
+
+The code and data used here are available on Github at [github.com/pratikunterwegs/atlas-best-practices](https://github.com/pratikunterwegs/atlas-best-practices), and on Zendo at [zenodo/atlas-best-practices](https://doi.org/10.5281/zenodo.4287462).

### 2021-10-07 supplement/06_bat_data.Rmd
correct figure
styler pass

--- 2021-10-06 supplement/06_bat_data.Rmd
@@
 # read in roosts and select the Har Gershom roost
-roosts = fread("data/Roosts.csv")
+roosts <- fread("data/Roosts.csv")
 setnames(roosts, "Species", "roost_name")
-roosts = roosts[roost_name == "Har Gershom"]
+roosts <- roosts[roost_name == "Har Gershom"]
 
 # define a simple distance function that is vectorised
-get_distance_adhoc = function(x1, y1, x2, y2) {
-  sqrt(((x2 - x1) ^2) + ((y2 - y1) ^2) )
+get_distance_adhoc <- function(x1, y1, x2, y2) {
+  sqrt(((x2 - x1)^2) + ((y2 - y1)^2))
 }
@@
   y2 = data_residence$Y
-)
-]
+)]
 ```
@@
 # plot for sanity check --- this plot is not shown
-fig_roost_exclude =
-  ggplot(data_residence)+
+fig_roost_exclude <-
+  ggplot(data_residence) +
   geom_point(
     aes(X, Y, col = distance_roost)
-  )+
+  ) +
   geom_point(
@@
     alpha = 0.5
-  )+
+  ) +
   scale_colour_viridis_c(
@@
     na.value = "lightblue"
-  )+
+  ) +
   ggspatial::annotation_scale(location = "br") +
@@
     crs = 2039
-  )+
+  ) +
   labs(
@@
     "hour > 20 | hour < 5",
-    "distance_roost > 1000")
+    "distance_roost > 1000"
+  )
 )
